---
title: Parameterized Reporting with Quarto
subtitle: "Registered `r params$pet_type` in Switzerland"
author: "Jadey Ryan"
date: "2024-03-27"
date-modified: last-modified
format: 
  aps-html:
    code-tools: true
categories: [quarto, params]
params:
  pet_type: "dogs"
  fave_breed: "Australian Shepherd"
other-links:
    - text: Download data
      href: data/pets.RDS
      icon: download
---

---------------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| message: false

# Attach packages
library(tidyverse)

# Load data
pets <- read_rds("data/pets.RDS")
```

## Breeds of `r params$pet_type` in Switzerland

```{r}
#| label: breeds-tbl

breeds <- pets |>
  filter(pet_type == params$pet_type)

# Top 5 per date
breeds_top_5 <- breeds |>
  slice_max(order_by = count, n = 5, by = date)

# Top 5 overall (average across all dates)
breeds_avg <- breeds_top_5 |>
  summarize(
    "Average Number" = format(
      mean(count),
      big.mark = ",",
      digits = 1,
      justify = "right",
      scientific = FALSE
    ),
    .by = "breed"
  ) |>
  rename("Breed" = breed) |>
  head(5)
```

The breed with the largest average population in Switzerland from 2016 - 2023 was
the **`r breeds_avg$Breed[[1]]`**.

```{r}
#| label: breeds-plot

breeds_top_5 |>
  ggplot(
    aes(x = date, y = count, color = breed)
  ) +
  geom_line(linewidth = 1) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  scale_y_continuous(
    labels = \(x) format(x, big.mark = ",", scientific = FALSE)
  ) +
  labs(
    title = "Evolution by breed",
    subtitle = str_wrap(
      paste(
        "Number of registered and living",
        params$pet_type,
        "by breed. The evolution of the largest populations is presented."
      ),
      80
    ),
    color = NULL,
    caption = "Source: Identitas AG Animal Statistics"
  ) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    plot.caption.position = "plot",
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

## My favorite breed: `r params$fave_breed`

```{r}
#| label: fave-breed
fave_breed <- breeds |>
  filter(breed == params$fave_breed)
```

```{r}
#| label: fave-breed-plot

fave_breed |>
  ggplot(
    aes(x = date, y = count)
  ) +
  geom_line(linewidth = 1) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  scale_y_continuous(
    labels = \(x) format(x, big.mark = ",", scientific = FALSE)
  ) +
  labs(
    title = paste(params$fave_breed, "population"),
    caption = "Source: Identitas AG Animal Statistics"
  ) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    plot.caption.position = "plot",
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
```

## Data source

[Identitas AG](https://tierstatistik.identitas.ch/en/index.html) is the leading
service provider for animal health, welfare and food safety in Switzerland.

Dogs are obliged to be registered for licensing purposes at a local government
level and this data is kept in the `Amicus` database. Other pets, such as pets,
rodents, birds or fish, may be registered and this data is stored in the `Anis`
database.

## Tutorial video

{{< video https://youtu.be/kQn82pa04jQ?si=b5ptcn2wVHFELaSJ >}}

{{< video https://www.youtube.com/watch?v=7dkufBboMi4&t=14s >}}
