---
title: "Autocoups and Democracy"
author: Zhu Qi

date: last-modified

format:
  pdf: 
    documentclass: article
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    #toc: true
    #toc-title: Contents
    #lof: true
    #lot: true
    #crossref: 
    #  lof-title: Figures
    #  lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  #docx: default
    
#categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
#include-before-body: title.tex
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue, plm, brglm2, flextable, broom)
```

## Abstract {.unnumbered}

This chapter investigates the impact of autocoups on political institutions,
comparing them with traditional coups through an analysis of variations in Polity
V scores. It advances two primary hypotheses: first, that incumbent leaders
frequently consolidate power by systematically undermining institutional
constraints in the period leading up to an autocoup, resulting in a decline in
Polity V scores attributable to the autocoup. Second, unlike traditional coups,
which exhibit a "U-shaped" trajectory in Polity V scores, autocoups precipitate a
persistent decline in these scores without subsequent recovery. This is
attributed to autocoup leaders’ deliberate intent to suppress opposition and
dismantle institutional checks and balances to secure prolonged tenure. Employing
a country-fixed effects model, this study demonstrates that Polity V scores
typically decline following autocoups, mirroring the magnitude of decline
observed after traditional coups. However, while traditional coups often lead to
an immediate reduction in Polity V scores followed by conditions conducive to
recovery over time, autocoups result in sustained democratic erosion. These
findings highlight the divergent political trajectories induced by coups and
autocoups. This research addresses a critical gap in the empirical analysis of
autocoups and contributes to academic and policy discussion by elucidating their
detrimental effects, particularly in terms of democratic backsliding and the
entrenchment of authoritarian governance.

**Keywords:** *Coups, Autocoups, Democratization*

\newpage

## Introduction

The decline in global political rights and civil liberties, as documented by
reports such as Freedom House's Freedom in the World 2024, marks the eighteenth
consecutive year of democratic backsliding worldwide [@freedomhouse2024freedom].
This sustained erosion naturally raises the question: What political mechanisms
primarily drive the decline in liberties and democratic quality?

One of the primary suspects historically linked to episodes of democratic
recession and the decline in global liberties since 2000 has been the coup
d'état—the violent, non-constitutional seizure of power. However, an emerging
paradox challenges the view of coups as the main contemporary perpetrator of
backsliding. Data from the most cited coup datasets suggest a secular decline in
the frequency of traditional coups; specifically, the number of coup attempts
between 2008 and 2017 represented the lowest ten-year total since at least 1960
[@powell2011a; @thyne2019]. Furthermore, despite ongoing scholarly debates, a
significant body of empirical literature contends that coups may exert a positive
or complex long-term effect on democratization by removing entrenched
dictatorships and breaking political logjams [@powell2014a; @thyne2014;
@dahl2023]. If traditional coups are decreasing and their long-term democratic
effect is ambiguous or even restorative, they may not be the principal driver of
the current, steady decline in global liberties. This critical gap compels us to
search for an alternative, more insidious mechanism of regime erosion.

In sharp contrast to the trend in traditional coups, a similar but distinct
political event—the autocoup—has increased notably since 2000 [@bermeo2016;
@baturo2022; @zhu2024]. An autocoup is defined as the extension of an incumbent
leader’s tenure in office beyond the originally mandated limit via
extra-constitutional manipulation in this study [@zhu2024]. While both coups and
autocoups disrupt established political orders, autocoups involve the insidious
erosion of democratic norms from within by the very leader sworn to uphold them.
Despite the growing prevalence and conceptual significance of this phenomenon,
its specific impact on democracy and regime transitions remains under-examined.

This article undertakes the first empirical investigation into the democratic
consequences of autocoups. Its primary objective is to determine whether
autocoups, in the current global context, entrench authoritarian rule, facilitate
democratization, or have no substantive impact on regime trajectories. Given the
conceptual and empirical parallels to coups, a secondary aim is to conduct a
focused comparative analysis of their respective effects on democratization to
clarify their broader political ramifications.

To address these questions, this study employs a fixed-effects model to evaluate
the respective impacts of coups and autocoups on democratic quality,
operationalized through Polity V scores. The findings demonstrate a critical
divergence:

While both coups and autocoups are associated with an immediate decline in
democratic quality, the impact differs significantly over time. Polity V scores
affected by coups typically exhibit a notable recovery within two years.
Conversely, democratic quality impacted by autocoups shows no such improvement
over the same period, indicating a longer and deeper impact on the erosion of
democracy.

This study makes two principal contributions to political science. Firstly, it
provides the first systematic empirical analysis of the impact of autocoups on
democratization, establishing them as a distinct political phenomenon and
addressing a critical gap in the literature on democratic backsliding. Secondly,
by comparing the effects of coups and autocoups, this research demonstrates the
more severe and sustained damage to democratic institutions caused by the latter.
This underscores the urgent need to treat autocoups not merely as a variant of
executive aggrandizement but as a distinct political pathology warranting focused
scholarly and policy attention.

The remainder of this article is structured as follows. Section 2 examines the
mechanisms through which autocoups impact democratic institutions, with
particular emphasis on their comparison with traditional coups. Section 3
outlines the research design, methodological approach, and variables employed.
Section 4 presents the empirical findings and discusses their broader
implications. Section 5 concludes by summarising the key findings and reflecting
on their significance for understanding and addressing autocoup dynamics.

## Theoretical Framework: Autocratic Intent and the Mechanisms of Sustained Damage

The search for the primary determinants of democracy and democratization has
traditionally revolved around two major explanatory poles: economic development
and the impact of coups. While both are crucial, their explanatory power for the
specific phenomenon of 21st-century democratic backsliding is increasingly
questioned.

The relationship between economic conditions and democracy remains complex and
inconclusive. Early influential work by Robert Barro, Nobel laureate in
economics, found that improvements in the standard of living—specifically per
capita GDP, primary schooling, and reduced gender gaps in education—predict
increases in democracy, as measured by electoral rights [@barro1999]. However,
other large-N studies present a counter-argument: analyzing 167 countries from
1875 to 2004, @miller2012 found that economic development can also strengthen
autocratic regimes, as indicated by a reduced likelihood of violent leader
removal. These mixed findings suggest that economic development alone is an
insufficient and sometimes contradictory predictor of regime trajectory.

The impact of traditional coups on democracy is similarly characterized by
conflicting results. Some scholars argue that coups may catalyze democratization.
@thyne2014 contend that both successful and failed coups can promote reform;
successful coups incentivize new leaders to democratize quickly for legitimacy,
while failed coups serve as credible signals that compel incumbents to enact
meaningful reforms. This view is supported by observations that the majority of
post-Cold War coups have been followed by competitive elections [@marinov2013].
However, this positive view is highly contextual, relying on the fact that most
coups occur in non-democracies. Furthermore, the outcome remains contingent on
external factors, as international reactions significantly determine whether
coup-installed leaders lean toward authoritarianism or democracy [@thyne2020].

These traditional explanations—economic growth and the impact of the coup
d'état—fail to fully account for the prolonged global democratic backsliding
observed since the 2000s, especially given the concurrent decrease in traditional
coup activity. This failure is compounded by the undeniable fact that prominent
personal dictators and strongmen in major autocracies (e.g., Putin in Russia, Xi
Jinping in China, Lukashenko in Belarus, Erdoğan in Turkey) have actively
maintained power through illegitimate or extra-constitutional means
[@baturo2019politics; @baturo2022]. Even in established democracies, attempts by
leaders to unlawfully retain office, such as President Trump's efforts to deny
the 2020 election results, have raised serious discussions about the occurrence
of "self-coups" in the US [@antonio2021; @pion-berlin2022].

Therefore, an analysis of the autocoup's impact on democracy is not only
necessary but crucial. It directly addresses the defining political trend of the
21st-century: the institutional erosion caused by incumbent leaders determined to
overstay their mandates. By filling this gap, this study can better explain the
mechanisms driving the eighteen consecutive years of decline in global liberties.

### Mechanisms of democratic damage by autocoups

A critical, yet often unmeasured, determinant of democratization is the
**inclination of incumbent leaders**: do they wish to consolidate democracy or
pursue authoritarian transition? Leaders, particularly in non-consolidated or
autocratic regimes, play a paramount role in shaping the political structure;
even in democracies, leaders can fundamentally alter institutions, as exemplified
by Hitler's ascent and dismantling of the Weimar Republic in the 1930s.

While leaders rarely admit to autocratic aspirations—many notorious authoritarian
regimes formally retain the title "democracy"—their true intentions can be
observed through their actions. The autocoup serves as a potent, unambiguous
**instrumental proxy** for a leader's intent toward autocracy. No matter the
justification, extending power through an autocoup is a clear, self-initiated
break from the constitutional process of power transition. A leader truly
committed to democratic rule would adhere to established succession limits;
therefore, the execution of an autocoup is definitive evidence of an intent
*opposite* to democratization. This makes the autocoup a valuable subject for
analyzing the roots of democratic backsliding.

Beyond merely revealing intent, the autocoup itself poses a distinct and acute
threat to democratic institutions. The process of achieving and maintaining
illicit tenure forces the leader into a self-reinforcing cycle of institutional
destruction. The democratic damage unfolds through three cumulative mechanisms:

The decline associated with autocoups begins **before the formal extension of
tenure**. To ensure a successful and uncontested breach of constitutional limits,
incumbents must first systematically neutralize potential veto players and
dismantle oversight mechanisms. This preparation involves **purging rival elites,
restricting media freedoms, weakening judicial independence, and harassing
opposition parties**. This is exemplified by Peru's 1992 *autogolpe*, where the
Polity V score collapsed from 8 to $\text{–4}$ following the dissolution of
Congress [@cameron1998]. These actions constitute the first wave of democratic
erosion.

The execution of the autocoup is a fundamental **unlawful violation of existing
constitutional constraints**. When incumbents circumvent foundational safeguards
like term limits, they destroy one of the most important institutional
protections against authoritarianism. The act permanently **weakens the
credibility of legal norms**, establishing a precedent that facilitates future
violations by subsequent leaders. Thus, the autocoup is both a symptom and a
powerful catalyst of institutional decline.

Following the illegal extension of power, leaders face a powerful structural
incentive for long-term control. The committed constitutional crime exposes the
leader to the threat of future prosecution and retribution should they be
removed. This structural necessity dictates a continuous high-pressure
environment over opponents. The autocoup leader cannot credibly commit to
liberalization without significantly increasing their personal risk. The result
is a self-reinforcing authoritarian drift, characterized by **sustained
repression and the indefinite postponement of democratic restoration**.

The immediate, sharp institutional disruption caused by the autocoup, despite its
gradual preparation, is thus hypothesized to be comparable to that of an abrupt
traditional coup:

***H1: Autocoups will result in a significant decline in Polity V scores
immediately following their occurrence, in a manner comparable to the effects
observed in traditional coups.***

### Comparative long-term outcomes: autocoups vs. coups

Despite the shared feature of immediate institutional disruption, compared to
traditional coups, the long-term impact of autocoups on democratization is
characterized by its **greater durability** and a **unidirectional trajectory**.
The effects almost exclusively manifest as a sustained descent towards autocracy,
rather than the mixed or potentially restorative outcomes often observed after
conventional coups.

Firstly, the successful rate of autocoups are much higher than that of coups.
According to @zhu2024, 83 autocoups since 1945 around the world, 64 of them are
successful, the rate of 77 per cent compared to coups' rate about 50 per cent
[@powell2011]. In the meantime, the fate of failed autocoup leaders and failed
coup plotters are different in most cases. Failed autocoup leaders usually only
lose their power, some of them even can finish their ongoing tenure. Few of them
suffer sever backfire from opponents [@baturo]. However, failed coup leaders will
not be so lucky, they will face prison, exile, or even death [@kebschull1994;
@easton2018; @easton2018a]. Hence comparably, autocoup leaders have more chances
to shape political institutions, because failed coup leaders cannot do anything
to influence the politics.

Secondly, previous research on autocoups has shown that the survival time of
post-autocoup leaders is, on average, roughly five years longer than that of
coup-installed leaders [@zhu2024]. This survival comparison has excluded
coup-installed leaders who survive less 180 days, since they have no enough time
to restructure the political institutions in such a short time. This exclusion
reduced the coup-installed leaders down to about 140 from 245 succeeded
coup-installed leaders [@powell2011a, @zhu2024]. Therefore, compared to
coup-installed leaders, autocoup leaders have much longer influence in shaping
institutions of their regimes. Most coup-installed leaders, in contrast, have
been ousted from power before they could do anything in democracy.

Thirdly, previous research on coups shows that coups produce highly variable
consequences. This is reasonable since not only failed coups might result in
different reaction from incumbent leaders and ruling groups, more importantly,
successful coups could also lead to different outcomes. As @thyne2016 discussed,
even failed coups as credible signals that leaders must enact meaningful reforms
to remain in power. Furthermore, succeed coups can have much more possibilities
than autocoups. For example, coups can act as catalysts for political renewal,
especially when they remove entrenched authoritarian leaders or aim to restore
constitutional order [@thyne2014; @miller2016]. In Niger, for example, President
Mamadou Tandja’s attempt in 2009 to amend the constitution to permit a third term
precipitated a military coup in 2010 [@miller2016]. Similarly, in Honduras the
same year, President Manuel Zelaya was removed from office by the military after
seeking to alter the constitution to allow immediate re-election
[@muñoz-portillo2019]. In these two cases, the coup leaders, the military
officers in the former and the supreme court in the latter, installed other civil
officials to replace the incumbent leaders. Although most of coup plotters do
seize the power at last, some of them still feel incentives to promise reforms to
build legitimacy and mitigate elite resistance. So they might have to make some
efforts to better off the current political pressure. Therefore, no matter failed
or succeeded, coups and the threat of coups can be a significant weapon in
fostering democracy [@derpanopoulos2016]. Of course, whether or not coups are
good or bad for democracy is still highly debating [@tansey2016; @miller2016;
@derpanopoulos2017], the empirical analyses show positive impact on democracy.

Autocoups, however, follow a fundamentally different logic. Because the act is
inherently anti-democratic and self-serving, the leader cannot later pursue
liberalization without endangering their own political survival. The structural
incentives created by the successful autocoup are thus to centralize power,
maintain long-term coercive control, and prohibit political openings. The result
is a unidirectional, durable authoritarian trajectory. Democratic decline
associated with autocoups tends not only to be sharp but also sustained, with
little evidence of recovery even in the medium term.

This difference leads to the second hypothesis:

***H2: Autocoups produce significant declines in Polity V scores that typically
do not rebound, whereas traditional coups often exhibit a U-shaped trajectory,
with initial deterioration followed by gradual democratic improvement.***

## Methodology and variables

### Methodology

As outlined above, autocoups are less likely to result in full regime
transitions—whether from democracy to autocracy or vice versa. Consequently,
evaluating their effects solely in terms of regime change or shifts across
democratic thresholds is analytically inappropriate. Instead, this study assesses
political change by examining variations in Polity V scores, which capture more
subtle shifts in institutional quality and democratic performance.

To differentiate between immediate and medium-term effects, the analysis
considers both event-year and two-year impacts of autocoups. The event-year
effect is measured as the change in Polity V score in the year of the autocoup
relative to the preceding year:

$$
Polity_{t} - Polity_{t-1}
$$

The three-year effect captures the change in Polity V score two years after the
event, relative to the year of the autocoup:

$$
Polity_{t+3} - Polity_t
$$

This three-year specification is intended to capture medium-term political
developments, as autocoups typically entrench existing power structures rather
than inducing immediate systemic change. Short-term fluctuations may not fully
reflect the institutional consequences of such events.

To empirically test the hypotheses, the study employs a linear fixed-effects
model at the country level. To distinguish between attempted and successful
autocoups, separate models are estimated using binary variables that code for
autocoup attempts and successes, respectively.

### Variables

The scholarly literature on coups' impact on democratization often evaluate
political outcomes through binary regime classifications—whether countries
democratize, autocratize, or remain stable [@thyne2014; @derpanopoulos2016]. This
framework is appropriate for traditional coups, which trigger abrupt leadership
replacement and elite rearrangement, making the political disruption easily
captured through dichotomous shifts.

However, this binary framework is inadequate in this study as it aims to measure
and compare the impact of autocoups. Unlike a classic coup d’état, an autocoup
retains the existing leader and governing coalition, meaning it rarely triggers
immediate changes in regime labels. The absence of nominal transition thus
obscures the true consequence: the subversion of institutional constraints that
regulate executive power.

For this reason, to evaluate the political impact of autocoups, a more sensitive
approach is required. This study employs the Polity V score (from the Polity5
dataset), which ranges from $\text{–10}$ to $+\text{10}$. This continuous measure
enables the detection of incremental degradation in executive constraints and
political participation, aligning with research on subtle democratic backsliding
[@dahl2023].

The analysis draws upon a global panel of country-year observations spanning from
1950 to 2020, resulting in approximately 9,100 observations. The primary
dependent variable is the change in Polity V score, calculated either as a
one-year or three-year difference, depending on the model specification. Polity V
scores range from −10 (full autocracy) to +10 (full democracy). To address
missing data caused by transitional codes (−66, −77, −88), these values are
replaced with the nearest valid Polity score to preserve temporal continuity and
reduce bias associated with listwise deletion.

The primary independent variable is the occurrence of an autocoup, as defined in
Chapter 2. The dataset includes 83 attempted and 64 successful autocoups. For
models analysing attempted autocoups, the variable is coded as 1 in the year of
the attempt and 0 otherwise. In the three-year specification, a decay function is
applied to measure the persistence of effects, following the approach of
@dahl2023. To account for temporal diffusion, a half-life of five years is
specified, allowing the model to capture both immediate and delayed consequences
from the year of the autocoup ( $y_t$ ) through to four years post-event (
$y_{t+4}$ ).

In addition, traditional coups are included as a secondary independent variable
for two reasons. First, they enable a comparative evaluation of the political
consequences of coups versus autocoups. Second, coups and autocoups may occur in
close proximity or in causal sequence, necessitating analytical disaggregation.
The coup data are drawn from @powell2011, and are coded in a manner consistent
with the autocoup variables—using a binary indicator for one-year effects and a
decay function for three-year impacts.

A set of control variables is included to account for alternative explanations.
These comprise: economic performance, proxied by GDP growth and GDP per capita;
political violence, to capture variations in political stability; and the
logarithm of population size, which serves as a proxy for state capacity and
scale effects. To mitigate concerns regarding reverse causality, all control
variables are lagged by one year, ensuring that their values precede the outcome
being measured.

Two additional dummy variables are incorporated:

**Non-democracy:** This variable captures regime type by distinguishing cases
with Polity V scores below −6 (already autocratic and less prone to further
decline) and above +6 (institutionally resilient to democratic erosion).

**Cold War:** A temporal dummy variable to account for the geopolitical context,
in line with previous studies on the relationship between coups and
democratisation [@thyne2014; @derpanopoulos2016; @dahl2023]. It captures broad
international trends, such as the stagnation or decline in democratic scores
during the Cold War (1960s–1990) and the more pronounced democratising trend
after 1990.

## Results and discussion

This section examines the democratic implications of autocoups by analysing their
effects on Polity V scores, both in the immediate aftermath and in the medium
term. @tbl-demomodel presents four models: Models 1 and 2 report results for
attempted autocoups, while Models 3 and 4 pertain to successful autocoups. Within
each group, Models 1 and 3 assess immediate effects (in the event year), whereas
Models 2 and 4 evaluate medium-term effects (three years after the event).

### Immediate democratic impact

Consistent with the first hypothesis, autocoups and coups are associated with
significant immediate declines in Polity V scores. In both Models 1 and 3,
autocoups—whether attempted or successful—lead to a statistically significant
reduction of approximately 1.3 points in Polity V scores in the event year, all
else equal. These effects are comparable in magnitude across both attempted and
successful autocoups, suggesting that the democratic damage materialises
irrespective of whether the attempt fully succeeds.

Traditional coups are associated with larger immediate declines. Model 1 shows
that attempted coups reduce Polity V scores by 1.31 points, while successful
coups, in Model 3, lead to a drop of 2.12 points, both significant at the $1\%$
level. These findings confirm that both types of irregular power grabs deliver
immediate shocks to democratic institutions, though coups—especially successful
ones—inflict greater disruption.

```{r}
#| label: coup_autocoup_data
coup_autocoup <- read.csv("https://raw.githubusercontent.com/reddylee/Datasets/refs/heads/main/coup_autocoup.csv")
```

```{r}
#| label: tbl-demomodel
#| tbl-cap: "The Impacts on Democratization(1950–2018): Autocoups vs Coups"
#| results: asis

coup_autocoup <- coup_autocoup |>
  mutate(
    GDP_pc = dplyr::lag(GDP_pc, 1),
    GDP_trend = dplyr::lag(GDP_trend, 1),
    pop_log = dplyr::lag(pop_log, 1),
    violence = dplyr::lag(violence, 1),
    .by = ccode
  )

demo_model1 <- coup_autocoup |>
  rename(Attemp = diff1, Autocoup = autocoup_attemp, Coup = coup_attemp) |>
  plm(
    Attemp ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )


demo_model2 <- coup_autocoup |>
  rename(Attemp = diff3, Autocoup = autocoup_effect_lag3, Coup = coup_effect_lag3) |>
  plm(
    Attemp ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )


demo_model3 <- coup_autocoup |>
  rename(Succeed = diff1, Autocoup = autocoup_succeed, Coup = coup_succeed) |>
  plm(
    Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )


demo_model4 <- coup_autocoup |>
  rename(Succeed = diff3, Autocoup = autocoup_seffect_lag3, Coup = coup_seffect_lag3) |>
  plm(
    Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )

stargazer(demo_model1, demo_model2, demo_model3, demo_model4,
  #type = "text",
  float = F,
  header = F,
  #dep.var.labels.include = F,
  dep.var.caption = "Dependent variable: Differences of Polity V scores",
  dep.var.labels = c("Attempted", "Succeeded"),
  column.labels = c("(1)", "(2)", "(3)","(4)"),
  model.numbers = F,
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Non-Democracy", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "30pt",
  # font.size = "footnotesize",
  df = F
)
```

### Medium-term divergence: coups vs. autocoups

In the medium term, however, the political trajectories begin to diverge: while
coups are followed by significant improvements in Polity V scores, autocoups
continue to exert a negative effect, albeit one that does not reach statistical
significance.

Models 2 and 4 evaluate changes in Polity V scores three years after the event.
The results indicate that autocoups have no statistically significant effect in
the medium term—whether attempted or successful—implying that the initial
democratic decline is not followed by subsequent institutional reform or
recovery. In contrast, attempted coups are associated with a significant increase
of 1.2 points, and successful coups show a particularly strong rebound of 1.87
points, both at the $1\%$ significance level.

These findings provide clear support for the second hypothesis. Whereas coups
tend to exhibit a “U-shaped” pattern—with democratic erosion followed by
recovery—autocoups demonstrate a consistent, unidirectional decline in democratic
quality, with no evidence of rebound.

The results suggest that autocoups exert their impact primarily in the short
term, as reflected in the immediate drop in Polity V scores, while offering no
potential for democratic revitalisation in the medium term. This contrasts with
coups, which, although initially disruptive, sometimes serve as catalysts for
institutional renewal, particularly in cases where they are followed by electoral
processes or popular mobilisation.

These findings reinforce the notion that autocoups function to entrench
incumbents, undermining constitutional safeguards and consolidating executive
power. By contrast, coups—particularly those that displace entrenched regimes—may
open space for institutional realignment or liberalisation, depending on the
post-coup political context.

The models incorporate a range of control variables to isolate the effects of
coups and autocoups:

GDP per capita is negatively and significantly associated with changes in Polity
V scores across all models. This counterintuitive negative association may
reflect the limited potential for democratic gains in already high-income
democracies, where Polity V scores are near their ceiling.

Log of population size is positively and significantly associated with Polity
score changes, suggesting that larger states may possess greater institutional
adaptability or reform potential.

The results for non-democratic regimes (defined as those with Polity V scores
below −6) reveal a temporal asymmetry in their effects on democratic outcomes. In
the event-year models (Models 1 and 3), non-democratic regimes are associated
with significant positive changes in Polity V scores. This likely reflects cases
where short-term liberalisation or reform efforts follow leadership crises or
institutional ruptures, producing modest democratic gains even within
authoritarian contexts. By contrast, in the three-year models (Models 2 and 4),
the effect reverses direction: non-democratic regimes are associated with
significant declines in Polity V scores over the medium term. This pattern
suggests that early signs of liberalisation often fail to consolidate and may be
followed by renewed authoritarian entrenchment. In essence, while non-democratic
regimes may exhibit initial democratic openings—whether symbolic or
procedural—these gains are frequently short-lived, with longer-term trajectories
reverting to autocratic norms. This dynamic underscores the fragility of
democratic progress in authoritarian contexts, where reforms introduced in the
aftermath of institutional disruption are often superficial or strategically
instrumental, lacking the structural support required for sustained
democratisation.

Cold War context is statistically significant only in the event-year models,
where it correlates with a decline in Polity V scores, reflecting the broader
global pattern of democratic suppression during the Cold War period.

Political violence and economic growth do not show consistent or significant
effects, indicating that immediate democratic outcomes are more sensitive to
regime characteristics and structural factors than to short-term economic or
security conditions.

Overall, the empirical results offer robust support for both hypotheses.
Autocoups and coups both lead to significant immediate declines in democratic
quality, with coups inflicting greater short-term damage. In the medium term,
coups are often followed by democratic recovery, whereas autocoups result in
persistent democratic erosion with no evidence of rebound.

These findings suggest that autocoups represent a particularly insidious form of
democratic backsliding, less dramatic than coups but ultimately more damaging in
their long-term effects. They reinforce the need for greater scholarly and policy
attention to constitutional manipulations by incumbents, which, although often
gradual and legally framed, can produce lasting democratic decay.

### Robustness tests

To assess the robustness of the main findings, a series of alternative model
specifications were estimated. The results confirm that the core conclusions
remain stable under these variations.

First, the operationalisation of the autocoup variable was modified: the decay
function used in the baseline analysis was replaced with a binary indicator
distinguishing between attempted and successful autocoups. Additionally, the
broad ‘non-democracy’ category was disaggregated into more specific regime
types—military, personalist, presidential, parliamentary, and ‘other’—with
dominant-party regimes serving as the reference category. This classification
mirrors the approach used in the determinants analysis of autocoups presented in
earlier chapters. The results of these robustness models are presented in Models
5 to 8 in @tbl-demomodel2.

```{r}
#| label: tbl-demomodel2
#| tbl-cap: "The Impact of Autocoups on Democratization: Binary Autocoups"
#| results: asis


demo_model1 <- plm(
  Attempted ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Attempted = diff1, Autocoup = autocoup_attemp, Coup = coup_attemp) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)

demo_model2 <- plm(
  Attempted ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Attempted = diff3, Autocoup = autocoup_attemp_lag3, Coup = coup_attemp_lag3) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)

demo_model3 <- plm(
  Succeeded ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Succeeded = diff1, Autocoup = autocoup_succeed, Coup = coup_succeed) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)
demo_model4 <- plm(
  Succeeded ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Succeeded = diff3, Autocoup = autocoup_succeed_lag3, Coup = coup_succeed_lag3) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)

stargazer(demo_model1, demo_model2, demo_model3, demo_model4,
  #type = "text",
  float = F,
  header = F,
  column.labels = c("(5)", "(6)", "(7)","(8)"),
  model.numbers = F,
  dep.var.caption = "Dependent variable: Differences of Polity V scores",
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Regime: Military", "\\hspace{1.5cm} Personal", "\\hspace{1.5cm} Presidential", "\\hspace{1.5cm} Parliamentary", "\\hspace{1.5cm} Other", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "20pt",
  # font.size = "footnotesize",
  df = F
)
```

Consistent with the main models, autocoups remain significantly associated with
negative changes in Polity V scores in the short term (Models 5 and 7), with
coefficients of −1.236 and −1.234, respectively (both significant at the $1\%$
level). However, in the three-year models (Models 6 and 8), the effect becomes
statistically insignificant, indicating that the negative effect of autocoups is
immediate but not sustained over time.

By contrast, coups continue to show a distinct “U-shaped” effect. In the
event-year models (Models 5 and 7), coups are associated with significant
declines in Polity V scores (−1.366 and −2.190), both at the $1\%$ level. Yet in
the three-year models (Models 6 and 8), the effect reverses direction: coups are
now associated with large positive changes in Polity V scores (+1.240 and +1.712,
also significant at the $1\%$ level). This confirms the earlier interpretation
that while coups may cause immediate democratic disruption, they are often
followed by democratic recovery in the medium term.

The disaggregated regime type variables provide additional insights. Military
regimes show significant positive effects in the event-year models (Models 5 and
7), with coefficients of +0.602 and +0.574, but become negative and significant
in the three-year models (−0.545 and −0.584 in Models 6 and 8). This reversal
suggests that initial post-event liberalisation in military regimes is not
sustained, and may even regress.

Personalist regimes are consistently associated with negative and significant
effects in the three-year models (Models 6 and 8: −0.532 and −0.526), but not in
the two-year models, suggesting that their democratic erosion becomes more
evident over time.

Presidential and parliamentary democracies follow a similar pattern: both show
significant negative effects in the short term (Models 5 and 7), and positive,
statistically significant effects in the medium term (Models 6 and 8). For
example, parliamentary democracies are associated with a drop of −0.475/−0.468 in
the short term but show a gain of +0.965/0.966 over three years. This pattern
supports the idea that democratic institutions may initially be shaken by
political disruption but recover when institutional mechanisms are strong.

“Other” regimes (likely transitional or provisional systems) show consistently
large and positive effects across all models, ranging from +0.999 to +1.115, all
significant at the $1\%$ level. This implies that these regimes tend to
transition toward more democratic forms over both short and medium time frames.

Several control variables also behave consistently with the baseline models. GDP
per capita is negatively and significantly associated with changes in Polity V
scores across all models, again likely reflecting ceiling effects in advanced
democracies with limited room for improvement. Log of population is positively
and significantly related to Polity changes, reinforcing earlier interpretations
that larger states may possess greater reform potential or be more likely to
register changes in democratic performance. Political violence becomes
statistically significant only in the three-year models (Models 6 and 8), where
it has a small positive effect (+0.044, +0.046), suggesting that prolonged unrest
may precede some form of institutional response or democratic opening. The Cold
War variable is significant only in the event-year models (Models 5 and 7), where
it is associated with small negative effects (−0.168 and −0.156), consistent with
broader patterns of democratic suppression during the Cold War period.

These robustness models confirm the main findings while offering additional
nuance. These results underscore the importance of both regime context and
temporal scope in evaluating the consequences of irregular power grabs.
Autocoups, unlike coups, represent a consistently negative force for democratic
institutions—one that undermines without paving the way for recovery.

## Conclusion

This chapter has examined the impact of autocoups on democratic institutions by
analysing changes in Polity V scores, with a comparative focus on traditional
coups. Two key hypotheses guided the analysis: first, that autocoups are
associated with consistent declines in democratic quality, particularly in the
short term; and second, that while coups often generate initial disruptions, they
tend to produce a “U-shaped” effect, marked by subsequent democratic recovery or
even advancement in the medium term.

The empirical results offer strong support for these hypotheses. Across multiple
model specifications, autocoups—whether attempted or successful—exhibit
significant negative effects on Polity V scores in the event year, but these
effects do not persist into the medium term. In contrast, coups are associated
with significant democratic improvement three years after the event, despite an
initial decline. This pattern is robust across models incorporating disaggregated
regime types, alternative lag structures, and extended time horizons.

The analysis further reveals important variation across regime types. Military
and personalist regimes, while sometimes exhibiting modest democratic gains in
the immediate aftermath, tend to experience declines in Polity V scores over
time, suggesting a return to entrenched authoritarianism. Presidential and
parliamentary democracies, by contrast, initially register democratic decline but
tend to recover within three years—consistent with institutional resilience.
Notably, transitional or provisional regimes (“other” types) display consistently
strong democratic gains, underscoring their potential for reform during periods
of flux.

The findings carry several theoretical and policy-relevant implications. While
coups are widely recognised as pivotal events in the study of regime change,
autocoups deserve greater scholarly attention. Unlike coups, which may at times
catalyse democratic transitions, autocoups represent a systematically
anti-democratic mechanism, typically employed to erode checks on executive power
and extend incumbents’ rule. Moreover, as shown in Chapter 4, autocoup leaders
tend to retain power for longer periods—nearly a decade on average—compared to
less than seven years for coup-installed leaders, implying more durable
institutional consequences.

This chapter also advances the methodological literature by emphasising the
importance of temporal framing. Many political shocks—particularly autocoups—are
preceded by elite purges, electoral manipulation, or institutional weakening.
Consequently, focusing solely on post-event changes risks overlooking the
cumulative nature of democratic decline. The findings thus support a more
longitudinal and process-oriented approach to studying regime erosion.

Nevertheless, limitations remain. Notably, coups and autocoups occasionally occur
in close temporal proximity, making it difficult to disentangle their respective
contributions to changes in Polity V scores. Future research should seek to
better isolate these overlapping effects, perhaps through finer-grained event
sequencing or qualitative case tracing.

In sum, this chapter reinforces the view that autocoups are a critical yet
underexplored driver of democratic backsliding. Their often-subtle execution
belies their long-term consequences. As such, they warrant continued empirical
scrutiny and deeper integration into both the comparative democratisation
literature and policy frameworks concerned with defending constitutional
governance.
