---
title: "Autocoups and Democracy"
author: Zhu Qi

date: last-modified

format:
  pdf: 
    documentclass: article
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    #toc: true
    #toc-title: Contents
    #lof: true
    #lot: true
    #crossref: 
    #  lof-title: Figures
    #  lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  #docx: default
    
#categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
#include-before-body: title.tex
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue, plm, brglm2, flextable, broom)
```

## Abstract {.unnumbered}

This chapter investigates the impact of autocoups on political institutions,
comparing them with traditional coups through an analysis of variations in Polity
V scores. It advances two primary hypotheses: first, that incumbent leaders
frequently consolidate power by systematically undermining institutional
constraints in the period leading up to an autocoup, resulting in a decline in
Polity V scores attributable to the autocoup. Second, unlike traditional coups,
which exhibit a "U-shaped" trajectory in Polity V scores, autocoups precipitate a
persistent decline in these scores without subsequent recovery. This is
attributed to autocoup leaders’ deliberate intent to suppress opposition and
dismantle institutional checks and balances to secure prolonged tenure. Employing
a country-fixed effects model, this study demonstrates that Polity V scores
typically decline following autocoups, mirroring the magnitude of decline
observed after traditional coups. However, while traditional coups often lead to
an immediate reduction in Polity V scores followed by conditions conducive to
recovery over time, autocoups result in sustained democratic erosion. These
findings highlight the divergent political trajectories induced by coups and
autocoups. This research addresses a critical gap in the empirical analysis of
autocoups and contributes to academic and policy discussion by elucidating their
detrimental effects, particularly in terms of democratic backsliding and the
entrenchment of authoritarian governance.

**Keywords:** *Coups, Autocoups, Democratization*

\newpage

## Introduction

The decline in global political rights and civil liberties, as documented by
reports such as Freedom House's Freedom in the World 2024, marks the eighteenth
consecutive year of democratic backsliding worldwide [@freedomhouse2024freedom].
This sustained erosion naturally raises the question: What political mechanisms
primarily drive the decline in liberties and democratic quality?

One of the primary suspects historically linked to episodes of democratic
recession and the decline in global liberties since 2000 has been the coup
d'état—the violent, non-constitutional seizure of power. However, an emerging
paradox challenges the view of coups as the main contemporary perpetrator of
backsliding. Data from the most cited coup datasets suggest a secular decline in
the frequency of traditional coups; specifically, the number of coup attempts
between 2008 and 2017 represented the lowest ten-year total since at least 1960
[@powell2011a; @thyne2019]. Furthermore, despite ongoing scholarly debates, a
significant body of empirical literature contends that coups may exert a positive
or complex long-term effect on democratization by removing entrenched
dictatorships and breaking political logjams [@powell2014a; @thyne2014;
@dahl2023]. If traditional coups are decreasing and their long-term democratic
effect is ambiguous or even restorative, they may not be the principal driver of
the current, steady decline in global liberties. This critical gap compels us to
search for an alternative, more insidious mechanism of regime erosion.

In sharp contrast to the trend in traditional coups, a similar but distinct
political event—the autocoup—has increased notably since 2000 [@bermeo2016;
@baturo2022; @zhu2024]. An autocoup is defined as the extension of an incumbent
leader’s tenure in office beyond the originally mandated limit via
extra-constitutional manipulation in this study [@zhu2024]. While both coups and
autocoups disrupt established political orders, autocoups involve the insidious
erosion of democratic norms from within by the very leader sworn to uphold them.
Despite the growing prevalence and conceptual significance of this phenomenon,
its specific impact on democracy and regime transitions remains under-examined.

This article undertakes the first empirical investigation into the democratic
consequences of autocoups. Its primary objective is to determine whether
autocoups, in the current global context, entrench authoritarian rule, facilitate
democratization, or have no substantive impact on regime trajectories. Given the
conceptual and empirical parallels to coups, a secondary aim is to conduct a
focused comparative analysis of their respective effects on democratization to
clarify their broader political ramifications.

To address these questions, this study employs a fixed-effects model to evaluate
the respective impacts of coups and autocoups on democratic quality,
operationalized through Polity V scores. The findings demonstrate a critical
divergence:

While both coups and autocoups are associated with an immediate decline in
democratic quality, the impact differs significantly over time. Polity V scores
affected by coups typically exhibit a notable recovery within two years.
Conversely, democratic quality impacted by autocoups shows no such improvement
over the same period, indicating a longer and deeper impact on the erosion of
democracy.

This study makes two principal contributions to political science. Firstly, it
provides the first systematic empirical analysis of the impact of autocoups on
democratization, establishing them as a distinct political phenomenon and
addressing a critical gap in the literature on democratic backsliding. Secondly,
by comparing the effects of coups and autocoups, this research demonstrates the
more severe and sustained damage to democratic institutions caused by the latter.
This underscores the urgent need to treat autocoups not merely as a variant of
executive aggrandizement but as a distinct political pathology warranting focused
scholarly and policy attention.

The remainder of this article is structured as follows. Section 2 examines the
mechanisms through which autocoups impact democratic institutions, with
particular emphasis on their comparison with traditional coups. Section 3
outlines the research design, methodological approach, and variables employed.
Section 4 presents the empirical findings and discusses their broader
implications. Section 5 concludes by summarising the key findings and reflecting
on their significance for understanding and addressing autocoup dynamics.

## Autocoups: Definition and Measurement of Democratic Impact

### Defining the Irregular Transition

Traditional political science has focused on the **coup d'état**, defined by
@powell2011 as "illegal and overt attempts by the military or other elites within
the state apparatus to unseat the sitting executive." This definition focuses on
**leadership turnover** and **external disruption**.

In contrast, the **autocoup** refers to the phenomenon of an incumbent leader's
refusal to relinquish power and has received comparatively less scholarly
attention, despite its growing frequency. Since the end of the Cold War, while
classic coups have declined, these "incumbent retention" or "overstay" strategies
have become more frequent [@ginsburg2010evasion; @baturo2014].

To conduct a valuable pioneer analysis, this study formally defines the
**autocoup as the extension of an incumbent leader’s tenure in office beyond the
originally mandated limit, achieved through extra-constitutional means.**

### Data Compilation

To operationalize this definition, **I compile a new autocoup dataset** based on
three complementary sources:

-   **Leader Identification:** I utilize the Archigos dataset [@goemans2009] and
    the Political Leaders’ Affiliation Database (PLAD) [@bomprezzi2024wedded] to
    identify *de facto* national leaders and their precise time in office
    (1945–2023).

-   **Event Inventory:** I use the Incumbent Takeover dataset [@baturo2022] as
    the primary inventory of potential autocoup events.

-   **Cross-Referencing:** Entries from the Incumbent Takeover dataset are
    cross-referenced with Archigos and PLAD to isolate only those events that
    meet the strict definition of an autocoup (illegal tenure **extension**).

### Measuring Impact on Democracy

The scholarly literature on the impact of traditional coups on democratization
often evaluates political outcomes through **binary regime classifications**
(e.g., democratize, autocratize) [@thyne2014; @derpanopoulos2016]. This framework
is appropriate for coups, which trigger abrupt leadership replacement and thus
easily captured dichotomous shifts.

However, this binary framework is **inadequate for measuring and comparing the
impact of autocoups**. An autocoup retains the existing leader, meaning it rarely
triggers immediate changes in regime labels. The absence of nominal transition
thus **obscures the true consequence: the subversion of institutional constraints
that regulate executive power.**

For this reason, to evaluate the political impact of autocoups, a **more
sensitive and continuous approach is required**. This study employs the **Polity
V score** (from the Polity5 dataset), which ranges from $\text{–10}$ (full
autocracy) to $+\text{10}$ (full democracy). This continuous measure enables the
detection of **incremental degradation** in executive constraints and political
participation, aligning with recent research on subtle democratic backsliding
[@dahl2023].

## Theoretical framework: **Autocoups as a Mechanism of Democratic Erosion**

### Beyond Traditional Determinants: Rethinking the Drivers of Democratic Decline

Scholars of democratization have long emphasized two broad explanatory traditions
for understanding regime trajectories: the structural effects of economic
development and the political consequences of coercive power transitions such as
coups d’état. Each of these literatures offers valuable insights, but neither
sufficiently accounts for the pattern of persistent, widespread democratic
backsliding since the early 2000s—a period in which global freedom scores have
declined for more than eighteen consecutive years.

Early structural approaches, rooted in modernization theory, argued that rising
levels of development foster democratic outcomes by expanding education,
generating a middle class, and promoting economic autonomy (Lipset 1959). Robert
Barro’s cross-national work, for example, finds that improvements in GDP per
capita, education levels, and human development indicators predict increases in
democratic rights [@barro1999]. Yet later research complicates this consensus.
Using a dataset spanning 167 countries from 1875 to 2004, @miller2012
demonstrates that economic development can stabilize authoritarian leaders by
reducing the likelihood they are violently removed from office. In this view,
modernization may bolster entrenched autocrats as much as it supports democratic
transitions. The mixed findings underscore that economic conditions alone cannot
explain the contemporary global shift toward illiberal governance.

A second major literature focuses on the role of coups d’état in shaping regime
outcomes. Historically, coups have been associated with heightened political
instability and democratic decline. Yet post–Cold War evidence is far more
ambiguous. Several scholars argue that coups—successful or even failed—can
sometimes catalyze democratization, particularly when they remove authoritarian
incumbents or pressure elites into reform [@thyne2014; @marinov2013]. This
interpretation is reinforced by the fact that most post–Cold War coups have been
followed by competitive multiparty elections. However, these positive pathways
are highly conditional on international incentives, domestic coalitions, and the
political motives of coup plotters. As a result, the long-term effects of coups
remain inconsistent and unpredictable.

While economic growth and coups remain important pieces of the democratization
puzzle, neither can explain the systematic pattern of institutional erosion
observed across diverse regions since the early 2000s, particularly given the
simultaneous decline in the frequency of traditional coups [@powell2011;
@thyne2020]. What is more striking—and comparatively understudied—is that many
contemporary leaders have prolonged their tenure not by forceful military
seizures of power, but by subverting constitutional procedures from within.
Prominent examples range from Vladimir Putin and Aleksandr Lukashenko to Xi
Jinping, Recep Tayyip Erdoğan, and Nayib Bukele [@baturo2019politics;
@baturo2022]. Even established democracies have faced such threats: Donald
Trump’s attempt to overturn the 2020 U.S. election results generated widespread
debate about the possibility of an American self-coup [@antonio2021;
@pion-berlin2022].

Against this backdrop, the autocoup—in which an incumbent illegally extends their
tenure by dismantling constitutional constraints—emerges as a critical yet
understudied mechanism of democratic erosion. The analysis of autocoups is not
only timely but necessary for understanding the most salient pattern of
21st-century regime change: the slow, deliberate, and internal dismantling of
democratic institutions by incumbents who seek to remain in power.

### Why Autocoups Matter: Leadership Agency and Authoritarian Intent

Existing democratization research often privileges structural
determinants—economic performance, resource endowments, regime type—while
treating the preferences of political leaders as uniform or unobservable. Yet
leaders play a decisive role in institutional trajectories, especially in hybrid
regimes or weak democracies. Autocratic consolidation frequently reflects
deliberate choices by incumbents who seek to extend their tenure and secure
immunity from political or legal threats.

Measuring leader intent, however, poses a substantial challenge. Political elites
rarely admit authoritarian ambitions openly, and even entrenched autocracies
often maintain democratic façades. Nevertheless, autocoups offer a uniquely
observable indicator of authoritarian preference. Extending tenure through
unconstitutional means is an unmistakable breach of democratic norms. A leader
committed to democratic rule could abide by term limits and step down; only a
leader with authoritarian ambitions circumvents constitutional constraints.

Thus, an autocoup is analytically significant in two respects. One is the
revealed preference, which provides a direct, observable signal of a leader’s
intention to subvert democracy. The other is the institutional mechanism, which
initiates and accelerates processes of institutional decay.

Unlike many political behaviors that are ambiguous or open to interpretation, an
autocoup constitutes a clear and deliberate repudiation of the institutional
framework of democratic succession.

### The Causal Mechanisms: How Autocoups Damage Democratic Institutions

Autocoups generate democratic erosion through a cumulative, multi-stage process
that begins well before the formal tenure extension and persists long after the
event. This process can be conceptualized through three mutually reinforcing
mechanisms: preparation, execution, and consolidation.

#### Pre-Autocoup Preparation: Weakening Constraints and Neutralizing Opposition

Democratic decline typically begins before the autocoup itself. To ensure a
successful and uncontested breach of constitutional limits, incumbents must first
eliminate or neutralize potential veto players. This preparatory stage often
involves: purging rival elites and reshuffling security leadership, restricting
media freedoms and discrediting independent journalism, weakening judicial
autonomy, harassing or outlawing opposition parties, and deploying state
resources to entrench patronage networks.

These actions gradually erode the institutional foundations of accountability and
transparency. For instance, Alberto Fujimori’s 1992 self-coup in Peru—where the
Polity V score collapsed from +8 to –4—was preceded by systematic efforts to
undermine oversight bodies and neutralize political opponents (Cameron 1998).
Such preemptive moves constitute the first wave of democratic degradation:
erosion that is deliberate, cumulative, and strategically designed to smooth the
path toward unconstitutional tenure extension.

#### Execution: Violating the Constitutional Order and Destroying Succession Norms

The autocoup itself represents a profound rupture in the democratic order. By
bypassing term limits or dissolving representative institutions, incumbents
breach the central norm regulating the peaceful rotation of power. Term limits
are foundational safeguards against personalist authoritarianism; violating them
eliminates a key constraint on executive power and undermines the credibility of
constitutional rules.

This rupture produces two major consequences. One is the institutional
de-legitimation. By breaking constitutional constraints, the autocoup signals to
elites and the public that legal rules are contingent and malleable. The other is
precedent for future violations. Once constraints are breached, future leaders
(or even the same leader) find it far easier to commit additional violations,
accelerating the process of democratic decay.

This moment represents the most visible—and often the most dramatic—expression of
democratic breakdown.

#### Post-Autocoup Consolidation: Structural Incentives for Repression

After successfully extending tenure, the autocoup leader faces a heightened
threat of retribution. Having committed an illegal act, the leader becomes
vulnerable to future prosecution or retaliation should they leave office. This
creates a powerful incentive for continued authoritarian consolidation.

The leader must maintain a system capable of deterring opponents and preventing
any political opening that could threaten their rule. This imperative generates a
self-reinforcing authoritarian drift, characterized by: sustained repression of
opponents, concentration of executive power, politicization of security forces,
permanent weakening of horizontal accountability, and indefinite postponement of
democratic restoration.

Thus, democratic decline accelerates not only because institutions are
dismantled, but because the leader’s survival now depends on maintaining the very
system that subverts them. The consolidation phase ensures that democratic
erosion is not merely abrupt but persistent.

Given this multi-stage process, the immediate institutional impact of an autocoup
is expected to be substantial. Although autocoups differ from coups in that
incumbents remain in power, both events produce sharp institutional
discontinuities by disrupting established rules of succession and governance.

***H1: Autocoups will result in a significant decline in Polity V scores
immediately following their occurrence, comparable in magnitude to the effects
observed after traditional coups.***

This reflects the fact that the autocoup represents an acute violation of
constitutional order, producing an immediate deterioration in formal democratic
institutions.

### Divergent Long-Term Trajectories: Why Autocoups Are More Damaging Than Coups

Although autocoups and coups can produce similar short-term institutional shocks,
their long-term consequences diverge sharply. The literature on coups identifies
wide variation in post-coup trajectories: some coups deepen authoritarian rule,
while others enable democratic transitions or liberalization. By contrast,
autocoups produce a consistent pattern of long-term democratic decline.

Autocoups succeed far more frequently than coups. Globally, approximately 77% of
autocoups succeed [@zhu2024], compared to only about half of traditional coups
[@powell2011]. Failed coup plotters face imprisonment, exile, or death, whereas
failed autocoup leaders often remain in office or complete their term
[@baturo2019]. Consequently, autocoup leaders enjoy far greater opportunities to
reshape institutions.

Autocoup leaders survive in office significantly longer—on average five
additional years—than coup-installed leaders [@zhu2024]. This expanded tenure
provides more time to institutionalize authoritarian practices, control elite
networks, and eliminate alternative centers of power. In contrast, many
coup-installed leaders are ousted before they can make durable institutional
changes.

Traditional coups can produce diverse institutional trajectories. Successful
coups sometimes facilitate democratic openings, particularly when removing deeply
authoritarian incumbents [@thyne2014; @miller2016]. Failed coups can force
incumbents to implement reforms. The motivations of coup plotters also differ:
many seek legitimacy and public support, creating incentives to promise
democratic restoration.

Autocoups, by contrast, follow a unidirectional logic. The leader has violated
constitutional order and therefore cannot credibly commit to liberalization
without increasing personal risk. The structural incentives following an autocoup
push the leader toward deeper authoritarian consolidation, not reform.

Because autocoups are driven by self-preservation and institutional manipulation
rather than political renewal, their long-term impact is systematically negative.

***H2: Autocoups generate sustained declines in Polity V scores that do not
rebound, whereas traditional coups often follow a U-shaped trajectory
characterized by an initial decline followed by gradual democratic
improvement.***

Thus, whereas coups may trigger democratization under certain conditions,
autocoups produce long-lasting—and often irreversible—democratic erosion.

## Methodology and variables

### Methodology

As outlined above, autocoups are less likely to result in full regime
transitions—whether from democracy to autocracy or vice versa. Consequently,
evaluating their effects solely in terms of regime change or shifts across
democratic thresholds is analytically inappropriate. Instead, this study assesses
political change by examining variations in Polity V scores, which capture more
subtle shifts in institutional quality and democratic performance.

To differentiate between immediate and medium-term effects, the analysis
considers both event-year and two-year impacts of autocoups. The event-year
effect is measured as the change in Polity V score in the year of the autocoup
relative to the preceding year:

$$
Polity_{t} - Polity_{t-1}
$$

The three-year effect captures the change in Polity V score two years after the
event, relative to the year of the autocoup:

$$
Polity_{t+3} - Polity_t
$$

This three-year specification is intended to capture medium-term political
developments, as autocoups typically entrench existing power structures rather
than inducing immediate systemic change. Short-term fluctuations may not fully
reflect the institutional consequences of such events.

To empirically test the hypotheses, the study employs a linear fixed-effects
model at the country level. To distinguish between attempted and successful
autocoups, separate models are estimated using binary variables that code for
autocoup attempts and successes, respectively.

### Variables

The analysis draws upon a global panel of country-year observations spanning from
1950 to 2020, resulting in approximately 9,100 observations. The primary
dependent variable is the change in Polity V score, calculated either as a
one-year or three-year difference, depending on the model specification. Polity V
scores range from −10 (full autocracy) to +10 (full democracy). To address
missing data caused by transitional codes (−66, −77, −88), these values are
replaced with the nearest valid Polity score to preserve temporal continuity and
reduce bias associated with listwise deletion.

The primary independent variable is the occurrence of an autocoup, as defined in
Chapter 2. The dataset includes 83 attempted and 64 successful autocoups. For
models analysing attempted autocoups, the variable is coded as 1 in the year of
the attempt and 0 otherwise. In the three-year specification, a decay function is
applied to measure the persistence of effects, following the approach of
@dahl2023. To account for temporal diffusion, a half-life of five years is
specified, allowing the model to capture both immediate and delayed consequences
from the year of the autocoup ( $y_t$ ) through to four years post-event (
$y_{t+4}$ ).

In addition, traditional coups are included as a secondary independent variable
for two reasons. First, they enable a comparative evaluation of the political
consequences of coups versus autocoups. Second, coups and autocoups may occur in
close proximity or in causal sequence, necessitating analytical disaggregation.
The coup data are drawn from @powell2011, and are coded in a manner consistent
with the autocoup variables—using a binary indicator for one-year effects and a
decay function for three-year impacts.

A set of control variables is included to account for alternative explanations.
These comprise: economic performance, proxied by GDP growth and GDP per capita;
political violence, to capture variations in political stability; and the
logarithm of population size, which serves as a proxy for state capacity and
scale effects. To mitigate concerns regarding reverse causality, all control
variables are lagged by one year, ensuring that their values precede the outcome
being measured.

Two additional dummy variables are incorporated:

**Non-democracy:** This variable captures regime type by distinguishing cases
with Polity V scores below −6 (already autocratic and less prone to further
decline) and above +6 (institutionally resilient to democratic erosion).

**Cold War:** A temporal dummy variable to account for the geopolitical context,
in line with previous studies on the relationship between coups and
democratisation [@thyne2014; @derpanopoulos2016; @dahl2023]. It captures broad
international trends, such as the stagnation or decline in democratic scores
during the Cold War (1960s–1990) and the more pronounced democratising trend
after 1990.

## Results and discussion

This section examines the democratic implications of autocoups by analysing their
effects on Polity V scores, both in the immediate aftermath and in the medium
term. @tbl-demomodel presents four models: Models 1 and 2 report results for
attempted autocoups, while Models 3 and 4 pertain to successful autocoups. Within
each group, Models 1 and 3 assess immediate effects (in the event year), whereas
Models 2 and 4 evaluate medium-term effects (three years after the event).

### Immediate democratic impact

Consistent with the first hypothesis, autocoups and coups are associated with
significant immediate declines in Polity V scores. In both Models 1 and 3,
autocoups—whether attempted or successful—lead to a statistically significant
reduction of approximately 1.3 points in Polity V scores in the event year, all
else equal. These effects are comparable in magnitude across both attempted and
successful autocoups, suggesting that the democratic damage materialises
irrespective of whether the attempt fully succeeds.

Traditional coups are associated with larger immediate declines. Model 1 shows
that attempted coups reduce Polity V scores by 1.31 points, while successful
coups, in Model 3, lead to a drop of 2.12 points, both significant at the $1\%$
level. These findings confirm that both types of irregular power grabs deliver
immediate shocks to democratic institutions, though coups—especially successful
ones—inflict greater disruption.

```{r}
#| label: coup_autocoup_data
coup_autocoup <- read.csv("https://raw.githubusercontent.com/reddylee/Datasets/refs/heads/main/coup_autocoup.csv")
```

```{r}
#| label: tbl-demomodel
#| tbl-cap: "The Impacts on Democratization(1950–2018): Autocoups vs Coups"
#| results: asis

coup_autocoup <- coup_autocoup |>
  mutate(
    GDP_pc = dplyr::lag(GDP_pc, 1),
    GDP_trend = dplyr::lag(GDP_trend, 1),
    pop_log = dplyr::lag(pop_log, 1),
    violence = dplyr::lag(violence, 1),
    .by = ccode
  )

demo_model1 <- coup_autocoup |>
  rename(Attemp = diff1, Autocoup = autocoup_attemp, Coup = coup_attemp) |>
  plm(
    Attemp ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )


demo_model2 <- coup_autocoup |>
  rename(Attemp = diff3, Autocoup = autocoup_effect_lag3, Coup = coup_effect_lag3) |>
  plm(
    Attemp ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )


demo_model3 <- coup_autocoup |>
  rename(Succeed = diff1, Autocoup = autocoup_succeed, Coup = coup_succeed) |>
  plm(
    Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )


demo_model4 <- coup_autocoup |>
  rename(Succeed = diff3, Autocoup = autocoup_seffect_lag3, Coup = coup_seffect_lag3) |>
  plm(
    Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
    index = c("ccode", "year"),
    model = "within", # fixed effect
    effect = "individual",
    data = _
  )

stargazer(demo_model1, demo_model2, demo_model3, demo_model4,
  #type = "text",
  float = F,
  header = F,
  #dep.var.labels.include = F,
  dep.var.caption = "Dependent variable: Differences of Polity V scores",
  dep.var.labels = c("Attempted", "Succeeded"),
  column.labels = c("(1)", "(2)", "(3)","(4)"),
  model.numbers = F,
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Non-Democracy", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "30pt",
  # font.size = "footnotesize",
  df = F
)
```

### Medium-term divergence: coups vs. autocoups

In the medium term, however, the political trajectories begin to diverge: while
coups are followed by significant improvements in Polity V scores, autocoups
continue to exert a negative effect, albeit one that does not reach statistical
significance.

Models 2 and 4 evaluate changes in Polity V scores three years after the event.
The results indicate that autocoups have no statistically significant effect in
the medium term—whether attempted or successful—implying that the initial
democratic decline is not followed by subsequent institutional reform or
recovery. In contrast, attempted coups are associated with a significant increase
of 1.2 points, and successful coups show a particularly strong rebound of 1.87
points, both at the $1\%$ significance level.

These findings provide clear support for the second hypothesis. Whereas coups
tend to exhibit a “U-shaped” pattern—with democratic erosion followed by
recovery—autocoups demonstrate a consistent, unidirectional decline in democratic
quality, with no evidence of rebound.

The results suggest that autocoups exert their impact primarily in the short
term, as reflected in the immediate drop in Polity V scores, while offering no
potential for democratic revitalisation in the medium term. This contrasts with
coups, which, although initially disruptive, sometimes serve as catalysts for
institutional renewal, particularly in cases where they are followed by electoral
processes or popular mobilisation.

These findings reinforce the notion that autocoups function to entrench
incumbents, undermining constitutional safeguards and consolidating executive
power. By contrast, coups—particularly those that displace entrenched regimes—may
open space for institutional realignment or liberalisation, depending on the
post-coup political context.

The models incorporate a range of control variables to isolate the effects of
coups and autocoups:

GDP per capita is negatively and significantly associated with changes in Polity
V scores across all models. This counterintuitive negative association may
reflect the limited potential for democratic gains in already high-income
democracies, where Polity V scores are near their ceiling.

Log of population size is positively and significantly associated with Polity
score changes, suggesting that larger states may possess greater institutional
adaptability or reform potential.

The results for non-democratic regimes (defined as those with Polity V scores
below −6) reveal a temporal asymmetry in their effects on democratic outcomes. In
the event-year models (Models 1 and 3), non-democratic regimes are associated
with significant positive changes in Polity V scores. This likely reflects cases
where short-term liberalisation or reform efforts follow leadership crises or
institutional ruptures, producing modest democratic gains even within
authoritarian contexts. By contrast, in the three-year models (Models 2 and 4),
the effect reverses direction: non-democratic regimes are associated with
significant declines in Polity V scores over the medium term. This pattern
suggests that early signs of liberalisation often fail to consolidate and may be
followed by renewed authoritarian entrenchment. In essence, while non-democratic
regimes may exhibit initial democratic openings—whether symbolic or
procedural—these gains are frequently short-lived, with longer-term trajectories
reverting to autocratic norms. This dynamic underscores the fragility of
democratic progress in authoritarian contexts, where reforms introduced in the
aftermath of institutional disruption are often superficial or strategically
instrumental, lacking the structural support required for sustained
democratisation.

Cold War context is statistically significant only in the event-year models,
where it correlates with a decline in Polity V scores, reflecting the broader
global pattern of democratic suppression during the Cold War period.

Political violence and economic growth do not show consistent or significant
effects, indicating that immediate democratic outcomes are more sensitive to
regime characteristics and structural factors than to short-term economic or
security conditions.

Overall, the empirical results offer robust support for both hypotheses.
Autocoups and coups both lead to significant immediate declines in democratic
quality, with coups inflicting greater short-term damage. In the medium term,
coups are often followed by democratic recovery, whereas autocoups result in
persistent democratic erosion with no evidence of rebound.

These findings suggest that autocoups represent a particularly insidious form of
democratic backsliding, less dramatic than coups but ultimately more damaging in
their long-term effects. They reinforce the need for greater scholarly and policy
attention to constitutional manipulations by incumbents, which, although often
gradual and legally framed, can produce lasting democratic decay.

### Robustness tests

To assess the robustness of the main findings, a series of alternative model
specifications were estimated. The results confirm that the core conclusions
remain stable under these variations.

First, the operationalisation of the autocoup variable was modified: the decay
function used in the baseline analysis was replaced with a binary indicator
distinguishing between attempted and successful autocoups. Additionally, the
broad ‘non-democracy’ category was disaggregated into more specific regime
types—military, personalist, presidential, parliamentary, and ‘other’—with
dominant-party regimes serving as the reference category. This classification
mirrors the approach used in the determinants analysis of autocoups presented in
earlier chapters. The results of these robustness models are presented in Models
5 to 8 in @tbl-demomodel2.

```{r}
#| label: tbl-demomodel2
#| tbl-cap: "The Impact of Autocoups on Democratization: Binary Autocoups"
#| results: asis


demo_model1 <- plm(
  Attempted ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Attempted = diff1, Autocoup = autocoup_attemp, Coup = coup_attemp) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)

demo_model2 <- plm(
  Attempted ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Attempted = diff3, Autocoup = autocoup_attemp_lag3, Coup = coup_attemp_lag3) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)

demo_model3 <- plm(
  Succeeded ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Succeeded = diff1, Autocoup = autocoup_succeed, Coup = coup_succeed) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)
demo_model4 <- plm(
  Succeeded ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Succeeded = diff3, Autocoup = autocoup_succeed_lag3, Coup = coup_succeed_lag3) |>
    mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other")
    )
  )
)

stargazer(demo_model1, demo_model2, demo_model3, demo_model4,
  #type = "text",
  float = F,
  header = F,
  column.labels = c("(5)", "(6)", "(7)","(8)"),
  model.numbers = F,
  dep.var.caption = "Dependent variable: Differences of Polity V scores",
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Regime: Military", "\\hspace{1.5cm} Personal", "\\hspace{1.5cm} Presidential", "\\hspace{1.5cm} Parliamentary", "\\hspace{1.5cm} Other", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "20pt",
  # font.size = "footnotesize",
  df = F
)
```

Consistent with the main models, autocoups remain significantly associated with
negative changes in Polity V scores in the short term (Models 5 and 7), with
coefficients of −1.236 and −1.234, respectively (both significant at the $1\%$
level). However, in the three-year models (Models 6 and 8), the effect becomes
statistically insignificant, indicating that the negative effect of autocoups is
immediate but not sustained over time.

By contrast, coups continue to show a distinct “U-shaped” effect. In the
event-year models (Models 5 and 7), coups are associated with significant
declines in Polity V scores (−1.366 and −2.190), both at the $1\%$ level. Yet in
the three-year models (Models 6 and 8), the effect reverses direction: coups are
now associated with large positive changes in Polity V scores (+1.240 and +1.712,
also significant at the $1\%$ level). This confirms the earlier interpretation
that while coups may cause immediate democratic disruption, they are often
followed by democratic recovery in the medium term.

The disaggregated regime type variables provide additional insights. Military
regimes show significant positive effects in the event-year models (Models 5 and
7), with coefficients of +0.602 and +0.574, but become negative and significant
in the three-year models (−0.545 and −0.584 in Models 6 and 8). This reversal
suggests that initial post-event liberalisation in military regimes is not
sustained, and may even regress.

Personalist regimes are consistently associated with negative and significant
effects in the three-year models (Models 6 and 8: −0.532 and −0.526), but not in
the two-year models, suggesting that their democratic erosion becomes more
evident over time.

Presidential and parliamentary democracies follow a similar pattern: both show
significant negative effects in the short term (Models 5 and 7), and positive,
statistically significant effects in the medium term (Models 6 and 8). For
example, parliamentary democracies are associated with a drop of −0.475/−0.468 in
the short term but show a gain of +0.965/0.966 over three years. This pattern
supports the idea that democratic institutions may initially be shaken by
political disruption but recover when institutional mechanisms are strong.

“Other” regimes (likely transitional or provisional systems) show consistently
large and positive effects across all models, ranging from +0.999 to +1.115, all
significant at the $1\%$ level. This implies that these regimes tend to
transition toward more democratic forms over both short and medium time frames.

Several control variables also behave consistently with the baseline models. GDP
per capita is negatively and significantly associated with changes in Polity V
scores across all models, again likely reflecting ceiling effects in advanced
democracies with limited room for improvement. Log of population is positively
and significantly related to Polity changes, reinforcing earlier interpretations
that larger states may possess greater reform potential or be more likely to
register changes in democratic performance. Political violence becomes
statistically significant only in the three-year models (Models 6 and 8), where
it has a small positive effect (+0.044, +0.046), suggesting that prolonged unrest
may precede some form of institutional response or democratic opening. The Cold
War variable is significant only in the event-year models (Models 5 and 7), where
it is associated with small negative effects (−0.168 and −0.156), consistent with
broader patterns of democratic suppression during the Cold War period.

These robustness models confirm the main findings while offering additional
nuance. These results underscore the importance of both regime context and
temporal scope in evaluating the consequences of irregular power grabs.
Autocoups, unlike coups, represent a consistently negative force for democratic
institutions—one that undermines without paving the way for recovery.

## Conclusion

This chapter has examined the impact of autocoups on democratic institutions by
analysing changes in Polity V scores, with a comparative focus on traditional
coups. Two key hypotheses guided the analysis: first, that autocoups are
associated with consistent declines in democratic quality, particularly in the
short term; and second, that while coups often generate initial disruptions, they
tend to produce a “U-shaped” effect, marked by subsequent democratic recovery or
even advancement in the medium term.

The empirical results offer strong support for these hypotheses. Across multiple
model specifications, autocoups—whether attempted or successful—exhibit
significant negative effects on Polity V scores in the event year, but these
effects do not persist into the medium term. In contrast, coups are associated
with significant democratic improvement three years after the event, despite an
initial decline. This pattern is robust across models incorporating disaggregated
regime types, alternative lag structures, and extended time horizons.

The analysis further reveals important variation across regime types. Military
and personalist regimes, while sometimes exhibiting modest democratic gains in
the immediate aftermath, tend to experience declines in Polity V scores over
time, suggesting a return to entrenched authoritarianism. Presidential and
parliamentary democracies, by contrast, initially register democratic decline but
tend to recover within three years—consistent with institutional resilience.
Notably, transitional or provisional regimes (“other” types) display consistently
strong democratic gains, underscoring their potential for reform during periods
of flux.

The findings carry several theoretical and policy-relevant implications. While
coups are widely recognised as pivotal events in the study of regime change,
autocoups deserve greater scholarly attention. Unlike coups, which may at times
catalyse democratic transitions, autocoups represent a systematically
anti-democratic mechanism, typically employed to erode checks on executive power
and extend incumbents’ rule. Moreover, as shown in Chapter 4, autocoup leaders
tend to retain power for longer periods—nearly a decade on average—compared to
less than seven years for coup-installed leaders, implying more durable
institutional consequences.

This chapter also advances the methodological literature by emphasising the
importance of temporal framing. Many political shocks—particularly autocoups—are
preceded by elite purges, electoral manipulation, or institutional weakening.
Consequently, focusing solely on post-event changes risks overlooking the
cumulative nature of democratic decline. The findings thus support a more
longitudinal and process-oriented approach to studying regime erosion.

Nevertheless, limitations remain. Notably, coups and autocoups occasionally occur
in close temporal proximity, making it difficult to disentangle their respective
contributions to changes in Polity V scores. Future research should seek to
better isolate these overlapping effects, perhaps through finer-grained event
sequencing or qualitative case tracing.

In sum, this chapter reinforces the view that autocoups are a critical yet
underexplored driver of democratic backsliding. Their often-subtle execution
belies their long-term consequences. As such, they warrant continued empirical
scrutiny and deeper integration into both the comparative democratisation
literature and policy frameworks concerned with defending constitutional
governance.
