---
format:
  pdf: 
    documentclass: report
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    toc: true
    toc-title: Contents
    lof: true
    lot: true
    crossref: 
      lof-title: Figures
      lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  docx: default
    
categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
include-before-body: title.tex
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue)
```

## Acknowledgements {.unnumbered}

The completion of this thesis marks the culmination of a remarkable journey,
filled with dedication, perseverance, and moments of profound joy. I am deeply
grateful to the numerous individuals who have supported and encouraged me
throughout this endeavour.

I would like to express my sincerest appreciation to my supervisor, Professor
Kristian Skrede Gleditsch, whose guidance, expertise, and unwavering support have
been instrumental in shaping this research. His constructive feedback and
encouragement have been invaluable, and I am deeply indebted to him for his
mentorship.

I am also grateful to Professor Han Dorussen, the chair of my board panel, for
his continuous support and thoughtful input. His insightful comments and
suggestions have significantly enhanced the quality and depth of my research.

I would like to acknowledge the important contributions of my initial
co-supervisors, Dr. Saurabh Pant and Professor David Siroky, who laid a strong
foundation for this work during the early stages of my research. Although they
are no longer at the University of Essex, their instruction and guidance were
instrumental in shaping the direction of this project.

I have been fortunate to receive feedback and guidance from several esteemed
scholars in the field, including Dr. Brian J Phillips, Dr. Prabin Khadka, and Dr.
Winnie Xia. Their expertise and insights have enriched this research, and I am
grateful for their contributions.

On a personal note, I would like to express my deepest gratitude to my family,
who have been a constant source of support and inspiration throughout this
journey. To my beloved wife, Ji Zhi, your patience, love, and encouragement have
been immeasurable. To my dear children, Siyan and Sisheng, your joy and curiosity
have motivated me to persevere and strive for excellence.

I am also deeply grateful to my father for his enduring support and belief in my
abilities. To the cherished memory of my late mother, your love, guidance, and
values continue to shape my path and inspire my endeavors. And to my three
brothers, whose support enabled me to pursue my PhD without worries, I am forever
grateful.

While many individuals have contributed to the success of this work, I take full
responsibility for any errors or shortcomings that may remain.

## Abstract {.unnumbered}

This dissertation presents a comprehensive analysis of irregular power
transitions, focusing on the dynamics of coups and autocoups, and their
implications for leadership survival and democratic stability. By employing a
comparative framework and innovative analytical tools, this study illuminates the
factors driving the success and frequency of these unconstitutional transfers of
power, as well as their consequences for democratic erosion and regime
durability.

The research first investigates the critical role of power dynamics, particularly
the influence of regime types, on the likelihood and success of coup attempts.
Utilizing a double probit model with sample selection, the study reveals that the
expected probability of coup success is a key driver of coup attempts, with
military regimes exhibiting heightened vulnerability due to their inherent power
structures.

Expanding beyond traditional coup analysis, this dissertation explores the
understudied phenomenon of autocoups, where incumbent leaders manipulate
institutional frameworks to extend their mandated terms. To address this
knowledge gap, the study introduces a refined definition of autocoups and
develops a novel dataset spanning from 1945 to 2024. This enables a rigorous
comparative analysis of coups and autocoups, demonstrating the significant impact
of both phenomena on democratic backsliding.

Through advanced survival analysis techniques, the research examines how the
method of power acquisition – whether through coups or autocoups – affects
leadership tenure. The findings indicate that coup-installed leaders typically
face shorter tenures and higher risks of removal compared to those who extend
their rule via autocoups. This stark contrast suggests that autocoups may
incentivize power personalization and contribute to the erosion of democratic
institutions.

This dissertation makes several significant contributions to political science
literature:

-   Conceptual advancement: Provides a refined definition of autocoups,
    addressing a critical gap in existing literature.

-   Data innovation: Introduces a comprehensive dataset on autocoups,
    facilitating more robust quantitative analyses.

-   Analytical framework: Develops a unified approach for studying both coups and
    autocoups, enabling comparative analysis of these pivotal political events.

-   Theoretical insights: Offers a nuanced understanding of the relationship
    between power acquisition methods, leadership survival, and democratic
    resilience.

The study's findings have important implications for scholars and policy-makers
concerned with democratic resilience, regime stability, and the dynamics of
political power. By elucidating the complex interplay between irregular power
transitions and institutional stability, this research contributes to a deeper
understanding of the challenges facing contemporary democracies and hybrid
regimes.

***Keywords:** Coups, Autocoups, Power transitions, Leadership survival,
Democratic resilience*

## Introduction

The study of irregular power transitions is a critical area of research in
political science, as these unconventional transfers of authority can have
significant and lasting impacts on political systems, democratic institutions,
and societal stability. This dissertation aims to provide a comprehensive
analysis of the dynamics of two key forms of irregular power transitions—**coups
and autocoups**—and their influence on leadership survival, regime stability, and
democratic resilience. By examining these phenomena within a unified framework,
this research contributes to a deeper understanding of the complex interplay
between power dynamics, institutional structures, and political outcomes in
contemporary governance.

### **Motivation and Research Questions**

The fundamental question driving this research is: **Why do some leaders face
premature removal from power, while others extend their rule far beyond
constitutionally mandated limits?** This puzzle is particularly relevant in
today’s global context, where democratic backsliding and the rise of
authoritarianism are increasingly apparent. By examining the factors that
influence leadership survival following unconstitutional power transitions, this
study aims to deepen our understanding of the complex relationships between power
transitions, leadership longevity, political stability, and democratic
resilience.

This study specifically addresses the following three research questions in its
main chapters:

-   How do power dynamics and regime types influence the likelihood and success
    of coup attempts?

-   How can autocoups be analysed within a comparative framework alongside
    traditional coups?

-   How does the method of power acquisition (coup vs. autocoup) affect a
    leader’s tenure and risk of removal?

### Comparative Framework: Coups and Autocoups

Coups have traditionally dominated the academic discourse on irregular power
transitions due to their frequency and visibility. The well-established
definition of coups, as the complete removal of incumbent leaders by elites
within the existing power structure, has facilitated the development of
comprehensive datasets and rigorous quantitative analyses.

However, another form of irregular power transition has been largely
overlooked—instances where incumbent leaders refuse to relinquish power and
extend their mandated terms. This phenomenon, which I term **"autocoups"**, has
received comparatively less attention, with a lack of consensus on terminology,
definition, and data.

To address this gap, this study proposes a unified framework for analysing both
coups and autocoups. Viewing these as related but distinct forms of irregular
power transitions enables a more nuanced understanding of the factors that shape
leadership survival following unconstitutional changes in government. This
comparative approach is crucial for several reasons:

-   Both coups and autocoups represent the most frequent means of irregular power
    transition and significantly influence democratic backsliding.

-   The similar nomenclature reflects a fundamental similarity: while a coup aims
    to replace the current leader, an autocoup seeks to prevent the succession of
    a future leader.

-   Integrating these phenomena within a unified analytical framework can
    contribute to a more comprehensive understanding of political stability,
    democratic erosion, and leadership longevity in various regime types.

-   This approach allows for a more holistic examination of the strategies
    employed by political actors to gain, maintain, or extend power outside of
    constitutional norms.

### Research Objectives and Contributions

This study addresses the critical gap in the literature by offering a unified
framework for analysing coups and autocoups. The key contributions are threefold:

-   **Emphasis on Power Dynamics and Regime Types:** The research highlights the
    significant role of power dynamics, particularly the influence of regime
    types, in determining the success and frequency of coup and autocoup
    attempts. This contribution enhances our understanding of how institutional
    structures and power distribution within different regime types shape the
    likelihood and outcomes of irregular power transitions. By focusing on the
    balance of power between incumbents and potential challengers, this study
    provides new insights into the strategic calculations that underpin coup
    attempts and their outcomes.

-   **Refined Definition and Novel Dataset for Autocoups:** The study introduces
    a refined definition of autocoups and develops a novel dataset covering
    events from 1945 to 2024, filling a significant gap in the existing
    literature and enabling a comparative analysis with classic coups. This
    dataset provides a valuable resource for future research on autocoups and
    their impact on political systems. By offering a clear conceptualization of
    autocoups and a new dataset, this study lays the groundwork for more rigorous
    quantitative analyses of this understudied phenomenon.

-   **Survival Analysis of Leaders from Different Entry Modes:** The research
    applies survival analysis to existing coup data and the new autocoup dataset,
    demonstrating how different modes of entry into power significantly affect
    leadership survival. This analysis offers insights into the long-term
    consequences of irregular power transitions on leadership tenure and
    political stability. By comparing the longevity of coup-installed leaders
    versus autocoup leaders, this study provides a nuanced understanding of the
    relationship between power acquisition methods and leadership survival.

These contributions collectively advance the field by providing a more holistic
understanding of irregular power transitions, offering new tools and data for
quantitative analysis of autocoups, and demonstrating the interconnectedness of
power acquisition methods and leadership survival. The findings have important
implications for understanding democratic backsliding, regime stability, and the
dynamics of political power.

### Implications

The examination of irregular power transitions provides a crucial perspective on
the interrelated phenomena of democratic backsliding, breakdown, and autocratic
intensification. This study’s findings offer logical explanations for several
political trends:

-   **Regression of Global Democracy Levels:** This study can explain why global
    democracy levels have regressed to pre-2000 levels
    [@freedomhouse2024freedom], as irregular power transitions inevitably violate
    democratic norms and disrupt the trajectory towards stable democracies.

-   **Within-Regime Democratic Erosion:** The research explains why democratic
    backsliding often occurs within regimes [@mechkova2017], with democracies
    becoming less liberal and autocracies becoming less competitive, particularly
    due to the prevalence of autocoups since 2000 [@bermeo2016].

-   **Prevalence of Autocoups Since 2000:** The analysis reveals that autocoups
    possess several strategic advantages for incumbent leaders: they have a
    significantly higher probability of success, the consequences of failure are
    relatively milder, and leaders who manage to extend their rule through an
    autocoup often enjoy considerably longer tenures compared to those who come
    to power through a traditional coup.

### Overview of the Thesis

This dissertation investigates the complex dynamics of irregular power
transitions and their implications for leadership survival and democratic
processes. The study examines three key aspects: the determinants of classic coup
attempts, the conceptualization and analysis of autocoups, and the impact of
power acquisition methods on leadership longevity.

#### Chapter 2: Determinants of Classic Coup Attempts {.unnumbered}

This chapter addresses the ongoing debate surrounding the determinants of coups
by shifting focus from pre-conditions to the strategic calculations of coup
plotters. While previous research has identified numerous potential predictors,
there remains a lack of consensus on the primary drivers of coup attempts. This
study emphasizes the severe consequences of failed coups as a critical factor for
potential perpetrators.

The chapter argues that coup plotters are unlikely to act unless success is
highly probable. To operationalise this concept, the study leverages regime types
as a proxy for the balance of power between coup perpetrators and incumbent
leaders. This approach is based on the understanding that regime types
fundamentally reflect the distribution of power within ruling groups.

To address the inherent selection bias in coup attempts, the chapter employs a
double probit model with sample selection. This sophisticated methodological
approach allows for a more accurate assessment of the factors influencing both
coup attempts and their likelihood of success. The analysis reveals that expected
success rates significantly influence coup attempts and provides strong evidence
for how the balance of power, shaped by regime types, determines the chances of
coup success and, consequently, coup attempts. A key finding is that military
regimes are substantially more vulnerable to coups than dominant-party regimes,
offering important insights into the relationship between institutional
structures and political stability.

#### Chapter 3: Conceptualising and Analysing Autocoups {.unnumbered}

This chapter addresses a significant gap in the literature by refining the
concept of autocoups and providing a comprehensive dataset for quantitative
analysis. While autocoups have gained increased scholarly attention since 2000,
previous research has been hampered by conceptual ambiguities and a lack of
systematic data.

The chapter begins by addressing two key issues in the existing literature: the
conflation of power expansion and power extension[^1], and the misalignment
between definitions of autocoups and traditional coups. By focusing specifically
on power extensions by incumbent leaders, this study offers a more precise and
theoretically consistent definition of autocoups.

Building on this conceptual clarification, the chapter introduces a novel dataset
covering autocoup events from 1945 to 2024, encompassing 110 attempts and 87
successes. This dataset represents a significant contribution to the field,
enabling more rigorous quantitative analysis of autocoups and their impacts. The
chapter demonstrates the utility of this dataset through case studies and
empirical analyses, highlighting the significance of autocoups in shaping
democratic backsliding and power personalization.

#### Chapter 4: Impact of Power Acquisition Methods on Leadership Longevity {.unnumbered}

Building on the foundations established in the previous chapters, this chapter
presents a comparative analysis of coups and autocoups within a unified
framework. The chapter focuses specifically on the impact of power acquisition
methods on leadership survival, comparing the tenure of leaders who come to power
through coups versus those who extend their rule via autocoups.

Employing advanced statistical techniques, including Cox proportional hazards
models and time-dependent Cox models, the analysis provides a nuanced examination
of how different modes of power acquisition affect leadership longevity. The
findings reveal a significant impact of power acquisition methods on leader
tenure, with autocoup leaders generally enjoying longer tenures than
coup-installed leaders. This insight contributes to our understanding of the
strategic incentives for different forms of irregular power transitions and their
long-term consequences for political stability.

#### Chapter 5: Conclusion and Future Directions {.unnumbered}

The final chapter synthesizes the key findings from each substantive chapter,
weaving together the insights gained from the analysis of coups, autocoups, and
leadership survival. It reflects on the broader implications of these findings
for our understanding of democratic backsliding, power personalization, and
political stability in diverse regime types.

The chapter also acknowledges the limitations of the current study and outlines
promising avenues for future research. It emphasizes the need for continued
exploration of irregular power transitions, particularly in light of evolving
global political dynamics. By highlighting the significance of this research for
understanding contemporary challenges to democratic governance, the conclusion
underscores the broader relevance of the study's findings for both scholars and
policy-makers concerned with promoting political stability and democratic
resilience.

## Power Dynamics and Coup Attempts: A Selection Mechanism Analysis {#sec-chapter2}

### Abstract {.unnumbered}

While existing literature on coup attempts has identified over a hundred
potential determinants, a consensus on their relative importance remains elusive.
This study proposes a novel approach by prioritizing determinants based on their
impact on coup success, positing that the expected outcomes of coups are a
critical factor in their occurrence. Employing a double probit model with sample
selection, I investigate the relationship between regime types and coup attempts,
with a focus on internal power dynamics. The main findings confirm that regime
type is a crucial determinant of coup likelihood. Military and personalist
regimes exhibit significantly higher susceptibility to coups, attributed to their
weaker institutional frameworks and increased vulnerability.

***Keywords:** Coups, Power transitions, Regime types,* *Sample selection*

\newpage

### Introduction

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| include: false
#load coup data

Coup <- fread("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt")

attempts <- length(Coup$country)
succeeded <- length(Coup$country[Coup$coup == 2])

coups <- Coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )



Coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

Coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

Coups d'état, defined as "illegal and overt attempts by the military or other
elites within the state apparatus to unseat the sitting executive" [@powell2011],
represent a critical threat to constitutional power transitions and political
stability. The frequency and success of these attempts vary significantly across
countries and regions, prompting essential questions about **why coups are more
prevalent in certain contexts and why some coups succeed while others fail**.

For instance, according to the Global Instances of Coups dataset (GIC)
[@powell2011], Latin America has seen notable variation in coup activity: Bolivia
experienced `r coups |> filter(country == "Bolivia") |> pull(2)` coups between
1950 and 1984, while Argentina saw
`r coups |> filter(country == "Argentina") |> pull(2)` coups during a similar
period. In Africa, Sudan endured
`r coups |> filter(country == "Sudan") |> pull(2)` coups between 1955 and 2023.
In stark contrast, countries such as Mexico and South Africa have remained
coup-free since 1950.

Despite decades of scholarly inquiry and the identification of over 100 potential
determinants [@gassebner2016], a consensus on the key factors driving coup
attempts and their outcomes remains elusive. The proliferation of variables
presents a significant challenge: How can we establish a framework that
prioritizes the most relevant factors, rather than sifting through an
ever-expanding list of possibilities?

This study proposes a different perspective on understanding coup dynamics by
shifting the focus from pre-coup conditions to the strategic calculations of
potential coup plotters. Central to this approach is the consideration of the
expected probability of coup success as a key determinant of coup initiation.
Several observations support this perspective:

-   **The high stakes of coups**: Failed attempts often lead to severe
    consequences for the perpetrators, including imprisonment, exile, or death.

-   **Selective coup attempts**: Despite 491 recorded attempts since 1950, these
    represent only about 4% of over 12,000 country-years during the same period
    (GIC).

-   **Relatively high success rates**: Approximately half of all coup attempts
    succeed, suggesting that plotters carefully select opportunities based on
    their likelihood of success (GIC).

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts


coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(99)
              ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*")) |> 
  as_raw_html()
```

Given the difficulty in directly observing the probability of coup success, this
study proposes using regime type as a crucial proxy for predicting coup outcomes.
The underlying premise is that the balance of power within a regime, which is
significantly influenced by its type, ultimately shapes the outcomes of coups. By
analysing power dynamics across various regime types, we can gain more profound
insights into the structural factors that affect both the likelihood of coup
attempts and their success.

To address the inherent selection bias in studying coup attempts, this research
employs a double probit model with sample selection, building on the
methodological approaches used by @powell2012 and @böhmelt2014. This method
enables the simultaneous analysis of factors influencing both the initiation and
success of coups.

This study aims to contribute to scholarly discourse and practical efforts in two
key ways:

-   **Emphasizing the expected chances of success**: By focusing on the
    probability of success as a driver of coup attempts, this research offers a
    more targeted approach to understanding coup dynamics, specifically the
    selective effect of anticipated outcomes on coup initiation.

-   **Highlighting the significance of regime type**: In the absence of perfect
    knowledge of internal power balances, this study leverages regime types as a
    proxy for power structures, shedding light on how these structures influence
    anticipated coup outcomes.

The remainder of this chapter is organized as follows: Section 2 explores the
dynamics of coup attempts and their outcomes. Section 3 outlines the research
design, methodology, and variables. Section 4 presents and discusses the
empirical findings. Section 5 concludes with key insights and their implications
for understanding and potentially mitigating coup risks.

### Dynamics of coup attempts and outcomes

Coup attempts represent a critical juncture in the political landscape of
regimes, driven by a complex interplay of factors. These factors can be broadly
categorized into two main components: the **disposition** of potential
challengers (their motivations and willingness to act) and their **capability**
(the resources and opportunities available to them). This analysis explores the
intricate dynamics of coup attempts, examining the underlying motivations,
factors influencing their success, and the role of regime types in shaping a
country's susceptibility to coups.

#### Motivations for coups

The decision to undertake a coup is rarely made lightly, given the high stakes
and potential consequences. Three primary categories of motivations emerge from
the literature:

-   **Personal Ambition:** At its core, the allure of absolute power serves as a
    potent motivator for many coup plotters. The prospect of seizing control
    offers the ability to shape national policies without constraint, control
    over significant resources and wealth, the power to make impactful decisions
    that affect millions, prestige and recognition on both national and
    international stages, and the opportunity to leave a lasting historical
    legacy.

-   **Purported National Interest:** Coup leaders often justify their actions as
    necessary interventions for the greater good of the nation. While such claims
    should be scrutinized, some examples demonstrate genuine attempts to address
    critical issues: Resolving constitutional crises, facilitating transitions to
    democracy, or addressing severe economic downturns or social unrest. A
    notable example is the 2010 coup in Niger, which ousted President Tandja
    after he attempted to secure an unconstitutional third term by dissolving the
    opposing court and calling a self-serving referendum [@ginsburg2019].

-   **Self-Preservation:** In some instances, coups serve as pre-emptive strikes
    against perceived existential threats. Motivations in this category include:
    Preventing elimination or political persecution by incumbent leaders,
    protecting the interests of specific military or political factions,
    safeguarding ideological or ethnic groups from marginalization. The 1971 coup
    led by Idi Amin against Ugandan President Obote exemplifies this motivation,
    as Amin acted to prevent his removal from a key military command position
    [@sudduth2017].

Despite these potential motivations, it's crucial to note that coups remain
relatively uncommon events. Since 1950, coups have occurred in only about 4% of
country-years. This rarity underscores the importance of the capability factor –
even the most motivated actors require substantial resources and favourable
circumstances to successfully execute a coup.

#### Capability for coups

While motivations provide the impetus for coup attempts, the decision to actually
launch a coup hinges on a calculated assessment of the chances of success. This
capability assessment is crucial in determining whether potential coup plotters
move from contemplation to action. Several key factors influence this assessment:

-   **Military Strength**: A clear advantage in military capabilities compared to
    the incumbent regime significantly increases the odds of a successful coup
    [@powell2018; @choulis2022]. This factor encompasses: Control over elite
    military units or special forces, access to advanced weaponry and technology,
    loyalty of key military commanders, and strategic positioning of supportive
    military units. The balance of military power is often the most critical
    determinant in coup success, as it directly affects the ability to seize and
    hold key government installations.

-   **Internal Divisions within the Regime**: Existing fractures within the
    government's power structure can be exploited by coup plotters to gain
    support from disgruntled factions. These divisions may manifest as:
    Ideological disagreements among ruling elites, competing centers of power
    within the government, ethnic or regional tensions in the political
    leadership, or dissatisfaction among mid-level bureaucrats or military
    officers. Coup plotters can leverage these divisions to build a coalition of
    support, promising benefits or addressing grievances of marginalized factions
    within the regime.

-   **Public Support**: Widespread discontent with the incumbent leaders,
    especially within the military or key sectors of society, can create a ripe
    environment for a successful coup. Factors contributing to public support
    include: Economic hardship or inequality, perceived corruption or
    mismanagement by the government, human rights abuses or political repression,
    or failure to address critical national issues. While public support alone
    may not be sufficient for coup success, it can provide legitimacy to the coup
    plotters and reduce resistance from the general population.

-   **Foreign Backing**: External support from powerful nations can provide
    resources, legitimacy, and even direct military intervention to tip the
    scales in favor of the coup plotters. This support may come in various forms:
    Financial assistance Intelligence sharing, diplomatic recognition, covert
    military aid or training, or threats of intervention against the incumbent
    regime. The role of foreign backing in coups has been particularly
    significant during the Cold War era and continues to shape geopolitical
    dynamics in many regions.

-   **Timing and Opportunity**: The ability to identify and exploit moments of
    vulnerability in the incumbent leaders is crucial for coup success. Opportune
    moments may include: National crises or emergencies, periods of political
    transition or uncertainty, major public events or celebrations, and times of
    internal conflict within the regime. Coup plotters must carefully assess
    these windows of opportunity and time their actions to maximize their chances
    of success.

#### The Selection Bias in Coup Analysis

While historical data might suggest a high success rate for coups, it is crucial
to consider selection bias in interpreting this information. We only observe
attempted coups, not the numerous plots that never materialize. This creates a
significant challenge in accurately assessing the true likelihood of coup
success.

Several factors contribute to this selection bias:

-   **Unobserved Deterrence**: Incumbent regimes may implement preventive
    measures that deter potential coups before they reach the planning stage.

-   **Self-Selection of Capable Plotters**: Those who attempt coups are likely to
    do so only when they believe they have a reasonable chance of success,
    leading to an overestimation of the overall success rate.

-   **Unreported Failed Attempts**: Many unsuccessful coup attempts, particularly
    those in the early planning stages, may go unreported or undetected.

-   **Varying Definitions of "Coup Attempt"**: Inconsistencies in how researchers
    define and classify coup attempts can further skew the data.

To address these challenges and understand coup attempts comprehensively, we need
to develop theoretical frameworks that account for this selection bias. A
frequently cited framework [@gassebner2016; @aidt2019] offers a structured
approach to assess the disposition and capability of coup attempts by evaluating
the anticipated benefits for coup plotters.

The expected pay-off of a coup can be represented by the equation:

$$
\begin{aligned}
E(U) = p \times B + (1 - p) \times (-C)
\end{aligned}
$$ {#eq-eq1}

Where:

-   $E(U)$**:** Expected utility or pay-off of the coup attempt;
-   $B$ represents the return of a successful coup;
-   $C$ signifies the cost of a failed coup;
-   $p$ represents the probability of coup success.

The condition for staging a coup is when the expected benefit is positive
($E(U) > 0$). Rearranging the equation, we get:

$$
\begin{aligned}
p \times B > (1 - p) \times C
\end{aligned}
$$ {#eq-eq2}

This implies that for a coup to be attempted, the expected benefits of success
must outweigh the expected costs of failure.

However, quantifying $B$ and $C$ is inherently difficult due to several factors.
The intangible nature of some costs and benefits (e.g., loss of life, personal
freedom). The variability of outcomes across different contexts. The subjective
valuation of power and risk by individual coup plotters.

Given the difficulty in precisely quantifying $B$ and $C$, we can treat them as
roughly equal for analytical purposes. This allows us to shift our focus to the
probability of success ($p$). The simplified equation becomes:

$$
\begin{aligned}
p > (1-p)
\end{aligned}
$$ {#eq-eq3}

This suggests that a success probability greater than 50% is necessary for a coup
to be attempted. While empirical data shows a slightly lower overall success rate
for coups since 1950 (49.9%, as shown in @tbl-coups), it is crucial to remember
that this is an average and does not reflect the specific probabilities assessed
by coup plotters beforehand.

Based on this discussion, we propose the following hypothesis:

> **H2-1: *The fundamental determinant of a coup attempt is the perceived chance
> of success. Coup plotters likely require a success threshold of at least
> 50%.***

This hypothesis shifts the focus of coup research from the prerequisites for
triggering a coup to the expected coup outcome as the fundamental determinant. It
emphasizes the importance of understanding the factors that influence coup
plotters' perceptions of success probability.

The next logical question is: What factors determine coup success and
consequently influence the decision to attempt one? The answer lies in
understanding the balance of power inherent in the different power dynamics
across various regime types.

#### Regime Types and Coup Susceptibility

The historical success rates of coups do not dictate the outcomes of individual
coup attempts, as each faces unique circumstances and conditions. Coup plotters
assess their chances based on their specific context. While numerous factors
influence a coup's success, the decisive element remains the relative strength
between incumbents and coup perpetrators, particularly in terms of military
power.

However, viewing military strength as the sole determinant of coup attempts
oversimplifies the complex reality of these events [@singh2016]. The clandestine
nature of coups necessitates small, secretive groups, making it challenging to
gauge the stance of other factions within the military. As @geddes1999 notes, a
coup's success often hinges on the reactions of these other factions, creating a
complex decision-making environment for potential plotters. Moreover, incumbents
frequently employ strategic division within the military as a coup-proofing
measure. @böhmelt2014 describe how leaders intentionally create rival groups
within the armed forces, establishing an artificial balance and structural
obstacles to coup attempts. Additionally, factors beyond military force, such as
internal divisions among ruling elites, public support, and foreign backing,
significantly shape the balance of power.

Consequently, a more holistic approach involves analysing the overall balance of
power within a political system. However, as noted in the introduction of this
chapter, directly observing this balance presents significant challenges for
outside observers. An alternative approach is to analyze the factors that are
decisive in shaping power dynamics. This method aligns more closely with academic
research goals, which focus on understanding the patterns and underlying factors
of coup attempts rather than identifying specific power challengers capable of
launching successful coups.

Among the various factors influencing power dynamics, regime type emerges as one
of the most crucial. Regime types are fundamentally classifications based on
power structures within a political system, and different regime types allocate
authority differently for critical decisions such as the deployment of military
forces, appointment of key officials and military officers, and formulation and
implementation of main policies.

@geddes2014 provide a classification of autocratic regime types based on their
power structures. This classification offers valuable insights into the power
dynamics within different autocracies and their relative susceptibility to coups.
The three main types of autocratic regimes, their characteristics, and their
vulnerability to coup attempts are explored as follows (@tbl-regimes1):

-   **Military Regimes:** Characterized by a junta – a group of military officers
    controlling leadership selection and policy formulation. Examples include
    regimes in Brazil (1964-1985), Argentina (1976-1983), and El Salvador
    (1948-1984) [@geddes1999].

    Despite their military nature, these regimes are surprisingly unstable due to
    internal power struggles within the junta. The absence of a clear final
    authority and the presence of multiple military factions increase the
    likelihood of resorting to force to resolve disputes, making these regimes
    the most vulnerable to coups.

-   **Personalist Regimes:** Power is concentrated in a single, charismatic
    leader who controls the military, policy, and succession. Examples include
    Rafael Trujillo's regime in the Dominican Republic (1930-1961), Idi Amin's
    regime in Uganda (1971-1979), and Jean-Bédel Bokassa's regime in the Central
    African Republic (1966-1979) [@geddes1999].

    Personalist regimes are relatively stable during the leader's tenure.
    However, they face a higher risk of coups due to unclear succession plans and
    vulnerabilities associated with the leader's personal weaknesses, health, and
    mortality.

-   **Dominant-Party Regimes:** Power resides within a well-organized ruling
    party, with leaders acting as its representatives. The party structure and
    ideology foster internal cohesion and a long-term vision. Examples include
    the Partido Revolucionario Institutional (PRI) in Mexico, the Revolutionary
    Party of Tanzania (CCM), and Leninist parties in various Eastern European
    countries [@geddes1999].

    Dominant-party regimes exhibit the greatest resilience against coups due to
    their institutionalized structures, unified leadership, clear ideology, and
    internal discipline.

Empirical data supports this theoretical framework. While military regimes
represent only 5.6% of country-years since 1950, they experience a
disproportionate share of coups (over 22%). Personalist regimes, constituting 13%
of country-years, account for 23% of coups. Conversely, dominant-party regimes,
representing 22.6% of country-years, account for only 16.7% of coups
(@tbl-regimes). These statistics demonstrate the disproportionate vulnerability
of military and personalist regimes to coups, while highlighting the relative
stability of dominant-party regimes.

The distinct power dynamics exhibited by these regime types significantly
influence their susceptibility to coups. This analysis leads to our second
hypothesis:

The susceptibility to coups varies significantly among different types of
autocratic regimes, with military regimes being the most vulnerable, followed by
personalist regimes, and dominant-party regimes being the least susceptible.

> **H2-2: *The susceptibility to coups varies significantly among different types
> of autocratic regimes, with military regimes being the most vulnerable,
> followed by personalist regimes, and dominant-party regimes being the least
> susceptible.***

\blandscape

```{r}
#| label: tbl-regimes1
#| tbl-cap: Main features of different types of regimes 

read.csv(here("data/regimes_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center",
              page.orientation = "landscape"
              ) |>
  cols_width(
    1:6 ~ px(100)
  ) |>
  tab_source_note(source_note = md("*Source: GWF & Author*")) |> 
  as_raw_html()
```

\clearpage

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and coups since 1950 


regimes <- REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949)

coup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/coup_model.csv")

regime_total <- coup_model |>
  select(regime_type = regime) |> 
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(country_year =n,
         share = percent(percent),
         .keep = "unused") 
  

coup1 <- Coup |>
  select(country, ccode, year, coup)

coup_by_regime <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup)) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  mutate(coups = n,
         `coups(%)` = percent(percent,accuracy = 0.1),
         .keep = "unused") 

coup_by_succeed <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup),
         coup == 2) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  select(regime, succeeded_coup = n) 

coup_regimes <-  coup_by_regime |> 
  left_join(coup_by_succeed) |> 
  mutate(success_rate = percent(succeeded_coup / coups, accuracy = 0.1))

regime_total |>
  left_join(coup_regimes, join_by(regime_type == regime)) |>
  mutate(regime_type = str_to_title(regime_type),
         likelihood = percent(coups / country_year, accuracy = 0.1)) |>
  select(-succeeded_coup) |> 
  clean_names(case = "title") |>
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN and GIC Datasets*")) |> 
  as_raw_html()
```

\elandscape

### Research Design

#### Double probit with sample selection model {#sec-coup231}

This study employs a sophisticated statistical approach to account for the
selective nature of coup attempts. While coup attempt rates vary across regimes,
success rates tend to be surprisingly consistent, hovering around 50% (as shown
in @tbl-regimes). This suggests that coup attempts are not random acts, but
rather strategically planned and undertaken only when the odds of success appear
favorable. A standard statistical model would not account for this selectivity,
potentially leading to biased results.

To address this issue, we utilize a double probit with sample selection model,
similar to the approach used by @powell2012 and @böhmelt2014. This model, known
as a Heckman probit model or bivariate probit model with sample selection,
consists of two parts:

-   **Selection Equation (Stage 1)**: This stage analyses the factors influencing
    whether a coup attempt occurs in a particular country-year.

-   **Outcome Equation (Stage 2)**: This stage focuses on the probability of
    success for those coup attempts that actually take place.

The selection equation (first stage) models the probability that a coup attempt
occurs:

$$
\begin{aligned}
y_{1i}^*&=\alpha_0 + \alpha_1 Regime_i + \mathbf{X}_i \boldsymbol{A} + \mu_{1i}
\\
\\
y_{1i} &= 
\begin{cases} 
1 &\text{if $y_{1i}^*>0$ (coup attempt occurs)} \\
\\
0 &\text{if $y_{1i}^*\le0$ (no coup attempt)}
\end{cases}
\end{aligned}
$$ {#eq-eq4}

The outcome equation (second stage) models the probability that a coup attempt is
successful, given that it occurs:

$$
\begin{aligned}
y_{2i}^*&=\beta_0 + \beta_1 Regime_i + \mathbf{Z}_i \boldsymbol{B} + \mu_{2i}
\\
\\
y_{2i} &= 
\begin{cases} 
1 &\text{if $y_{2i}^*>0$ (coup succeeds)} \\
\\
0 &\text{if $y_{2i}^*\le0$ (coup failes)}
\end{cases}
\end{aligned}
$$ {#eq-eq5}

Where:

-   $y_{1i}^*$ *and* $y_{2i}^*$ are latent variables

-   $\text{Regime}_i$ is a categorical variable (military, personalist, or
    dominant-party)

-   $\mathbf{X}_i$ and $\mathbf{Z}_i$ are vectors of control variables

-   $\mu_{1i}$ and $\mu_{2i}$ are error terms, assumed to follow a bivariate
    normal distribution with correlation $\rho$

The model assumes:

$$
\binom{\mu_{1 i}}{\mu_{2 i}} \sim N\left(\binom{0}{0},\left(\begin{array}{ll}
1 & \rho \\
\rho & 1
\end{array}\right)\right)
$$

The probability equations are:

$$
\begin{aligned}
P\left(y_{1 i}=1\right) & =\Phi\left(\alpha_0+\alpha_1 \operatorname{Regime}_i+\mathbf{X}_i \boldsymbol{A}\right)
\end{aligned}
$$ {#eq-eq7}

$$
\begin{aligned}
P\left(y_{2 i}=1 \mid y_{1 i}=1\right) & =\Phi\left(\beta_0+\beta_1 \operatorname{Regime}_i+\mathbf{Z}_i \boldsymbol{B}\right)
\end{aligned}
$$ {#eq-eq8}

Where $\Phi(\cdot)$ is the cumulative distribution function of the standard
normal distribution.

#### Variables

##### Dependent variable

The analysis utilizes data on coup attempts and outcomes from @powell2011. A
successful coup is defined as one where the incumbent leader is removed from
power for more than seven days. The dataset covers the period from 1950 to 2023
and includes information on 491 coup attempts, with roughly half (245) being
successful. Descriptive statistics for these coup attempts and regime types can
be found in @tbl-coups and @tbl-regimes.

-   **Coup Attempt**: Binary variable indicating whether a coup attempt
    occurred (1) or not (0) in a given country-year.

-   **Coup Success**: Binary variable indicating whether a coup attempt was
    successful (1) or failed (0), conditional on a coup attempt occurring.

##### Key Independent Variable: Regime Type {#sec-chapter2322}

I categorize regime types following @geddes2014 (GWF), focusing on military,
personalist, and dominant-party regimes, with democracies and monarchies included
for comparison. Descriptive statistics for regime types are presented in
@tbl-regimes.

##### Control variables {.unnumbered}

The control variables are chosen based on the research of @gassebner2016. They
analysed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

```{r}
#| label: variables
#| results: asis
#| echo: false
gdppc <- "-   **Economic Level:** Represented by GDP per capita. This measure provides an
    indication of the overall economic health and standard of living in a
    country. We use GDP per capita data (in constant 2017 international 1000
    dollars, PPP) from the V-Dem dataset by @fariss2022."

ct <- "-   **Economic Performance:** Measured using the current-trend ($CT$) ratio
    developed by @krishnarajan2019. This ratio compares a country's current GDP
    per capita to the average GDP per capita over the previous five years. A
    higher $CT$ ratio indicates stronger economic performance. For a country $i$
    at year $t$, the $CT$ ratio is calculated as follows:"

glue(gdppc)
glue(ct)
```

$$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
$$ {#eq-eq6}

-   **Political Stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    variable “actotal” in the Major Episodes of Political Violence dataset
    [@marshall2005current], with 0 representing the most stable conditions (no
    violence at all) and 18 representing the most unstable.
-   **Previous coups:** Included in the selection equation as either: a) The
    number of previous coups in a country (Model 1), or b) The time since the
    last coup attempt (Model 2 for robustness check).

### Results and Discussion

```{r}
#| label: tbl-coupmodel
#| tbl-cap: "Sample Selection Model of Regime Type and Coup Success, 1950-2019"
#| results: asis


model_1 <- coup_model|>
  filter(regime != "other") |>
  select(coup, regime, GDP_trend, GDP_pc, violence, coup_dummy = pre_coups) |> 
  mutate(regime = fct_relevel(regime, "dominant-party")) |> 
  selection(coup > 0 ~ regime + GDP_trend + GDP_pc + violence + coup_dummy,
            coup > 1 ~ regime + GDP_trend + GDP_pc + violence,
            #method = "2step",
            data = _) 

model_1_selection <- model_1
model_1_selection$param$index$betaO <- model_1$param$index$betaS
model_1_selection$param$index$betaS <- model_1$param$index$betaO

model_2 <- coup_model |>
  filter(regime != "other") |>
  select(Coup = coup, regime, GDP_trend, GDP_pc, violence, coup_dummy, time_since_last_coup) |> 
  mutate(regime = fct_relevel(regime, "dominant-party")) |> 
  selection(Coup > 0 ~ regime + GDP_trend + GDP_pc + violence + coup_dummy*time_since_last_coup,
            Coup > 1 ~ regime + GDP_trend + GDP_pc + violence,
            data = _) 

model_2_selection <- model_2
model_2_selection$param$index$betaO <- model_2$param$index$betaS
model_2_selection$param$index$betaS <- model_2$param$index$betaO

stargazer(model_1_selection, model_1, model_2_selection, model_2,
          #type = "text",
          float = F,
          header = F,
          dep.var.caption = "",
          dep.var.labels = c("Model 1", "Model 2"),
          multicolumn = T,
          dep.var.labels.include = T, 
          column.labels = c("Coup Attempts", "Coup Outcome","Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant", "Regime: Democracy", "\\hspace{1.6cm}Military", "\\hspace{1.6cm}Monarchy", "\\hspace{1.6cm}Personalist", "Economic trend", "GDP per capita", "Political violence", "Previous coups (P)", "Yrs since coup (Y)", "Interaction term: P * Y"),
          summary.logical = F,
          column.sep.width = "7pt",
          font.size = "small"
          )
```

The double probit model with sample selection, estimated using the
***sampleSelection*** package [@sampleSelection-2] in R, provides valuable
insights into the factors influencing coup attempts and their outcomes across
different regime types from 1950 to 2019 (@tbl-coupmodel). I present two models
that differ slightly in their treatment of previous coups: Model 1 incorporates
the number of previous coups, while Model 2 utilizes the time elapsed since the
last coup.

#### Selection Model: Coup Attempts

In the selection model (Column 1), military and personalist regimes exhibit
significant positive coefficients at the 1% level, indicating a higher likelihood
of experiencing coup attempts compared to dominant-party regimes. This aligns
with our theoretical expectations regarding internal power struggles within
military juntas and succession vulnerabilities in personalist regimes.

```{r}
#| label: tbl-mfx1
#| tbl-cap: Average marginal effects of coup attempts (Selection of Model 1)
#| echo: false

avg_comparisons(model_1) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_1, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups",
                          term == "GDP_trend" ~ "Economic trend",
                          term == "GDP_pc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Democracy",
                          str_detect(contrast, "military") ~  "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_options(
    table.width = pct(99)
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  ) |> 
  as_raw_html()
```

@tbl-mfx1 clarifies the regime effects using Average Marginal Effects (AME) and
ratio percentages. The military regime's marginal effect of 0.07 indicates that
the probability of coup attempts in military regimes is 7 percentage points (pp)
higher than in dominant-party regimes, ceteris paribus. This translates to
military regimes being about 277.7% more likely to encounter coups than
dominant-party regimes. Similarly, personalist regimes show a 2.4 pp higher
probability, about 94% more likely compared to dominant-party regimes.

Control variables show effects in expected directions but with weaker magnitudes.
Stronger economic performance, indicated by higher economic growth trends and GDP
per capita levels, correlates with a lower risk of coup attempts. Political
violence shows a positive effect, indicating that higher levels of instability
increase the likelihood of coups. The positive coefficient for the number of
previous coups suggests a "copycat" effect from earlier incidents.

#### Outcome Model: Coup Success

The outcome model (Columns 2 and 4 in @tbl-coupmodel) reveals determinants of
coup success. Military regimes demonstrate a higher probability of coup success
compared to dominant-party regimes, aligning with expectations that military
regimes face higher coup risks due to their increased chances of success.
Personalist and monarchical regimes show slight positive effects on coup success,
but these effects are not statistically significant.

Control variables exhibit different patterns in the outcome model compared to the
selection model. Both GDP per capita and political violence maintain a weak
influence, similar to their effects in the selection model. However, the economic
trend shows a less significant negative effect on coup success.

#### Model Comparison (Model 1 vs Model 2)

Model 2 employs years since the last coup instead of the number of previous
coups, with an interaction term between previous coups (as a binary variable) and
years since the last coup. Generally, Model 2 shows results in the same direction
as Model 1, albeit with relatively lower coefficients (@tbl-mfx2).

The differences between Model 1 and Model 2 suggest that while the recency of
coups matters, the overall history of coups in a country may have a stronger
influence on future coup attempts.

```{r}
#| label: tbl-mfx2
#| tbl-cap: Average marginal effects of coup attempts (Selection of Model 2)
#| echo: false


avg_comparisons(model_2) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_2, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups (P)",
                          term == "time_since_last_coup" ~ "Yrs since coup (Y)" ,
                          term == "coup_dummy*time_since_last_coup" ~ "P * Y",
                          term == "GDP_trend" ~ "Economic trend",
                          term == "GDP_pc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Democracy",
                          str_detect(contrast, "military") ~ "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups (P)",
  "Yrs since coup (Y)",
   "P * Y"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_options(
    table.width = pct(99)
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  ) |> 
  as_raw_html()
```

#### Discussion of key findings

The $\rho$ values of 0.898 in Model 1 and 0.386 in Model 2, significant at 1% and
10% levels respectively, indicate strong correlation between unobserved factors
influencing coup attempts and coup success. This supports the appropriateness of
the sample selection model and underscores the importance of considering both
stages in the analysis.

The significant coefficients with theoretically consistent directions suggest the
model effectively captures key aspects of coup dynamics. The observed disparity
between coup attempt rates and success rates across regimes points towards
selection bias, further validating the use of the sample selection model.

#### Theoretical Implications

These results strongly support my theoretical framework, highlighting the crucial
role of regime structure in determining coup vulnerability. The findings
underscore that coups are strategic actions undertaken when odds appear
favourable, rather than random events.

### Conclusion

This study addresses the lack of consensus in empirical research on coup
predictors by introducing a novel approach that prioritizes determinants based on
their impact on coup success. By analysing coup success rates, I posit that the
expected outcomes of coups are critical determinants of their occurrence.
Employing a double probit model with sample selection, the research investigates
and confirms a strong and robust relationship between regime types and coup
attempts.

The main findings reveal that regime type plays a pivotal role in the likelihood
of coup attempts. Military and personalist regimes, characterized by weaker
institutional frameworks and higher vulnerability, demonstrate significantly
higher susceptibility to coups compared to dominant-party regimes. This
underscores the importance of supporting initiatives that establish regimes with
constitutional institutions rather than those dependent on military power or
personal authority, as the latter prove more volatile and coup-prone.

The research also indicates that stronger economic performance, while not as
influential as regime type, is associated with a lower risk of coups. This
suggests that policies promoting economic development can be effective in
reducing coup risk. Given that regime type is often determined during regime
formation and is difficult to alter, promoting economic growth might be the most
viable coup-proofing strategy available to incumbents.

However, these findings present a paradox for autocratic leaders, particularly
dictators and military juntas. While institutionalization of regimes could
enhance stability and reduce coup risk, few such leaders are willing to implement
these reforms. This reluctance stems from the potential constraints on their
power or shortened terms that such changes might entail. Thus, while institutions
may benefit the regime's longevity, they do not necessarily serve the immediate
interests of individual leaders.

This study opens avenues for further research, including:

-   Investigating the long-term effects of regime types on political stability,
    including democratization or personalization.

-   Examining how international factors might influence regime choices and coup
    dynamics when internal factors are too weak to prevent coups.

In conclusion, this chapter provides robust evidence for the critical role of
regime type in determining coup risk, while also acknowledging the complex
realities of political power. By illuminating the strategic calculus behind coup
attempts, it offers valuable insights for scholars seeking to understand
political instability and policymakers working to promote democratic transitions
and political stability in diverse global contexts.

## Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders {#sec-chapter3}

### Abstract {.unnumbered}

This study aims to clarify the concept of autocoups, specifically focusing on
power extensions by incumbent leaders. By distinguishing autocoups from the
broader and more ambiguous concepts of self-coups or executive takeovers, which
encompass both executive power aggrandizement and power extension, this research
redefines the concept of autocoups. Based on this refined definition, I introduce
a novel dataset of autocoup events from 1945 to 2024. Using this newly compiled
dataset, the research includes three types of case studies that provide
qualitative insights into the dynamics of autocoups. Additionally, an empirical
analysis of the determinants of autocoup attempts and success is offered to
demonstrate how the autocoup dataset can be employed for more quantitative
research. This study contributes to the existing literature by providing a
clearer conceptual framework and a novel dataset of autocoups. It enhances our
understanding of the mechanisms and motivations behind power extensions by
incumbent leaders and examines the implications for democratic backsliding,
democratic breakdown, and autocratic deterioration. The insights gained from this
study could draw more attention to the effects of autocoups on power transitions,
political stability, and democratic resilience.

***keywords**: Autocoups, Coups, Power transitions*

\newpage

### Introduction

The study of irregular power transitions has predominantly focused on coups due
to their frequency and significant impact. This emphasis has led to substantial
scholarly attention and the creation of comprehensive datasets. However, as noted
by @gassebner2016, the multitude of potential factors proposed to explain coups
has often resulted in increased confusion rather than a clearer understanding of
coup dynamics.

In contrast, another form of irregular power transition—the incumbent leader's
refusal to relinquish power—has received relatively less attention despite its
significance. Recent decades have witnessed a decline in traditional coups and a
rise in this "incumbent retention or overstay" type of irregular power
transition, particularly since the end of the Cold War (@ginsburg2010evasion;
@baturo2014; @versteeg2020law).

This chapter will redefine and clarify this type of irregular power transition,
where leaders overstay their mandated term limits, as "autocoup." Although
analyses related to autocoups are not rare, the existing literature exhibits
several notable shortcomings:

-   **Terminological Ambiguity:** Terms such as self-coups, autocoups,
    autogolpes, incumbent takeovers, executive aggrandizement, overstay, and
    continuismo are used without a clear, universally accepted definition. This
    lack of standardization leads to confusion and inconsistent application
    [@marsteintredet2019; @baturo2022].

-   **Limited Data:** Due to the conceptual ambiguity surrounding autocoups, data
    collection remains in its early stages compared to the rich datasets
    available for traditional coups.

-   **Methodological Gaps:** Research on autocoups has largely relied on case
    studies [@cameron1998; @antonio2021; @pion-berlin2022], with few studies
    employing quantitative analysis.

Moreover, analyses of autocoups are often not integrated with those of
traditional coups. As a distinct category of coup, autocoups lack a corresponding
definition in relation to classic coups. Consequently, traditional coups and
autocoups are frequently analysed separately, despite their related nature, and
there is a paucity of comparative analyses.

Studying autocoups is important for several reasons. Firstly, autocoups can
undermine the rule of law, weaken institutions, and contribute to democratic
backsliding or authoritarian personalization. Secondly, like traditional coups,
successful autocoups increase the likelihood of future irregular power
transitions. For example, since 1945, approximately 62% of leaders who extended
their terms through autocoups in non-democratic countries were either ousted or
assassinated while in office [@baturo2019]. Thirdly, failed autocoups often lead
to instability, inciting protests, violence, and even civil wars.

This chapter aims to address these gaps by focusing on autocoups, aiming to
clarify terminology, refine concepts and definitions, enhance data collection,
and explore determinants through empirical analysis, contributing in three key
areas:

-   **Conceptual Clarification:** The term "autocoup" will be redefined and
    clarified, with a focus on power extension.

-   **Data Collection:** A new dataset of autocoups since 1945 will be introduced
    based on this refined definition.

-   **Empirical Analysis:** Utilizing this dataset, a quantitative analysis of
    the factors influencing leaders' decisions to attempt autocoups will be
    conducted.

The structure of this chapter is as follows: Section 2 will review definitions
related to power expansions and extensions, leading to a precise definition of
autocoups. Section 3 will present the new autocoup dataset. Sections 4 and 5 will
explore the determinants of autocoup attempts through case studies and
demonstrate the application of the dataset in empirical analysis. The conclusion
will summarize key findings and suggest directions for future research.

Definitions of power expansion and power extension in political science can often
be ambiguous or overlapping. To ensure clarity in this study, we propose distinct
definitions for these concepts:

-   Power Expansion: This refers to the process by which an incumbent leader
    acquires additional authority or control over state apparatuses beyond their
    original mandate. This may involve centralizing power, reducing checks and
    balances, or assuming roles typically held by other branches of government.

-   Power Extension: This concept specifically relates to the temporal aspect of
    an incumbent's rule. It describes situations where a leader prolongs their
    tenure beyond the originally mandated term in office, often through
    constitutional amendments, cancellation of elections, or other means of
    circumventing term limits.

By distinguishing between these two concepts, we aim to provide a more nuanced
analysis of how leaders consolidate and maintain their positions. While power
expansion and extension can often occur simultaneously, differentiating between
them allows for a more precise examination of the strategies employed by leaders
to entrench their rule.

This distinction is crucial for our study as it enables us to: 1. Analyze the
different mechanisms leaders use to solidify their power 2. Examine the potential
consequences of each type of power consolidation 3. Explore the relationship
between power expansion and extension in various political contexts

Understanding these nuances is essential for comprehending the complex dynamics
of political power and leadership longevity in both democratic and authoritarian
systems."

### Autocoups: A literature review and clarification of definitions

The concept of autocoups, or the illegitimate seizure of power by incumbent
leaders, remains a complex and contested area of study. Unlike coups, which are
generally understood as illegal attempts by elites to overthrow the government,
autocoups lack a clear and consistent definition. This ambiguity hinders our
ability to accurately identify, analyze, and compare instances of this
phenomenon.

#### Terminology

The most common term in autocoup literature is "self-coup," or "autogolpe" in
Spanish [@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017;
@marsteintredet2019]. This term gained academic prominence after Peruvian
President Alberto Fujimori dissolved Congress, temporarily suspended the
constitution, and ruled by decree in 1992 [@mauceri1995; @cameron1998]. However,
as @marsteintredet2019 point out, the term "self-coup" can be misleading, as it
implies a coup against oneself, which is inaccurate since it targets other state
institutions or apparatus.

Another approach to describe coups staged by incumbents is to use terms with
adjectives or modifiers, such as "presidential coup," "executive coup,"
"constitutional coup," "electoral coup," "judicial coup," "slow-motion coup,"
"soft coup," and "parliamentary coup" [@marsteintredet2019]. While these terms
can be helpful in specific contexts, their proliferation often adds to the
overall confusion rather than providing clarification. Most of these terms focus
on the specific methods used by coup perpetrators but fail to clearly identify
the perpetrator, necessitating further explanation. In fact, many of these
methods could be employed either by or against executive leaders.

A third alternative involves terms like "incumbent takeover," "executive
takeover," or "overstay." Incumbent takeover refers to "an event perpetuated by a
ruling executive that significantly reduces the formal and/or informal
constraints on his/her power" [@baturo2022, p.374], based on earlier research
[@svolik2014]. Meanwhile, overstay is defined as "staying longer than the maximum
term as it stood when the candidate originally came into office"
[@ginsburg2011evasion, p.1844]. These terms identify the perpetrator (the
incumbent) and/or the nature of the event (overstaying/extending power). However,
they do not highlight the illegality or illegitimacy of these actions. Therefore,
they cannot serve as a direct counterpart to "coup," which clearly denotes the
illegality of leadership ousters, while "takeover" or "overstay" diminish the
severity.

As these terms often lack precision, focusing on specific methods rather than the
core act of power usurpation, this study proposes "autocoup" as the most suitable
term for this phenomenon. Unlike other terms, 'autocoup' clearly identifies the
perpetrator and the illegitimate nature of the power grab, distinguishing it from
traditional coups.

#### Definition

While terminology is important, another issue arises with the previous definition
of autocoups: What is the emphasis of an autocoup—power expansion, power
extension, or both?

Definitions of power expansion and power extension in political science can often
be ambiguous or overlapping. To ensure clarity in the study of autocoups, I
propose distinct definitions for these concepts:

-   **Power Expansion:** This refers to the process by which an incumbent leader
    acquires additional authority or control over state apparatuses beyond their
    original mandate. This may involve centralizing power, reducing checks and
    balances, or encroaching on the authority of other branches like the
    legislature or judiciary.

-   **Power Extension:** This describes situations where a leader prolongs their
    tenure beyond the originally mandated term in office, often through
    constitutional amendments, cancellation of elections, or other means of
    circumventing term limits.

Existing definitions of autocoups or related concepts often either are ambiguous
between power expansion and extension, or focus more on power expansion, which
has several drawbacks.

Firstly, defining autocoups primarily in terms of power expansion does not align
well with the definition of a coup. The focus of a classical coup is clearly on
the ouster of the current leader, not merely a limitation or restriction on their
power. Using the same logic, a more appropriate definition of an autocoup should
prioritize the tenure extension of executive leadership. Power restriction on
incumbents would not be coded as a coup as long as they remain in office.
Similarly, an executive leader acquiring more power from other branches could be
coded as power aggrandizement, but not an autocoup, as long as they step down
when their term expires.

Secondly, emphasizing power expansion in autocoups often neglects the ultimate
purpose of incumbents. It is irrational for an incumbent to expand executive
power and then pass the powerful role to future leaders. Although the term
"self-coup" gained prominence from the 1992 Fujimori case in Peru, which
initially involved seizing power from other institutions, it is important to note
that Fujimori ultimately extended his term limits through constitutional
amendments. The 1993 Constitution allowed Fujimori to run for a second term,
which he won with popularity in April 1995. Shortly after Fujimori began his
second term, his supporters in Congress passed a law of "authentic
interpretation" which effectively allowed him to run for another term in 2000,
which he won amid suspicions and rumors. However, he did not survive the third
term. In 2000, facing charges of corruption and human rights abuses, Fujimori
fled Peru and took refuge in Japan [@ezrow2019].

Thirdly, measuring the extent of power expansion to qualify as an autocoup can be
challenging. As @cameron1998a defined, a self-coup is "a temporary suspension of
the constitution and dissolution of congress by the executive, who rules by
decree until new legislative elections and a referendum can be held to ratify a
political system with broader executive power" (p. 220). "Broader executive
power" is difficult to define, and it would be problematic and disputable no
matter how it is defined.

Therefore, this study argues that a more accurate definition of autocoups should
prioritize power extension as the core characteristic. It is easy to identify the
event and the outcome in the first place. In most cases, autocoups in terms of
power extension involve power expansions as their prerequisite and foreshadowing.

Based on these criteria, I define **an autocoup as the illegitimate extension of
an incumbent leader's term in office beyond the originally mandated limits
through unconstitutional means**. This definition emphasizes the core
characteristic of power extension while acknowledging the potential for power
expansion as a related phenomenon. Three key points need to be highlighted for
this definition.

Firstly, this definition refers to the actual leaders of the country, regardless
of their official titles. Typically, this would be the president; however, in
some cases, such as in Germany, the primary leader is the premier, as the
president is a nominal head of state.

Secondly, while the primary characteristic of an autocoup is extending the term
in office, this definition does not exclude instances of power expansion. Both
aspects can coexist, but the extension of the term is the central element.

Thirdly, autocoups, by their nature, subvert legal norms and established power
transfer mechanisms. No matter how legitimate they claim to be, their
illegitimacy is not beyond a reasonable doubt as long as the incumbents are the
direct beneficiaries. This critical aspect will be explored further in Section 3.

### Introduction to the Autocoup Dataset

#### Defining the scope

Categorizing political events as autocoups inevitably involves challenging
borderline cases. To maintain consistency and avoid ambiguity, this study adopts
a broad coding approach: All instances of incumbents extending their original
mandated term in office are coded as autocoups, regardless of the apparent
legality of the extension.

This approach is justified because truly legitimate amendments to power
transition institutions should apply only to subsequent leaders, not the
incumbent. Even when extension procedures appear legal, the legitimacy is
questionable when the incumbent is the direct beneficiary.

#### Classifying autocoups {#sec-classify}

Autocoups manifest in various forms. I categorize them based on several key
factors:

-   **Methods Employed**: Specific strategies used by incumbents (e.g.,
    constitutional amendments, election cancellation).

-   **Degree of Legality**: Extent of deviation from established legal norms.

-   **Duration of Extension**: Length of time the incumbent remains in office
    beyond designated term limits.

-   **Outcomes**: Whether the autocoup attempt succeeds or fails.

This study primarily focuses on the methods employed, while coding for other
aspects when information is available.

##### Evasion of term limits {.unnumbered}

Evasion of term limits is a common tactic employed in autocoups. Incumbents often
resort to seemingly legal manoeuvres to extend their hold on power. These
manoeuvres primarily involve manipulating constitutional provisions through
various means. The incumbents may pressure legislative bodies (congress) or
judicial institutions (Supreme Court) to reinterpret existing term limits, amend
the constitution to extend terms, or even replace the constitution altogether.
This might also involve popular vote through referendums, or a combination of
these approaches. The extension can range from a single term to indefinite rule.

These manoeuvres primarily involve manipulating constitutional provisions through
various means.

-   **Changing Term Length:** Incumbents might lengthen the official term
    duration (e.g., from 4 to 6 years) to stay in office longer, even if the
    number of allowed terms remains unchanged. Examples, in the dataset, include
    Presidents Dacko (CAR, 1962), Kayibanda (Rwanda, 1973), and Pinochet (Chile,
    1988).

-   **Enabling re-election:** This approach involves incumbents modifying legal
    or constitutional frameworks to permit themselves to run for leadership
    again, despite initial restrictions. These restrictions might include
    prohibitions on re-election, bans on immediate re-election, or term limits
    that the incumbents have already reached. An illustrative example is
    President Menem of Argentina in 1993, who leveraged this tactic to extend his
    tenure.

-   **Removing Term Limits Altogether:** This approach, as seen with President Xi
    Jinping of China in 2018, technically allows the leader to rule for life,
    although they may still need to participate in elections (a formality in such
    cases).

-   **Declaring** **Leader for Life:** This differs from removing term limits as
    the leader still faces elections (although potentially rigged or
    uncontested). An example is Indonesia's President Sukarno, who attempted to
    declare himself president for life in 1963 (ultimately unsuccessful).

These methods are often used in combination. Initially, the duration of a term is
extended, followed by amendments to allow re-election, then the removal of term
limits, and finally, the declaration of the leader for life. For example, Haitian
President François Duvalier amended the constitution in 1961 to permit immediate
re-election and then declared himself president for life in 1964.

##### Election Manipulation or Rigging {.unnumbered}

Election manipulation or rigging is the second most commonly used tactic to
extend an incumbent's tenure.

-   **Delaying or Removing Elections:** Delaying or removing scheduled elections
    without legitimate justification is a frequent method used by incumbents to
    maintain power. For instance, Chadian President François Tombalbaye delayed
    general elections until 1969 after assuming power in 1960. Similarly, Angolan
    President José Eduardo dos Santos suspended elections throughout his rule
    from 1979 to 2017.

-   **Refusing Unfavourable Election Results:** Incumbents may refuse to accept
    unfavourable election results and attempt to overturn them through
    illegitimate means. For example, President Donald Trump of the United States
    refused to accept the results of the 2020 election and tried to overturn
    them.

-   **Rigging Elections:** Winning elections with an extraordinarily high
    percentage of votes is highly questionable. This study will code elections
    where the incumbent wins more than 90% of the vote as autocoups. For
    instance, President Teodoro Obiang of Equatorial Guinea has consistently won
    elections with over 95% of the vote in multi-party elections since 1996,
    indicating election rigging.

-   **Excluding Opposition in Elections:** Manipulating the electoral process by
    excluding opposition parties or candidates from participation, effectively
    creating a one-candidate race, clearly signifies an autocoup.

##### **Use of Figurehead** {.unnumbered}

To circumvent term limits, some incumbents might choose a close associate to act
as a figurehead, taking the office publicly while the incumbent retains real
power behind the scenes. This can be achieved through seemingly subordinate
positions.

One example is Russia in 2008. Facing term limits, President Putin selected
Dmitry Medvedev to run for president. After the election, Medvedev appointed
Putin as Prime Minister. However, most analysts believe Putin wielded the true
power throughout this period.

##### Reassigning supreme authority to a new role {.unnumbered}

This tactic involves an incumbent leader manipulating the constitution or legal
framework to create a new position of power, or elevate an existing one, before
stepping down from their current role. They then strategically take on this new
position, effectively retaining significant control despite appearing to
relinquish power. For example, in 2017, Recep Tayyip Erdoğan, the Prime Minister
of Turkey, spearheaded a constitutional referendum that transitioned the country
from a parliamentary system to a presidential one. This new system concentrated
significant executive power in the presidency. Following the referendum's
approval, Erdoğan successfully ran for the newly established presidency,
effectively retaining control under a different title.

##### One-Time Arrangement for Current Leaders {.unnumbered}

This strategy involves special arrangements that extend the term or tenure of
current leaders without altering the underlying institutions. For example,
Lebanon extended President Émile Lahoud's term by three years in 2004 through a
one-time arrangement.

#### Data Coding

The autocoup dataset is built upon existing studies and datasets, ensuring a
comprehensive and reliable foundation. @tbl-source outlines the main sources used
for coding the autocoup dataset.

The Archigos dataset [@goemans2009] and the Political Leaders' Affiliation
Database (PLAD) [@bomprezzi2024wedded] provide comprehensive data on all leaders
from 1875 to 2023, although our coding only includes autocoups since 1945. These
datasets are invaluable for identifying actual rulers, distinguishing them from
nominal heads of state.

The Incumbent Takeover dataset [@baturo2022] integrates data from 11 related
datasets, offering a broad spectrum of cases where leaders significantly reduced
constraints on their power. This dataset includes both power expansions and
extensions, necessitating cross-referencing with Archigos to verify
qualifications for autocoups.

```{r}
#| label: tbl-source
#| tbl-cap: Main Data Sources for Coding the Autocoup Dataset
#| echo: false
#| results: asis

read.csv(here("data/autocoup_source.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(
    table.width = pct(100),
    table.align = "center") |>
  as_raw_html()
```

```{r}
#| label: data-autocoups
#| echo: false

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))
```

In total, `r nrow(Autocoup)` observations were coded, with
`r nrow(Autocoup |> filter(!is.na(source)))` overlapping with the candidate data
from Incumbent Takeover. The remaining
`r nrow(Autocoup) - nrow(Autocoup |> filter(!is.na(source)))` events were newly
coded by the author through verification with other sources such as Archigos,
PLAD and news reports.

The main deviation from the Incumbent Takeover dataset arises from excluding
power expansions that do not involve attempts to extend tenure.

The dataset encompasses a total of 14 variables along with the *notes* field.

-   **Country Identification:** Country code (*ccode*) and country name
    (*country*) from Correlates of War project [@stinnett2002].

-   **Leader Information:** Name of the de facto leader (*leader_name*,coded
    following Archigos and PLAD datasets).

-   **Timeline Variables:** Date the leader assumed power (*entry_date*), date
    the leader left office(*exit_date*), date of the significant event marking
    the autocoup (*autocoup_date*), and Start date of the leader's additional
    term acquired through the autocoup (*extending_date*).

-   **Power Transition Methods:** Categorical variable for how the leader entered
    power (*entry_method*), categorical variable for how the leader exited power
    (*exit_method*), dummy variable indicating regular (1) or irregular (0) entry
    (*entry_regular*), and dummy variable indicating regular (1) or irregular (0)
    exit (*exit_regular*).

-   **Autocoup Details:** Key variable capturing methods used to extend power
    (*autocoup_method*) and outcome of the autocoup attempt (*autocoup_outcome*,
    "fail and lose power", "fail but complete original tenure", or "successful").
    For successful coups, the additional term length can be calculated from the
    difference between *exit_date* and *extending_date*.

-   **Data Source:** Identifies the dataset source used for coding (*source*).

-   **Additional Notes:** Provides context for exceptional cases (*notes*).

There are a few coding challenges and decisions worth mention. For cases where
extensions happen incrementally, the *autocoup_date* reflects a significant event
marking the extension, such as a legislative vote or successful referendum. In
cases where a leader undertook multiple autocoup attempts, details are recorded
in the notes field. Care was taken to differentiate between cases of power
expansion and actual attempts to extend tenure, which required cross-referencing
multiple sources. Determining the success or failure of an autocoup attempt often
required in-depth research, especially for less documented cases.

#### Data descriptions

```{r}
#| label: data_descriptions

country <- Autocoup |> 
  distinct(ccode) |> 
  summarise(count = n())

autocoup_desc <- Autocoup |> 
  summarise(attempted = n(),
            succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method) |> 
  arrange(-attempted)
```

The primary coding has identified `r nrow(Autocoup)` autocoup cases from 1945 to
2024, involving `r country$count` countries. This comprehensive dataset provides
a rich source of information for analysing trends and patterns in autocoup
attempts across different political contexts.

@tbl-autocoup_method presents a breakdown of the autocoup methods employed by
leaders:

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup methods and success rates (1945-2021)

autocoup_desc |> 
  adorn_totals() |> 
  mutate(Success_rate = percent(round(succeeded/attempted, 3))) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(99)
    ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: Autocoup dataset*")) |> 
  as_raw_html()
```

The most common autocoup method is "enabling re-election", accounting for
`r autocoup_desc$attempted[1]` events. This is followed by "removing term limits
altogether" (`r autocoup_desc$attempted[2]` cases), and then "delaying elections"
and "declaring the leader for life" (each with `r autocoup_desc$attempted[3]`
cases).

The overall success rate of autocoups is 79%, which is significantly higher than
the approximately 50% success rate of classical coups. This high success rate can
be attributed to several factors:

-   Incumbent Advantage: Leaders already in power have access to resources and
    institutional mechanisms that can be leveraged to their advantage.

-   Gradual Implementation: Unlike sudden coups, autocoups can be implemented
    gradually, allowing leaders to build support and legitimacy over time.

-   Legal Facade: Many autocoup methods operate within a veneer of legality,
    making them harder to oppose openly.

-   Control of State Apparatus: Incumbents often have significant control over
    state institutions, which can be used to facilitate their autocoup attempts.

However, success rates vary significantly across different methods.

-   **100% Success Rate**: Removing term limits, delaying elections, declaring
    the leader for life, and cancelling elections all have perfect success rates.
    This suggests that once these processes are set in motion, they are difficult
    to reverse.

-   **Lower Success Rates:** Refusing to accept election results has the lowest
    success rate, with only 1 out of 4 attempts succeeding. Although the sample
    size is limited (only 4 cases in total), this trend might suggest several
    factors at play. These include greater democratic resilience in systems where
    general elections are regularly held, heightened international scrutiny and
    pressure in response to blatant manipulation of election results, and
    stronger domestic opposition to such overt power grabs.

### Determinants of Autocoup Attempts: Case Studies

#### High Frequency and Success Rate of Autocoups in Post-Communist Countries

Analysis of our dataset reveals a notably high frequency and success rate of
autocoups in post-communist countries. These nations, formerly communist regimes
prior to the collapse of the Soviet Union, have largely evolved into 'hybrid
regimes' [@nurumov2019], with only a few retaining their communist status. The
data documents 12 cases of autocoups aimed at prolonging incumbency in these
countries, with only two attempts failing. Examination of these cases highlights
several distinctive characteristics:

-   **Inherited Authoritarian Systems**: Despite most of these 12 countries
    transitioning from communist to non-communist governments (with the exception
    of China), they retained many authoritarian systems from their communist
    past.

-   **Continuity of Former Elites**: The transitions did not result in the
    removal or overthrow of previous ruling groups. Instead, former communist
    elites often maintained their positions of power.

-   **Subverted Democratic Processes**: While general elections and term limits
    were introduced in most of these countries, the legacy of former communist
    regimes frequently led to the circumvention of term limits and manipulation
    of elections [@nurumov2019].

##### Case 1: Lifelong Ruler–Alexander Lukashenko in Belarus {.unnumbered}

Alexander Lukashenko, a former member of the Supreme Soviet of the Byelorussian
Soviet Socialist Republic, became the head of the interim anti-corruption
committee of the Supreme Council of Belarus following the dissolution of the
Soviet Union. Elected as Belarus's first president in 1994, he has maintained
this position ever since. Initially, the 1994 constitution limited presidents to
two successive terms. However, Lukashenko removed this restriction in 2004.
International monitors have not regarded Belarusian elections as free and fair
since his initial victory. Despite significant protests, Lukashenko has
consistently claimed to win with a high vote share, often exceeding 80% in each
election. This pattern is evident across all five Central Asian countries of the
former Soviet Union, where post-dissolution leaders were typically high officials
or heads of the former Soviet republics who continued their leadership in the
presidency.

##### Case 2: Transferring Power to a Handpicked Successor–Nursultan Nazarbayev in Kazakhstan {.unnumbered}

Nursultan Nazarbayev served as the first president of Kazakhstan from 1991 until
2019. Prior to the dissolution of the Soviet Union, he held de facto leadership
as the First Secretary of the Communist Party of Kazakhstan. Following
independence, he was elected as the first president and retained office until
2019 through various means, including resetting term limits due to the
implementation of new constitutions. Notably, Nazarbayev did not officially
eliminate term limits but instead created an exemption for the "First President"
[@nurumov2019]. Unlike Lukashenko, who remains the incumbent of Belarus,
Nazarbayev transferred the presidency to a designated successor, Kassym-Jomart
Tokayev, in 2019. However, he retained significant influence as the Chairman of
the Security Council of Kazakhstan until 2022.

```{r}
#| label: latin
#| include: false

library(countrycode)
latin <- codelist |> 
  filter(region == "Latin America & Caribbean") |> 
  select(region, cown) |>
  drop_na() |> 
  pull(cown)

Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = country) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = autocoup_method) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = c(autocoup_method,autocoup_outcome)) |>
  adorn_totals()
```

#### Autocoups for immediate re-election: Cases of Latin America

Latin America has a long-standing tradition of maintaining term limit
conventions. Simón Bolívar, the founding father of Bolivia, was initially a
strong advocate for term limits, stating in 1819, "Nothing is as dangerous as
allowing the same citizen to remain in power for a long time... That's the origin
of usurpation and tyranny" [@ginsburg2019, p.38]. Although Bolívar eventually
modified his stance, arguing in his 1826 Constitution Assembly speech that "a
president for life with the right to choose the successor is the most sublime
inspiration for the republican order," term limits became a convention in Latin
America. Approximately 81% of Latin American constitutions between independence
and 1985 imposed some form of term limits on the presidency
[@marsteintredet2019a].

An analysis of cases in Latin American countries reveals two notable patterns.

##### Often Successful at Breaking Non-re-election or Non-immediate Re-election Restrictions {.unnumbered}

Unlike other presidential systems where two terms are more common,
non-re-election or non-immediate re-election used to be prevalent in Latin
America. According to @marsteintredet2019a, non-consecutive re-election was
mandated in about 64.9% of all constitutions between independence and 1985, while
5.9% banned re-election entirely.

However, adherence to these conventions has varied across the region. Since
Mexico introduced non-re-election institutions in 1911 at the start of the
Mexican Revolution, they have remained inviolate [@klesner2019]. Similarly,
Panama and Uruguay have never altered their re-election rules, and Costa Rica has
only experienced a brief period (1897-1913) permitting immediate presidential
re-election since prohibiting it in 1859 [@marsteintredet2019a]. In many other
countries, however, constitutions have been frequently amended or violated.

The pursuit of re-election or consecutive re-election, therefore, has been a
significant trigger for autocoups aimed at power extension in this region. Our
research documents 32 autocoup cases, with over 50% (17 cases) attempting to
enable re-election or immediate re-election, and about 59% (10 cases out of 17)
being successful.

Unlike those who attempt to overstay in office indefinitely, many Latin American
leaders exit after their second term expires. Examples include President Fernando
Henrique Cardoso of Brazil (1995-2003), President Danilo Medina of the Dominican
Republic (2012-2020), and President Juan Orlando Hernández of Honduras
(2014-2022) [@ginsburg2019; @marsteintredet2019a; @landau2019; @baturo2019;
@neto2019].

##### Failing to Further Extend Tenure {.unnumbered}

This trend does not imply that none of these leaders attempted further
extensions, but rather that most accepted their unsuccessful outcomes without
abusing their power to manipulate the process. While autocoups aimed at securing
one additional term are often successful, attempts to overstay beyond this are
frequently unsuccessful.

Two contrasting cases illustrate the varied outcomes of term limit challenges:

-   **Unsuccessful Extension – Carlos Menem (Argentina)**: President Menem
    successfully extended his tenure by one term through a 1994 constitutional
    amendment allowing one executive re-election. He was subsequently re-elected
    in 1995. However, his attempt to reset his term count, arguing that his first
    term (1988-1995) should not count as it was under previous constitutions, was
    unanimously rejected by the Supreme Court in March 1999 [@llanos2019]. A
    similar scenario unfolded with President Álvaro Uribe of Colombia (2002-2010)
    [@baturo2019].

-   **Successful Extension – Daniel Ortega (Nicaragua)**: In contrast, Daniel
    Ortega, the incumbent president of Nicaragua, successfully extended his
    presidency. In 2009, the Supreme Court of Justice of Nicaragua permitted his
    re-election in 2011. Subsequently, in 2014, the National Assembly of
    Nicaragua approved constitutional amendments abolishing presidential term
    limits, allowing Ortega to run for an unlimited number of five-year terms. As
    a result, he has held the presidency since 2007 [@close2019].

#### As common as classical coups: Cases of African countries

```{r}
#| label: africa
#| include: false

africa <- codelist |> 
  filter(continent == "Africa") |> 
  select(continent, cown) |>
  drop_na() |> 
  pull(cown)

Autocoup |>
  filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = country) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% africa) |>
  distinct(country)
Autocoup |>
  filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = autocoup_method) |>
  adorn_totals()
Autocoup |>
  #filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = autocoup_outcome) |>
  adorn_totals()
```

Classical coups have been prevalent in Africa, accounting for approximately 45%
of all global coups (219 out of 491 cases) since 1950, involving 45 out of 54
African countries (GIC dataset). While autocoups are less frequent compared to
traditional coups, they maintain a significant presence in Africa. Among 110
documented autocoup cases globally, 46% (51 cases) occurred in Africa, involving
36 countries. Notably, the success rate of autocoups in Africa is over 84% (43
out of 51 attempts), which surpasses both the success rate of classical coups in
the region (roughly 50%) and the global average success rate of autocoups (79%).

Identifying a clear pattern of autocoups in Africa is challenging, mirroring the
complexity observed with classical coups. Various factors have been proposed to
explain this phenomenon:

-   **Natural Resources**: Countries rich in natural resources, particularly oil
    or diamonds, may see leaders more likely to attempt and succeed in extending
    their terms [@posner; @cheeseman2015; @cheeseman2019a].

-   **Quality of Democracy**: The quality of democracy is a critical factor
    influencing respect for term limits [@reyntjens2016].

-   **International Influence**: International aid or donor influence can play a
    significant role in discouraging attempts at power extension [@brown2001;
    @tangri2010].

-   **Organized Opposition and Party Unity**: The extent of organized opposition
    and the president's ability to enforce unity within the ruling party are
    crucial factors [@cheeseman2019].

Utilizing the Africa Executive Term Limits (AETL) dataset, @cassani2020
highlights human rights abuses and the desire for impunity as main drivers for
incumbents to cling to power. The more authoritarian a leader, the more likely
they are to attempt to break term limits and overstay in office. A leader's
ability to secure the loyalty of the armed forces through public investment
increases the chances of success in overstaying.

Despite both coups and autocoups being prevalent, there has been a noticeable
shift since the end of the Cold War in 1991: Traditional coups have decreased in
frequency while autocoups have become more prevalent.

This trend can be partially attributed to the introduction of multi-party
elections in Africa in the 1990s, which also brought in term limits for
executives [@cassani2020; @cheeseman2019]. Before 1991, personal or military rule
was more common, and term limits were less frequent. Post-1991, with more term
limits introduced, challenges to these limits have increased. However, it is
crucial to note that this increase in challenges does not necessarily imply that
violations are more common than adherence to term limits, because total power
transitions have increased compared to the past.

### Empirical Analysis: An Example of Utilizing the Autocoup Dataset

The availability of the autocoup dataset has made it feasible to conduct
quantitative analyses that extend beyond traditional case studies. This section
provides a straightforward example of how to utilize this dataset effectively. To
analyze the determinants of autocoup attempts, I employ a probit regression
model. This approach differs from the double probit model with sample selection
used in [Chapter @sec-chapter2] for coup attempts and success analyses. Instead,
I use two separate probit models. Due to the high probability of success in
autocoups, they do not exhibit the typical sample selection characteristics that
necessitated the use of a sample selection model in our earlier analysis of
traditional coups.

#### Dependent Variables

-   **Autocoup Attempt**: Binary variable indicating whether an autocoup attempt
    occurred (1) or not (0) during the tenure of an incumbent leader.

-   **Autocoup Success**: Binary variable indicating whether an autocoup attempt
    was successful (1) or failed (0), conditional on an autocoup attempt
    occurring.

#### Independent Variables

The selection of independent variables are consistent with the coup analysis in
[Chapter @sec-chapter2], plus the population size and the leader's age.

-   **Population Size:** To account for its potential impact on leaders' tenures,
    we consider the log of the population size. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on power transitions. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's Age:** The age of the leader is included as an additional variable
    in the analysis, offering insights into potential correlations with
    leadership strength. Older leaders may have different experiences, networks,
    and health considerations that could influence their ability to maintain
    power. This data is sourced from Archigos and PLAD datasets.

Unlike the analysis of coup determinants, which could theoretically occur in any
given year, I assume that an autocoup happens only once during an incumbent
leader's tenure, as a successful autocoup negates the need for another attempt.
However, this assumption does not always reflect reality, as leaders might
attempt further extensions or try again after a failed attempt. For simplicity, I
overlook these possibilities in our analysis.

Therefore, in our probit model, the unit of analysis for autocoups is the entire
tenure of a leader, rather than a country-year. I establish a base year for the
variables: for leaders who staged an autocoup, we use the year of their first
attempt as the base year; for leaders who did not attempt to overstay, I use the
middle year of their tenure as the base year.

#### Results and discussions

```{r}
#| label: tbl-autocoupmodel
#| tbl-cap: "Determinants of autocoup attempts and success (1945-2018)"
#| results: asis
#| echo: false

autocoup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_model.csv")


model_attempt <- autocoup_model |> 
  filter(!regime %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential")) |>
  glm(autocoup_attempt ~ regime + GDP_pc + GDP_trend + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

model_success <- autocoup_model |> 
  filter(!regime  %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential"),
         autocoup_outcome = factor(autocoup_outcome)) |>
  filter(autocoup_attempt == 1) |> 
  glm(autocoup_outcome ~ regime + GDP_pc + GDP_trend + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

stargazer(model_attempt, model_success, 
          header = F,
          float = F,
          column.labels = c("Autocoup Attempts", "Autocoup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant",
                               "Regime: Dominant-party",
                               "\\hspace{1.6cm}Military",
                               "\\hspace{1.6cm}Personalist",
                               "GDP per capita",
                               "Economic trend",
                               "Political stability",
                               "Age",
                               "Population(log)"),
          dep.var.caption = "",
          dep.var.labels.include = F,
          summary.logical = T,
          column.sep.width = "50pt"
          )
#view(autocoup_model)
```

@tbl-autocoupmodel summarizes the findings from the probit regression models
based on our analysis of the determinants of autocoup attempts and their success.

Model 1, which examines autocoup attempts, reveals only one significant predictor
besides the constant term. Among the regime types, personalist regimes
significantly increase the likelihood of autocoup attempts, all else being equal.
This suggests that leaders in personalist regimes are more prone to attempt to
extend their power through autocoups compared to leaders in democratic regimes
(reference regime). Leaders in dominant-party and military regimes, however, show
no significant difference in the likelihood of attempting an autocoup compared to
democratic leaders.

The model for autocoup success (Model 2) shows similar dynamics. Personalist
regimes again have a strong positive and significant effect on the success of
autocoups compared to democratic leaders. Dominant-party regimes also show a
positive and marginally significant effect. However, a detailed examination
reveals that about half of the successful autocoups in dominant-party regimes (9
out of 20) exhibit a personalist style, such as "party-personal-military"
regimes.

This outcome is logical since personalist leaders are typically much more
powerful than other types of leaders, making them more inclined and capable of
overstaying in power.

Other factors play an insignificant role in determining the attempts and outcomes
of autocoups. This aligns with our conclusions on the determinants of classic
coups. Both coups and autocoups are significantly affected by power dynamics. As
power transitions involve the struggle between seizing and maintaining power, the
balance of power status quo inevitably matters in both coups and autocoups. This
also explains the high success rate of autocoups. Compared to power challengers,
incumbents are in an obviously advantageous position. Incumbent leaders can use
state power to their benefit, which is difficult to counteract. Even the abuse of
power is often unchecked under a powerful leader's rule.

The empirical analysis of autocoups yields significant implications for
real-world politics. In particular, the high overall success rate of autocoups
highlights the vulnerability of democratic institutions to gradual erosion by
incumbent leaders. The threshold for ousting or impeaching an incumbent leader
through constitutional means is exceptionally high, with success often requiring
more than a simple majority and substantial support across various sectors.
Resorting to illegal means, such as a coup, presents even greater challenges due
to high costs, severe consequences, and a low likelihood of success.

Conversely, political dynamics, whether in democracies or autocracies, tend to
favour incumbents even when they act unconstitutionally. Incumbents can leverage
state resources to achieve their political ambitions, benefiting from a high
probability of success and minimal consequences in case of failure. This
asymmetry in power and risk creates a concerning scenario: for incumbents who do
not respect constitutional institutions, the opportunity to launch an autocoup
appears sufficiently low-risk to warrant an attempt.

### Conclusion

This chapter presents a comprehensive analysis of autocoups, focusing
specifically on political events where incumbent leaders illegitimately extend
their tenure in power. By refining the concept and distinguishing it from broader
definitions such as 'self-coups', 'autogolpes', and 'executive takeovers', I
introduce a novel dataset of autocoups spanning from 1945 to 2024. Through this
more precise definition and dataset, this research expands the analysis of
irregular power dynamics from irregular transition to irregular retention,
offering a broader and more nuanced review of the phenomenon.

The findings reveal that personalist regimes are significantly more likely to
experience autocoup attempts and succeed in these attempts compared to
democracies. Dominant-party systems, often exhibiting personalist
characteristics, also show an association with successful autocoups. While regime
type significantly influences autocoups, other factors appear less impactful,
mirroring classic coups where the balance of power is a more essential
determinant. The high success rate of autocoups can be attributed to the inherent
advantages incumbents possess, such as control over or abuse of state power and
the difficulty of removing or impeaching them through legal or illegal means.

However, several limitations warrant consideration for future research. Firstly,
the definition of an autocoup requires further commentary and discussion to gain
wider acceptance in the academic community. Despite efforts to maintain
objectivity, some coding decisions may involve subjective judgements,
particularly in borderline cases. Secondly, due to the nature of autocoups, which
are less frequent than classic coups (491 coups versus 110 autocoups during the
same period), the quantitative analysis cannot be conducted as a country-year
variable as in coup studies. This raises the issue of choosing an appropriate
base year. For instance, when analysing how GDP level or growth rate, political
stability, or the age of the leader affects autocoup attempts, we must decide
which year's value should be used. In this study, I chose the middle year of a
leader's tenure or the year they staged the autocoup, but determining the most
appropriate year requires further discussion and potentially sensitivity
analyses.

Despite these limitations, this research significantly enhances our understanding
of the mechanisms and motivations behind autocoups, contributing to the
literature on political stability and democratic resilience. The findings
highlight the vulnerability of political systems, particularly democracies, to
erosion from within by incumbent leaders.

Future studies could build on this work by employing the dataset to explore more
nuanced power dynamics or examine the long-term impacts of these events on
political systems. Particularly fruitful areas for investigation include the
relationship between autocoups and democratic backsliding, democratic breakdown,
and the personalization of power. Additionally, comparative analyses between
autocoups and traditional coups could yield insights into the evolving nature of
power consolidation strategies in different political contexts.

In conclusion, this study not only provides a valuable resource for future
research but also contributes to our understanding of the complex interplay
between leadership, institutional structures, and political stability. As
autocratic tendencies continue to challenge democratic norms globally, the
insights gained from this analysis of autocoups become increasingly relevant for
both scholars and policy-makers concerned with preserving and strengthening
democratic institutions.

## Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-Entry and Autocoup Leaders

### Abstract {.unnumbered}

This chapter examines the relationship between methods of power acquisition and
the longevity of leaders who assume power through irregular means, specifically
coup-entry and autocoup leaders. The central hypothesis posits that the mode of
accession significantly influences leader tenure. Employing Cox proportional
hazards and time-dependent Cox models, this study provides robust evidence of
divergent survival times between these two leader types. Findings reveal that
coup-entry leaders face a substantially higher risk of removal from office
compared to their autocoup counterparts. These results have significant
implications for political stability and democratic processes, suggesting that
the perceived low costs and high rewards associated with autocoups may
incentivize incumbents to extend their tenure through this mechanism, potentially
contributing to democratic backsliding. This research makes a notable
contribution to the academic literature by leveraging a newly developed dataset
on autocoups. By doing so, it offers valuable insights into the dynamics of
irregular leadership transitions and enhances our understanding of the complex
interplay between power acquisition methods and political longevity.

***keywords**: Coups, Autocoups, Leadership Survival, Cox Model*

\newpage

### Introduction

The longevity of political leaders has long captivated scholars and observers.
Why do some rulers endure for decades while others are ousted within months or
even days? This question has spurred extensive research in political science,
primarily focusing on the survival of political leaders. However, within this
broader framework, a specific subset of leaders—those who ascend to power through
coups or extend their tenure through autocoups—has received comparatively less
attention. Examining the tenures of these leaders is crucial, as it illuminates
the dynamics of irregular leadership transitions and their implications for
political stability and democratic processes.

Leaders who attain power through conventional means often follow predictable
patterns of ascension, tenure, and departure. In contrast, those who rise through
irregular channels, such as coups or autocoups, present more complex and
intriguing cases for study. Archigos dataset underscores the prevalence of such
irregular transitions: between 1945 and 2015, over half of leaders who assumed
power irregularly (158 out of 308) also exited irregularly. This rate
significantly surpasses that of leaders who accessed office through regular
channels, of whom only 14.5% (213 out of 1,472) experienced irregular departures.

Coup-entry and autocoup leaders constitute a substantial portion of these
irregular cases. Archigos dataset notes that of 374 leaders who exited
irregularly, 246 (65.8%) were ousted through coups. @frantz2016 further
demonstrate that coup-related exits account for approximately one-third of all
exits in autocracies, surpassing any other transition type. Additionally, the
autocoup dataset, introduced in [Chapter 3](#sec-chapter3), documents 110
autocoup attempts between 1945 and 2024, of which 87 were successful.

```{r}
#| label: data
#| include: false
survival_cox_ph_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_ph_model.csv") |> 
  mutate(survival_year = survival_days / 365)

survival_cox_td_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_td_model.csv")

avg <- survival_cox_ph_model |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)
```

Measuring the survival tenure of coup-entry and autocoup leaders presents
challenges due to the inherent irregularity and uncertainty of their positions.
Nevertheless, a comparative analysis reveals that leaders who extend power
through autocoups tend to have longer average post-autocoup tenures
(approximately `r avg$avg[1]` years) compared to coup-entry leaders
(approximately `r avg$avg[2]` years), suggesting a potential tenure gap of over
five years.

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders


# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = survival_cox_ph_model)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = survival_cox_ph_model, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("Autocoup leaders", "Coup-entry leaders"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 12)
```

A preliminary log-rank test in survival analysis, as illustrated in @fig-logrank,
demonstrates a statistically significant difference between the tenures of
autocoup and coup-entry leaders. The survival curve for autocoup leaders
consistently exceeds that of coup-entry leaders, indicating longer survival times
and a reduced risk of ouster for autocoup leaders.

This study posits that the method of accession significantly influences
leadership longevity. Coup-entry leaders likely confront greater challenges to
their rule, resulting in shorter average tenures compared to autocoup leaders.
The analysis, employing Cox proportional hazards and time-dependent Cox models,
supports this hypothesis, demonstrating that autocoup leaders generally
experience longer tenures than coup-entry leaders.

This research offers two primary contributions to the field. First, it highlights
an understudied factor in leadership survival analysis: the impact of the method
of accession to power. My findings suggest that leader survival is influenced not
only by ruling strategies but also by the initial method of acquiring power.
Second, by employing survival models, this study provides empirical evidence of
the significant difference in tenure duration between autocoup and coup-entry
leaders. This insight may explain the increasing prevalence of tenure extensions
through autocoups since 2000, as more incumbents observe and potentially emulate
successful precedents.

The remainder of this chapter is structured as follows: Section 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Section 3 explores the factors influencing the survival of
coup and autocoup leaders. Section 4 outlines the methodology and data used,
including the application of survival models to analyse the determinants of
leadership longevity. Section 5 presents the analysis findings and a detailed
discussion of the results. Finally, Section 6 concludes by synthesizing key
takeaways and exploring their broader implications for political stability and
democratic processes.

### Literature review

The longevity of political leaders has been a central focus in political science
research for decades, driven by the wide-ranging variations observed across
regimes, countries, and historical periods. This field encompasses two
interconnected aspects: regime survival and individual leader survival. Regime
survival focuses on the endurance of political systems, such as monarchies,
political parties, or specific ideological structures, while leader survival
concerns the duration of individual leaders' time in office.

These concepts often exhibit contrasting patterns. For instance, parliamentary
democracies like Japan or the United Kingdom may experience prolonged periods of
party dominance coupled with frequent leadership changes, whereas communist
regimes typically demonstrate enduring party rule with more frequent leadership
transitions. Presidential systems, such as the United States or many military
regimes, tend to exhibit more frequent changes in both ruling party or junta and
leader. This study specifically investigates the dynamics of individual leader
survival, focusing on factors influencing the duration of leaders' time in
office.

The existing literature on leader survival is extensive and multifaceted. Some
studies explore specific mechanisms influencing leadership longevity within
particular regimes, such as democracies [@svolik2014] or autocracies
[@davenport2021], while others aim to develop more generalizable theoretical
frameworks explaining leader survival across different political systems
[@buenodemesquita2003]. Although a universal theory remains an aspirational goal,
the complexities of leadership survival across diverse regime types present
significant challenges.

Power transition mechanisms vary substantially across different regimes,
particularly between democracies and autocracies. Autocratic systems often
feature closed leadership selection processes, restricted to a narrow pool of
individuals such as royal families, military elites, or ruling party members.
While some autocracies may hold elections, significant barriers to entry for
legitimate challengers typically persist. Potential rivals may face threats like
assassination, imprisonment, or exile. Moreover, the opacity of selection
processes makes it difficult to assess genuine levels of public support compared
to democracies. Consequently, conceptualizing selectorates or winning coalitions,
as proposed by @buenodemesquita2003, becomes problematic in many autocratic
contexts.

Given these complexities, focusing research on specific regimes or leader types
may be more fruitful. While regular leadership changes offer valuable insights,
they provide limited opportunities to explore the dynamics of leader survival. In
contrast, the study of irregular leaders, such as those who ascend to power
through coups or overstay in office through autocoups, offers a more compelling
avenue for research due to the inherent complexities and uncertainties
surrounding their leadership trajectories.

Two primary perspectives have emerged to explain the dynamics of leader survival.
The first emphasizes objective factors and resources, such as personal competence
[@yu2016], societal stability [@arriola2009], economic development [@palmer1999;
@williams2011], natural resource endowments [@smith2004; @quirozflores2012;
@wright2013], and external support [@licht2009; @wright2008; @thyne2017]. The
second focuses on subjective factors and strategies, including political
policies, responses to opposition, and tactics for consolidating power
[@gandhi2007; @morrison2009; @escribà-folch2013; @davenport2021].

Coups have received considerable scholarly attention, with research examining
coup prevention strategies, the impact of coups on leadership, and the subsequent
actions of coup leaders. Existing research delves into strategies for thwarting
coups [@powell2017; @sudduth2017; @debruin2020] and how leaders extend their
tenures after surviving coup attempts [@easton2018]. @sudduth2017 examines
post-coup actions of dictators, focusing on purge strategies, while @sudduth2018
investigates how leaders' entry methods affect their removal in dictatorships.

However, a significant gap exists in the literature regarding the comparison of
leadership survival between coup-entry and autocoup leaders. This study aims to
address this gap by investigating and comparing the duration of leadership
survival for these two leader types, contributing to a more nuanced understanding
of political survival in irregular leadership transitions.

### Survival dynamics of autocoup and coup-entry leaders

The study of leadership survival in political systems presents inherent
challenges due to the opacity and diverse mechanisms of power transitions.
However, these challenges underscore the significance of this research, as it
illuminates understudied dynamics in political leadership. While the survival of
political leaders exhibits complexity and variation, it is not entirely devoid of
patterns. Leaders of similar types often display significant comparability.

#### Key Definitions and Scope

Before delving into the comparison, it is essential to clarify several key
terminologies:

-   **Coup and Autocoup**: These terms adhere to the definitions established in
    previous chapters.

-   **Tenure Length Threshold**: To ensure meaningful analysis, this study
    focuses on leaders with substantial periods in power, applying a six-month
    threshold to both autocoup and coup-entry leaders.

-   **Autocoup Leader**: An incumbent leader who successfully uses illegitimate
    or unconstitutional means to extend their tenure in power.

-   **Coup-Entry Leader**: The individual who assumes power after a successful
    coup, regardless of their role in the coup itself.

This study focuses on comparing the post-autocoup tenure of autocoup leaders with
the post-coup tenure of coup-entry leaders, motivated by the relevance and
similarity of these leader types in terms of illegitimacy, uncertainty, and
instability.

#### Challenges in Power Consolidation

Both autocoup and coup-entry leaders face distinct challenges in consolidating
their power, primarily due to differences in the intensity of issues related to
illegitimacy, uncertainty, and instability. This disparity creates an uneven
playing field in terms of power dynamics, with coup-entry leaders at a
significant disadvantage. @tbl-leaders compares the main features between
autocoup and coup-entry leaders.

\blandscape

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-entry leaders 
#| echo: false
#| results: asis

read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(150),
    2:3 ~ px(300)
  )

# |> data_color(rows = c(2, 4), colors = "#edede9")
```

\elandscape

##### Illegitimacy {.unnumbered}

While both types of leaders suffer from a legitimacy deficit, the nature of this
deficit differs:

-   **Coup-Entry Leaders**: Their illegitimacy is explicit due to the open
    seizure of power.

-   **Autocoup Leaders**: They employ a deceptive strategy, manipulating legal
    processes to create a façade of democratic legitimacy.

##### Uncertainty {.unnumbered}

The irregular paths to power create uncertainty regarding their reigns and
eventual departures. However, the levels of uncertainty differ.

Coup-entry leaders face three major uncertainties: Unclear who will assume
leadership after the coup, uncertain tenure length, and ambiguity regarding
future successors.

Autocoup leaders present a clearer picture: No ambiguity about who will rule
after an autocoup, many seek to extend their rule indefinitely or incrementally.

##### Instability {.unnumbered}

The awareness of shaky legitimacy and persistent uncertainty breeds insecurity
and a sense of crisis.

Coup-entry leaders must reshape power dynamics and often purge potential
adversaries. They need to create a new equilibrium, often disrupting established
structures. They also face potential backlash even from close allies and hence
must compromise with internal or external power challengers.

Autocoup leaders encounter fewer abrupt changes in their regimes. They face less
pressure to dismantle existing ruling paradigms and have more time to implement
changes gradually.

#### Empirical Evidence and Hypothesis

Empirical evidence supports the disadvantage faced by coup-entry leaders. Data
reveals a correlation between the frequency of coup attempts in a country and the
likelihood of future coups. Over a third of coups occur in the top ten countries
with the most attempts since 1950 (@tbl-coups). The average survival period
following an autocoup is approximately five years longer than that of coup-entry
leaders (@fig-logrank).

The distinct challenges faced by autocoup and coup-entry leaders in consolidating
power create a self-perpetuating cycle that influences their tenure length.
Coup-entry leaders, facing greater challenges, struggle to attract and retain
strong support, making them more vulnerable to internal and external challenges.
Conversely, autocoup leaders, often benefiting from a veneer of legitimacy and a
stronger initial position, are better able to consolidate power and attract
supporters, leading to potentially longer tenures.

Based on these observations, I propose the following hypothesis:

***H4-1: Political leaders who successfully extend their tenure through autocoups
are more likely to survive longer compared to coup-entry leaders.***

### Research Design

This study employs survival analysis to test the hypothesis that autocoup leaders
have longer survival times in office compared to coup-entry leaders. The research
design utilizes Cox models to analyze the survival tenures of these two types of
leaders, accounting for multiple factors that may influence their time in power.

#### Methodology: Survival analysis

Two Cox models will be employed to analyze the survival tenures of coup-entry and
autocoup leaders:

-   **Cox Proportional Hazards (PH) Model**: This model uses only the variables
    present at the entry year, without considering changes over time.

-   **Time-Dependent Cox Model**: This model accounts for variations in
    time-dependent control variables such as economic performance and political
    stability.

The Cox model is preferred over the Kaplan-Meier model as it allows for the
estimation of multiple factors' impacts. While it does not directly estimate the
duration of tenure in office, it evaluates the hazard rate associated with being
ousted from power. This approach captures different facets of the same
phenomenon: as a leader's cumulative hazard of being ousted increases, their
probability of survival in office decreases.

#### Data and Variables

The dependent variables include survival time and end point status:

-   **Survival Time:** The duration of a leader's tenure, measured in days. For
    coup-entry leaders, the survival time begins on the day they assume power
    through a coup. For autocoup leaders, the survival time starts on the
    expiration date of their original legitimate term. For example, Xi Jinping
    assumed power in 2013 and removed term limits in 2018. His original
    legitimate tenure was set to end in 2023, so his survival time begins in
    2023, marking the start of his post-autocoup tenure. The survival time
    concludes on the day the leader exits office, applicable to both coup-entry
    and autocoup leaders.

-   **End point status:** This variable indicates the manner in which the
    leader's tenure concluded, categorized as follows:

    **0 = Censored:** This status is assigned to leaders who leave office through
    regular means other than being ousted. This includes leaders transferring
    power to their designated successors, leaving office as their terms expire,
    losing in general elections, voluntarily leaving office due to health issues,
    or dying of natural causes.

    **1 = Ousted:** This status is assigned to leaders who are forced to leave
    office. This includes leaders resigning under pressure, being ousted by coups
    or other forces, or being assassinated.

The key independent variable is the leader type, which categorizes leaders into
two distinct groups:

-   **Group A = Autocoup Leader**: Leaders who extend their tenure through
    autocoups.
-   **Group B = Coup-Entry Leader**: Leaders who assume power through coups.

This variable is the primary independent variable of interest, serving as the
basis for comparing the survival time between these two types of leaders.

The data for both dependent and independent variables are sourced from the
autocoup dataset introduced in this study, Archigos, and PLAD.

Control variables include economic performance, political stability, population
size, and the leader's age, which are consistent with the autocoup analysis in
[Chapter @sec-chapter3].

### Results and discussion

#### Model results

Using the `survival` package in R [@survival], I present the regression results
for both the Cox Proportional Hazards (Cox PH) model and the time-dependent Cox
model in @tbl-cox.

```{r}
#| label: model

Mydata1 <- survival_cox_ph_model |> 
   mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")))

model_cox1 <- Mydata1 |> 
  coxph(
    Surv(survival_days, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age,
    data = _
  ) 

tbl1 <- model_cox1|> 
   tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T,
                         hide_se = F, 
                         hide_ci = T,
                         thresholds = c(0.01, 0.05, 0.1),
                         ) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 

Mydata2 <- survival_cox_td_model |> 
  mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")),
         survt = (tstop - tstart) * 365)

model_cox2 <- Mydata2 |> 
  coxph(
    Surv(T1, T2, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age + cluster(id),
    data = _
  ) 
tbl2 <- model_cox2|>
  tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T,
                         hide_se = F, 
                         hide_ci = T,
                         thresholds = c(0.01, 0.05, 0.1),
                         ) |> 
  bold_p(t = 0.05) |> 
  bold_labels()
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox models for survival time of different types of leaders


tbl_merge(list(tbl1, tbl2),
          tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**"))
```

Both the Cox PH model and the time-dependent Cox model analyses revealed a
statistically significant association between leadership type and the hazard of
removal from power. Since time-dependent Cox model use the control variables
which change over time, I interpret the main findings based on time-dependent
model.

Coup-entry leaders were found to have a hazard ratio of 2.23 in the
time-dependent model compared to autocoup leaders (reference group), assuming all
other variables in the model are held constant. This suggests that coup-entry
leaders face a significantly greater risk of removal from power compared to
autocoup leaders. At any given time during their tenure, coup-entry leaders are
2.23 times more likely to be ousted from power compared to autocoup leaders, all
else being equal in the model.

The control variables perform differently in the two models. Economic level (GDP
per capita) exhibits statistically significant effects in both models. In the
time-dependent model, the hazard ratio of 0.95 indicates that for each unit
increase in GDP per capita (measured in units of \$10,000), the hazard (or risk)
of being ousted at any given time is reduced by 5%, assuming all other variables
in the model are held constant.

GDP growth trend demonstrates a more substantial effect in reducing the risk of
coups. Specifically, a 1 percentage point higher economic growth trend is
associated with an 80% reduction in the risk of being ousted, although this
effect is only statistically significant at the 10% level. This suggests a
possible trend where positive economic performance might mitigate the risk of
removal from power, but the evidence is not robust enough to confirm this
conclusively.

Political stability, as measured by the violence index, shows that a 1-point
increase in the index correlates with an 11% higher risk of being ousted.
However, this effect is also only statistically significant at the 10% level,
indicating a weaker but potentially important relationship between increased
violence and the risk of removal from office.

#### Discussion

```{r}
#| label: fig-coxSurv
#| fig-cap: Survival curves for Cox Model
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

testdata1 <- Mydata1 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox1 |> 
  survfit(newdata = testdata1) |>
  ggsurvplot(data = testdata1,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))

testdata2 <- Mydata2 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox2 |> 
  survfit(newdata = testdata2) |>
  ggsurvplot(data = testdata2,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))
```

The survival curves depicted in @fig-coxSurv illustrate the survival rates for
leaders of both types. Both the Cox PH model and the time-dependent Cox model
produce similar plots. Notably, the survival curve for coup-entry leaders
exhibits a significantly lower trajectory compared to that of autocoup leaders.
The steeper drop at the early stage for coup-entry leaders indicates they are
more likely to be ousted shortly after assuming power. Additionally, the survival
curve for coup-entry leaders crosses the median survival line much earlier (about
3,000 days) than that of autocoup leaders (about 8,500 days). This disparity
suggests that autocoup leaders tend to remain in power for longer durations than
their coup-entry counterparts.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard ratios and 95% CIs for Leader Ousting
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw() 

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |> 
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw() 
```

@fig-coxHR displays the hazard ratios and corresponding 95% confidence intervals
for the variables incorporated in the Cox model. Both the Cox Proportional
Hazards (PH) model and the time-dependent model produce similar plots,
reinforcing the robustness of the findings. Key points to note include:

-   The closer the hazard ratio (represented by the dots) is to 1, the less
    impact the variable has on the risk of being ousted. A hazard ratio of 1
    indicates no effect.

-   The whiskers extending from the dots represent the 95% confidence intervals.
    If these whiskers cross the vertical blue line at 1, it indicates that the
    variable is not statistically significant at the 5% level.

-   The hazard ratio for coup-entry leaders is significantly greater than 1 and
    statistically significant at the 5% level. This indicates that coup-entry
    leaders face a substantially higher risk of being ousted compared to autocoup
    leaders.

-   Most other variables have hazard ratios close to 1, suggesting that a
    one-unit increase in these variables does not significantly affect the risk
    of being ousted.

-   Although the hazard ratio for GDP growth trend is considerably less than 1 in
    the time-dependent model, indicating a potential protective effect, it is not
    statistically significant at the 5% level. However, it is statistically
    significant at the 10% level, suggesting that better economic performance may
    help to consolidate the rule of the incumbents to some extent, albeit the
    evidence is not as strong.

#### Assessing the Proportional Hazards Assumption

Assessing the proportional hazards assumption is crucial for the validity of the
Cox model results. To evaluate this, we used the chi-square test based on
Schoenfeld residuals to determine whether the covariate effects remain constant
(proportional) over time. Although the Cox PH model violates the proportional
hazards assumption, our primary analysis relies on the time-dependent Cox model,
which does not show strong evidence of violating the proportional hazards
assumption for any covariate. The global p-value of 0.416 is much greater than
the 5% significance level, indicating that the proportional hazards assumption is
reasonably met for the time-dependent Cox model.

```{r}
#| label: robust_test
#| include: false
test <- cox.zph(model_cox2)
```

### Conclusion

This chapter examined the survival durations of political leaders who come to
power through irregular means, specifically coups and autocoups. I hypothesized
that the mode of accession significantly influences leader tenure. Employing
survival analysis techniques, including the Cox proportional hazards model and a
time-dependent Cox model, I found strong evidence that autocoup leaders generally
enjoy longer tenures than coup-entry leaders.

The findings revealed a significant difference in average tenure, with
post-autocoup leaders averaging approximately 11 years in power compared to 5.6
years for coup-entry leaders. The time-dependent Cox model further indicated that
coup-entry leaders are 2.23 times more likely to be ousted from power at any
given time compared to autocoup leaders, all else being equal.

These results highlight the importance of understanding the phenomenon of
autocoups, where leaders extend their rule by manipulating legal frameworks. Due
to the relative ease and potential benefits of autocoups, this method of power
retention might incentivize more leaders to employ it. Consequently, democratic
backsliding could become more prevalent as autocoups weaken democratic
institutions and constitutional norms, particularly in nascent democracies or
those transitioning from autocracy.

This study contributes to the field of leadership survival by demonstrating that
the mode of accession significantly impacts leader tenure, a factor previously
under-explored in the literature. By utilizing both Cox models, the research
offers robust analytical techniques for studying political leadership survival
and provides strong evidence of divergent tenure lengths between these two types
of irregular-entry leaders.

However, limitations exist. The study relies heavily on the autocoup dataset
collected and coded by the author. The concept and data itself are relatively
novel within academia. Future research should refine and establish wider
recognition for the term "autocoup," leading to more accurate and comprehensive
data collection efforts. Expanding the dataset to include more cases and
integrating it with data on other irregular leadership transitions could yield a
more holistic understanding of political survival in such contexts.

Overall, this chapter underscores the need for more nuanced approaches to
studying political tenure and the mechanisms of irregular power retention,
contributing valuable insights into the dynamics of political stability and the
risks associated with different forms of non-democratic leadership succession.

## Conclusion

### Main Findings

This study delves into the dynamics and implications of irregular power
transitions, focusing on coups and autocoups. The findings illuminate the complex
interplay between incumbents and challengers fighting for power.

Firstly, our analysis reveals that the expected success rate of a coup attempt
significantly influences its likelihood. This success rate is heavily influenced
by the balance of power between the incumbent regime and challengers, which is
largely determined by regime type. We find that military regimes, although with
more control over their own military forces, face a higher risk of coups compared
to dominant-party regimes.

Secondly, the study introduces a redefined concept: the autocoup. Defined as an
incumbent leader's refusal to relinquish power as mandated, this research
distinguishes autocoups from broader terms like self-coups. Based on this
definition, we present the first publicly available dataset of autocoup events
from 1945 to 2022, encompassing 110 attempts and 87 successful autocoups. Case
studies and empirical analyses demonstrate the dataset's utility for quantitative
research, providing a robust foundation for further analysis on autocoups.

Thirdly, employing survival analysis techniques, the study finds clear
differences in leader longevity between those who come to power through coups and
those who extend their rule through autocoups. The results indicate that
coup-installed leaders face a significantly higher risk of removal compared to
autocoup leaders who manipulate the system to extend their rule.

### Limitations and directions for future research

This study offers a novel framework for analysing irregular power transitions,
but some limitations require further exploration:

-   **Data refinement:** Defining and classifying autocoups is a new approach.
    Future research should validate this classification system through additional
    studies and expert evaluations.

-   **Data harmonization:** The current analysis faces challenges due to
    mismatched units (country-year vs. leader) between coup and autocoup
    datasets. Future efforts should explore data harmonization techniques for
    more robust comparisons.

-   **Democratic backsliding:** While this study establishes a connection between
    irregular power transitions and democratic backsliding, further empirical
    evidence is needed to solidify this link.

Several avenues exist for future research:

-   **Terminology and data collection:** Refining the "autocoup" concept and
    achieving wider recognition will facilitate more accurate and comprehensive
    data collection.

-   **Dataset expansion:** Expanding the autocoup dataset with more cases and
    integrating it with data on other irregular leadership transitions can
    provide a more holistic view of political survival after these events.

-   **Power dynamics and long-term impacts:** Utilizing this dataset, future
    studies can delve deeper into power dynamics at play and explore the
    long-term consequences of irregular transitions on political systems,
    particularly regarding democratic backsliding, breakdown, and personalization
    of power.

In conclusion, this study sheds light on the dynamics of irregular power
transitions, specifically focusing on coups and autocoups. By redefining
autocoups, classifying the dataset, analysing determinants, and comparing leader
longevity, we establish a framework for understanding irregular transitions and
leader survival. This work contributes to a deeper understanding of democratic
resilience and political stability. Future research can build upon this
foundation by conducting further empirical analyses based on the novel autocoup
dataset and continuing to refine the framework.

```{r}
#| label: references
#| results: asis
#| echo: false

cat("\\newpage
## References")
```

[^1]: The definitions and concepts of power expansion and power extension can
    often be ambiguous. In this study, we define power expansion as an incumbent
    acquiring additional authority from other branches or apparatuses of the
    state. Conversely, power extension refers to an incumbent prolonging their
    tenure beyond the originally mandated term in office.
