---
format:
  pdf: 
    documentclass: report
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    toc: true
    toc-title: Contents
    lof: true
    lot: true
    crossref: 
      lof-title: Figures
      lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  docx: default
    
categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
include-before-body: title.tex
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false

pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue, plm, brglm2)
```

## Acknowledgements {.unnumbered}

The completion of this thesis marks the culmination of a remarkable journey,
filled with dedication, perseverance, and moments of profound joy. I am deeply
grateful to the numerous individuals who have supported and encouraged me
throughout this endeavour.

I would like to express my sincerest appreciation to my supervisor, Professor
Kristian Skrede Gleditsch, whose guidance, expertise, and unwavering support have
been instrumental in shaping my research. His constructive feedback and
encouragement have been invaluable, and I am profoundly grateful for his
mentorship.

I am also grateful to Professor Han Dorussen, the chair of my board panel, for
his continuous support and thoughtful input. His insightful comments and
suggestions have significantly enhanced the quality and depth of my research.

I would like to acknowledge the important contributions of my initial
co-supervisors, Dr. Saurabh Pant and Professor David Siroky, who laid a strong
foundation for this work during the early stages of my research. Although they
are no longer at the University of Essex, their instruction and guidance were
instrumental in shaping the direction of this project.

I have been fortunate to receive feedback and guidance from several esteemed
scholars in the field, including Dr. Brian J Phillips, Dr. Prabin Khadka, and Dr.
Winnie Xia. Their expertise and insights have enriched this research, and I am
grateful for their contributions.

On a personal note, I would like to express my deepest gratitude to my family,
who have been a constant source of support and inspiration throughout this
journey. To my beloved wife, Ji Zhi, your patience, love, and encouragement have
been immeasurable. To my dear children, Siyan and Sisheng, your joy and curiosity
have motivated me to persevere and strive for excellence.

I am also deeply grateful to my father for his enduring support and belief in my
abilities. To the cherished memory of my late mother, your love, guidance, and
values continue to shape my path and inspire my endeavours. And to my three
brothers, whose support enabled me to pursue my PhD without worries, I am forever
grateful.

While many individuals have contributed to the success of this work, I take full
responsibility for any errors or shortcomings that may remain.

## Abstract {.unnumbered}

This thesis addresses a significant gap in the study of irregular leadership
transitions by systematically incorporating autocoups—events in which incumbent
leaders extend their constitutionally mandated term, either through legal
manipulation or overtly illegal means. It refines the conceptual definition of
autocoups by resolving existing ambiguities, thereby aligning them more closely
with traditional coup frameworks. Based on this refined definition, the thesis
introduces a novel global dataset of autocoup events from 1945 to 2023,
encompassing 109 documented cases, 87 of which were successful.

Leveraging this dataset, the study undertakes a large-N empirical analysis to
explore the structural determinants of autocoups. It finds that personalist
regimes are significantly more prone to employing autocoups as a means of power
retention compared to other autocratic subtypes. The analysis then turns to
leadership survival, employing survival analysis techniques to compare the
political longevity of leaders installed via classic coups with those who
extended power through autocoups. The findings reveal that coup-installed leaders
are 2.23 times more likely to be forcibly removed from office than autocoup
leaders. On average, autocoup leaders enjoy post-event tenures approximately five
years longer than their coup-installed counterparts.

The thesis also evaluates the broader institutional implications of these power
seizures, particularly their impact on democratization. Using Polity scores as a
proxy for democratic quality and applying a country-fixed effects model, the
analysis shows that autocoups are associated with a steady decline in democratic
institutions both before and after their occurrence. In contrast, traditional
coups often produce a sharp initial decline in democratic quality, but may be
followed by partial recovery over time.

Together, these findings underscore the distinct trajectories and consequences of
coups and autocoups. This research makes several important contributions: it
clarifies the conceptual boundaries of autocoups, offers a new empirical
foundation for their study, and provides robust comparative insights into how
different modes of irregular power transition affect leadership survival and
institutional development. The implications are substantial for both academic
inquiry and policy-making, particularly in the context of global democratic
backsliding and the resilience of political institutions.

***Keywords:** Coups, Autocoups, Leadership transitions, Leadership survival,
Democratic resilience*

## Introduction

At the heart of contemporary political dynamics lies a fundamental question: *Why
are some political leaders prematurely removed from power, while others succeed
in extending their tenure beyond constitutional limits?* Furthermore, *how does
the manner of their survival or ouster shape political stability and democratic
institutions?* This thesis seeks to address these pressing questions by analysing
the structural and strategic underpinnings of irregular leadership transitions.

### Motivations

The stability and resilience of political systems depend heavily on the orderly
transfer of power. When leadership transitions occur within established
institutional frameworks, they reinforce legitimacy and bolster regime
durability. In contrast, the breakdown of conventional mechanisms for political
succession often leads to instability, violence, and institutional erosion. Among
the most disruptive of these breakdowns are irregular leadership transitions,
which cast long institutional shadows and fundamentally reshape the political
trajectory of regimes. Understanding the causes and consequences of such events
remains a central concern in the study of political order and regime change.

Extant literature identifies a range of catalysts for irregular leadership exits,
including civil wars [@kokkonen2019], international conflict [@demesquita1995],
ethnic tensions [@londregan1995], economic downturns [@miller2012;
@krishnarajan2019], and natural disasters [@quirozflores2012]. Among these, coups
d’état stand out as perhaps the most consequential, given both their frequency
and their direct targeting of incumbents. In autocracies, coups are responsible
for nearly one-third of all leadership exits—surpassing regular transitions,
which account for just over one-fifth [@frantz2016]. Moreover, over 63% of
non-constitutional removals in dictatorships result from coups [@svolik2009].

As a result, coups have received substantial scholarly attention. A robust body
of literature examines their causes, outcomes, and long-term impacts on democracy
and development [@thyne2019]. Scholars have proposed nearly one hundred potential
determinants of coups, yet consensus on a baseline explanatory model remains
elusive [@gassebner2016].

By contrast, another form of irregular power transition remains comparatively
neglected in the literature: the autocoup. In this phenomenon, an incumbent
leader extends their tenure by circumventing or violating constitutional term
limits—either through legal manipulation or overtly illegal acts. While autocoups
do not involve immediate leadership change, they constitute a profound disruption
of institutional succession norms and thwart expected transitions of power. As
such, they should be understood as a critical, albeit underexamined, variant of
irregular leadership transition.

This thesis contends that autocoups warrant systematic analysis alongside
traditional coups within a unified analytical framework. Despite their distinct
mechanisms, both coups and autocoups represent extra-constitutional attempts to
retain or seize power, and both can produce profound consequences for leadership
survival, regime durability, and institutional integrity. Comparing these two
forms of irregular transitions can illuminate shared drivers, divergent
consequences, and implications for democratic resilience.

The urgency of this research is underscored by the severe risks associated with
irregular transitions. Coups and autocoups alike can spark immediate
crises—ranging from institutional paralysis to civil conflict—and leave
long-lasting institutional scars. More fundamentally, they often dismantle
constitutional checks, disrupt electoral processes, and accelerate democratic
erosion or authoritarian entrenchment.

Historical examples highlight these dangers vividly. Ghana’s turbulent period
from 1979 to 1984 exemplifies the destabilizing effects of classic coups:
following Jerry Rawlings’s 1979 coup, eight individuals, including three former
heads of state, were executed [@pieterse1982]. Rawlings staged another coup in
1981 and subsequently suppressed three further coup attempts [@haynes2022d]. In
contrast, the 1992 autocoup in Peru by President Alberto Fujimori demonstrates
how incumbents can dismantle democratic institutions without leadership change:
Fujimori dissolved Congress, suspended the constitution, and ruled by decree
[@mauceri1995; @cameron1998].

These patterns are increasingly relevant in today’s global political context.
According to Freedom House’s Freedom in the World 2024 report, global political
rights and civil liberties declined for the 18th consecutive year in 2023, with
setbacks in 52 countries and improvements in only 21 [@freedomhouse2024freedom].
The persistence of democratic erosion underscores the urgent need to understand
the mechanisms that facilitate it—including both coups and autocoups.

This thesis aims to advance theoretical and empirical understanding of irregular
leadership transitions, offering insights with significant implications for both
academic scholarship and policy-making in fragile and transitioning regimes.

### Research objectives and contributions

To address the pressing challenges posed by irregular leadership transitions,
this study undertakes a comprehensive comparative analysis guided by four core
research objectives. First, to refine the conceptual definition of autocoups and
introduce a novel global dataset suitable for large-N empirical analysis. Second,
to identify the structural and institutional determinants of autocoups through
systematic quantitative investigation. Third, to compare the survival prospects
of leaders who come to power through traditional coups with those who extend
their tenure via autocoups. Finally, to assess the divergent impacts of coups and
autocoups on democratization trajectories and the resilience of political
institutions.

By examining both coups and autocoups from 1950 to 2023, this thesis addresses a
significant gap in political science by developing and applying a unified
analytical framework to treat these events as distinct yet interrelated forms of
extra-constitutional power transitions. Through this lens, the study makes four
key contributions to the literature on leadership dynamics, regime stability, and
institutional development.

#### Conceptual Clarification and Empirical Foundation for Autocoups {.unnumbered}

The thesis advances conceptual clarity by situating autocoups within the broader
typology of irregular power transitions. It offers a refined definition of
autocoups—centred on the extension of tenure—and distinguishes them from both
legal executive aggrandizement and traditional coups. Based on this
reconceptualization, the study introduces a novel global dataset of autocoups
spanning from 1945 to 2023, documenting 109 events, of which 87 were successful.
This dataset addresses a long-standing empirical deficit and provides the
foundational basis for systematic comparative analysis, enabling further research
into a previously understudied form of institutional disruption.

#### Determinants of Autocoups {.unnumbered}

Drawing on this new dataset, the thesis conducts an empirical analysis of the
conditions under which autocoups are likely to occur. It finds that leaders
operating within personalist regimes are significantly more prone to extending
their tenure through autocoups than their counterparts in other autocratic
configurations. These findings contribute to the literature on regime type and
irregular power retention, shedding light on how institutional structure
influences leaders’ strategic decisions to subvert term limits.

#### Quantifying the Impact of Transition Type on Leader Longevity {.unnumbered}

This research contributes to the study of leadership survival by comparing tenure
lengths of leaders who seize power via coups with those who entrench themselves
through autocoups. Employing survival analysis on both coup and autocoup
datasets, the study demonstrates that the method of power acquisition is a
significant predictor of leadership duration. Specifically, leaders installed
through coups are 2.23 times more likely to be forcibly removed than autocoup
leaders, who on average serve five years longer following the event. These
findings underscore the importance of power transition mode in shaping regime
durability and elite turnover.

#### Disaggregating the Institutional Consequences of Irregular Transitions {.unnumbered}

The thesis further explores how different types of irregular transitions affect
democratic institutions. Using country-fixed effects regression models and Polity
scores as a measure of democratic quality, it finds that autocoups are
consistently associated with gradual and sustained democratic decline—both before
and after the event. In contrast, coups produce more varied institutional
outcomes: although they often cause immediate setbacks to democratic norms, in
some cases they allow for subsequent democratic recovery. This disaggregation
reveals distinct regime trajectories and institutional consequences stemming from
the two forms of power seizure.

Collectively, these contributions offer a more nuanced and empirically grounded
understanding of irregular leadership transitions. By conceptualizing coups and
autocoups within a shared analytical framework, providing original data, and
conducting rigorous comparative analyses, this thesis enhances our knowledge of
how and why leaders subvert institutional norms—and with what consequences. The
findings hold important implications for both academic theory and contemporary
policymaking, particularly in contexts marked by democratic fragility, elite
opportunism, and institutional decay.

### Policy implications

While scholarly debate continues over the potential for coups to inadvertently
catalyze democratization in certain contexts [@thyne2014; @derpanopoulos2016;
@miller2016], a strong policy consensus holds that coups are inherently
illegitimate mechanisms for political change. As violent disruptions of
constitutional order, they tend to inflict immediate institutional damage,
generate instability, and usher in unpredictable political trajectories.
Consequently, international and domestic policy responses have rightly
prioritized prevention—most notably through "coup-proofing" strategies aimed at
insulating regimes from military intervention or elite defection [@quinlivan1999;
@pilster2012; @powell; @albrecht2014; @carey2015; @brown2015; @sudduth2017]. Yet
these approaches have well-documented limitations [@albrecht2014a; @reiter2020],
and as this thesis demonstrates, deeper structural power dynamics within regimes
are often the most decisive factors in shaping vulnerability to both coups and
autocoups.

From this perspective, the study’s findings yield several important policy
implications.

The first and most direct implication concerns the design of political
institutions. Regimes that feature robust horizontal accountability
mechanisms—such as independent legislatures, judiciaries, and oversight
bodies—are inherently more resilient to irregular power seizures. A
well-calibrated balance of power reduces both the opportunity and incentive for
executive actors to bypass constitutional constraints. Institutional designs that
disperse authority, limit executive overreach, and embed strong checks and
balances can therefore serve as a first line of defense against both coups and
autocoups. Supporting good governance practices that prevent excessive power
concentration—through constitutional term limits, civil society engagement, and
transparent succession rules—is essential to reducing the risks of authoritarian
entrenchment.

A second major policy implication lies in the role of external actors and
international institutions. Regional and global organizations—such as MERCOSUR,
the Organization of American States (OAS), the African Union (AU), ECOWAS, the
European Union, and the Council of Europe—have played important roles in
deterring unconstitutional leadership changes. By establishing legal frameworks
that mandate democratic governance and deploying tools like sanctions, diplomatic
isolation, and suspension of membership, these bodies can exert pressure on
regimes that emerge from military coups or other overt violations of
constitutional norms [@wobig2014; @shannon2014; @thyne2017].

However, the effectiveness of these mechanisms is often limited when dealing with
autocoups. Unlike traditional coups, autocoups are typically executed through
procedural manipulations—such as emergency decrees, manipulated referenda, or
constitutional revisions—rather than overt military force. This legalistic veneer
complicates international responses: the absence of visible violence or abrupt
leadership change frequently leads to delayed or diluted intervention, hindered
by political ambiguity and lack of consensus.

Given the rising incidence of autocoups since the early 2000s and their central
role in democratic backsliding [@bermeo2016; @mechkova2017], international actors
must adapt their monitoring and response frameworks accordingly. A third policy
recommendation is therefore clear: international and regional organizations must
shift from reactive to proactive approaches, developing criteria that capture
gradual institutional erosion—a hallmark of modern authoritarian consolidation.
Rather than waiting for clear-cut disruptions, policymakers should employ
existing democratic monitoring tools—such as Polity scores, Freedom House
ratings, and V-Dem indices—to track sustained declines in institutional quality,
media freedom, judicial independence, and electoral competitiveness. When
patterns of democratic regression emerge, they should serve as early warning
signals, prompting calibrated responses: diplomatic pressure, conditional aid
adjustments, public condemnation, or even targeted sanctions. This would allow
the international community to act before constitutional structures are fully
hollowed out.

In sum, this thesis underscores that modern threats to democracy are increasingly
procedural rather than violent, and policy responses must evolve accordingly.
While coups remain a clear and present danger to political order, autocoups pose
a subtler but equally corrosive threat—eroding democracy from within rather than
toppling it outright. To preserve constitutional governance in an era of creeping
authoritarianism, both domestic institutions and international frameworks must be
retooled to detect and resist the slow unraveling of democratic norms, not just
their sudden collapse.

### Limitations and future research

While this study offers a novel analytical framework for understanding coups and
autocoups, their effects on leadership survival, and their broader institutional
implications, several limitations suggest important avenues for future research
and refinement.

First, although the definition and classification of autocoups presented here
represent a significant conceptual advance, broader scholarly consensus on these
parameters remains underdeveloped. Both the definitional boundaries and the
empirical dataset introduced in this study serve as foundational contributions,
yet they are necessarily provisional. Future research should aim to further
elaborate and refine these components to enhance the reliability and
comparability of findings across studies.

A central challenge lies in the conceptual ambiguity surrounding the very
definition of autocoups—particularly in borderline cases where incumbents extend
power through legal or quasi-legal means. Future work could investigate the
normative and analytical trade-offs involved in including such cases within the
autocoup category. Comparative analyses of "legal" versus "extra-legal" tenure
extensions may help clarify whether these phenomena should be treated as variants
of autocoups or as distinct processes altogether.

This definitional ambiguity directly impacts dataset construction. Minor
adjustments in inclusion criteria can substantially alter the number and
classification of identified cases, particularly where intentions, timing, and
legal interpretation are contested. A useful illustrative example is the case of
President Manuel Zelaya of Honduras in 2009. Zelaya proposed a referendum to
amend the 1965 Constitution, which banned re-election, ostensibly to allow future
leaders to seek another term. The referendum was scheduled for the same day as
the national presidential election, meaning that even if it had passed, Zelaya
would not have immediately benefited. However, questions remain about his
underlying intent: by proposing a reform that appeared to exclude him, Zelaya may
have been attempting to build public support for a subsequent reinterpretation or
amendment to extend his own rule. Ultimately, he was ousted in a military coup,
triggering a profound constitutional crisis [@muñoz-portillo2019]. This example
illustrates the analytical complexity of identifying autocoups and the importance
of improving coding criteria and interpretive clarity in future datasets.

Accordingly, both the conceptual framework and empirical foundation introduced in
this study would benefit from continued theoretical debate and methodological
refinement. Several promising research directions follow.

Given the long-term decline in traditional coups and the concurrent rise of
autocoups, intensified scholarly focus on the latter is essential. While this
study centres on tenure extensions as the defining feature of autocoups, broader
forms of executive power expansion—whether within or beyond constitutional
limits—warrant further scrutiny. These dynamics are both frequent and damaging to
institutional integrity, yet remain insufficiently explored in comparative
politics. To capture the full spectrum of these behaviors, the development of a
dedicated dataset on power expansions is a critical next step.

Moreover, as the incidence of overt and dramatic regime changes has decreased
since the early 2000s, so too has the number of clean transitions between
democracy and autocracy. This trend underscores the need for more sensitive
measurement tools that can detect incremental shifts within regimes. Future
empirical work should focus on capturing these subtler transformations, such as
small but sustained changes in Polity scores or other fine-grained institutional
indicators. Doing so will enable scholars to better track the slow erosion or
recovery of democratic governance in contexts where abrupt transitions are no
longer the norm.

### Overview of the thesis

This thesis investigates the complex power dynamics inherent in **coups and
autocoups**, focusing on their consequences for leadership survival and the
institutional integrity of political regimes. It develops a unified analytical
framework to analyze these phenomena as distinct yet interrelated forms of
irregular power transitions. Each chapter contributes to this overarching inquiry
by offering conceptual refinements, empirical innovations, and comparative
insights.

#### Chapter 2: Conceptualising autocoups and data introduction {.unnumbered}

Despite the increasing prevalence of autocoups—particularly in the post–Cold War
period—their systematic study remains underdeveloped. Existing scholarship
suffers from conceptual fragmentation, with a proliferation of overlapping and
inconsistently defined terms ("self-coup," "autogolpe," "executive
aggrandizement," etc.) [@marsteintredet2019; @baturo2022]. This ambiguity
complicates data collection and analysis, as many datasets fail to distinguish
between tenure extension and other forms of executive power consolidation—an
essential distinction for this study. Consequently, methodological progress has
been constrained, with most studies relying on qualitative case analyses
[@cameron1998; @antonio2021; @pion-berlin2022] rather than much-needed large-N
approaches.

This chapter addresses these shortcomings by offering a more precise and
theoretically grounded definition of the autocoup. It argues that autocoups
should be defined as instances where incumbents extend their constitutionally
mandated terms through legal manipulation or overtly illegal means. By centering
the concept on tenure extension, the definition excludes broader forms of
executive aggrandizement that occur within constitutional timeframes and aligns
autocoups conceptually with classic coups, both of which involve a disruption of
constitutionally prescribed leadership succession.

Based on this refined definition, the chapter introduces a major empirical
contribution: a newly compiled global dataset of autocoups spanning from 1945 to
2023, identifying 109 distinct events, of which 87 were successful. This dataset
enables robust quantitative analysis and opens new avenues for comparative
research on irregular power retention.

#### Chapter 3: Determinants of autocoup attempts {.unnumbered}

Due to long-standing conceptual and empirical limitations, previous discussions
of autocoups have relied predominantly on case studies [@baturo;
@marsteintredet2019; @baturo2022]. The dataset introduced in Chapter 2 enables,
for the first time, large-N analysis of the structural conditions associated with
autocoup attempts.

Drawing on insights from coup literature, this chapter examines a range of
potential predictors—including economic performance, succession rules, military
variables, protest levels, and media freedom. While these variables have been
explored in the context of traditional coups, they often fail to account for
persistent variation across regime types or explain the limited success of
"coup-proofing" strategies [@albrecht2014a; @reiter2020]. Additionally, studies
that incorporate regime type often rely on overly simplistic dichotomies (e.g.,
democracy vs. autocracy, or civilian vs. military), obscuring important variation
within autocratic subtypes [@hiroi2013; @schiel2019].

This chapter proposes that autocoup risk is fundamentally shaped by the
structural balance of power inherent in a regime’s foundational type.
Specifically, the likelihood of an autocoup depends on the equilibrium between
incumbent leaders and potential institutional challengers—a balance largely
locked in at regime inception [@geddes2014]. To test this, the chapter utilizes
regime typologies as proxies for internal power configurations.

Employing a probit model, the analysis finds that personalist regimes exhibit a
significantly higher likelihood of autocoup attempts, even when controlling for
other variables. Leaders in dominant-party and military regimes, by contrast, do
not differ significantly from democratic leaders in their propensity to attempt
an autocoup. These findings underscore the importance of regime structure in
shaping strategic decisions around irregular tenure extension.

#### Chapter 4: Power acquisition methods and leadership Survival {.unnumbered}

A substantial literature has examined the tenure survival of leaders who rise to
power via coups [@gandhi2007; @sudduth2017; @easton2018], yet the absence of a
comparable dataset on autocoups has historically prevented rigorous comparison
between coup-installed leaders and those who extend their rule through autocoups.
This chapter closes that gap by conducting the first comparative survival
analysis of these two categories within a unified theoretical framework.

The chapter argues that leaders who seize power through classic coups face
heightened legitimacy deficits, political uncertainty, and immediate
institutional instability, in contrast to autocoup leaders, who maintain
institutional continuity while neutralizing constraints. These differences shape
divergent paths to power consolidation.

Using Cox proportional hazards and time-dependent Cox models, the study finds
that leaders installed via coups are, on average, 2.23 times more likely to be
forcibly removed from office than those who extend their tenure through
autocoups. Moreover, autocoup leaders enjoy, on average, five additional years in
power post-event. These findings offer a compelling explanation for the rising
prevalence of autocoups and the relative decline of classic coups, particularly
since the early 2000s.

#### Chapter 5: Coups, autocoups, and democracy {.unnumbered}

While the impact of coups on democratization has been the subject of extensive
research [@clayton2000; @powell2014a; @thyne2020], the effects of autocoups
remain underexplored due to the historical lack of relevant data. This chapter
addresses that gap through a quantitative analysis of how coups and autocoups
affect democratic institutions.

Unlike coups, which may result in leadership change or regime transition,
autocoups typically involve incumbents dismantling institutional constraints
without changing the ruling coalition. As such, their impacts are evaluated not
through regime-type shifts but via changes in Polity scores, which provide a
continuous measure of democratic quality.

Two key arguments are advanced. First, leaders often begin undermining
institutional checks in anticipation of an autocoup, leading to declines in
Polity scores even before the event. Second, while coups generate mixed
democratic outcomes—occasionally enabling transitions—autocoups almost uniformly
result in democratic erosion or deeper authoritarian consolidation.

The empirical analysis, using a country-fixed effects model, confirms that
autocoups correlate with consistent declines in democratic quality both before
and after their occurrence. In contrast, coups often produce an immediate drop in
Polity scores, though some regimes subsequently recover. These findings
demonstrate that autocoups pose a uniquely insidious threat to democracy, often
proceeding incrementally and under a guise of legality.

#### Chapter 6: Conclusion and future research directions {.unnumbered}

The final chapter synthesizes the findings of the preceding chapters,
highlighting the structural, strategic, and institutional forces that shape
irregular leadership transitions. It emphasizes that coups and autocoups are not
merely disruptive events, but strategic tools wielded by political elites to
recalibrate or entrench their power. Their institutional legacies diverge: while
coups often destabilize regimes, autocoups tend to consolidate authoritarian
rule.

The chapter identifies key implications for understanding the resilience of
autocracy, the fragility of democratic institutions, and the strategic behavior
of leaders under threat. It also lays out several avenues for future research.

First, the broader phenomenon of executive power expansion—whether or not it
involves tenure extension—warrants systematic study. Given its growing frequency
and corrosive institutional impact, the development of a dedicated dataset on
power expansions is recommended.

Second, as full regime transitions become less common, future research should
focus on detecting subtle changes in political quality. Fine-grained tools—such
as small shifts in Polity scores—will be increasingly important for tracking
democratic backsliding and recovery within hybrid or semi-authoritarian regimes.

## Autocoups: Conceptual Clarification of Power Extensions by Incumbent Leaders {#sec-chapter3}

### Abstract {.unnumbered}

This chapter introduces a refined conceptualization of **autocoups**, defined as
instances in which incumbent leaders extend their constitutionally mandated
tenure by circumventing or violating term limits—either through legal
manipulation or overtly illegal means. By critically reviewing and synthesizing
overlapping terms such as *self-coup*, *autogolpe*, and *executive takeover*, the
chapter clarifies the boundaries of the concept, emphasizing its defining
feature: the extension of leadership tenure. In distinguishing autocoups from
broader and more ambiguous forms of executive aggrandizement, this chapter
presents a more analytically precise framework for studying irregular power
extensions. Building on this conceptual foundation, it introduces a novel global
dataset of autocoup events spanning 1945 to 2023, thereby enabling systematic,
large-N analysis of this increasingly salient form of authoritarian
consolidation.

***keywords**: Autocoups, Coups, Irregular Power Transitions, Leadership Tenure,
Dataset*

### Introduction

The stability and resilience of political systems fundamentally depend on the
orderly transfer of power. When leadership successions occur within established
constitutional frameworks, they reinforce the legitimacy and durability of
governing institutions. Conversely, when the norms and mechanisms governing
succession break down, they often trigger political violence, institutional
decay, and long-term instability.

In political practice, while many leadership transitions proceed smoothly, a
significant number do not. In particular, authoritarian and semi-authoritarian
regimes frequently witness two types of irregular leadership outcomes: the
premature removal of incumbents, and the illegitimate extension of power beyond
constitutional limits.

The former—forced removals of leaders before their terms expire—has been
extensively studied under the broader category of irregular leadership
transitions. These events cast long shadows, with profound consequences for
regime stability, democratic legitimacy, and institutional development. As such,
understanding the causes and consequences of irregular transitions remains
central to political science.

Extant literature identifies a range of triggers for these disruptions, including
civil war [@kokkonen2019], international conflict [@demesquita1995], ethnic
divisions [@londregan1995], poor economic performance [@miller2012;
@krishnarajan2019], and natural disasters [@quirozflores2012]. Among these,
however, coups d’état stand out as the most frequent and consequential cause of
leadership change—especially in autocracies, where they account for roughly
one-third of all leader exits, surpassing even regular transitions [@frantz2016].
In fact, over 63% of non-constitutional removals in dictatorships are the result
of coups [@svolik2009].

Defined as illegal and overt attempts by the military or state elites to unseat a
sitting executive [@powell2011], coups have attracted extensive scholarly
attention. The literature has explored their structural and immediate causes,
their aftermath, and their effects on democratic consolidation and economic
development [@thyne2019]. However, this focus on traditional coups risks
overlooking a distinct but increasingly significant form of irregular transition:
the autocoup.

An autocoup, as defined in this chapter, refers to an incumbent leader extending
their tenure by subverting or bypassing constitutional term limits—either through
legal manipulation (such as judicial reinterpretation or referenda) or clearly
illegal means (such as emergency decrees). Despite their growing
frequency—particularly since the end of the Cold War—autocoups remain
understudied. The field suffers from conceptual fragmentation, with a
proliferation of overlapping and inconsistently applied terms such as self-coup,
autogolpe, and executive aggrandizement [@marsteintredet2019; @baturo2022]. This
ambiguity hampers both definitional clarity and reliable data collection.
Existing datasets, while valuable, often conflate tenure extensions with other
forms of executive power consolidation, thus failing to capture the specific
dynamics this study seeks to analyze [@baturo2022]. These conceptual and
empirical gaps have, in turn, led to a reliance on qualitative case studies
[@cameron1998; @antonio2021; @pion-berlin2022] at the expense of broader
comparative insights.

This chapter argues that such limitations obscure a crucial dimension of
contemporary authoritarianism and proposes a unified framework for analysing
coups and autocoups together as distinct but comparable strategies for
undermining constitutional norms governing leadership succession. This
comparative approach is justified on three grounds.

First, both coups and autocoups constitute fundamental breaches of constitutional
order, with significant implications for democratic resilience, political
legitimacy, and institutional integrity. Analyzing them in tandem allows for a
systematic exploration of how different modes of irregular power transition shape
political development and democratization trajectories.

Second, although both disrupt established norms of succession, they operate in
opposite directions relative to the incumbent: while coups forcibly truncate an
incumbent’s tenure, autocoups unlawfully extend it. This contrast—between the
premature termination and the prolongation of rule—provides a compelling lens
through which to examine the divergent mechanisms of authoritarian survival and
entrenchment.

Third, this comparative lens helps illuminate key contemporary puzzles. For
instance, how can the decline in coup frequency since the 1990s [@bermeo2016] be
reconciled with the ongoing erosion of democratic governance, now in its 18th
consecutive year of global decline [@freedomhouse2024freedom]? By including
autocoups in the analytical framework, this research highlights the growing
importance of incremental, procedural subversions of democracy, often spearheaded
by incumbents from within the existing legal architecture.

In addressing these gaps, this chapter contributes in two principal ways. First,
it provides conceptual clarification by redefining autocoups as a subtype of
irregular power transitions explicitly centered on unconstitutional tenure
extension. This definitional clarity distinguishes autocoups from broader
executive aggrandizement. Second, it introduces a new global dataset of autocoup
events from 1945 to 2023, compiled according to this refined conceptualization,
enabling systematic quantitative analysis for the first time.

The remainder of the chapter is organized as follows. Section 2 reviews existing
definitions related to power extension and executive aggrandizement, culminating
in a refined conceptualization of autocoups. Section 3 introduces the newly
compiled global dataset of autocoup events, outlining its scope, coding criteria,
and methodological foundations. Section 4 provides an initial exploration of the
dataset through descriptive statistics and illustrative case studies. The
conclusion synthesizes the chapter’s key contributions and proposes avenues for
future research.

### literature review and clarification of definitions

A key limitation in the study of irregular leadership transitions is the lack of
integration between research on classic coups and autocoups. Although both
represent critical mechanisms of extra-constitutional power change, they have
typically been studied in isolation, with limited attention to their conceptual
and empirical intersections.

This divide stems primarily from two factors: the historical underrecognition of
autocoups as a distinct subtype of irregular transition, and the persistent
conceptual ambiguity surrounding their definition. While classic coups are marked
by the abrupt removal of incumbents, autocoups generally refer to incumbent-led
efforts to expand or extend power by circumventing constitutional constraints.
Yet the inconsistent use of overlapping terms—such as self-coup, autogolpe, and
executive aggrandizement—has further obscured these distinctions.

Clarifying the definition of autocoups is therefore a crucial first step toward
building a comparative framework capable of capturing the full spectrum of
irregular power transitions. This section undertakes that task by distinguishing
autocoups from broader forms of executive power consolidation and aligning them
conceptually with traditional coups through their shared violation of
constitutional norms.

#### Terminology {.unnumbered}

Existing studies that examine autocoups employ a range of terms to describe the
expansion of power or extension of tenure by incumbent leaders. The most commonly
used term is ‘self-coup’, or autogolpe in Spanish [@przeworski2000;
@cameron1998a; @bermeo2016; @helmke2017; @marsteintredet2019]. This label gained
prominence in scholarly discourse following the 1992 actions of Peruvian
President Alberto Fujimori, who dissolved Congress, temporarily suspended the
constitution, and ruled by decree [@mauceri1995; @cameron1998]. However, as
@marsteintredet2019 astutely observes, the term self-coup can be misleading, as
it implies that the leader is acting against themselves, when in fact such
actions are typically directed at other branches of government or state
institutions.

A second set of terms seeks to describe these events using modifiers or
adjectives, such as 'presidential coup', 'executive coup', 'constitutional coup',
'electoral coup', 'judicial coup', 'slow-motion coup', 'soft coup', and
'parliamentary coup' [@marsteintredet2019]. While these expressions may be useful
in capturing specific mechanisms or contexts, their proliferation often leads to
greater conceptual confusion. Many focus on how the power grab is executed, but
fail to consistently identify who the perpetrator is. Moreover, such
mechanisms—judicial rulings, legislative actions, or administrative decrees—may
be employed either by or against executive leaders, further complicating
definitional clarity.

A third group of terms includes expressions such as ‘incumbent takeover’,
‘executive takeover’, and ‘overstay’. For instance, incumbent takeover refers to
“an event perpetuated by a ruling executive that significantly reduces the formal
and/or informal constraints on his/her power” [@baturo2022, p.374], building on
earlier research by @svolik2014. Similarly, overstay is defined as “staying
longer than the maximum term as it stood when the candidate originally came into
office” [@ginsburg2011evasion, p.1844]. These terms are useful in identifying the
actor (the incumbent) and the behaviour (extending or consolidating power).
However, they fall short in conveying the illegality or unconstitutionality of
such actions. Unlike coup, which inherently implies an unlawful seizure of power,
terms such as takeover or overstay may understate the severity of the act.

Given that many of these terms prioritise procedural mechanisms over normative
content, or fail to distinguish between legal and extra-legal conduct, this study
adopts ‘autocoup’ as the most appropriate and analytically coherent term. It
offers several key advantages:

**Clarity and precision:** Autocoup is a specific and unambiguous term that
encapsulates the essential character of the event—namely, the unlawful extension
of an incumbent leader’s tenure—while clearly distinguishing it from traditional
coups and other forms of executive overreach.

**Normative and analytical weight:** The suffix -coup conveys the gravity of the
event, highlighting its unconstitutional and destabilising nature, and affirming
its relevance to both academic inquiry and policy discourse on par with, or
exceeding, that of classic coups.

**Identification of the perpetrator:** The prefix auto- unambiguously identifies
the incumbent leader as the instigator, thereby differentiating autocoups from
traditional coups, which are typically carried out by actors external to the
executive.

**Terminological symmetry:** Sharing a common root with coup allows for
conceptual alignment and comparative analysis. This symmetry facilitates the
study of coups and autocoups as distinct but related forms of irregular power
transitions, thereby enhancing analytical coherence and explanatory scope.

In sum, autocoup represents the most effective and analytically robust term for
describing this phenomenon. It captures both the actor and the act, while
signalling its irregularity, seriousness, and theoretical proximity to classic
coups, thereby supporting the unified framework proposed in this thesis.

#### Definition {#sec-definition .unnumbered}

While the use of precise terminology is undoubtedly important, another
significant issue in previous definitions of autocoups concerns the
identification of their primary characteristic: is the central feature power
expansion, tenure extension, or a combination of both? This question arises from
the notable ambiguity in existing definitions of autocoups and related terms.

Within the field of political science, the concepts of power expansion and tenure
extension often overlap or are ambiguously applied, contributing to potential
confusion. To facilitate greater conceptual clarity, it is necessary to
distinguish more rigorously between these two frameworks. Power expansion refers
to the process by which an incumbent leader accrues authority beyond their
original constitutional mandate—typically through the centralisation of power,
the weakening of checks and balances, or encroachment upon other state
institutions such as the legislature or judiciary. Tenure extension, by contrast,
refers to efforts by a leader to remain in power beyond the term originally
prescribed, often through constitutional amendments, the cancellation or
manipulation of elections, or other mechanisms used to circumvent term limits.

Existing definitions of autocoups frequently conflate these two dynamics or place
disproportionate emphasis on power expansion. For instance, @cameron1998a defines
a self-coup as “a temporary suspension of the constitution and dissolution of
congress by the executive, who rules by decree until new legislative elections
and a referendum can be held to ratify a political system with broader executive
power” (p. 220). Yet the concept of “broader executive power” is inherently vague
and contestable. Similarly, the term incumbent takeover, defined as “an event
perpetuated by a ruling executive that significantly reduces the formal and/or
informal constraints on his/her power” [@baturo2022, p. 374], draws on earlier
work by @svolik2014 and centres on power expansion. However, the dataset based on
this definition includes cases involving both power expansion and tenure
extension. In contrast, the term overstay is clearly defined as “staying longer
than the maximum term as it stood when the candidate originally came into office”
[@ginsburg2011evasion, p. 1844], thereby focusing specifically on tenure
extension.

This thesis contends that tenure extension should be the primary and defining
characteristic of an autocoup, for several reasons.

First, this focus aligns autocoups conceptually with traditional coups. A classic
coup is typically defined by the forcible and premature removal of a sitting
executive; it does not necessarily involve restrictions on the leader’s powers,
but rather a disruption of tenure. By the same logic, an autocoup should be
defined by the prolongation of tenure, not merely by the expansion of executive
power. An incumbent may experience a reduction in authority without being removed
from office—such a case would not be coded as a coup. Similarly, a leader who
consolidates power without exceeding term limits would fall under the category of
executive aggrandisement, but not an autocoup.

Second, in practice, power expansion often serves as a strategic means to
facilitate tenure extension. The widely cited case of President Alberto Fujimori
in Peru illustrates this dynamic. Although his 1992 movement involved the
suspension of the constitution and the dissolution of Congress, the ultimate
objective was to secure continued rule. The 1993 Constitution allowed him to seek
a second term, which he won in 1995. Subsequently, a law of “authentic
interpretation” passed by his allies in Congress enabled him to run again in
2000—a move shrouded in controversy. Although he secured re-election, Fujimori's
regime ultimately collapsed in 2000 amidst corruption and human rights scandals,
prompting his flight to Japan [@ezrow2019]. In this light, it is illogical for
incumbents to expand their powers unless they intend to remain in office; doing
so only enhances the authority of their successors, whose interests may diverge
from their own.

Third, measuring power expansion poses greater methodological challenges than
identifying tenure extension. For example, @cameron1998a defines a self-coup as
involving both constitutional suspension and congressional dissolution. Yet it is
unclear whether either act alone constitutes an autocoup, or whether both are
required, or even whether they should be coded as separate events. Although the
difficulty of measurement should not preclude consideration of power expansion, a
clear starting point is needed. This thesis therefore prioritises tenure
extension as the definitional core of autocoups, leaving the broader discussion
of power expansion (and its inverse, power contraction) for future research.

Based on these considerations, this study defines an **autocoup** as:

> *The extension of an incumbent leader’s tenure in office beyond the originally
> mandated limit, achieved through either legal manipulation or overtly illegal
> means.*

This definition centres on tenure extension as the core feature, while
acknowledging the possible coexistence of power expansion.

First, the term refers to the actual national leader, regardless of formal
office. For instance, although Vladimir Putin formally stepped down as President
of Russia in 2008 and assumed the premiership, real political power remained with
him. During this period, the presidency—held by Dmitry Medvedev—functioned
largely as a symbolic office under Putin’s continued control [@chaisty2019]. To
avoid arbitrariness, this study adopts the Archigos dataset [@goemans2009] to
determine whether an incumbent has effectively remained in power.

Second, while tenure extension is the definitional cornerstone, this approach
does not exclude concurrent power expansion. Both may occur simultaneously, but
the act of overextending one’s time in office is the decisive criterion. In the
Fujimori case, for example, the 1992 actions were not coded as an autocoup until
the 1993 constitutional amendment that enabled his re-election.

Third, autocoups may be carried out through both legal and illegal means. For
instance, Chadian President François Tombalbaye postponed general elections until
1969 following his rise to power in 1960. Likewise, Angolan President José
Eduardo dos Santos suspended elections throughout his nearly four-decade rule
[@baturo]. These represent clear violations of constitutional norms. However,
other cases—such as Putin’s 2008 manoeuvre—may not be overtly illegal, yet
undermine the constitutional spirit intended to limit consecutive terms. Hence,
this definition emphasises the functional illegitimacy of such actions,
regardless of formal legality, particularly when the incumbent is the clear and
direct beneficiary.

Finally, an incumbent leader who seeks re-election in accordance with the
existing constitution is clearly not engaging in an autocoup. However, if they
subsequently refuse to accept an electoral defeat and remain in power beyond
their legitimate mandate, such an action would qualify as an autocoup.

By clarifying these definitional boundaries, this chapter sets the conceptual
foundation for the autocoup dataset introduced in the next section.

### Introduction to the autocoup dataset

#### Defining the scope {.unnumbered}

Classifying political events as autocoups often requires addressing borderline
cases. To ensure consistency and minimise ambiguity, this study employs a coding
strategy grounded in the definition advanced in the previous section.
Specifically, only those instances in which incumbent leaders extend their
originally mandated term in office are coded as autocoups. Cases involving solely
power expansion without any corresponding tenure extension are excluded from the
dataset.

The temporal scope of the dataset spans from 1945 to the end of 2023, reflecting
the most recent data available at the time of compilation. The geographical
coverage is global, encompassing leaders from all countries and regions
worldwide.

#### Classifying autocoups {.unnumbered}

In categorising autocoups, this study primarily focuses on the methods employed
by incumbents, with the outcomes considered as a secondary classificatory
dimension. Other aspects are coded when relevant information is available.

##### Evasion of term limits {.unnumbered}

One of the most common tactics employed in autocoups is the evasion of term
limits. Incumbents often use ostensibly legal mechanisms to extend their hold on
power, primarily through the manipulation of constitutional provisions. Such
manoeuvres may involve pressuring legislatures or judicial bodies to reinterpret
term limits, amending the constitution to extend terms, or even replacing the
constitution entirely. In some cases, referendums are used to give these actions
a veneer of legitimacy. These extensions may range from a single additional term
to indefinite rule. Key methods include:

**Changing term length:** Incumbents may increase the duration of a single term
(e.g., from four to six years) without altering the number of permitted terms.
Examples include President David Dacko (Central African Republic, 1962),
President Grégoire Kayibanda (Rwanda, 1973), and President Augusto Pinochet
(Chile, 1988).

**Enabling re-election:** This involves amending legal or constitutional
frameworks to allow previously barred re-election. For instance, President Carlos
Menem of Argentina in 1993 modified the constitution to permit himself to run
again, thereby extending his tenure.

**Removing term limits altogether:** An example is President Paul Biya of
Cameroon, who in 2008 successfully abolished presidential term limits, allowing
himself to run for re-election indefinitely.

**Declaring** **leader for life:** Unlike the removal of term limits, this
declaration often retains the appearance of electoral competition, albeit through
rigged or uncontested elections. For example, Indonesia’s President Sukarno
attempted to declare himself president for life in 1963, though ultimately
unsuccessfully.

These strategies are often used in combination. For instance, Haitian President
François Duvalier first amended the constitution in 1961 to permit immediate
re-election and later declared himself president for life in 1964.

##### Election manipulation or rigging {.unnumbered}

The second most frequently employed tactic in autocoups involves the manipulation
of electoral processes to ensure incumbents remain in office.

**Delaying or cancelling elections:** Postponing scheduled elections without
legitimate justification is a frequent strategy. President François Tombal baye
of Chad, for example, delayed general elections until 1969 after assuming power
in 1960. Similarly, Angolan President José Eduardo dos Santos suspended elections
throughout his rule from 1979 to 2017.

**Rejecting unfavourable election outcomes**: Incumbents may refuse to accept
electoral defeat and seek to remain in power through unconstitutional means. A
high-profile case is President Donald Trump of the United States, who refused to
accept the 2020 election results and attempted to overturn them.

**Electoral rigging:** Securing re-election through implausibly high vote shares
is a key indicator of electoral manipulation. This study codes elections in which
incumbents receive over 90% of the vote as indicative of an autocoup. President
Teodoro Obiang of Equatorial Guinea has routinely secured over 95% of the vote in
multiparty elections since 1996.

**Exclusion of opposition:** Preventing opposition parties or candidates from
participating in elections—thereby turning them into uncontested
contests—constitutes a clear sign of an autocoup.

##### Figurehead Installation {.unnumbered}

Some incumbents circumvent term limits by installing a trusted proxy or
figurehead while retaining de facto control over state affairs. This strategy
enables them to relinquish formal office while continuing to govern from behind
the scenes.

A prominent example is the 2008 Russian presidential transition. Confronted with
constitutional term limits, President Vladimir Putin endorsed Dmitry Medvedev as
his successor, who was subsequently elected. Medvedev then appointed Putin as
Prime Minister. Despite this formal role reversal, most analysts concur that
Putin maintained substantial influence, effectively rendering Medvedev a
figurehead.

It is important to note that the identification of such cases must be anchored in
objective criteria to avoid arbitrary classification. Accordingly, this thesis
relies on the Archigos dataset to determine leadership status. If a former
officeholder is not recorded as the country’s leader in Archigos—despite
retaining informal power—they are not coded as engaging in a figurehead-style
autocoup within this dataset.

##### Reassignment of supreme authority {.unnumbered}

This strategy involves restructuring the legal or constitutional system to create
a new, more powerful role, which the incumbent then assumes after formally
leaving office.

In 2017, Turkish Prime Minister Recep Tayyip Erdoğan spearheaded a constitutional
referendum that transformed the country from a parliamentary to a presidential
system. The presidency, under the new system, was endowed with significantly
expanded executive powers. Erdoğan then ran for, and won, the newly empowered
presidency, thus maintaining control under a different institutional guise.

##### One-time extension arrangements {.unnumbered}

In certain cases, special arrangements are enacted to extend an incumbent’s
tenure without altering the broader institutional framework. These extensions are
explicitly tailored to the current officeholder, with constitutional rules on
term limits or tenure intended to resume their normal application for future
leaders. For instance, in 2004, Lebanon extended President Émile Lahoud’s term by
three years through a one-time legal provision that applied solely to his
incumbency.

#### Data coding {.unnumbered}

The autocoup dataset is constructed upon a foundation of established datasets and
scholarly work, ensuring both reliability and comprehensiveness. The primary
sources used for coding are outlined in @tbl-source.

The Archigos dataset [@goemans2009] and the Political Leaders' Affiliation
Database (PLAD) [@bomprezzi2024wedded] provide detailed records of national
leaders from 1875 to 2023. Although this study focuses on events from 1945
onwards, these datasets are instrumental in identifying de facto leaders,
distinguishing them from nominal heads of state.

The Incumbent Takeover dataset [@baturo2022], which integrates data from 11
separate sources, offers a broad collection of cases where executives
significantly reduced constraints on their authority. As this dataset includes
both power expansion and tenure extension cases, cross-referencing with Archigos
and PLAD was necessary to determine whether each case met the criteria for an
autocoup.

```{r}
#| label: tbl-source
#| tbl-cap: Main Data Sources for Coding the Autocoup Dataset
#| echo: false
#| results: asis

read.csv(here("data/autocoup_source.csv")) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(100),
    table.align = "center"
  ) |>
  as_raw_html()
```

```{r}
#| label: data-autocoups
#| echo: false

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))

# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))
# nrow(Autocoup) - nrow(Autocoup |> filter(!is.na(source)))
```

In total, 109 events were coded as autocoups. Of these, 94 overlap with entries
in the Incumbent Takeover dataset. The remaining 15 cases were newly identified
and coded by the author through verification with supplementary sources such as
Archigos, PLAD, and contemporary news reports.

Although the majority of observations originate from the Incumbent Takeover
dataset, this study is not a replication of that work. Over two-thirds of entries
in Incumbent Takeover were excluded, as they involved power expansions
unaccompanied by any attempt to extend tenure—thereby falling outside the
definitional scope of autocoups in this thesis. This distinction represents the
principal departure from the Incumbent Takeover dataset.

The final dataset includes 14 structured variables in addition to a free-text
notes field. The variables are as follows:

-   **Country identification:** Country code (ccode) and country name (country)
    follow the Correlates of War project standard [@stinnett2002].

-   **Leader information:** Name of the de facto leader (leader_name), coded in
    line with the conventions of the Archigos and PLAD datasets.

-   **Timeline variables:** Date the leader assumed power (entry_date), date of
    departure (exit_date), date of the autocoup-defining event (autocoup_date),
    and start date of the extended term (extending_date).

-   **Power transition methods:** Entry method (entry_method), exit method
    (exit_method), and dummy variables for regular or irregular entry
    (entry_regular) and exit (exit_regular).

-   **Autocoup characteristics:** Method of power extension (autocoup_method) and
    outcome of the attempt (autocoup_outcome), categorised as "failed and removed
    from office", "failed but completed original tenure", or "successful". For
    successful cases, the length of the additional term is calculated from the
    difference between exit_date and extending_date.

-   **Data source:** The primary dataset used to code each case (source).

-   **Additional notes:** Contextual explanations for exceptional or borderline
    cases (notes).

Several coding challenges and decisions merit further explanation. In cases where
tenure extensions occurred incrementally, the autocoup_date reflects a defining
event—such as the passage of a constitutional amendment, legislative vote, or
successful referendum. Where leaders made multiple attempts at autocoups, details
are consolidated in the notes field. Special attention was paid to
differentiating between mere power consolidation and clear attempts to prolong
tenure, requiring triangulation across multiple sources. Additionally,
determining the success or failure of an autocoup, particularly in underreported
cases, often necessitated in-depth background research.

#### Data descriptions {.unnumbered}

```{r}
#| label: data_descriptions

country <- Autocoup |>
  distinct(ccode) |>
  summarise(count = n())

autocoup_desc <- Autocoup |>
  summarise(
    attempted = n(),
    succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method
  ) |>
  arrange(-attempted)
```

The primary coding process identified 109 cases of autocoups between 1945 and
2023, spanning 73 countries. This comprehensive dataset provides a robust
foundation for analysing trends and patterns in autocoup attempts across diverse
political and institutional contexts.

A breakdown of the methods employed by incumbents to extend their tenure is
presented in @tbl-autocoup_method.

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup methods and success rates (1945-2023)

autocoup_desc |>
  adorn_totals() |>
  mutate(Success_rate = percent(round(succeeded / attempted, 3))) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(99)
  ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |>
  tab_source_note(source_note = md("*Source: Autocoup dataset*"))
```

The most prevalent method is enabling re-election, accounting for 45 cases. This
is followed by removing term limits (14 cases), and delaying elections and
declaring the leader for life, each with 9 cases. Other methods, such as
cancelling elections or rejecting electoral defeat, are less frequently observed.

Autocoups demonstrate a notably high overall success rate of 80%, in stark
contrast to the approximate 50% success rate of classical coups. Several factors
may account for this disparity. First, incumbents possess direct access to state
resources and institutional levers that can be mobilised in their favour. Second,
unlike the abrupt nature of traditional coups, autocoups are often gradual and
strategic, allowing time to consolidate elite support and cultivate public
legitimacy. Third, many autocoup tactics are cloaked in legal or constitutional
procedures, reducing overt resistance and complicating opposition efforts.
Finally, incumbents typically exercise considerable control over state
institutions—including the judiciary, legislature, and security apparatus—which
can facilitate the execution and consolidation of such moves.

Nevertheless, success rates vary significantly depending on the method employed.
Certain strategies appear to be consistently effective. For example, removing
term limits, delaying elections, declaring oneself leader for life, and
cancelling elections all demonstrate a 100% success rate within the dataset.
Notably, these methods represent some of the most overt violations of
constitutional norms governing leadership succession. This pattern suggests that
success is influenced more by the underlying balance of power than by the degree
of legal or constitutional violation. In other words, it is not the illegality of
the act that determines success, but rather the extent of the incumbent's control
over political institutions. Leaders who command overwhelming authority are often
willing and able to blatantly defy established processes, precisely because their
dominance over the state apparatus shields them from effective resistance.

By contrast, refusing to accept election results exhibits the lowest success
rate, with only one out of four attempts proving successful. Although the sample
size is small, this pattern may reflect the greater institutional resilience of
electoral democracies, stronger civil society mobilisation, heightened
international scrutiny, and the inherently high-risk nature of attempting to
overturn electoral outcomes. These factors may increase the likelihood of failure
for incumbents who pursue this path.

### Case studies

#### High frequency and success rate of autocoups in post-communist regimes

The dataset reveals a notably high frequency and success rate of autocoup
attempts in post-communist countries. These states, formerly governed under
communist rule prior to the collapse of the Soviet Union, have predominantly
transitioned into ‘hybrid regimes’ [@nurumov2019], with only a small number—such
as China—retaining their communist identity. Within these contexts, the dataset
records 12 autocoup attempts aimed at prolonging incumbency, of which only two
failed. An examination of these cases highlights several shared structural and
political characteristics.

First, many post-communist regimes inherited authoritarian institutional
frameworks. Although most transitioned formally to non-communist governments,
they often retained significant authoritarian features from their communist past,
particularly in the consolidation of executive power.

Second, elite continuity has been a defining feature of these transitions. Rather
than involving a decisive rupture with the previous regime, political transitions
in many post-communist countries saw former communist elites maintain their
positions of influence, often rebranding themselves within newly formed or
nominally democratic institutions.

Third, post-communist regimes have frequently subverted democratic processes.
While elections and constitutional term limits were introduced during
democratisation efforts, the institutional legacies of communist rule have often
facilitated the manipulation of electoral outcomes and the circumvention of term
limits [@nurumov2019].

##### Case 1: Lifelong ruler: Alexander Lukashenko in Belarus {.unnumbered}

Alexander Lukashenko, a former member of the Supreme Soviet of the Byelorussian
SSR, rose to national prominence as the head of Belarus's interim anti-corruption
committee following the dissolution of the Soviet Union. In 1994, he was elected
as the first president of Belarus, a position he has retained ever since. The
original 1994 constitution imposed a two-term limit on the presidency; however,
this restriction was abolished in 2004 via constitutional amendment.

Since his first electoral victory, international observers have consistently
deemed Belarusian elections neither free nor fair. Despite widespread domestic
opposition and recurring protests, Lukashenko has repeatedly claimed re-election
victories, often securing over 80% of the vote. This pattern is indicative of a
broader regional trend across Central Asia, where many post-Soviet
leaders—previously high-ranking officials within the communist
apparatus—transitioned into the presidency and maintained power with limited
institutional challenge.

##### Case 2: Dynastic succession: Nursultan Nazarbayev in Kazakhstan {.unnumbered}

Nursultan Nazarbayev served as the first president of independent Kazakhstan from
1991 to 2019. Prior to this, he held de facto leadership as First Secretary of
the Communist Party of Kazakhstan, placing him at the helm of political power
both before and after the dissolution of the Soviet Union. Following
independence, Nazarbayev was elected president and remained in power through
various constitutional and legal manoeuvres, including the implementation of new
constitutions that effectively reset term limits.

Significantly, Nazarbayev did not eliminate term limits outright. Instead, a
constitutional exemption was introduced exclusively for the “First President”,
allowing him to bypass term constraints while maintaining the appearance of legal
continuity [@nurumov2019]. In contrast to Lukashenko, who remains in office,
Nazarbayev formally stepped down in 2019, transferring the presidency to his
designated successor, Kassym-Jomart Tokayev. Nevertheless, Nazarbayev retained
substantial political influence, particularly in his role as Chairman of the
Security Council, a position he held until 2022—underscoring the persistence of
informal executive dominance even after formal succession.

```{r}
#| label: latin
#| include: false

library(countrycode)
latin <- codelist |>
  filter(region == "Latin America & Caribbean") |>
  select(region, cown) |>
  drop_na() |>
  pull(cown)

Autocoup |>
  filter(ccode %in% latin) |>
  summarise(
    count = n(),
    .by = country
  ) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(
    count = n(),
    .by = autocoup_method
  ) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(
    count = n(),
    .by = c(autocoup_method, autocoup_outcome)
  ) |>
  adorn_totals()
```

#### Autocoups for immediate re-election: Cases of Latin America

Latin America has a long-standing tradition of constitutional term limits on
executive power. Simón Bolívar, regarded as a founding father of several Latin
American republics, initially championed this principle, declaring in 1819 that
“nothing is as dangerous as allowing the same citizen to remain in power for a
long time... That is the origin of usurpation and tyranny” [@ginsburg2019, p.
38]. Although Bolívar later altered his position—asserting in his 1826 speech to
the Constitutional Assembly that "a president for life with the right to choose
the successor is the most sublime inspiration for the republican order"—the
notion of term limits became embedded in the region’s political culture. Indeed,
approximately 81% of Latin American constitutions between independence and 1985
included some form of presidential term limitation [@marsteintredet2019a].

An analysis of autocoup cases in the region reveals two notable patterns
concerning re-election dynamics.

##### Frequent success in breaking non-re-election rules {.unnumbered}

Unlike in many presidential systems where two consecutive terms are the norm,
Latin American constitutions historically favoured more restrictive models.
According to @marsteintredet2019a, 64.9% of constitutions in the region between
independence and 1985 prohibited immediate re-election, while 5.9% forbade
re-election altogether.

Nonetheless, adherence to these rules has varied. Countries such as Mexico, which
introduced a strict non-re-election clause in 1911 at the outset of the Mexican
Revolution, have upheld this principle consistently [@klesner2019]. Similarly,
Panama and Uruguay have never amended their re-election provisions, while Costa
Rica has allowed immediate re-election only briefly (1897–1913) since banning it
in 1859 [@marsteintredet2019a]. In contrast, other states have frequently amended
or circumvented constitutional term limits.

The pursuit of re-election or immediate re-election has been a key catalyst for
autocoup attempts aimed at extending executive power. This study identifies 32
autocoup cases in Latin America, of which 17 (over 50%) involved attempts to
enable re-election or immediate re-election. Among these, 10 were successful—a
success rate of approximately 59%.

Importantly, not all such leaders sought indefinite rule. Many stepped down after
securing and completing a second term. Notable examples include: Fernando
Henrique Cardoso (Brazil, 1995–2003); Danilo Medina (Dominican Republic,
2012–2020); Juan Orlando Hernández (Honduras, 2014–2022) [@ginsburg2019;
@marsteintredet2019a; @landau2019; @baturo2019; @neto2019].

##### Resistance to further extensions {.unnumbered}

The relative restraint shown by many leaders should not be interpreted as an
absence of ambition for further tenure. Rather, it suggests that additional
extension attempts often failed, and that incumbents accepted these outcomes
without resorting to overt manipulation or repression.

While autocoups enabling a single additional term tend to be relatively
successful, attempts to prolong tenure beyond two terms face greater resistance
and are more prone to failure.

Two contrasting cases illustrate these divergent trajectories:

**Unsuccessful extension:** Carlos Menem (Argentina): President Menem
successfully secured a second term through a 1994 constitutional amendment that
allowed one re-election. He was re-elected in 1995. However, his subsequent
attempt to reset his term count—arguing that his first term (1988–1995) fell
under a previous constitutional framework—was unanimously rejected by the Supreme
Court in 1999 [@llanos2019]. A similar case occurred with President Álvaro Uribe
in Colombia (2002–2010) [@baturo2019].

**Successful extension:** Daniel Ortega (Nicaragua): In contrast, Nicaraguan
President Daniel Ortega successfully extended his tenure through a series of
judicial and legislative actions. In 2009, the Supreme Court of Justice
authorised his re-election in 2011. Subsequently, in 2014, the National Assembly
passed constitutional amendments abolishing presidential term limits, enabling
Ortega to seek unlimited five-year terms. He has remained in power since 2007
[@close2019].

#### As common as classical coups: Autocoups in African countries

```{r}
#| label: africa
#| include: false

africa <- codelist |>
  filter(continent == "Africa") |>
  select(continent, cown) |>
  drop_na() |>
  pull(cown)

Autocoup |>
  filter(ccode %in% africa) |>
  summarise(
    count = n(),
    .by = country
  ) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% africa) |>
  distinct(country)
Autocoup |>
  filter(ccode %in% africa) |>
  summarise(
    count = n(),
    .by = autocoup_method
  ) |>
  adorn_totals()
Autocoup |>
  # filter(ccode %in% africa) |>
  summarise(
    count = n(),
    .by = autocoup_outcome
  ) |>
  adorn_totals()
```

Classical coups have historically been widespread across the African continent,
accounting for approximately 45% of all coups globally—219 out of 491 recorded
incidents since 1950—affecting 45 out of 54 African states [@powell2011].
Although autocoups occur less frequently than traditional coups, they nonetheless
represent a significant political phenomenon in Africa. Of the 110 documented
autocoup cases worldwide, 46% (51 cases) have taken place in Africa, involving 36
countries. Remarkably, the success rate of autocoups in Africa exceeds 84% (43
out of 51), surpassing not only the regional success rate of classical coups
(approximately 50%) but also the global average success rate of autocoups (79%).

Identifying a clear and consistent pattern for autocoups in Africa proves
difficult—mirroring the analytical complexity associated with classical coups.
Several explanatory factors have been proposed in the literature:

**Natural resource wealth:** Countries rich in oil, diamonds, or other strategic
commodities often see incumbents more inclined to pursue and succeed in term
extensions [@posner; @cheeseman2015; @cheeseman2019a].

**Democratic quality:** The degree of democratic consolidation strongly
influences respect for constitutional term limits [@reyntjens2016].

**International influence:** External actors, including donors and international
organisations, may exert pressure to deter leaders from violating term limits
[@brown2001; @tangri2010].

**Opposition strength and party cohesion:** The ability of the opposition to
organise effectively, and the president’s success in maintaining unity within the
ruling party, are also decisive factors [@cheeseman2019].

Drawing on the Africa Executive Term Limits (AETL) dataset, @cassani2020
identifies human rights abuses and the desire to evade accountability as key
drivers behind attempts to overstay in power. The more authoritarian the
leadership style, the more likely the incumbent is to challenge constitutional
limits. Additionally, a leader’s ability to command loyalty from the military,
often through strategic public investment, significantly increases the likelihood
of successful tenure extension.

Although both coups and autocoups remain prevalent, a notable shift has occurred
since the end of the Cold War in 1991. The frequency of traditional coups has
declined, while autocoups have become increasingly common. This trend is partly
attributable to the introduction of multiparty electoral systems in the 1990s,
which were frequently accompanied by the institutionalisation of presidential
term limits [@cassani2020; @cheeseman2019].

Prior to 1991, personalist or military rule predominated across much of the
continent, and term limits were rarely codified. In the post-Cold War period, the
proliferation of democratic frameworks led to an increase in term-limit
provisions—and, correspondingly, a rise in attempts to circumvent them. It is
important to emphasise, however, that this uptick in term-limit challenges does
not necessarily indicate a decline in compliance. On the contrary, overall
turnover in executive leadership has increased compared to previous decades,
suggesting that while term-limit violations attract attention, adherence to
constitutional constraints has also become more common.

### Summary

This chapter has presented a comprehensive analysis of autocoups, focusing on
political events in which incumbent leaders extend their tenure beyond
constitutional limits. By refining existing definitions and distinguishing
autocoups from related concepts—such as self-coups, autogolpes, and executive
takeovers—the study introduces a novel dataset cataloguing 109 autocoup cases
from 1945 to 2023, of which 87 were successful.

The revised conceptual framework, together with the newly compiled dataset,
enables a broader and more systematic analysis of irregular leadership
transitions. Whereas traditional research has largely concentrated on the
premature termination of leadership through coups, this study expands the
analytical lens to include irregular tenure extensions. This approach offers a
more nuanced understanding of how incumbent leaders may subvert constitutional
norms and democratic processes to entrench their authority.

Nonetheless, several limitations should be acknowledged. First, the proposed
definition of an autocoup requires further scholarly engagement and debate before
it can be widely adopted. Despite efforts to ensure objectivity, certain coding
decisions—particularly in ambiguous or borderline cases—may unavoidably involve
subjective judgement. Second, while this thesis focuses on tenure extensions,
events involving the expansion of executive power—without a formal extension of
tenure—also merit deeper conceptual and empirical exploration.

Despite these constraints, the research makes a substantive contribution to the
literature on political stability, leadership dynamics, and democratic
resilience. The dataset offers a valuable empirical foundation for future
research into the mechanisms and motivations behind autocoups.

Several promising avenues for further inquiry emerge. Future studies could
leverage the dataset to explore long-term institutional impacts, including the
role of autocoups in democratic backsliding, authoritarian consolidation, and the
personalisation of power. In addition, comparative analyses of autocoups and
classical coups may shed light on the evolving strategies of power consolidation
in different regime types and political contexts.

## Power Dynamics and Autocoup Attempts {#sec-chapter2}

### Abstract {.unnumbered}

This chapter investigates the determinants of autocoup attempts in order to
better understand the political dynamics underpinning tenure extensions by
incumbent leaders. Employing a probit regression model, the analysis explores the
relationship between regime type and the likelihood of autocoup occurrence, with
particular attention to the balance of power between incumbents and institutional
constraints or potential challengers from within the state apparatus.

The findings indicate that regime type constitutes a significant predictor of
autocoup attempts. In particular, personalist regimes are shown to be
substantially more likely than dominant-party regimes to engage in autocoups,
controlling for other factors. This increased propensity is attributed to the
weaker institutional constraints, greater executive dominance, and heightened
intra-elite rivalries that typify personalist rule.

***Keywords:** Autocoups, Coup, Regime types, Tenure Extension, Authoritarianism*

\newpage

### Introduction

As discussed in Chapter 2, the study of autocoups has been hindered by conceptual
ambiguity and a lack of systematic data, resulting in limited empirical analysis
of this phenomenon. To address this gap, the present chapter seeks to make a
substantive contribution by offering a quantitative analysis of the determinants
of autocoup attempts. Following the precedent set by empirical research on
classic coups, which has primarily focused on the factors precipitating coup
attempts [@gassebner2016], this chapter similarly investigates why some incumbent
leaders choose to extend their tenure through autocoups, while others do not.

The rationale for investigating the determinants of autocoups is threefold.
First, autocoups represent one of the most frequent forms of irregular leadership
transition, with over 100 documented cases since 1945 (as detailed in Chapter 2).
Their prevalence has increased notably since 2000, coinciding with the apparent
global decline in classic coups [@bermeo2016; @thyne2019]. Second, autocoups play
a critical role in shaping political stability and democratic trajectories,
frequently resulting in long-term institutional erosion. Third, understanding the
factors that drive autocoup attempts constitutes a foundational step for further
enquiry into their consequences. Without identifying the conditions under which
autocoups occur, it is difficult to assess how to prevent them or mitigate their
negative effects.

While autocoups differ fundamentally from classic coups—primarily in terms of the
actor initiating the event (the incumbent, rather than a power challenger)—the
two forms of power disruption share important features. Consequently, methods
commonly applied to the study of traditional coups can be usefully adapted for
the analysis of autocoups. Nevertheless, despite an extensive literature on coup
dynamics [@gassebner2016], many studies treat regime type as a mere background
condition or control variable, rather than as a central determinant of power
disruptions.

This chapter advances the argument that the risk of autocoup attempts is
significantly shaped by the structural balance of power embedded within a
regime's foundational type. It contends that the likelihood of an autocoup is
determined by the perceived power equilibrium between the incumbent leader and
potential institutional or elite challengers. This power configuration is
relatively stable over time, as it is both shaped by and constitutive of the
regime type itself [@geddes2014]. Accordingly, this study employs regime type as
a proxy for underlying power dynamics, enabling a more systematic investigation
of the structural conditions that facilitate or constrain autocoup attempts.

To empirically test this proposition, the chapter employs a probit regression
model to examine the strategic calculus of incumbent leaders. It argues that the
balance of power, as inferred from regime typology, is a key determinant of
whether an incumbent will initiate an autocoup.

Given the inherent difficulty in directly measuring the internal balance of power
within regimes, regime type is used as an operational proxy. The premise
underpinning this approach is that regime type fundamentally conditions the
distribution of authority, the strength of institutional constraints, and the
capacity of incumbents to subvert constitutional norms. By analysing variation
across regime types, this study provides deeper insight into the structural
factors that influence both the occurrence and outcome of autocoup attempts.

In light of the limited quantitative research on autocoups in the existing
literature, this thesis aims to make a pioneering contribution to their empirical
study by offering a theoretically grounded and methodologically robust analysis
of their determinants.

The remainder of this chapter is organised as follows. Section 2 examines the
dynamics and patterns of autocoup attempts and their outcomes. Section 3 outlines
the research design, methodology, and variables used in the analysis, offering a
clear account of the empirical strategy. Section 4 presents and interprets the
findings, identifying key trends and implications. Section 5 concludes the
chapter by summarising the main insights and considering their relevance for
understanding and mitigating the risks of autocoups.

### Dynamics of autocoup attempts

Much like traditional coup attempts, autocoups are driven by two fundamental
factors: the disposition of incumbent leaders—namely, their motivations and
willingness to act—and their capability, referring to the resources and
opportunities available to them. However, autocoups exhibit two notable features
that distinguish them from classical coups. First, whereas coups predominantly
occur in autocracies [@thyne2014], approximately one-third of recorded autocoups
have taken place in democracies, as documented in Chapter 2. Second, while the
success rate of traditional coups hovers around 50%, more than 80% of autocoup
attempts have succeeded, according to the autocoup dataset introduced in Chapter
2. These distinctions suggest that the disposition and capability dynamics
underpinning autocoups differ substantially from those governing traditional
coups. This section examines the complex dynamics of autocoup attempts, focusing
particularly on how the motivations of incumbents, the factors influencing
success, and the institutional structures of regime types shape a state's
vulnerability to such actions.

#### Motivations for autocoups {.unnumbered}

Incumbent leaders seeking to extend their tenure may be motivated by a variety of
considerations, which can broadly be categorised into three principal domains:
personal ambition, purported national interest, and self-preservation.

First, the pursuit of personal power serves as a powerful incentive for many
leaders. The ability to govern without institutional constraints grants them
control over national policy, access to significant state resources, influence
over the judiciary and legislature, and the prestige associated with continued
office. For some, the desire to secure a lasting political legacy—to be
remembered as a transformative figure in national history—further reinforces the
appeal of prolonged rule.

Second, incumbents often justify tenure extensions as being in the national
interest. A common argument posits that a single term is insufficient to complete
long-term reform or development agendas. In this framing, remaining in office is
presented as necessary for the successful realisation of ongoing projects,
thereby positioning the autocoup as a sacrifice for the greater good rather than
a personal power grab.

Third, autocoups can function as a mechanism for self-preservation. Leaders who
face potential prosecution for corruption, human rights abuses, or other crimes
may view tenure extension as a means of maintaining immunity. Moreover, those who
have accumulated significant political enemies during their time in power may
fear retaliation upon stepping down. In such cases, extending one’s rule is not
only about ambition but also about survival—shielding oneself from legal or
political retribution.

#### Power dynamics and autocoups {.unnumbered}

While motivations may spark an incumbent’s decision to initiate an autocoup, it
is often the capability to execute and sustain such an action that proves more
decisive. The relatively high frequency and strikingly high success rate of
autocoups—over 80%, compared to around 50% for traditional coups—suggest that
incumbents enjoy structural advantages when seeking to extend their tenure. These
advantages are not confined to autocracies but are also observable in
democracies, highlighting that the resources and institutional leverage available
to incumbents vary significantly across regimes.

This observation necessitates a closer examination of state power
structures—particularly the distribution of control over the armed forces.
Military allegiance is a critical determinant of autocoup success. If the armed
forces remain loyal to the incumbent, opposition—whether from civil society, the
judiciary, or the legislature—can be neutralised or marginalised. Conversely,
open defiance by military leaders or a refusal to intervene on behalf of the
executive can render an autocoup untenable.

However, it would be reductive to assume that nominal authority as
commander-in-chief translates into uncontested control. Just as it is simplistic
to attribute the occurrence of traditional coups solely to military power
[@singh2016], it is equally misleading to presume that incumbents uniformly wield
the loyalty of the armed forces. Formal titles belie the often complex and
fragile relationships that underpin military loyalty.

In autocratic contexts, the military may not adhere to constitutional norms, but
nor is it inherently loyal to the head of state. Executives rely on military
officers to implement commands, yet those officers may possess their own
ambitions or rival loyalties. For example, in 1971, Ugandan President Milton
Obote attempted to dismiss General Idi Amin. In response, Amin capitalised on his
influence within the armed forces to stage a successful coup, ousting Obote
[@sudduth2017].

In contrast, consolidated democracies typically institutionalise military
allegiance to the constitution rather than to individual officeholders. In the
United States, following the 2020 presidential election, General Mark Milley,
Chairman of the Joint Chiefs of Staff, publicly reaffirmed the military’s
commitment to constitutional principles:

> “We are unique among militaries. We do not take an oath to a king or a queen, a
> tyrant or a dictator. We do not take an oath to an individual. We take an oath
> to the Constitution.”
>
> — (US Army Museum, 12 November 2020[^1])

In hybrid or fragile democracies, attempts to extend executive tenure can carry
higher risks. For example, in Niger (2009), President Mamadou Tandja’s efforts to
amend the constitution and secure a third term provoked a military coup in 2010
[@miller2016]. Similarly, in Honduras (2009), President Manuel Zelaya was removed
from office by the military after attempting to modify the constitution to allow
immediate re-election [@muñoz-portillo2019].

#### Regime types and autocoups {.unnumbered}

Given these complexities, a more effective analytical strategy involves assessing
the broader balance of power within political systems. Direct observation of this
balance is inherently difficult, so this study employs regime type as a proxy—a
method consistent with established approaches in comparative politics. Regime
types reflect the institutional architecture of power distribution, especially
regarding authority over military command, political appointments, and
policy-making.

Drawing on the classification by @geddes2014, autocratic regimes can be divided
as follows:

**Military regimes** are governed by a junta, typically composed of senior
military officers who collectively determine leadership and policy. Examples
include Brazil (1964–1985), Argentina (1976–1983), and El Salvador (1948–1984)
[@geddes1999].

**Personalist regimes** are centred around a dominant individual who exercises
unchecked authority over military, policy, and succession processes. Notable
cases include Rafael Trujillo in the Dominican Republic (1930–1961), Idi Amin in
Uganda (1971–1979), and Jean-Bédel Bokassa in the Central African Republic
(1966–1979) [@geddes1999].

**Dominant-party regimes** entrust authority to a structured political party,
with the leader operating within, or as the head of, that organisation. Prominent
examples include the PRI in Mexico, CCM in Tanzania, and Leninist parties in
Eastern Europe [@geddes1999].

Among these, personalist regimes are particularly conducive to autocoups. The
concentration of power reduces institutional resistance, and loyalty from the
armed forces is often cultivated through personal patronage. Military regimes,
though rooted in armed power, frequently experience internal instability, making
them more vulnerable to traditional coups than to autocoups. Dominant-party
regimes are more ambiguous: while the leader may be constrained by party
structures, exceptionally strong party leaders may nonetheless initiate
autocoups—as evidenced by Xi Jinping (2018), operating under dominant-party
regimes.

Monarchies, though classified as autocracies, typically render autocoups
irrelevant, as monarchs usually serve for life by design.

An important clarification arises at this point: why would leaders in personalist
regimes feel compelled to extend their tenure, given the extensive authority they
already wield? The answer lies in the distinction between the scope of power and
its duration. While personalist rulers often exercise substantial control over
state institutions, many originally assume office through constitutional or legal
means and must incrementally consolidate power. In this context, autocoups serve
as formal mechanisms to institutionalise their dominance, converting de facto
authority into de jure permanence. This dynamic is exemplified by the multiple
tenure extensions pursued by Vladimir Putin and Alexander Lukashenko.

In post-Soviet Russia, President Boris Yeltsin oversaw the transformation of a
parliamentary system into a personalist regime. However, Yeltsin himself did not
overstay his term; instead, he selected Vladimir Putin as his successor. Upon
assuming office in 2000, Putin progressively extended his rule and has remained
in power since, utilising constitutional amendments and legal manoeuvres to
circumvent term limits.

Similarly, in Belarus, Alexander Lukashenko was elected president in 1994 under a
party-based regime. Within a year, he dismantled the existing political structure
and established a personalist system. Since then, Lukashenko has remained in
power through successive extensions of his tenure, further entrenching his
authority over time.

In democracies, autocoups exclusively occur in presidential systems. This
reflects the institutional leverage enjoyed by presidents, who are directly
elected, often command the armed forces, and may override or bypass legislative
decisions. Prime ministers in parliamentary systems, by contrast, are more
constrained. They are typically beholden to legislative confidence and may be
dismissed via no-confidence votes. Military command is often institutionally
separated from their authority. Consequently, prime ministers face more frequent
leadership turnover and fewer opportunities to extend their terms. For example,
in the UK, three prime ministers served in 2022; in Japan, there have been 36
prime ministers since 1945—an average of one every two years. By contrast, only
14 presidents have served in the United States over the same period, reflecting
greater institutional stability. Given these distinctions, presidential systems
provide a more conducive environment for autocoups—even in established
democracies—due to their centralised executive authority and command over the
military.

From this analysis, the following hypothesis is proposed:

> **H2-1: *The likelihood of autocoup attempts varies significantly by regime
> type, with personalist autocracies and presidential democracies being the most
> susceptible, relative to other regime types.***

### Research design

#### Methodology {.unnumbered}

Given the binary nature of the dependent variable—whether an autocoup is staged
or not—this section initially employs a probit regression model to analyse the
determinants of autocoup attempts. This approach enables the identification of
statistically significant factors influencing the likelihood of such events, as
well as the direction and magnitude of their effects.

However, since the dataset comprises over 5,000 observations with only
approximately 100 instances of autocoups, the dependent variable qualifies as a
rare event. Standard maximum likelihood estimation methods such as probit or
logit may underestimate the probability of occurrence in such cases. To address
this issue and ensure more reliable inference, the analysis additionally applies
Firth’s Bias-Reduced Penalised Maximum Likelihood Estimation, as proposed by
@firth1993David.

#### Data and variables {.unnumbered}

The principal dataset, covering autocoups and regime types, spans the period from
1945 to 2023. However, due to issues in data matching, there is some truncation,
and the actual range of usable data extends from 1945 to 2018. The dataset
encompasses approximately 6,000 observations, of which around 100 are instances
of autocoup attempts.

##### Dependent variable {.unnumbered}

The analysis utilises the autocoup dataset introduced in Chapter 2. This dataset
covers the period from 1945 to 2023 and records information on 107 autocoup
attempts. Descriptive statistics for these attempts, along with regime
classifications, are presented in Chapter 2.

**Autocoup attempt**: A binary variable indicating whether an autocoup attempt
occurred (coded as 1) or did not occur (coded as 0) in each country-year
observation.

##### Independent variables {.unnumbered}

As discussed, regime type constitutes the primary explanatory focus of the model.
Regimes are categorised according to the typology developed by @geddes2014 (GWF
dataset), distinguishing between military, personalist, and dominant-party
regimes, with democracies serving as the reference category.

Control variables are selected based on previous research into coup determinants,
predominantly encompassing factors such as economic performance, political
violence, and tenure in office. In addition, population size and a Cold War dummy
variable are included as further controls.

**Economic Level:** Measured by GDP per capita, representing the general economic
wellbeing of a country. Data are sourced from the V-Dem dataset by @fariss2022,
expressed in constant 2017 international dollars (PPP, per thousand).

**Economic Performance:** Assessed through the current-trend ($CT$) ratio, as
developed by @krishnarajan2019. This metric compares a country’s current GDP per
capita with the average GDP per capita over the preceding five years. A higher
$CT$ ratio signifies stronger economic growth. Formally:

$$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
$$

**Political stability:** Captured by a violence index reflecting the incidence of
internal and interstate conflict, derived from the “actotal” variable in the
Major Episodes of Political Violence dataset [@marshall2005current]. Scores range
from 0 (complete stability) to 18 (maximum instability).

**Days in office (log):** This variable records the natural logarithm of an
incumbent leader’s cumulative days in office in each year. Longer tenure is
posited to facilitate power consolidation, thereby making autocoup attempts more
feasible. Data are drawn from the Archigos dataset [@goemans2009] and the
Political Leaders' Affiliation Database (PLAD) [@bomprezzi2024wedded].

**Population size:** The natural logarithm of national population is included to
control for the potential impact of population size on leadership stability.
Larger populations may pose greater governance challenges and amplify opposition.
Population data are sourced from the V-Dem dataset.

**Cold War:** Following the approach of prior studies [@thyne2014;
@derpanopoulos2016; @dahl2023], a dummy variable, Cold War, is included to
capture differences across historical periods. This variable reflects the
relative scarcity of autocoup events between the 1960s and 1990, with a notable
increase thereafter.

Unlike the analysis of coup determinants, which could theoretically occur in any
given year, I assume that an autocoup happens only once during an incumbent
leader's tenure, as a successful autocoup negates the need for another attempt.
However, this assumption does not always reflect reality, as leaders might
attempt further extensions or try again after a failed attempt. For simplicity, I
overlook these possibilities in our analysis.

Therefore, in our probit model, the unit of analysis for autocoups is the entire
tenure of a leader, rather than a country-year. I establish a base year for the
variables: for leaders who staged an autocoup, we use the year of their first
attempt as the base year; for leaders who did not attempt to overstay, I use the
middle year of their tenure as the base year.

#### Results and discussions

```{r}
#| label: tbl-autocoupmodel
#| tbl-cap: "Determinants of autocoup attempts and success (1945-2018)"
#| results: asis
#| echo: false

autocoup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_model.csv")

model_logit <- glm(
  autocoup_attempt ~ regime_autocoup + GDP_pc + GDP_trend + log(cumulative_days_in_office) + violence + factor(cold_war) + polity5,
  family = binomial,
  data = autocoup_model |>
    mutate(regime_autocoup = relevel(factor(regime_autocoup), ref = "military"))
)
summary(model_logit)

model_logit <- glm(
  autocoup_attempt ~ regime_autocoup + GDP_pc + GDP_trend + violence + pop_log + log(cumulative_days_in_office) + factor(cold_war),
  family = binomial,
  data = autocoup_model |>
    mutate(regime_autocoup = relevel(factor(regime_autocoup), ref = "presidential"))
)

model_brg <- glm(
  autocoup_attempt ~ regime_autocoup + GDP_pc + GDP_trend + violence + pop_log + log(cumulative_days_in_office) + factor(cold_war),
  family = binomial(link = "logit"),
  method = "brglmFit",
  data = autocoup_model |>
    mutate(regime_autocoup = relevel(factor(regime_autocoup), ref = "presidential"))
)

summary(model_brg)



stargazer(model_logit, model_brg,
  warning = FALSE,
  header = FALSE,
  type = "text"
)

stargazer(model_attempt, model_success,
  header = F,
  float = F,
  column.labels = c("Autocoup Attempts", "Autocoup Outcome"),
  intercept.bottom = F,
  covariate.labels = c(
    "Constant",
    "Regime: Dominant-party",
    "\\hspace{1.6cm}Military",
    "\\hspace{1.6cm}Personalist",
    "GDP per capita",
    "Economic trend",
    "Political stability",
    "Age",
    "Population(log)"
  ),
  dep.var.caption = "",
  dep.var.labels.include = F,
  summary.logical = T,
  column.sep.width = "50pt"
)
# view(autocoup_model)

autocoup_model |>
  select(regime_autocoup) |>
  tabyl(regime_autocoup)
distinct(regime_autocoup, .keep_all = TRUE)
```

@tbl-autocoupmodel summarizes the findings from the probit regression models
based on our analysis of the determinants of autocoup attempts and their success.

Model 1, which examines autocoup attempts, reveals only one significant predictor
besides the constant term. Among the regime types, personalist regimes
significantly increase the likelihood of autocoup attempts, all else being equal.
This suggests that leaders in personalist regimes are more prone to attempt to
extend their power through autocoups compared to leaders in democratic regimes
(reference regime). Leaders in dominant-party and military regimes, however, show
no significant difference in the likelihood of attempting an autocoup compared to
democratic leaders.

The model for autocoup success (Model 2) shows similar dynamics. Personalist
regimes again have a strong positive and significant effect on the success of
autocoups compared to democratic leaders. Dominant-party regimes also show a
positive and marginally significant effect. However, a detailed examination
reveals that about half of the successful autocoups in dominant-party regimes (9
out of 20) exhibit a personalist style, such as "party-personal-military"
regimes.

This outcome is logical since personalist leaders are typically much more
powerful than other types of leaders, making them more inclined and capable of
overstaying in power.

Other factors play an insignificant role in determining the attempts and outcomes
of autocoups. This aligns with our conclusions on the determinants of classic
coups. Both coups and autocoups are significantly affected by power dynamics. As
power transitions involve the struggle between seizing and maintaining power, the
balance of power status quo inevitably matters in both coups and autocoups. This
also explains the high success rate of autocoups. Compared to power challengers,
incumbents are in an obviously advantageous position. Incumbent leaders can use
state power to their benefit, which is difficult to counteract. Even the abuse of
power is often unchecked under a powerful leader's rule.

The empirical analysis of autocoups yields significant implications for
real-world politics. In particular, the high overall success rate of autocoups
highlights the vulnerability of democratic institutions to gradual erosion by
incumbent leaders. The threshold for ousting or impeaching an incumbent leader
through constitutional means is exceptionally high, with success often requiring
more than a simple majority and substantial support across various sectors.
Resorting to illegal means, such as a coup, presents even greater challenges due
to high costs, severe consequences, and a low likelihood of success.

Conversely, political dynamics, whether in democracies or autocracies, tend to
favour incumbents even when they act unconstitutionally. Incumbents can leverage
state resources to achieve their political ambitions, benefiting from a high
probability of success and minimal consequences in case of failure. This
asymmetry in power and risk creates a concerning scenario: for incumbents who do
not respect constitutional institutions, the opportunity to launch an autocoup
appears sufficiently low-risk to warrant an attempt.

## Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-installed and Autocoup Leaders

### Abstract {.unnumbered}

This chapter examines the relationship between the methods of power acquisition
and the tenure of leaders who ascend to power through unconventional means, with
a particular focus on coup-installed and autocoup leaders. The central hypothesis
posits that the mode of accession has a profound impact on leadership longevity.
Utilizing Cox proportional hazards and time-dependent Cox models, this study
provides robust empirical evidence of disparate survival times between these two
leader types. The findings reveal that, on average, coup-installed leaders are
2.23 times more likely to be ousted from power than autocoup leaders, all else
being equal. These results have far-reaching implications for political stability
and democratic processes, suggesting that the perceived low costs and high
rewards associated with autocoups may incentivize incumbents to prolong their
tenure through this means, potentially contributing to democratic erosion. This
research makes a notable contribution to the academic literature by offering
nuanced insights into the dynamics of irregular leadership transitions and
enhances our understanding of the complex interplay between power acquisition
methods and leadership longevity.

***keywords**: Coups, Autocoups, Leadership Survival, Cox Model*

\newpage

### Introduction

The enduring fascination with the longevity of political leaders has sparked
extensive research in political science, with scholars seeking to understand why
some leaders maintain power for decades while others are ousted in a matter of
months or even days. However, a specific subset of leaders—those who ascend to
power through coups or extend their tenure through autocoups—has received
relatively limited attention. Examining the tenures of these leaders is crucial,
as it sheds light on the dynamics of irregular leadership transitions and their
implications for political stability and democratic processes.

In contrast to leaders who attain power through conventional means, those who
rise through irregular channels, such as coups or autocoups, present more complex
and intriguing cases for study. The Archigos dataset highlights the prevalence of
irregular power transitions. Between 1945 and 2015, over half of leaders who
assumed power irregularly also exited irregularly, a rate significantly higher
than that of leaders who accessed office through regular channels.

Coup-installed and autocoup leaders constitute a substantial portion of these
irregular cases. The Archigos dataset notes that of 374 leaders who exited
irregularly, 246 (65.8%) were ousted through coups. Furthermore, research by
@frantz2016 demonstrates that coup-related exits account for approximately
one-third of all exits in autocracies, surpassing any other transition type.
Additionally, the autocoup dataset, introduced in [Chapter 3](#sec-chapter3),
documents 110 autocoup attempts between 1945 and 2023, of which 87 were
successful.

```{r}
#| label: data
#| include: false
survival_cox_ph_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_ph_model.csv") |>
  mutate(survival_year = survival_days / 365)

survival_cox_td_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_td_model.csv")

avg <- survival_cox_ph_model |>
  summarise(
    avg = round(mean(survival_year), 1),
    .by = group
  )
```

Measuring the tenure of coup-installed and autocoup leaders poses challenges due
to the inherent irregularity and uncertainty of their positions. Nevertheless, a
comparative analysis reveals that leaders who extend power through autocoups tend
to have longer average post-autocoup tenures (approximately `r avg$avg[1]` years)
compared to coup-installed leaders (approximately `r avg$avg[2]` years),
suggesting a potential tenure gap of over five years.

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of autocoup and coup-installed leaders


# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = survival_cox_ph_model)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit,
  data = survival_cox_ph_model,
  risk.table = TRUE,
  risk.table.height = 0.3,
  pval = TRUE,
  surv.median.line = "hv",
  break.x.by = 5,
  legend.title = "Leader Type",
  legend.labs = c("Autocoup leaders", "coup-installed leaders"),
  xlab = "Survival Years",
  # title = "Survival Curves of overstay and coup-installed leaders",
  font.title = 12
)
```

A preliminary log-rank test in survival analysis, as illustrated in @fig-logrank,
demonstrates a statistically significant difference between the tenures of
autocoup and coup-installed leaders. The survival curve for autocoup leaders
consistently exceeds that of coup-installed leaders, indicating longer survival
times and a reduced risk of ouster for autocoup leaders.

This study posits that the method of accession significantly influences
leadership longevity. Coup-installed leaders likely confront greater challenges
to their rule, resulting in shorter average tenures compared to autocoup leaders.
The analysis, employing Cox proportional hazards and time-dependent Cox models,
supports this hypothesis, demonstrating that autocoup leaders generally
experience longer tenures than coup-installed leaders.

This research offers two primary contributions to the field. First, it highlights
an understudied factor in leadership survival analysis: the impact of the method
of accession to power. The findings suggest that leader survival is influenced
not only by ruling strategies but also by the initial method of acquiring power.
Second, by employing survival models, this study provides empirical evidence of
the significant difference in tenure duration between autocoup and coup-installed
leaders. This insight may explain the increasing prevalence of tenure extensions
through autocoups since 2000, as more incumbents observe and potentially emulate
successful precedents.

The remainder of this chapter is structured as follows: Section 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Section 3 explores the factors influencing the survival of
coup and autocoup leaders. Section 4 outlines the methodology and data used,
including the application of survival models to analyze the determinants of
leadership longevity. Section 5 presents the analysis findings and a detailed
discussion of the results. Finally, Section 6 concludes by synthesizing key
takeaways and exploring their broader implications for political stability and
democratic processes.

### Literature review

The longevity of political leaders, which varies widely across different regimes,
countries, and historical periods, has been a longstanding focus of research in
political science. This field can be divided into two interconnected areas:
regime survival and individual leader survival, which are distinct but related
concepts. Regime survival focuses on the endurance of political systems, such as
monarchies, political parties, or specific ideological structures, while leader
survival concerns the duration of individual leaders' time in office.

Political survival patterns vary widely across different systems. Parliamentary
democracies (e.g., Japan, United Kingdom) often experience prolonged periods of
party dominance coupled with frequent leadership changes. Similarly, communist
regimes (e.g., China) typically demonstrate enduring party rule with more
frequent leadership transitions. Presidential systems (e.g., United States) and
many military regimes tend to exhibit more frequent changes in both ruling party
or junta and leader.

The existing literature on leader survival is extensive and multifaceted. Some
studies explore specific mechanisms influencing leadership longevity within
particular regimes, such as democracies [@svolik2014] or autocracies
[@davenport2021]. Others aim to develop more generalizable theoretical frameworks
explaining leader survival across different political systems
[@buenodemesquita2003]. While a universal theory remains an aspirational goal,
the complexities of leadership survival across diverse regime types present
significant challenges.

Power transition mechanisms vary substantially across different types of regimes,
particularly between democracies and autocracies. Autocratic systems often
feature closed leadership selection processes, restricted to a narrow pool of
individuals. While some autocracies may hold elections, significant barriers to
entry for legitimate challengers typically persist. The opacity of selection
processes in autocracies makes it difficult to assess genuine levels of public
support compared to democracies. Conceptualizing selectorates or winning
coalitions, as proposed by @buenodemesquita2003, becomes problematic in many
autocratic contexts.

Given these complexities, focusing research on specific regimes or leader types
may be more fruitful. The study of irregular leaders, such as those who ascend to
power through coups or extend their tenures through autocoups, offers a
compelling avenue for research due to the inherent complexities and uncertainties
surrounding their leadership trajectories.

Two primary perspectives have emerged to explain the dynamics of leader survival.
The first emphasizes objective factors and resources, such as personal competence
[@yu2016], societal stability [@arriola2009], economic development [@palmer1999;
@williams2011], natural resource endowments [@smith2004; @quirozflores2012;
@wright2013], and external support [@licht2009; @wright2008; @thyne2017]. The
second focuses on subjective factors and strategies, including political
policies, responses to opposition, and tactics for consolidating power
[@gandhi2007; @morrison2009; @escribà-folch2013; @davenport2021].

Coups, a significant aspect of irregular leadership transitions, have received
considerable scholarly attention. Research has examined coup prevention
strategies [@powell2017; @sudduth2017; @debruin2020]. Studies have explored the
impact of coups on leadership and the subsequent actions of coup leaders
[@sudduth2017, @sudduth2018, @easton2018].

However, a significant gap remains in the literature regarding the comparison of
leadership survival between coup-installed and autocoup leaders. This study aims
to address this gap by investigating and comparing the duration of leadership
survival for these two leader types.

By focusing on the comparison between coup-installed and autocoup leaders, this
study seeks to contribute to a more nuanced understanding of political survival
in irregular leadership transitions. This approach may offer valuable insights
into the complex dynamics of leader longevity across different political
contexts.

### Survival dynamics of autocoup and coup-installed leaders

Studying leadership survival in political systems is challenging due to the
opacity and diverse mechanisms of power transitions. These challenges, however,
underscore the significance of this research, as it illuminates understudied
dynamics in political leadership. Although the survival of political leaders is
complex and varied, some patterns do emerge. Leaders of similar types often
exhibit similar characteristics, allowing for meaningful analysis.

#### Key definitions and scope

Before delving into the comparison, it is essential to clarify several key
terminologies:

-   **Coup and autocoup**: These terms are defined consistently with previous
    chapters to ensure clarity and consistency throughout the study.

-   **Tenure length threshold**: To ensure meaningful analysis, this study
    focuses on leaders with substantial periods in power, applying a six-month
    threshold to both autocoup and coup-installed leaders. This criterion filters
    out ephemeral leadership episodes, allowing for a more robust examination of
    survival dynamics.

-   **Autocoup leader**: An incumbent leader who successfully employs
    illegitimate or unconstitutional means to extend their tenure in power. This
    definition encompasses various methods of power consolidation that circumvent
    established democratic processes or constitutional limits.

-   **Coup-installed leader**: The individual who assumes power after a
    successful coup, regardless of their role in the coup itself. This broad
    definition allows for the inclusion of both coup instigators and those
    selected to lead post-coup, providing a comprehensive view of leadership
    dynamics following forceful regime change.

This study focuses on comparing the post-autocoup tenure of autocoup leaders with
the post-coup tenure of coup-installed leaders. This comparative approach is
motivated by the relevance and similarity of these leader types in terms of
illegitimacy, uncertainty, and instability. By examining these parallel yet
distinct paths to power, we can gain insights into the factors that influence
leadership longevity in irregular leadership transitions.

#### Challenges in power consolidation

Both autocoup and coup-installed leaders confront distinct challenges in
consolidating their power, primarily stemming from the varying intensity of
issues related to illegitimacy, uncertainty, and instability. This disparity
creates an uneven playing field in terms of power dynamics, placing
coup-installed leaders at a significant disadvantage. @tbl-leaders provides a
comparative overview of the main features of autocoup and coup-installed leaders,
highlighting these key differences.

\blandscape

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-installed leaders
#| echo: false
#| results: asis

read.csv(here("data/leader_comparison.csv")) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    column_labels.font.weight = "bold",
    table.width = pct(100),
    table.align = "center"
  ) |>
  cols_width(
    1 ~ px(150),
    2:3 ~ px(300)
  )

# |> data_color(rows = c(2, 4), colors = "#edede9")
```

\elandscape

##### Illegitimacy {.unnumbered}

While both types of leaders suffer from a legitimacy deficit, the nature and
perception of this deficit differ significantly:

-   **Coup-installed leaders**: Their illegitimacy is blatant and unambiguous,
    stemming from the overt and often violent seizure of power. This overt act
    undermines pre-existing norms and institutions, generating immediate domestic
    and international condemnation.

-   **Autocoup leaders**: In contrast, autocoup leaders employ a more subtle and
    deceptive strategy, manipulating legal processes and institutions to create a
    façade of democratic legitimacy. This veneer of legality, while often thin,
    can provide a degree of cover and buy time for consolidating power.

##### Uncertainty {.unnumbered}

The irregular paths to power for both types of leaders create uncertainty about
the longevity of their rule and the mechanisms of their eventual departure.
However, the levels and sources of this uncertainty differ significantly.

-   **Coup-installed leaders:** These leaders face a trifecta of uncertainties.
    First, the immediate aftermath of a coup often involves a struggle for power
    within the junta or ruling coalition, creating ambiguity about who will
    ultimately consolidate control. Second, the tenure of coup-installed leaders
    is inherently precarious, subject to internal rivalries, popular uprisings,
    or counter-coups. Third, the lack of established succession mechanisms
    further amplifies uncertainty, making it difficult to predict the transfer of
    power and potentially triggering future instability.

-   **Autocoup leaders:** While not immune to uncertainty, autocoup leaders
    generally present a clearer picture. The question of who will rule
    post-autocoup is largely settled, as the incumbent retains power.
    Furthermore, many autocoup leaders openly aspire to extend their rule
    indefinitely or incrementally, attempting to establish a sense of permanence.
    This perceived stability, whether real or manufactured, can contribute to a
    more predictable political environment, at least in the short term.

##### Instability {.unnumbered}

The awareness of shaky legitimacy and persistent uncertainty inevitably breeds
insecurity and a sense of crisis, forcing both autocoup and coup-installed
leaders to prioritize stabilization measures. However, the nature and intensity
of these challenges differ:

-   **Coup-installed leaders:** These leaders face the daunting task of rapidly
    reshaping power dynamics, often resorting to purges and crackdowns to
    eliminate potential adversaries and consolidate control. This process of
    dismantling existing structures and building new ones generates significant
    instability, potentially alienating former allies and triggering resistance
    from various segments of society. The need to appease powerful actors both
    domestically and internationally further limits their options, forcing them
    into compromises that can undermine their authority and long-term stability.

-   **Autocoup leaders:** In contrast, autocoup leaders often benefit from a
    degree of continuity in regime personnel and institutions. This relative
    stability allows them to implement changes gradually, minimizing disruptions
    and mitigating potential backlash. While they may still face opposition, they
    are less likely to confront immediate and existential threats to their rule,
    providing them with more time and leverage to consolidate power.

By understanding these contrasting challenges, we can better appreciate the
relative advantages and disadvantages faced by autocoup and coup-installed
leaders. This comparative perspective provides a nuanced framework for analyzing
the strategies these leaders employ to consolidate power and navigate the
perilous terrain of irregular leadership transitions.

#### Empirical evidence and hypothesis

Empirical evidence substantiates the disadvantage faced by coup-installed
leaders, revealing a complex interplay between historical precedent, power
consolidation challenges, and leadership longevity. This section presents key
data points and introduces the central hypothesis guiding this study.

Data analysis shows a significant correlation between the frequency of coup
attempts in a country and the likelihood of future coups. Notably, over a third
of coups have occurred in the top ten countries with the most attempts since 1950
(As shown in @tbl-coups). This pattern suggests a self-reinforcing cycle of
political instability, where each successful coup increases the probability of
subsequent attempts, creating an environment of persistent uncertainty for
coup-installed leaders.

The disparity in leadership longevity between autocoup and coup-installed leaders
is starkly illustrated by survival data. As depicted in @fig-logrank, the average
survival period following an autocoup is approximately five years longer than
that of coup-installed leaders. This substantial difference in tenure length
underscores the divergent challenges faced by these two types of leaders in
maintaining their grip on power.

The distinct challenges faced by autocoup leaders and coup-installed leaders in
consolidating power create a self-perpetuating cycle that significantly
influences their tenure length:

-   Coup-installed leaders: Face greater legitimacy challenges and internal
    instability; Struggle to attract and retain strong support; More vulnerable
    to internal and external challenges; Shorter average tenures reinforce
    perception of instability.

-   Autocoup leaders: Often benefit from a veneer of legitimacy and a stronger
    initial position; Better able to consolidate power and attract supporters;
    Face less immediate threat of overthrow; Longer average tenures contribute to
    perception of stability.

This cycle suggests that the initial method of power acquisition or extension has
far-reaching consequences for a leader's ability to maintain their position over
time.

Based on these observations and the theoretical framework outlined earlier, I
propose the following hypothesis:

***H4-1: Political leaders who successfully extend their tenure through autocoups
are more likely to survive longer extended tenure compared to coup-installed
leaders.***

This hypothesis encapsulates the expected outcome of the divergent challenges and
advantages faced by autocoup and coup-installed leaders. By testing this
hypothesis, I aim to quantify the impact of the method of power acquisition or
extension on leadership longevity, contributing to a more nuanced understanding
of political survival in contexts of irregular transitions.

### Research design

This section employs survival analysis to test the hypothesis that autocoup
leaders have longer survival times in office compared to coup-installed leaders.
This study uses Cox models to analyze the survival tenures of autocoup and
coup-installed leaders, controlling for various factors that may affect their
time in office.

#### Methodology: Survival analysis

Two Cox models will be employed to analyze the survival tenures of coup-installed
and autocoup leaders:

-   **Cox proportional hazards (PH) model**: This model uses only the variables
    present at the entry year, without considering changes over time.

-   **Time-dependent Cox model**: This model accounts for variations in
    time-dependent control variables such as economic performance and political
    stability.

The Cox model is preferred over the Kaplan-Meier model because it enables the
estimation of the impact of multiple factors. Although it does not directly
estimate the duration of tenure, it assesses the hazard rate associated with
being ousted from power. This approach captures different facets of the same
phenomenon: as a leader's cumulative hazard of being ousted increases, their
probability of survival in office decreases.

#### Data and variables

The dependent variables include survival time and end point status:

-   **Survival time:** Survival time refers to the duration of a leader's tenure,
    measured in days. For coup-installed leaders, the survival time begins on the
    day they assume power through a coup. For autocoup leaders, the survival time
    starts on the expiration date of their original legitimate term. For example,
    Russia's president Vladimir Putin assumed power in 2000 and, after serving
    two terms, stepped down in 2008. However, he remained in a powerful position
    as the prime minister and hand-picked Dmitry Medvedev to succeed him as
    president, while continuing to control the power behind the scenes. In this
    case, Putin's survival time begins in 2008, marking the start of his
    post-autocoup tenure. The survival time concludes on the day the leader
    finally exits office, applicable to both coup-installed and autocoup leaders.

-   **End point status:** This variable indicates the manner in which the
    leader's tenure concluded, categorized as follows:

    **0 = Censored:** This status is assigned to leaders who leave office through
    regular means other than being ousted. This includes leaders transferring
    power to their designated successors, leaving office as their terms expire,
    losing in general elections, voluntarily leaving office due to health issues,
    or dying of natural causes.

    **1 = Ousted:** This status is assigned to leaders who are forced to leave
    office. This includes leaders resigning under pressure, being ousted by coups
    or other forces, or being assassinated.

The key independent variable is the leader type, which categorizes leaders into
two distinct groups:

-   **Group A = Autocoup leader**: Leaders who extend their tenure through
    autocoups.
-   **Group B = Coup-installed leader**: Leaders who assume power through coups.

This variable is the primary independent variable of interest, serving as the
basis for comparing the survival time between these two types of leaders.

The data for both dependent and independent variables are sourced from the
autocoup dataset introduced in this study, Archigos, and PLAD.

Control variables include economic performance, political stability, population
size, and the leader's age, which are consistent with the autocoup analysis in
[Chapter @sec-chapter3].

### Results and discussion

#### Model results

Using the ***surviavl*** package in R [@survival], I present the regression
results for both the Cox Proportional Hazards model (Cox PH) and the
time-dependent Cox model in @tbl-cox.

```{r}
#| label: model

Mydata1 <- survival_cox_ph_model |>
  mutate(group = factor(group,
    labels = c("Autocoup leaders", "Coup-installed leaders")
  ))

model_cox1 <- Mydata1 |>
  coxph(
    Surv(survival_days, status) ~
      group + GDP_trend + GDP_pc + pop_log +
      polity5 + violence + age,
    data = _
  )

tbl1 <- model_cox1 |>
  tbl_regression(
    exponentiate = T,
    conf.int = F,
    add_estimate_to_reference_rows = T,
    label = list(
      group ~ "Leader Type",
      GDP_trend ~ "GDP Growth Trend",
      GDP_pc ~ "GDP per capita",
      pop_log ~ "Population: log",
      polity5 ~ "Polity 5",
      violence ~ "Political stability",
      age ~ "Age"
    )
  ) |>
  add_n(location = "level") |>
  add_nevent(location = "level") |>
  add_significance_stars(
    hide_p = T,
    hide_se = F,
    hide_ci = T,
    thresholds = c(0.01, 0.05, 0.1),
  ) |>
  bold_p(t = 0.05) |>
  bold_labels()

Mydata2 <- survival_cox_td_model |>
  mutate(
    group = factor(group,
      labels = c("Autocoup leaders", "Coup-installed leaders")
    ),
    survt = (tstop - tstart) * 365
  )

model_cox2 <- Mydata2 |>
  coxph(
    Surv(T1, T2, status) ~
      group + GDP_trend + GDP_pc + pop_log +
      polity5 + violence + age + cluster(id),
    data = _
  )
tbl2 <- model_cox2 |>
  tbl_regression(
    exponentiate = T,
    conf.int = F,
    add_estimate_to_reference_rows = T,
    label = list(
      group ~ "Leader Type",
      GDP_trend ~ "GDP Growth Trend",
      GDP_pc ~ "GDP per capita",
      pop_log ~ "Population: log",
      polity5 ~ "Polity 5",
      violence ~ "Political stability",
      age ~ "Age"
    )
  ) |>
  add_n(location = "level") |>
  add_nevent(location = "level") |>
  add_significance_stars(
    hide_p = T,
    hide_se = F,
    hide_ci = T,
    thresholds = c(0.01, 0.05, 0.1),
  ) |>
  bold_p(t = 0.05) |>
  bold_labels()
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox models for survival time of different types of leaders


tbl_merge(list(tbl1, tbl2),
  tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**")
)
```

Both models showed a statistically significant relationship between leadership
type and the hazard of removal from power. Since time-dependent Cox model use the
control variables which change over time, I interpret the main findings based on
time-dependent model.

Coup-installed leaders were found to have a hazard ratio of 2.23 in the
time-dependent model compared to autocoup leaders (reference group), assuming all
other variables in the model are held constant. This suggests that coup-installed
leaders face a significantly greater risk of removal from power compared to
autocoup leaders. At any given time during their tenure, coup-installed leaders
are 2.23 times more likely to be ousted from power compared to autocoup leaders,
all else being equal in the model.

The control variables perform differently in the two models. Economic level (GDP
per capita) exhibits statistically significant effects in both models. In the
time-dependent model, the hazard ratio of 0.95 indicates that for each unit
increase in GDP per capita (measured in units of \$10,000), the hazard (or risk)
of being ousted at any given time is reduced by 5%, assuming all other variables
in the model are held constant.

GDP growth trend demonstrates a more substantial effect in reducing the risk of
coups. Specifically, a 1 percentage point higher economic growth trend is
associated with an 80% reduction in the risk of being ousted, although this
effect is only statistically significant at the 10% level. This suggests a
possible trend where positive economic performance might mitigate the risk of
removal from power, but the evidence is not robust enough to confirm this
conclusively.

Political stability, as measured by the violence index, shows that a 1-point
increase in the index correlates with an 11% higher risk of being ousted.
However, this effect is also only statistically significant at the 10% level,
indicating a weaker but potentially important relationship between increased
violence and the risk of removal from office.

#### Discussion

```{r}
#| label: fig-coxSurv
#| fig-cap: Survival curves for Cox Model
#| fig-subcap:
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

testdata1 <- Mydata1 |>
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox1 |>
  survfit(newdata = testdata1) |>
  ggsurvplot(
    data = testdata1,
    xlab = "Days",
    surv.median.line = "hv",
    legend.title = "Leader Type",
    legend.labs = c("Autocoup leaders", "Coup-installed leaders")
  )

testdata2 <- Mydata2 |>
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox2 |>
  survfit(newdata = testdata2) |>
  ggsurvplot(
    data = testdata2,
    xlab = "Days",
    surv.median.line = "hv",
    legend.title = "Leader Type",
    legend.labs = c("Autocoup leaders", "Coup-installed leaders")
  )
```

The survival curves in @fig-coxSurv displays the survival rates for both types of
leaders, highlighting the differences in their survival curves. Both the Cox PH
model and the time-dependent Cox model yield similar results. Notably, the
survival curve for coup-installed leaders has a significantly lower trajectory
than that of autocoup leaders. The steeper drop at the early stage for
coup-installed leaders indicates they are more likely to be ousted shortly after
assuming power. Additionally, the survival curve for coup-installed leaders
crosses the median survival line much earlier (about 3,000 days) than that of
autocoup leaders (about 8,500 days). This disparity suggests that autocoup
leaders tend to remain in power for longer durations than their coup-installed
counterparts.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard ratios and 95% CIs for Leader Ousting
#| fig-subcap:
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c(
    "Coup-installed leaders",
    "GDP Growth Trend",
    "GDP per capita",
    "Population: log",
    "Polity 5",
    "Political stability",
    "Age"
  ))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "skyblue"
  ) +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw()

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |>
  mutate(term = c(
    "Coup-installed leaders",
    "GDP Growth Trend",
    "GDP per capita",
    "Population: log",
    "Polity 5",
    "Political stability",
    "Age"
  ))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "skyblue"
  ) +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw()
```

@fig-coxHR displays the hazard ratios and corresponding 95% confidence intervals
for the variables incorporated in the Cox model. Both the Cox Proportional
Hazards (PH) model and the time-dependent model produce similar plots,
reinforcing the robustness of the findings. Key points to note include:

-   The closer the hazard ratio (represented by the dots) is to 1, the less
    impact the variable has on the risk of being ousted. A hazard ratio of 1
    indicates no effect.

-   The whiskers extending from the dots represent the 95% confidence intervals.
    If these whiskers cross the vertical blue line at 1, it indicates that the
    variable is not statistically significant at the 5% level.

-   The hazard ratio for coup-installed leaders is significantly greater than 1
    and statistically significant at the 5% level. This indicates that
    coup-installed leaders face a substantially higher risk of being ousted
    compared to autocoup leaders.

-   Most other variables have hazard ratios close to 1, suggesting that a
    one-unit increase in these variables does not significantly affect the risk
    of being ousted.

-   Although the hazard ratio for GDP growth trend is considerably less than 1 in
    the time-dependent model, indicating a potential protective effect, it is not
    statistically significant at the 5% level. However, it is statistically
    significant at the 10% level, suggesting that better economic performance may
    help to consolidate the rule of the incumbents to some extent, albeit the
    evidence is not as strong.

#### Assessing the proportional hazards assumption

Assessing the proportional hazards assumption is crucial for the validity of the
Cox model results. To evaluate this, we used the chi-square test based on
Schoenfeld residuals to determine whether the covariate effects remain constant
(proportional) over time. Although the Cox PH model violates the proportional
hazards assumption, our primary analysis relies on the time-dependent Cox model,
which does not show strong evidence of violating the proportional hazards
assumption for any covariate. The global p-value of 0.416 is much greater than
the 5% significance level, indicating that the proportional hazards assumption is
reasonably met for the time-dependent Cox model.

```{r}
#| label: robust_test
#| include: false
test <- cox.zph(model_cox2)
```

### Summary

This chapter explores the survival durations of political leaders who come to
power through unconventional means, specifically focusing on coups and autocoups.
Based on the hypothesis that the mode of accession affects leader tenure, I use
survival analysis techniques, including the Cox proportional hazards model and a
time-dependent Cox model, to investigate this phenomenon. The findings suggest
that autocoup leaders tend to have longer tenures than coup-installed leaders.

Empirical analysis reveals a significant disparity in tenure length: leaders who
assume power via autocoup remain in office for an average of 11 years, compared
to just 5.6 years for those installed by coups. Moreover, the time-dependent Cox
model indicates that coup-installed leaders are 2.23 times more likely to be
ousted from power at any given time compared to their autocoup counterparts, all
other factors being equal. These findings underscore the importance of
understanding the autocoup as a mechanism through which leaders extend their rule
by manipulating legal frameworks and weakening institutional constraints.

The implications of these findings are profound. The relative ease and potential
rewards of an autocoup could incentivize more leaders to resort to this method of
power retention, particularly in fragile democracies or transitioning regimes.
Consequently, democratic backsliding may become more prevalent, as autocoups
erode democratic institutions and undermine constitutional norms.

This study contributes significantly to the literature on political leadership
survival by showing that the mode of accession has a substantial impact on leader
tenure, an aspect that has received limited attention in previous research.
Methodologically, this work advances the field by applying robust survival
analysis techniques, including both Cox models, to provide a nuanced
understanding of the dynamics that influence leadership stability.

However, the study is not without limitations. The analysis relies on an autocoup
dataset that was collected and coded by the author, a relatively novel concept in
the academic sphere. As the understanding and recognition of "autocoup" as a term
continue to evolve, future research should refine and expand the dataset.
Incorporating additional cases and cross-referencing with other forms of
irregular leadership transitions would contribute to a more comprehensive view of
political survival under such conditions.

In conclusion, this chapter highlights the need for more refined approaches to
studying political tenure and irregular power retention. By offering valuable
insights into the dynamics of political stability and the risks associated with
non-democratic leadership transitions, this research emphasizes the importance of
continued investigation into the complex relationships between power, legitimacy,
and survival in political leadership.

## Coups, Autocoups, and Democracy

### Abstract {.unnumbered}

This chapter explores the impact of autocoups on political institutions, drawing
comparisons with traditional coups through an analysis of changes in Polity
scores. I contend that, first, incumbent leaders frequently consolidate power by
undermining established institutions in anticipation of an autocoup, resulting in
a decline in Polity scores even prior to the event. Second, unlike coups, which
produce mixed outcomes regarding democratization, autocoups almost invariably
lead to democratic backsliding or authoritarian entrenchment, as they are
specifically designed to dismantle institutional checks, enabling leaders to
maintain power for significantly longer periods than those installed via coups.

Employing a country-fixed effects model and utilizing datasets on both autocoups
and coups, this study reveals that Polity scores decline both preceding and
following an autocoup, while coups tend to trigger an immediate decrease that may
allow for some degree of democratic recovery over time. These findings underscore
the divergent political trajectories associated with coups and autocoups. This
research not only addresses a critical gap in the empirical analysis of autocoups
but also significantly raises awareness within academic and policy-making circles
regarding their potential adverse effects, including democratic backsliding and
further authoritarian deterioration.

### Introduction

In the preceding chapters, I clarified the definition of an autocoup, introduced
a novel dataset on autocoups, conducted empirical analyses on the determinants of
autocoup attempts, and compared the post-event survival times of leaders
established through coups versus autocoup leaders. A natural follow-up question
emerges: What are the broader impacts of autocoups? Specifically, from a
political science perspective, how do autocoups affect the process of
democratization?

As previously noted, due to the absence of a widely recognized dataset on
autocoups, most existing discussions on their impact have relied on case studies
[@baturo; @baturo2022]. To move beyond case-specific analyses and adopt a more
systematic and comparative approach, this chapter aims to pioneer empirical
research on the democratic consequences of autocoups. The first objective of this
chapter, therefore, is to examine whether autocoups reinforce authoritarianism,
promote democratization, or have no significant effect on political regimes.

Given the conceptual and empirical connections between coups and autocoups,
another key objective is to compare their respective effects on democratization.
While both events disrupt existing political orders, their immediate and
long-term consequences may vary significantly. Understanding these distinctions
is essential for evaluating their broader implications.

To address these questions, this study leverages data from a widely recognized
coup dataset alongside a newly compiled dataset on autocoups. Employing a
fixed-effects model, it evaluates their effects on democracy, as measured by the
Polity Index. The findings reveal that both coups and autocoups lead to an
immediate decline in democratic levels. However, coups have a more pronounced
negative short-term impact. Notably, three years after these events, democracies
affected by coups tend to show significant recovery, whereas those experiencing
autocoups exhibit no meaningful improvement.

This study offers two significant contributions to the field of political
science. First, it provides the inaugural empirical analysis of the impact of
autocoups on democratization, effectively addressing a critical gap in the
existing literature. Second, by directly comparing the effects of coups and
autocoups, this research highlights autocoups as a distinct phenomenon that
demands increased scholarly inquiry and policy focus.

The remainder of this chapter is structured as follows: Section 2 analyzes the
impact of autocoups on democratization, by comparing the effects with traditional
coups. Section 3 outlines the research design, methodology, and variables used in
this study. Section 4 presents and interprets the empirical findings, followed by
a discussion of their implications. Finally, Section 5 concludes by summarizing
the key results and exploring their significance for understanding and mitigating
the occurrence of autocoups.

### Impact of autocoups on political change

According to the definition in @sec-definition, an autocoup refers to an
incumbent leader extending their tenure in power beyond the originally mandated
limits, whether through legal or illegal means. While the leader may assume a
different title or position, the individual in power remains unchanged.
Therefore, unlike traditional coups, an autocoup does not result in real
leadership turnover, elite restructuring, or regime change. In other words, the
fundamental ruling structure remains intact.

This distinction has important implications. Since regime change rarely occurs
following an autocoup, its impact on politics cannot be assessed using
conventional methods. Typically, studies on coups and democratization measure
their effects by estimating the probability of regime transition—either from
autocracy to democracy or vice versa—as seen in previous research on coup
outcomes [@thyne2014; @derpanopoulos2016; @miller2016]. However, this approach is
not suitable for autocoups, as they do not directly trigger regime transfers.

Although autocoups rarely lead to formal regime change, this does not mean they
have no impact on political dynamics. In fact, they inevitably shape political
trajectories in various ways and, on occasion, can even lead to significant
transformations. Therefore, a more appropriate method for assessing the political
impact of autocoups is to analyze democratic indices, such as those measured by
Polity5 [@p1]. The Polity Score, ranging from -10 (full autocracy) to +10 (full
democracy), captures gradual shifts in political regime characteristics rather
than abrupt transitions. Thus, even if an autocoup does not result in a formal
regime change, subtle shifts in political openness and institutional constraints
can still be assessed by examining variations in Polity scores before and after
the event. This approach has also been employed in previous studies [@dahl2023].

Although autocoups may not lead to significant regime change, their impact on
democratization should not be overlooked. However, their influence differs from
that of traditional coups in at least two key ways.

#### **The pre-emptive effects of autocoups on political dynamics**

First and foremost, unlike coups, which are marked by clear, decisive events—such
as the removal of a leader—autocoups typically unfold gradually, through
incremental steps rather than a singular, dramatic event. Incumbent leaders
seeking to extend their tenure often lay the groundwork well in advance before
executing their final move to remain in power. To reduce resistance and
opposition, they engage in extensive preparation, which may include purging
officials, suppressing political opposition, cracking down on dissent and
protests, and restricting press freedom. Without such measures, an autocoup might
face strong internal resistance, and in the worst case, provoke a backlash that
not only derails the leader’s attempt to extend their rule but also results in
their immediate removal from office.

However, once an incumbent successfully secures an extension of their rule,
continued repression is not always necessary. On the contrary, some leaders relax
political pressure to ease internal dissent and mitigate opposition from external
actors. This adaptive approach helps maintain stability after the autocoup is
complete.

As a result, the primary impact of autocoups on political change is often
reflected in shifts in political scores before the final stage of the autocoup is
enacted. Once the process is completed, further shifts may be minimal. In
contrast, coup plotters, unlike incumbents, lack the ability to influence
political institutions beforehand, meaning their impact on politics is often felt
in the aftermath rather than before the event.

This distinction is evident in empirical cases of autocoups.

One of the most frequently cited examples of an autocoup is Peru’s 1992 case, in
which President Alberto Fujimori dissolved Congress, temporarily suspended the
1979 Constitution, and ruled by decree until November of that year, when a
*Democratic Constituent Congress* was elected to draft a new constitution
[@cameron1998]. However, these moves did not immediately grant Fujimori a longer
tenure in office.

Under the 1979 Peruvian Constitution, immediate presidential re-election was
prohibited. To bypass this restriction, Fujimori initiated a constitutional
overhaul, leading to the adoption of a new constitution in 1993, which permitted
re-election. Consequently, he secured a second term in 1995 [@baturo2019].

An analysis of Peru’s Polity5 scores during this process reflects these political
changes. Upon taking office in 1990, Peru’s Polity score was 8, remaining
unchanged in 1991. However, a dramatic shift occurred in 1992, when Fujimori
dissolved Congress, causing the Polity score to plummet from 8 to -4.
Interestingly, when he formally extended his rule by amending the constitution in
1993, the score rebounded slightly to -1. This -1 score remained unchanged
throughout Fujimori’s tenure until 2000, indicating a lack of further
institutional transformation after the constitutional amendment.

A similar pattern emerges in Belarus under Alexander Lukashenko. Upon assuming
office as president in 1994, Belarus had a Polity5 score of 8. However, in 1995,
when Lukashenko moved to hold a referendum—defying opposition in the Supreme
Council and threatening to suspend its activities—the score dropped sharply to 0.
Following the 1996 referendum, which extended his term by two additional years,
Lukashenko officially overstayed his tenure. Consequently, the Polity5 score
further declined to -7 in 1996, where it has remained ever since, despite two
additional term extensions [@ash2014; @baturo].

These cases illustrate a broader pattern: the impact of autocoups on political
change is often reflected before the final stage of the autocoup is enacted,
whereas the impact of coups tends to materialize afterwards, as have been fully
discussed by previous studies.

Based on this analysis, I propose the first hypothesis:

> *H1*: Autocoups primarily shape political change in advance, whereas coups
> typically drive political change only after they are executed.

#### **The singular nature of autocoups**

Secondly, in contrast to the ambiguous nature of coups [@dahl2023], the impact of
autocoups on political change rarely contributes to democratization.

The influence of coups on democratization has been extensively examined in
existing literature. Some scholars argue that coups—and even the mere threat of
them—can act as catalysts for democratization. One argument suggests that coups
deliver a political "shock" that may create opportunities for liberalization that
would not have otherwise materialized [@thyne2014]. In a critical examination,
@derpanopoulos2016 questioned the role of coups in promoting democracy, engaging
in multiple rounds of debate with @miller2016. More recently, @dahl2023 further
explored this ongoing discussion, arguing that both democratic and autocratic
transitions are likely to follow a coup, with popular mobilization playing a
decisive role in shaping post-coup trajectories.

A frequently cited example of a “pro-democracy coup” occurred in February 2010,
when Nigerien troops ousted President Mamadou Tandja after he extended his rule
autocratically. The Supreme Council for the Restoration of Democracy (CSRD) took
control, pledging democratic reforms. Their actions were widely celebrated, with
both citizens and political opposition viewing the coup as an opportunity to
restore democracy. The CSRD fulfilled its commitment by overseeing free and fair
elections in 2011, resulting in Mahamadou Issoufou assuming the presidency
[@miller2016].

While the debate on the democratic consequences of coups remains ongoing, it is
evident that their impact is not uniform. In contrast, autocoups almost never
lead to democratic transitions, nor do they even marginally enhance political
freedoms. This stems from the intrinsic nature of autocoups, which disrupt the
established process of political leadership transition, particularly term limits.

Term limits are constitutional provisions that restrict the maximum duration a
leader can remain in office. They play a crucial role in both democracies and
autocracies by preventing the excessive concentration of power and ensuring
political stability. In democracies, term limits promote accountability,
leadership renewal, and reduce the risks of corruption and authoritarian
entrenchment. In autocracies, when enforced, they can curb indefinite rule,
mitigate succession crises, and provide rare opportunities for political
transitions. However, without term limits, leaders can entrench themselves in
power, undermining institutions and hindering political progress.

As shown in @tbl-autocoup_method in Chapter 3, autocoups are executed through
either façade legal mechanisms or blatantly illegal methods. This includes
amending or disregarding term limits, delaying or cancelling elections, rigging
electoral outcomes, or outright refusing to accept election results. While many
autocoups maintain a veneer of legality, their defining characteristic is the
violation of term limits, which are intended as safeguards against prolonged and
unchecked rule.

As previously discussed, before an incumbent leader formally overstays their
term, they often inflict significant damage on political institutions.
Furthermore, to secure their prolonged rule, they are unlikely to fully restore
political freedoms even if they temporarily ease political repression once they
have consolidated power.

Case studies from Peru and Belarus illustrate how autocoups lead to a decline in
Polity scores, reflecting democratic backsliding. However, most autocoups occur
in already autocratic regimes where Polity scores are low. This aligns with
trends observed in coups, which also primarily occur in autocracies.

For instance, in China’s 2018 constitutional amendment, Xi Jinping abolished
presidential term limits, effectively allowing himself to remain in power
indefinitely[^2]. However, China’s Polity score remained unchanged at -7 before
and after the amendment. This pattern is common in autocratic regimes with Polity
scores below -6, as they already lack significant democratic features, leaving
little room for further decline.

While most autocoups occur in low-Polity-score countries, some have taken place
in relatively democratic settings, where the Polity score remains stable despite
term limit extensions. This is particularly evident in Latin America, where
presidents have amended "no immediate re-election" rules to allow for a second
consecutive term, but voluntarily stepped down afterwards. Examples include:
Argentina (1993), Polity score remained at 7; Brazil (1997), Polity score
remained at 8, Colombia (2004), Polity score remained at 7. In these cases,
leaders extended their tenure within a structured political framework without
further dismantling democratic institutions [@baturo2019].

Across all cases—whether in Peru, Belarus, China, Argentina, Brazil, or
Colombia—there is no instance where a Polity score increased following an
autocoup. In the autocoup dataset which I introduced in Chapter 3, only four
cases—Guinea-Bissau (1988), Burkina Faso (1997), Congo-Brazzaville (2001), and
Lebanon (2004)—saw minor increases in Polity scores, but the changes were
insignificant.

Thus, unlike some coup leaders who justify their actions by claiming to restore
democracy (as seen in Niger’s 2010 case), leaders who execute autocoups lack any
democratic justification. If their true intention were to advance democracy, they
would transfer power peacefully rather than violate term limits.

Based on this analysis, I propose the second hypothesis:

> *H2*: Autocoups are more likely to entrench autocracy than coups.

### Methodology and variables

#### Methodology

As discussed earlier, autocoups are less likely to result in full regime
transitions—either from democracy to autocracy or vice versa. Therefore, it is
inappropriate to assess their effects solely based on regime transitions or
changes that cross a critical threshold. Instead, this study examines political
changes through fluctuations in Polity5 scores.

Unlike coup analysis, which primarily focuses on post-event effects, this study
examines both pre-event and post-event impacts of autocoups. Specifically,
pre-event effects are measured by the change in Polity5 scores three years before
the autocoup compared to the year of the autocoup, expressed as:

$Polity_t - Polity_{t-3}$

Post-event effects are measured by the change in Polity5 scores three years after
the autocoup compared to the year of the autocoup, expressed as:

$Polity_{t+3} - Polity_t$

The three-year window is chosen for two key reasons. Pre-event political changes
typically occur incrementally, as incumbents consolidate power gradually over
several years rather than through a single, abrupt action. Post-event analysis
focuses on medium-term effects rather than short-term shocks, as autocoups rarely
trigger immediate regime transitions and instead reinforce existing political
structures. Short-term fluctuations may be too minor to capture meaningful
institutional change empirically.

To estimate how political institutions change before and after autocoups, I
employ a linear model with country-case fixed effects. For pre-event effects, I
analyze attempted autocoups, since before the event, no one knows whether the
autocoup will succeed or fail.

For post-event effects, I analyze only successful autocoups for three reasons:
The majority of autocoups succeed (87 out of 110 cases); failed autocoups tend to
produce immediate political shocks, rather than medium- or long-term effects;
failed autocoups are often followed by major disruptive events, such as coups,
insurrections, or mass protests, making it difficult to isolate their impact. For
example, in Niger's case, a failed autocoup in 2009 was followed by a coup in
2010, creating overlapping political effects. In contrast, successful autocoups
provide a clearer analytical framework, as their effects are less entangled with
other disruptions, allowing for a more systematic assessment of institutional
change.

#### Variables

This study uses a global sample of all country-year data from 1950 to 2020,
applying a linear model to examine the effects of autocoups on political change.
The dependent variable is the change in Polity scores, while the main independent
variable is autocoups. The dataset includes approximately 7,500 observations.

The dependent variable measures political change using three-year differences in
Polity scores. Model 1 (Pre-event effects): $Polity_t − Polity_{t − 3}$. Model 2
(Post-event effects): $Polity_{t+3} - Polity_t$.

The Polity score ranges from -10 (full autocracy) to 10 (full democracy). Some
values in the dataset, such as -66, -77, and -88, represent transitional regimes
or special periods. To prevent excessive data loss, I replace these values with
the closest valid Polity scores. This approach ensures that the model captures
all changes in Polity scores, rather than focusing solely on transitions that
cross a democratic threshold.

The main independent variable is autocoups, as introduced in Chapter 3. The
dataset includes110 attempted autocoups (used for pre-event analysis) and 87
successful autocoups (used for post-event analysis). For pre-event analysis, the
autocoup variable is binary, where: 1 indicates the presence of an attempted
autocoup and 0 indicates no autocoup event.

For post-event analysis, I apply decay functions to account for both immediate
and delayed effects, following the methodology of @dahl2023. To assess the
persistence of autocoup effects, I consider a half-life specification of five
years, analysing the impact from the autocoup year ($y_t$) to four years after
($y_{t+4}$).

I also include traditional coups as a secondary independent variable for two
reasons. Comparative significance: It is essential to compare the effects of
autocoups and coups. Overlapping events: In many cases, coups and autocoups are
interconnected, as autocoups can trigger coups. Distinguishing their effects is
necessary for a clear empirical analysis.

As in previous chapters, I use the coup dataset from @powell2011. To maintain
consistency, I apply the same methodological approach to coups as to autocoups:
Binary coding for pre-event effects and decay function coding for post-event
effects.

Control variables include economic performance, political stability, and
population size, all of which have been analysed in previous chapters.
Additionally, I incorporate two dummy variables. The first,
**"**non_democracy**,"** accounts for regime type, recognizing that
non-democratic regimes with Polity scores below -6 have limited room for further
decline, while democracies with scores above 6 are less likely to experience
significant increases. The second, **"**cold_war**,"** follows the approach used
in prior studies on the impact of coups on democratization [@thyne2014;
@derpanopoulos2016; @dahl2023] and accounts for the Cold War period. This
variable captures the observable trend of declining Polity scores from the 1960s
to 1990, followed by an upward shift after 1990.

### Results and discussion

#### Pre-event effects

Initially, I analyse the trajectory of Polity scores leading up to these events.
As presented in @tbl-demomodel, columns 1 and 2 display the empirical results for
pre-event effects, examining changes over two years (model 1) and three years
(model 2) prior to the event. Consistent with the first hypothesis, Polity scores
exhibit a significant decline in the 2–3 years preceding an autocoup. Notably,
this downward trend is both more pronounced and statistically significant for
autocoups compared to traditional coups.

```{r}
#| label: coup_autocoup_data
coup_autocoup <- read.csv("https://raw.githubusercontent.com/reddylee/Datasets/refs/heads/main/coup_autocoup.csv")
```

```{r}
#| label: tbl-demomodel
#| tbl-cap: "The Impact of Autocoups on Democratization(1950–2018): OLS with country-fixed effects"
#| results: asis

demo_model1 <- plm(
  Attemp ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Attemp = diff2, Autocoup = autocoup_attemp_3, Coup = coup_attemp_3)
)

demo_model2 <- plm(
  Attemp ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Attemp = diff3, Autocoup = autocoup_attemp_3, Coup = coup_attemp_3)
)

demo_model3 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff3, Autocoup = autocoup_effect, Coup = coup_effect)
)

demo_model4 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff3, Autocoup = autocoup_seffect, Coup = coup_seffect)
)

stargazer(demo_model1, demo_model2, demo_model3, demo_model4,
  # type = "text",
  float = F,
  header = F,
  dep.var.caption = "Dependent variable: Differences of Polity scores",
  dep.var.labels = c("Pre-event effects", "Post-event effects"),
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Non-Democracy", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "30pt",
  # font.size = "footnotesize",
  df = F
)
```

Column 1 examines the pre-event changes, specifically the difference between
Polity scores at time $t-1$ and $t-3$ ($Polity_{t-1} - Polity_{t-3}$). The
results reveal a statistically significant decrease in Polity scores prior to
autocoups. On average, Polity scores decline by 0.45 in the two years before an
autocoup, holding other factors constant. In contrast, no significant changes are
observed before traditional coups, indicating that pre-coup periods do not
measurably affect democratization levels.

Column 2 assesses the cumulative impact from three years prior to the event year,
calculated as $Polity_{t} - Polity_{t-3}$. As previously discussed, the year of
an autocoup or coup witnesses a substantial shock to political institutions.
Consequently, both types of events result in a decline in Polity scores relative
to three years earlier. However, the negative effect of autocoups is more severe.
Polity scores decline by an average of 1.53 in the three years leading up to an
autocoup, compared to 1.27 for coups, all else being equal. This disparity
reinforces our hypothesis that autocoups have a more detrimental pre-event effect
on democratic institutions as incumbents overextend their legitimate tenure.

#### Post-event effects

Columns 3 and 4 present the empirical results for post-event effects, analysing
changes in Polity scores following attempted (column 3) and successful (column 4)
autocoups.

Column 3 examines the impact of attempted autocoups. The results indicate that
attempted autocoups do not have a statistically significant effect on Polity
scores. This contrasts with coup attempts, which lead to an average increase of
0.48 in Polity scores over three years, all else being equal.

Column 4 evaluates the impact of successful autocoups. In contrast to attempted
autocoups, successful autocoups demonstrate a significant negative effect on
Polity scores, with an average decline of 0.33 over three years, holding other
factors constant. Conversely, successful coups continue to exhibit positive
effects on democratization, resulting in an average increase of 0.72.

These findings yield several key insights. First, while both successful and
attempted coups influence Polity scores, only successful autocoups produce
statistically significant effects—and exclusively in a negative direction.
Second, the impact of coups on democratization is more nuanced. Although
post-coup Polity scores may improve, the pre-coup period often involves
significant democratic backsliding. My models indicate that, on average, the
democratic gains following a coup fail to offset the losses incurred before it.

#### Effects of control variables

To ensure robust results, I incorporated several control variables into the
models. While economic trends and political violence do not exhibit statistically
significant effects on Polity scores, other factors warrant further discussion.

The impact of the Cold War is relatively straightforward. As outlined in the
research design section, global democracy experienced a general decline during
the Cold War period. Consistent with this trend, all four models indicate that
the Cold War era is associated with an average decrease of 0.23 in Polity scores.

The effects of GDP per capita, population size, and regime type, however, present
a more complex pattern. Counter-intuitively, higher GDP per capita correlates
with lower Polity scores, whereas non-democratic regimes and larger populations
correspond with positive changes in Polity scores. This pattern may be explained
by considering baseline differences between democracies and non-democracies. In
established democracies, Polity scores are already high, leaving limited room for
further increases. These countries also tend to have higher GDP per capita and
lower birth rates, further reinforcing stability in their democratic scores. In
contrast, non-democracies start from lower Polity scores, providing greater
potential for upward movement. Given that authoritarian regimes often exhibit
weaker economic performance and higher population growth rates, their Polity
scores may increase as they undergo political transitions or reforms.

Taken together, these results support both hypotheses. First, the primary effects
of autocoups on Polity scores manifest before the events occur. Second, whereas
coups have ambiguous consequences for democratization, the effects of autocoups
are unidirectional—consistently negative.

#### Robustness tests

To assess the sensitivity of my key findings to model specifications, I conduct a
series of robustness tests. The results indicate that the findings remain
consistent across these variations.

First, I compare the effects of autocoups over a period ranging from one to five
years after the event. The results show that the effects of coups are positive
and statistically significant throughout all five years, with a general trend of
increasing magnitude as time progresses. In contrast, the effects of autocoups
are negative across all five years but reach statistical significance only three
years after the event. This finding aligns with the earlier hypothesis: autocoups
never contribute to an increase in Polity scores.

```{r}
#| label: tbl-demomodel1
#| tbl-cap: "The Impact of Autocoups on Democratization: one to five years"
#| results: asis

demo_model1 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff1, Autocoup = autocoup_seffect, Coup = coup_seffect)
)

demo_model2 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff2, Autocoup = autocoup_seffect, Coup = coup_seffect)
)

demo_model3 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff3, Autocoup = autocoup_seffect, Coup = coup_seffect)
)

demo_model4 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff4, Autocoup = autocoup_seffect, Coup = coup_seffect)
)

demo_model5 <- plm(
  Succeed ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + factor(non_democracy) + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |> rename(Succeed = diff5, Autocoup = autocoup_seffect, Coup = coup_seffect)
)

stargazer(demo_model1, demo_model2, demo_model3, demo_model4, demo_model5,
  # type = "text",
  float = F,
  header = F,
  dep.var.caption = "Dependent variable: Differences of Polity scores",
  dep.var.labels = "Years after the event",
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Non-Democracy", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "10pt",
  # font.size = "footnotesize",
  df = F
)
```

```{r}
#| label: tbl-demomodel2
#| tbl-cap: "The Impact of Autocoups on Democratization: Dummy autocoups"
#| results: asis


demo_model1 <- plm(
  Attempted ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Attempted = diff2, Autocoup = autocoup_attemp, Coup = coup_attemp) |>
    filter(regime != "other")
)

demo_model2 <- plm(
  Attempted ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Attempted = diff3, Autocoup = autocoup_attemp, Coup = coup_attemp) |>
    filter(regime != "other")
)

demo_model3 <- plm(
  Succeeded ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Succeeded = diff2, Autocoup = autocoup_succeed, Coup = coup_succeed) |>
    filter(regime != "other")
)
demo_model4 <- plm(
  Succeeded ~ Autocoup + Coup + GDP_pc + GDP_trend + pop_log + violence + regime + factor(cold_war),
  index = c("ccode", "year"),
  model = "within", # fixed effect
  effect = "individual",
  data = coup_autocoup |>
    rename(Succeeded = diff3, Autocoup = autocoup_succeed, Coup = coup_succeed) |>
    filter(regime != "other")
)


stargazer(demo_model1, demo_model2, demo_model3, demo_model4,
  # type = "text",
  float = F,
  header = F,
  dep.var.caption = "Dependent variable: Differences of Polity scores",
  covariate.labels = c(
    "Autocoup", "Coup", "GDP per Capita",
    "Economic Trend", "Log Population", "Political Violence",
    "Regime: Dominant-party", "\\hspace{1.5cm} Military", "\\hspace{1.5cm} Monarchy", "\\hspace{1.5cm} Personal", "Cold War"
  ),
  summary.logical = F,
  column.sep.width = "20pt",
  # font.size = "footnotesize",
  df = F
)
```

Second, I refine the treatment of autocoups by replacing the decay effects of
autocoups with a dummy variable that distinguishes between attempted and
successful autocoups. Additionally, I disaggregate the ‘Non-democracy’ category
into specific regime types—democracy, dominant-party, military, monarchy, and
personal—consistent with the analysis of coup determinants, setting democracy as
the reference category.

Columns 1 and 2 in @tbl-demomodel2 examine the effects of attempted autocoups,
measured two and three years after the event, respectively, while Columns 3 and 4
focus on successful autocoups. As in previous models, these adjustments do not
alter the core findings across all four models. However, while they lead to
differences in regression coefficients, the results consistently show that Polity
scores decline following autocoups in all models, in contrast to the increase
observed after coups.

Comparing Columns 3 and 4 in @tbl-demomodel2 and @tbl-demomodel, it becomes
evident that replacing the decay factor for autocoups and coups with a dummy
variable significantly amplifies the estimated effects—nearly doubling them. This
result aligns with expectations, as the decay factor distributes the influence of
events over subsequent years, whereas the event dummy captures their impact
exclusively in the year of occurrence.

Furthermore, @tbl-demomodel2 reveals distinct regime-specific variations.
Military regimes exhibit the largest increase in Polity scores, followed by
monarchies and personalist regimes, with dominant-party regimes showing the
smallest but still statistically significant positive effect compared to
democracies. This finding corroborates the results in Chapter 2, which indicate
that military and personalist regimes are more prone to coups. The analysis in
this chapter proves that Polity scores increase significantly following coups.

### Summary

This chapter examines the impact of autocoups on political institutions,
particularly in contrast to coups, by analysing their effects on changes in
Polity scores. It tests two central hypotheses: first, that unlike coups, the
effects of autocoups on Polity scores manifest primarily before the event,
indicating that leaders consolidate power in anticipation of their autocoups; and
second, that while coups produce ambiguous effects—sometimes leading to
democratization and other times reinforcing authoritarianism, as suggested by
previous literature—autocoups consistently result in democratic backsliding or
authoritarian entrenchment.

To evaluate these hypotheses, the chapter employs multiple robustness checks,
including varying time horizons, alternative model specifications, and different
variable treatments. A key finding is that Polity scores begin to decline in the
years leading up to an autocoup, underscoring a pre-emptive process of
authorization. In contrast, coups initially cause a sharp drop in Polity scores
due to the shock of leadership change, but over time, scores often rise,
suggesting that in some cases, coups can contribute to an improvement in the
level of democracy. These findings highlight the fundamentally different
political trajectories triggered by coups and autocoups.

The implications of these results are significant for both academic research and
political practice. While coups have long been a focal point of democratization
studies, this chapter argues that autocoups demand equal, if not greater,
attention due to their systematic role in reversing democratic progress. Unlike
coups, which can sometimes serve as catalysts for political reform, autocoups
almost always reinforce authoritarianism, weakening institutions and eroding
democratic governance. This underscores the urgent need for scholars and
policy-makers to closely monitor the conditions that enable autocoups and their
broader consequences for democratic stability.

Methodologically, this chapter contributes to the study of political transitions
by demonstrating the importance of pre-event trends in analysing regime change.
It also highlights the necessity of distinguishing between different types of
irregular power transitions when assessing their long-term effects. While the
findings reinforce the study’s core arguments, they also raise questions for
future research—particularly regarding leader survival. As shown in Chapter 4,
autocoup leaders tend to survive in power for nearly 11 years on average, while
coup-installed leaders last only about 5 years. This suggests that coups and
autocoups not only differ in their immediate effects but also have distinct
long-term political consequences.

In conclusion, this chapter strengthens the argument that autocoups are a
critical yet under-explored mechanism of authoritarian survival, one that
warrants further investigation to fully understand its implications for global
democratization and authoritarian resilience.

## Conclusion

### Main findings

This study provides critical insights into the dynamics and implications of
irregular power transitions, with a specific focus on coups and autocoups. The
research illuminates the complex interplay between incumbents and challengers
vying for power, yielding three key findings.

-   **Coup attempt determinant**: The expected success rate significantly
    influences the likelihood of a coup attempt. This success rate is largely
    determined by the balance of power between incumbent leaders and challengers,
    which varies by regime type. Notably, the findings show that military regimes
    are approximately 277.7% more likely, and personalist regimes 94% more
    likely, to experience coups compared to dominant-party regimes, all else
    being equal.

-   **Autocoup concept and dataset**: I introduce a refined concept of
    "autocoup", defined as an incumbent leader's refusal to relinquish power as
    mandated. We present the first publicly available dataset of autocoup events
    from 1945 to 2023, encompassing 110 attempts and 87 successful autocoups.
    Case studies and empirical analyses demonstrate the dataset's utility for
    quantitative research.

-   **Leader longevity**: Survival analysis techniques reveal clear differences
    in leader longevity between coup-installed leaders and autocoup leaders. The
    findings reveal that, on average, coup-installed leaders are 2.23 times more
    likely to be ousted from power than autocoup leaders, all else being equal.

### Policy implications

The examination of irregular power transitions and leadership survival offers a
crucial perspective on the interrelated phenomena of democratic backsliding,
breakdown, and autocratic intensification. The findings of this study provide
logical explanations for several political trends:

-   **Global democracy regression:** This study elucidates why global freedom has
    declined for the 18th consecutive year. Irregular power transitions, whether
    through coups or autocoups, inherently violate democratic norms and disrupt
    the trajectory toward stable democracies.

-   **Within-regime democratic erosion:** The research explains why democratic
    backsliding often occurs within regimes [@mechkova2017], rather than through
    regime change. Democracies are becoming less liberal and autocracies less
    competitive, particularly due to the prevalence of autocoups since 2000
    [@bermeo2016]. As discussed in Chapter 3, autocoups extend the tenure of
    incumbent leaders without overturning the regime itself.

-   **Rise of autocoups since 2000:** The analysis also clarifies why autocoups
    have been on the rise since 2000. Incumbent leaders possess several strategic
    advantages: firstly, they have a significantly higher probability of success
    due to their incumbent vantages compared to coup plotters. Secondly, the
    consequences of failed autocoups are relatively milder than those for failed
    coup plotters, resulting in lower costs even if they fail. Lastly, leaders
    who manage to extend their rule through an autocoup often enjoy considerably
    longer tenures compared to coup-installed leaders, thus benefiting more
    substantially.

-   **Role of external pressure:** Due to the challenges of internal opposition
    to autocoups, where power is concentrated in the hands of incumbent leaders,
    external pressure from regional or international communities may play a vital
    role in encouraging adherence to constitutional processes of power
    transition. For instance, after the general election in Venezuela on July 29,
    2024, at least nine Latin American countries rejected the election results
    and called for dialogue[^3]. Although this pressure might not be effective in
    every case, it showcases the potential influence of the international
    community in discouraging future autocoup attempts.

### Limitations and directions for future research

While the study offers a novel framework for analysing irregular leadership
transitions, several limitations require further exploration:

-   **Data refinement:** Defining and classifying autocoups is a new approach.
    Future research should validate this classification system through additional
    studies and expert evaluations.

-   **Data harmonization:** The current analysis faces challenges due to
    mismatched units (country-year vs. leader) between coup and autocoup
    datasets. Future efforts should explore data harmonization techniques for
    more robust comparisons.

-   **Democratic backsliding:** While this study establishes a connection between
    irregular power transitions and democratic backsliding, further empirical
    evidence is needed to solidify this link.

Future research avenues include:

-   **Terminology and data collection:** Refining the "autocoup" concept and
    achieving wider recognition will facilitate more accurate and comprehensive
    data collection.

-   **Dataset expansion:** Expanding the autocoup dataset with more cases and
    integrating it with data on other irregular leadership transitions can
    provide a more holistic view of political survival after these events.

-   **Power dynamics and long-term impacts:** Utilizing this dataset, future
    studies can delve deeper into power dynamics at play and explore the
    long-term consequences of irregular transitions on political systems,
    particularly regarding democratic backsliding, breakdown, and personalization
    of power.

In conclusion, this study significantly contributes to our understanding of
irregular leadership transitions, focusing on coups and autocoups. By redefining
autocoups, classifying the dataset, analysing determinants, and comparing leader
longevity, I establish a robust framework for understanding irregular power
transitions and leadership survival. This work deepens our comprehension of
democratic resilience and political stability, providing a foundation for future
research to conduct further empirical analyses based on the novel autocoup
dataset and continue refining the framework.

## References {.unnumbered}

::: {#refs}
:::

## Appendix**: Datasets** {.unnumbered}

-   **Coup Model Dataset**

    -   **Dataset Name:** **`coup_model.csv`**

    -   **Description:** This dataset is specifically cleaned for the coup model
        and contains the relevant data points necessary for analysis.

-   **Autocoup Dataset**

    -   **Dataset Name:** **`autocoup.csv`**

    -   **Description:** This dataset is an original contribution of this thesis,
        compiled and curated by the author to support the research objectives.

-   **Autocoup Model Dataset**

    -   **Dataset Name:** **`autocoup_model.csv`**

    -   **Description:** This dataset is cleaned for the autocoup model and
        includes the data required for the modelling process.

-   **Cox Proportional Hazards (Cox PH) Model Dataset**

    -   **Dataset Name:** **`survival_cox_ph_model.csv`**

    -   **Description:** This dataset is used for the Cox Proportional Hazards
        model and contains the data necessary for analysing survival rates and
        hazard ratios.

-   **Time-Dependent Cox Model Dataset**

    -   **Dataset Name:** **`survival_cox_td_model.csv`**

    -   **Description:** This dataset is cleaned for the time-dependent Cox
        model, incorporating variables that account for time-dependent effects in
        survival analysis.

[^1]: CNN. *Top US General Stands Firm Amid Pentagon Turmoil.* 12 November 2020.
    Available at:
    <https://edition.cnn.com/2020/11/12/politics/mark-milley-pentagon-turmoil/index.html>
    \[Accessed 24 April 2025\].

[^2]: **BBC News,** "China's Xi Allowed to Remain 'President for Life' as Term
    Limits Removed," *BBC News*, March 11, 2018,
    <https://www.bbc.co.uk/news/world-asia-china-43361276>, accessed March 14,
    2025.

[^3]: While the world waited for the outcome, nine Latin American countries
    released a joint statement urging transparency and recognition of the voters’
    will. The nine countries are Argentina, Costa Rica, the Dominican Republic,
    Ecuador, Guatemala, Panama, Paraguay, Peru, and Uruguay. On the morning after
    the election, the same group released a second statement demanding a complete
    review of the results in the presence of independent electoral observers
    ([AS/COA](https://www.as-coa.org/articles/how-have-international-leaders-responded-venezuelas-2024-election),
    accessed on September 9, 2024).
