---
title: "Power Acquisition and Leadership Survival: A Comparative Analysis of Autocoup and Coup-installed Leaders"
author: Zhu Qi

date: last-modified
format:
  pdf: 
    documentclass: article
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    number-depth: 2
    #shift-heading-level-by: -1
    # toc: true
    # toc-title: Contents
    # lof: true
    # lot: true
    # crossref: 
    #   lof-title: Figures
    #   lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      

#include-before-body: title.tex
bibliography: references.bib

execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue)
```

## Abstract {.unnumbered}

This study examines political leader survival following irregular transitions,
comparing classic coups d'état with autocoup (incumbent power extension).
Employing a rigorous time-dependent Cox model on a new dataset, I test if the
mode of accession affects tenure length. The analysis reveals that the initial
hypothesis—that autocoup leaders survive longer—is rejected. Once dynamic
contextual variables are controlled for, the method of irregular entry (coup vs.
autocoup) does not independently predict survival. Instead, longevity is
overwhelmingly determined by regime type and economic performance. Military and
personalist regimes exhibit high instability, while GDP growth provides a strong
protective effect. The research concludes that the institutional and economic
environment, not the specific method of power seizure, is the principal
determinant of political longevity following irregular transitions.

**Keywords:** *Coups, Autocoups, Leadership Survival, Cox Model*

\newpage

## Introduction

Why some political leaders achieve lasting longevity while others are quickly
removed remains a central question in political science. While general theories
of political survival are rare, existing scholarship has emphasized a broad
spectrum of factors, ranging from objective conditions—such as economic
performance [@palmer1999; @williams2011], natural resource wealth [@smith2004],
and societal stability [@arriola2009]—to strategic behaviours, including elite
co-optation, repression, and institutional manipulation [@gandhi2007;
@morrison2009; @davenport2021].

Among the mechanisms of leadership turnover, coups d'état are the most
consequential and, consequently, the most studied form of irregular transition
[@goemans2009; @svolik2014]. Coup-driven exits account for nearly two-thirds of
all irregular leader removals between 1945 and 2015 [@frantz2016].

Yet, despite extensive research on classic coups, a crucial and growing subset of
irregular transitions remains systematically underexamined: the autocoup. Defined
as an incumbent leader dismantling institutional constraints to illegally
entrench or extend their rule, this form of self-engineered power extension has
received far less attention than traditional coups, even as its empirical
relevance rises. Since 1945, at least 64 successful autocoups have been
identified, constituting a substantial share of irregular leadership trajectories
[@zhu2024]. Crucially, the success rate of autocoups (77 percent) significantly
exceeds that of traditional coups (approximately 50 percent) [@powell2011],
suggesting they represent a uniquely effective pathway to prolonged power.

Comparative analysis of coup-installed and autocoup leaders is particularly
scarce. This omission is surprising because preliminary evidence suggests
fundamentally different survival dynamics. Despite the shared challenge of
irregular accession, leaders who achieved power via autocoups exhibit
significantly longer average post-event tenures—a gap that can exceed five
years—compared to those installed by coups. A log-rank survival test confirms a
statistically significant divergence, with autocoup leaders consistently facing
lower hazards of removal (@fig-logrank).

```{r}
#| label: data
#| include: false
survival_cox_ph_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_ph_model.csv") |> 
  mutate(survival_year = survival_days / 365)

survival_cox_td_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_td_model.csv")

avg <- survival_cox_ph_model |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)
```

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of autocoup and coup-installed leaders


# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = survival_cox_ph_model |> 
                 filter(group != "A"))
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = survival_cox_ph_model, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("Autocoup leaders", "coup-installed leaders"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-installed leaders",
           font.title = 12)
```

Building on these observations, this article argues that the mode of irregular
accession is a fundamental determinant of leadership survival. Coup-installed
leaders typically confront heightened and immediate challenges from competing
military factions and entrenched societal forces. In stark contrast, autocoup
leaders consolidate control from a position of incumbency, allowing them to
neutralize institutional veto players and rivals *before* the irregular
transition is complete.

Using Cox proportional hazards and time-dependent Cox models, this study tests
this claim. While preliminary models suggested that the method of accession
impacted longevity, the more robust analysis reveals no statistically significant
difference in survival risk between autocoup leaders and coup-installed leaders.
This finding indicates that the institutional and economic context, rather than
the initial mode of irregular entry, is the crucial factor shaping post-accession
tenure.

This study makes two core contributions. First, methodologically, it provides a
rigorous systematic empirical investigation into the survival of autocoup
leaders, a crucial but understudied subset of irregular transitions, by employing
superior time-dependent survival models. Second, substantively, by showing that
the initial method of accession is subsumed by contextual factors (regime type
and economic dynamics), this research refines our understanding of authoritarian
durability. It redirects scholarly focus from the specific event of the coup or
autocoup toward the structural conditions that truly enable or curtail political
longevity in the 21st century.

The remainder of this article proceeds as follows. Section 2 introduces the
definition and data coding of autocoups. Section 3 reviews the theoretical
foundations of political survival. Section 4 details the research design and
methodology. Section 5 presents and interprets the empirical findings. Section 6
concludes by discussing how the method of accession reshapes our understanding of
authoritarian durability and the future of irregular transitions.

## Autocoup: Definition and Dataset

While studies of irregular leadership transitions traditionally focus on coups
d'état due to their frequency and impact, this article examines a distinct form:
an incumbent leader's refusal to relinquish power. This phenomenon has received
comparatively less scholarly attention, despite its growing importance. Since the
end of the Cold War, classic coups have declined, while these "incumbent
retention" or "overstay" strategies have become more frequent
[@ginsburg2010evasion; @baturo2014; @versteeg2020law].

### Terminology

The literature employs a diverse terminology for this phenomenon. The most
prevalent term is ‘self-coup’ (or its Spanish equivalent, autogolpe)
[@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017; @marsteintredet2019].
This term gained prominence after Peruvian President Alberto Fujimori dissolved
Congress and suspended the constitution in 1992 [@mauceri1995; @cameron1998].
However, 'self-coup' can be misleading; as @marsteintredet2019 observes, the
leader acts against state institutions, not against themselves.

Other terms use modifiers to specify the mechanism, such as "presidential,"
"executive," "constitutional," or "judicial" coups, while others describe the
process, like "slow-motion" or "soft" coups [@marsteintredet2019]. Another
prominent concept is ‘incumbent takeover’, defined as "an event perpetuated by a
ruling executive that significantly reduces the formal and/or informal
constraints on his/her power" [@svolik2014; @baturo2022, p.374].

These varying terms often create conceptual ambiguity, focusing on procedural
mechanisms or conflating legal and extra-legal actions. This study adopts the
term ‘**autocoup**’ as the most analytically coherent. It precisely identifies
both the actor (the incumbent, auto-) and the act (a coup), clearly conveys the
illegitimate nature of the behaviour, and establishes a theoretical link to
conventional coups. It thus provides a robust term for the unified analytical
framework this study seeks to establish.

### Definition

Existing definitions of irregular power retention, particularly the encompassing
term 'self-coup', often fail to adequately distinguish between two fundamentally
different concepts: *power expansion* and *tenure extension*. Power expansion
refers to an incumbent's acquisition of greater authority or control over state
apparatuses during their mandated term. In contrast, Tenure extension
specifically denotes a leader's illegitimate act of prolonging their time in
office beyond the originally mandated constitutional limit.

Most previous research on autocoups does not sufficiently distinguish between two
related but fundamentally distinct concepts: power expansion and tenure
extension. However, this study contends that a precise definition of the autocoup
must prioritize tenure extension as its defining characteristic. This focus
offers greater conceptual clarity and is more easily operationalised. While power
expansion frequently serves as a prerequisite for or accompaniment to tenure
extension, it is the extension itself, not the expansion of authority, that
constitutes the autocoup event.

I therefore define an autocoup as ***the extension of an incumbent leader’s
tenure in office beyond the originally mandated limit, achieved through
extra-constitutional means***.

This definition places the violation or evasion of mandated term limits at the
centre of the concept, providing a more precise and consistent framework for
analysis.

### Data coding

The autocoup dataset is constructed by integrating and refining several
established political science datasets to ensure reliability and
comprehensiveness (see @tbl-source).

I use the Archigos dataset [@goemans2009] and the Political Leaders’ Affiliation
Database (PLAD) [@bomprezzi2024wedded] to identify de facto national leaders and
their precise time in office from 1945 to 2023. These sources are essential for
distinguishing de facto rulers from nominal heads of state.

The Incumbent Takeover dataset [@baturo2022], which synthesizes eleven sources,
serves as my primary inventory of potential events. This dataset catalogs a broad
range of cases where executives curtailed institutional constraints on their
authority. However, because that dataset includes both power expansion and tenure
extension cases, I cross-referenced its entries with Archigos and PLAD to isolate
only those events that meet our specific definition of an autocoup (i.e., those
involving tenure extension).

```{r}
#| label: tbl-source
#| tbl-cap: Main Data Sources for Coding the Autocoup Dataset
#| echo: false
#| results: asis

read.csv(here("data/autocoup_source.csv")) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(100),
    table.align = "center"
  )
```

```{r}
#| label: data-autocoups
#| echo: false

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))

# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))
# nrow(Autocoup) - nrow(Autocoup |> filter(!is.na(source)))
```

In total, I identified and coded 83 autocoup events: 50 were adapted from cases
within the Incumbent Takeover dataset and 33 were newly identified by the author
through cross-verification of Archigos, PLAD, and contemporary news sources.

While the dataset builds on the valuable Incumbent Takeover project, it is not a
replication. The principal point of departure is conceptual: of the 279 cases
catalogued in the Incumbent Takeover dataset, I excluded 229 because they
entailed power consolidation without an accompanying attempt to extend the
leader’s tenure. Such events fall outside our operational definition of an
autocoup.

## Theoretical Framework: Survival Dynamics of Autocoup and Coup-Installed Leaders

While the literature on leadership survival is extensive, studies of irregular
transitions often either focus heavily on coup-installed leaders or aggregate
various types of non-constitutional rulers. This article addresses a critical gap
by disaggregating these categories, conducting a focused comparison between two
distinct types of irregular leaders: those installed by a coup and those who
execute an autocoup. Based on preliminary evidence from @fig-logrank, I argue
that despite both paths representing an irregular transition, the method of entry
or retention fundamentally alters a leader's subsequent survival prospects. This
section establishes the theoretical framework for this comparison, beginning with
key definitions, then analyzing the divergent challenges these leaders face, and
concluding with a testable hypothesis.

### Key definitions and scope

To ensure conceptual clarity, I first operationalise key terms. Following
@powell2011, a coup is defined as "illegal and overt attempts by the military or
other elites within the state apparatus to unseat the sitting executive." As
defined previously, an autocoup is "the extension of an incumbent leader’s tenure
in office beyond the originally mandated limit, achieved through
extra-constitutional means."

From these definitions, I also identify two leader types. Coup-Installed Leader
is the individual who assumes power *following* a successful coup, regardless of
their direct role in its execution. This broad definition includes both the
primary instigators and those selected to lead the post-coup regime. Autocoup
Leader is an incumbent leader who successfully uses extra-constitutional means to
extend their tenure, effectively beginning a new, irregular term in office.

This study compares the post-autocoup tenure of autocoup leaders with the
post-coup tenure of coup-installed leaders. To ensure a meaningful analysis of
survival dynamics and filter out ephemeral episodes, I apply a six-month tenure
threshold to both groups. This comparative focus is motivated by the distinct
challenges of illegitimacy, uncertainty, and instability that both leader types
face, allowing for a nuanced analysis of how their different starting positions
influence longevity.

### Challenges in power consolidation

Both autocoup and coup-installed leaders must consolidate power in the face of
significant challenges. However, the nature and intensity of these challenges
differ markedly across the three key domains of illegitimacy, uncertainty, and
instability. As @tbl-leaders outlines, these differences place coup-installed
leaders at a significant relative disadvantage.

\blandscape

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-installed leaders 
#| echo: false
#| results: asis

read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(150),
    2:3 ~ px(300)
  )
```

\elandscape

#### Illegitimacy

While both leaders lack democratic legitimacy, its manifestation differs
significantly. Coup-Installed Leaders face immediate and unambiguous
illegitimacy. Their power originates from an overt, often violent, rupture of the
established order. This act generates immediate domestic and international
condemnation, undermining pre-existing norms and institutions.

Autocoup Leaders, in contrast, typically employ a strategy of manipulation. They
leverage a procedural façade, bending or breaking legal institutions to create a
veneer of constitutionality. While this "democratic" cover may be thin, it
obscures the irregular nature of the transition, mitigates immediate backlash,
and buys the leader critical time to consolidate power.

#### Uncertainty

The irregular path to power creates uncertainty regarding tenure and succession
for both types. Coup-Installed Leaders confront a threefold uncertainty. First,
the immediate aftermath involves a power struggle within the new ruling coalition
or junta. Second, their tenure is inherently precarious, subject to internal
rivalries, popular unrest, or counter-coups. Third, the lack of established
succession mechanisms amplifies this ambiguity.

Autocoup Leaders present a more certain picture. The central question of *who*
rules is already settled: the incumbent. By retaining power, the autocoup leader
signals an intention to rule indefinitely or on an extended timeline, which can,
in itself, create a perception of stability and predictability, at least in the
short term.

#### Instability

Precarious legitimacy and uncertainty breed instability, but the stabilization
tasks for each leader type are distinct. Coup-Installed Leaders must rapidly
reshape the state's power dynamics. This often requires purges of the old elite
and crackdowns on opponents, generating significant instability and alienating
potential allies. They must simultaneously build a new support structure while
managing domestic and international pressures, limiting their options and
undermining long-term stability.

Autocoup Leaders benefit from institutional and personnel continuity. Because
they are already in power, they can implement changes more gradually, minimizing
disruption and mitigating backlash. They face opposition, but they are less
likely to confront an immediate, existential threat, affording them more time and
leverage to solidify their new grip on power.

### Hypothesis Development

Preliminary empirical evidence supports this theoretical distinction. Analysis of
survival data, as depicted in @fig-logrank, illustrates that the average survival
period following an autocoup is approximately five years longer than that of a
leader installed by a conventional coup.

This survival gap can be explained by the self-perpetuating cycles created by
their differing consolidation challenges. Coup-installed leaders begin from a
position of profound weakness. Their overt illegitimacy and the instability from
the coup itself make it difficult to attract reliable support. This environment
of persistent uncertainty is compounded by historical precedent; as @powell2011
notes, coups often occur in cycles, with over a third of all attempts since 1950
occurring in the ten most coup-prone countries. This history reinforces the
leader's precarity, inviting further challenges and shortening their expected
tenure.

Autocoup leaders, conversely, leverage their initial advantages. The veneer of
legality and the continuity of state institutions provide a stronger foundation.
They are better positioned to attract and maintain support, consolidate power
incrementally, and face fewer immediate threats of overthrow. This initial
stability contributes to longer tenures, which in turn reinforces the perception
of their durability.

Based on this theoretical framework and the divergent challenges identified, this
study proposes the following hypothesis:

> **H1: Political leaders who successfully extend their tenure through autocoups
> will have a longer post-event tenure than leaders who are installed by a coup
> d'état.**

Testing this hypothesis will quantify the impact of the method of irregular power
acquisition on leadership longevity, offering a more nuanced understanding of
political survival.

## Research design

To test the central hypothesis that autocoup leaders exhibit longer survival in
office than coup-installed leaders, I utilize survival analysis—a set of
statistical methods designed to model the time until an event occurs (in this
case, the time until a leader is removed from office). Specifically, I employ Cox
proportional hazards and time-dependent Cox models to estimate the effect of the
primary explanatory factor (leader type) on tenure length while controlling for
relevant covariates.

### Methodology: Survival analysis

I employ two variants of the Cox model to analyse leadership survival. Cox
Proportional Hazards (PH) model incorporates time-invariant covariates (e.g.,
leader's age, country's coup history) measured at the *start* of the irregular
tenure. It assumes the covariates have a constant proportional effect on the
hazard rate over time.

Time-dependent Cox model allows for the inclusion of covariates whose values vary
over time, such as annual economic performance or levels of political violence.
This approach offers a more dynamic and robust analysis of survival.

The Cox model is preferred over the Kaplan-Meier estimator due to its capacity to
account for multiple explanatory variables simultaneously. The model estimates
the hazard ratio, which reflects the relative risk of a tenure-ending event
(ouster) at any given time. A higher hazard ratio corresponds to a lower
probability of survival, thus capturing the critical dynamics of leadership
vulnerability.

### Data and Variables

The analysis relies on the following operationalised variables.

#### Dependent Variables

The dependent variable for survival analysis consists of two components:

**Survival Time:** This measures the duration of the leader's tenure in days.

-   For coup-installed leaders, this period begins on their day of accession.

-   For autocoup leaders, it begins on the date their original, legitimate term
    was set to expire, marking the start of their irregular tenure.

**End Point Status (Event):** This is a binary variable indicating how the tenure
concluded:

-   **0 = Censored:** The leader's tenure ended through regular or 'natural'
    means (e.g., term expiration, natural death, electoral loss, voluntary
    retirement).

-   **1 = Ousted (Event):** The leader was forcibly removed from office (e.g.,
    via a subsequent coup, popular uprising, resignation under pressure, or
    assassination).

#### Key independent variable

The primary independent variable of interest is **Leader Type**, a categorical
variable with three levels:

-   Regular Leader (Reference Group): A leader who assumed power through regular,
    constitutional means.

-   Autocoup Leader: An incumbent who extended their tenure through
    extra-constitutional means.

-   Coup-Installed Leader: A leader who assumed power following a coup d'état.

#### Control variables

To isolate the effect of leader type, the models control for key covariates as
influencing leadership stability.

**Regime Type**: A categorical variable (democracy, hybrid, autocracy) to account
for broad institutional differences in leadership stability and turnover norms.

**Polity V Score**: Used to control for the specific institutional
characteristics and degree of democratic or autocratic constraints on the
executive.

**Economic Performance**: Measured by macroeconomic indicators (e.g., GDP growth)
that influence a leader's resource base and popular support.$$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
$$ {#eq-eq6}

**Political Violence**: Accounts for the extent of civil conflict or unrest that
can directly threaten a leader's tenure.

**Population Size**: Included (often in log form) to control for structural
differences and governance challenges across states.

#### Data Sources

Data for all variables are compiled from several sources:

**The Archigos dataset** [@goemans2009] and the Political Leaders’ Affiliation
Database (PLAD) [@bomprezzi2024wedded] provide the core data on leader tenure and
exit types.

**The Autocoup Dataset** (introduced in this study) is used to identify autocoup
leaders and the start date of their irregular tenure.

**Control variable data** is sourced from standard datasets (e.g., Polity V,
World Bank, and datasets on political violence).

## Results and discussion

### Model results

The regression estimates from both the Cox Proportional Hazards (PH) model and
the time-dependent Cox model are presented in @tbl-cox. The two specifications
yield divergent findings regarding this study's central hypothesis.

```{r}
#| label: model

common_labels <- list(
  group ~ "Leader Type",
  regime ~ "Regime Types",
  GDP_trend ~ "GDP Growth Trend",
  GDP_pc ~ "GDP per capita",
  pop_log ~ "Population: log",
  polity5 ~ "Polity V score",
  violence ~ "Political violence"
)

model_cox1 <- survival_cox_ph_model |>
  mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other"),
      labels = c("Dominant-party", "Military", "Personal", "Presidential", "Parliamentary", "Other")
    ),
    group = factor(group,
                   labels = c("Non-coup leaders", "Autocoup leaders", "Coup-installed leaders")
    )
  ) |>
  coxph(
    Surv(survival_days, status) ~
      group + regime + GDP_trend + GDP_pc + pop_log +
      polity5 + violence,
    data = _
  )

tbl1 <- model_cox1 |>
  tbl_regression(
    exponentiate = T,
    conf.int = F,
    add_estimate_to_reference_rows = T,
    label = common_labels
  ) |>
  add_n(location = "level") |>
  add_nevent(location = "level") |>
  add_significance_stars(
    hide_p = T,
    hide_se = F,
    hide_ci = T,
    thresholds = c(0.01, 0.05, 0.1),
  ) |>
  bold_p(t = 0.05) |>
  bold_labels()


model_cox2 <- survival_cox_td_model |>
  mutate(
    regime = factor(regime,
      levels = c("dominant-party", "military", "personal", "presidential", "parliamentary", "other"),
      labels = c("Dominant-party", "Military", "Personal", "Presidential", "Parliamentary", "Other")
    ),
    group = factor(group,
                   labels = c("Non-coup leaders", "Autocoup leaders", "Coup-installed leaders")
    ),
    survt = (tstop - tstart) * 365
  ) |>
  coxph(
    Surv(T1, T2, status) ~
      group + regime + GDP_trend + GDP_pc + pop_log +
      polity5 + violence + cluster(id),
    data = _
  )

tbl2 <- model_cox2 |>
  tbl_regression(
    exponentiate = T,
    conf.int = F,
    add_estimate_to_reference_rows = T,
    label = common_labels
  ) |>
  add_n(location = "level") |>
  add_nevent(location = "level") |>
  add_significance_stars(
    hide_p = T,
    hide_se = F,
    hide_ci = T,
    thresholds = c(0.01, 0.05, 0.1),
  ) |>
  bold_p(t = 0.05) |>
  bold_labels()
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox Models for Survival Time of Different Types of Leaders


tbl_merge(list(tbl1, tbl2),
  tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**")
) |> 
  as_gt() |>
  tab_options(
    table.font.size = px(13),
    data_row.padding = px(8)
  )
```

The standard Cox PH model, which uses only time-fixed covariates, identifies a
statistically significant difference in removal risk. In this model,
coup-installed leaders face a 77% higher hazard of removal compared to non-coup
leaders (HR = 1.77, p \< 0.01). The hazard ratio for autocoup leaders (1.21) is
not statistically significant.

However, the time-dependent Cox model, which is theoretically superior as it
accounts for evolving conditions (like economic performance and political
violence), contradicts this finding. In this more robust model, no statistically
significant difference in removal risk is found between leader types. The hazard
ratios for both autocoup (HR = 1.22) and coup-installed (HR = 1.26) leaders are
statistically indistinguishable from the non-coup baseline.

Given the superior specification of the time-dependent model, we base our
principal interpretation on its results. This leads to a clear rejection of the
initial hypothesis (H1). Once key time-varying contextual factors are controlled
for, the manner of power acquisition (coup vs. autocoup) does not appear to have
an independent, statistically significant effect on leader survival.

Instead, the results indicate that survival is primarily determined by
institutional and contextual factors:

-   Regime Type: This emerges as the most powerful predictor. Compared to the
    baseline of dominant-party regimes, leaders in military regimes face a 217%
    higher risk of removal (HR = 3.17, p \< 0.01). Leaders in personalist regimes
    (HR = 1.78, p \< 0.01) and "Other" regimes (often transitional or
    provisional) (HR = 2.10, p \< 0.05) are also significantly more vulnerable.

-   Economic Performance: Economic conditions are strongly associated with
    survival. Sustained GDP growth provides a powerful protective effect; a
    one-unit increase (e.g., 1% growth above the 5-year average) is associated
    with an 87% reduction in the hazard of removal (HR = 0.13, p \< 0.01). Higher
    GDP per capita also has a modest but significant protective effect (HR =
    0.96, p \< 0.01).

-   Political Instability: Political violence is a destabilising force. A
    one-unit increase in the political violence index is associated with a 6%
    increase in the removal hazard (HR = 1.06, p \< 0.05).

Other variables, including population size (log-transformed) and Polity V scores,
did not achieve statistical significance in the time-dependent model.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard Ratios and 95% CIs for Leader Ousting
#| fig-subcap:
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c(
    "Autocoup leaders",
    "Coup-installed leaders",
    "Military",
    "Personal",
    "Presidential",
    "Parliamentary",
    "Ohter",
    "GDP Growth Trend",
    "GDP per capita",
    "Population: log",
    "Polity 5",
    "Political violence"
  ))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "skyblue"
  ) +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw()

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |>
  mutate(term = c(
    "Autocoup leaders",
    "Coup-installed leaders",
    "Military",
    "Personal",
    "Presidential",
    "Parliamentary",
    "Ohter",
    "GDP Growth Trend",
    "GDP per capita",
    "Population: log",
    "Polity 5",
    "Political violence"
  ))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "skyblue"
  ) +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw()
```

@fig-coxHR displays the hazard ratios (HRs) with 95% confidence intervals for all
covariates in the Cox Proportional Hazards (PH) model (left panel) and the
time-dependent Cox model (right panel). Dots represent estimated HRs, and
horizontal bars indicate the 95% confidence intervals. A value of 1 denotes no
effect on the hazard of removal. Covariates whose intervals cross the vertical
reference line at 1 are not statistically significant at the 5% level. Because it
captures time-varying conditions, the time-dependent Cox model forms the primary
basis for interpretation.

### Discussion

This result carries an important theoretical implication: after an irregular
seizure of power, the initial symbolic or procedural differences between coups
(external overthrow) and autocoups (internal institutional manipulation) lose
explanatory leverage. Both leader types immediately confront similar structural
vulnerabilities—contested legitimacy, heightened elite uncertainty, and limited
institutionalized authority. Their subsequent survival depends less on how they
entered office and more on how effectively they navigate this shared environment
of instability.

The rejection of the initial hypothesis, however, does not diminish the study's
value; rather, it shifts the focus to the proximate political and economic
contexts that truly dictate longevity. The analysis underscores that
institutional and economic conditions—not accession type—are the primary
determinants of survival. Regime type stands out as the strongest predictor:
leaders in military, personalist, and transitional regimes face substantially
higher risks of removal. This finding is consistent with theories linking high
turnover to weak institutions and poorly regulated elite competition. In these
settings, authority rests on fragile bargains and coercive capacity rather than
stable rules of succession, leaving leaders highly exposed to internal rivals.

Two important contextual notes must be clarified regarding the findings. First,
these results are conditional on the exclusion of leaders who survived for fewer
than 180 days. This methodological choice is analytically justified, as
short-lived leaders (disproportionately found among coup-installed regimes) often
fail to achieve even minimal consolidation and would artificially skew the hazard
rates. Second, the statistical finding that accession type is not a predictive
factor does not negate the observational reality that autocoup leaders still
enjoy longer average tenures than coup-installed leaders. The result simply
excludes the mode of accession as the causal predictive factor once time-varying
structural and performance variables are introduced into the model.

### Methodological Validation

As a final methodological check, the proportional hazards assumption was
validated using Schoenfeld residuals. The test results confirm that the
assumption is satisfied in both the standard Cox PH model (Global $p=0.12$) and
the time-dependent Cox model (Global $p=0.23$), ensuring the reliability of the
hazard ratio estimates.

```{r}
#| label: robust_test
#| include: false
test <- cox.zph(model_cox2)
```

## Conclusion

This study provided a systematic analysis of the survival prospects of political
leaders who assumed office through irregular transitions (coups and autocoups),
employing robust survival analysis techniques, including the time-dependent Cox
model.

The principal finding of this study is that the mode of power acquisition does
not independently predict leadership duration once dynamic contextual variables
are considered. Although the simpler standard Cox model suggested a higher risk
of removal for coup-installed leaders, this effect dissipated in the more
rigorous time-dependent model. This result indicates that leadership type itself
is not a significant determinant of political survival, challenging the initial
hypothesis.

Instead, leader survival is overwhelmingly determined by structural and
performance-based factors:

-   Regime Type: This emerged as the most influential predictor. Leaders in
    military, personalist, and transitional (“other”) regimes face significantly
    higher hazards of removal than those in dominant-party systems. This
    highlights the profound institutional instability and elite factionalism
    associated with less consolidated political structures.

-   Economic Performance: Economic development is critically linked to stability.
    While higher GDP per capita is associated with greater leadership stability,
    GDP growth exerts an especially strong protective effect. Even modest
    increases in growth rates substantially reduce the risk of removal,
    reinforcing the link between performance-based legitimacy and political
    tenure.

-   Political Violence: Conversely, political violence consistently increases the
    likelihood of ousting, underscoring the destabilising impact of societal
    unrest and conflict on a leader's ability to maintain control.

Other structural variables, such as population size and Polity V scores, did not
attain statistical significance, suggesting that under conditions of irregular
accession, proximate and dynamic factors are more consequential than long-term
institutional attributes.

The robust findings regarding the factors that truly predict leadership
survival—or removal—carry clear policy implications for all leaders, regardless
of their path to power. The results show unequivocally that personalist rule,
military regimes, and poor economic management significantly heighten the risk of
violent overthrow. Therefore, for any leader seeking long-term stability and
reduced risk of removal, the evidence strongly suggests two key priorities:

-   Promote Institutionalized Governance: Leaders should strive to move away from
    highly personalist rule or military governance, as these regimes inherently
    raise the hazard of removal. Even under non-democratic conditions, fostering
    more rules-based, inclusive, and institutionalized governance systems can
    reduce the elite uncertainty and factionalism that fuel coups.

-   Prioritize Economic Performance: Maintaining high economic growth and
    improving GDP per capita is paramount. Leaders must prioritize policies that
    deliver tangible economic benefits, as sustained poor economic performance is
    one of the clearest signals that significantly raises the probability of them
    being violently ousted. The study suggests that focusing on these structural
    and performance metrics is far more effective for ensuring political
    longevity than attempting to manipulate the initial accession process.

In sum, this study concludes that regime characteristics and economic dynamics,
rather than the mode of accession, are the principal determinants of political
survival following irregular transitions to power.

Methodologically, this work advances the field by demonstrating the analytical
necessity of time-dependent survival models in leadership research.
Substantively, it provides one of the first systematic empirical investigations
into the survival of autocoup leaders, contributing valuable data to the growing
literature on irregular leadership transitions, while also pointing to the
necessity of future refinement and expansion of the newly constructed autocoup
dataset. These findings contribute to broader debates on authoritarian
resilience, executive instability, and the structural foundations of political
longevity.

```{r}
#| label: references
#| results: asis
cat("\\newpage
## References")
```
