```{r}
library(tidyverse)
library(janitor)
library(gtsummary)
library(gt)
library(democracyData)
gwf_all |> 
  view()
gwf_all_extended |> 
  view()
library(splines)

REIGN |> 
  head()
```

gwf_party Binary indicator of party regime type (groups party-based,
party-personal, party-military, party-personal-military, oligarchy, and Iran
1979-2010)

gwf_personal Binary indicator of personalist regime type

gwf_military Binary indicator of military regime type (groups military,
military-personal, indirect military)

gwf_monarchy Binary indicator of monarchy regime type

```{r}
REIGN |> 
  distinct(gwf_regimetype) 
```

```{r}
REIGN |> 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant_party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) |> 
  # filter(regime_type == "other") |> 
  filter(year > 1949) |> 
  # view()
  tabyl(regime_type) |> 
  filter(regime_type != "other") |>
  mutate(percent = percent(percent),
         regime_type = str_to_title(regime_type),
         no_of_country_year = n,
         .before = n) |>
  select(-n) |>
  add_row(total = "Total",
          sum_no_of_country_year = sum(no_of_country_year),
          sum_percent = sum(percent)) |> 
  arrange(-no_of_country_year) |>
  clean_names(case = "title") |>
  gt() |> 
  tab_options(table.width = pct(90),
              column_labels.font.size = 16,
              column_labels.font.weight = "bold") |>
  opt_row_striping() |>
  tab_source_note(source_note = md("***Source: REIGN dataset***"))
```

```{r}
gwf_all |> 
  mutate(regimetype = case_when(
    gwf_party == 1 ~ "Dominant_Party",
    gwf_military == 1 ~ "Military",
    gwf_personal == 1 ~ "Personal",
    gwf_monarchy == 1 ~ "Monarchy",
    gwf_nonautocracy == "democracy" ~ "Democracy",
    TRUE ~ "other"
  )) |>
  select(1:5, regimetype) |>
  filter(year > 1949) |>
  tabyl(regimetype)
```

```{r}
REIGN %>% 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant-party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) %>% 
  filter(year > 1949) %>%
  tabyl(regime_type) %>% 
  arrange(-n) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                         across(where(is.character), ~'Total'))) %>%
  mutate(percent = percent(percent),
         regime_type = str_to_title(regime_type),
         counts_of_country_year = n) %>%
  select(-n) %>%
  clean_names(case = "title") %>%
  gt() %>% 
  tab_options(table.width = pct(90),
              column_labels.font.size = 16,
              column_labels.font.weight = "bold") %>%
  opt_row_striping() %>%
  tab_source_note(source_note = md("***Source: REIGN dataset***"))

```

```{r}

```

Dominant party regimes

In a typical dominant party regime, the ruling party holds supreme power, with
the incumbent leader serving as the representative of the party and subject to
the collective leadership of the party. Examples such as the Partido
Revolucionario Institucional (PRI) in Mexico, the Revolutionary Party of Tanzania
(CCM), and Leninist parties in various East European countries illustrate this
dynamic (Geddes 1999). The party's power is enshrined in its well-structured and
highly disciplined organization, with internal institutions, regulations, and
ordinations governing its members.

The selection of the incumbent is controlled by the party, ensuring their support
from within. Any power challenger faces not only the incumbent and their allies
but the entire party apparatus. This balance of power makes success elusive for
coup perpetrators, as seen in Mexico's history, where three power challengers
from the dominant party, PRI, ran for the presidency under different party
banners in 1940, 1952, and 1988 (Langston 2002). Despite considerable support,
they all ultimately failed, highlighting the difficulty of challenging the
dominant party's rule.

Although not coups, these examples demonstrate the similar balance of power
between challengers and the dominant party. Moreover, leadership in dominant
party regimes typically has term limits, providing opportunities for succession
through legitimate means. The collective leadership of the party selects
successors, aiming to maintain governance stability by choosing the most
competent and suitable candidates. This reduces the likelihood of successful
coups, as challengers are less likely to garner support when a respected
successor has been appointed.

Mexico's PRI serves as a notable example of a dominant party regime, where the
president wields extensive power during the authoritarian period. Despite this,
strict adherence to term limits prevented any coups during PRI's rule. This
example illustrates how the balance of power within dominant party regimes
mitigates coup attempts.

In summary, the power balance inherent in dominant party regimes reduces the
likelihood of coup success and consequently leads to fewer coup attempts compared
to military and personal regimes.

However, as @powell2012 posits, while measures aimed at pacifying the military
may diminish the likelihood of a coup being attempted, greater military resources
may simultaneously enhance the probability of a coup's success.

There are three dimensions when we analyse coup outcomes. The first dimension is
straightforward: why some coups succeed while others fail [@quinlivan1999;
@kebschull1994]. The second dimension focuses on the potential influences: what
happens after failed coups and successful coups. In particular, whether it helps
the shifts towards democracy following successful coups. This debate is
exemplified by the arguments between Michael Miller [@miller2016] and
Derpanopoulos, Frantz & Geddes [@derpanopoulos2016; @derpanopoulos2017]. However,
the third dimension, which explores how the expected success of coups influences
coup attempts, remains underexplored.

Among the limited discussions on the influence of coup success on coup attempts,
@powell2012 utilizes a selection model with global data spanning 1961 to 2000.
Powell's findings suggest that coup-proofing strategies, such as balancing and
providing resources to the military, reduce the likelihood of coups but increase
their chances of success.

However, few of the aforementioned studies have explored the interdependence
between coup attempts and coup success. In other words, as demonstrated in
@eq-eq1 and @eq-eq2, they do not consider the likelihood of coup success as a
primary determinant of coup attempts.

If the likelihood of coup success is predetermined, what factors influence or
potentially determine the success rates of coups? Much attention is
understandably given to military strength, as it is believed that the control of
military forces often determines the success of coups.

In addition to military factors, research delves into other determinants as well.
For instance, @bell2016 contends that democracies are less susceptible to coups,
yet coup attempts against democracies are more likely to succeed.

Let's take a look at the data and variables used in this study. \## Welcome to my
political science research page. As a Ph.D. student specializing in political
science, my focus is on democratization, with a particular interest in coups.
Here, you'll find a collection of my work, where I employ various methodologies
such as experiments, surveys, and observational data analysis. Currently, I'm
engaged in a project exploring coups, self-coups, and survival analyses of
political leaders of successful coups or self-coups. Explore my research to delve
deeper into these fascinating topics.

Consectetur sapien justo dictum: metus at turpis. Quis non et sollicitudin
curabitur posuere mattis! Blandit dignissim a vitae nam phasellus integer blandit
eros, felis nibh vitae velit! Faucibus hac urna cras interdum massa et lacus
scelerisque; convallis parturient ridiculus – phasellus feugiat auctor arcu
faucibus nisi vehicula.

Lorem ad pretium: ultrices litora turpis mus viverra quisque, aliquam nam quis
aenean. Aenean lobortis varius, ultrices metus sagittis primis mauris. Risus
pharetra inceptos fringilla mollis nullam volutpat; litora sollicitudin diam
facilisi. Parturient felis nullam rutrum iaculis, at egestas tortor inceptos
porttitor.

Consectetur interdum porttitor egestas quam vivamus, tellus libero – euismod
dignissim. Ultricies nibh luctus gravida; vestibulum libero nisi mus rhoncus
fermentum platea tristique venenatis? Neque ac sollicitudin augue mattis ante
dignissim, hendrerit nostra quis. Risus mi ridiculus ante, eget ornare lacus duis
quis aptent commodo. Luctus quam sociosqu quam. Metus nunc elementum lectus
rutrum nunc dis euismod mus eu faucibus malesuada per, imperdiet mollis.

Adipiscing duis, vivamus quisque, commodo: ligula molestie felis fusce himenaeos
placerat magna nec cubilia! Vel duis – id diam porttitor phasellus tristique
erat, sed lectus! Orci laoreet rhoncus duis magna dis. Tortor nibh lobortis donec
nostra – id, porttitor semper libero curae, mollis molestie aliquam facilisis hac
integer.

Lorem inceptos dictum ante scelerisque commodo. Urna bibendum ante dignissim id
felis accumsan dis, at magnis malesuada sodales conubia. Pretium viverra luctus
vehicula: dis purus pharetra nibh pellentesque natoque. Dignissim aliquet lectus,
himenaeos sociis laoreet volutpat class?

Dolor ac varius cum natoque torquent conubia – est, blandit enim inceptos. Eu
facilisis pulvinar, natoque metus iaculis id – eu iaculis ridiculus. Praesent leo
felis elementum vel suscipit a sociis placerat. Praesent mus cum elementum fusce
ligula, pellentesque dui: magnis accumsan metus, erat potenti.

Sit eleifend nisi mauris tempor, pretium, augue inceptos rhoncus mi; consequat
varius dui posuere. Leo nascetur integer eros etiam, mauris egestas semper –
curae vulputate elementum. Dui sodales semper dis, potenti mi dis sodales ad.
Tempor id, ante mi vehicula parturient natoque potenti inceptos, nunc urna
laoreet.

Consectetur vestibulum ornare dignissim fusce sociosqu quam, malesuada, torquent
cursus aliquet. Aliquam malesuada et viverra pretium sociis habitant quisque
ridiculus! Suspendisse metus magnis imperdiet facilisis suspendisse, faucibus;
primis leo eros? Accumsan pharetra justo cras nibh fringilla suscipit, sed urna
eleifend.

Sit donec lacinia scelerisque, himenaeos morbi mollis etiam? Suspendisse dictumst
ad dignissim cum: himenaeos aenean cursus! Euismod fermentum tincidunt natoque
convallis interdum faucibus? Vel sociis pretium ullamcorper vestibulum lacus
habitasse cum, potenti, odio libero torquent ultrices et.

Elit condimentum, proin cum enim, odio venenatis pulvinar? Quisque venenatis mi
nisl porttitor pretium mattis id consequat fusce? Aptent consequat semper posuere
ullamcorper, feugiat platea litora. Imperdiet scelerisque senectus hendrerit,
semper nulla etiam auctor neque, vitae imperdiet ligula habitant primis libero
commodo justo?

```{r}
# Example in R
# Install the package if you haven't
# install.packages("splines")

# Load the package
# library(splines)

# Create a data frame
set.seed(123)
x <- 1:10
y <- rnorm(10, mean = 5 * sin(x), sd = 0.2)
data <- data.frame(x, y)

# Fit a model using natural splines
model <- lm(y ~ ns(x, df = 4), data = data)

# Plot the data and the fitted spline
plot(x, y)
lines(x, predict(model, newdata = data), col = "blue")
```

## MFX

```{r}
probit_selection <- reg_data |>
  filter(regime_type != "other") |>
  mutate(regime = fct_relevel(regime_type, "dominant-party"),
         ect = (ct -1)*100) |>
  glm(coup >0 ~ regime + ect + gdppc + violence + pre_coups, 
      family = binomial(link = "probit"), data = _)

selection_margin <- margins(probit_selection)

probit_outcome <- reg_data |>
  filter(regime_type != "other",
         coup != 0) |>
  mutate(regime = fct_relevel(regime_type, "dominant-party"),
         ect = (ct -1)*100) |>
  glm(coup > 1 ~ regime + ect + gdppc + violence, 
      family = binomial(link = "probit"), data = _)

outcome_margin <- margins(probit_outcome)
marginal_means(probit_selection, type = "response")
export_summs(
  probit_selection,
  probit_outcome,
  selection_margin,
  outcome_margin
)
stargazer(type = "text",
  probit_selection,
  probit_outcome,
  selection_margin,
  outcome_margin
)
export_summs(
  lpm,
  selection_margin,
  outcome_margin
)
library(marginaleffects)
library(insight)
marginal_effects(probit_selection, type = "response") |> 
  summary() 
marginal_effects(probit_outcome) |> 
  summary() 

avg_slopes(model_2step)
```

### clarify pacakge

```{r}
library(clarify)


s <- sim(model_2step1, n = 1000)
s
```

```{r}
sim_ame(s, var = "ect", verbose = F) |> 
  summary()
```
