```{r}
library(tidyverse)
library(janitor)
library(gtsummary)
library(gt)
library(democracyData)
gwf_all |> 
  view()
gwf_all_extended |> 
  view()
```

gwf_party Binary indicator of party regime type (groups party-based, party-personal, party-military, party-personal-military, oligarchy, and Iran 1979-2010)

gwf_personal Binary indicator of personalist regime type

gwf_military Binary indicator of military regime type (groups military, military-personal, indirect military)

gwf_monarchy Binary indicator of monarchy regime type

```{r}
REIGN |> 
  distinct(gwf_regimetype) 
```

```{r}
REIGN |> 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant_party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) |> 
  # filter(regime_type == "other") |> 
  filter(year > 1949) |> 
  # view()
  tabyl(regime_type) |> 
  filter(regime_type != "other") |>
  mutate(percent = percent(percent),
         regime_type = str_to_title(regime_type),
         no_of_country_year = n,
         .before = n) |>
  select(-n) |>
  add_row(total = "Total",
          sum_no_of_country_year = sum(no_of_country_year),
          sum_percent = sum(percent)) |> 
  arrange(-no_of_country_year) |>
  clean_names(case = "title") |>
  gt() |> 
  tab_options(table.width = pct(90),
              column_labels.font.size = 16,
              column_labels.font.weight = "bold") |>
  opt_row_striping() |>
  tab_source_note(source_note = md("***Source: REIGN dataset***"))
```

```{r}
gwf_all |> 
  mutate(regimetype = case_when(
    gwf_party == 1 ~ "Dominant_Party",
    gwf_military == 1 ~ "Military",
    gwf_personal == 1 ~ "Personal",
    gwf_monarchy == 1 ~ "Monarchy",
    gwf_nonautocracy == "democracy" ~ "Democracy",
    TRUE ~ "other"
  )) |>
  select(1:5, regimetype) |>
  filter(year > 1949) |>
  tabyl(regimetype)
```

```{r}
REIGN %>% 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant-party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) %>% 
  filter(year > 1949) %>%
  tabyl(regime_type) %>% 
  arrange(-n) %>%
  bind_rows(summarise(., across(where(is.numeric), sum),
                         across(where(is.character), ~'Total'))) %>%
  mutate(percent = percent(percent),
         regime_type = str_to_title(regime_type),
         counts_of_country_year = n) %>%
  select(-n) %>%
  clean_names(case = "title") %>%
  gt() %>% 
  tab_options(table.width = pct(90),
              column_labels.font.size = 16,
              column_labels.font.weight = "bold") %>%
  opt_row_striping() %>%
  tab_source_note(source_note = md("***Source: REIGN dataset***"))

```

```{r}

```

Dominant party regimes

In a typical dominant party regime, the ruling party holds supreme power, with the incumbent leader serving as the representative of the party and subject to the collective leadership of the party.
Examples such as the Partido Revolucionario Institucional (PRI) in Mexico, the Revolutionary Party of Tanzania (CCM), and Leninist parties in various East European countries illustrate this dynamic (Geddes 1999).
The party's power is enshrined in its well-structured and highly disciplined organization, with internal institutions, regulations, and ordinations governing its members.

The selection of the incumbent is controlled by the party, ensuring their support from within.
Any power challenger faces not only the incumbent and their allies but the entire party apparatus.
This balance of power makes success elusive for coup perpetrators, as seen in Mexico's history, where three power challengers from the dominant party, PRI, ran for the presidency under different party banners in 1940, 1952, and 1988 (Langston 2002).
Despite considerable support, they all ultimately failed, highlighting the difficulty of challenging the dominant party's rule.

Although not coups, these examples demonstrate the similar balance of power between challengers and the dominant party.
Moreover, leadership in dominant party regimes typically has term limits, providing opportunities for succession through legitimate means.
The collective leadership of the party selects successors, aiming to maintain governance stability by choosing the most competent and suitable candidates.
This reduces the likelihood of successful coups, as challengers are less likely to garner support when a respected successor has been appointed.

Mexico's PRI serves as a notable example of a dominant party regime, where the president wields extensive power during the authoritarian period.
Despite this, strict adherence to term limits prevented any coups during PRI's rule.
This example illustrates how the balance of power within dominant party regimes mitigates coup attempts.

In summary, the power balance inherent in dominant party regimes reduces the likelihood of coup success and consequently leads to fewer coup attempts compared to military and personal regimes.

However, as @powell2012 posits, while measures aimed at pacifying the military may diminish the likelihood of a coup being attempted, greater military resources may simultaneously enhance the probability of a coup's success.

There are three dimensions when we analyse coup outcomes.
The first dimension is straightforward: why some coups succeed while others fail [@quinlivan1999; @kebschull1994].
The second dimension focuses on the potential influences: what happens after failed coups and successful coups.
In particular, whether it helps the shifts towards democracy following successful coups.
This debate is exemplified by the arguments between Michael Miller [@miller2016] and Derpanopoulos, Frantz & Geddes [@derpanopoulos2016; @derpanopoulos2017].
However, the third dimension, which explores how the expected success of coups influences coup attempts, remains underexplored.

Among the limited discussions on the influence of coup success on coup attempts, @powell2012 utilizes a selection model with global data spanning 1961 to 2000.
Powell's findings suggest that coup-proofing strategies, such as balancing and providing resources to the military, reduce the likelihood of coups but increase their chances of success.

However, few of the aforementioned studies have explored the interdependence between coup attempts and coup success.
In other words, as demonstrated in @eq-eq1 and @eq-eq2, they do not consider the likelihood of coup success as a primary determinant of coup attempts.

If the likelihood of coup success is predetermined, what factors influence or potentially determine the success rates of coups?
Much attention is understandably given to military strength, as it is believed that the control of military forces often determines the success of coups.

In addition to military factors, research delves into other determinants as well.
For instance, @bell2016 contends that democracies are less susceptible to coups, yet coup attempts against democracies are more likely to succeed.

Let's take a look at the data and variables used in this study.
## Welcome to my political science research page.
As a Ph.D. student specializing in political science, my focus is on democratization, with a particular interest in coups.
Here, you'll find a collection of my work, where I employ various methodologies such as experiments, surveys, and observational data analysis.
Currently, I'm engaged in a project exploring coups, self-coups, and survival analyses of political leaders of successful coups or self-coups.
Explore my research to delve deeper into these fascinating topics.
