---
title: "Autocoups Redefined: A Clarification of Concept and Introduction of A Novel Dataset"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: last-modified
date-modified: last-modified
format:
  aps-pdf: 
    keep-tex: true


bibliography: references.bib

abstract: |
  This article provides a critical reexamination of autocoups, an under-explored phenomenon in which incumbent leaders manipulate institutions to extend their tenure. By distinctly separating autocoups from the broader and more ambiguous concepts of self-coups or executive takeovers, the research introduces a refined and more nuanced definition of autocoups. Building on this clarified conceptual framework, the study unveils a novel dataset of autocoup events from 1945 to 2023, greatly enhancing the empirical foundation for future research. A robust mixed-methods approach is employed, combining three types of qualitative case studies that offer in-depth insights into the dynamics of autocoups, alongside a comprehensive quantitative analysis of the determinants influencing both autocoup attempts and their outcomes. This rigorous empirical inquiry not only underscores the utility and versatility of the new dataset but also makes a significant contribution to the existing body of literature.
  
keywords: [Coups, Autocoups, Political Leadership, Power Transitions]

execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue)
```

## Introduction

While the study of irregular leadership transitions has predominantly focused on
coups due to their frequency and significant impact, another form—the incumbent
leader's refusal to relinquish power—has received comparatively less attention
despite its importance. Recent decades, particularly since the end of the Cold
War, have witnessed a decline in classic coups and a concomitant rise in this
incumbent retention or overstay type of irregular leadership transition
[@ginsburg2010evasion; @baturo2014; @versteeg2020law].

This research aims to redefine and clarify this type of irregular leadership
transition, where leaders overstay their mandated term limits, as an
***autocoup***. Although analyses related to autocoups are not uncommon, the
existing literature exhibits several notable shortcomings:

-   **Terminological ambiguity:** The use of terms like "self-coups",
    "autocoups", "autogolpes", "incumbent takeovers", "executive aggrandizement",
    "overstay", and "continuismo" in different literature lacks clear,
    universally accepted definitions, leading to confusion and inconsistent
    application [@marsteintredet2019; @baturo2022]. This terminological ambiguity
    hinders accurate analysis and comparison across studies

-   **Limited dataset:** Due to the conceptual ambiguity surrounding autocoups,
    data collection remains in its nascent stages compared to the rich datasets
    available for classic coups.

-   **Methodological gaps:** The study of autocoups has been hindered by a
    limited dataset, resulting in a reliance on in-depth case studies
    [@cameron1998; @antonio2021; @pion-berlin2022] to explore this phenomenon.
    Notably, quantitative analysis has been underutilized in this field, with few
    studies employing statistical methods to examine autocoups.

More importantly, analyses of autocoups are often not integrated with those of
classic coups, despite their interconnected nature. As a distinct category of
coup, autocoups lack a clear and differentiated definition in relation to
traditional coups. Classic coups are typically defined as the complete removal of
incumbent leaders, with a focus on the termination of their tenure. In contrast,
autocoups often focus more on incumbent leaders consolidating power by seizing
control from other state institutions, rather than on extending their tenure. As
a result, coups and autocoups are frequently analysed in isolation. This
separation has led to a dearth of comparative analyses, hindering a more
comprehensive understanding of the complex dynamics between these two types of
irregular leadership transitions.

Examining autocoups, particularly in conjunction with classic coups, is essential
for several compelling reasons. Firstly, both coups and autocoups represent
significant and frequent means of irregular leadership transitions, underscoring
the need for a comprehensive understanding of these phenomena. Secondly,
autocoups, much like coups, have a profoundly detrimental impact on governance,
as they undermine the rule of law, erode institutional capacity, and contribute
to democratic backsliding or the personalization of authoritarian power. Thirdly,
successful autocoups, akin to successful coups, create a precedent that increases
the likelihood of future irregular power transitions, thereby perpetuating a
cycle of instability. For instance, since 1945, a striking 62% of leaders who
extended their terms through autocoups in non-democratic countries ultimately met
a tumultuous end, either being ousted or assassinated while in office
[@baturo2019]. Lastly, failed autocoups, similar to failed coups, often
precipitate instability, inciting widespread protests, violence, and even civil
wars, which can have far-reaching and devastating consequences for the affected
country and its citizens.

This chapter addresses these gaps by focusing on autocoups, aiming to clarify
terminology, refine concepts and definitions, enhance data collection, and
explore determinants through empirical analysis, contributing in three key areas:

-   **Conceptual clarification:** The term autocoup will be redefined and
    clarified, with a focus on power extension.

-   **Data collection:** A new dataset of autocoups since 1945 will be introduced
    based on this refined definition.

-   **Empirical analysis:** Utilizing this dataset, a quantitative analysis of
    the factors influencing leaders' decisions to attempt autocoups will be
    conducted.

The structure of this chapter is as follows: Section 2 will review definitions
related to power expansions and extensions, leading to a precise definition of
autocoups. Section 3 will present the new autocoup dataset. Sections 4 and 5 will
explore the determinants of autocoup attempts through case studies and
demonstrate the application of the dataset in empirical analysis. The conclusion
will summarize key findings and suggest directions for future research.

## Autocoups: A literature review and clarification of definitions

A significant limitation in the study of irregular leadership transition is the
conspicuous lack of integration between research on autocoups and classic coups.
Despite both coups and autocoups being crucial mechanisms of irregular leadership
change, the existing literature has largely treated these two phenomena in
isolation, neglecting to explore their interconnectedness and the nuanced
dynamics that govern their occurrence.

This separation is attributable to two primary factors. Firstly, previous
research has often overlooked autocoups as a distinct form of irregular
leadership transition. Secondly, a persistent conceptual ambiguity has hindered
the development of a clear and consistent definition and categorization of
autocoups.

Classic coups are typically characterized by the abrupt and comprehensive removal
of incumbent leaders, focusing on the swift termination of their tenure. In
contrast, autocoups are often defined as events wherein incumbent leaders
consolidate their authority by systematically usurping power from other state
institutions, rather than merely extending their own tenure.

The absence of a clear, differentiated definition of autocoups in relation to
their classic counterparts has exacerbated this divide. While classic coups have
been extensively studied and well-defined, autocoups remain a distinct yet
under-explored category. This definitional ambiguity has resulted in a dearth of
comparative analyses that could illuminate the complex interplay between these
two forms of power subversion.

Integrating the study of autocoups with that of classic coups is crucial for
bridging this analytical gap. Such an approach would offer a more nuanced and
comprehensive perspective on the various mechanisms by which political power can
be subverted or consolidated. By examining these phenomena in tandem, researchers
can better understand the full spectrum of irregular power transitions, from
outright removal to internal power grabs, providing valuable insights into the
nature of political instability and regime change.

To strengthen the analysis of autocoups, a crucial first step is to establish a
clear and consistent terminology, followed by a refinement of the definition of
autocoup to mitigate ambiguity and clarify its distinct characteristics.

### Terminology

The most prevalent term in autocoup literature is "self-coup," or "autogolpe" in
Spanish [@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017;
@marsteintredet2019]. This term gained academic prominence following Peruvian
President Alberto Fujimori's actions in 1992, when he dissolved Congress,
temporarily suspended the constitution, and ruled by decree [@mauceri1995;
@cameron1998]. However, as @marsteintredet2019 astutely points out, the term
"self-coup" can be misleading, as it implies a coup against oneself, which is
inaccurate since the action typically targets other state institutions or
apparatus.

Another approach to describing coups staged by incumbents involves using terms
with adjectives or modifiers, such as "presidential coup," "executive coup,"
"constitutional coup," "electoral coup," "judicial coup," "slow-motion coup,"
"soft coup," and "parliamentary coup" [@marsteintredet2019]. While these terms
can be useful in specific instances, their proliferation often creates more
confusion than clarity. Most of these terms focus on the specific methods
employed by coup perpetrators but fail to clearly identify the perpetrator,
necessitating further explanation. Moreover, many of these methods could be
employed either by or against executive leaders, further muddying the waters.

A third alternative involves terms like "incumbent takeover," "executive
takeover," or "overstay." Incumbent takeover refers to "an event perpetuated by a
ruling executive that significantly reduces the formal and/or informal
constraints on his/her power" [@baturo2022, p.374], building on earlier research
by [@svolik2014]. Meanwhile, overstay is defined as "staying longer than the
maximum term as it stood when the candidate originally came into office"
[@ginsburg2011evasion, p.1844]. These terms effectively identify the perpetrator
(the incumbent) and/or the nature of the event (overstaying/extending power).
However, they fall short in highlighting the illegality or illegitimacy of these
actions. Consequently, they cannot serve as a direct counterpart to "coup," which
clearly denotes the illegality of leadership ousters, while "takeover" or
"overstay" diminish the severity of the act.

Given that these terms often lack precision, focusing on specific methods rather
than the core act of power usurpation, this study proposes "autocoup" as the most
suitable term for this phenomenon. Unlike other terms, 'autocoup' clearly
identifies the perpetrator and the illegitimate nature of the power grab,
distinguishing it from classic coups while maintaining a parallel structure in
terminology.

### Definition

While precise terminology is undoubtedly crucial, another issue arises with
previous definitions of autocoups: what is the primary emphasis—power expansion,
power extension, or a combination of the two?

Definitions of power expansion and power extension within the field of political
science can often be ambiguous or overlapping, presenting a potential source of
confusion. To ensure greater clarity in the study of autocoups, it is necessary
to distinguish these two distinct conceptual frameworks more clearly:

-   **Power expansion:** This refers to the process by which an incumbent leader
    acquires additional authority or control over state apparatuses beyond their
    original mandate. This may involve centralizing power, reducing checks and
    balances, or encroaching on the authority of other branches such as the
    legislature or judiciary.

-   **Power extension:** This describes situations where a leader prolongs their
    tenure beyond the originally mandated term in office, often through
    constitutional amendments, cancellation of elections, or other means of
    circumventing term limits.

Existing definitions of autocoups or related concepts often suffer from ambiguity
between power expansion and extension, or they focus more on power expansion,
which has several drawbacks.

Firstly, defining autocoups primarily in terms of power expansion does not align
well with the traditional definition of a coup. A classical coup is clearly
focused on the ouster of the current leader, not merely a limitation or
restriction on their power. Using the same logic, a more appropriate definition
of an autocoup should prioritize the tenure extension of executive leadership.
Power restrictions on incumbents would not be coded as a coup as long as they
remain in office. Similarly, an executive leader acquiring more power from other
branches could be coded as power aggrandizement but not an autocoup, as long as
they step down when their term expires.

Secondly, emphasizing power expansion in autocoups often neglects the ultimate
purpose of incumbents. It is irrational for an incumbent to expand executive
power only to pass the powerful role to future leaders. Although the term
"self-coup" gained prominence from the 1992 Fujimori case in Peru, which
initially involved seizing power from other institutions, it is important to note
that Fujimori ultimately extended his term limits through constitutional
amendments. The 1993 Constitution allowed Fujimori to run for a second term,
which he won in April 1995. Shortly after Fujimori began his second term, his
supporters in Congress passed a law of "authentic interpretation" that
effectively allowed him to run for another term in 2000, which he won amid
suspicions and rumors. However, he did not survive the third term; in 2000,
facing charges of corruption and human rights abuses, Fujimori fled Peru and took
refuge in Japan [@ezrow2019].

Thirdly, measuring the extent of power expansion to qualify as an autocoup can be
challenging. As @cameron1998a defined, a self-coup is "a temporary suspension of
the constitution and dissolution of congress by the executive, who rules by
decree until new legislative elections and a referendum can be held to ratify a
political system with broader executive power" (p. 220). However, defining
"broader executive power" is inherently problematic and disputable.

Therefore, this study argues that a more accurate definition of autocoups should
prioritize power extension as its core characteristic. This approach is
straightforward and easy to identify in practice. In most cases, autocoups
involving power extension also involve power expansions as their prerequisite and
foreshadowing.

Based on these criteria, I define **an autocoup as the illegitimate extension of
an incumbent leader's term in office beyond the originally mandated limits
through unconstitutional means**. This definition emphasizes the core
characteristic of power extension while acknowledging the potential for power
expansion as a related phenomenon:

-   **Leadership Focus**: This definition refers to the actual leaders of the
    country, regardless of their official titles. Typically, this would be the
    president; however, in some cases, such as in Germany, the primary leader is
    the premier, as the president serves as a nominal head of state.

-   **Primary Characteristic**: While the primary characteristic of an autocoup
    is extending the term in office, this definition does not exclude instances
    of power expansion. Both aspects can coexist, but the extension of the term
    is the central element.

-   **Illegitimacy**: Autocoups, by their nature, subvert legal norms and
    established leadership transfer mechanisms. No matter how legitimate they
    claim to be, their illegitimacy is not beyond a reasonable doubt as long as
    the incumbents are the direct beneficiaries. This critical aspect will be
    explored further in Section 3.

By clarifying these definitions, this study aims to provide a more precise and
consistent framework for understanding and analysing autocoups, thereby enhancing
the clarity and rigor of research in this field.

## Introduction to the autocoup dataset

### Defining the scope

Classifying political events as autocoups often involves addressing borderline
cases. To maintain consistency and avoid ambiguity, this study adopts a broad
coding approach: All instances of incumbents extending their original mandated
term in office are coded as autocoups, regardless of the apparent legality of the
extension.

This approach is justified because truly legitimate amendments to power
transition institutions should apply only to subsequent leaders, not the
incumbent. Even when extension procedures appear legal, the legitimacy is
questionable when the incumbent is the direct beneficiary.

### Classifying autocoups {#sec-classify}

To maintain consistency, autocoups are categorized based on several key factors.

-   **Methods employed**: Specific strategies used by incumbents (e.g.,
    constitutional amendments, election cancellation).

-   **Degree of legality**: Extent of deviation from established legal norms.

-   **Duration of extension**: Length of time the incumbent remains in office
    beyond designated term limits.

-   **Outcomes**: Whether the autocoup attempt succeeds or fails.

This study primarily focuses on the methods employed, while coding for other
aspects when information is available.

#### Evasion of term limits

Evasion of term limits is a common tactic employed in autocoups. Incumbents often
resort to seemingly legal manoeuvres to extend their hold on power. These
manoeuvres primarily involve manipulating constitutional provisions through
various means. The incumbents may pressure legislative bodies (congress) or
judicial institutions (Supreme Court) to reinterpret existing term limits, amend
the constitution to extend terms, or even replace the constitution altogether.
This might also involve popular vote through referendums, or a combination of
these approaches. The extension can range from a single term to indefinite rule.

These manoeuvres primarily involve manipulating constitutional provisions through
various means.

-   **Changing term length:** Incumbents might lengthen the official term
    duration (e.g., from 4 to 6 years) to stay in office longer, even if the
    number of allowed terms remains unchanged. Examples, in the dataset, include
    Presidents Dacko (CAR, 1962), Kayibanda (Rwanda, 1973), and Pinochet (Chile,
    1988).

-   **Enabling re-election:** This approach involves incumbents modifying legal
    or constitutional frameworks to permit themselves to run for leadership
    again, despite initial restrictions. These restrictions might include
    prohibitions on re-election, bans on immediate re-election, or term limits
    that the incumbents have already reached. An illustrative example is
    President Menem of Argentina in 1993, who leveraged this tactic to extend his
    tenure.

-   **Removing term limits altogether:** This approach was implemented by
    President Paul Biya of Cameroon in 2008. Biya, who had been in power since
    1982, successfully pushed for a constitutional amendment that abolished
    presidential term limits. This change allowed him to run for re-election
    indefinitely, effectively opening the possibility for him to rule for life.

-   **Declaring** **leader for life:** This differs from removing term limits as
    the leader still faces elections (although potentially rigged or
    uncontested). An example is Indonesia's President Sukarno, who attempted to
    declare himself president for life in 1963 (ultimately unsuccessful).

These methods are often used in combination. Initially, the duration of a term is
extended, followed by amendments to allow re-election, then the removal of term
limits, and finally, the declaration of the leader for life. For example, Haitian
President François Duvalier amended the constitution in 1961 to permit immediate
re-election and then declared himself president for life in 1964.

#### Election manipulation or rigging

Election manipulation or rigging is the second most commonly used tactic to
extend an incumbent's tenure.

-   **Delaying or removing elections:** Delaying or removing scheduled elections
    without legitimate justification is a frequent method used by incumbents to
    maintain power. For instance, Chadian President François Tombalbaye delayed
    general elections until 1969 after assuming power in 1960. Similarly, Angolan
    President José Eduardo dos Santos suspended elections throughout his rule
    from 1979 to 2017.

-   **Refusing unfavourable election results:** Incumbents may refuse to accept
    unfavourable election results and attempt to overturn them through
    illegitimate means. For example, President Donald Trump of the United States
    refused to accept the results of the 2020 election and tried to overturn
    them.

-   **Rigging elections:** Winning elections with an extraordinarily high
    percentage of votes is highly questionable. This study will code elections
    where the incumbent wins more than 90% of the vote as autocoups. For
    instance, President Teodoro Obiang of Equatorial Guinea has consistently won
    elections with over 95% of the vote in multi-party elections since 1996,
    indicating election rigging.

-   **Excluding opposition in elections:** Manipulating the electoral process by
    excluding opposition parties or candidates from participation, effectively
    creating a one-candidate race, clearly signifies an autocoup.

#### Figurehead Installation

One strategy employed by incumbents to evade term limits is to install a trusted
associate as a figurehead, allowing the incumbent to maintain de facto control
while formally relinquishing office. This can be achieved through the creation of
seemingly subordinate positions, which in reality serve as conduits for the
incumbent's continued influence.

A notable example of this tactic is the 2008 Russian presidential transition.
Confronted with constitutional term limits, President Vladimir Putin hand-picked
Dmitry Medvedev to succeed him as president. Following Medvedev's election, he
appointed Putin as Prime Minister, ostensibly reversing their roles. However,
most observers and analysts concur that Putin continued to wield significant
behind-the-scenes influence, effectively rendering Medvedev a proxy leader.

#### Reassigning supreme authority to a new role

This tactic involves an incumbent leader manipulating the constitution or legal
framework to create a new position of power, or elevate an existing one, before
stepping down from their current role. They then strategically take on this new
position, effectively retaining significant control despite appearing to
relinquish power. For example, in 2017, Recep Tayyip Erdoğan, the Prime Minister
of Turkey, spearheaded a constitutional referendum that transitioned the country
from a parliamentary system to a presidential one. This new system concentrated
significant executive power in the presidency. Following the referendum's
approval, Erdoğan successfully ran for the newly established presidency,
effectively retaining control under a different title.

#### One-time arrangement for current leaders

This strategy involves special arrangements that extend the term or tenure of
current leaders without altering the underlying institutions. For example,
Lebanon extended President Émile Lahoud's term by three years in 2004 through a
one-time arrangement.

### Data coding

The autocoup dataset is built upon existing studies and datasets, ensuring a
comprehensive and reliable foundation. @tbl-source outlines the main sources used
for coding the autocoup dataset.

The Archigos dataset [@goemans2009] and the Political Leaders' Affiliation
Database (PLAD) [@bomprezzi2024wedded] provide comprehensive data on all leaders
from 1875 to 2023, although our coding only includes autocoups since 1945. These
datasets are invaluable for identifying actual rulers, distinguishing them from
nominal heads of state.

The Incumbent Takeover dataset [@baturo2022] integrates data from 11 related
datasets, offering a broad spectrum of cases where leaders significantly reduced
constraints on their power. This dataset includes both power expansions and
extensions, necessitating cross-referencing with Archigos to verify
qualifications for autocoups.

```{r}
#| label: tbl-source
#| tbl-cap: Main Data Sources for Coding the Autocoup Dataset
#| echo: false
#| results: asis

read.csv(here("data/autocoup_source.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(
    table.width = pct(100),
    table.align = "center") |>
  as_raw_html()
```

```{r}
#| label: data-autocoups
#| echo: false

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))
```

In total, `r nrow(Autocoup)` observations were coded, with
`r nrow(Autocoup |> filter(!is.na(source)))` overlapping with the candidate data
from Incumbent Takeover. The remaining
`r nrow(Autocoup) - nrow(Autocoup |> filter(!is.na(source)))` events were newly
coded by the author through verification with other sources such as Archigos,
PLAD and news reports.

The main deviation from the Incumbent Takeover dataset arises from excluding
power expansions that do not involve attempts to extend tenure.

The dataset encompasses a total of 14 variables along with the *notes* field.

-   **Country identification:** Country code (*ccode*) and country name
    (*country*) from Correlates of War project [@stinnett2002].

-   **Leader information:** Name of the de facto leader (*leader_name*,coded
    following Archigos and PLAD datasets).

-   **Timeline variables:** Date the leader assumed power (*entry_date*), date
    the leader left office(*exit_date*), date of the significant event marking
    the autocoup (*autocoup_date*), and Start date of the leader's additional
    term acquired through the autocoup (*extending_date*).

-   **Power transition methods:** Categorical variable for how the leader entered
    power (*entry_method*), categorical variable for how the leader exited power
    (*exit_method*), dummy variable indicating regular (1) or irregular (0) entry
    (*entry_regular*), and dummy variable indicating regular (1) or irregular (0)
    exit (*exit_regular*).

-   **Autocoup details:** Key variable capturing methods used to extend power
    (*autocoup_method*) and outcome of the autocoup attempt (*autocoup_outcome*,
    "fail and lose power", "fail but complete original tenure", or "successful").
    For successful coups, the additional term length can be calculated from the
    difference between *exit_date* and *extending_date*.

-   **Data source:** Identifies the dataset source used for coding (*source*).

-   **Additional notes:** Provides context for exceptional cases (*notes*).

There are a few coding challenges and decisions worth mention. For cases where
extensions happen incrementally, the *autocoup_date* reflects a significant event
marking the extension, such as a legislative vote or successful referendum. In
cases where a leader undertook multiple autocoup attempts, details are recorded
in the notes field. Care was taken to differentiate between cases of power
expansion and actual attempts to extend tenure, which required cross-referencing
multiple sources. Determining the success or failure of an autocoup attempt often
required in-depth research, especially for less documented cases.

### Data descriptions

```{r}
#| label: data_descriptions

country <- Autocoup |> 
  distinct(ccode) |> 
  summarise(count = n())

autocoup_desc <- Autocoup |> 
  summarise(attempted = n(),
            succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method) |> 
  arrange(-attempted)
```

The primary coding has identified `r nrow(Autocoup)` autocoup cases from 1945 to
2023, involving `r country$count` countries. This comprehensive dataset provides
a rich source of information for analysing trends and patterns in autocoup
attempts across different political contexts.

@tbl-autocoup_method presents a breakdown of the autocoup methods employed by
leaders:

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup methods and success rates (1945-2021)

autocoup_desc |> 
  adorn_totals() |> 
  mutate(Success_rate = percent(round(succeeded/attempted, 3))) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(99)
    ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: Autocoup dataset*"))
```

The most common autocoup method is "enabling re-election", accounting for
`r autocoup_desc$attempted[1]` events. This is followed by "removing term limits"
(`r autocoup_desc$attempted[2]` cases), and then "delaying elections" and
"declaring the leader for life" (each with `r autocoup_desc$attempted[3]` cases).

Autocoups have a success rate of 79%, compared to the 50% success rate of
classical coups. This high success rate can be attributed to several factors:

-   Incumbent Advantage: Leaders already in power have access to resources and
    institutional mechanisms that can be leveraged to their advantage.

-   Gradual Implementation: Unlike sudden coups, autocoups can be implemented
    gradually, allowing leaders to build support and legitimacy over time.

-   Legal Facade: Many autocoup methods operate within a veneer of legality,
    making them harder to oppose openly.

-   Control of State Apparatus: Incumbents often have significant control over
    state institutions, which can be used to facilitate their autocoup attempts.

However, success rates vary significantly across different methods.

-   **100% success rate**: Removing term limits, delaying elections, declaring
    the leader for life, and cancelling elections all have perfect success rates.
    This suggests that once these processes are set in motion, they are difficult
    to reverse.

-   **Lower success rates:** Refusing to accept election results has the lowest
    success rate, with only 1 out of 4 attempts succeeding. Although the sample
    size is limited (only 4 cases in total), this trend might suggest several
    factors at play. These include greater democratic resilience in systems where
    general elections are regularly held, heightened international scrutiny and
    pressure in response to blatant manipulation of election results, and
    stronger domestic opposition to such overt power grabs.

## Determinants of autocoup attempts: Case studies

### High frequency and success rate of autocoups in post-communist regimes

Our dataset shows a high frequency and success rate of autocoups in
post-communist countries. These nations, formerly communist regimes prior to the
collapse of the Soviet Union, have largely evolved into 'hybrid regimes'
[@nurumov2019], with only a few retaining their communist status. The data
documents 12 cases of autocoups aimed at prolonging incumbency in these
countries, with only two attempts failing. Examination of these cases highlights
several distinctive characteristics:

-   **Inherited authoritarian systems**: Despite most of these 12 countries
    transitioning from communist to non-communist governments (with the exception
    of China), they retained many authoritarian systems from their communist
    past.

-   **Continuity of former elites**: The transitions did not result in the
    removal or overthrow of previous ruling groups. Instead, former communist
    elites often maintained their positions of power.

-   **Subverted democratic processes**: While general elections and term limits
    were introduced in most of these countries, the legacy of former communist
    regimes frequently led to the circumvention of term limits and manipulation
    of elections [@nurumov2019].

#### Case 1: Lifelong ruler–Alexander Lukashenko in Belarus

Alexander Lukashenko, a former member of the Supreme Soviet of the Byelorussian
Soviet Socialist Republic, became the head of the interim anti-corruption
committee of the Supreme Council of Belarus following the dissolution of the
Soviet Union. Elected as Belarus's first president in 1994, he has maintained
this position ever since. Initially, the 1994 constitution limited presidents to
two successive terms. However, Lukashenko removed this restriction in 2004.
International monitors have not regarded Belarusian elections as free and fair
since his initial victory. Despite significant protests, Lukashenko has
consistently claimed to win with a high vote share, often exceeding 80% in each
election. This pattern is evident across all five Central Asian countries of the
former Soviet Union, where post-dissolution leaders were typically high officials
or heads of the former Soviet republics who continued their leadership in the
presidency.

#### Case 2: Transferring power to a handpicked successor–Nursultan Nazarbayev in Kazakhstan

Nursultan Nazarbayev served as the first president of Kazakhstan from 1991 until
2019. Prior to the dissolution of the Soviet Union, he held de facto leadership
as the First Secretary of the Communist Party of Kazakhstan. Following
independence, he was elected as the first president and retained office until
2019 through various means, including resetting term limits due to the
implementation of new constitutions. Notably, Nazarbayev did not officially
eliminate term limits but instead created an exemption for the "First President"
[@nurumov2019]. Unlike Lukashenko, who remains the incumbent of Belarus,
Nazarbayev transferred the presidency to a designated successor, Kassym-Jomart
Tokayev, in 2019. However, he retained significant influence as the Chairman of
the Security Council of Kazakhstan until 2022.

```{r}
#| label: latin
#| include: false

library(countrycode)
latin <- codelist |> 
  filter(region == "Latin America & Caribbean") |> 
  select(region, cown) |>
  drop_na() |> 
  pull(cown)

Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = country) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = autocoup_method) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = c(autocoup_method,autocoup_outcome)) |>
  adorn_totals()
```

### Autocoups for immediate re-election: Cases of Latin America

Latin America has a long-standing tradition of maintaining term limit
conventions. Simón Bolívar, the founding father of Bolivia, was initially a
strong advocate for term limits, stating in 1819, "Nothing is as dangerous as
allowing the same citizen to remain in power for a long time... That's the origin
of usurpation and tyranny" [@ginsburg2019, p.38]. Although Bolívar eventually
modified his stance, arguing in his 1826 Constitution Assembly speech that "a
president for life with the right to choose the successor is the most sublime
inspiration for the republican order," term limits became a convention in Latin
America. Approximately 81% of Latin American constitutions between independence
and 1985 imposed some form of term limits on the presidency
[@marsteintredet2019a].

An analysis of cases in Latin American countries reveals two notable patterns.

#### Often successful at breaking non-re-election or non-immediate re-election restrictions

Unlike other presidential systems where two terms are more common,
non-re-election or non-immediate re-election used to be prevalent in Latin
America. According to @marsteintredet2019a, non-consecutive re-election was
mandated in about 64.9% of all constitutions between independence and 1985, while
5.9% banned re-election entirely.

However, adherence to these conventions has varied across the region. Since
Mexico introduced non-re-election institutions in 1911 at the start of the
Mexican Revolution, they have remained inviolate [@klesner2019]. Similarly,
Panama and Uruguay have never altered their re-election rules, and Costa Rica has
only experienced a brief period (1897-1913) permitting immediate presidential
re-election since prohibiting it in 1859 [@marsteintredet2019a]. In many other
countries, however, constitutions have been frequently amended or violated.

The pursuit of re-election or consecutive re-election, therefore, has been a
significant trigger for autocoups aimed at power extension in this region. Our
research documents 32 autocoup cases, with over 50% (17 cases) attempting to
enable re-election or immediate re-election, and about 59% (10 cases out of 17)
being successful.

Unlike those who attempt to overstay in office indefinitely, many Latin American
leaders exit after their second term expires. Examples include President Fernando
Henrique Cardoso of Brazil (1995-2003), President Danilo Medina of the Dominican
Republic (2012-2020), and President Juan Orlando Hernández of Honduras
(2014-2022) [@ginsburg2019; @marsteintredet2019a; @landau2019; @baturo2019;
@neto2019].

#### Failing to further extend tenure

This trend does not imply that none of these leaders attempted further
extensions, but rather that most accepted their unsuccessful outcomes without
abusing their power to manipulate the process. While autocoups aimed at securing
one additional term are often successful, attempts to overstay beyond this are
frequently unsuccessful.

In contrast to the previous examples, two contrasting cases illustrate the varied
outcomes of term limit challenges:

-   **Unsuccessful extension – Carlos Menem (Argentina)**: President Menem
    successfully extended his tenure by one term through a 1994 constitutional
    amendment allowing one executive re-election. He was subsequently re-elected
    in 1995. However, his attempt to reset his term count, arguing that his first
    term (1988-1995) should not count as it was under previous constitutions, was
    unanimously rejected by the Supreme Court in March 1999 [@llanos2019]. A
    similar scenario unfolded with President Álvaro Uribe of Colombia (2002-2010)
    [@baturo2019].

-   **Successful extension – Daniel Ortega (Nicaragua)**: In contrast, Daniel
    Ortega, the incumbent president of Nicaragua, successfully extended his
    presidency. In 2009, the Supreme Court of Justice of Nicaragua permitted his
    re-election in 2011. Subsequently, in 2014, the National Assembly of
    Nicaragua approved constitutional amendments abolishing presidential term
    limits, allowing Ortega to run for an unlimited number of five-year terms. As
    a result, he has held the presidency since 2007 [@close2019].

### As common as classical coups: Cases of African countries

```{r}
#| label: africa
#| include: false

africa <- codelist |> 
  filter(continent == "Africa") |> 
  select(continent, cown) |>
  drop_na() |> 
  pull(cown)

Autocoup |>
  filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = country) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% africa) |>
  distinct(country)
Autocoup |>
  filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = autocoup_method) |>
  adorn_totals()
Autocoup |>
  #filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = autocoup_outcome) |>
  adorn_totals()
```

Classical coups have been prevalent in Africa, accounting for approximately 45%
of all global coups (219 out of 491 cases) since 1950, involving 45 out of 54
African countries (GIC dataset). While autocoups are less frequent compared to
traditional coups, they maintain a significant presence in Africa. Among 110
documented autocoup cases globally, 46% (51 cases) occurred in Africa, involving
36 countries. Notably, the success rate of autocoups in Africa is over 84% (43
out of 51 attempts), which surpasses both the success rate of classical coups in
the region (roughly 50%) and the global average success rate of autocoups (79%).

Identifying a clear pattern of autocoups in Africa is challenging, mirroring the
complexity observed with classical coups. Various factors have been proposed to
explain this phenomenon:

-   **Natural Resources**: Countries rich in natural resources, particularly oil
    or diamonds, may see leaders more likely to attempt and succeed in extending
    their terms [@posner; @cheeseman2015; @cheeseman2019a].

-   **Quality of democracy**: The quality of democracy is a critical factor
    influencing respect for term limits [@reyntjens2016].

-   **International influence**: International aid or donor influence can play a
    significant role in discouraging attempts at power extension [@brown2001;
    @tangri2010].

-   **Organized opposition and party unity**: The extent of organized opposition
    and the president's ability to enforce unity within the ruling party are
    crucial factors [@cheeseman2019].

Utilizing the Africa Executive Term Limits (AETL) dataset, @cassani2020
highlights human rights abuses and the desire for impunity as main drivers for
incumbents to cling to power. The more authoritarian a leader, the more likely
they are to attempt to break term limits and overstay in office. A leader's
ability to secure the loyalty of the armed forces through public investment
increases the chances of success in overstaying.

Despite both coups and autocoups being prevalent, there has been a noticeable
shift since the end of the Cold War in 1991: Traditional coups have decreased in
frequency while autocoups have become more prevalent.

This trend can be partially attributed to the introduction of multi-party
elections in Africa in the 1990s, which also brought in term limits for
executives [@cassani2020; @cheeseman2019]. Before 1991, personal or military rule
was more common, and term limits were less frequent. Post-1991, with more term
limits introduced, challenges to these limits have increased. However, it is
crucial to note that this increase in challenges does not necessarily imply that
violations are more common than adherence to term limits, because total power
transitions have increased compared to the past.

## Empirical analysis: An example of utilizing the autocoup dataset

The autocoup dataset opens up opportunities for quantitative analysis that extend
beyond traditional case studies. This section presents a clear example of how to
effectively leverage this dataset. To examine the factors driving autocoup
attempts, I employ a probit regression model, demonstrating its applicability for
investigating the determinants of such events. This example highlights the
dataset's potential for robust empirical inquiry and provides a foundation for
future research on the conditions that influence autocoup attempts.

### Dependent variables[^1]

-   **Autocoup attempt**: Binary variable indicating whether an autocoup attempt
    occurred (1) or not (0) during the tenure of an incumbent leader.

-   **Autocoup success**: Binary variable indicating whether an autocoup attempt
    was successful (1) or failed (0), conditional on an autocoup attempt
    occurring.

### Key independent variable: regime type

I categorize regime types following @geddes2014 (GWF), focusing on military,
personalist, and dominant-party regimes, with democracies and monarchies included
for comparison. Descriptive statistics for regime types are presented in
@tbl-regimes.

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types since 1950 

REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949) |> 
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(regime_type =str_to_title(regime_type),
         country_year =n,
         share = percent(percent),
         .keep = "unused") |>
  clean_names(case = "title") |>
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN Datasets*")) |> 
  as_raw_html()
```

### Control variables

The control variables are chosen based on the research of @gassebner2016. They
analysed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

```{r}
#| label: variables
#| results: asis
gdppc <- "-   **Economic Level:** Represented by GDP per capita. This measure provides an
    indication of the overall economic health and standard of living in a
    country. We use GDP per capita data (in constant 2017 international 1000
    dollars, PPP) from the V-Dem dataset by @fariss2022."

ct <- "-   **Economic Performance:** Measured using the current-trend ($CT$) ratio
    developed by @krishnarajan2019. This ratio compares a country's current GDP
    per capita to the average GDP per capita over the previous five years. A
    higher $CT$ ratio indicates stronger economic performance. For a country $i$
    at year $t$, the $CT$ ratio is calculated as follows:"

glue(gdppc)
glue(ct)
```

$$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
$$ {#eq-eq6}

-   **Political stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    variable “actotal” in the Major Episodes of Political Violence dataset
    [@marshall2005current], with 0 representing the most stable conditions (no
    violence at all) and 18 representing the most unstable.

-   **Previous coups:** Included in the selection equation as either: a) The
    number of previous coups in a country (Model 1), or b) The time since the
    last coup attempt (Model 2 for robustness check).

-   **Population size:** To account for its potential impact on leaders' tenures,
    we consider the log of the population size. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on power transitions. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's age:** The age of the leader is included as an additional variable
    in the analysis, offering insights into potential correlations with
    leadership strength. Older leaders may have different experiences, networks,
    and health considerations that could influence their ability to maintain
    power. This data is sourced from Archigos and PLAD datasets.

Unlike the analysis of classic coup determinants, which could theoretically occur
in any given year, I assume that an autocoup happens only once during an
incumbent leader's tenure, as a successful autocoup negates the need for another
attempt. However, this assumption does not always reflect reality, as leaders
might attempt further extensions or try again after a failed attempt. For
simplicity, I overlook these possibilities in our analysis.

Therefore, in our probit model, the unit of analysis for autocoups is the entire
tenure of a leader, rather than a country-year. I establish a base year for the
variables: for leaders who staged an autocoup, we use the year of their first
attempt as the base year; for leaders who did not attempt to overstay, I use the
middle year of their tenure as the base year.

### Results and discussions

```{r}
#| label: tbl-autocoupmodel
#| tbl-cap: "Determinants of autocoup attempts and success (1945-2018)"
#| results: asis
#| echo: false

autocoup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_model.csv")


model_attempt <- autocoup_model |> 
  filter(!regime %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential")) |>
  glm(autocoup_attempt ~ regime + GDP_pc + GDP_trend + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

model_success <- autocoup_model |> 
  filter(!regime  %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential"),
         autocoup_outcome = factor(autocoup_outcome)) |>
  filter(autocoup_attempt == 1) |> 
  glm(autocoup_outcome ~ regime + GDP_pc + GDP_trend + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

stargazer(model_attempt, model_success, 
          header = F,
          float = F,
          column.labels = c("Autocoup Attempts", "Autocoup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant",
                               "Regime: Dominant-party",
                               "\\hspace{1.6cm}Military",
                               "\\hspace{1.6cm}Personalist",
                               "GDP per capita",
                               "Economic trend",
                               "Political stability",
                               "Age",
                               "Population(log)"),
          dep.var.caption = "",
          dep.var.labels.include = F,
          summary.logical = T,
          column.sep.width = "50pt"
          )
#view(autocoup_model)
```

@tbl-autocoupmodel summarizes the findings from the probit regression models
based on our analysis of the determinants of autocoup attempts and their success.

Model 1, which examines autocoup attempts, reveals only one significant predictor
besides the constant term. Among the regime types, personalist regimes
significantly increase the likelihood of autocoup attempts, all else being equal.
This suggests that leaders in personalist regimes are more prone to attempt to
extend their power through autocoups compared to leaders in democratic regimes
(reference regime). Leaders in dominant-party and military regimes, however, show
no significant difference in the likelihood of attempting an autocoup compared to
democratic leaders.

The model for autocoup success (Model 2) shows similar dynamics. Personalist
regimes again have a strong positive and significant effect on the success of
autocoups compared to democratic leaders. Dominant-party regimes also show a
positive and marginally significant effect. However, a detailed examination
reveals that about half of the successful autocoups in dominant-party regimes (9
out of 20) exhibit a personalist style, such as "party-personal-military"
regimes.

This outcome is logical since personalist leaders are typically much more
powerful than other types of leaders, making them more inclined and capable of
overstaying in power.

Other factors play an insignificant role in determining the attempts and outcomes
of autocoups. This aligns with our conclusions on the determinants of classic
coups. Both coups and autocoups are significantly affected by power dynamics. As
power transitions involve the struggle between seizing and maintaining power, the
balance of power status quo inevitably matters in both coups and autocoups. This
also explains the high success rate of autocoups. Compared to power challengers,
incumbents are in an obviously advantageous position. Incumbent leaders can use
state power to their benefit, which is difficult to counteract. Even the abuse of
power is often unchecked under a powerful leader's rule.

The empirical analysis of autocoups yields significant implications for
real-world politics. In particular, the high overall success rate of autocoups
highlights the vulnerability of democratic institutions to gradual erosion by
incumbent leaders. The threshold for ousting or impeaching an incumbent leader
through constitutional means is exceptionally high, with success often requiring
more than a simple majority and substantial support across various sectors.
Resorting to illegal means, such as a coup, presents even greater challenges due
to high costs, severe consequences, and a low likelihood of success.

Conversely, political dynamics, whether in democracies or autocracies, tend to
favour incumbents even when they act unconstitutionally. Incumbents can leverage
state resources to achieve their political ambitions, benefiting from a high
probability of success and minimal consequences in case of failure. This
asymmetry in power and risk creates a concerning scenario: for incumbents who do
not respect constitutional institutions, the opportunity to launch an autocoup
appears sufficiently low-risk to warrant an attempt.

## Conclusion

This study conducts a thorough and comprehensive analysis of autocoups, with a
specific focus on political events where incumbent leaders illegitimately extend
their tenure in power. By refining the existing definition and distinguishing
autocoups from related concepts such as "self-coups", "autogolpes", and
"executive takeovers", this research introduces a novel dataset that catalogues
autocoups from 1945 to 2023. This refined definition and the accompanying dataset
enable the study to broaden its analysis of irregular leadership transitions.
While traditional analyses often concentrate on the abrupt termination of tenure
through coups, this research expands the scope to include the irregular extension
of tenure through autocoups. This approach provides a more comprehensive and
nuanced understanding of the phenomenon, highlighting the various mechanisms by
which incumbent leaders can subvert democratic processes to maintain their power.

The findings reveal that personalist regimes are significantly more likely to
experience autocoup attempts and succeed in these attempts compared to
democracies. Dominant-party systems, often exhibiting personalist
characteristics, also show an association with successful autocoups. While regime
type significantly influences autocoups, other factors appear less impactful,
mirroring classic coups where the balance of power is a more essential
determinant. The high success rate of autocoups can be attributed to the inherent
advantages incumbents possess, such as control over or abuse of state power and
the difficulty of removing or impeaching them through legal or illegal means.

However, several limitations warrant consideration for future research. Firstly,
the definition of an autocoup requires further commentary and discussion to gain
wider acceptance in the academic community. Despite efforts to maintain
objectivity, some coding decisions may involve subjective judgments, particularly
in borderline cases. Secondly, due to the nature of autocoups, which are less
frequent than classic coups (491 coups versus 110 autocoups during the same
period), the quantitative analysis cannot be conducted as a country-year variable
as in coup studies. This raises the issue of choosing an appropriate base year
for the analysis, which requires further discussion and potentially sensitivity
analyses.

Despite these limitations, this research significantly enhances our understanding
of the mechanisms and motivations behind autocoups, contributing to the
literature on political stability and democratic resilience. The findings
highlight the vulnerability of political systems, particularly democracies, to
erosion from within by incumbent leaders.

Future studies could build on this work by employing the dataset to explore more
nuanced power dynamics or examine the long-term impacts of these events on
political systems. Particularly fruitful areas for investigation include the
relationship between autocoups and democratic backsliding, democratic breakdown,
and the personalization of power. Additionally, comparative analyses between
autocoups and traditional coups could yield insights into the evolving nature of
power consolidation strategies in different political contexts.

In conclusion, this study enhances our understanding of autocoups by clarifying
terminology, refining definitions, and providing a comprehensive dataset. Future
research could explore the relationship between autocoups and democratic
backsliding, democratic breakdown, and the personalization of power.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: The dataset used in this article is publicly available and can be accessed
    at: <https://github.com/reddylee/Datasets>.
