---
title: "Autocoups: Determinants and Outcomes of Irregular Overtays of Political Leadership"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-14
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: autocoup.html

categories: [Article, Coups, Autocoups]
bibliography: references.bib

abstract: |
  This study aims to provide a comprehensive analysis of the factors influencing political leaders' decisions to either voluntarily step down or attempt to overstay in power. By employing a sample selection model, the research explores the effects of economic performance, the post-Cold War era, regime types, and levels of democracy on overstay attempts. Furthermore, it examines the influence of incumbency advantages on the likelihood of successful overstay attempts. Through a quantitative analysis based on a new dataset of overstays, this study contributes to the existing literature by offering valuable insights into the determinants of overstay attempts and their outcomes.
  
keywords: [Coups, Autocoups, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

## Introduction

When political leaders assume power, they can exit in three main ways: being
compelled to leave office before their term expires (ousted or forced to resign),
successfully completing their full term and stepping down[^1], or overstaying in
power through illegitimate means.

Unsurprisingly, irregular exits, particularly the first type, have garnered
significant academic attention. According to @goemans2009, out of 374 leaders who
exited irregularly, 246 were ousted through coups, accounting for 65.8% of these
cases. @frantz2016 shows that coup-related exits account for roughly one-third of
all exits in autocracies, surpassing any other type, including regular
transitions. Given the prevalence of coups as the primary form of irregular
exits, much research has focused on this phenomenon.

While previous literature has primarily concentrated on irregular entries and
exits, there has been comparatively less attention given to **irregular
overstays**, where leaders overstay their term despite legal requirements such as
term limits, losing elections, or impeachment. Therefore, it is worth studying
irregular overstays to analyse why some political leaders exit regularly
according to their term limits, while others attempt to overstay in office, and
why some leaders successfully overstay in power, while others fail in their
attempts. The significance of this topic is based on the following reasons:

Firstly, coups have become less common since 2000. Since 1990, there has been a
significant decline in cases of incumbent political leaders being ousted before
their legal term limits. The risk of coups, for example, has been remarkably
reduced [@bermeo2016; @thyne2019]. As shown in @tbl-coup_year_counts, the top 10
least frequent years of coups, except for 1950, are all after 1990, while the top
10 most frequent years of coups, except for 1991, are all before 1990.

However, irregular overstays have become more frequent since 2000. In July 2020,
Russia overwhelmingly passed a referendum amending the Russian Constitution,
allowing President Vladimir Putin, who has been in power since 1999, to remain in
office until 2036[^2]. Similarly, in March 2018, China's National People's
Congress (NPC) removed the two-term limit on the presidency from China's
Constitution, effectively enabling Xi Jinping to potentially remain in power
indefinitely[^3]. Even the United States, often considered a well-established
democracy, faced a potential risk of self-coup following the 2020 election
[@antonio2021; @pion-berlin2022]. Previous studies indicate that roughly 20 to 30
percent of political leaders have extended their terms through various
means[@ginsburg2010evasion; @baturo2014]. More recent research highlights an even
more concerning trend, showing that since 2000, approximately half of the rulers
in non-democratic countries have attempted to overstay in their positions
[@versteeg2020law]. Alarmingly, a considerable number of these overstay attempts
have been successful.

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups frequency by year (1950-2023)"
#| echo: !expr format == "html"

#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")


t1 <- coup |> 
  summarise(least_counts = n(),
            .by = year) |> 
  rename(least_frequency_years = year) |>
  arrange(least_counts) |>
  #slice(c(head(row_number(), 8), tail(row_number(), 7))) |> 
  head(10)
t2 <- coup |> 
  summarise(most_counts = n(),
            .by = year) |> 
  rename(most_frequency_years = year) |>
  arrange(-most_counts) |>
  #slice(c(head(row_number(), 8), tail(row_number(), 7))) |> 
  head(10)

cbind(t1, t2) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "center"
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***")) 
```

Secondly, while the negative outcomes of coups, such as societal turmoil and
democratic decline, have been widely discussed, the negative outcomes of
irregular overstays have not received enough attention, despite being at least as
significant. Although coups can sometimes end the worst regimes or chaos and have
been debated for their potential benefits to democracy [@derpanopoulos2016;
@miller2016; @derpanopoulos2017; @lumjiak2018], for example, the coup in Niger in
2010 resolved a constitutional crisis, ended turmoil, and promoted democracy
[@derpanopoulos2016].

Irregular overstays, however, are typically accompanied by the abuse of power,
defiance of legitimate authority, violation of the rule of law, destruction of
political conventions, and damage to checks and balances. Successful overstays
almost always lead to democratic backsliding or autocratic deterioration. Even
unsuccessful overstays often spark protests, strikes, riots, insurgencies,
rebellions, coups, or, even worse, civil wars. As overstays undermine formal
institutions or violate informal norms of power transition, they significantly
increase the risk of irregular power transitions in the future. Research
indicates that since 1945, approximately 62 percent of irregular stayed leaders
in non-democratic countries were either ousted or assassinated while in office
[@baturo2019].

Despite its high frequency and severe consequences, irregular overstays has
received comparatively less attention compared to the extensive research
conducted on the irregular exits of political leaders since the early 2000s
[@thyne2019]. The limited existing studies on irregular overstays have primarily
relied on case studies [@cameron1998; @antonio2021; @pion-berlin2022], and few
have been focused on quantitative analyses.

This study aims to address these issues and contribute in three areas. Firstly,
it clarifies and redefines the concept of autocoups in terms of irregular
overstays, establishing a new dataset of autocoups since 1945. Secondly, it
highlights the significant academic value of autocoups in the context of
irregular overstays, similar to the value of coups in irregular entry and exits.
Thirdly, it enhances the existing body of knowledge by conducting a quantitative
analysis on the determinants of autocoups.

The subsequent section will provide a comprehensive overview of the definition
and typology of the dataset on overstays. This will be followed by a presentation
of relevant theories and hypotheses that examine the factors influencing overstay
attempts and their outcomes. The research design will be outlined in Part IV,
detailing the data and methodology employed to test the hypotheses. The results
of the analysis will be discussed in Part V, providing a thorough examination of
the findings. Finally, the paper will conclude in Part VI, summarizing the key
insights gained from the study and suggesting avenues for future research.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: For lifelong rulers, such as those in monarchies or de facto hereditary
    regimes, leaders typically exit power only upon their death.

[^2]: Referendum In Russia Passes, Allowing Putin To Remain President Until 2036:
    <https://www.npr.org/2020/07/01/886440694/referendum-in-russia-passes-allowing-putin-to-remain-president-until-2036>,
    accessed on October 15, 2022.

[^3]: China's Xi allowed to remain 'president for life' as term limits removed:
    <https://www.bbc.co.uk/news/world-asia-china-43361276>, accessed on Octbor
    15, 2022.
