---
title: "Power Dynamics and Autocoup Attempts"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: zhu-qi@hotmail.com

date: last-modified
format:
  aps-pdf: 
    number-depth: 2

bibliography: references.bib

abstract: |
  This study explores the determinants of autocoup attempts, aiming to deepen understanding of the political dynamics that underpin tenure extensions by incumbent leaders. Addressing a notable gap in the existing literature, the study contends that the balance of power plays a critical role in shaping the likelihood of autocoup events. In contrast to classical coups—which are often triggered by unstable or fragmented power structures—autocoups tend to arise in contexts characterised by stable and concentrated power.
  To operationalise the concept of power balance in an observable manner, regime type is employed as a proxy, reflecting the structural distribution of power between incumbents and potential institutional constraints or elite challengers.
  Using a bias-reduced logistic regression model, the analysis finds that regime type is a significant predictor of autocoup attempts. Leaders operating within regimes marked by concentrated power are more prone to extend their tenure unconstitutionally. In particular, presidential democracies and personalist autocracies are found to be significantly more susceptible to autocoup attempts than dominant-party regimes.
  The study contributes to the broader literature on irregular leadership transitions by offering a more systematic and empirically grounded account of the conditions under which incumbents seek to subvert constitutional term limits.
  
keywords: [Autocoups, Coup, Regime types, Tenure Extension, Authoritarianism]

execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue, brglm2)
```

## Introduction

Why are some political leaders removed from office prematurely, while others
succeed in extending their tenure beyond constitutionally mandated limits? The
extant literature has extensively examined the former issue—namely, the forced
removal of leaders prior to the completion of their terms. Most such instances
are attributed to classical coups, defined as illegal and overt attempts by
military or state elites to depose a sitting executive [@powell2011], due to
their frequency and immediate impact on incumbent displacement. For example, in
autocratic regimes, classical coups account for nearly one-third of all
leadership exits, surpassing regular transitions, which comprise just over
one-fifth [@frantz2016]. Moreover, over $63\%$ of non-constitutional removals in
dictatorships are attributable to coups [@svolik2009].

By contrast, the latter part of the question—why some leaders are able to extend
their tenure beyond constitutional limits—remains comparatively underexplored.
This oversight stems primarily from conceptual ambiguity and the lack of
systematic data on such political events, specifically autocoups, which has
impeded rigorous empirical investigation. Nonetheless, the importance of studying
autocoups should not be underestimated. Firstly, autocoups represent one of the
most prevalent forms of irregular leadership transition, with over 80 documented
cases since 1945. Their frequency has notably increased since 2000, coinciding
with a marked global decline in classical coups [@bermeo2016; @thyne2019].
Secondly, autocoups exert profound consequences on political stability and
democratic development, frequently leading to long-term institutional erosion.
Thirdly, identifying the drivers of autocoup attempts is essential for
understanding their broader political impact; without a clear grasp of the
conditions under which autocoups occur, efforts to prevent or mitigate their
harmful effects remain severely constrained.

Although autocoups differ fundamentally from classical coups—in particular, in
being orchestrated by incumbents rather than external challengers—the two
phenomena share essential features as disruptions to the established political
order. Accordingly, the methodological tools commonly applied to the study of
traditional coups may be fruitfully adapted to the analysis of autocoups.
However, despite the extensive literature on coup dynamics [@gassebner2016],
regime type is frequently treated as a background condition or control variable,
rather than as a central explanatory factor.

This research contends that the likelihood of an autocoup attempt is
significantly influenced by the structural distribution of power embedded within
regime type. Unlike classical coups, which often arise in contexts marked by
contested or fragmented authority, autocoups tend to occur in regimes
characterised by concentrated and stabilised power. Given the difficulty of
directly measuring internal power dynamics, regime type is employed as a proxy
variable. The underlying assumption is that regime type reflects core
institutional arrangements—such as the distribution of authority, the strength of
constitutional constraints, and the incumbent’s capacity to subvert democratic
norms. Examining variation across regime types thus facilitates a more nuanced
understanding of the institutional foundations that condition the risk of
autocoups. These power structures are generally stable over time, as they both
shape and are shaped by the regime’s overarching institutional design
[@geddes2014].

Drawing upon a refined conceptualisation of autocoups and an original global
dataset encompassing autocoup events from 1945 to 2023, this study aims to make a
substantive contribution through a quantitative analysis of the determinants of
autocoup attempts.

To empirically assess this proposition, the study utilises both a standard
logistic regression model and a bias-reduced logistic regression model to
evaluate how regime type influences the probability of incumbents seeking to
extend their tenure through extra-constitutional means. Given the scarcity of
quantitative research on autocoups, this investigation offers a potentially
pioneering contribution to the literature by providing a theoretically grounded
and methodologically robust analysis of their underlying determinants.

The remainder of the paper is structured as follows: Section 2 introduces the
refined conceptualisation of autocoups and the original autocoup dataset; Section
3 explores the dynamics and outcomes of autocoup attempts; Section 4 outlines the
research design, including methodological approach and variable selection;
Section 5 presents and interprets the empirical findings, highlighting key
patterns and implications; and Section 6 concludes by summarising the main
insights and reflecting on their broader relevance for understanding and
mitigating the risks associated with autocoups.

## Autocoup Conceptualisation and Dataset

### Terminology

The academic discourse on autocoups utilises a diverse lexicon to describe the
extension of power or tenure by incumbent leaders. The most commonly employed
term is ‘self-coup’, or its Spanish counterpart ‘autogolpe’ [@przeworski2000;
@cameron1998a; @bermeo2016; @helmke2017; @marsteintredet2019]. This term rose to
prominence following the actions of Peruvian President Alberto Fujimori in 1992,
who dissolved Congress, suspended the constitution, and governed by decree
[@mauceri1995; @cameron1998]. However, as @marsteintredet2019 notes, ‘self-coup’
can be misleading, as it suggests the leader is acting against their own
interests, whereas such actions typically target other state institutions or
constitutional constraints.

A second category of terminology includes descriptors such as ‘presidential
coup’, ‘executive coup’, ‘constitutional coup’, ‘electoral coup’, ‘judicial
coup’, ‘slow-motion coup’, ‘soft coup’, and ‘parliamentary coup’
[@marsteintredet2019]. Whilst these terms may highlight specific mechanisms or
contexts, their proliferation often leads to conceptual ambiguity. Many focus on
the method of power acquisition but fail consistently to identify the
perpetrator. Furthermore, such mechanisms—judicial rulings, legislative
manoeuvres, or administrative decrees—may be employed either by or against
executive actors, thereby complicating classification further.

A third group of terms encompasses phrases such as ‘incumbent takeover’,
‘executive takeover’, and ‘overstay’. For example, ‘incumbent takeover’ is
defined as "an event perpetrated by a ruling executive that significantly reduces
the formal and/or informal constraints on his/her power" [@baturo2022, p. 374],
building on @svolik2014. Similarly, ‘overstay’ refers to "staying longer than the
maximum term as it stood when the candidate originally came into office"
[@ginsburg2011evasion, p. 1844]. These terms clearly delineate the actor (the
incumbent) and the action (power consolidation or term extension) but often fail
to convey the illegality or unconstitutionality of such actions. Unlike ‘coup’,
which inherently denotes an unlawful seizure of power, terms such as ‘takeover’
or ‘overstay’ may inadvertently downplay the normative severity of these events.

Given that many existing terms prioritise procedural mechanisms over normative
considerations or conflate legal and extra-legal practices, this study proposes
‘autocoup’ as the most precise and analytically robust term. It accurately
identifies both the actor and the act, clearly conveys the illegitimacy and
gravity of the behaviour, and establishes theoretical connections to conventional
coups. Consequently, ‘autocoup’ serves as the most effective term for capturing
and analysing this phenomenon, aligning seamlessly with the unified analytical
framework this study aims to establish.

### Definition

The conceptualisation of autocoups in political science literature is marked by
ambiguity, particularly around whether their defining feature is the expansion of
executive power, the extension of tenure, or a combination of both. Power
expansion refers to an incumbent accumulating authority beyond their original
constitutional remit—often through centralisation or weakening of institutional
constraints—while tenure extension involves efforts to remain in office beyond
the term originally prescribed, frequently by way of constitutional amendments or
electoral manipulation.

Many existing definitions conflate or overemphasise power expansion. For example,
@cameron1998a defines a self-coup as a suspension of the constitution and
dissolution of congress, resulting in rule by decree until new elections and a
referendum can establish a more empowered executive (p. 220). However, the phrase
“broader executive power” remains vague. Likewise, the notion of an “incumbent
takeover” as used by @baturo2022, drawing on @svolik2014, centres on the erosion
of formal or informal checks on executive power (p. 374), though the associated
dataset includes both power expansion and tenure extension. In contrast,
@ginsburg2011evasion offers a more precise definition of “overstay” as exceeding
the original term limit (p. 1844), focusing explicitly on tenure.

This study argues that tenure extension—rather than power expansion—constitutes
the core of an autocoup. Conceptually, this aligns autocoups with traditional
coups, which involve the disruption of executive tenure rather than a reduction
in power. Just as a coup entails premature removal from office, an autocoup
should be understood as the illegtimate prolongation of tenure. An executive who
loses power but remains in office does not qualify; likewise, a leader who
consolidates power while respecting term limits is better described in terms of
executive aggrandisement, not autocoup.

In practice, power expansion often serves as a strategic means to achieve tenure
extension. The widely cited case of President Alberto Fujimori in Peru
illustrates this point. His 1992 suspension of the constitution and dissolution
of Congress ultimately enabled him to remain in power: the 1993 Constitution
allowed a second term, and a controversial “authentic interpretation” by his
allies enabled a third run in 2000 [@ezrow2019]. These actions were not ends in
themselves but instrumental to prolonging his rule.

Finally, tenure extension offers a clearer empirical basis for classification
than the often ambiguous and difficult-to-measure concept of power expansion.
Definitions like @cameron1998a’s leave open questions about whether one or both
of certain actions (e.g., suspending the constitution or dissolving congress) are
necessary conditions for an autocoup.

In light of these considerations, this study defines an autocoup as ***the
extension of an incumbent leader’s tenure in office beyond the originally
mandated limit, achieved through extra-constitutional means***.

### Autocoup dataset

The autocoup dataset is constructed with reference to established datasets and
scholarly literature, thereby ensuring both reliability and comprehensiveness.
The principal sources used for case coding are listed in @tbl-source. The
Archigos dataset [@goemans2009] and the Political Leaders’ Affiliation Database
(PLAD) [@bomprezzi2024wedded] provide detailed records of national leaders from
1875 to 2023. Although the temporal scope of this study is restricted to events
occurring from 1945 onwards, these datasets remain essential for identifying de
facto leaders and distinguishing them from nominal heads of state.

```{r}
#| label: tbl-source
#| tbl-cap: Main Data Sources for Coding the Autocoup Dataset
#| echo: false
#| results: asis

read.csv(here("data/autocoup_source.csv")) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(100),
    table.align = "center"
  )
```

The Incumbent Takeover dataset [@baturo2022], which synthesises information from
eleven distinct sources, offers an extensive inventory of cases in which
executive actors significantly reduced institutional constraints on their
authority. As this dataset encompasses both instances of power consolidation and
tenure extension, cross-referencing with Archigos [@goemans2009] and PLAD
[@bomprezzi2024wedded] was undertaken to assess whether individual cases met the
definitional criteria for an autocoup.

In total, 83 events were identified and coded as autocoups. Of these, 50
correspond to entries found within the Incumbent Takeover dataset, while the
remaining 33 were newly identified and coded by the author through
cross-verification with additional sources, including Archigos [@goemans2009],
PLAD [@bomprezzi2024wedded], and contemporary news media.

The primary coding process yielded 83 instances of autocoups between 1945 and
2023, spanning 63 countries. This dataset provides a robust empirical foundation
for analysing patterns and trajectories in autocoup activity across a diverse
range of political and institutional contexts.

A breakdown of the methods employed by incumbents to extend their tenure is
provided in @tbl-autocoup_method. The most commonly observed strategy is the
legalisation or reintroduction of re-election provisions, accounting for 37
cases. This is followed by the removal of term limits (10 cases) and the
declaration of the incumbent as president for life (7 cases). Other tactics—such
as the cancellation of scheduled elections or refusal to concede electoral
defeat—are less frequently observed. Electoral manipulation is recorded in only
one case, primarily due to the inherent difficulty of verifying such practices
with certainty, despite strong circumstantial indications in many instances.

Importantly, and in contrast to classical coups—which occur predominantly within
autocratic regimes—a significant proportion of autocoups have taken place in
democratic contexts. Of the 83 identified cases, 30 occurred in democracies, of
which 29 were in presidential systems. This constitutes approximately $36\%$ of
the total. By contrast, traditional coups have been markedly rarer in democratic
settings, with only 99 out of 493 recorded cases ( $20\%$ ) occurring under
democratic rule.

```{r}
#| label: data-autocoups
#| echo: false

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
country <- Autocoup |>
  distinct(ccode) |>
  summarise(count = n())

autocoup_desc <- Autocoup |>
  summarise(
    attempted = n(),
    succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method
  ) |>
  arrange(-attempted)
```

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup Methods and Success Rates (1945-2023)

autocoup_desc |>
  adorn_totals() |>
  mutate(Success_rate = percent(round(succeeded / attempted, 3))) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(99)
  ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |>
  tab_source_note(source_note = md("*Source: Autocoup dataset*"))
```

## Dynamics of autocoup attempts

Like traditional coup attempts, autocoups are shaped by two fundamental elements:
the disposition of incumbent leaders—referring to their motivations and
willingness to act—and their capability, defined by the resources and
opportunities at their disposal. However, autocoups exhibit two notable features
that distinguish them from classical coups. First, whereas traditional coups
occur predominantly in autocracies [@thyne2014], over one-third of documented
autocoups have taken place in democratic regimes, as outlined in Chapter 2.
Second, while the success rate of traditional coups hovers around $50\%$, more
than $77\%$ of autocoup attempts have resulted in success, according to the
dataset introduced in Chapter 2. These distinctions indicate that the dynamics of
disposition and capability underlying autocoups differ significantly from those
of traditional coups.

This section explores the complex dynamics of autocoup attempts, with particular
emphasis on how the motivations of incumbents, the determinants of success, and
the institutional frameworks of various regime types shape the vulnerability of
states to such extra-constitutional power extensions.

### Motivations for autocoups

Incumbents seeking to prolong their tenure may be driven by a range of
motivations, broadly falling into three principal categories: personal ambition,
appeals to national interest, and self-preservation.

First, the pursuit of personal power constitutes a compelling incentive for many
leaders. The capacity to govern free from institutional constraints enables
incumbents to exercise dominance over national policy-making, access state
resources, influence the judiciary and legislature, and retain the prestige
associated with holding high office. For some, the aspiration to secure a lasting
political legacy—to be remembered as a transformative figure—further amplifies
the appeal of extended rule.

Second, tenure extensions are often justified by incumbents in the name of the
national interest. A commonly advanced rationale suggests that a single term is
insufficient for the completion of long-term reforms or development initiatives.
Within this narrative, remaining in power is portrayed as essential to ensuring
the continuity and success of ongoing projects. The autocoup is thus framed not
as an act of self-interest, but as a necessary step for the greater good.

Third, autocoups may serve as mechanisms of self-preservation. Incumbents facing
the prospect of prosecution for corruption, human rights violations, or other
transgressions may view continued tenure as a means of preserving legal immunity.
Additionally, those who have amassed significant political adversaries during
their rule may fear retribution upon leaving office. In such cases, the extension
of power is not merely a product of ambition but also a strategy for
survival—intended to shield the leader from legal or political repercussions.

### Power dynamics and autocoups

While motivations may initiate an incumbent’s decision to pursue an autocoup, the
decisive factor often lies in their ability to implement and sustain such an
action. The relatively high frequency and remarkable success rate of
autocoups—over $77\%$, compared to approximately $50\%$ for classical
coups—suggest that incumbents benefit from notable structural advantages when
attempting to consolidate power. These advantages are not limited to autocracies
but are also evident in democratic systems, underscoring the variation in
institutional leverage available to incumbents across different regime types.

This reality necessitates a closer examination of state power structures,
particularly the allocation of control over the military. The allegiance of the
armed forces is a critical determinant of autocoup outcomes. If the military
remains loyal to the executive, resistance—whether from civil society, the
judiciary, or the legislature—can be suppressed or marginalised. Conversely, open
defiance or refusal by the military to support the incumbent may render an
autocoup untenable.

Nevertheless, it is reductive to assume that formal authority as
commander-in-chief guarantees unqualified control. Just as it is overly
simplistic to attribute the success of traditional coups solely to the presence
of military force [@singh2016], it is equally erroneous to presume that
incumbents invariably enjoy the unconditional loyalty of the armed forces.
Nominal titles often obscure the complex and sometimes precarious dynamics
underpinning military allegiance.

In autocratic regimes, while the military may not be bound by constitutional
principles, it is not inherently loyal to the head of state. Executives depend on
military officers to execute their commands; however, these officers may harbour
independent political ambitions or competing loyalties. A case in point is Uganda
in 1971, when President Milton Obote attempted to dismiss General Idi Amin. In
response, Amin exploited his influence within the armed forces to mount a
successful coup, ousting Obote [@sudduth2017].

By contrast, in consolidated democracies, military loyalty is typically
institutionalised through allegiance to the constitution rather than to
individual officeholders. For example, in the United States, following the 2020
presidential election, General Mark Milley, Chairman of the Joint Chiefs of
Staff, publicly reaffirmed the military’s constitutional commitment: “We are
unique among militaries. We do not take an oath to a king or a queen, a tyrant or
a dictator. We do not take an oath to an individual. We take an oath to the
Constitution.” (US Army Museum, 12 November 2020\[\^1\])

In hybrid regimes or fragile democracies, attempts to prolong executive tenure
may entail significant political risks. In Niger, for example, President Mamadou
Tandja’s attempt in 2009 to amend the constitution to permit a third term
precipitated a military coup in 2010 [@miller2016]. Similarly, in Honduras the
same year, President Manuel Zelaya was removed from office by the military after
seeking to alter the constitution to allow immediate re-election
[@muñoz-portillo2019].

### Regime types and autocoups

Given the complexities discussed, a more effective analytical strategy entails
evaluating the broader balance of power within political systems. As direct
observation of this balance is inherently challenging, this study adopts regime
type as a proxy—an approach consistent with established methodologies in
comparative politics. Regime types encapsulate the institutional architecture of
power distribution, particularly with respect to control over the military,
political appointments, and policy-making authority.

Following the typology developed by @geddes2014, autocratic regimes can be
categorised as follows:

**Military regimes** are governed by a junta, typically comprising senior
military officers who collectively determine leadership and policy direction.
Notable examples include Brazil (1964–1985), Argentina (1976–1983), and El
Salvador (1948–1984) [@geddes1999].

**Personalist regimes** revolve around a dominant individual who wields unchecked
authority over the military, policy decisions, and succession processes.
Prominent instances include Rafael Trujillo in the Dominican Republic
(1930–1961), Idi Amin in Uganda (1971–1979), and Jean-Bédel Bokassa in the
Central African Republic (1966–1979) [@geddes1999].

**Dominant-party regimes** concentrate authority within a structured political
party, with the leader operating either as part of or at the helm of the party
apparatus. Illustrative cases include the PRI in Mexico, CCM in Tanzania, and the
Leninist parties of Eastern Europe [@geddes1999].

Among these regime types, personalist autocracies are particularly conducive to
autocoups. The concentration of power in a single individual weakens
institutional checks and fosters loyalty—particularly from the military—through
mechanisms of personal patronage. While military regimes are rooted in coercive
power, they are often beset by internal factionalism, rendering them more
susceptible to traditional coups than to autocoups. Dominant-party regimes occupy
a more ambiguous position: although party structures can constrain executive
action, exceptionally powerful party leaders may still initiate autocoups, as
exemplified by Xi Jinping’s constitutional amendments in 2018 within a
dominant-party framework.

Monarchies, though technically autocratic, generally render autocoups redundant,
as monarchs typically rule for life by constitutional design.

A key clarification is warranted at this juncture: why might leaders in
personalist regimes—already possessing extensive authority—feel compelled to
extend their tenure further? The answer lies in distinguishing between the scope
and duration of power. While such leaders may exercise considerable de facto
control over state institutions, many initially assume office via legal or
constitutional channels, necessitating a gradual process of consolidation. In
this context, autocoups function as formal mechanisms to institutionalise
existing dominance—transforming informal power into legally sanctioned
permanence. This dynamic is exemplified by the repeated tenure extensions pursued
by Vladimir Putin and Alexander Lukashenko.

In post-Soviet Russia, President Boris Yeltsin presided over the transformation
of a parliamentary system into a personalist regime. However, Yeltsin himself did
not overstay his term; instead, he designated Vladimir Putin as his successor.
Upon assuming office in 2000, Putin progressively entrenched his authority,
employing constitutional amendments and legal strategies to circumvent term
limits and extend his rule indefinitely.

Likewise, in Belarus, Alexander Lukashenko was elected president in 1994 under a
party-based system. Within a year, he dismantled the existing institutional
framework and established a personalist regime. Since then, he has remained in
power through successive tenure extensions, steadily consolidating his control
over the state apparatus.

In democratic contexts, autocoups are found exclusively in presidential systems.
This reflects the institutional leverage enjoyed by presidents, who are directly
elected, typically command the armed forces, and may possess the capacity to
override or circumvent legislative opposition. By contrast, prime ministers in
parliamentary systems are considerably more constrained. Their tenure depends on
maintaining legislative confidence and they may be removed through votes of no
confidence. Moreover, they often lack direct control over the military, which is
institutionally separated from their office. As a result, prime ministers are
subject to more frequent leadership turnover and face fewer opportunities to
unilaterally extend their mandates. For instance, the United Kingdom saw three
prime ministers serve in 2022 alone, while Japan has had 36 prime ministers since
1945—an average of one every two years. In contrast, only 14 presidents have
served in the United States over the same period, reflecting greater
institutional continuity. These structural distinctions render presidential
systems more conducive to autocoups—even within well-established democracies—due
to their centralised executive authority and command over the military.

From this analysis, the following hypothesis is proposed:

The likelihood of autocoup attempts is significantly shaped by regime type, with
regimes characterised by concentrated and stable executive power—namely,
personalist autocracies and presidential democracies—being the most susceptible,
relative to other regime types.

***H3-1: The likelihood of autocoup attempts is significantly shaped by regime
type, with regimes characterised by concentrated and stable executive
power—namely, personalist autocracies and presidential democracies—being the most
susceptible, relative to other regime types.***

## Research design

### Methodology

Given the binary nature of the dependent variable—namely, whether an autocoup is
attempted in a given country-year—the study initially employs a logistic
regression model to investigate the determinants of autocoup attempts. This
method enables the identification of statistically significant factors
influencing the likelihood of such events, as well as the direction and magnitude
of their effects.

Nevertheless, the rarity of autocoup incidents—83 cases out of over 9,000
observations—poses a methodological challenge. Standard maximum likelihood
estimation techniques, including conventional logit and probit models, are prone
to underestimating the probability of rare events. To mitigate this limitation
and improve the robustness of statistical inference, the analysis also employs
Firth’s Bias-Reduced Penalised Maximum Likelihood Estimation (commonly referred
to as Bias-Reduced Logit), as outlined by @firth1993.

### Data and variables

The primary dataset, which incorporates information on autocoups and regime
types, spans the period from 1945 to 2023. However, due to data alignment
limitations, the usable data range extends from 1945 to 2018. The dataset
comprises approximately 9,400 country-year observations, of which 83 represent
recorded autocoup attempts.

#### Dependent variable

The analysis draws upon the autocoup dataset introduced in Chapter 2, which
covers the period from 1945 to 2023 and includes 83 documented autocoup attempts.
Summary statistics for these events, as well as the corresponding regime
classifications, are presented in Chapter 2.

**Autocoup attempt**: A binary variable indicating whether an autocoup attempt
occurred (coded as 1) or did not occur (coded as 0) in each country-year
observation.

#### Independent variables

The principal independent variable in this analysis is regime type, reflecting
the central analytical focus of the study. Regime classifications are drawn from
the typology developed by @geddes2014 (GWF dataset), which distinguishes among
military, personalist, and dominant-party regimes within autocratic systems. For
democratic systems, regimes are categorised as either parliamentary or
presidential. A residual category—labelled “other”—captures regimes that are
provisional, transitional, or otherwise not easily classified within the primary
typology.

In addition to regime type, a range of control variables is included, selected on
the basis of established scholarship on the determinants of coups. These controls
account for factors such as economic performance, political violence, and the
tenure of incumbents. Further controls comprise the level of democracy,
population size, and a Cold War dummy variable, which captures temporal variation
in the global political environment.

**Economic Level:** Measured by GDP per capita, this variable reflects the
overall economic wellbeing of a country. Data are sourced from the V-Dem dataset
[@fariss2022] and are expressed in constant 2017 international dollars (PPP, per
thousand).

**Economic Performance:** Operationalised via the Current-Trend (CT) ratio
developed by @krishnarajan2019, this measure compares current GDP per capita with
the average of the previous five years. Higher CT values indicate stronger
economic growth. Formally:

$$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
$$

**Political violence:** Measured using a violence index based on the “actotal”
variable from the Major Episodes of Political Violence dataset
[@marshall2005current], this index captures both internal and interstate
conflict. Scores range from 0 (complete stability) to 18 (maximum instability).

**Days in office (log):** The natural logarithm of an incumbent leader’s
cumulative days in office is included as a proxy for power consolidation. Longer
tenures are hypothesised to facilitate the conditions necessary for an autocoup.
Data are drawn from the Archigos dataset [@goemans2009] and the Political
Leaders’ Affiliation Database (PLAD) [@bomprezzi2024wedded].

**Democratic level:** This variable employs the Polity V score to measure the
degree of democracy in a country, ranging from -10 (fully autocratic) to +10
(fully democratic). The index, developed by the Centre for Systemic Peace,
assesses regime characteristics such as the competitiveness of political
participation, executive recruitment, and constraints on executive authority
[@marshall2005current].

**Population size:** The natural logarithm of a country’s population is included
to account for the potential effects of demographic scale on governance. Larger
populations may present more complex administrative challenges and generate
greater opposition. Data are sourced from the V-Dem dataset.

**Cold War:** Following the precedent of earlier studies [@thyne2014;
@derpanopoulos2016; @dahl2023], a dummy variable is included to distinguish the
Cold War period (approximately 1960–1990) from the post-Cold War era. This
distinction reflects the relative paucity of autocoup events during the Cold War
and their increased frequency thereafter.

## Results and discussion

This chapter employs logistic regression techniques to examine the structural and
contextual factors that influence the probability of autocoup attempts. Given the
dichotomous nature of the dependent variable—whether or not an autocoup attempt
occurred in a specific country-year—and the rarity of such events (78 out of
9,434 observations) in certain categories, the analysis includes both a standard
logit model and a bias-reduced logit model. The latter is particularly
well-suited for rare events data, as it corrects for the small-sample bias often
encountered with conventional maximum likelihood estimation. Consequently, the
interpretation of results prioritises estimates from the bias-reduced model. Odds
Ratios (ORs) are reported to facilitate an intuitive understanding of effect
sizes.

@tbl-autocoupmodel presents the model estimates. The core hypothesis posits that
regime type is a key predictor of autocoup incidence, particularly that
personalist regimes and presidential democracies are significantly more prone to
such events than dominant-party regimes, which serve as the reference category.

```{r}
#| label: tbl-autocoupmodel
#| tbl-cap: "Determinants of Autocoup Attempts(1945-2018)"
#| results: asis
#| echo: false

autocoup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_model.csv") |>
  filter(days_in_office > 180) |>
  mutate(
    regime = case_when(
      regime == "personal" ~ "Personal",
      regime == "dominant-party" ~ "Dominant Party",
      regime == "presidential" ~ "Presidential",
      regime == "military" ~ "Military",
      regime == "parliamentary" ~ "Parliamentary",
      TRUE ~ "Other"
    ),
    regime = factor(regime,
      levels = c("Dominant Party", "Personal", "Presidential", "Military", "Parliamentary", "Other")
    ),
  )

common_labels <- list(
  "(Intercept)" ~ "Constant",
  regime ~ "Regime Type",
  GDP_pc ~ "GDP per capita",
  GDP_trend ~ "GDP growth trend",
  violence ~ "Political violence",
  pop_log ~ "Log of Population",
  polity5 ~ "Polity V scores",
  "log(cumulative_days_in_office)" ~ "Log of days in office",
  "factor(cold_war)" ~ "Cold war"
)

model_formula <- autocoup_attempt ~ regime + GDP_pc + GDP_trend + violence + pop_log +
  polity5 + log(cumulative_days_in_office) + factor(cold_war)

model_logit <- autocoup_model |>
  glm(
    model_formula,
    family = binomial,
    data = _
  ) |>
  tbl_regression(
    label = common_labels,
    show_single_row = "factor(cold_war)",
    conf.int = T,
    intercept = T
  ) |>
  add_n(location = "level") |>
  add_nevent(location = "level") |>
  add_significance_stars(
    hide_p = TRUE,
    thresholds = c(0.01, 0.05, 0.1)
  ) |>
  modify_column_hide(columns = std.error)


model_logit_od <- glm(
  model_formula,
  family = binomial,
  data = autocoup_model
) |>
  tbl_regression(
    exponentiate = TRUE,
    label = common_labels,
    show_single_row = "factor(cold_war)",
    conf.int = T,
    intercept = T,
    digits = 1
  ) |>
  add_significance_stars(
    hide_p = TRUE,
    thresholds = c(0.01, 0.05, 0.1)
  ) |> 
  modify_fmt_fun(
    std.error = function(x) style_number(x, digits = 2)
  )


model_brg <- glm(
  model_formula,
  family = binomial(link = "logit"),
  method = "brglmFit",
  data = autocoup_model
) |>
  tbl_regression(
    label = common_labels,
    show_single_row = "factor(cold_war)",
    conf.int = F,
    intercept = T
  ) |>
  add_significance_stars(
    hide_p = TRUE,
    thresholds = c(0.01, 0.05, 0.1)
  ) |>
  modify_column_hide(columns = std.error)

model_brg_od <- glm(
  model_formula,
  family = binomial(link = "logit"),
  method = "brglmFit",
  data = autocoup_model
) |>
  tbl_regression(
    exponentiate = TRUE,
    label = common_labels,
    show_single_row = "factor(cold_war)",
    conf.int = T,
    intercept = T
  ) |>
  add_significance_stars(
    hide_p = TRUE,
    thresholds = c(0.01, 0.05, 0.1)
  ) |> 
  modify_fmt_fun(
    std.error = function(x) style_number(x, digits = 2)
  )

tbl_merge(
  tbls = list(model_logit, model_logit_od, model_brg, model_brg_od),
  tab_spanner = c("Standard Logit", "Standard Logit", "Bias-reduced Logit", "Bias-reduced Logit")
) |>
  as_gt() |>
  tab_options(
    table.font.size = px(13),
    data_row.padding = px(8)
  )
```

The empirical results provide robust support for this hypothesis. In the
bias-reduced model, the odds of an autocoup occurring in a personalist regime are
more than twice as high as in a dominant-party regime (OR = 2.08, p \< 0.05). In
presidential democracies, the odds are nearly five times greater (OR = 4.87, p \<
0.01). To illustrate the magnitude of these effects, I compute predicted
probabilities for a prototypical country in the dataset, holding all other
covariates at their mean values. In dominant-party regimes, the predicted
probability of an autocoup in a given year is approximately $0.99\%$. In
personalist regimes, the probability rises to around $2.0\%$. In presidential
democracies, the likelihood increases further to approximately $4.7\%$. While
these probabilities are low in absolute terms—reflecting the rarity of autocoup
events—the relative differences are substantial. Leaders in presidential
democracies, for instance, are nearly five times more likely to attempt an
autocoup than those in dominant-party systems, holding other factors constant.
This underscores the structural vulnerability of executive-centric political
systems, particularly when institutional checks on executive power are weak.

Among other regime types, military and parliamentary democracies do not show
statistically significant differences in autocoup likelihood relative to
dominant-party regimes. The residual "other" category does reach marginal
significance (OR = 0.34, p \< 0.1), suggesting lower odds, although the
heterogeneity within this group warrants cautious interpretation.

Turning to the control variables, several findings warrant closer attention. The
Polity V score, which proxies the level of democratic institutionalisation, is
significantly associated with reduced odds of autocoup (OR = 0.91, p \< 0.01).
Substantively, this indicates that for each one-point increase in Polity score,
the odds of an autocoup decrease by approximately $9\%$, holding other factors
constant. This underscores the protective role of democratic institutions against
executive overreach.

The Cold War indicator also emerges as significant (OR = 0.45, p \< 0.01),
suggesting that autocoups were $55\%$ less likely during the Cold War era than in
the post-Cold War period. This aligns with historical interpretations that view
the Cold War as imposing external constraints on authoritarian innovation, often
via superpower influence.

The analysis reveals a statistically significant, albeit marginal, negative
relationship between the log of population size and the incidence of autocoups
(OR = 0.86, p \< 0.1). This suggests that as a country's population grows, the
odds of an autocoup tend to decrease. This finding aligns with theoretical
arguments positing that larger, more populous states may exhibit greater
organizational complexity and higher visibility of executive power dynamics,
thereby increasing the difficulty and scrutiny associated with an executive power
grab.

Conversely, indicators of economic performance—including GDP per capita, GDP
growth, and political violence—do not exhibit statistically significant
relationships with the outcome variable. Likewise, the log of days in office for
the incumbent does not significantly predict autocoup attempts, suggesting that
tenure alone is not a sufficient condition for extra-constitutional moves.

In sum, the analysis confirms that regime type—particularly personalist and
presidential systems—is a critical structural condition influencing the
likelihood of autocoup attempts. The inclusion of predicted probabilities and
percentage changes in odds ratios serves to clarify the substantive significance
of these patterns, beyond their statistical robustness. These results point to
the institutional fragility of regimes where executive authority is highly
centralised and unchecked. Additionally, the protective effects of democratic
institutions and Cold War-era international structures warrant greater attention
in discussion of executive stability and regime resilience. The implications of
these findings for policy and democratic governance will be explored in detail in
the final chapter.

## Summary

This chapter provides a quantitative analysis of the determinants of autocoup
attempts, addressing a notable gap in the existing literature, which has often
been hindered by conceptual imprecision and the absence of systematic empirical
data. It advances the central argument that the likelihood of autocoup attempts
is shaped significantly by the structural configuration of political power within
regimes, operationalised through regime type. Employing both standard logistic
regression and Firth’s bias-reduced logit model, the analysis demonstrates that
personalist autocracies and presidential democracies are markedly more
susceptible to autocoup attempts than dominant-party regimes. Specifically, the
odds of an autocoup are estimated to be approximately three times higher in
personalist autocracies and nearly five times higher in presidential democracies
relative to the baseline category.

These findings lend empirical support to the hypothesis that such regime types
possess structural vulnerabilities that facilitate extra-constitutional power
consolidation by incumbents. In addition to regime type, the analysis identifies
several other statistically significant covariates: population size, the degree
of democratic institutionalisation, and the broader historical context of the
Cold War all exert discernible effects on the probability of autocoup occurrence.
By examining the strategic incentives confronting incumbent leaders across
diverse institutional contexts, the study contributes to a more nuanced
understanding of irregular leadership transitions.

Nevertheless, the analysis also underscores several conceptual and methodological
challenges that warrant further investigation. Unlike traditional coups—which may
occur at various stages of a regime’s lifespan and are often subject to
recurrence—autocoups appear to follow distinct temporal patterns. For instance,
their likelihood may be comparatively low during the initial stages of a leader’s
tenure, increasing as the conclusion of a constitutional term approaches.
Moreover, while a successful extension of tenure may reduce the short-term risk
of subsequent attempts, empirical cases such as those of Presidents Putin and
Lukashenko suggest that incumbents may engage in serial autocoup behaviour.

To render the analysis tractable, this study adopts the simplifying assumption
that an autocoup attempt occurs only once per leadership tenure. While
analytically expedient, this assumption highlights the need for future research
to explore the temporal dynamics and sequencing of autocoup activity. Such
inquiries would usefully complement the present findings by offering deeper
insights into the long-term patterns of institutional adaptation, authoritarian
durability, and democratic erosion.
