---
title: "Irregular Leadership Transitions and Political Stability: A Survival Analysis of Coups and Autocoups"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: survival1.html
categories: [Article, Coups]
bibliography: references.bib

abstract: |
  This study explores and compares the survival time of two categories of irregular-entry political leaders: those who ascend to power through coups and those who overstay their legitimate term limits (autocoups). The central hypothesis posits that the mode of accession to power influences the duration of their tenures. Utilizing survival analysis, the results indicate that leaders who overstay their terms through autocoups generally remain in power longer than those who come to power through coups.

keywords: [Coups, Autocoups, Survival, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(survival)
library(survminer)
library(ggsurvfit)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(stargazer)
library(here)
```

## Introduction

The survival of political leaders has been a well-studied topic in political
science. However, two specific types of leaders - those who come to power through
coups and those who overstay their legitimate terms (autocoup leaders) - have
received comparatively less attention. Examining the tenures of these irregularly
ascended leaders holds particular significance for two main reasons.

Firstly, leaders who ascend through irregular means constitute the majority of
irregular exits from power. According to @goemans2009, between 1945 and 2015, out
of 1,472 leaders who assumed office through regular channels, approximately 213
exited irregularly (14.5%). Conversely, out of 308 leaders who assumed office
through irregular means, roughly 158 (51.3%) experienced irregular exits.

Secondly, among irregularly ascended leaders, the majority gained power through
launching coups or overstaying in office. As per @goemans2009, out of 374 leaders
who exited irregularly, 246 were ousted through coups, constituting 65.8% of
these cases. Additionally, between 1945 and 2020, there were 106 attempts to
overstay in power, of which 86 were successful [@Zhu2024]. Leaders who extend
their legitimate term limits essentially stage coups against legitimate
successors, thereby qualifying as autocoup leaders. Hence, it is both relevant
and insightful to delve into and compare the tenures of coup-entry leaders and
autocoup leaders.

While it is expected that autocoup leaders survive longer than coup-entry leaders
when considering their total tenures, a more insightful comparison arises when
examining the tenures following autocoups in contrast to those of coup-entry
leaders. Conducting a log-rank test in survival analysis on the leaders dataset
[@goemans2009] and the author's autocoup dataset @zhu2024\] reveals a distinct
contrast between autocoup leaders and coup-entry leaders.

```{r}
#| label: data
#| echo: !expr format == "html"
#| include: !expr format == "html"

Mydata1 <- read.csv(here("data/mydata1.csv")) |> 
  mutate(survival_year = survival_time / 365)

avg <- Mydata1 |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)

avg
```

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders
#| echo: !expr format == "html"

# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = Mydata1)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = Mydata1, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("entry by autocoup", "entry by coup"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 14)
```

Preliminary findings from survival analysis (see @fig-logrank) indicate a
significant difference. Leaders who consolidate power through autocoups appear to
have a longer average tenure after the autocoup itself (approximately
`r avg$avg[1]`) compared to coup leaders (approximately `r avg$avg[2]`). This
suggests a potential shortfall of 5 years in the average tenure of coup leaders.

Thus, we hypothesize that the method of accession significantly influences
leadership longevity. Coup leaders likely face greater challenges to their
legitimacy, leading to a more precarious initial period in power. They may
encounter higher levels of uncertainty, instability, and pressure to share power,
all of which could contribute to a shorter average tenure compared to autocoup
leaders.

By employing the Cox proportional hazards model and the extended Cox model, the
results suggest that leaders who exceed their term limits generally experience
longer tenures compared to those who come to power through coups.

This study offers two key contributions. First, it highlights a critical factor
often overlooked in leadership survival studies: the impact of accession method.
Leaders' survival is not solely determined by their actions after taking power,
but also by the way they acquired power in the first place. This research
underscores the significant difference in tenure duration between autocoup and
coup leaders. Second, it provides empirical measurements to compare the tenure
duration of these two irregularly ascended leaders, offering insights into their
distinct impacts on leadership longevity.

The remainder of the paper delves deeper into this topic. Chapter 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Chapter 3 explores the factors influencing the survival of
coup and autocoup leaders. Chapter 4 outlines the methodology and data used,
including the application of survival models to analyze the determinants of
leadership longevity. Chapter 5 presents the findings of the analysis and a
detailed discussion of the results. Finally, Chapter 6 concludes by synthesizing
the key takeaways and exploring their broader implications.

```{r}
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
