---
title: "Leadership Transitions and Survival:\n Coups, Autocoups, and Power Dynamics"

author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk
  
format:
  aps-html: default
  aps-pdf: 
    documentclass: report
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    toc: true
    toc-title: Contents
    lof: true
    lot: true
    crossref: 
      lof-title: Figures
      lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  docx: default
    
categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
#include-before-body: title.tex
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(data.table)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(readxl)
library(here)
library(stargazer)
library(survival)
library(sampleSelection)
library(marginaleffects)
```

## Acknowledgements {.unnumbered}

The completion of this thesis has been a significant journey, filled with hard
work, learning, and moments of joy. Throughout this time, I have received support
and encouragement from many individuals, without whom this dissertation would not
have been possible.

First and foremost, I would like to express my deepest gratitude to my great
supervisor, Professor Kristian Skrede Gleditsch, for his invaluable guidance,
unwavering support, and insightful feedback throughout this journey. His
expertise and encouragement have been instrumental in shaping this dissertation.
I would also like to extend my heartfelt thanks to the chair of my board panel,
Professor Han Dorussen, for his continuous support and constructive criticism,
which have significantly enhanced the quality of my research.

I am profoundly grateful for the comments, advice, and suggestions from several
esteemed scholars who have contributed to this work. Dr. Brian J Phillips, Dr.
Prabin Khadka, and Dr. Winnie Xia, their expertise and thoughtful input have been
greatly appreciated and have enriched this dissertation.

Finally, I want to thank my family for their unwavering support and love. To my
beloved wife, Ji Zhi, who has been my rock throughout this journey, and to my
dear daughter, Siyan, and son, Sisheng, who have been my source of joy and
motivation. I am deeply thankful to my father for his enduring support, and to
the memory of my late mother, whose love and guidance continue to inspire me.

All errors and faults are my own.

## Abstract {.unnumbered}

This dissertation examines the dynamics of irregular power transitions,
particularly coups and autocoups, and their influence on leader survival. It
highlights the critical role of power dynamics, shaped by **regime type**, in
determining coup success rates and attempt frequency. Utilizing a **double probit
model with sample selection**, the study reveals that expected coup success
significantly influences attempts, with military regimes facing a heightened
vulnerability due to their power structure.

While often understudied, autocoups are shown to have a substantial impact on
democratic trends. This research introduces a refined definition of autocoups
alongside a novel dataset encompassing events from 1945 to 2022, enabling a more
robust quantitative analysis.

Employing survival analysis, the study compares the longevity of leaders who rise
to power through coups versus autocoups. The findings demonstrate that
coup-installed leaders face a significantly shorter tenure and higher risk of
removal. This contrasts with autocoup leaders who manipulate the system to extend
their rule, suggesting the potential for autocoups to incentivize power grabs and
contribute to democratic backsliding.

This work contributes significantly to the political science literature by:

-   Defining key concepts: It establishes a clear definition of autocoups, a
    previously understudied phenomenon.

-   Introducing a novel dataset: This dataset enables researchers to conduct more
    comprehensive quantitative analyses of autocoups.

-   Establishing a general framework: The framework provides a comparative
    approach to studying the dynamics of irregular power transitions and their
    impact on democratic stability.

**keywords:** *Coups, Autocoups, Power transitions, Leadership Survival*

## Introduction

### Research question

```{r}
#| label: leader_data
#| include: false

download.file("https://raw.githubusercontent.com/reddylee/Datasets/main/PLAD.csv","data/Plad.csv")
Plad <- fread(here("data/Plad.csv"))
download.file("https://raw.githubusercontent.com/reddylee/Datasets/main/archigos.csv", "data/Archigos.csv")
Archigos <- fread(here("data/Archigos.csv"))

Plad |>
  filter(startdate > 1945) |> 
  tabyl(exit)
Archigos |>
  filter(startdate > 1945,
         !exitcode  %in% c("Regular", "Still in Office")) |> 
  tabyl(exitcode) |> 
  adorn_totals()
```

Irregular power transitions, marked by a disregard for constitutional procedures,
are a critical area of study in political science. They not only disrupt
established rules but often require unconstitutional tactics to secure power.
Furthermore, these transitions can inspire copycat behaviour among other
ambitious leaders.

Despite their central role in political science and the extensive research
conducted on irregular power transitions, a long-standing question continues to
intrigue political scientists: ***Why are some leaders ousted before their terms
expire, while others complete their full terms or even overstay beyond their
originally mandated limits?*** In other words, why do some leaders survive for
decades while others last for only years, months, or even days? This dissertation
focuses on this question and seeks to provide a comprehensive analysis, dedicated
to understanding how leaders come to power through unconstitutional means and
what factors determine the duration of a leader's rule following an irregular
ascent.

### Analyses on coups and autocoups in a general framework

When discussing irregular power transitions, the concepts that often come to mind
are irregular entries or exits, such as coups, assassinations, rebellions,
protests, and foreign interventions. Among these methods, coups hold a prominent
position due to their frequent occurrence. According to the Archigos dataset
[@goemans2009], from 1945 to 2015, there were approximately 145 instances of
irregular leader exits, with coups[^1] accounting for more than half (79
leaders). The often-cited Global Instances of Coups (GIC)[^2] dataset
[@powell2011] records even more leaders (245 cases) removed by coups from 1950 to
2023.

Given their prevalence and substantial influence on political systems, coups have
been extensively studied, particularly since 2000 [@thyne2019]. Consequently, the
concept of a coup is comparatively clear and widely accepted in academic circles.
Many scholars, including this study, follow the definition by @powell2011, which
describes coups as "illegal and overt attempts by the military or other elites
within the state apparatus to unseat the sitting executive... \[a coup is
successful\] if the perpetrators seize and hold power for at least seven days"
(P. 252). Although debates persist, two elements are clear: first, the
perpetrators are elites within the ruling group, and the victims of coups are
incumbent executive leaders. Second, the strategy or aim of a coup involves
completely removing the incumbents, not merely seizing part of their power or
forcing them to concede on specific policies. Beyond defining coups, several
datasets have been developed for quantitative analyses, such as the Global
Instances of Coups [@powell2011], the Cline Centre Coup d’État Project Dataset
[@peyton2024], and the Colpus Dataset [@chin2021]. These datasets are
well-developed and frequently used in political science research.

However, irregular power transitions are not limited to irregular entries and
exits but should also include irregular "overstays." Using illegal means to
overthrow an incumbent leader before their term expires is undoubtedly an
irregular power transition. Similarly, an incumbent using illegitimate means to
extend their term beyond term limits is also an irregular power transition.

Although academic attention to irregular retention of power has increased since
the 1990s, especially after Peru’s President Alberto Fujimori's self-coup in
1992, it remains comparatively understudied and has several shortcomings. First,
there is no universally accepted terminology for this "overstaying in power" type
of irregular power transition, unlike the clear term "coup." Consequently,
various terms such as self-coup, autogolpe, and executive coup are used by
different scholars. This dissertation will use 'autocoup' to refer to this type
of irregular power transition, which will be thoroughly discussed in Chapter 3.
Second, there is no consensus on the definition of an autocoup. Existing
definitions remain vague, often conflating power expansions and power
extensions[^3]. For example, @cameron1998a defines an autogolpe as a temporary
suspension of the constitution and dissolution of Congress by the executive, who
then rules by decree. This definition focuses on power expansion instead of power
extension, leading to conceptual confusion and misalignment with the definition
of a classic coup. Third, a consensus autocoup dataset is lacking. While several
related datasets exist, as discussed by @baturo2022 in coding their Incumbent
Takeover dataset, the terminologies, definitions, and coverage years vary,
lacking wide acknowledgement and extensive academic exploration. In summary,
autocoup has not been analysed in a comparative manner connected with coups.

Analysing coups and autocoups separately is less problematic. However, from a
comprehensive framework perspective on irregular power transitions and leader
survival, coups and autocoups should be, and can be, analysed within the same
framework. Both coup and autocoup significantly influence democratic backsliding
and are the most frequent means of irregular power transition. Furthermore, as
both are called "coups," classic coups and autocoups are very similar since a
coup is launched to replace the current leader, while an autocoup is staged to
replace the future leader.

### Academic Contributions

This study addresses a critical gap in the literature by offering a comprehensive
framework for analysing both coups and autocoups, which are the most common forms
of irregular power transitions. While existing research often examines these
topics separately with varying terminologies, definitions, methods, and datasets,
this dissertation integrates these elements to provide a unified perspective on
irregular power transitions and leader survival.

Our contributions are threefold:

-   **Emphasis on power dynamics and regime types**: We highlight the significant
    role of power dynamics, particularly the influence of regime types, in
    determining the success and frequency of coup attempts. Our analysis
    underscores how the expected chances of coup success motivate such attempts,
    with military regimes being notably susceptible.

-   **Refined definition and novel dataset for autocoups:** We introduce a
    refined definition of autocoups and develop a novel dataset covering events
    from 1945 to 2022. This enables a comparative analysis with classic coups,
    providing clearer insights into the nature and impact of autocoups on
    political systems.

-   **Survival analysis of leaders from different entry modes:** By applying
    survival analysis to existing coup data and our new autocoup dataset, we
    demonstrate how different modes of entry into power significantly affect
    leader survival. Our findings reveal that leaders who come to power through
    coups typically have shorter tenures and face higher removal risks compared
    to those who extend their rule through autocoups.

Our analysis of irregular power transitions is particularly relevant to
understanding democratic backsliding. These transitions violate democratic norms
and disrupt the path towards stable democracy. Leaders who gain power through
irregular means often employ undemocratic tactics, such as suppressing
opposition, to consolidate their illegitimate hold on power. This creates a
vicious cycle where the erosion of democratic institutions is both a cause and
consequence of efforts to maintain power.

### Overview of the thesis

This study is structured into three main chapters beyond the introduction, each
addressing key aspects of irregular power transitions and their implications for
political stability and democratic processes.

**Chapter 2** examines the determinants of classic coup attempts. While extensive
research exists on coups, most studies focus on observable factors before coups,
such as economic performance, political stability, previous coups, and
coup-proofing strategies. This chapter, however, emphasizes the less observable
but crucial factor of expected chances of coup success, which have been often
overlooked. Utilizing the double probit model with sample selection, the analysis
reveals that expected success rates significantly influence coup attempts. These
success rates are primarily shaped by the balance of power between incumbents and
challengers, which is largely determined by regime type. The findings indicate
that military regimes face a much higher risk of coups compared to dominant-party
regimes.

**Chapter 3** focuses on the concept of autocoups, specifically on power
extensions by incumbent leaders. It distinguishes autocoups from broader concepts
like self-coups or executive coups by redefining them as instances where
incumbent leaders refuse to transition power as mandated, thereby overstaying in
office. Based on this refined definition, a novel dataset of autocoup events from
1945 to 2022 is introduced, encompassing 110 attempts and 87 successes. The
chapter includes case studies and empirical analyses that demonstrate the utility
of this dataset for quantitative research, providing a basis for empirical
analysis on autocoups.

**Chapter 4** investigates how the method of power acquisition impacts the
longevity of leaders who come to power through coups versus those who extend
their rule through autocoups. The hypothesis is that the method of accession
significantly affects leader tenure. Using the Cox proportional hazards model and
a time-dependent Cox model, the chapter provides evidence of differing survival
times between these two types of leaders. The results indicate that leaders who
come to power through coups face a significantly higher risk of removal compared
to those who extend their rule through autocoups. This finding highlights the
implications for political stability and democratic processes, suggesting that
the relatively low cost and high returns of autocoups could incentivize
incumbents to seize power in this manner, potentially leading to democratic
backsliding and the personalization of power.

In **Chapter 5**, the study concludes by summarizing the main findings,
discussing policy implications, and acknowledging the limitations of the
research. It also outlines directions for future research, emphasizing the need
for further exploration of irregular power transitions, particularly coups and
autocoups.

## Power Dynamics and Coup Attempts: A Selection Mechanism Analysis

### Introduction

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data

download.file("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt", "data/Coup.txt")
Coup <- fread(here("data/Coup.txt"))

coups <- Coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(Coup$country)
succeeded <- length(Coup$country[Coup$coup == 2])

Coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

Coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

Coups, defined as "illegal and overt attempts by the military or other elites
within the state apparatus to unseat the sitting executive" [@powell2011, P.252],
occur with varying frequency across countries, with some experiencing them more
often than others. According to GIC dataset, Latin American countries such as
Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups
between 1950 and 1984, while Argentina experienced
`r coups |> filter(country == "Argentina") |> pull(2)` during a similar time
frame. However, Mexico's authoritarian period from 1917 to 2000 saw no coups at
all. In Africa, Sudan endured `r coups |> filter(country == "Sudan") |> pull(2)`
coups between 1955 and 2023, whereas South Africa has not experienced any coup
since 1950. Similar patterns are observed in the Middle East and South Asia.

The varying frequency of coup attempts has captivated political scientists for
decades, leading to extensive research on the subject. As highlighted by
@gassebner2016, despite approximately one hundred potential determinants of coups
being suggested, no consensus has been reached. In an effort to address this
issue, they have tested 66 factors proposed in previous literature using three
million model permutations in an extreme bounds analysis (ibid.).

Examining previous research, which has tested around 100 variables as potential
determinants of coups, raises an important question beyond simply understanding
why coups are more frequent in some countries than others. The critical question
is: Can we establish a framework to help scholars focus on the most relevant
factors of coups, rather than sifting through over 100 variables without reaching
a consensus?

Reviewing previously proposed variables of coups, it is evident that all focus on
pre-coup conditions, with no consideration given to post-coup factors. This means
more attention is paid to factors observable before a coup, such as military
force or spending, economic performance, political stability, and previous coups.

However, coups are high-stakes gambles with an all-or-nothing nature. Due to its
illegality, the consequences of a failed coup can be severe, with perpetrators
risking imprisonment, exile, or even death. In some instances, repercussions
extend to their families, which means coup plotters would not take the risk
without some assurance of success.

Historical coup attempts and their success rates provide valuable hints about
coup plotters' decision-making processes. Despite the significant risks
associated with coups, as shown in @tbl-coups, there have been `r attempts` coups
worldwide since 1950. Importantly, about half of these coups have been
successful. At first glance, coups appear to be a high-success-rate political
venture. However, compared to over 12,000 country-years since 1950, the
occurrence of `r attempts` coups is relatively rare, accounting for only about 4%
(GIC).

The low occurrence rate and high success rate clearly indicate that the
initiation of coups is highly selective. In other words, the likelihood of a coup
occurring depends greatly on its probability of success. However, the probability
of success is not a factor that can be observed before a coup, at least not to
outsiders and researchers, but coup plotters surely have more information on the
possible outcome of coups.

Since coup plotters meticulously assess potential outcomes before staging a coup,
coup researchers should also analyse what factors might affect the outcome of
coups. Otherwise, we ran the risk of selection bias. When considering the factors
that most affect the outcomes of coups, the current literature predominantly
identifies military power as the decisive factor in the success of coups.
Ultimately, military power is determined by the power structure within the
regime.

Unlike coup plotters, who have a clearer understanding of power structures and
dynamics, including their supporters, followers, bystanders, and opponents,
researchers are not able to accurately observe this information. However, we can
analyse the most relevant factors which determine the balance of power and are
observable to us. This leads us to **regime type**, which reflects the
distribution of power within a government, encompassing who controls the
military, sets policy, and appoints officials. Since coup plotters consider the
balance of power within the regime before acting, this study argues that regime
type plays a crucial role in shaping coup attempts. Analysing which regime types
are more susceptible to coups can offer valuable insights.

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| include: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold",
    table.layout = "fixed"
              ) |> 
  # cols_width(1 ~ px(120),
  #            2:4 ~ px(88)
  # ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

We employ a **double probit model with sample selection** to address the
selection bias. This model allows us to analyze both the factors influencing coup
success and the factors related to the decision to initiate a coup attempt in the
first place.

This study contributes to the literature in two key ways. First, it emphasizes
the importance of focusing on expected success rates as a driver of coup
attempts, offering a more targeted approach compared to past studies. Second, it
highlights the significance of regime type as a factor influencing coup
likelihood, even when researchers lack perfect knowledge of a regime's internal
power dynamics.

The subsequent section (Section 2) explores the dynamics of coup attempts and
their outcomes. Section 3 delves into the research design, outlining the
methodology and variables used in the analysis. Section 4 presents and discusses
the empirical findings. Finally, Section 5 concludes the chapter by summarizing
the key insights and their implications.

### Dynamics of coup attempts and outcomes

Coup attempts are driven by a complex interplay of factors, including motivations
(**disposition**) and the resources and opportunities available to succeed
(**capability**).

#### Motivations for coups

This section focuses on the motivations that compel challengers to undertake
coups. We can categorize coup motivations into three main types:

**Personal Ambition:** Personal ambition is a significant motivator for the
majority of coup plotters, driven by the allure of absolute power, prestige, and
wealth. The prospect of seizing absolute power provides an unparalleled
opportunity to shape national policies, control resources, and make significant
decisions without constraints. The pursuit of prestige and recognition, along
with the potential for immense economic gain and wealth, further incentivizes
individuals. Additionally, the desire to leave a lasting legacy and make a
historical impact can compel individuals to undertake the risky venture of
staging a coup.

**Purported National Interest:** Sometimes coups are justified as necessary
interventions to address national crises, uphold the constitution, or facilitate
a transition to democracy. While the motivations behind such claims require
scrutiny, genuine examples do exist. For instance, the 2010 coup in Niger ousted
President Tandja, who attempted an unconstitutional third term by dissolving the
opposing court and calling a self-serving referendum [@ginsburg2019].

**Self-Preservation:** In some cases, coups serve as pre-emptive strikes against
perceived threats. Coup leaders might not necessarily seek power, but rather fear
elimination or political persecution by the incumbent regime. An example is Idi
Amin's 1971 coup against Ugandan President Obote, who was attempting to remove
Amin from his military command position [@sudduth2017].

These motivations are often most prevalent in autocratic regimes, where
justifications under the guise of national interest or self-preservation can be
used to mask personal agendas. Stable democracies rarely face the same level of
constitutional crises or political persecution that might necessitate a coup.
However, new established democracies can be vulnerable to instability, economic
downturns, and democratic backsliding, creating opportunities for coup plotters
to exploit these weaknesses and justify their actions.

Despite the potential motivations outlined above, coups remain relatively
uncommon, occurring in only about 4% of country-years since 1950. The main reason
is that coup perpetrators face inherent disadvantages compared to incumbent
leaders, highlighting the importance of capability. Even the most motivated
plotters need the resources and opportunities to succeed. The next section will
explore the concept of capability in greater detail.

#### Capability for coups

For coup plotters, the decision to act hinges not only on their motivations but
also on a calculated assessment of their chances of success. Several factors can
influence this threshold:

-   **Military Strength**: A clear advantage in military capabilities compared to
    the incumbent regime significantly increases the odds of a successful coup.

-   **Internal Divisions within the Regime**: Existing fractures within the
    government's power structure can present opportunities for coup plotters to
    exploit weaknesses and gain support from disgruntled factions.

-   **Public Support**: Widespread discontent with the incumbent regime,
    especially within the military or key sectors of society, can create an
    environment ripe for a successful coup.

-   **Foreign Backing**: External support from powerful nations can provide
    resources, legitimacy, and even direct military intervention to tip the
    scales in favour of the coup plotters.

The high success rate of coups since 1950, as shown in @tbl-coups, might suggest
that coups are not particularly difficult to succeed. However, it is important to
consider selection bias. We only observe attempted coups, not the numerous
dispositions and conspiracies that never came to light. Analysing launched coup
data alone can be misleading.

To gain a more comprehensive understanding of coup attempts and their likelihood,
we need to move beyond historical data and employ a theoretical framework that
accounts for selection bias. This framework will allow us to analyze the factors
influencing both the success of coups and the decision to attempt a coup in the
first place. The next section will delve into such a framework and its
implications for our understanding of coup attempts.

#### Framework of coup success

An oft-cited framework [@gassebner2016; @aidt2019] provides a structured approach
to assess the disposition and capability of coup attempts by evaluating the
anticipated benefits for coup plotters. The expected pay-off of coups can be
represented by the equation:

$$
\begin{aligned}
E(U) = p \times B + (1 - p) \times (-C)
\end{aligned}
$$ {#eq-eq1}

Here, $\mathbf B$ represents the return of a successful coup, $\mathbf C$
signifies the cost of a failed coup, and $p$ represents the probability of coup
success. The condition for staging a coup is when the expected benefit is
positive, meaning that the expected pay-off is greater than 0. Rearranging the
equation, we get:

$$
\begin{aligned}
p \times B > (1 - p) \times C
\end{aligned}
$$ {#eq-eq2}

@eq-eq2 implies that for @eq-eq1 to hold, the expected benefits earned from
successful coups must outweigh the expected cost of failed coups.

While seemingly clear, the equation faces practical challenges. Quantifying
$\mathbf B$ (the value of a successful coup) and $\mathbf C$ (the cost of
failure) is difficult. The loss of life, freedom, or loved ones after a failed
coup, as well as the value of assuming leadership after a successful coup, are
intangible concepts that defy precise measurement. As evidenced by the 1979 coup
in Ghana[^4], the fate of the coup leader(s) hangs in the balance; they are high
likely to be killed if the coup fails, or to execute others if the coup succeeds.

However, these challenges do not render the framework useless. Firstly, its core
logic remains valuable, offering insights into how coup plotters might assess the
return and cost of their actions. Secondly, given the significant and elusive
nature of precise values for $\mathbf B$ and $\mathbf C$, they can be treated as
roughly equal. Consequently, there is no need to fret over how to measure and
compare these values precisely. Instead, we can shift our focus from $\mathbf B$
and $\mathbf C$, to the probability of success ($p$), simplifying @eq-eq2 to:

$$
\begin{aligned}
p > (1-p)
\end{aligned}
$$ {#eq-eq3}

@eq-eq3 suggests that, to hold @eq-eq2 true, a success probability greater than
$50\%$ is necessary. Interestingly, empirical data on coups since 1950 somewhat
supports this notion. As shown in @tbl-coups, the overall success rate is
$49.9\%$. While this falls short of the $50\%$ threshold, it's important to
consider two factors. Firstly, this is an average rate, not necessarily
reflective of the probabilities assessed by coup plotters beforehand. Secondly,
outliers such as irrational actors and coups driven by self-preservation may not
prioritize success probabilities. Taking these points into account, we can
propose our first hypothesis:

> ***H1: The fundamental determinant of a coup attempt is the perceived chance of
> success. Coup plotters likely require a success threshold of at least 50%.***

This leads us to the next crucial question: What factors determine coup success,
influencing the very decision to attempt one? We will discuss it in the next
section.

#### Regime types and power dynamics

The high success rate in historical data is merely a post-event statistical
description, not indicative of the probability of any specific coup. Importantly,
coup plotters do not rely on past success rates as their guide; instead, they
assess their chances based on their unique context.

Military strength undeniably plays a critical role in coup attempts. Control of
the armed forces offers a significant advantage, explaining why military coups
dominate discussions on the topic. Much of the literature treats "coup" and
"military coup" interchangeably, with scholars like @powell2011 finding that half
of 14 studies attribute coups solely to the military. Consequently, significant
focus, from both researchers and policy-makers, centres on the balance of power
between civilian and military authorities, or among military factions themselves.
Strategies like "keeping the military content" [@aidt2019] or "providing them
with resources" [@huntington1991democratization] aim to reduce military
intervention. Empirical research informs coup-proofing strategies that either
decrease the military's desire for coups or raise barriers to success
[@leon2013a; @powell2018].

However, while military power is decisive, it complicates the analysis. As
@tbl-regimes demonstrates, military regimes, despite concentrated military
control, exhibit surprising instability, experiencing the most frequent coup
attempts. This highlights a crucial issue: the intra-military component. Treating
the military as a monolithic entity ignores its complex internal dynamics
[@singh2016]. Any military comprises diverse groups with their own hierarchies,
fostering suspicion, competition, and vigilance rather than unity. The
clandestine nature of coups necessitates small, secretive groups. Plotters are
unsure of other factions' stances and fear their opposition or intervention, as
exemplified by the swiftly thwarted 2021 Niger coup[^5]. The success of a coup
hinges heavily on other military factions' reactions [@geddes1999].

Moreover, military force is not the only factor shaping the balance of power.
Internal divisions within the ruling elites, public support, and foreign backing
also play important roles. Since the balance of power is not a clearly observable
variable, especially difficult to measure for outsiders and academic scholars, we
can shift our focus from who controls power in a specific context to what factors
shape the balance of power. This leads us to regime type, as the classification
of regime type is based mainly on the power structure. For example,
"...definition of regimes emphasizes the rules that identify the group from which
leaders can come and determine who influences leadership choice and policy"
[@geddes2014, P. 314]. We will leverage this framework to categorize autocracies
based on leadership origin and decision-making, classifying regimes into three
main categories: military, personalist, and dominant-party.

-   **Military Regimes:** Characterized by the dominance of a junta—a group of
    military officers who control the regime's power structure, including
    leadership selection and policy formulation. Examples include the Brazilian
    regime (1964-1985), the Argentine regime (1976-1983), and the Salvadoran
    regime (1948-1984) [@geddes1999].

-   **Personalist Regimes:** Power resides with a single, charismatic leader who
    controls policy, the military, and succession. Regimes like Rafael Trujillo's
    in the Dominican Republic (1930-1961), Idi Amin's in Uganda (1971-1979), and
    Jean-Bédel Bokassa's in the Central African Republic (1966-1979) exemplify
    personalist rule (ibid.).

-   **Dominant-Party Regimes:** Power rests within a well-organized ruling party,
    with leaders acting as its representatives. The party structure and ideology
    foster internal cohesion and a long-term vision. Examples include the Partido
    Revolucionario Institucional (PRI) in Mexico, the Revolutionary Party of
    Tanzania (CCM), and Leninist parties in various Eastern European countries
    (ibid.).

The critical distinction between regime types lies in the unique power balance
established during their seizure of power, often through irregular and violent
means, frequently involving military confrontations. This power struggle
typically results in the emergence of the most competent group—be it a military
junta, a political party, or a strongman—as the core leadership. Furthermore, the
early stages of new regimes are often marked by internal purges to consolidate
power and eliminate potential rivals [@sudduth2017; @roessler2011].

Following these external challenges and internal purges, new power dynamics
emerge in three primary forms: dominant-party regimes, personalist regimes, and
military regimes. These contrasting power dynamics significantly influence a
regime's susceptibility to coups. Dominant-party regimes, characterized by a
well-organized party with clear ideology, hierarchy, and discipline, exhibit the
greatest resilience against coups due to their institutionalized structures and
unified leadership. Personalist regimes, where power centers around a single
strong leader, are relatively stable during the leader's tenure. However, they
face a higher risk of coups, particularly due to the lack of clear succession
plans and vulnerabilities associated with the leader's personal weaknesses,
health issues, and finite lifespan. Military regimes, characterized by power
struggles within a junta, are the most vulnerable to coups. The absence of a
clear final authority in military regimes often leads to internal conflicts, and
the presence of multiple military factions increases the likelihood of resolving
disputes through force, making these regimes the least stable. @tbl-regimes1
summarises the key characteristics of the three main regime types.

These contrasting power dynamics significantly influence a regime's
susceptibility to coups. As @tbl-regimes confirms, military regimes, despite
representing only 5.6% of country-years, experience a disproportionate share of
coups, accounting for over 22% of all occurrences. Personalist regimes face a
similarly high coup risk, constituting 23% of coups while representing only 13%
of country-years. Conversely, dominant-party regimes, with their
institutionalized structures and unified leadership, exhibit the greatest
resilience. They represent 22.6% of country-years but account for only 16.7% of
coups. The column "Coup Likelihood" clearly illustrates the varying risks:
military regimes have the highest likelihood at 17.2%, followed by personalist
regimes at 7.7%, and dominant-party regimes at 3.2% (excluding *Other* regime
types).

> ***H2: Due to their balance of power dynamics, military regimes are more prone
> to coups, followed by personalist regimes, while dominant-party regimes are the
> least likely to experience coups among the three.***

\newpage

\blandscape

```{r}
#| label: tbl-regimes1
#| tbl-cap: Main features of different types of regimes 
#| echo: !expr format == "html"

read.csv(here("data/regimes_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center",
              page.orientation = "landscape"
              ) |>
  cols_width(
    1:6 ~ px(100)
  ) |>
  tab_source_note(source_note = md("*Source: GWF & Author*"))
```

\newpage

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and coups since 1950 
#| echo: !expr format == "html"

regimes <- REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949)

regime_total <- regimes |>
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(country_year =n,
         share = percent(percent),
         .keep = "unused") 
  

coup1 <- Coup |>
  select(country, ccode, year, coup)

coup_by_regime <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup)) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  mutate(num_of_coups = n,
         percent_of_coups = percent(percent,accuracy = 0.1),
         .keep = "unused") 

coup_by_succeed <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup),
         coup == 2) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  select(regime, succeeded_coup = n) 

coup_regimes <-  coup_by_regime |> 
  left_join(coup_by_succeed) |> 
  mutate(success_rate = percent(succeeded_coup / num_of_coups, accuracy = 0.1))

regime_total |>
  left_join(coup_regimes, join_by(regime_type == regime)) |>
  mutate(regime_type = str_to_title(regime_type),
         coup_likelihood = percent(num_of_coups / country_year, accuracy = 0.1)) |>
  select(-succeeded_coup) |> 
  clean_names(case = "title") |>
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN and GIC Datasets*"))
```

\elandscape

### Research Design

#### Double probit with sample selection model

This study employs a sophisticated statistical approach to account for the
selective nature of coup attempts. While coup attempt rates vary across regimes
(as discussed previously), success rates tend to be surprisingly consistent,
hovering around 50% (as shown in @tbl-regimes). This suggests that coup attempts
are not random acts, but rather strategically planned and undertaken only when
the odds of success appear favourable. A standard statistical model would not
account for this selectivity, potentially leading to biased results.

To address this issue, we utilize a two-stage sample selection model, similar to
the approach used by @powell2012. This model has two parts:

-   **Selection Equation (Stage 1):** This stage analyzes the factors influencing
    whether a coup attempt occurs in a particular regime. The primary explanatory
    variable here is regime type, as previously discussed. Additional control
    variables may also be included, denoted by $\mathbf{XB}$.

-   **Outcome Equation (Stage 2):** This stage focuses on the probability of
    success for those coup attempts that actually take place.

The primary explanatory variables are regime types, as previously discussed.
Control variables are included in $\mathbf{XB}$. The selection equation (first
stage) models the probability that a coup attempt occurs and can be expressed as
follows:

$$
\begin{aligned}
y_1^*=\alpha_0 + \alpha_1 Regime_i + \mathbf {XA} + \mu_{1i}
\end{aligned}
$$ {#eq-eq4}

Here, ${y_1}^*$ is an unobserved variable, which may be known to coup plotters.
$Regime_i$ is a categorical variable (*military*, *personalist*, or
*dominant-party*). $\mathbf{XB}$ captures other control variables, such as the
economic crisis index, previous coups, military expenditure, etc.

The observed binary outcome $\mathbf{y_1}$ is:

$$
\begin{aligned}
y_1 = 
\begin{cases} 
1 &\text{if $y_1^*>0$ (coup attempt occurs)} \\
0 &\text{if $y_1^*\le0$ (no coup attempt)}
\end{cases}
\end{aligned}
$$\
In the first stage, if $y_1^*\le0$, no coup attempt occurs in a given
country-year, indicating that the unobserved variable does not reach the
threshold. If $y_1^*>0$, at least one coup attempt is made in a country-year,
indicating that the unobserved variable surpasses the threshold. The probability
is expressed as:

$$
\begin{aligned}
Prob(y_1 =1)&=Prob(y_{1}^*>0) \\
&=\Phi(\alpha_0 + \alpha_1 Regime_i + \mathbf{XA})
\end{aligned}
$$ {#eq-eq4a}

Similarly, the outcome equation (second stage) models the probability that a coup
attempt is successful, given that it occurs:

$$
\begin{aligned}
y_2^*=\beta_0 + \beta_1 Regime_i + \mathbf {XB} + \mu_{2i}
\end{aligned}
$$ {#eq-eq5}

The observed outcome $y_2$ is:

$$
\begin{aligned}
y_2 = 
\begin{cases} 
1 &\text{if $y_2^*>0$ (coup succeeds)} \\
0 &\text{if $y_2^*\le0$ (coup fails)}
\end{cases}
\end{aligned}
$$

The probability equations is:

$$
\begin{aligned}
Prob(y_2=1|y_1 =1)=\Phi(\beta_0 + \beta_1 Regime_i + \mathbf{XB})
\end{aligned}
$$ {#eq-eq5a}

#### Variables

-   Dependent variable

Our analysis utilizes data on coup attempts and outcomes from @powell2011. A
successful coup is defined as one where the incumbent leader is removed from
power for more than seven days. The dataset covers the period from 1950 to 2023
and includes information on 491 coup attempts, with roughly half (245) being
successful. Descriptive statistics for these coup attempts and regime types can
be found in @tbl-coups and @tbl-regimes.

-   Key Independent Variable: Regime Type

The core variable of interest is regime type, categorized following the
classification system of @geddes2014 (GWF). We focus on military, personalist,
and dominant-party regimes, with democracies and monarchies included for
comparison. Descriptive statistics for regime types are presented in
@tbl-regimes.

-   Control variables

Our control variables are chosen based on the research of @gassebner2016. They
analyzed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

**Economic Performance:** We measure economic performance using the current-trend
($CT$) ratio developed by @krishnarajan2019. This ratio compares a country's
current GDP per capita to the average GDP per capita over the previous five
years. A higher $CT$ ratio indicates stronger economic performance. We use GDP
per capita data (in constant 2017 international 1000 dollars, PPP) from the V-Dem
dataset by @fariss2022, lagged by one year to reflect the prior year's economic
impact. For a country $i$ at year $t$, the $CT$ ratio is calculated as follows:
$$
\begin{aligned}
CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
\end{aligned}
$$

**Political Violence:** We capture overall regime stability by including a
violence index that encompasses all types of internal and interstate wars and
violence. This data comes from the Major Episodes of Political Violence dataset
by Marshall [@marshall2005current].

**Previous coups:** The number of previous coups in a country is included in the
first-stage (selection) model to assess its influence on the likelihood of a coup
attempt. However, it is excluded from the second-stage model (outcome) because
the number of past coups may not directly impact the outcome of a specific coup
attempt. Recognizing that previous coups from a long time ago might not
significantly affect the current situation, I ran another regression using the
time distance since the last coup instead of the number of previous coups for
comparison. This approach accounts for the diminishing impact of past events over
time, providing a potentially more accurate assessment of how historical coup
activity influences the likelihood of current coup attempts.

### Results and Discussion

```{r}
#| label: regression
#| results: asis
#| echo: !expr format == "html"
download.file("https://raw.githubusercontent.com/reddylee/Datasets/main/political_violence.csv", "data/political_violence.csv")
political_violence <- fread(here("data/political_violence.csv")) |> 
  dplyr::select(ccode, year, violence = actotal)

gdp <- vdem |> 
  dplyr::select(ccode = COWcode, year, gdppc = e_gdppc) |> 
  filter(year > 1940) |> 
  mutate(ct = gdppc / lag(rollapply(gdppc, width = 5, FUN = mean,
                               align = 'right', fill = NA)),
         .by = c(ccode)) 

download.file("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_ccode_year.txt", "data/coup_all.csv")
coup_all <- fread(here("data/coup_all.csv")) |> 
  mutate(coup = case_when(
    coup1 == 0 ~ 0,
    coup1 == 2 | coup2 == 2 | coup3 == 2 | coup4 == 2 ~ 2,
    TRUE ~ 1
  )) |>
  group_by(ccode) %>%
  arrange(year) %>%
  mutate(pre_coups = lag(cumsum(coup %in% c(1, 2)), default = 0),
         last_coup_year = lag(case_when(coup %in% c(1, 2)  ~ year,
                                        TRUE ~ NA_integer_)),
         last_coup_year = case_when(year == min(year) ~ year,
                                    TRUE ~ last_coup_year)) |> 
  fill(last_coup_year, .direction = "down") |>
  mutate(time_since_last_coup = year - last_coup_year,
         coup_dummy = case_when(pre_coups == 0 ~ 0,
                                TRUE ~ 1)) |>
  dplyr::select(ccode, year, coup, pre_coups, time_since_last_coup, coup_dummy) 

reg_data <- regimes |> 
  dplyr::select(country = extended_country_name,
         ccode = reign_cowcode, year, regime_type) |>
  left_join(coup_all, by = join_by(ccode, year)) |> 
  left_join(political_violence, by = join_by(ccode, year)) |>
  left_join(gdp, by = join_by(ccode, year)) |> 
  filter(!is.na(gdppc)) |> 
  mutate(violence = na.locf(violence, .by = ccode)) 

model_1 <- reg_data |>
  filter(regime_type != "other") |>
  mutate(regime = fct_relevel(regime_type, "dominant-party"),
         ect = (ct-1)*100) |>
  selection(coup > 0 ~ regime + ect + gdppc + violence + coup_dummy*time_since_last_coup,
            coup > 1 ~ regime + ect + gdppc + violence,
            data = _) 

model_1_selection <- model_1
model_1_selection$param$index$betaO <- model_1$param$index$betaS
model_1_selection$param$index$betaS <- model_1$param$index$betaO

model_2 <- reg_data |>
  filter(regime_type != "other") |>
  mutate(regime = fct_relevel(regime_type, "dominant-party"),
         ect = (ct-1)*100,
         Coup = coup) |>
  select(Coup, regime, ect, gdppc, violence, coup_dummy = pre_coups) |> 
  selection(Coup > 0 ~ regime + ect + gdppc + violence + coup_dummy,
            Coup > 1 ~ regime + ect + gdppc + violence,
            #method = "2step",
            data = _) 

model_2_selection <- model_2
model_2_selection$param$index$betaO <- model_2$param$index$betaS
model_2_selection$param$index$betaS <- model_2$param$index$betaO

stargazer(model_2_selection, model_2, model_1_selection, model_1,
          #type = "text",
          header = F,
          dep.var.caption = "",
          dep.var.labels = c("Model 1", "Model 2"),
          #dep.var.labels.include = T,
          #multicolumn = T,
          title = "Sample Selection Model of Regime Types and Coups, 1950-2019",
          column.labels = c("Coup Attempts", "Coup Outcome","Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          label = "results",
          covariate.labels = c("Constant", "Regime: Democracy", "\\hspace{1.6cm}Military", "\\hspace{1.6cm}Monarchy", "\\hspace{1.6cm}Personalist", "Economic trend", "GDP per capita", "Political violence", "Previous coups (P)", "Years since last coup (Y)", "Interactive term: P * Y"),
          summary.logical = F,
          column.sep.width = "7pt",
          font.size = "small"
          )
```

The double `probit` model with sample selection, estimated using the
`sampleSelection` package [@sampleSelection-2] in R, offers valuable insights
into the factors influencing coup attempts and their outcomes across different
regime types from 1950 to 2019 (Table \ref{results}). Models 1 and 2 use the
numbers of previous coups

\newpage

#### The Selection Model: Coup Attempts

```{r}
#| label: tbl-mfx1
#| tbl-cap: Average marginal effects of coup attempts (Selection Model)
#| echo: !expr format == "html"
avg_comparisons(model_2) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_2, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups",
                          term == "ect" ~ "Economic trend",
                          term == "gdppc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Regime: Democracy",
                          str_detect(contrast, "military") ~ "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Regime: Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_style(
    style = cell_text(indent = px(63)),
    locations = cells_body(
      columns = Term,
      rows = 2:4
    )
  ) |> 
  tab_options(
    table.width = pct(95),
    # table.font.size = px(14),
    column_labels.font.weight = "bold"
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  )
#avg_predictions(model_2, variables = "regime")
# slopes(model_2, newdata = "mean")
# predictions(model_2, newdata = "mean") |>
#   view()
  
```

In the selection model (Column 1), military and personalist regimes exhibit
significant positive coefficients at the 1% level, indicating a higher likelihood
of experiencing coup attempts compared to dominant-party regimes. Using a
marginal effect package [@marginaleffects], as shown in @tbl-mfx1, the
probability of coup attempts in military regimes is, on average, 7 percentage
points (pp) higher than in dominant-party regimes, while personalist regimes are
2.4 pp higher, holding other factors constant. Monarch regimes also display a
positive effect similar to personalist regimes, reflecting the fact that
monarchies are a subset of personalist regimes with royal titles.

While a 7 pp increase might initially seem modest, it is substantial in context.
The average probability of a coup attempt in dominant-party regimes is only about
2.53%. A 7 pp increase means that military regimes are approximately 2.8 times
more likely to encounter coups than dominant-party regimes, which represents a
significant change.

The results aligns with theoretical expectations regarding the internal power
struggles within military juntas and the succession vulnerabilities in
personalist regimes, underscoring the importance of regime structure in
understanding coup likelihood.

Control variables also show effects in the expected directions. Stronger economic
performance, as indicated by higher economic growth trends and GDP per capita
levels, correlates with a lower risk of coup attempts. This suggests that better
economic performance and relatively higher living standards diminish the
incentives for coups. Political violence has a positive and significant effect on
coup attempts, indicating that higher levels of instability increase the
likelihood of coups. The positive coefficient for the number of previous coups
suggests a "copycat" effect from earlier examples. However, the control variables
are less substantial in predicting coup attempts, as the average marginal effect
shown, a one-unit increase in these factors' value results in only 0.2 pp change
in coup probability, ceteris paribus. As the average probability of coup attempt
is about 2%, 0.2 pp change means the average marginal effects is only 10% change,
which is not substantial.

Control variables exhibit effects in expected directions. Stronger economic
performance, indicated by higher economic growth trends and GDP per capita
levels, correlates with a lower risk of coup attempts. This suggests that better
economic conditions and relatively higher living standards reduce incentives for
coups. Political violence shows a positive and significant effect on coup
attempts, indicating that higher levels of instability increase the likelihood of
coups. The positive coefficient for the number of previous coups suggests a
"copycat" effect from earlier incidents.

However, these control variables have less substantial impacts in predicting coup
attempts. As shown by the average marginal effects (AME), a one-unit increase in
these factors' values results in only a 0.2 pp change in coup probability,
ceteris paribus. Given that the average probability of a coup attempt is
approximately 2%, a 0.2 pp change represents only a 10% relative change, which is
not substantial.

#### The Outcome Model: Coup Success

The outcome model (Column 2 in Table \ref{results}) reveals determinants of coup
success. Military regimes have a higher probability of coup success compared to
dominant-party regimes, meeting expectations that military regimes face higher
coup risks due to their better chances of success. Personalist and monarch
regimes show a slight positive effect on coup success, but these are not
statistically significant.

Control variables exhibit different effects in the outcome model. Both GDP per
capita and political violence maintain a weak influence similar to the selection
model, while the economic trend shows a less significant negative effect.

These results indicate that regime type remains a significant determinant of coup
attempts and successes, even after controlling for other factors, strongly
supporting the theoretical framework.

```{r}
#| label: fig-margins
#| fig-cap: "Probability of Coup Attempts + Coup Success"
#| echo: !expr format == "html"
plot_predictions(model_2, by = "regime")+
  labs(
    x = "Regimes",
    y = "Total probability")

# plot_comparisons(model_2, 
#                  variables = "ect",
#                  by = "regime")
```

#### Discussion

The $\rho$ value of 0.898, which is highly significant (p \< 0.01), is a crucial
parameter in the sample selection model. This value represents the correlation
between the error terms of the selection equation (coup attempts) and the outcome
equation (coup outcomes). A high and significant $\rho$ suggests that unobserved
factors influencing the likelihood of a coup attempt are strongly correlated with
those influencing the likelihood of a successful coup. In practical terms, this
means that the selection model is appropriate and that accounting for the
selection bias (i.e., the fact that only coups with high chances of success will
be attempted) is critical to obtaining unbiased estimates. The high $\rho$ value
indicates that the same underlying conditions that lead to a coup attempt also
affect the success of the coup, underscoring the importance of considering both
stages in the analysis.

The results strongly support the choice of the sample selection model.
Significant coefficients with theoretically consistent directions suggest the
model effectively captures key aspects of coup dynamics. Regimes with weaker
institutional structures are more vulnerable to coup attempts, while better
economic conditions make coups less likely overall. The model effectively
addresses the non-random nature of coup attempts by treating selection and
outcome as separate processes.

The observed disparity between coup attempt rates and success rates across
regimes points towards selection bias, further validating the use of the sample
selection model. This model acknowledges that coups are not random events, but
rather strategic actions undertaken when the odds appear favorable.

In summary, the double `probit` model with sample selection proves to be a
well-suited approach for this research. It provides robust insights into the
factors influencing both the likelihood of coup attempts and their success rates
across different regime types. The findings highlight the crucial role of regime
structure and the selective nature of coup attempts, supporting the theoretical
framework and empirical strategy employed in this study.

### Conclusion

Motivated by the lack of consensus despite numerous empirical studies on the
determinants of coups, this study introduces a novel approach that prioritizes
determinants based on their impact on coup success. By analysing coup success
rates, the study hypothesizes that the expected outcomes of coups are critical
determinants of their occurrence. Utilizing a double `probit` model with sample
selection, I investigate and confirm the relationship between regime types and
coup attempts.

The findings suggest that regime type plays a significant role in the likelihood
of coup attempts. Military and personalist regimes, characterized by weaker
institutional frameworks and higher vulnerability during power transitions, are
more susceptible to coups. This underscores the importance of supporting
initiatives that strengthen constitutional institutions within these regimes.

The research also finds that stronger economic performance is associated with a
lower risk of coups, suggesting that policies promoting economic development can
be effective in reducing coup risk.

The study shows that the most efficient coup-proofing strategies involve the
establishment of strong institutions. In contrast, purges, random shifting of
military officers, or increased military expenditures are less effective.
However, few autocratic leaders, particularly dictators or military juntas, are
willing to institutionalize their regimes, as such reforms may constrain their
power or shorten their terms. While institutions benefit the regime, they do not
necessarily benefit the leaders themselves.

Future research could explore specific institutional reforms that are most
effective in improving stability across different regimes.

## Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders

## Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-Entry and Autocoup Leaders

## Conclusion

### Main Findings

This study delves into the dynamics and implications of irregular power
transitions, focusing on coups and autocoups. The findings illuminate the complex
interplay between incumbents and challengers fighting for power.

Firstly, our analysis reveals that the expected success rate of a coup attempt
significantly influences its likelihood. This success rate is heavily influenced
by the balance of power between the incumbent regime and challengers, which is
largely determined by regime type. We find that military regimes, although with
more control over their own military forces, face a higher risk of coups compared
to dominant-party regimes.

Secondly, the study introduces a redefined concept: the autocoup. Defined as an
incumbent leader's refusal to relinquish power as mandated, this research
distinguishes autocoups from broader terms like self-coups. Based on this
definition, we present the first publicly available dataset of autocoup events
from 1945 to 2022, encompassing 110 attempts and 87 successful autocoups. Case
studies and empirical analyses demonstrate the dataset's utility for quantitative
research, providing a robust foundation for further analysis on autocoups.

Thirdly, employing survival analysis techniques, the study finds clear
differences in leader longevity between those who come to power through coups and
those who extend their rule through autocoups. The results indicate that
coup-installed leaders face a significantly higher risk of removal compared to
autocoup leaders who manipulate the system to extend their rule.

### Policy Implications

The findings of this study offer valuable insights for policy-makers concerned
with promoting and protecting global democracy, which has faced increasing
challenges despite a general post-WWII trend towards democratization. Notably,
the "third wave" of democratization [@huntington1991democratization] witnessed a
surge in democratic transitions in the late 20th century. Since the Cold War's
end, democratic nations have outnumbered non-democratic ones (@fig-democracy)
with the gap widening.

```{r}
#| label: fig-democracy
#| fig-cap: Comparison of the number of democratic and non-democratic countries (1945-2020)
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| echo: !expr format == "html"
REIGN |>
  filter(year > 1945 & year < 2021) |> 
  mutate(democracy = case_when(
    gwf_regimetype %in% c("parliamentary", "presidential") ~ "Democracy",
    .default = "Non-democracy")) |> 
  summarise(Count = n(), 
            .by = c(year, democracy)) |> 
  #view()
  ggplot(aes(year, Count, color = democracy)) +
  geom_point() +
  labs(color = 'Regime type',
       x = "Years: 1945 ~ 2020",
       y = "No. of Countries") +
  theme(legend.position = "top") +
  annotate("text",
    label = "Data source: REIGN (Regime data)",
    x = 2010,
    y = 30,
    fontface = "italic",
    size = 3
    )
```

However, a "democratic recession" has emerged in recent years [@diamond2008].
Freedom House reports an 18th consecutive year of global freedom decline in 2023
[@freedomhouse2024freedom]. While few countries have completely regressed to
autocracy, the average global democracy level has fallen back to pre-2000 levels.
Notably, democratic backsliding often occurs within regimes, with democracies
becoming less liberal and autocracies becoming less competitive [@mechkova2017].

This research highlights irregular power transitions as a significant factor in
democratic backsliding within regimes. These transitions, often coups or
autocoups, violate democratic norms and disrupt the path towards stable
democracies. Leaders who gain power through irregular means often resort to
undemocratic tactics to maintain control, creating a vicious cycle of eroding
democratic institutions.

Our findings suggest that the shorter lifespans and potentially severe
consequences associated with coups may deter potential coup leaders. Conversely,
autocoups appear to be a more tempting option for power-hungry leaders due to
their higher success rates, seemingly moderate consequences, and extended leader
tenure after the autocoup. This trend may explain the decline in classic coups
since the 1990s alongside the rise of autocoups [@bermeo2016].

### Limitations and directions for future research

This study offers a novel framework for analysing irregular power transitions,
but some limitations require further exploration:

-   **Data refinement:** Defining and classifying autocoups is a new approach.
    Future research should validate this classification system through additional
    studies and expert evaluations.

-   **Data harmonization:** The current analysis faces challenges due to
    mismatched units (country-year vs. leader) between coup and autocoup
    datasets. Future efforts should explore data harmonization techniques for
    more robust comparisons.

-   **Democratic backsliding:** While this study establishes a connection between
    irregular power transitions and democratic backsliding, further empirical
    evidence is needed to solidify this link.

Several avenues exist for future research:

-   **Terminology and data collection:** Refining the "autocoup" concept and
    achieving wider recognition will facilitate more accurate and comprehensive
    data collection.

-   **Dataset expansion:** Expanding the autocoup dataset with more cases and
    integrating it with data on other irregular leadership transitions can
    provide a more holistic view of political survival after these events.

-   **Power dynamics and long-term impacts:** Utilizing this dataset, future
    studies can delve deeper into power dynamics at play and explore the
    long-term consequences of irregular transitions on political systems,
    particularly regarding democratic backsliding, breakdown, and personalization
    of power.

In conclusion, this study sheds light on the dynamics of irregular power
transitions, specifically focusing on coups and autocoups. By redefining
autocoups, classifying the dataset, analysing determinants, and comparing leader
longevity, we establish a framework for understanding irregular transitions and
leader survival. This work contributes to a deeper understanding of democratic
resilience and political stability. Future research can build upon this
foundation by conducting further empirical analyses based on the novel autocoup
dataset and continuing to refine the framework.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false

cat("\\newpage
## References")
```

[^1]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^2]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^3]: The definitions and concepts of power expansion and power extension can be
    vague. In this study, we define power expansion as an incumbent acquiring
    additional authority from other state apparatuses, whereas power extension
    refers to an incumbent prolonging their tenure beyond the designated term in
    office.

[^4]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^5]: The definitions and concepts of power expansion and power extension can be
    vague. In this study, we define power expansion as an incumbent acquiring
    additional authority from other state apparatuses, whereas power extension
    refers to an incumbent prolonging their tenure beyond the designated term in
    office.
