---
title: "Power Dynamics and Coup Attempts: A Selection Mechanism Analysis"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: coups1.html
categories: [Article, Coups]
bibliography: references.bib

abstract: |
  Despite extensive research identifying around one hundred potential determinants of coup attempts, no consensus has been reached. This study introduces a novel approach that prioritizes determinants based on their impact on coup success. By analyzing coup success rates, the study hypothesizes that the expected outcomes of coups are critical determinants of their occurrence. Utilizing a double probit model with sample selection, the research investigates the relationship between regime types and coup attempts. The findings suggest that regime type, by shaping internal power dynamics, is a crucial determinant of coup likelihood.

keywords: [Coup attempts, Coup outcomes, Regime types, Sample selection]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(coup$country)
succeeded <- length(coup$country[coup$coup == 2])

coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

## Introduction

Coups occur with varying frequency across different countries, with some
experiencing them more frequently than others. According to the Global Instances
of Coups (GIC)[^1] dataset [@powell2011], Latin American countries such as
Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups
between $1950$ and $1984$, while Argentina experienced
`r coups |> filter(country == "Argentina") |> pull(2)` during a similar time
frame. However, Mexico's authoritarian period from $1917$ to $2000$ saw no coups
at all. In Africa, Sudan endured
`r coups |> filter(country == "Sudan") |> pull(2)` coups between $1955$ and
$2023$, whereas South Africa has not experienced any coup since $1950$. Similar
patterns are observed in the Middle East and South Asia.

The varying frequency of coup attempts has captivated political scientists for
decades, leading to extensive research on the subject. As highlighted by
@gassebner2016, despite approximately one hundred potential determinants of coups
being suggested, no consensus has been reached. In an effort to address this
issue, @gassebner2016 tested 66 factors proposed in previous literature using
three million model permutations in an extreme bounds analysis.

Examining previous research, which has tested around 100 variables as potential
determinants of coups, raises an important question beyond simply understanding
why coups are more frequent in some countries than others. The critical question
is: Can we establish a method to help scholars focus on the most relevant factors
of coups, rather than sifting through over 100 variables without reaching a
consensus?

Reviewing previously proposed variables of coups, it is evident that all focus on
pre-coup conditions, with no consideration given to post-coup factors. However,
coups are high-stakes gambles with an all-or-nothing nature. As defined by
@powell2011, coups are "illegal and overt attempts by the military or other
elites within the state apparatus to unseat the sitting executive" [@powell2011,
P.252]. Due to their illegality, the consequences of a failed coup can be severe,
with perpetrators risking imprisonment, exile, or even death. In some instances,
repercussions extend to the families of the coup perpetrators. Therefore, no coup
plotters would stage a coup without some assurance of success.

Historical coup attempts and their success rates provide valuable insights.
Despite the significant risks associated with coups since 1950, as shown in
@tbl-coups, there have been `r attempts` coups worldwide. Importantly, about half
of these coups have been successful. At first glance, coups appear to be a
high-success-rate political venture. However, compared to over 12,000
country-years since 1950, the occurrence of `r attempts` coups is relatively
rare, accounting for about 4% [@powell2011].

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold"
              ) |> 
  # cols_width(1 ~ px(120),
  #            2:4 ~ px(88)
  # ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

The low occurrence rate and high success rate clearly indicate that the
initiation of coups is highly selective. In other words, the likelihood of a coup
occurring depends greatly on its potential success rate. Since coup plotters
meticulously assess potential outcomes, we should also analyze what factors most
affect these outcomes when discussing the key determinants of coups. This
approach allows us to focus on the most relevant factors and disregard those less
related.

When considering the factors that most affect the outcomes of coups, the current
literature predominantly identifies military power as the decisive factor in the
success of coups. This necessitates an analysis of power dynamics within regimes,
as military power is ultimately shaped by power dynamics.

Because coup attempts are self-selective rather than random, this study employs a
double `probit` model with sample selection to examine factors influencing coup
success rates and, consequently, the likelihood of coup attempts. I posit that
regime type, by shaping internal power dynamics among coup plotters, incumbents,
and other ruling elites, is a crucial determinant of coup likelihood.

This study makes two key contributions to the existing literature. First, it
underscores the importance of regime type as a crucial determinant of coup
attempts. Previous studies often treat regime type as a control variable,
overlooking that variations in many other variables are fundamentally rooted in
different regime types. More importantly, this study establishes a systematic
approach for identifying the most relevant factors, thereby avoiding sifting
through over 100 variables.

The subsequent sections of this paper explore the dynamics of coup attempts and
their outcomes. In Part 3, I detail the research design, outlining the
methodology and variables used in the analysis. Part 4 presents and discusses the
empirical findings. Finally, Part 5 concludes the study, summarizing the key
insights and their implications.

## Dynamics of coup attempts and outcomes

Coup attempts are driven by a complex interplay of factors, with two key elements
attracting significant scholarly attention: **disposition** (the motivations
behind the attempt) and **capability** (the resources and opportunities to
succeed).

### Motivations for coups

This section focuses on disposition, exploring the primary motivations that
compel individuals to undertake the significant risks associated with a coup. We
can categorize coup motivations into three main types:

**Personal Ambition:** The allure of absolute power, prestige, and wealth is a
significant motivator for some coup plotters. For example, @wintrobe2019
distinguishes between totalitarian and tinpot dictators based on their use of
power. While both prioritize personal gain, totalitarian leaders seek complete
control over every aspect of society, whereas tinpot leaders focus on enriching
themselves through extravagant lifestyles.

**Purported National Interest:** Coups are sometimes justified as necessary
interventions to save a nation from crisis, uphold the constitution, or
facilitate a transition to democracy. While scepticism is warranted due to the
potential for self-serving justifications, legitimate cases do exist. For
instance, the 2010 coup in Niger ousted President Tandja, who attempted an
unconstitutional third term by dissolving the opposing court and calling a
self-serving referendum [@ginsburg2019].

**Self-Preservation:** In some instances, coups are pre-emptive strikes against
imminent political persecution or repression. Coup leaders might not be motivated
by a desire for power, but rather a fear of elimination by the incumbent regime.
A notable example is Idi Amin's 1971 coup against Ugandan President Obote, who
was attempting to remove Amin from his military command position [@sudduth2017].

These motivations can arise in any regime, but autocracies are particularly
susceptible, especially for coups framed under the guise of national interest or
self-preservation. Stable democracies, on the other hand, rarely face the same
level of constitutional crises or political persecution that might necessitate a
coup. However, new democracies can be vulnerable to instability, economic
downturns, and democratic backsliding, creating opportunities for coup plotters
to exploit these weaknesses and justify their actions.

Despite the potential motivations outlined above, coups remain relatively
uncommon events, occurring in only about 4% of country-years since 1950. This low
frequency highlights the importance of the second key element – capability. Even
the most motivated plotters require the resources and opportunities to succeed.
No rational actor attempts a guaranteed failure; the next section will explore
the concept of capability in greater detail.

### Capability for coups

While many ambitious individuals may covet supreme power, only a select few
possess the capability to orchestrate a successful coup. This capability hinges
not just on their desire, but on overcoming inherent disadvantages compared to
the incumbent leaders.

Firstly, coups are inherently clandestine operations due to their illegality.
Plotters require a tight-knit group to minimize leaks and maximize the element of
surprise. This secrecy restricts their ability to openly recruit supporters, a
privilege enjoyed by incumbents who can implement "coup-proofing" measures.

Secondly, coup plotters face uncertainty about the reactions of other powerful
factions within the regime, those who could tip the scales of power. Incumbents,
however, have a deeper understanding of these dynamics and proactively work to
solidify their own position. While they may not know who exactly might attempt a
coup, they are attuned to potential threats and adapt their strategies
accordingly.

Thirdly, coup plotters face a significant challenge in securing unwavering
loyalty from potential co-conspirators. The risks associated with a coup are
substantial, with uncertain rewards even in the event of success. Promises made
by coup leaders might not be kept, and post-coup purges are a common tactic to
eliminate future coup threats. Defecting to the incumbent leader can often be a
safer option, offering predictable rewards and less risk.

Given these inherent obstacles, rational coup plotters are unlikely to gamble on
a low-probability attempt. They may choose to abandon their plans altogether or
bide their time for a more opportune timing. Therefore, when coup plotters do
take action, it is because they have meticulously assessed their chances of
success and believe the risks are outweighed by the potential gains.

But what is the threshold for a "good enough" chance of success? Before diving
into a theoretical framework, let's examine historical data to gain some
perspective. Surprisingly, coups since 1950 boast a rather high success rate,
with nearly half ending in victory (as shown in @tbl-coups).

### Framework of coup success

An oft-cited framework [@gassebner2016; @aidt2019] provides a structured approach
to assess the disposition and capability of coup attempts by evaluating the
anticipated benefits for coup plotters. The expected payoff of coups can be
represented by the equation:

$$
\begin{aligned}
E(U) = p \times B + (1 - p) \times (-C)
\end{aligned}
$$ {#eq-eq1}

Here, $\mathbf B$ represents the return of a successful coup, $\mathbf C$
signifies the cost of a failed coup, and $p$ represents the probability of coup
success. The condition for staging a coup is when the expected benefit is
positive, meaning that the expected pay-off is greater than 0. Rearranging the
equation, we get:

$$
\begin{aligned}
p \times B > (1 - p) \times C
\end{aligned}
$$ {#eq-eq2}

@eq-eq2 implies that for @eq-eq1 to hold, the expected benefits earned from
successful coups must outweigh the expected cost of failed coups.

While seemingly clear, the equation faces practical challenges. Quantifying
$\mathbf B$ (the value of a successful coup) and $\mathbf C$ (the cost of
failure) is difficult. The loss of life, freedom, or loved ones after a failed
coup, as well as the value of assuming leadership after a successful coup, are
intangible concepts that defy precise measurement. As evidenced by the 1979 coup
in Ghana[^2], the fate of the coup leader(s) hangs in the balance; they are high
likely to be killed if the coup fails, or to execute others if the coup succeeds.

However, these challenges do not render the framework useless. Firstly, its core
logic remains valuable, offering insights into how coup plotters might assess the
return and cost of their actions. Secondly, given the significant and elusive
nature of precise values for $\mathbf B$ and $\mathbf C$, they can be treated as
roughly equal. Consequently, there is no need to fret over how to measure and
compare these values precisely. Instead, we can shift our focus from $\mathbf B$
and $\mathbf C$, to the probability of success ($p$), simplifying @eq-eq2 to:

$$
\begin{aligned}
p > (1-p)
\end{aligned}
$$ {#eq-eq3}

@eq-eq3 suggests that, to hold @eq-eq2 true, a success probability greater than
$50\%$ is necessary. Interestingly, empirical data on coups since 1950 somewhat
supports this notion. As shown in @tbl-coups, the overall success rate is
$49.9\%$. While this falls short of the $50\%$ threshold, it's important to
consider two factors. Firstly, this is an average rate, not necessarily
reflective of the probabilities assessed by coup plotters beforehand. Secondly,
outliers such as irrational actors and coups driven by self-preservation may not
prioritize success probabilities. Taking these points into account, we can
propose our first hypothesis:

> ***H1: The fundamental determinant of a coup attempt is the perceived chance of
> success. Coup plotters likely require a success threshold of at least 50%.***

This leads us to the next crucial question: what factors determine a coup's
success, influencing the very decision to attempt one? While specifics may vary,
the core element hinges on the power dynamic between coup plotters and the
incumbent leaders. Logically, the more powerful entity holds a greater advantage
in this high-stakes struggle for control.

### Regime types and power dynamics

Military strength undeniably plays a critical role in coup attempts. Control of
the armed forces offers a significant advantage, explaining why military coups
dominate discussions on the topic. Much of the literature treats "coup" and
"military coup" interchangeably, with scholars like @powell2011 finding half of
14 studies attribute coups solely to the military. Consequently, significant
focus, from both researchers and policymakers, centers on the balance of power
between civilian and military authorities, or among military factions themselves.
Strategies like "keeping the military content" [@aidt2019] or "providing them
with resources" [@huntington1991democratization] aim to reduce military
intervention. Empirical research informs coup-proofing strategies that either
decrease the military's desire for coups or raise barriers to success
[@leon2013a; @powell2018].

However, while military power is decisive, previous literature often
oversimplifies its nature. As @tbl-regimes will demonstrate, military regimes,
despite concentrated military control, exhibit surprising instability. Military
regimes experience most frequent coup attempts. This highlights a crucial
oversight: the intra-military component. Treating the military as a monolithic
entity ignores the complex internal dynamics [@singh2016]. Regardless of size,
any military comprises diverse groups with their own hierarchies, fostering
suspicion, competition, and vigilance rather than unity. The clandestine nature
of coups necessitates small, secretive groups. Plotters are unsure of other
factions' stances and fear their opposition or intervention, as exemplified by
the swiftly thwarted 2021 Niger coup[^3]. The success of a coup hinges heavily on
other military factions' reactions [@geddes1999].

Furthermore, the relationship between government and military varies across
regimes. In democracies, civilian authority reigns supreme. The military is a
national institution bound by the constitution, not individual leaders, ensuring
political neutrality (e.g., the U.S. Armed Forces). Conversely, non-democracies
display a less clear power structure. Identifying the true leader of the military
depends on the regime type. We will leverage framework of @geddes2014 to
categorize autocracies based on leadership origin and decision-making. This
framework classifies regimes into three main categories: military, personalist,
and dominant-party.

**Military regimes** are characterized by the dominance of a junta – a group of
military officers who control the regime's power structure, including leadership
selection and policy formulation. Examples include the Brazilian regime
(1964-1985), the Argentine regime (1976-1983), and the Salvadoran regime
(1948-1984) [@geddes1999]. In **personalist regimes**, power resides with a
single, charismatic leader who controls policy, the military, and succession.
Regimes like Rafael Trujillo's in the Dominican Republic (1930-1961), Idi Amin's
in Uganda (1971-1979), and Jean-Bédel Bokassa's in the Central African Republic
(1966-1979) exemplify personalist rule [@geddes1999]. In **dominant-party
regimes,** power rests within a well-organized ruling party, with leaders acting
as its representatives. The party structure and ideology foster internal cohesion
and a long-term vision. Examples include the Partido Revolucionario Institucional
(PRI) in Mexico, the Revolutionary Party of Tanzania (CCM), and Leninist parties
in various Eastern European countries [@geddes1999].

The critical distinction between regime types lies in the unique power balance
established during the initial power seizure. The most competent group, be it a
military junta, a political party, or a strongman, typically prevails due to the
challenges of seizing control. This power grab is often accompanied by purges of
potential rivals, solidifying the newly established regime [@sudduth2017;
@roessler2011].

Following these internal purges and external challenges, a new power dynamic
emerges, typically solidifying into one of three main structures: military
regimes, personalist regimes, or dominant-party regimes.

-   **Dominant-Party Regimes:** These regimes boast the greatest stability due to
    their institutionalized structure. A dominant party, with its shared ideology
    and goals, fosters internal cohesion and a long-term vision. Power resides
    within the party, not with any single individual, and the military aligns
    with the party itself, contributing to greater stability. Formalized
    succession rules further bolster stability by ensuring a smooth transfer of
    power [@frantz2016].

-   **Personalist Regimes:** These regimes exhibit a degree of initial stability
    as dictators, having emerged from intense competition, are typically tough
    and competent. The purging of rivals creates a temporary status quo within
    the dictator's inner circle. However, the lack of a clear succession plan
    creates a vulnerability. The dictator's sudden death can plunge the regime
    into chaos, as potential successors scramble for power, creating a prime
    opportunity for coups.

-   **Military Regimes:** These regimes are often the least stable. Power is
    typically shared among a junta, leading to mistrust and internal conflicts
    over benefits and policies. The absence of a single authority figure hinders
    decisive action, as exemplified by the power struggles within the Chilean
    junta after the 1973 coup [@arriagadaherrera1988].

```{r}
#| label: tbl-regimes1
#| tbl-cap: Main features of different types of regimes 
#| echo: !expr format == "html"

read.csv(here("data/regimes_comparison.csv")) |> 
  mutate(institutionalized_leadership_succession = percent(institutionalized_leadership_succession)) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    4 ~ px(180),
    3 ~ px(80),
    1:2 ~ px(30)
  ) |> 
  tab_source_note(source_note = md("*Source: GWF & Author*"))
```

These contrasting power dynamics significantly influence a regime's
susceptibility to coups. As @tbl-regimes confirms, military regimes, despite
representing only 5.6% of country-years, experience a disproportionate share of
coups (over 22%). Personalist regimes follow a similar pattern, facing a higher
coup risk (23% of coups) despite constituting only 13% of country-years.
Conversely, dominant-party regimes, with their institutionalized structures and
unified leadership, exhibit the greatest resilience. They represent 22.6% of
country-years but experience a lower incidence of coups (only 16.7%).

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and coups since 1950 
#| echo: !expr format == "html"

regimes <- REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949)

regime_total <- regimes |>
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(country_year =n,
         share = percent(percent),
         .keep = "unused") 
  

coup1 <- coup |>
  select(country, ccode, year, coup)

coup_by_regime <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup)) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  mutate(num_of_coups = n,
         percent_of_coups = percent(percent,accuracy = 0.1),
         .keep = "unused") 

coup_by_succeed <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup),
         coup == 2) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  select(regime, succeeded_coup = n) 

coup_regimes <-  coup_by_regime |> 
  left_join(coup_by_succeed) |> 
  mutate(success_rate = percent(succeeded_coup / num_of_coups, accuracy = 0.1))

regime_total |>
  left_join(coup_regimes, join_by(regime_type == regime)) |>
  mutate(regime_type = str_to_title(regime_type)) |>
  select(-succeeded_coup) |> 
  clean_names(case = "title") |>
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN and GIC Datasets*"))
```

> ***H2: Due to their balance of power dynamics, military regimes are more prone
> to coups, followed by personalist regimes, while dominant-party regimes are the
> least likely to experience coups among the three.***

## Research Design

### Double probit with sample selection model

This study employs a sophisticated statistical approach to account for the
selective nature of coup attempts. While coup attempt rates vary across regimes
(as discussed previously), success rates tend to be surprisingly consistent,
hovering around 50% (as shown in @tbl-regimes). This suggests that coup attempts
are not random acts, but rather strategically planned and undertaken only when
the odds of success appear favourable. A standard statistical model would not
account for this selectivity, potentially leading to biased results.

To address this issue, we utilize a two-stage sample selection model, similar to
the approach used by @powell2012. This model has two parts:

-   **Selection Equation (Stage 1):** This stage analyzes the factors influencing
    whether a coup attempt occurs in a particular regime. The primary explanatory
    variable here is regime type, as previously discussed. Additional control
    variables may also be included, denoted by $\mathbf{XB}$.

-   **Outcome Equation (Stage 2):** This stage focuses on the probability of
    success for those coup attempts that actually take place.

The primary explanatory variables are regime types, as previously discussed.
Control variables are included in $\mathbf{XB}$. The selection equation (first
stage) models the probability that a coup attempt occurs and can be expressed as
follows:

$$
\begin{aligned}
y_1^*=\alpha_0 + \alpha_1 Regime_i + \mathbf {XA} + \mu_{1i}
\end{aligned}
$$ {#eq-eq4}

Here, ${y_1}^*$ is an unobserved variable, which may be known to coup plotters.
$Regime_i$ is a categorical variable (*military*, *personalist*, or
*dominant-party*). $\mathbf{XB}$ captures other control variables, such as the
economic crisis index, previous coups, military expenditure, etc.

The observed binary outcome $\mathbf{y_1}$ is:

$$
\begin{aligned}
y_1 = 
\begin{cases} 
1 &\text{if $y_1^*>0$ (coup attempt occurs)} \\
0 &\text{if $y_1^*\le0$ (no coup attempt)}
\end{cases}
\end{aligned}
$$\
In the first stage, if $y_1^*\le0$, no coup attempt occurs in a given
country-year, indicating that the unobserved variable does not reach the
threshold. If $y_1^*>0$, at least one coup attempt is made in a country-year,
indicating that the unobserved variable surpasses the threshold. The probability
is expressed as:

$$
\begin{aligned}
Prob(y_1 =1)&=Prob(y_{1}^*>0) \\
&=\Phi(\alpha_0 + \alpha_1 Regime_i + \mathbf{XA})
\end{aligned}
$$ {#eq-eq4a}

Similarly, the outcome equation (second stage) models the probability that a coup
attempt is successful, given that it occurs:

$$
\begin{aligned}
y_2^*=\beta_0 + \beta_1 Regime_i + \mathbf {XB} + \mu_{2i}
\end{aligned}
$$ {#eq-eq5}

The observed outcome $y_2$ is:

$$
\begin{aligned}
y_2 = 
\begin{cases} 
1 &\text{if $y_2^*>0$ (coup succeeds)} \\
0 &\text{if $y_2^*\le0$ (coup fails)}
\end{cases}
\end{aligned}
$$

The probability equations is:

$$
\begin{aligned}
Prob(y_2=1|y_1 =1)=\Phi(\beta_0 + \beta_1 Regime_i + \mathbf{XB})
\end{aligned}
$$ {#eq-eq5a}

### Variables

#### Dependent variable

Our analysis utilizes data on coup attempts and outcomes from @powell2011. A
successful coup is defined as one where the incumbent leader is removed from
power for more than seven days. The dataset covers the period from 1950 to 2023
and includes information on 491 coup attempts, with roughly half (245) being
successful. Descriptive statistics for these coup attempts and regime types can
be found in @tbl-coups and @tbl-regimes.

#### Key Independent Variable: Regime Type

The core variable of interest is regime type, categorized following the
classification system of @geddes2014 (GWF). We focus on military, personalist,
and dominant-party regimes, with democracies and monarchies included for
comparison. Descriptive statistics for regime types are presented in
@tbl-regimes.

#### Control variables

Our control variables are chosen based on the research of @gassebner2016. They
analyzed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

**Economic Performance:** We measure economic performance using the current-trend
($CT$) ratio developed by @krishnarajan2019. This ratio compares a country's
current GDP per capita to the average GDP per capita over the previous five
years. A higher $CT$ ratio indicates stronger economic performance. We use GDP
per capita data (in constant 2017 international dollars, PPP) from the V-Dem
dataset by @fariss2022, lagged by one year to reflect the prior year's economic
impact. For a country $i$ at year $t$, the $CT$ ratio is calculated as follows:
$$
\begin{aligned}
CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
\end{aligned}
$$

**Political Violence:** We capture overall regime stability by including a
violence index that encompasses all types of internal and interstate wars and
violence. This data comes from the Major Episodes of Political Violence dataset
by Marshall [@marshall2005current].

**Previous coups:** The number of previous coups in a country is included in the
first-stage (selection) model to assess its influence on the likelihood of a coup
attempt. However, it is excluded from the second-stage model (outcome) because
the number of past coups may not directly impact the outcome of a specific coup
attempt.

## Results and Discussion

### Interpretation and Discussion

```{r}
#| label: regression
#| results: asis
#| echo: false

political_violence <- read_excel(here("data/MEPVv2018.xls")) |> 
  dplyr::select(ccode, year, violence = actotal)

gdp <- vdem |> 
  dplyr::select(ccode = COWcode, year, gdppc = e_gdppc) |> 
  filter(year > 1944) |> 
  mutate(ct = gdppc / lag(rollapply(gdppc, width = 5, FUN = mean,
                               align = 'right', fill = NA)),
         .by = c(ccode)) 

coup_all <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_ccode_year.txt", sep = "\t") |> 
  mutate(coup = case_when(
    coup1 == 0 ~ 0,
    coup1 == 2 | coup2 ==2 | coup3 == 2 | coup4 == 2 ~ 2,
    TRUE ~ 1
  )) |>
  mutate(pre_coups = cumsum(lag(coup)  %in% c(1,2))) |> 
  dplyr::select(ccode, year, coup, pre_coups)

reg_data <- regimes |> 
  dplyr::select(country = extended_country_name,
         ccode = reign_cowcode, year, regime_type) |>
  left_join(coup_all, by = join_by(ccode, year)) |> 
  left_join(political_violence, by = join_by(ccode, year)) |>
  left_join(gdp, by = join_by(ccode, year)) |> 
  filter(!is.na(gdppc)) |> 
  mutate(violence = na.locf(violence, .by = ccode)) 

model_2step <- reg_data |>
  filter(regime_type != "other") |>
  mutate(regime = relevel(factor(regime_type), "dominant-party")) |>
  selection(coup > 0 ~ regime + ct + gdppc + violence + pre_coups,
            coup > 1 ~ regime + ct + gdppc + violence,
            data = _) 

model_selection <- model_2step
model_selection$param$index$betaO <- model_2step$param$index$betaS
model_selection$param$index$betaS <- model_2step$param$index$betaO

stargazer(model_selection, model_2step,
          header = F,
          column.labels = c("Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          title = "Sample Selection Model of Regime Types and Coups, 1950-2019",
          label = "results",
          covariate.labels = c("Constant", "Regime: Democracy", "\\hspace{1.6cm}Military", "\\hspace{1.6cm}Monarchy", "\\hspace{1.6cm}Personalist", "Economic trend", " GDP per capita", "Political violence",  "Previous coups"),
          dep.var.caption = "",
          dep.var.labels.include = F,
          summary.logical = T,
          column.sep.width = "50pt"
          )
```

The double `probit` model with sample selection, estimated using the
`sampleSelection` package in R, offers valuable insights into the factors
influencing coup attempts and their success rates across different regime types
(Table \ref{results}).

As expected, military and personalist regimes are significantly more likely to
experience coup attempts compared to dominant-party regimes (all coefficients
positive and significant at the 1% level). This aligns with our theoretical
expectations regarding the internal power struggles within military juntas and
the vulnerability of personalist regimes during succession crises. These findings
highlight the importance of regime structure in understanding coup likelihood.

The control variables show the expected effects. Stronger economic performance
(higher economic trend and GDP per capita) is associated with a lower risk of
coups. Interestingly, the number of previous coups has a negative coefficient,
suggesting a possible deterrent effect. However, this finding is not
substantially significant. Overall, even after considering these factors, regime
type remains a significant determinant of coup attempts, underscoring the
robustness of the model.

Most coefficients in the coup outcome equation are not statistically significant.
This supports the hypothesis that coup attempts are strategically planned and
undertaken only when the perceived chance of success is high. The selection
process (whether a coup attempt occurs) seems to play a more critical role than
these variables in determining the outcome. The negative and significant
coefficient for GDP per capita suggests that stronger economies may bolster
support for incumbents, making successful coups less likely.

The correlation coefficient ($\rho$) between the error terms is positive (0.709)
but not statistically significant (p-value = 0.445). While a positive correlation
might suggest that factors increasing coup attempts also increasing their
success, the lack of statistical significance weakens this connection. This
finding implies that selection bias, while present, may not be as strong as
initially anticipated.

The results strongly support the choice of the sample selection model.
Significant coefficients with theoretically consistent directions suggest the
model effectively captures key aspects of coup dynamics. Regime types with weaker
institutional structures are more vulnerable to coup attempts, while better
economic conditions make coups less likely overall. The model effectively
addresses the non-random nature of coup attempts by treating selection and
outcome as separate processes.

The observed disparity between coup attempt rates and success rates across
regimes points towards selection bias, further validating the use of the sample
selection model. This model acknowledges that coups are not random events, but
rather strategic actions undertaken when the odds appear favourable.

To summary, the double `probit` model with sample selection proves to be a
well-suited approach for this research. It provides robust insights into the
factors influencing both the likelihood of coup attempts and their success rates
across different regime types. The findings highlight the crucial role of regime
structure and the selective nature of coup attempts, supporting the theoretical
framework and empirical strategy employed in this study.

## Conclusion

Motivated by the lack of consensus despite numerous empirical studies on the
determinants of coups, this study introduces a novel approach that prioritizes
determinants based on their impact on coup success. By analysing coup success
rates, the study hypothesizes that the expected outcomes of coups are critical
determinants of their occurrence. Utilizing a double `probit` model with sample
selection, I investigate and confirm the relationship between regime types and
coup attempts.

The findings suggest that regime type plays a significant role in the likelihood
of coup attempts. Military and personalist regimes, characterized by weaker
institutional frameworks and higher vulnerability during power transitions, are
more susceptible to coups. This underscores the importance of supporting
initiatives that strengthen constitutional institutions within these regimes.

The research also finds that stronger economic performance is associated with a
lower risk of coups, suggesting that policies promoting economic development can
be effective in reducing coup risk.

The study shows that the most efficient coup-proofing strategies involve the
establishment of strong institutions. In contrast, purges, random shifting of
military officers, or increased military expenditures are less effective.
However, few autocratic leaders, particularly dictators or military juntas, are
willing to institutionalize their regimes, as such reforms may constrain their
power or shorten their terms. While institutions benefit the regime, they do not
necessarily benefit the leaders themselves.

Future research could explore specific institutional reforms that are most
effective in improving stability across different regimes.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: <https://www.uky.edu/~clthyn2/coup_data/home.htm>, accessed on
    `r Sys.Date()`

[^2]: In the case of the Ghanaian coup, flight lieutenant Jerry John Rawlings
    narrowly avoided execution after his initial failure, being freed by mutinous
    soldiers. Three weeks later, following Rawlings' successful overthrow of the
    government, the deposed leader, General Fred Akuffo, was executed along with
    many other senior members of his government.

[^3]: Niger: Attack on presidential palace an 'attempted coup'. Source: Al
    Jazeera. Retrieved from
    <https://www.aljazeera.com/news/2021/3/31/heavy-gunfire-heard-near-nigers-presidency.>
    Accessed on `r Sys.Date()`.
