---
title: "Prospective Impact of Expected Coup Outcomes on Coup Attempts: A Selection Mechanism Analysis"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf:
    pdf-engine: pdflatex
  docx: default
  aps-revealjs:
    output-file: coups1.html
categories: [Article, Coups]
bibliography: references.bib
abstract: |
  A substantial body of research has examined coups, with much of it focusing on the factors that lead to coup attempts. However, consensus remains elusive regarding why coups are more prevalent in certain countries while less so in others. Previous scholarship exploring the determinants of coup attempts has often overlooked the crucial aspect of coup success. Given the severe consequences of a failed coup, coup plotters are unlikely to proceed unless they perceive a high chance of success. Thus, the expected outcome of a coup—whether successful or unsuccessful—is not merely incidental but serves as a pivotal determinant of coup attempts. The decision to stage a coup is a self-selected variable contingent upon the anticipated success rate of coups. This study employs a sample selection model (specifically, a two-stage probit model) to elucidate why coups are more common in some autocratic countries but rare in others. I contend that coup attempts are largely shaped by the likelihood of coup success, which, in turn, hinges on the power dynamics between coup perpetrators and incumbents. These power dynamics are influenced by the regime type and their distinct responses to internal and external shocks.
keywords: [Coup, Autocracy, Regime types, Sample selection]
---

```{r}
#| label: setup
#| message: false
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(democracyData)
```

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(coup$country)
succeeded <- length(coup$country[coup$coup == 2])

coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

## Introduction

Coups occur with varying frequency across different countries, with some experiencing them more frequently than others.
According to the Global Instances of Coups (GIC)[^1] dataset [@powell2011], Latin American countries such as Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups between $1950$ and $1984$, while Argentina experienced `r coups |> filter(country == "Argentina") |> pull(2)` during a similar time frame.
However, Mexico's authoritarian period from $1917$ to $2000$ saw no coups at all.
In Africa, Sudan endured `r coups |> filter(country == "Sudan") |> pull(2)` coups between $1955$ and $2023$, whereas South Africa has not experienced any coup since $1950$.
Similar patterns are observed in the Middle East and South Asia.
The question of why coups occur more frequently in certain regions, countries, and periods, while being less common in others, has captivated scholars for decades.
Consequently, scholars have delved into extensive research on coups.
Despite numerous efforts in past studies to shed light on these disparities, a definitive model or set of determinants for analysing coups remains elusive.
As highlighted by @gassebner2016, although approximately one hundred potential determinants of coups have been suggested, the fundamental question remains unanswered.

[^1]: <https://www.uky.edu/~clthyn2/coup_data/home.htm>, accessed on `r Sys.Date()`

However, despite extensive research by scholars, the impact of anticipated outcomes on coup initiation has not received sufficient attention.
When analysing the determinants of coups, it's crucial not to overlook the most significant characteristic of coups themselves.
As noted by @powell2011, coups are "illegal and overt attempts by the military or other elites within the state apparatus to unseat the sitting executive."[@powell2011, P.252].
Due to their illegality, the consequences of a failed coup could be severe, with perpetrators risking imprisonment, exile, or even death.
In some instances, the repercussions extend to the families of the coup perpetrators.

Despite the significant risks associated with coups, as shown in @tbl-coups since $1950$, there have still been as many as `r attempts` coups worldwide.
Furthermore, more importantly, half of these coups have been successful.
At first glance, coups seem to be a high-success-rate, high-reward political venture and speculation.
However, compared to over $12,000$ country-years since $1950$, the occurrence of `r attempts` coups appears rather rare, accounting for less than $4\%$.

The low occurrence rate and high success rate indicate that the initiation of coups is highly selective.
In other words, the likelihood of a coup occurring depends greatly on its potential success rate.
Coup plotters carefully assess their chances before staging a coup.
If they decide to proceed, it suggests that the conditions are relatively ripe, hence the fifty-fifty success rate.
Plans with immature conditions and low chances of success are automatically filtered out.
Fundamentally, launching a coup is not like participating in a general election, where failure in one election allows for another attempt in the next election.
A coup is a high-stakes gamble where success brings substantial rewards, while failure may result in martyrdom.
Failure, put differently, means permanent and absolute defeat.
Furthermore, the outcomes of coups are typically decided within a matter of days, sometimes even mere hours.
Unlike prolonged conflicts where victories or defeats may shift over time, the success of a coup is often sealed at its inception.
Every detail must be meticulously planned and arranged beforehand.

\clearpage

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    #table.width = pct(90),
    column_labels.font.weight = "bold",
    quarto.use_bootstrap = TRUE
              ) |> 
  cols_width(1 ~ px(120),
             2:4 ~ px(88)
  ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

Hence, the factors influencing the success rates of coups play a significant role in shaping coup attempts.
This study employs a sample selection model to examine the factors affecting the success rates of coups and, consequently, the likelihood of coup attempts.
I posit that the power dynamics among coup perpetrators, incumbents, and other ruling elites are pivotal in determining the success of coups.
These dynamics are largely contingent on regime types, highlighting the pivotal role of regime types in shaping coup attempts.

This study offers two potential contributions to the existing literature: firstly, it highlights the significance of power dynamics within various regime types as critical factors influencing coup attempts; secondly, it demonstrates how sample selection models can enhance our understanding of coup attempts by taking into account the success rates of coups.

The subsequent section of this paper delves into previous research on coups.
Following that, in Part $3$, I present the research framework and propose hypotheses.
Part 4 provides insights into the data and variables utilized in the study.
The testing results are discussed in Part $5$, followed by the conclusion in Part $6$.

## Dynamics of coup attempts and outcomes

When contemplating a coup, plotters grapple with two crucial factors that also occupy coup scholars: ***disposition*** (why they take the risk) and ***capability*** (whether they can succeed).

### Disposition: Motivations for Coups

Coup motivations can be categorized into three main types:

**Personal Ambition:** The allure of absolute power, prestige, and wealth for themselves, family, and close associates is a significant motivator.
@wintrobe2019 distinguishes between totalitarian and tinpot dictators based on their use of power.
Totalitarian leaders control every aspect of life, relishing their authority.
Tinpot leaders, however, prioritize personal enrichment, indulging in extravagant lifestyles.


**Purported National Interest:** Sometimes, coups are justified as rescuing a nation in crisis, upholding the constitution, or facilitating a democratic transition.
However, scepticism is warranted, as self-serving plotters often use such rhetoric.
Nevertheless, legitimate cases exist.
In 2010, a coup in Niger ousted President Tandja, who attempted an unconstitutional third term.
His actions, including dissolving the opposing Constitutional Court and calling a self-serving referendum, triggered a coup to protect the constitution [@ginsburg2019].


**Self-Preservation:** In rare instances, coups serve as a last resort against imminent political persecution.
Coup leaders might not seek power but fear elimination by the incumbent leader.
Consequently, the coup becomes a pre-emptive strike rather than an ambitious power grab.
An example is Idi Amin, who, in 1971, staged a coup against Ugandan President Obote, who was attempting to remove Amin from his position as army commander-in-chief [@sudduth2017].


While these motivations can arise in any regime, autocracies are more susceptible, particularly for the latter two types.
Stable democracies rarely face such constitutional crises or political persecutions that necessitate coups to protect them.
However, new democracies can be vulnerable to instability, economic downturns, and democratic backsliding which coup plotters can then use as a pretext to justify their actions.

Despite these potential motivations, coups are relatively uncommon, occurring in only $4\%$ of country-years since 1950.
This is because even the most determined coup plotters require the capability to succeed, which we will explore in the next section.
No rational actor attempts a guaranteed failure.

### **Capability**

When coup plotters conspire to stage a coup, their foremost consideration is the likelihood of success.
They might have many incentives to launch a coup, for instance, to fulfil their political ambitions, to implement policies that they prefer, to protect themselves from purge, or to prevent the destruction of the regime, but they are less likely to put the plot into practice if the chance of success is low.
Rational coup plotters would not risk their lives to stage a low successful rate coup.
They either abort their scheme or wait until the chance of success is high enough.
Therefore, when perpetrators of a coup decide to take an action, they must have calculated their chance of success, and evaluated their cost and revenue of
the coup.
They believe their chance of success is high and expect to gain profit from the successful coup.
As long as they succeed, whether the coup is illegal or not would not be a problem at all.
They can declare that their action is just and is in the interest of the people and country.

However, although there may be numerous ambitious political figures aspiring to seize supreme power for various reasons, only a select few have the opportunity to orchestrate a successful coup.
There are many apparent obstacles in coup plotters side compared to the incumbent leaders.

First, as we have mentioned, coups are illegal actions against the incumbent leaders.
Therefore, it has to be discussed secretly in a very small core group.
That is to say, they cannot openly recruit supporters.
While on the other hand, the incumbent leaders could easily and openly take some coup-proving strategies.

Secondly, they are not certain how the other ruling groups would react to their action, especially those forces that are powerful enough to change the situation.
However, the incumbent leaders know well the balance of power and take actions to adjust the balance in favour of their interests.
That is to say, although the current leaders do not know who is likely to betray them, they know who has the ability to betray them, and thus adjust and change this ability according to the situation.

Thirdly, coup plotters face a serious loyalty, or an unreliable commitment problem.
Coup participants face great risks, and their expected benefits are highly uncertain.
This is not only because of the risk of a failed coup, but also because even if the coup succeeds, it is still unknown whether the coup leaders will deliver on their previous promises.
Moreover, after the coup leaders succeed, they often purge a group of capable assistants to eliminate the risk of being overthrown by another coup.
In this case, betraying the coup and joining the current leader appears to be relatively less risky and the benefits are stable and reliable.

It is hard to know the threshold chance of success, because different coup plotters might have different risk preferences.
However, we can get a rough clue from the coups that are staged, which are expected to succeed by perpetrators.
The
data of previous coups show that the chances of success of staged coups are quite
satisfactory, higher than 40% except in democracies ( Table 1 ).
The latest coups
staged in 2021, with 5 successful coups out of 7 staged coups, even have higher
successful rate.

To understand how the perpetrators
of coups evaluate the cost and benefit of coups, we can explain the mechanism of
coups through a typical model.

As previously mentioned,

According to @eq-eq3, if the probability of success, $\sigma$, exceeds $0.5$, the equation holds true.
This assertion finds some support in empirical data on coups since $1950$.
As indicated in @tbl-coups, the overall success rate of coups stands at $49.9\%$.
Although this figure falls just below the threshold of $0.5$, it is important to acknowledge that this is an average rate calculated post-coups.
In reality, coup plotters typically require more certainty about success than a mere fifty-fifty chance.
Furthermore, outliers do exist, such as irrational political careerists or coup attempts driven more by self-protection instincts than strategic planning.
In such cases, coup perpetrators may not prioritize or even consider the potential outcomes, as the consequences of being purged are no different from those of a failed coup.

Given the pivotal role of coup success in determining coup attempts, the factors influencing coup success inevitably become the determinants of coup attempts themselves.
While the specific determinants may vary, the fundamental factor lies in the balance of power between the coup perpetrators and the incumbent leaders.
I will delve into the factors that could potentially tip the scales of power among different political factions.

### Basic framework

An oft-cited framework for assessing the disposition and opportunity of coup attempts is formalized by @leon2013 to evaluate the anticipated benefits for coup plotters.
This framework can be outlined as follows:

$$
\begin{aligned}
\text{Expected Payoff of coups} = \sigma V_S + (1 - \sigma)  (V_I - C)
\end{aligned}
$$ {#eq-eq1}

Here, $V_S$ represents the return of a successful coup, $V_I$ denotes the normal value paid by the incumbent in the status quo, $C$ signifies the cost of the failed coup, and $\sigma$ represents the probability of coup success.
By rearranging the equation, we can derive:

$$
\begin{aligned}
\sigma(V_S - V_I) > (1-\sigma)C
\end{aligned}
$$ {#eq-eq2}

Drawing from these equations, @aidt2019 suggests that the decision to initiate a coup hinges on whether the expected benefits of a successful coup outweigh the potential costs of failure.
Both the disposition to attempt a coup and the opportunity to stage a successful coup are encompassed within the equation.
The disposition is captured in the gain from a successful coup, denoted as $V_S-V_I$.
Meanwhile, the opportunity is represented by the probability of a successful coup, denoted as $\sigma$.
Although this logical framework seems plausible, it immediately poses practical challenges in quantifying the values of $V_S$ and $C$, which are not easily measurable in real-world terms.
How can one accurately measure the cost of losing freedom, life, or even beloved family members if a coup fails?
Similarly, how can the value of assuming leadership of a country be quantified if the coup succeeds?
These are not commodities that can be exchanged in a market.

While $V_S$ and $C$ may be impossible to measure in precise terms, this does not render the equations useless.
Firstly, the underlying logic of the equations remains robust and provides a valuable framework for understanding the decision-making process of coup plotters.
Secondly, there is no need for precise measurements of $V_S$ and $C$.
It is sufficient to consider them as invaluable, with both potentially considered roughly equal.
Additionally, coup plotters are driven by factors beyond mere economic gains or losses.
As long as the coup succeeds, they win anyway.
This suggests that in @eq-eq2, the values of $V_S-V_I$ and $C$ can be disregarded, and the focus can solely be on the probability of the coup's success, $\sigma$:

$$
\begin{aligned}
\sigma > (1-\sigma)
\end{aligned}
$$ {#eq-eq3}

### Regime types and balance of power

When discussing the balance of power, the first aspect that comes to mind is military strength.
It's evident that control over the military gives individuals the upper hand in coup attempts.
This explains why military coups are often at the forefront of political science discussions.
In much of the literature on coups, the terms "coup" and "military coup" are used interchangeably, with little exploration into why this is the case.
Consequently, more attention is directed towards preventing military involvement in politics.
Strategies such as "*Keeping the military happy*" [@aidt2019, P. 15] or "*giving them toys*" [@huntington1991democratization, P. 252] have been proposed to mitigate military intervention.
Many coup-proofing strategies, informed by empirical research, aim to either diminish the military's inclination to stage coups or create obstacles to their success.
For example, studies by @leon2013a suggest that nations with lower military spending as a percentage of GDP are more susceptible to coups.
Similarly, @powell2018 argues that increased military expenditures may reduce the likelihood of coups among military factions.

However, previous studies have often oversimplified the intricate balance of power within military forces.
They have frequently assumed, without substantial evidence, that the military operates as a monolithic entity, uniformly and decisively.
Such a notion fails to capture the complexities of real-world dynamics.

It is crucial to acknowledge that the military is not always a unified entity.
Regardless of its size, any military force is composed of various groups or factions, each with its own chain of command.
Within these factions, mutual suspicion, competition, and vigilance are common, while moments of unity are rare.
Due to the clandestine nature of coups and the need for secrecy, coup attempts are often orchestrated within small, tight-knit groups.
Coup plotters are uncertain about the stances and intentions of other factions beforehand, and they are particularly worried that once the coup is initiated, other factions may not only refuse to support it but also actively oppose, intervene in, or even suppress it.
Therefore, the success of a coup heavily depends on the reactions of other military factions [@geddes1999].

Furthermore, the structure of military forces, especially their relationships with other state departments, particularly civilian officials, varies across different regime types.
In democratic countries, the military is nationalized and bound by constitutional authority rather than individual leaders.
For instance, the United States Armed Forces are under the command of the President but remain loyal to the Constitution, refraining from intervening in politics.
In military regimes, power is often centralized in the hands of top generals or a junta, while in personal dictatorships, the leader retains ultimate authority over the military.
In dominant-party regimes, the military serves the interests of the ruling party.
Thus, the balance of power within military forces is deeply influenced by the nature of the regime in question.

To gain a thorough understanding of military power dynamics across diverse political landscapes, it is imperative to explore their intricate nuances.
This section adopts the foundational concepts articulated by @geddes2014 (GWF), which categorize regimes based on the origins of their leadership and the factors influencing their policy decisions.
These regimes are broadly classified into four categories: military regimes, personalist regimes (including monarchies), dominant-party regimes, and democratic regimes.

**Military regimes**, as defined by GWF, are characterized by the dominance of a junta composed of military officers who exercise control over the regime's power structure, including leadership selection and policy formulation.
Examples of such regimes include the Brazilian regime (1964–1985), the Argentine regime (1976–1983), and the Salvadoran regime (1948–1984) [@geddes1999].
It's noteworthy that within military regimes, political parties may persist, but they typically serve as conduits for the military junta or align with the directives of military officers to avoid being categorized as dominant-party regimes.
Additionally, while a junta may appoint a senior officer as the executive or enlist civilian administrators and technocrats as intermediaries, their authority over other officers remains restricted, distinguishing military regimes from personalist ones.

In a typical **personalist regime**, supreme power rests with the dictator, encompassing policy making, control over the military, and the authority to nominate top officials and successors.
Personalist rules often emerge under charismatic leaders, particularly founding fathers of newly independent nations.
Regimes such as Rafael Trujillo’s in the Dominican Republic (1930–1961), Idi Amin’s in Uganda (1971–1979), and Jean-Bédel Bokassa’s in the Central African Republic (1966–1979) serve as illustrative examples of personalist rule [@geddes1999].
In such regimes, either political parties are absent or they are established by or subservient to the dictator.
Importantly, the dictator may or may not be a military officer themselves, but the military is invariably under the dictator's control.

In a typical **dominant-party regime**, supreme power lies within the party, with the incumbent serving as its representative and subject to collective party leadership.
Examples of dominant-party regimes include the Partido Revolucionario Institucional (PRI) in Mexico, the Revolutionary Party of Tanzania (CCM), and Leninist parties in various East European countries [@geddes1999].
The party functions as a well-organized and highly disciplined entity, with its own ideology and political agenda to unite and mobilize its members and supporters.
While powerful leaders may emerge within such regimes, such as Stalin in the Soviet Union (1924-1953 and Mao Zedong in China (1949-1976), they do not wield enough power to supplant or replace the party with a new one.

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and their distribution since 1950 
#| echo: !expr format == "html"

REIGN |> 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant-party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) |> 
  filter(year > 1949) |>  
  summarise(counts = n(),
            .by = regime_type) |>
  arrange(desc(counts)) |>
  mutate(regime_type = str_to_title(regime_type),
         share = counts / sum(counts),
         share = percent(share)) |>
  adorn_totals(where = "row", fill = "100%") |>
  clean_names(case = "title") |> 
  gt() |>
  tab_options(
    column_labels.font.weight = "bold",
    table.width = pct(90)
              ) |>
  # cols_width(1 ~ px(200),
  #            2:3 ~ px(100)
  # ) |>
  tab_source_note(source_note = md("*Source: REIGN dataset*"))
```

## Research Framework and Hypotheses

## Data and Variables

## Results and Discussion

## Conclusion

```{r}
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
