---
title: "Leadership Transitions and Survival:\n Coups, Autocoups, and Power Dynamics"

author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk
  
format:
  aps-html: default
  aps-pdf: 
    documentclass: report
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    toc: true
    toc-title: Contents
    lof: true
    lot: true
    crossref: 
      lof-title: Figures
      lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  docx: default
    
categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
#include-before-body: title.tex
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

pacman::p_load(tidyverse, data.table, gt, gtsummary, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects)
```

## Acknowledgements {.unnumbered}

The completion of this thesis has been a significant journey, filled with hard
work, learning, and moments of joy. Throughout this time, I have received support
and encouragement from many individuals, without whom this dissertation would not
have been possible.

First and foremost, I would like to express my deepest gratitude to my great
supervisor, Professor Kristian Skrede Gleditsch, for his invaluable guidance,
unwavering support, and insightful feedback throughout this journey. His
expertise and encouragement have been instrumental in shaping this dissertation.
I would also like to extend my heartfelt thanks to the chair of my board panel,
Professor Han Dorussen, for his continuous support and constructive criticism,
which have significantly enhanced the quality of my research.

I am profoundly grateful for the comments, advice, and suggestions from several
esteemed scholars who have contributed to this work. Dr. Brian J Phillips, Dr.
Prabin Khadka, and Dr. Winnie Xia, their expertise and thoughtful input have been
greatly appreciated and have enriched this dissertation.

Finally, I want to thank my family for their unwavering support and love. To my
beloved wife, Ji Zhi, who has been my rock throughout this journey, and to my
dear daughter, Siyan, and son, Sisheng, who have been my source of joy and
motivation. I am deeply thankful to my father for his enduring support, and to
the memory of my late mother, whose love and guidance continue to inspire me.

All errors and faults are my own.

## Abstract {.unnumbered}

This dissertation examines the dynamics of irregular power transitions,
particularly coups and autocoups, and their influence on leader survival. It
highlights the critical role of power dynamics, shaped by **regime type**, in
determining coup success rates and attempt frequency. Utilizing a **double probit
model with sample selection**, the study reveals that expected coup success
significantly influences attempts, with military regimes facing a heightened
vulnerability due to their power structure.

While often understudied, autocoups are shown to have a substantial impact on
democratic trends. This research introduces a refined definition of autocoups
alongside a novel dataset encompassing events from 1945 to 2022, enabling a more
robust quantitative analysis.

Employing survival analysis, the study compares the longevity of leaders who rise
to power through coups versus autocoups. The findings demonstrate that
coup-installed leaders face a significantly shorter tenure and higher risk of
removal. This contrasts with autocoup leaders who manipulate the system to extend
their rule, suggesting the potential for autocoups to incentivize power grabs and
contribute to democratic backsliding.

This work contributes significantly to the political science literature by:

-   Defining key concepts: It establishes a clear definition of autocoups, a
    previously understudied phenomenon.

-   Introducing a novel dataset: This dataset enables researchers to conduct more
    comprehensive quantitative analyses of autocoups.

-   Establishing a general framework: The framework provides a comparative
    approach to studying the dynamics of irregular power transitions and their
    impact on democratic stability.

**keywords:** *Coups, Autocoups, Power transitions, Leadership Survival*

## Introduction

### Research question

Irregular power transitions, marked by a disregard for constitutional procedures,
are a critical area of study in political science. They not only disrupt
established rules but often require unconstitutional tactics to secure power.
Furthermore, these transitions can inspire copycat behaviour among other
ambitious leaders.

Despite their central role in political science and the extensive research
conducted on irregular power transitions, a long-standing question continues to
intrigue political scientists: ***Why are some leaders ousted before their terms
expire, while others complete their full terms or even overstay beyond their
originally mandated limits?*** In other words, why do some leaders survive for
decades while others last for only years, months, or even days? This dissertation
focuses on this question and seeks to provide a comprehensive analysis, dedicated
to understanding how leaders come to power through unconstitutional means and
what factors determine the duration of a leader's rule following an irregular
ascent.

### Analyses on coups and autocoups in a general framework

When discussing irregular power transitions, the concepts that often come to mind
are irregular entries or exits, such as coups, assassinations, rebellions,
protests, and foreign interventions. Among these methods, coups hold a prominent
position due to their frequent occurrence. According to the Archigos dataset
[@goemans2009], from 1945 to 2015, there were approximately 145 instances of
irregular leader exits, with coups[^1] accounting for more than half (79
leaders). The often-cited Global Instances of Coups (GIC)[^2] dataset
[@powell2011] records even more leaders (245 cases) removed by coups from 1950 to
2023.

Given their prevalence and substantial influence on political systems, coups have
been extensively studied, particularly since 2000 [@thyne2019]. Consequently, the
concept of a coup is comparatively clear and widely accepted in academic circles.
Many scholars, including this study, follow the definition by @powell2011, which
describes coups as "illegal and overt attempts by the military or other elites
within the state apparatus to unseat the sitting executive... \[a coup is
successful\] if the perpetrators seize and hold power for at least seven days"
(P. 252). Although debates persist, two elements are clear: first, the
perpetrators are elites within the ruling group, and the victims of coups are
incumbent executive leaders. Second, the strategy or aim of a coup involves
completely removing the incumbents, not merely seizing part of their power or
forcing them to concede on specific policies. Beyond defining coups, several
datasets have been developed for quantitative analyses, such as the Global
Instances of Coups [@powell2011], the Cline Centre Coup d’État Project Dataset
[@peyton2024], and the Colpus Dataset [@chin2021]. These datasets are
well-developed and frequently used in political science research.

However, irregular power transitions are not limited to irregular entries and
exits but should also include irregular "overstays." Using illegal means to
overthrow an incumbent leader before their term expires is undoubtedly an
irregular power transition. Similarly, an incumbent using illegitimate means to
extend their term beyond term limits is also an irregular power transition.

Although academic attention to irregular retention of power has increased since
the 1990s, especially after Peru’s President Alberto Fujimori's self-coup in
1992, it remains comparatively understudied and has several shortcomings. First,
there is no universally accepted terminology for this "overstaying in power" type
of irregular power transition, unlike the clear term "coup." Consequently,
various terms such as self-coup, autogolpe, and executive coup are used by
different scholars. This dissertation will use 'autocoup' to refer to this type
of irregular power transition, which will be thoroughly discussed in Chapter 3.
Second, there is no consensus on the definition of an autocoup. Existing
definitions remain vague, often conflating power expansions and power
extensions[^3]. For example, @cameron1998a defines an autogolpe as a temporary
suspension of the constitution and dissolution of Congress by the executive, who
then rules by decree. This definition focuses on power expansion instead of power
extension, leading to conceptual confusion and misalignment with the definition
of a classic coup. Third, a consensus autocoup dataset is lacking. While several
related datasets exist, as discussed by @baturo2022 in coding their Incumbent
Takeover dataset, the terminologies, definitions, and coverage years vary,
lacking wide acknowledgement and extensive academic exploration. In summary,
autocoup has not been analysed in a comparative manner connected with coups.

Analysing coups and autocoups separately is less problematic. However, from a
comprehensive framework perspective on irregular power transitions and leader
survival, coups and autocoups should be, and can be, analysed within the same
framework. Both coup and autocoup significantly influence democratic backsliding
and are the most frequent means of irregular power transition. Furthermore, as
both are called "coups," classic coups and autocoups are very similar since a
coup is launched to replace the current leader, while an autocoup is staged to
replace the future leader.

### Academic Contributions

This study addresses a critical gap in the literature by offering a comprehensive
framework for analysing both coups and autocoups, which are the most common forms
of irregular power transitions. While existing research often examines these
topics separately with varying terminologies, definitions, methods, and datasets,
this dissertation integrates these elements to provide a unified perspective on
irregular power transitions and leader survival.

Our contributions are threefold:

-   **Emphasis on power dynamics and regime types**: We highlight the significant
    role of power dynamics, particularly the influence of regime types, in
    determining the success and frequency of coup attempts. Our analysis
    underscores how the expected chances of coup success motivate such attempts,
    with military regimes being notably susceptible.

-   **Refined definition and novel dataset for autocoups:** We introduce a
    refined definition of autocoups and develop a novel dataset covering events
    from 1945 to 2022. This enables a comparative analysis with classic coups,
    providing clearer insights into the nature and impact of autocoups on
    political systems.

-   **Survival analysis of leaders from different entry modes:** By applying
    survival analysis to existing coup data and our new autocoup dataset, we
    demonstrate how different modes of entry into power significantly affect
    leader survival. Our findings reveal that leaders who come to power through
    coups typically have shorter tenures and face higher removal risks compared
    to those who extend their rule through autocoups.

Our analysis of irregular power transitions is particularly relevant to
understanding democratic backsliding. These transitions violate democratic norms
and disrupt the path towards stable democracy. Leaders who gain power through
irregular means often employ undemocratic tactics, such as suppressing
opposition, to consolidate their illegitimate hold on power. This creates a
vicious cycle where the erosion of democratic institutions is both a cause and
consequence of efforts to maintain power.

### Overview of the thesis

This study is structured into three main chapters beyond the introduction, each
addressing key aspects of irregular power transitions and their implications for
political stability and democratic processes.

**Chapter 2** examines the determinants of classic coup attempts. While extensive
research exists on coups, most studies focus on observable factors before coups,
such as economic performance, political stability, previous coups, and
coup-proofing strategies. This chapter, however, emphasizes the less observable
but crucial factor of expected chances of coup success, which have been often
overlooked. Utilizing the double probit model with sample selection, the analysis
reveals that expected success rates significantly influence coup attempts. These
success rates are primarily shaped by the balance of power between incumbents and
challengers, which is largely determined by regime type. The findings indicate
that military regimes face a much higher risk of coups compared to dominant-party
regimes.

**Chapter 3** focuses on the concept of autocoups, specifically on power
extensions by incumbent leaders. It distinguishes autocoups from broader concepts
like self-coups or executive coups by redefining them as instances where
incumbent leaders refuse to transition power as mandated, thereby overstaying in
office. Based on this refined definition, a novel dataset of autocoup events from
1945 to 2022 is introduced, encompassing 110 attempts and 87 successes. The
chapter includes case studies and empirical analyses that demonstrate the utility
of this dataset for quantitative research, providing a basis for empirical
analysis on autocoups.

**Chapter 4** investigates how the method of power acquisition impacts the
longevity of leaders who come to power through coups versus those who extend
their rule through autocoups. The hypothesis is that the method of accession
significantly affects leader tenure. Using the Cox proportional hazards model and
a time-dependent Cox model, the chapter provides evidence of differing survival
times between these two types of leaders. The results indicate that leaders who
come to power through coups face a significantly higher risk of removal compared
to those who extend their rule through autocoups. This finding highlights the
implications for political stability and democratic processes, suggesting that
the relatively low cost and high returns of autocoups could incentivize
incumbents to seize power in this manner, potentially leading to democratic
backsliding and the personalization of power.

In **Chapter 5**, the study concludes by summarizing the main findings,
discussing policy implications, and acknowledging the limitations of the
research. It also outlines directions for future research, emphasizing the need
for further exploration of irregular power transitions, particularly coups and
autocoups.

## Power Dynamics and Coup Attempts: A Selection Mechanism Analysis

### Introduction

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data

Coup <- fread("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt")

attempts <- length(Coup$country)
succeeded <- length(Coup$country[Coup$coup == 2])

coups <- Coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )



Coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

Coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

Coups, defined as "illegal and overt attempts by the military or other elites
within the state apparatus to unseat the sitting executive" [@powell2011, P.252],
occur with varying frequency across countries, with some experiencing them more
often than others. According to GIC dataset, Latin American countries such as
Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups
between 1950 and 1984, while Argentina experienced
`r coups |> filter(country == "Argentina") |> pull(2)` during a similar time
frame. However, Mexico's authoritarian period from 1917 to 2000 saw no coups at
all. In Africa, Sudan endured `r coups |> filter(country == "Sudan") |> pull(2)`
coups between 1955 and 2023, whereas South Africa has not experienced any coup
since 1950. Similar patterns are observed in the Middle East and South Asia.

The varying frequency of coup attempts has captivated political scientists for
decades, leading to extensive research on the subject. As highlighted by
@gassebner2016, despite approximately one hundred potential determinants of coups
being suggested, no consensus has been reached. In an effort to address this
issue, they have tested 66 factors proposed in previous literature using three
million model permutations in an extreme bounds analysis (ibid.).

Examining previous research, which has tested around 100 variables as potential
determinants of coups, raises an important question beyond simply understanding
why coups are more frequent in some countries than others. The critical question
is: Can we establish a framework to help scholars focus on the most relevant
factors of coups, rather than sifting through over 100 variables without reaching
a consensus?

Reviewing previously proposed variables of coups, it is evident that all focus on
pre-coup conditions, with no consideration given to post-coup factors. This means
more attention is paid to factors observable before a coup, such as military
force or spending, economic performance, political stability, and previous coups.

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold",
    table.layout = "fixed"
              ) |> 
  # cols_width(1 ~ px(120),
  #            2:4 ~ px(88)
  # ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

However, coups are high-stakes gambles with an all-or-nothing nature. Due to its
illegality, the consequences of a failed coup can be severe, with perpetrators
risking imprisonment, exile, or even death. In some instances, repercussions
extend to their families, which means coup plotters would not take the risk
without some assurance of success.

Historical coup attempts and their success rates provide valuable hints about
coup plotters' decision-making processes. Despite the significant risks
associated with coups, as shown in @tbl-coups, there have been `r attempts` coups
worldwide since 1950. Importantly, about half of these coups have been
successful. At first glance, coups appear to be a high-success-rate political
venture. However, compared to over 12,000 country-years since 1950, the
occurrence of `r attempts` coups is relatively rare, accounting for only about 4%
(GIC).

The low occurrence rate and high success rate clearly indicate that the
initiation of coups is highly selective. In other words, the likelihood of a coup
occurring depends greatly on its probability of success. However, the probability
of success is not a factor that can be observed before a coup, at least not to
outsiders and researchers, but coup plotters surely have more information on the
possible outcome of coups.

Since coup plotters meticulously assess potential outcomes before staging a coup,
coup researchers should also analyse what factors might affect the outcome of
coups. Otherwise, we ran the risk of selection bias. When considering the factors
that most affect the outcomes of coups, the current literature predominantly
identifies military power as the decisive factor in the success of coups.
Ultimately, military power is determined by the power structure within the
regime.

Unlike coup plotters, who have a clearer understanding of power structures and
dynamics, including their supporters, followers, bystanders, and opponents,
researchers are not able to accurately observe this information. However, we can
analyse the most relevant factors which determine the balance of power and are
observable to us. This leads us to **regime type**, which reflects the
distribution of power within a government, encompassing who controls the
military, sets policy, and appoints officials. Since coup plotters consider the
balance of power within the regime before acting, this study argues that regime
type plays a crucial role in shaping coup attempts. Analysing which regime types
are more susceptible to coups can offer valuable insights.

We employ a **double probit model with sample selection** to address the
selection bias. This model allows us to analyze both the factors influencing coup
success and the factors related to the decision to initiate a coup attempt in the
first place.

This study contributes to the literature in two key ways. First, it emphasizes
the importance of focusing on expected success rates as a driver of coup
attempts, offering a more targeted approach compared to past studies. Second, it
highlights the significance of regime type as a factor influencing coup
likelihood, even when researchers lack perfect knowledge of a regime's internal
power dynamics.

The subsequent section (Section 2) explores the dynamics of coup attempts and
their outcomes. Section 3 delves into the research design, outlining the
methodology and variables used in the analysis. Section 4 presents and discusses
the empirical findings. Finally, Section 5 concludes the chapter by summarizing
the key insights and their implications.

### Dynamics of coup attempts and outcomes

Coup attempts are driven by a complex interplay of factors, including motivations
(**disposition**) and the resources and opportunities available to succeed
(**capability**).

#### Motivations for coups

This section focuses on the motivations that compel challengers to undertake
coups. We can categorize coup motivations into three main types:

**Personal Ambition:** Personal ambition is a significant motivator for the
majority of coup plotters, driven by the allure of absolute power, prestige, and
wealth. The prospect of seizing absolute power provides an unparalleled
opportunity to shape national policies, control resources, and make significant
decisions without constraints. The pursuit of prestige and recognition, along
with the potential for immense economic gain and wealth, further incentivizes
individuals. Additionally, the desire to leave a lasting legacy and make a
historical impact can compel individuals to undertake the risky venture of
staging a coup.

**Purported National Interest:** Sometimes coups are justified as necessary
interventions to address national crises, uphold the constitution, or facilitate
a transition to democracy. While the motivations behind such claims require
scrutiny, genuine examples do exist. For instance, the 2010 coup in Niger ousted
President Tandja, who attempted an unconstitutional third term by dissolving the
opposing court and calling a self-serving referendum [@ginsburg2019].

**Self-Preservation:** In some cases, coups serve as pre-emptive strikes against
perceived threats. Coup leaders might not necessarily seek power, but rather fear
elimination or political persecution by the incumbent regime. An example is Idi
Amin's 1971 coup against Ugandan President Obote, who was attempting to remove
Amin from his military command position [@sudduth2017].

These motivations are often most prevalent in autocratic regimes, where
justifications under the guise of national interest or self-preservation can be
used to mask personal agendas. Stable democracies rarely face the same level of
constitutional crises or political persecution that might necessitate a coup.
However, new established democracies can be vulnerable to instability, economic
downturns, and democratic backsliding, creating opportunities for coup plotters
to exploit these weaknesses and justify their actions.

Despite the potential motivations outlined above, coups remain relatively
uncommon, occurring in only about 4% of country-years since 1950. The main reason
is that coup perpetrators face inherent disadvantages compared to incumbent
leaders, highlighting the importance of capability. Even the most motivated
plotters need the resources and opportunities to succeed. The next section will
explore the concept of capability in greater detail.

#### Capability for coups

For coup plotters, the decision to act hinges not only on their motivations but
also on a calculated assessment of their chances of success. Several factors can
influence this threshold:

-   **Military Strength**: A clear advantage in military capabilities compared to
    the incumbent regime significantly increases the odds of a successful coup.

-   **Internal Divisions within the Regime**: Existing fractures within the
    government's power structure can present opportunities for coup plotters to
    exploit weaknesses and gain support from disgruntled factions.

-   **Public Support**: Widespread discontent with the incumbent regime,
    especially within the military or key sectors of society, can create an
    environment ripe for a successful coup.

-   **Foreign Backing**: External support from powerful nations can provide
    resources, legitimacy, and even direct military intervention to tip the
    scales in favour of the coup plotters.

The high success rate of coups since 1950, as shown in @tbl-coups, might suggest
that coups are not particularly difficult to succeed. However, it is important to
consider selection bias. We only observe attempted coups, not the numerous
dispositions and conspiracies that never came to light. Analysing launched coup
data alone can be misleading.

To gain a more comprehensive understanding of coup attempts and their likelihood,
we need to move beyond historical data and employ a theoretical framework that
accounts for selection bias. This framework will allow us to analyze the factors
influencing both the success of coups and the decision to attempt a coup in the
first place. The next section will delve into such a framework and its
implications for our understanding of coup attempts.

#### Framework of coup success

An oft-cited framework [@gassebner2016; @aidt2019] provides a structured approach
to assess the disposition and capability of coup attempts by evaluating the
anticipated benefits for coup plotters. The expected pay-off of coups can be
represented by the equation:

$$
\begin{aligned}
E(U) = p \times B + (1 - p) \times (-C)
\end{aligned}
$$ {#eq-eq1}

Here, $\mathbf B$ represents the return of a successful coup, $\mathbf C$
signifies the cost of a failed coup, and $p$ represents the probability of coup
success. The condition for staging a coup is when the expected benefit is
positive, meaning that the expected pay-off is greater than 0. Rearranging the
equation, we get:

$$
\begin{aligned}
p \times B > (1 - p) \times C
\end{aligned}
$$ {#eq-eq2}

@eq-eq2 implies that for @eq-eq1 to hold, the expected benefits earned from
successful coups must outweigh the expected cost of failed coups.

While seemingly clear, the equation faces practical challenges. Quantifying
$\mathbf B$ (the value of a successful coup) and $\mathbf C$ (the cost of
failure) is difficult. The loss of life, freedom, or loved ones after a failed
coup, as well as the value of assuming leadership after a successful coup, are
intangible concepts that defy precise measurement. As evidenced by the 1979 coup
in Ghana[^4], the fate of the coup leader(s) hangs in the balance; they are high
likely to be killed if the coup fails, or to execute others if the coup succeeds.

However, these challenges do not render the framework useless. Firstly, its core
logic remains valuable, offering insights into how coup plotters might assess the
return and cost of their actions. Secondly, given the significant and elusive
nature of precise values for $\mathbf B$ and $\mathbf C$, they can be treated as
roughly equal. Consequently, there is no need to fret over how to measure and
compare these values precisely. Instead, we can shift our focus from $\mathbf B$
and $\mathbf C$, to the probability of success ($p$), simplifying @eq-eq2 to:

$$
\begin{aligned}
p > (1-p)
\end{aligned}
$$ {#eq-eq3}

@eq-eq3 suggests that, to hold @eq-eq2 true, a success probability greater than
$50\%$ is necessary. Interestingly, empirical data on coups since 1950 somewhat
supports this notion. As shown in @tbl-coups, the overall success rate is
$49.9\%$. While this falls short of the $50\%$ threshold, it's important to
consider two factors. Firstly, this is an average rate, not necessarily
reflective of the probabilities assessed by coup plotters beforehand. Secondly,
outliers such as irrational actors and coups driven by self-preservation may not
prioritize success probabilities. Taking these points into account, we can
propose our first hypothesis:

> ***H1: The fundamental determinant of a coup attempt is the perceived chance of
> success. Coup plotters likely require a success threshold of at least 50%.***

This leads us to the next crucial question: What factors determine coup success,
influencing the very decision to attempt one? We will discuss it in the next
section.

#### Regime types and power dynamics

The high success rate in historical data is merely a post-event statistical
description, not indicative of the probability of any specific coup. Importantly,
coup plotters do not rely on past success rates as their guide; instead, they
assess their chances based on their unique context.

Military strength undeniably plays a critical role in coup attempts. Control of
the armed forces offers a significant advantage, explaining why military coups
dominate discussions on the topic. Much of the literature treats "coup" and
"military coup" interchangeably, with scholars like @powell2011 finding that half
of 14 studies attribute coups solely to the military. Consequently, significant
focus, from both researchers and policy-makers, centres on the balance of power
between civilian and military authorities, or among military factions themselves.
Strategies like "keeping the military content" [@aidt2019] or "providing them
with resources" [@huntington1991democratization] aim to reduce military
intervention. Empirical research informs coup-proofing strategies that either
decrease the military's desire for coups or raise barriers to success
[@leon2013a; @powell2018].

However, while military power is decisive, it complicates the analysis. As
@tbl-regimes demonstrates, military regimes, despite concentrated military
control, exhibit surprising instability, experiencing the most frequent coup
attempts. This highlights a crucial issue: the intra-military component. Treating
the military as a monolithic entity ignores its complex internal dynamics
[@singh2016]. Any military comprises diverse groups with their own hierarchies,
fostering suspicion, competition, and vigilance rather than unity. The
clandestine nature of coups necessitates small, secretive groups. Plotters are
unsure of other factions' stances and fear their opposition or intervention, as
exemplified by the swiftly thwarted 2021 Niger coup[^5]. The success of a coup
hinges heavily on other military factions' reactions [@geddes1999].

Moreover, military force is not the only factor shaping the balance of power.
Internal divisions within the ruling elites, public support, and foreign backing
also play important roles. Since the balance of power is not a clearly observable
variable, especially difficult to measure for outsiders and academic scholars, we
can shift our focus from who controls power in a specific context to what factors
shape the balance of power. This leads us to regime type, as the classification
of regime type is based mainly on the power structure. For example,
"...definition of regimes emphasizes the rules that identify the group from which
leaders can come and determine who influences leadership choice and policy"
[@geddes2014, p. 314]. We will leverage this framework to categorize autocracies
based on leadership origin and decision-making, classifying regimes into three
main categories: military, personalist, and dominant-party.

-   **Military Regimes:** Characterized by the dominance of a junta—a group of
    military officers who control the regime's power structure, including
    leadership selection and policy formulation. Examples include the Brazilian
    regime (1964-1985), the Argentine regime (1976-1983), and the Salvadoran
    regime (1948-1984) [@geddes1999].

-   **Personalist Regimes:** Power resides with a single, charismatic leader who
    controls policy, the military, and succession. Regimes like Rafael Trujillo's
    in the Dominican Republic (1930-1961), Idi Amin's in Uganda (1971-1979), and
    Jean-Bédel Bokassa's in the Central African Republic (1966-1979) exemplify
    personalist rule (ibid.).

-   **Dominant-Party Regimes:** Power rests within a well-organized ruling party,
    with leaders acting as its representatives. The party structure and ideology
    foster internal cohesion and a long-term vision. Examples include the Partido
    Revolucionario Institucional (PRI) in Mexico, the Revolutionary Party of
    Tanzania (CCM), and Leninist parties in various Eastern European countries
    (ibid.).

The critical distinction between regime types lies in the unique power balance
established during their seizure of power, often through irregular and violent
means, frequently involving military confrontations. This power struggle
typically results in the emergence of the most competent group—be it a military
junta, a political party, or a strongman—as the core leadership. Furthermore, the
early stages of new regimes are often marked by internal purges to consolidate
power and eliminate potential rivals [@sudduth2017; @roessler2011].

Following these external challenges and internal purges, new power dynamics
emerge in three primary forms: dominant-party regimes, personalist regimes, and
military regimes. These contrasting power dynamics significantly influence a
regime's susceptibility to coups. Dominant-party regimes, characterized by a
well-organized party with clear ideology, hierarchy, and discipline, exhibit the
greatest resilience against coups due to their institutionalized structures and
unified leadership. Personalist regimes, where power centers around a single
strong leader, are relatively stable during the leader's tenure. However, they
face a higher risk of coups, particularly due to the lack of clear succession
plans and vulnerabilities associated with the leader's personal weaknesses,
health issues, and finite lifespan. Military regimes, characterized by power
struggles within a junta, are the most vulnerable to coups. The absence of a
clear final authority in military regimes often leads to internal conflicts, and
the presence of multiple military factions increases the likelihood of resolving
disputes through force, making these regimes the least stable. @tbl-regimes1
summarises the key characteristics of the three main regime types.

These contrasting power dynamics significantly influence a regime's
susceptibility to coups. As @tbl-regimes confirms, military regimes, despite
representing only 5.6% of country-years, experience a disproportionate share of
coups, accounting for over 22% of all occurrences. Personalist regimes face a
similarly high coup risk, constituting 23% of coups while representing only 13%
of country-years. Conversely, dominant-party regimes, with their
institutionalized structures and unified leadership, exhibit the greatest
resilience. They represent 22.6% of country-years but account for only 16.7% of
coups. The column "Coup Likelihood" clearly illustrates the varying risks:
military regimes have the highest likelihood at 17.2%, followed by personalist
regimes at 7.7%, and dominant-party regimes at 3.2% (excluding *Other* regime
types).

> ***H2: Due to their balance of power dynamics, military regimes are more prone
> to coups, followed by personalist regimes, while dominant-party regimes are the
> least likely to experience coups among the three.***

\newpage

\blandscape

```{r}
#| label: tbl-regimes1
#| tbl-cap: Main features of different types of regimes 
#| echo: !expr format == "html"

read.csv(here("data/regimes_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center",
              page.orientation = "landscape"
              ) |>
  cols_width(
    1:6 ~ px(100)
  ) |>
  tab_source_note(source_note = md("*Source: GWF & Author*"))
```

\newpage

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and coups since 1950 
#| echo: !expr format == "html"

regimes <- REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949)

coup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/coup_model.csv")

regime_total <- coup_model |>
  select(regime_type = regime) |> 
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(country_year =n,
         share = percent(percent),
         .keep = "unused") 
  

coup1 <- Coup |>
  select(country, ccode, year, coup)

coup_by_regime <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup)) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  mutate(num_of_coups = n,
         percent_of_coups = percent(percent,accuracy = 0.1),
         .keep = "unused") 

coup_by_succeed <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup),
         coup == 2) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  select(regime, succeeded_coup = n) 

coup_regimes <-  coup_by_regime |> 
  left_join(coup_by_succeed) |> 
  mutate(success_rate = percent(succeeded_coup / num_of_coups, accuracy = 0.1))

regime_total |>
  left_join(coup_regimes, join_by(regime_type == regime)) |>
  mutate(regime_type = str_to_title(regime_type),
         coup_likelihood = percent(num_of_coups / country_year, accuracy = 0.1)) |>
  select(-succeeded_coup) |> 
  clean_names(case = "title") |>
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN and GIC Datasets*"))
```

\elandscape

### Research Design

#### Double probit with sample selection model

This study employs a sophisticated statistical approach to account for the
selective nature of coup attempts. While coup attempt rates vary across regimes
(as discussed previously), success rates tend to be surprisingly consistent,
hovering around 50% (as shown in @tbl-regimes). This suggests that coup attempts
are not random acts, but rather strategically planned and undertaken only when
the odds of success appear favourable. A standard statistical model would not
account for this selectivity, potentially leading to biased results.

To address this issue, we utilize a two-stage sample selection model, similar to
the approach used by @powell2012. This model has two parts:

-   **Selection Equation (Stage 1):** This stage analyses the factors influencing
    whether a coup attempt occurs in a particular regime. The primary explanatory
    variable here is regime type, as previously discussed. Additional control
    variables may also be included, denoted by $\mathbf{XB}$.

-   **Outcome Equation (Stage 2):** This stage focuses on the probability of
    success for those coup attempts that actually take place.

The primary explanatory variables are regime types, as previously discussed.
Control variables are included in $\mathbf{XB}$. The selection equation (first
stage) models the probability that a coup attempt occurs and can be expressed as
follows:

$$
\begin{aligned}
y_1^*=\alpha_0 + \alpha_1 Regime_i + \mathbf {XA} + \mu_{1i}
\end{aligned}
$$ {#eq-eq4}

Here, ${y_1}^*$ is an unobserved variable, which may be known to coup plotters.
$Regime_i$ is a categorical variable (*military*, *personalist*, or
*dominant-party*). $\mathbf{XB}$ captures other control variables, such as the
economic crisis index, previous coups, military expenditure, etc.

The observed binary outcome $\mathbf{y_1}$ is:

$$
\begin{aligned}
y_1 = 
\begin{cases} 
1 &\text{if $y_1^*>0$ (coup attempt occurs)} \\
0 &\text{if $y_1^*\le0$ (no coup attempt)}
\end{cases}
\end{aligned}
$$\
In the first stage, if $y_1^*\le0$, no coup attempt occurs in a given
country-year, indicating that the unobserved variable does not reach the
threshold. If $y_1^*>0$, at least one coup attempt is made in a country-year,
indicating that the unobserved variable surpasses the threshold. The probability
is expressed as:

$$
\begin{aligned}
Prob(y_1 =1)&=Prob(y_{1}^*>0) \\
&=\Phi(\alpha_0 + \alpha_1 Regime_i + \mathbf{XA})
\end{aligned}
$$ {#eq-eq4a}

Similarly, the outcome equation (second stage) models the probability that a coup
attempt is successful, given that it occurs:

$$
\begin{aligned}
y_2^*=\beta_0 + \beta_1 Regime_i + \mathbf {XB} + \mu_{2i}
\end{aligned}
$$ {#eq-eq5}

The observed outcome $y_2$ is:

$$
\begin{aligned}
y_2 = 
\begin{cases} 
1 &\text{if $y_2^*>0$ (coup succeeds)} \\
0 &\text{if $y_2^*\le0$ (coup fails)}
\end{cases}
\end{aligned}
$$

The probability equations is:

$$
\begin{aligned}
Prob(y_2=1|y_1 =1)=\Phi(\beta_0 + \beta_1 Regime_i + \mathbf{XB})
\end{aligned}
$$ {#eq-eq5a}

#### Variables

-   Dependent variable

Our analysis utilizes data on coup attempts and outcomes from @powell2011. A
successful coup is defined as one where the incumbent leader is removed from
power for more than seven days. The dataset covers the period from 1950 to 2023
and includes information on 491 coup attempts, with roughly half (245) being
successful. Descriptive statistics for these coup attempts and regime types can
be found in @tbl-coups and @tbl-regimes.

-   Key Independent Variable: Regime Type

The core variable of interest is regime type, categorized following the
classification system of @geddes2014 (GWF). We focus on military, personalist,
and dominant-party regimes, with democracies and monarchies included for
comparison. Descriptive statistics for regime types are presented in
@tbl-regimes.

-   Control variables

Our control variables are chosen based on the research of @gassebner2016. They
analyzed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

**Economic Performance:** We measure economic performance using the current-trend
($CT$) ratio developed by @krishnarajan2019. This ratio compares a country's
current GDP per capita to the average GDP per capita over the previous five
years. A higher $CT$ ratio indicates stronger economic performance. We use GDP
per capita data (in constant 2017 international 1000 dollars, PPP) from the V-Dem
dataset by @fariss2022, lagged by one year to reflect the prior year's economic
impact. For a country $i$ at year $t$, the $CT$ ratio is calculated as follows:
$$
\begin{aligned}
CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
\end{aligned}
$$ {#eq-eq6}

**Political Violence:** We capture overall regime stability by including a
violence index that encompasses all types of internal and interstate wars and
violence. This data comes from the Major Episodes of Political Violence dataset
by Marshall [@marshall2005current].

**Previous coups:** The number of previous coups in a country is included in the
first-stage (selection) model to assess its influence on the likelihood of a coup
attempt. However, it is excluded from the second-stage model (outcome) because
the number of past coups may not directly impact the outcome of a specific coup
attempt. Recognizing that previous coups from a long time ago might not
significantly affect the current situation, I ran another regression using the
time distance since the last coup instead of the number of previous coups for
comparison. This approach accounts for the diminishing impact of past events over
time, providing a potentially more accurate assessment of how historical coup
activity influences the likelihood of current coup attempts.

### Results and Discussion

```{r}
#| label: tbl-coupmodel
#| tbl-cap: "Sample Selection Model of Regime Type and Coup Success, 1950-2019"
#| results: asis
#| echo: !expr format == "html"

model_1 <- coup_model|>
  filter(regime != "other") |>
  select(coup, regime, GDP_trend, GDP_pc, violence, coup_dummy = pre_coups) |> 
  mutate(regime = fct_relevel(regime, "dominant-party")) |> 
  selection(coup > 0 ~ regime + GDP_trend + GDP_pc + violence + coup_dummy,
            coup > 1 ~ regime + GDP_trend + GDP_pc + violence,
            #method = "2step",
            data = _) 

model_1_selection <- model_1
model_1_selection$param$index$betaO <- model_1$param$index$betaS
model_1_selection$param$index$betaS <- model_1$param$index$betaO

model_2 <- coup_model |>
  filter(regime != "other") |>
  select(Coup = coup, regime, GDP_trend, GDP_pc, violence, coup_dummy, time_since_last_coup) |> 
  mutate(regime = fct_relevel(regime, "dominant-party")) |> 
  selection(Coup > 0 ~ regime + GDP_trend + GDP_pc + violence + coup_dummy*time_since_last_coup,
            Coup > 1 ~ regime + GDP_trend + GDP_pc + violence,
            data = _) 

model_2_selection <- model_2
model_2_selection$param$index$betaO <- model_2$param$index$betaS
model_2_selection$param$index$betaS <- model_2$param$index$betaO

stargazer(model_1_selection, model_1, model_2_selection, model_2,
          #type = "text",
          float = F,
          header = F,
          dep.var.caption = "",
          dep.var.labels = c("Model 1", "Model 2"),
          multicolumn = T,
          dep.var.labels.include = T, 
          column.labels = c("Coup Attempts", "Coup Outcome","Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant", "Regime: Democracy", "\\hspace{1.6cm}Military", "\\hspace{1.6cm}Monarchy", "\\hspace{1.6cm}Personalist", "Economic trend", "GDP per capita", "Political violence", "Previous coups (P)", "Yrs since coup (Y)", "Interaction term: P * Y"),
          summary.logical = F,
          column.sep.width = "7pt",
          font.size = "small"
          )
```

The double *`probit`* model with sample selection, estimated using the
*`sampleSelection`* package [@sampleSelection-2] in R, provides valuable insights
into the factors influencing coup attempts and their outcomes across different
regime types from 1950 to 2019 (@tbl-coupmodel). We present two models that
differ slightly in their treatment of previous coups: Model 1 incorporates the
number of previous coups, while Model 2 utilizes the time elapsed since the last
coup.

#### The Selection Model: Coup Attempts

In the selection model (Column 1), military and personalist regimes exhibit
significant positive coefficients at the 1% level, indicating a higher likelihood
of experiencing coup attempts compared to dominant-party regimes. Control
variables also exhibit effects in expected directions. Stronger economic
performance, indicated by higher economic growth trends and GDP per capita
levels, correlates with a lower risk of coup attempts. This suggests that better
economic conditions and relatively higher living standards reduce incentives for
coups. Political violence shows a positive and significant effect on coup
attempts, indicating that higher levels of instability increase the likelihood of
coups. The positive coefficient for the number of previous coups suggests a
"copycat" effect from earlier incidents.

```{r}
#| label: tbl-mfx1
#| tbl-cap: Average marginal effects of coup attempts (Selection of Model 1)
#| echo: !expr format == "html"

avg_comparisons(model_1) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_1, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups",
                          term == "GDP_trend" ~ "Economic trend",
                          term == "GDP_pc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Regime: Democracy",
                          str_detect(contrast, "military") ~ "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Regime: Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_style(
    style = cell_text(indent = px(63)),
    locations = cells_body(
      columns = Term,
      rows = 2:4
    )
  ) |> 
  tab_options(
    table.width = pct(95),
    # table.font.size = px(14),
    column_labels.font.weight = "bold"
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  )
```

While interpreting probit model results directly is not intuitive, @tbl-mfx1,
generated using the `marginaleffects` package [@marginaleffects], helps clarify
the regime effects. The Average Marginal Effect (AME) column shows changes in
probability when a variable increases by one unit (for numeric variables) or
compared to a reference category (for categorical variables). In the selection
equation of Model 1, the military regime's marginal effect of 0.07 indicates that
the probability of coup attempts in military regimes is 7 percentage points (pp)
higher than in dominant-party regimes, ceteris paribus. For GDP per capita, the
marginal effect of -0.002 suggests that a unit increase (\$1000) reduces the
probability of coup attempts by 0.2 pp under average conditions.

The Ratio Percent column displays the relative change in percentage terms. While
a 7 pp increase might seem modest, it is actually substantial. Given that the
average probability of a coup attempt in dominant-party regimes is approximately
2.53%, the 7 pp increase means military regimes are about 277.7% more likely to
encounter coups than dominant-party regimes—more than two times higher.
Similarly, personalist regimes show a 2.4 pp higher probability, about 94% more
likely compared to dominant-party regimes, nearly doubling the risk. Monarchies
display a positive effect similar to personalist regimes, reflecting that
monarchies are essentially a subset of personalist regimes with royal titles.

Control variables, however, show weak effects in predicting coup attempts. None
of their marginal effects reach 1 pp in probability, and all are less than 7% in
ratio, which is negligible given the baseline probabilities.

These results align with our theoretical expectations regarding internal power
struggles within military juntas and succession vulnerabilities in personalist
regimes, underscoring the importance of regime structure in understanding coup
likelihood.

```{r}
#| label: tbl-mfx2
#| tbl-cap: Average marginal effects of coup attempts (Selection of Model 2)
#| echo: !expr format == "html"
avg_comparisons(model_2) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_2, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups (P)",
                          term == "time_since_last_coup" ~ "Yrs since coup (P)" ,
                          term == "coup_dummy*time_since_last_coup" ~ "Interaction term: P * Y",
                          term == "GDP_trend" ~ "Economic trend",
                          term == "GDP_pc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Regime: Democracy",
                          str_detect(contrast, "military") ~ "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Regime: Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups (P)",
  "Yrs since coup (Y)",
   "Interaction term: P * Y"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_style(
    style = cell_text(indent = px(63)),
    locations = cells_body(
      columns = Term,
      rows = 2:4
    )
  ) |> 
  tab_options(
    table.width = pct(95),
    # table.font.size = px(14),
    column_labels.font.weight = "bold"
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  )
```

Model 2 employs years since the last coup instead of the number of previous
coups. Due to most countries not experiencing coups, an interaction term between
previous coups (as a binary variable indicating presence or absence of coups) and
years since the last coup is used. Generally, Model 2 shows results in the same
direction as Model 1, albeit with relatively lower coefficients (@tbl-mfx2).

#### The Outcome Model: Coup Success

The outcome model (Columns 2 and 4 in @tbl-coupmodel) reveals determinants of
coup success. Military regimes demonstrate a higher probability of coup success
compared to dominant-party regimes, aligning with expectations that military
regimes face higher coup risks due to their increased chances of success.
Personalist and monarchical regimes show slight positive effects on coup success,
but these effects are not statistically significant.

Control variables exhibit different patterns in the outcome model compared to the
selection model. Both GDP per capita and political violence maintain a weak
influence, similar to their effects in the selection model. However, the economic
trend shows a less significant negative effect on coup success.

These results indicate that regime type remains a significant determinant of both
coup attempts and successes, even after controlling for other factors, strongly
supporting the proposed theoretical framework.

```{r}
#| label: fig-margins
#| fig-cap: "Probability of Coup Attempts + Coup Success"
#| echo: !expr format == "html"
#| include: false
plot_predictions(model_1, by = "regime") +
  labs(
    x = "Regimes",
    y = "Total probability")

# plot_comparisons(model_2, 
#                  variables = "ect",
#                  by = "regime")
```

#### Discussion

The $\rho$ values of 0.898 in Model 1 and 0.386 in Model 2, significant at 1% and
10% levels respectively, are crucial parameters in the sample selection model.
These values represent the correlation between the error terms of the selection
equation (coup attempts) and the outcome equation (coup outcomes). A high and
significant $\rho$ suggests that unobserved factors influencing the likelihood of
a coup attempt are strongly correlated with those influencing the likelihood of a
successful coup. Practically, this indicates that the selection model is
appropriate and that accounting for selection bias (i.e., the fact that only
coups with high chances of success will be attempted) is critical for obtaining
unbiased estimates. The high $\rho$ value indicates that the same underlying
conditions that lead to a coup attempt also affect its success, underscoring the
importance of considering both stages in the analysis.

The results strongly support the choice of the sample selection model.
Significant coefficients with theoretically consistent directions suggest the
model effectively captures key aspects of coup dynamics. Regimes with weaker
institutional structures are more vulnerable to coup attempts, while better
economic conditions make coups less likely overall. The model effectively
addresses the non-random nature of coup attempts by treating selection and
outcome as separate processes.

The observed disparity between coup attempt rates and success rates across
regimes points towards selection bias, further validating the use of the sample
selection model. This model acknowledges that coups are not random events, but
rather strategic actions undertaken when the odds appear favourable.

In summary, the double `probit` model with sample selection proves to be a
well-suited approach for this research. It provides robust insights into the
factors influencing both the likelihood of coup attempts and their success rates
across different regime types. The findings highlight the crucial role of regime
structure and the selective nature of coup attempts, supporting the theoretical
framework and empirical strategy employed in this study.

#### Implications

The finding that regime type plays a crucial role in determining coup attempts,
while not surprising, has received relatively little attention as a primary
research focus. However, even if this result is acknowledged more widely by
academics or politicians, it is unlikely to be implemented as a coup-proofing
strategy.

Firstly, regime type is not a policy that can be easily altered. It is largely
established during the formation of the regime, resulting from power struggles
within the ruling group. As previously discussed, there are specific reasons why
one regime type is chosen over another. Regimes established through irregular
means are unlikely to transition to a more stable regime type peacefully and
smoothly.

Secondly, even if a particular regime type is more effective at preventing coups,
this may hold little significance for those in power. Autocratic leaders often
prioritize personal power and survival over long-term regime stability, as more
stable systems may not benefit them personally. Democratic norms and constraints
may be seen as obstacles to their goals rather than as safeguards against
instability.

While this conclusion might seem discouraging, it is not entirely pessimistic.
The power game involves not only the top leaders but also the majority of other
elites. While top leaders may prioritize personal gains, other elites, including
the military, economic elites, and civil society, have a vested interest in a
stable political environment. Therefore, it is possible to reach a consensus for
a political system that is less susceptible to coups, potentially benefiting a
broader spectrum of society.

### Conclusion

Motivated by the lack of consensus despite numerous empirical studies on the
determinants of coups, this study introduces a novel approach that prioritizes
determinants based on their impact on coup success. By analysing coup success
rates, the study hypothesizes that the expected outcomes of coups are critical
determinants of their occurrence. Utilizing a double `probit` model with sample
selection, I investigate and confirm the relationship between regime types and
coup attempts.

The findings suggest that regime type plays a significant role in the likelihood
of coup attempts. Military and personalist regimes, characterized by weaker
institutional frameworks and higher vulnerability during power transitions, are
more susceptible to coups. This underscores the importance of supporting
initiatives that strengthen constitutional institutions within these regimes.

The research also finds that stronger economic performance is associated with a
lower risk of coups, suggesting that policies promoting economic development can
be effective in reducing coup risk.

The study shows that the most efficient coup-proofing strategies involve the
establishment of strong institutions. In contrast, purges, random shifting of
military officers, or increased military expenditures are less effective.
However, few autocratic leaders, particularly dictators or military juntas, are
willing to institutionalize their regimes, as such reforms may constrain their
power or shorten their terms. While institutions benefit the regime, they do not
necessarily benefit the leaders themselves.

Future research could explore specific institutional reforms that are most
effective in improving stability across different regimes.

## Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders

### Abstract {.unnumbered}

This study aims to clarify the concept of autocoups, specifically focusing on
power extensions by incumbent leaders. By distinguishing autocoups from the
broader and more ambiguous concepts of self-coups or executive takeover, which
encompass both executive power aggrandizement and power extension, this research
redefines the concept of autocoups. Based on this refined definition, I introduce
a novel dataset of autocoup events from 1945 to 2022. Using the newly compiled
dataset, the research includes three types of case studies that provide
qualitative insights into the dynamics of autocoups. Additionally, an empirical
analysis on the determinants of autocoup attempts and success is offered to
demonstrate how the autocoup dataset can be employed for more quantitative
research. This study contributes to the existing literature by providing a
clearer conceptual framework and a novel dataset of autocoups. It enhances our
understanding of the mechanisms and motivations behind power extensions by
incumbent leaders and examines the implications for democratic backsliding,
democratic breakdown, personalization, and autocratic deterioration. The insights
gained from this study could draw more attention to the effects of autocoups on
power transitions, political stability, and democratic resilience.

keywords: *Coups, Autocoups, Political Leadership*

\newpage

### Introduction

The study of irregular power transitions, encompassing both irregular entries and
exits from office, has long been a central topic in political science. However,
one form of irregular power transition - the incumbent leader's refusal to
relinquish power - remains relatively understudied. This scenario, where leaders
overstay their mandated term limits for one or more terms, or even indefinitely,
lacks a universally accepted term in the existing literature. To address this
gap, this study adopts the term "autocoup" to denote this phenomenon.

Coups, being the primary form of irregular power transitions, have received
significant scholarly attention. Research by @goemans2009 indicate that coups
account for a substantial portion of such transitions, with roughly two-thirds
(65.8%) of 374 irregular exits occurring through coups. Additionally, @frantz2016
demonstrate that coups are the most common form of exit in autocracies,
surpassing regular transitions by a third.

However, while coups were once frequent, recent decades have witnessed a decline
in their occurrence [@bermeo2016; @thyne2019]. Conversely, autocoups appear to be
on the rise, particularly since the Cold War's end [@ginsburg2010evasion;
@baturo2014; @versteeg2020law]. Despite the significant political impact and
rising prevalence of autocoups, their study lags behind that of classical coups,
which are extensively documented due to their historical frequency [@thyne2019].

This research aims to bridge this gap by focusing on the understudied phenomenon
of autocoups as they are at least equally important and warrant comprehensive
analysis. Firstly, autocoups are typically accompanied by a disregard for the
rule of law, weakened institutions, and potential democratic backsliding or
authoritarian personalization. Secondly, like traditional coups, successful
autocoups increase the risk of future irregular power transitions. Research
indicates that since 1945, approximately 62 percent of leaders who overstayed
their term limits through autocoups in non-democratic countries were either
ousted or assassinated while in office [@baturo2019]. Thirdly, failed autocoups
often trigger instability that fuels protests, violence, and even civil wars.

Despite their significance, autocoups are understudied compared to traditional
coups. Although several relevant concepts and terminologies such as self-coups,
autocoups, autogolpes, incumbent takeovers, executive aggrandizement, overstay,
and continuismo have been proposed and discussed [@marsteintredet2019;
@baturo2022], there is no clear and widely accepted term to describe these
events. Different terms are often used interchangeably or inconsistently, leading
to confusion. Moreover, due to the lack of clear concepts, the collection of
autocoup datasets is still in its early stages compared to the rich datasets of
traditional coups. Hence, the limited existing studies on autocoups have
primarily relied on case studies [@cameron1998; @antonio2021; @pion-berlin2022],
with few focusing on quantitative analyses.

To fully understand irregular political leadership transitions and survival, we
need to redefine and clarify the term "autocoup" first. Based on a clear
definition, we can introduce an autocoup dataset that can be used for further
empirical analysis. This study aims to address these knowledge gap surrounding
autocoups and contribute in three areas. First, it clarifies the terminology by
defining autocoups based mainly on power extension. Second, it introduces a new
dataset of autocoups since 1945 based on this refined definition. Finally, the
research utilizes this dataset for a quantitative analysis of the factors that
influence leaders' decisions to attempt autocoups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of
autocoups. This will be followed by an introduction to the new autocoup dataset
in Section 3. In Section 4 and 5, I will explain the determinants of autocoup
attempts through two case studies and demonstrate how the novel autocoup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude this chapter in Section, summarizing the key insights gained from the
study and suggesting avenues for future research.

### Autocoups: A literature review and clarification of definitions

Compared to coups, which are clearly defined and widely accepted as illegal
attempts by elites within the ruling group to overthrow the leadership
[@powell2011], the concept of autocoups (also known as self-coups or autogolpes)
suffers from a lack of consistency and clarity. This ambiguity hinders our
understanding and study of a critical phenomenon in irregular power transitions.
To address this gap, we need to tackle two key issues: terminology and
definition.

#### Terminology

The most common term in autocoup literature is self-coup, or autogolpe in Spanish
[@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017; @marsteintredet2019].
This term gained academic prominence after Peruvian President Alberto Fujimori
dissolved Congress, temporarily suspended the constitution, and ruled by decree
in 1992 [@mauceri1995; @cameron1998]. However, as @marsteintredet2019 point out,
the term "self-coup" can be misleading, as it implies a coup against oneself,
which is inaccurate since it is not self-directed but targets other state
institutions or apparatus.

Another approach to describe coups staged by incumbents is to use terms with
adjectives or modifiers, such as presidential coup, executive coup,
constitutional coup, electoral coup, judicial coup, slow-motion coup, soft coup,
and parliamentary coup [@marsteintredet2019]. While these terms can be helpful in
specific contexts, their proliferation often adds to the overall confusion rather
than providing clarification. Most of these terms focus on the specific methods
used by coup perpetrators but fail to clearly identify the perpetrator,
necessitating further explanation. In fact, many of these methods could be
employed either by or against executive leaders.

A third alternative involves terms like "incumbent takeover" or "overstay".
Incumbent takeover referrs to "an event perpetuated by a ruling executive that
significantly reduces the formal and/or informal constraints on his/her power"
[@baturo2022, p.374], based on earlier research [@svolik2014]. While overstay is
defined as "staying longer than the maximum term as it stood when the candidate
originally came in office” [@ginsburg2011evasion, p.1844]. These terms identify
the perpetrator (the incumbent) and/or the nature of the event
(overstaying/extending power). However, they do not highlight the illegality or
illegitimacy of these actions. Therefore, they cannot serve as a direct
counterpart to "coup," which clearly denotes the illegality of leadership
ousters, while "takeover" or "overstay" diminish the severity.

Therefore, this study strongly advocates for "autocoup" as the most accurate and
appropriate term. It avoids the pitfalls of other terms by clearly conveying the
essence of the phenomenon as a coup-like power grab and effectively
distinguishing autocoups from classical coups by highlighting the perpetrator.

#### Definition

While terminology is important, another issue arises with the previous definition
of autocoups: should we emphasize power expansion or power extension? Power
Expansion refers to situations where an executive branch expands its power beyond
its original boundaries, often by encroaching on the authority of other branches
like the legislature or judiciary. Power Extension, on the other hand, refers to
situations where a leader extends their tenure in office beyond the originally
mandated term limits. Existing definitions of self-coup or autocoup sometimes
touch upon power extensions, but they primarily focus on power expansions, which
has several drawbacks.

Firstly, defining autocoups primarily in terms of power expansion does not align
well with the definition of a coup. When we define a classical coup, the focus is
clearly on the ouster of the current leader, not merely a limitation or
restriction on their power. Using the same logic, a more appropriate definition
of an autocoup should prioritize the extension of executive leadership. We would
not classify an event as a coup if a vice president seizes some power from the
president, as long as the president remains in office. Similarly, an executive
leader acquiring more power from other branches should not automatically qualify
as an autocoup if their term limits remain unchanged.

Secondly, emphasizing power expansion raises issues regarding the purpose of
launching an autocoup. As @cameron1998a defined, a self-coup is "a temporary
suspension of the constitution and dissolution of congress by the executive, who
rules by decree until new legislative elections and a referendum can be held to
ratify a political system with broader executive power" [@cameron1998a, p.220].
However, without extending their time in office, the leader faces significant
risks once their term expires, as they could face legal or political
repercussions for the autocoup itself, leaving behind an overly powerful
executive branch for the successor. Although the term "self-coup" gained
prominence from the 1992 Fujimori case in Peru, which initially involved seizing
power from other institutions, it is important to note that Fujimori ultimately
extended his term limits through constitutional amendments. The 1993 Constitution
allowed Fujimori to run for a second term, which he won with popularity in April
1995. Shortly after Fujimori began his second term, his supporters in Congress
passed a law of "authentic interpretation" which effectively allowed him to run
for another term in 2000, which he won amid suspicions and rumors. However, he
did not survive the third term. In 2000, facing charges of corruption and human
rights abuses, Fujimori fled Peru and took refuge in Japan [@ezrow2019].

Thirdly, measuring the extent of power expansion to qualify as an autocoup can be
challenging. Power expansion might occur as a sudden, significant event, as seen
with Fujimori in 1992, or it could be gradual and incremental, as in the case of
Putin since assuming power in 1999 [@chaisty2019]. In the latter case, it is
unclear whether each step should be considered an autocoup or if the entire
process together should be labelled as such. If the latter, determining the point
at which it becomes an autocoup is again problematic.

Therefore, this study argues that a more accurate definition of autocoups should
prioritize power extension as the core characteristic.

In light of this discussion, we formally define an autocoup as ***the practice of
current political leaders extending their time in office beyond their originally
mandated term limits through illegitimate means*****.** Three key points need to
be highlighted for this definition. Firstly, this definition refers to the actual
leaders of the country, regardless of their official titles. Typically, this
would be the president; however, in some cases, such as in Germany, the primary
leader is the premier, as the president is a nominal head of state. Secondly,
while the primary characteristic of an autocoup is extending the term in office,
this definition does not exclude instances of power expansion. Both aspects can
coexist, but the extension of the term is the central element. Thirdly,
autocoups, by their nature, subvert legal norms and established power transfer
mechanisms. While they may employ seemingly legal tactics, their essence is
illegitimacy. This critical aspect will be explored further in Section 3.

### Introduction to the Autocoup Dataset

#### Defining the scope

Defining any concept or term inevitably involves borderline cases that are
difficult to categorize. In the context of autocoups, it is particularly
challenging to determine whether a specific instance of overstaying should be
coded as an autocoup. To avoid ambiguity and dilemmas, we will code all instances
of incumbents extending their original mandated term in office as autocoups,
regardless of whether the extension is patently illegal or superficially legal.

This approach is taken because a truly legitimate amendment to power transition
institutions should only apply to subsequent leaders, not the incumbent. If the
incumbent changes the rules to their own benefit, the legitimacy of such changes
is questionable, even if the amendment procedures appear legal. We have valid
reasons to question their motivations since they are the beneficiaries of these
reforms.

Even so, some cases remain unclear. For example, consider a two-term limit
presidency where the leader wins a second term. Allegations of election rigging
might surface, raising suspicions about the fairness of the election results. If
such suspicions can be verified with evidence or judicial verdict, the case would
undoubtedly be coded as an autocoup. However, proving election rigging in
countries with powerful executives can be difficult. In such instances, we will
not classify it as an autocoup unless a formal judicial verdict confirms the
rigging.

Even in contexts where leaders lack clear term limits or the constitution allows
indefinite terms, identifying autocoups remains possible. We will discuss and
explain this in the next section.

#### Classifying autocoups {#sec-classify}

Autocoups can manifest in various forms. To understand this phenomenon better, we
can categorize them based on several key factors:

-   **Methods Employed:** This refers to the specific strategies incumbents use
    to extend their hold on power. Examples include constitution amending or
    reinterpreting, election cancelling, delaying or rigging.

-   **Degree of Legality:** This dimension captures the extent to which an
    autocoup deviates from established legal norms. It can range from minor
    rule-bending to outright constitutional violations.

-   **Duration of Extension:** This category focuses on the length of time the
    incumbent leader remains in office beyond their designated term limits. It
    could be a single additional term, multiple terms, or even a lifetime hold on
    power.

-   **Outcomes:** This factor examines whether the autocoup attempt is successful
    or ultimately fails.

This study will primarily focus on the **methods** employed by incumbents to
stage autocoups. However, we will also code for other relevant aspects like the
degree of legality, duration of extension, and outcomes whenever information is
available. This multifaceted approach allows for a more nuanced understanding of
the variations within autocoups.

##### Evasion of term limits {.unnumbered}

Evasion of term limits is a common tactic employed in autocoups. Incumbents often
resort to seemingly legal manoeuvres to extend their hold on power. These
manoeuvres primarily involve manipulating constitutional provisions through
various means. The incumbents may pressure legislative bodies (congress) or
judicial institutions (Supreme Court) to reinterpret existing term limits, amend
the constitution to extend terms, or even replace the constitution altogether.
This might also involve popular vote through referendums, or a combination of
these approaches. The extension can range from a single term to indefinite rule.

These manoeuvres primarily involve manipulating constitutional provisions through
various means.

**Changing Term Length:** Incumbents might lengthen the official term duration
(e.g., from 4 to 6 years) to stay in office longer, even if the number of allowed
terms remains unchanged. Examples, in the dataset, include Presidents Dacko (CAR,
1962), Kayibanda (Rwanda, 1973), and Pinochet (Chile, 1988).

-   **Enabling re-election:** This approach involves incumbents modifying legal
    or constitutional frameworks to permit themselves to run for leadership
    again, despite initial restrictions. These restrictions might include
    prohibitions on re-election, bans on immediate re-election, or term limits
    that the incumbents have already reached. An illustrative example is
    President Menem of Argentina in 1993, who leveraged this tactic to extend his
    tenure.

-   **Removing Term Limits Altogether:** This approach, as seen with President Xi
    Jinping of China in 2018, technically allows the leader to rule for life,
    although they may still need to participate in elections (a formality in such
    cases).

-   **Leader for Life:** This differs from removing term limits as the leader
    still faces elections (although potentially rigged or uncontested). An
    example is Indonesia's President Sukarno, who attempted to declare himself
    president for life in 1963 (ultimately unsuccessful).

These methods are often used in combination. Initially, the duration of a term is
extended, followed by amendments to allow re-election, then the removal of term
limits, and finally, the declaration of the leader for life. For example, Haitian
President François Duvalier amended the constitution in 1961 to permit immediate
re-election and then declared himself president for life in 1964.

##### Election Manipulation or Rigging {.unnumbered}

Election manipulation or rigging is the second most commonly used tactic to
extend an incumbent's tenure.

-   **Delaying or Removing Elections:** Delaying or removing scheduled elections
    without legitimate justification is a frequent method used by incumbents to
    maintain power. For instance, Chadian President François Tombalbaye delayed
    general elections until 1969 after assuming power in 1960. Similarly, Angolan
    President José Eduardo dos Santos suspended elections throughout his rule
    from 1979 to 2017.

-   **Refusing Unfavourable Election Results:** Incumbents may refuse to accept
    unfavourable election results and attempt to overturn them through
    illegitimate means. For example, President Donald Trump of the United States
    refused to accept the results of the 2020 election and tried to overturn
    them.

-   **Rigging Elections:** Winning elections with an extraordinarily high
    percentage of votes is highly questionable. This study will code elections
    where the incumbent wins more than 90% of the vote as autocoups. For
    instance, President Teodoro Obiang of Equatorial Guinea has consistently won
    elections with over 95% of the vote in multi-party elections since 1996,
    indicating election rigging.

-   **Excluding Opposition in Elections:** Manipulating the electoral process by
    excluding opposition parties or candidates from participation, effectively
    creating a one-candidate race, clearly signifies an autocoup.

##### **Figurehead** {.unnumbered}

To circumvent term limits, some incumbents might choose a close associate to act
as a figurehead, taking the office publicly while the incumbent retains real
power behind the scenes. This can be achieved through seemingly subordinate
positions.

One example is Russia in 2008. Facing term limits, President Putin selected
Dmitry Medvedev to run for president. After the election, Medvedev appointed
Putin as Prime Minister. However, most analysts believe Putin wielded the true
power throughout this period.

##### Reassigning supreme authority to a new role {.unnumbered}

This tactic involves an incumbent leader manipulating the constitution or legal
framework to create a new position of power, or elevate an existing one, before
stepping down from their current role. They then strategically take on this new
position, effectively retaining significant control despite appearing to
relinquish power. For example, in 2017, Recep Tayyip Erdoğan, the Prime Minister
of Turkey, spearheaded a constitutional referendum that transitioned the country
from a parliamentary system to a presidential one. This new system concentrated
significant executive power in the presidency. Following the referendum's
approval, Erdoğan successfully ran for the newly established presidency,
effectively retaining control under a different title.

##### One-Time Arrangement for Current Leaders {.unnumbered}

This strategy involves special arrangements that extend the term or tenure of
current leaders without altering the underlying institutions. For example,
Lebanon extended President Émile Lahoud's term by three years in 2004 through a
one-time arrangement.

#### Data Coding

```{r}
#| label: tbl-source
#| tbl-cap: Sources of coding autocoup dataset 
#| echo: !expr format == "html"
#| results: asis

read.csv(here("data/autocoup_source.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1:4 ~ px(80),
    5 ~ px(200)
  ) |>
   data_color(rows = c(1, 3), colors = "#edede9") 
```

The autocoup dataset, like most dataset coding procedures, is based on existing
studies and datasets. @tbl-source outlines the main sources for coding the
autocoup dataset. The Archigos dataset [@goemans2009] and the Political Leaders’
Affiliation Database (PLAD) dataset [@bomprezzi2024wedded] provide comprehensive
data on all leaders from 1875 to 2023, although our coding only includes
autocoups since 1945. These datasets help identify the actual rulers of
countries, saving time in distinguishing real leaders from nominal heads of
state.

The Incumbent Takeover dataset [@baturo2022], which integrates data from 11
related datasets, offers a broad spectrum of cases where leaders significantly
reduce the constraints on their power, encompassing both power expansions and
extensions. Since some executive takeovers do not qualify as autocoups due to the
lack of term extensions, I cross-referenced the Archigos dataset, which includes
detailed trajectories of leaders' entries and exits from power, to verify the
qualifications for autocoups.

```{r}
#| label: data-autocoups
#| echo: !expr format == "html"

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))
```

In total, I coded `r nrow(Autocoup)` observations, with
`r nrow(Autocoup |> filter(!is.na(source)))` overlapping with the candidate data
from Incumbent Takeover. The remaining
`r nrow(Autocoup) - nrow(Autocoup |> filter(!is.na(source)))` events were newly
coded by the author through verification with other sources such as Archigos, and
news reports.The main deviation from the Incumbent Takeover dataset arises from
excluding power expansions that do not involve attempts to extend tenure.

The main variables included in the autocoup dataset are:

-   **Country Identification:**

    -   `ccode` and `country`: These variables come from the Correlates of War
        project [@stinnett2002] and identify the countries included in the
        dataset, which are widely used in political science datasets, ensuring
        consistency and compatibility across different studies and datasets.

-   **Leader Information:**

    -   `leader_name`: This variable follows the Archigos dataset's coding and
        records the name of the de facto leader who wielded power in the country.

-   **Timeline Variables:**

    -   `entry_date`: Captures the date the leader assumed power.

    -   `exit_date`: Records the date the leader left office.

    -   `autocoup_date`: Indicates the date the autocoup is considered to have
        occurred. Since extensions often happen incrementally, this date reflects
        a significant event marking the extension, such as a legislative vote or
        successful referendum.

    -   `extending_date`: Represents the start date of the leader's additional
        term acquired through the autocoup.

    The `entry_date` and `exit_date` come from the Archigos and PLAD datasets,
    while the other two dates are coded by the dataset creator.

-   **Power Transition Methods:**

    -   `entry_method` and `exit_method`: These categorical variables record how
        the leader entered and exited power (e.g., election, coup, death).

    -   `entry_regular` and `exit_regular`: These dummy variables indicate
        whether the entry and exit methods were regular (e.g., election) or
        irregular (e.g., coup).

-   **Autocoup Details:**

    -   `autocoup_method`: This key variable captures the various methods leaders
        use to extend their stay in power (see @sec-classify for details).

    -   `autocoup_outcome`: This variable indicates the outcome of the autocoup
        attempt: "fail and lose power", "fail but complete original tenure", or
        "successful". For successful coups, the additional term length can be
        calculated from the difference between `exit_date` and `extending_date`.

-   **Data Source:**

    -   `source`: This variable identifies the dataset source used for coding,
        primarily differentiating among "Incumbent Takeovers" or other sources.

-   **Additional Notes:**

    -   `notes`: This variable provides context for exceptional cases. For
        instance, if a leader undertook multiple autocoup attempts, details are
        recorded here.

The dataset encompasses a total of 14 variables along with the `notes` field.

#### Data descriptions

```{r}
#| label: data_descriptions
#| echo: !expr format == "html"

country <- Autocoup |> 
  distinct(ccode) |> 
  summarise(count = n())

autocoup_desc <- Autocoup |> 
  summarise(autocoup_attempted = n(),
            autocoup_succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method) |> 
  arrange(-autocoup_attempted)
```

According to our primary coding, we have identified `r nrow(Autocoup)` autocoup
cases from 1945 to 2022, involving `r country$count` countries. As shown in
@tbl-autocoup_method, the most common autocoup method is enabling re-election,
accounting for `r autocoup_desc$autocoup_attempted[1]` events. This is followed
by removing term limits altogether, with `r autocoup_desc$autocoup_attempted[2]`
cases, and then delaying elections and declaring the leader for life, each with
`r autocoup_desc$autocoup_attempted[3]` cases.

Examining the success rates of autocoups, the total success rate is 79%, which is
significantly higher than the roughly 50% success rate of classical coups. This
suggests that incumbents are in an advantageous position to expand or extend
their powers as executive leaders. They can do so openly and gradually, whereas
coup plotters must operate in secrecy and face numerous challenges such as
promissory issues and the risk of betrayal or exposure.

However, the success rates vary significantly across different methods. Removing
term limits, delaying elections, declaring the leader for life, and cancelling
elections are all 100% successful. In contrast, there are only 4 cases of
refusing election results, with just one succeeding. Although this sample is
small, it suggests that in regimes where general elections are held and
incumbents lose, the system is relatively more democratic. In such cases,
incumbents must run for office without the ability to rig elections, making it
less likely for them to overturn election results.

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup methods and success rates (1945-2021)
#| echo: !expr format == "html"

autocoup_desc |> 
  adorn_totals() |> 
  mutate(Success_rate = percent(autocoup_succeeded/autocoup_attempted)) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(180),
    2:3 ~ px(140),
    4 ~ px(100)
  )

#autocoup_desc$autocoup_attempted[1]
```

### Determinants of Autocoup Attempts: Case Studies

#### High Frequency and Success Rate of Autocoups in Post-Communist Countries

From the dataset, we observe that in post-communist countries, both the frequency
and success rate of autocoups are notably high. Post-communist countries refer to
those that were communist regimes before the collapse of the Soviet Union, while
most of them developed into 'hybrid regimes' [@nurumov2019] and only a few remain
communist regimes after the collapse. In these countries, there are 12 documented
cases of autocoups aimed at prolonging incumbency, with only 2 of these attempts
failing. Examining the cases in post-communist countries, several characteristics
stand out:

-   **Inherited Authoritarian Systems**: Although most of these countries
    transitioned from communist regimes to non-communist governments (with the
    exception of China), they inherited the authoritarian systems of their
    communist past.

-   **Continuity of Former Elites**: The transitions did not result in the
    removal or overthrow of the previous ruling groups. Instead, the former
    communist elites remained in power.

-   **Subverted Democratic Processes**: Despite the introduction of general
    elections and term limits in most of these countries, the legacy of the
    former communist regimes often led to term limits being ignored and elections
    being rigged [@nurumov2019].

For example, Alexander Lukashenko was a member of the Supreme Soviet of the
Byelorussian Soviet Socialist Republic before the dissolution of the Soviet
Union. After the dissolution, he became head of the interim anti-corruption
committee of the Supreme Council of Belarus. Elected as the first president of
Belarus in 1994, he has held the office ever since. Initially, the 1994
constitution set a maximum of two successive presidential terms, but Lukashenko
removed this limit in 2004. Furthermore, international monitors have not regarded
Belarusian elections as free and fair, except for his initial win. Despite
significant protests against him, Lukashenko claimed to win with a high vote
share, often exceeding 80% in each election. This pattern is evident in all five
Central Asian countries of the former Soviet Union. Similarly, the
post-dissolution leaders of these countries were high officials or heads of the
former Soviet republics who continued their leadership in the presidency.

Another long-ruling example is Nursultan Nazarbayev, who was the first president
of Kazakhstan from 1991 until 2019. He had been the real leader as the First
Secretary of the Communist Party of Kazakhstan before the dissolution of the
Soviet Union. After independence, he was elected as the first president and held
the office until 2019, through various means like resetting the term limits due
to the implementation of new constitutions. However, he did not officially
eliminate the term limits but made an exemption for the First President,
Nazarbayev [@nurumov2019]. Unlike Lukashenko, who is still the incumbent of
Belarus, Nazarbayev passed the presidency to Kassym-Jomart Tokayev, a specially
designated successor, in 2019. However, he retained significant influence as the
Chairman of the Security Council of Kazakhstan until 2022.

```{r}
#| label: region
#| echo: !expr format == "html"

library(countrycode)
latin <- codelist |> 
  filter(region == "Latin America & Caribbean") |> 
  select(region, cown) |>
  drop_na() |> 
  pull(cown)
africa <- codelist |> 
  filter(continent == "Africa") |> 
  select(continent, cown) |>
  drop_na() |> 
  pull(cown)
```

#### Autocoups for immediate re-election: Cases of Latin American countries

Latin America has a long history of maintaining term limit conventions. Simón
Bolívar, the founding father of Bolivia, was a strong advocate for term limits,
stating in 1819, “Nothing is as dangerous as allowing the same citizen to remain
in power for a long time... That’s the origin of usurpation and tyranny”
[@ginsburg2019]. Although Bolívar eventually changed his stance, arguing in his
1826 Constitution Assembly speech that “a president for life with the right to
choose the successor is the most sublime inspiration for the republican order,”
term limits became a convention in Latin America. Approximately 81% of Latin
American constitutions between independence and 1985 imposed some form of term
limits on the presidency [@marsteintredet2019a].

Reviewing the cases in Latin American countries, we notice that:

-   **Striving for Re-election:** Non-re-election or non-immediate re-election
    has been common in Latin America, unlike other presidential countries where
    two terms are more popular. Autocoup leaders in Latin America often attempt
    to overstay a consecutive term.

-   **Resisting long extensions:** Autocoups for one more term are often
    successful, while attempts to overstay beyond this are frequently
    unsuccessful.

According to @marsteintredet2019a, non-consecutive re-election was mandated in
about 64.9% of all constitutions between independence and 1985, while 5.9% banned
re-election entirely. However, adherence to these conventions has varied across
the region. Since Mexico introduced non-re-election institutions in 1911 at the
start of the Mexican Revolution, they have never been violated since then
[@klesner2019]. Panama, along with Uruguay, has never changed the rules of
re-election, and since Costa Rica prohibited immediate presidential re-election
in 1859, the country has only experienced a brief period between 1897 and 1913 in
which the incumbent president could be re-elected [@marsteintredet2019a]. In many
other countries, however, constitutions have been frequently amended or violated.
The pursuit of re-election or consecutive re-election has been a significant
trigger for autocoups aimed at power extension in Latin American countries. There
are 32 documented autocoup cases, with over 50% (17 cases) attempting to enable
re-election.

Unlike those who attempt to overstay in office indefinitely, many Latin American
leaders exit after their second term expires. Examples include President Menem of
Argentina (1988-1999), President Fernando Cardoso of Brazil (1995-2003),
President Danilo Medina of the Dominican Republic (2012-2020), and President Juan
Orlando Hernández of Honduras (2014-2022) [@ginsburg2019; @marsteintredet2019a;
@landau2019; @baturo2019; @neto2019]. This does not mean none of them tried to
extend even longer, but most did not manipulate the process by abusing their
power and accepted their unsuccessful outcomes soon.

For instance, President Menem of Argentina successfully extended one term by
amending the constitution in 1994 to allow one executive re-election and was
re-elected in 1995. However, when he attempted to reset his term, arguing that
his first term from 1988 to 1995 did not count since it was under previous
constitutions, his appeal was unanimously ruled out by the Supreme Court in March
1999 [@llanos2019]. A similar scenario occurred with President Álvaro Uribe of
Colombia (2002-2010) [@baturo2019]. In contrast, Daniel Ortega, the incumbent
president of Nicaragua, successfully extended his presidency. In 2009, the
Supreme Court of Justice of Nicaragua allowed his re-running in 2011. In 2014,
the National Assembly of Nicaragua approved constitutional amendments that
abolished term limits for the presidency, allowing Ortega to run for an unlimited
number of five-year terms, making him president since 2007 [@close2019].

#### As common as classical coups: Cases of African countries

Classical coups have been very common in Africa, accounting for about 45% of all
coups (219 out of 491) globally since 1950, involving 45 out of 54 African
countries [@powell2011]. Autocoups, although less common compared to coups, still
have a significant presence in Africa. Among 113 documented autocoup cases, 46%
(52 cases) occurred in Africa, involving 36 countries. The success rate of coups
in Africa is roughly 50%, while the success rate of autocoups is about 83%, which
is higher than both the success rate of coups and the average global success rate
of autocoups, approximately 78%.

Identifying a clear pattern of autocoups in Africa is challenging, similar to the
case with coups. Various factors have been proposed:

-   **Natural Resources**: Countries rich in natural resources, particularly oil
    or diamonds, may see leaders more likely to attempt and succeed in extending
    their terms [@posner; @cheeseman2015; @cheeseman2019a].

-   **Quality of Democracy**: The quality of democracy is a critical factor
    influencing respect for term limits [@reyntjens2016].

-   **International Influence**: International aid or donor influence can play a
    significant role in discouraging attempts at power extension [@brown2001;
    @tangri2010].

-   **Organized Opposition and Party Unity**: The extent of organized opposition
    and the president's ability to enforce unity within the ruling party are
    crucial factors [@cheeseman2019].

Using the Africa Executive Term Limits (AETL) dataset, @cassani2020 highlights
human rights abuses and the desire for impunity as main drivers for incumbents to
cling to power. The more authoritarian a leader, the more likely they are to
attempt to break term limits and overstay in office. Additionally, a leader's
ability to secure the loyalty of the armed forces through public investment
increases the chances of success in overstaying.

Despite both coups and autocoups being prevalent, there has been a noticeable
shift since the end of the Cold War in 1991. Coups have decreased, while
autocoups have increased. This trend is partly due to the introduction of
multi-party elections in Africa in the 1990s, which also brought in term limits
for executives [@cassani2020; @cheeseman2019]. Before 1991, personal or military
rule was more common, and term limits were less frequent. Post-1991, with more
term limits introduced, challenges to these limits have increased. However, this
does not imply that violations are more common than adherence to term limits.

### Empirical Analysis: A Simple Example of Utilizing the Autocoup Dataset

With the availability of the autocoup dataset, we can now conduct quantitative
analyses that extend beyond traditional case studies. This section provides a
straightforward example of how to utilize this dataset effectively. To analyse
the determinants of autocoup attempts, we employ a probit regression model. This
model is particularly appropriate for binary dependent variables, allowing us to
estimate the probability of an autocoup attempt given a set of independent
variables.

However, our approach here differs from the double probit model with sample
selection used in Chapter 2 for coup attempts and success analyses. Instead, we
use two separate probit models. This methodological choice is justified by
several factors specific to autocoups:

-   **Incumbent Advantage**: Autocoups have a significantly higher success rate
    compared to traditional coups, often exceeding 80%. This high probability of
    success fundamentally alters the decision-making calculus for potential
    perpetrators.

-   **Reduced Risk**: Even in cases where autocoup attempts fail, the
    consequences for the instigators are typically less severe than those faced
    by leaders of failed traditional coups. This lower risk profile further
    distinguishes autocoups from other types of coup attempts.

Due to these unique features, autocoup attempts do not exhibit the typical sample
selection characteristics that necessitated the use of a sample selection model
in our earlier analysis of traditional coups. Given these considerations, a
sample selection model might not be appropriate in this context. By using
separate probit models, we can independently examine the factors that influence
autocoup attempts and their outcomes, without assuming a strong selection effect.

#### Dependent Variable

-   **Autocoup Attempts**: This binary variable indicates whether an autocoup was
    attempted during the tenure of an incumbent leader, based on the data
    introduced in this study.

-   **Autocoup Success:** This binary variable indicates whether an attempted
    autocoup of the incumbent leader succeeded, based on the data introduced in
    this study.

#### Independent Variables

-   **Economic Performance:** This variable is measured using two indicators:
    economic level and economic growth trend.

    **Economic Level:** Represented by GDP per capita. This measure provides an
    indication of the overall economic health and standard of living in a
    country.

    **Economic Growth Trend:** Assessed using the current-trend (CT) ratio,
    developed by @krishnarajan2019, which is consistent with @eq-eq6.

The GDP per capita data, expressed in constant 2017 international dollars (PPP)
and measured in units of \$10,000, is sourced from the *V-Dem* dataset by
@fariss2022. To account for the economic impact of the previous year, this data
is lagged by one year.

-   **Political Stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    Major Episodes of Political Violence dataset by Marshall. This index provides
    a comprehensive measure of the level of violence and conflict within a
    country, which can significantly impact power transitions.
    [@marshall2005current].

-   **Degree of Democracy:** The level of democracy is gauged using Polity 5
    scores at the entry year for each respective country. These scores range from
    -10 (fully autocratic) to +10 (fully democratic), capturing the extent of
    democratic versus autocratic governance. This dataset is sourced from the
    Centre for Systemic Peace (CSP)[^6] and provides an essential measure of
    political regime type, which might influence the political stability and
    power transitions.

-   **Population Size:** To account for its potential impact on leaders' tenures,
    the log of the population size is considered. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on power transitions. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's Age:** The age of the leader is included as an additional variable
    in the analysis, offering insights into potential correlations with
    leadership strength. Older leaders may have different experiences, networks,
    and health considerations that could influence their ability to maintain
    power. This data is sourced from the leaders dataset by [@goemans2009].

Different from analysing the determinants of coups, which theoretically could
happen in each year, we assume that autocoup happens only once during the tenure
of the incumbent. Because once the incumbent succeeds an autocoup, there is no
necessary to stage another one. Of course, this is not the case in real time. A
leader who successfully overstayed one term could try to overstay additional
term, or even strive for leadership for life. A leader who failed in the first
attempt might make another try if they are not removed due the first failure. To
simplify the analysis, we overlook those possibilities.

Therefore, in the *`probit`* model, the unit of analysis in autocoups is the
whole tenure of the leader, not country-year. So, we need to settle a base-year
for the variables. For leaders who staged an autocoup, we take the year of their
first attempts as the base-year, while for leaders who did not attempt to
overstay, we take the middle year of their tenure as the base-year.

Unlike the analysis of coup determinants, which could theoretically occur in any
given year, we assume that an autocoup happens only once during an incumbent
leader's tenure. Once an incumbent succeeds in staging an autocoup, there is no
need for another attempt. However, this assumption does not always reflect
reality. A leader who successfully extends their term might attempt further
extensions, striving for lifelong leadership. Conversely, a leader who fails in
their first attempt may try again if they are not removed from power after the
initial failure. For simplicity, we overlook these possibilities in our analysis.

Therefore, in our *`probit`* model, the unit of analysis for autocoups is the
entire tenure of a leader, rather than a country-year. We establish a base year
for the variables: for leaders who staged an autocoup, we use the year of their
first attempt as the base year; for leaders who did not attempt to overstay, we
use the middle year of their tenure as the base year.

#### Results and discussions

```{r}
#| label: tbl-autocoupmodel
#| tbl-cap: "Determinants of autocoup attempts and success (1945-2018)"
#| results: asis
#| echo: !expr format == "html"

autocoup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_model.csv")

model_attempt <- autocoup_model |> 
  filter(!regime %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential")) |>
  glm(autocoup_attempt ~ regime + GDP_pc + GDP_trend + polity5 + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

model_success <- autocoup_model |> 
  filter(!regime  %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential"),
         autocoup_outcome = factor(autocoup_outcome)) |>
  filter(autocoup_attempt == 1) |> 
  glm(autocoup_outcome ~ regime + GDP_pc + GDP_trend + polity5 + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

stargazer(model_attempt, model_success, 
          header = F,
          float = F,
          column.labels = c("Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant",
                               "Regime: Dominant-party",
                               "\\hspace{1.6cm}Military",
                               "\\hspace{1.6cm}Personalist",
                               "GDP per capita",
                               "Economic trend",
                               "Political 5",
                               "Political violence",
                               "Age",
                               "Population(log)"),
          dep.var.caption = "",
          dep.var.labels.include = F,
          summary.logical = T,
          column.sep.width = "50pt"
          )
```

@tbl-autocoupmodel summarizes the findings from the *`probit`* regression models
based on our analysis of the determinants of autocoup attempts and their success.

Model 1, which examines autocoup attempts, reveals only one significant predictor
besides the constant term. Among the regime types, personalist regimes
significantly increase the likelihood of autocoup attempts, all else being equal.
This suggests that leaders in personalist regimes are more prone to attempt to
extend their power through autocoups compared to leaders in democratic regimes.
Leaders in dominant-party and military regimes, however, show no significant
difference in the likelihood of attempting an autocoup compared to democratic
leaders.

The model for autocoup success (Model 2) shows similar dynamics. Personalist
regimes again have a strong positive and significant effect on the success of
autocoups compared to democratic leaders. Dominant-party regimes also show a
positive and marginally significant effect. However, a detailed examination
reveals that about half of the successful autocoups in dominant-party regimes (9
out of 20) exhibit a personalist style, such as "party-personal-military"
regimes.

This is logical since personalist leaders are typically much more powerful than
other types of leaders, making them more inclined and capable of overstaying in
power.

Other factors play an insignificant role in determining the attempts and outcomes
of autocoups. This aligns with our conclusions on the determinants of classic
coups. Both coups and autocoups are significantly affected by power dynamics. As
power transitions involve the struggle between seizing and maintaining power, the
balance of power status quo inevitably matters in both coups and autocoups. This
also explains the high success rate of autocoups. Compared to power challengers,
incumbents are in an obviously advantageous position. Incumbent leaders can use
state power to their benefit, which is difficult to counteract. Even the abuse of
power is often unchecked under a powerful leader's rule.

The threshold for removing or impeaching an incumbent leader is very high.
Without more than a majority or even greater support, it is very difficult to
succeed. Using illegal means such as a coup is even more challenging, as the
costs are high, the consequences are severe, and the likelihood of success is
very low.

### Conclusion

This study provides a comprehensive analysis of autocoups, specifically focusing
on political events where incumbent leaders illegitimately extend their tenure in
power. By refining the concept and distinguishing it from broader definitions,
such as 'self-coups', 'autogolpes', 'executive takeover', we introduce a novel
dataset of autocoups (1945-2022). Through this more precise definition and
dataset, we enhance the understanding and methods of analyzing the dynamics of
irregular power transitions and the survival of political leaders.

Our findings reveal that personalist regimes are far more likely to experience
autocoup attempts and are more likely to succeed compared to democracies.
Dominant-party systems, often exhibiting personalist characteristics, also show
an association with successful autocoups. While regime type significantly
influences autocoups, other factors appear less impactful, mirroring classic
coups where the balance of power is a more essential determinant. The high
success rate of autocoups can be attributed to the inherent advantages incumbents
possess, such as control over or abuse of state power and the difficulty of
removing or impeaching them through legal or illegal means.

However, there are several limitations that future research could address.
Firstly, the definition of an autocoup requires further commentary and discussion
to gain wider acceptance. Accordingly, the dataset of autocoups requires further
case-by-case verification to improve accuracy, although debates on some ambiguous
cases will likely persist. Secondly, due to the nature of autocoups, which are
less frequent than classic coups (491 coups versus 110 autocoups during the same
period), the quantitative analysis cannot be conducted as a country-year variable
as in coups. This raises the issue of choosing a base year. For example, when
analysing how GDP level or growth rate, the democratic index (such as Polity 5),
or the age of the leader affect autocoup attempts, we need to decide which year's
value should be used. In this study, we chose the middle year of a leader's
tenure or the year they staged the autocoup, but determining the most appropriate
year requires further discussion.

Despite these limitations, this research enhances our understanding of the
mechanisms and motivations behind autocoups, contributing to the literature on
political stability and democratic resilience. Future studies could build on this
work by employing the dataset to explore more power dynamics or examine the
long-term impacts of these events on political systems, particularly on
democratic backsliding, democratic breakdown, and personalization of power.

## Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-Entry and Autocoup Leaders

### Abstract {.unnumbered}

This study examines how the method of power acquisition impacts the longevity of
two main types of irregular-entry political leaders: coup-entry leaders and
autocoup leaders. We hypothesize that the method of accession significantly
affects leader tenure. Utilizing the Cox proportional hazards model and a
time-dependent Cox model, we provide compelling evidence of differing survival
times between these two leader types. The findings indicate that coup-entry
leaders face a significantly higher risk of removal compared to autocoup leaders.
Specifically, coup-entry leaders are found to have a much shorter average tenure
and a higher probability of being ousted. This study underscores the implications
of these findings for political stability and democratic processes, suggesting
that the relatively low cost and high returns of autocoups could incentivize more
incumbents to seize power in this manner, potentially leading to democratic
backsliding. Furthermore, this research contributes to the academic literature by
effectively utilizing a newly developed dataset on autocoups, offering valuable
insights into the dynamics of irregular leadership transitions.

keywords: ***Coups, Autocoups, Survival, Political Leadership, Cox Model***

\newpage

### Introduction

Why do some leaders rule for decades while others stay in power for only years,
months, or even days? This question has garnered substantial attention, making
the survival of political leaders a well-explored topic in political science.
Although the general framework of political survival has been extensively studied
and numerous theories have been proposed, specific types of leaders—particularly
coup-entry leaders and autocoup leaders (defined in more detail in Section
3)—have received comparatively less scrutiny. Nonetheless, examining the tenures
of these specific leader types is particularly significant, as it provides
insights into the dynamics of irregular leadership transitions and their impact
on political stability and democratic processes.

The processes of entry, tenure, and exit are more predictable for leaders who
come to power through regular channels, making their survival easier to
anticipate and thus less compelling to study. In contrast, leaders who ascend
through irregular means, such as coups or autocoups, attract more interest from
both academics and journalists due to the unpredictable and often tumultuous
nature of their tenures. According to @goemans2009, leaders with irregular
entries or exits are quite common. Between 1945 and 2015, more than half of the
leaders who entered power irregularly (158 out of 308) eventually exited
irregularly as well. This rate is significantly higher than that of leaders who
assumed office through regular channels, of whom only 14.5% (213 out of 1,472)
experienced irregular exits. While this percentage is still notable, it is
considerably lower than that for irregular-entry leaders.

Among leaders with irregular entry or exit, coup-entry and autocoup leaders
constitute the majority. According to @goemans2009, out of 374 leaders who exited
irregularly, 246 were ousted through coups, accounting for 65.8% of these cases.
@frantz2016 show that coup-related exits account for roughly one-third of all
exits in autocracies, surpassing any other type, including regular transitions.
Additionally, between 1945 and 2020, there were 106 autocoup attempts, with 86
being successful [@zhu2024].

```{r}
#| label: data
#| include: false
survival_cox_ph_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_ph_model.csv") |> 
  mutate(survival_year = survival_days / 365)

survival_cox_td_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_td_model.csv")

avg <- survival_cox_ph_model |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)
```

Due to their irregularity and inherent uncertainty, precisely measuring the
survival tenure of coup-entry and autocoup leaders presents a significant
challenge. Nonetheless, a comparative analysis of the tenures of these two types
of leaders is both feasible and valuable. Leaders who consolidate power through
autocoups have a longer average tenure post-autocoup (approximately
`r avg$avg[1]` years) compared to coup-entry leaders (approximately
`r avg$avg[2]` years). This suggests a potential shortfall of about 5 years in
the average tenure of coup-entry leaders.

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders
#| echo: !expr format == "html"

# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = survival_cox_ph_model)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = survival_cox_ph_model, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("Autocoup leaders", "Coup-entry leaders"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 12)
```

A preliminary analysis using a log-rank test in survival analysis, as shown in
@fig-logrank, reveals a distinct contrast in the tenures of autocoup leaders
versus coup-entry leaders. The survival curve for autocoup leaders is
consistently higher than that for coup-entry leaders, and the difference is
statistically significant. This indicates that autocoup leaders generally have a
longer survival tenure and a lower risk of being ousted compared to coup-entry
leaders.

We argue that the method of accession significantly influences leadership
longevity. Coup-entry leaders are likely to face greater challenges to their
rule, resulting in a shorter average tenure compared to autocoup leaders.

Using the Cox proportional hazards model and the time-dependent Cox model, our
results indicate that autocoup leaders generally experience longer tenures
compared to coup-entry leaders.

This study makes two potential contributions. Firstly, it sheds light on an
understudied factor in leadership survival analysis: the impact of the method of
accession to power. Leaders' survival is influenced not only by their ruling
strategies after taking power but also by how they acquired power initially.
Secondly, by utilizing survival models, this research provides empirical evidence
of the significant difference in tenure duration between autocoup and coup-entry
leaders. This finding may help explain the increasing prevalence of overstaying
in power through autocoups since 2000. As more incumbents observe these
precedents, they may be encouraged to adopt similar tactics, with ruling elites
tending to follow and support such actions to benefit from a relatively longer
tenure.

The remainder of this chapter delves deeper into this topic. Section 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Section 3 explores the factors influencing the survival of
coup and autocoup leaders. Section 4 outlines the methodology and data used,
including the application of survival models to analyze the determinants of
leadership longevity. Section 5 presents the findings of the analysis and a
detailed discussion of the results. Finally, Section 6 concludes by synthesizing
the key takeaways and exploring their broader implications.

### Literature review

The topic of political survival has been a cornerstone of political science
research for decades. This enduring interest arises from the wide-ranging
variations observed across regimes, countries, and historical periods. Previous
studies have identified two crucial yet interconnected aspects of political
survival: regime survival and individual leader survival.

Regime survival focuses on the longevity of political systems, such as
monarchies, political parties, or specific ideological structures. Leader
survival, on the other hand, is concerned with the duration of individual
leaders' time in office. These two concepts often exhibit contrasting patterns.
For example, in parliamentary democracies like Japan or the UK, specific
political parties may hold power for extended periods while individual leaders
(Prime Ministers) change frequently. Similarly, communist regimes typically see
long-lasting parties in power, with leadership transitions occurring at a more
frequent pace. In contrast, presidential systems like the United States or some
military regimes experience more frequent changes in both the ruling party or
junta and the leader of the country. This study specifically investigates the
dynamics of individual leader survival, focusing on the factors influencing how
long leaders remain in power.

The existing literature on leader survival is vast and multifaceted, offering a
wealth of insightful perspectives. Some studies delve into the specific
mechanisms that influence leadership longevity within particular regimes, such as
democracies [@svolik2014] or autocracies [@davenport2021]. Others seek to develop
more generalizable theoretical frameworks that can explain leader survival across
diverse political systems [@buenodemesquita2003]. While the development of a
universal theory remains an alluring goal, it is important to acknowledge the
inherent challenges in creating a single model that encompasses the complexities
of leadership survival across all regime types.

The mechanisms governing power transitions vary significantly across different
regimes, particularly between democracies and autocracies. In many autocratic
systems, leadership selection is a closed affair. Potential candidates are often
restricted to a narrow pool, such as royal families, military elites, or members
of the ruling party. While political competition and elections may exist in some
autocracies, there are often significant barriers to entry for legitimate
challengers. Potential rivals may face threats such as assassination,
imprisonment, or exile. Moreover, selection processes are shrouded in secrecy,
with outsiders unable to participate or obtain information on the selections, let
alone express dissent or complain about the results. This lack of inclusivity and
transparency makes it challenging to gauge true levels of public support compared
to democracies. Consequently, calculating selectorates or winning coalitions, as
explored by @buenodemesquita2003, becomes a near-impossible task in autocracies.

Therefore, focusing research on more specific regimes or types of leaders may be
more appropriate. While regular and anticipated leadership changes are important
as well, they offer less fertile ground for exploring the dynamics of leader
longevity, as the vast majority of leaders who assume power through established
channels also exit power through established mechanisms [@goemans2009]. In
contrast, the study of political survival among irregular leaders is particularly
captivating due to the intricacies and uncertainties associated with irregular
leadership transitions.

The diverse mechanisms influencing political leadership survival, as discussed
above, have prompted scholars to delve deeper into more specific determinants.
Two primary perspectives have emerged to explain the dynamics of leader survival.

The first perspective considers objective factors and resources available to
leaders. These include elements such as personal competence [@yu2016], societal
stability [@arriola2009], economic development [@palmer1999; @williams2011],
access to natural resources [@smith2004; @quirozflores2012; @wright2013], or
external support networks [@licht2009; @wright2008; @thyne2017].

The second perspective explores subjective factors, focusing on the strategies
leaders employ to consolidate their power. This encompasses both the formulation
and implementation of political policies and the leaders' responses to
opposition, challenges, or even coups and rebellions [@gandhi2007; @morrison2009;
@escribà-folch2013; @davenport2021].

Unsurprisingly, coups have garnered significant scholarly attention due to their
pivotal role in removing leaders [@svolik2009; @frantz2016]. Existing research
delves into strategies for thwarting coups [@powell2017; @sudduth2017;
@debruin2020], as well as how leaders extend their tenures after surviving coup
attempts [@easton2018]. For instance, @sudduth2017 examines the post-coup actions
of dictators, focusing on purge strategies. They argue that coup leaders
initially wield more power than elites within the regime, rendering them less
susceptible to subsequent coups, challenging the conventional view of new leaders
as inherently weak [@roessler2011]. Meanwhile, @sudduth2018 investigates how
leaders' entry methods affect their removal in dictatorships, positing that
irregular entry does not necessarily increase the likelihood of removal; in fact,
some forms of irregular entry may offer protection.

While scholars have extensively analysed leader survival across various contexts,
including universal frameworks, autocratic regimes, and the aftermath of failed
challenges, a significant gap persists. There is a lack of research specifically
exploring and comparing the survival tenures of leaders who extend their reigns
through autocoups compared to coup-entry leaders. This study aims to address this
gap by investigating and comparing the duration of leadership survival between
these two distinct leader types.

### Survival dynamics of autocoup and coup-entry leaders

#### Autocoup leaders versus coup-entry leaders

As emphasized in previous section, delving into leadership survival poses
inherent challenges, given factors such as the opacity and diverse mechanisms of
power transitions. However, these challenges underscore the significance of this
research, as it sheds light on understudied dynamics.

While the survival of political leaders manifests complexity and variation, it is
not devoid of patterns entirely. Leaders of similar types often exhibit
significant comparability. Before delving into the comparison, it is necessary to
clarify several relative terminologies.

Firstly, we define an autocoup as a situation where an incumbent leader utilizes
illegitimate or unconstitutional methods to extend their tenure and remain in
power. An autocoup is deemed successful if this power extension lasts for at
least six months. Coups, on the other hand, are defined as illegal and overt
attempts by the military or other powerful groups within the state to unseat the
sitting leader [@powell2011]. Similar to coups, successful autocoups in this
study will be measured by a power usurpation lasting at least six months,
differing from the seven-day duration coded by Powell and Thyne.

Secondly, it is also crucial to clarify the distinction between an autocoup
leader and a coup-entry leader, as the survival of leadership is the main concern
of this study.

-   **Autocoup leader:** This refers to an incumbent leader who successfully uses
    illegitimate or unconstitutional means to extend their tenure in power. In an
    autocoup, the leader orchestrates the power grab and continues to rule
    afterwards.

-   **Coup-entry leader:** This term designates the individual who assumes power
    after a successful coup. The coup leader and the coup-entry leader may or may
    not be the same person. Unlike in autocoups, coups often involve multiple
    leaders (individuals or groups) who overthrow the incumbent leader, but
    typically only one of them assumes supreme power. In some instances, coup
    leaders may support someone outside the coup plot to become the new leader.
    For example, military officers might return power to civilians after a coup
    or support a new general election. Regardless of the specific scenario, a
    coup-entry leader in this study refers to the individual who assumes formal
    leadership following a successful coup.

Given that autocoup leaders typically exhibit longer overall tenures compared to
coup-entry leaders, this study focuses on a more nuanced comparison.
Specifically, we will analyse the **post-autocoup** tenure of autocoup leaders
and contrast it with the **post-coup** tenure of coup-entry leaders. The
examination of the survival tenures of coup-entry leaders and autocoup leaders is
motivated by their relevance and similarity in terms of illegitimacy,
uncertainty, and instability.

#### Hypothesis

Previous research emphasizes that skilful power retention is the ultimate
determinant of leader longevity. Leaders who can maintain control or manipulate
the balance of power tend to stay in office longer. However, as discussed
earlier, although both coup-entry and autocoup leaders face similar challenges,
the intensity of their challenges related to illegitimacy, uncertainty, and
instability differs. These discrepancies create an uneven playing field in terms
of power dynamics, with coup-entry leaders at a significant disadvantage. This
disparity in power equilibrium shapes a nuanced landscape that profoundly impacts
the duration of leader tenures.

-   **Illegitimacy:** Both leader types lack legitimacy, though it manifests
    differently. Coup leaders seize power through force or the threat of force,
    making their illegitimacy explicit. Autocoups often employ seemingly legal
    procedures, but these processes are frequently manipulated by incumbents
    leveraging their control [@zhu2024]. This perception of illegitimacy can be
    used to justify the removal of autocoup and coup-entry leaders, even if the
    means of removal are themselves illegitimate.

    While both coup-entry and autocoup leaders lack genuine legitimacy, the
    nature of this illegitimacy differs significantly. Coups represent a more
    blatant disregard for legal processes, often involving force or the threat of
    force to seize power. In contrast, autocoups employ a façade of legality
    through tactics such as manipulating constitutional interpretations,
    engineering parliamentary votes, influencing court decisions, and even
    holding referendums. Despite this veneer, these manoeuvres do not erase the
    underlying lack of genuine legitimacy, hence this study categorizes them as
    autocoups. However, there is considerable debate among scholars, with some
    arguing for less severe terms like "incumbent overstay" or "executive
    takeover" due to the perceived legality.

    This perceived legitimacy can provide a temporary advantage for autocoup
    leaders. Challengers are often constrained to operate within legal
    frameworks, making it difficult to directly confront the incumbent.

-   **Uncertainty:** The tumultuous paths to power undertaken by coup-initiators
    and autocratic leaders cast a shadow of uncertainty over their reigns and
    eventual departure. Their ascension through irregular means undermines
    established power transition norms, leaving doubts lingering over their
    commitment to constitutional succession protocols. This uncertainty not only
    unsettles elites and citizens but also plagues the leaders themselves, who
    grapple with the perpetual ambiguity surrounding the transfer of power –
    when, how, and to whom. Historical analyses underscore this predicament, with
    data revealing that more than two-thirds of irregular exits from leadership
    stem from coup-related upheavals [@goemans2009].

    Coup-entry and autocoup leaders face different levels of uncertainty
    immediately following their rise to or overstay power.

    After a coup, at least three major uncertainties arise regarding leadership
    and its tenure. First, it is unclear who will assume leadership. Although
    coup leaders often take power, some may return or promise to return power to
    civilian leaders. Even among coup leaders, determining who will lead can be
    problematic, as coup plotters are sometimes a group without a clear core
    leader. For instance, following the 1973 Chilean coup, the initial plan for a
    rotating presidency among military leaders was abandoned when General
    Pinochet consolidated control and remained in power until 1990 [@svolik2014].
    Second, the duration of the coup leader's rule is uncertain. Leaders like
    Gamal Abdel Nasser in Egypt (1954 coup), Muammar Gaddafi in Libya (1969
    coup), and Idi Amin in Uganda (1971 coup) aimed to retain power for life
    [@geddes2018], but their ability to do so was uncertain. Others promise to
    transfer power to civilian authorities, but the timing and fulfillment of
    these promises are unclear. For example, Myanmar's military junta (2021 coup)
    has repeatedly extended a state of emergency, clinging to power beyond the
    promised time-frame[^7]. Conversely, after the 2010 coup in Niger, the
    military honoured their promise by restoring civilian rule within the same
    year [@ginsburg2019]. Third, the successors of coup leaders are uncertain.
    Some may designate successors from their inner circle, including family
    members, while others may support general elections, though whether this will
    be fulfilled as intended remains uncertain.

    In contrast, autocoup leaders present a clearer picture regarding leadership
    and tenure. There is no ambiguity about who will rule after an autocoup. In
    the medium term, autocoup leaders typically hold office themselves. Many,
    like Putin in Russia and Xi Jinping in China, seek to extend their rule
    indefinitely and are unlikely to relinquish power voluntarily. Others attempt
    to extend their terms incrementally, such as President Menem of Argentina,
    who overstayed until 1993 but failed in his bid for another term in 1999
    [@llanos2019].

-   **Instability:** Moreover, the awareness of their shaky legitimacy and the
    persistent uncertainty breeds insecurity and a perpetual sense of crisis
    among coup-entry leaders and autocoup leaders. In a bid to solidify their
    grip on power, they often resort to reshaping power dynamics or purging
    potential adversaries. Paradoxically, these attempts to bolster stability and
    fortify their rule frequently boomerang, unleashing even greater turmoil and
    instability.

    The stability of a regime, particularly in an autocracy, hinges on
    maintaining a balance of power. Coups, however, inevitably disrupt this
    balance, even when they are bloodless, necessitating the creation of a new
    equilibrium. The ousting of previous rulers requires dismantling the
    established governing structure and reshuffling high-ranking officials,
    actions that inherently generate instability and create adversaries for the
    new leadership. This makes restoring order and establishing a balanced power
    structure notably challenging. Studies show that new leaders often purge
    rival elite groups to consolidate their power at the outset of their tenure
    [@sudduth2017; @roessler2011]. Such actions can provoke backlash even from
    close allies. For instance, in Uganda, President Obote's attempt to undermine
    the army commander-in-chief, Idi Amin, led to Amin gaining the army's support
    and ultimately ousting Obote in a 1971 coup. Similarly, in Pakistan in 1999,
    shortly after Prime Minister Sharif dismissed powerful army chief General
    Pervez Musharraf, Sharif himself was ousted in a coup orchestrated by
    Musharraf and his supporters [@sudduth2017].

    To consolidate power, coup-entry leaders often have to compromise with
    internal or external power challengers. However, these compromises are
    frequently unstable and easily broken. The situation becomes even more
    complex when there is a risk of civil war. Leaders may attempt to reduce the
    likelihood of subsequent coups, potentially increasing the chances of
    societal rebellions and civil wars [@roessler2011].

    Moreover, instability extends beyond leadership to policies. A new leadership
    group often brings new policies, and coups are sometimes triggered by
    disagreements over significant policies. Major policy shifts can instigate
    dissent or grievances from various ruling factions, communities, regions,
    ethnicities, or religions.

    In contrast, autocoup leaders encounter far fewer of these issues, as their
    regimes experience fewer abrupt changes. They face less pressure to dismantle
    the existing ruling paradigm and establish a new order. Even when adjustments
    are necessary, they have more time to implement changes gradually.

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-entry leaders 
#| echo: !expr format == "html"
#| results: asis
# library(kableExtra)
read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(100),
    2:3 ~ px(250)
  ) |>
   data_color(rows = c(2, 4), colors = "#edede9") 
```

In conclusion, coup-entry leaders face a significantly greater degree of the
challenges discussed earlier, placing them at a substantial disadvantage in the
power dynamic compared to autocoup leaders. This disadvantage creates a
self-perpetuating cycle. Weaker leaders struggle to attract and retain strong
support, making them more vulnerable to internal and external challenges. The
perception of risk discourages potential allies, further eroding their power
base.

Empirical evidence bolsters this dynamic. Data reveals a correlation between the
frequency of coup attempts in a country and the likelihood of future coups (e.g.,
@tbl-coups shows over a third of coups occurring in the top ten countries with
the most attempts since 1950). This suggests that the more coups occur in a
country, the more likely additional coups are to happen in the future.

Conversely, autocoup leaders, often benefiting from a veneer of legitimacy and a
stronger initial position, are better able to consolidate power and attract
supporters. This advantage can be self-reinforcing, as a strong power base
discourages challenges and fosters loyalty. This dynamic is evident in cases like
China (2018), where the National People's Congress granted Xi Jinping the
potential to rule for life[^8], and Russia (2020), where constitutional changes
allow Putin to potentially remain in power until 2036[^9].

These features and the balance of power contribute to a shorter expected tenure
for coup-entry leaders compared to the relatively longer tenures of autocoup
leaders. The average survival period following an autocoup is approximately five
years longer than that of coup-entry leaders (@fig-logrank). Based on these
observations and discussions, I propose the following hypothesis:

> ***H1: Political leaders who successfully extend their tenure through autocoups
> are more likely to experience longer leadership survival compared to coup-entry
> leaders.***

In the following section, I will delineate the research methodology employed in
this chapter. To assess the persistence and robustness of the hypothesis, I will
introduce a series of control variables.

### Research Design

#### Methodology: Survival analysis

To test the hypothesis, I will employ two Cox models to analyse the survival
tenures of coup-entry leaders and autocoup leaders. Unlike the Kaplan-Meier
model, the Cox model enables the estimation of the impacts of multiple factors.
Although it does not directly estimate the duration of tenure in office, it
evaluates the hazard rate associated with being ousted from power. Essentially,
this represents different facets of the same phenomenon: as a leader's cumulative
hazard of being ousted increases, their probability of survival in office
decreases.

The first model will utilize the Cox proportional hazards model (referred to as
the Cox PH model), where I will only use the variables present at the entry year,
without considering the changes in these variables over the leaders' survival
times.

However, apart from the primary variable of interest in this research—the leader
type—control variables such as economic performance, Polity5 scores, and
political stability do change over time. Therefore, the second model will account
for these variations by using the time-dependent Cox model.

#### Dependent variables

-   **Survival Time:** The duration of a leader's tenure, measured in days. For
    coup-entry leaders, the survival time begins on the day they assume power
    through a coup. For autocoup leaders, the survival time starts on the
    expiration date of their original legitimate term. For example, Xi Jinping
    assumed power in 2013 and removed term limits in 2018. His original
    legitimate tenure was set to end in 2023, so his survival time begins in
    2023, not 2013 or 2018, marking the start of his post-autocoup tenure. The
    survival time concludes on the day the leader exits office, applicable to
    both coup-entry and autocoup leaders.

-   **End point status:** This variable indicates the manner in which the
    leader's tenure concluded, categorized as follows:

    -   **0 = Censored:** This status is assigned to leaders who leave office
        through means other than being ousted. This includes leaders who appoint
        their successors, opt for democratic transitions, or leave office due to
        health issues or natural death.

    -   **1 = Ousted:** This status is assigned to leaders who are forced out of
        office. This includes cases where leaders "voluntarily" resign under
        pressure, reflecting a de facto ousting by force.

#### Key Independent variable: Leader type

This variable categorizes leaders into two distinct groups:

-   Group A = Autocoup Leader: Leaders who extend their tenure through autocoups.
-   Group B = Coup-Entry Leader: Leaders who assume power through coups.This
    variable is the primary independent variable of interest, serving as the
    basis for comparing the survival time between these two types of leaders.

The data for both dependent and independent variables are sourced from @zhu2024
and @goemans2009.

#### Control variables

-   **Economic Performance:** This variable is measured using two indicators:
    economic level and economic growth rate.

    -   **Economic Level:** Represented by GDP per capita. This measure provides
        an indication of the overall economic health and standard of living in a
        country.

    -   **Economic Growth Trend:** Assessed using the current-trend (CT) ratio,
        developed by @krishnarajan2019, which is consistent with @eq-eq6.

The GDP per capita data, expressed in constant 2017 international dollars (PPP)
and measured in units of \$10,000, is sourced from the V-Dem dataset by
@fariss2022. To account for the economic impact of the previous year, this data
is lagged by one year.

-   **Political Stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    Major Episodes of Political Violence dataset by Marshall. This index provides
    a comprehensive measure of the level of violence and conflict within a
    country, which can significantly impact leadership survival.
    [@marshall2005current].

-   **Degree of Democracy:** The level of democracy is gauged using Polity 5
    scores at the entry year for each respective country. These scores range from
    -10 (fully autocratic) to +10 (fully democratic), capturing the extent of
    democratic versus autocratic governance. This dataset is sourced from the
    Center for Systemic Peace (CSP) and provides an essential measure of
    political regime type, which can influence the stability and survival of
    leaders in power.

-   **Population Size:** To account for its potential impact on leaders' tenures,
    the log of the population size is considered. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on leadership survival. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's Age:** The age of the leader at the entry year is included as an
    additional variable in the analysis, offering insights into potential
    correlations with leadership survival. Older leaders may have different
    experiences, networks, and health considerations that could influence their
    ability to maintain power. This data is sourced from the leaders dataset by
    [@goemans2009].

### Results and discussion

#### Model results

Using the @survival package in R, we present the regression results for both the
Cox PH model and the Time-dependent Cox model in Table @tbl-cox.

```{r}
#| label: model
#| echo: !expr format == "html"

Mydata1 <- survival_cox_ph_model |> 
   mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")))

model_cox1 <- Mydata1 |> 
  coxph(
    Surv(survival_days, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age,
    data = _
  ) 

tbl1 <- model_cox1|> 
   tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T, hide_se = F, hide_ci = T) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 

Mydata2 <- survival_cox_td_model |> 
  mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")),
         survt = (tstop - tstart) * 365)

model_cox2 <- Mydata2 |> 
  coxph(
    Surv(T1, T2, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age + cluster(id),
    data = _
  ) 
tbl2 <- model_cox2|>
  tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T, hide_se = F, hide_ci = T) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox models for survival time of different types of leaders
#| echo: !expr format == "html"

tbl_merge(list(tbl1, tbl2),
          tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**")) |> 
  as_gt()
```

Generally speaking, both the Cox PH model and time-dependent Cox model analyses
revealed a statistically significant association between leadership type and the
hazard of removal from power. Coup-entry leaders were found to have a hazard
ratio of 2.42 (in time-dependent Cox model) compared to autocoup leaders
(reference group), assuming all other variables in the model are held constant.
The analysis suggests that coup-entry leaders face a significantly greater risk
of removal from power compared to autocoup leaders. At any given time during
their tenure, coup-entry leaders are 2.42 times more likely to be ousted from
power compared to autocoup leaders, all else being equal in the model.

All other control variables in Cox PH model are not statistically significant. In
time-dependent Cox model, however, GDP per capita and the log of population are
both statistically significant at 5% level.

-   **GDP per capita:** The hazard ratio of 0.94 indicates that for each unit
    increase in GDP per capita (measured in constant 2017 international dollars,
    PPP, in units of \$10,000), the hazard (or risk) of being ousted at any given
    time is reduced by 6%, assuming all other variables in the model are held
    constant. Similarly, it indicates that for each unit increase in the log of
    population, the hazard (or risk) of being oustedat any given time is reduced
    by 14%, assuming all other variables in the model are held constant.

-   **Population:** The hazard ratio of 0.86 indicates that for each unit
    increase in the logarithm of the population (one percent increase of
    population), the hazard (or risk) of being ousted at any given time decreases
    by 14%, assuming all other variables in the model are held constant.

#### Discussion

```{r}
#| label: fig-coxSurv
#| fig-cap: Survival curves for Cox Model
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: !expr format == "html"

testdata1 <- Mydata1 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox1 |> 
  survfit(newdata = testdata1) |>
  ggsurvplot(data = testdata1,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))

testdata2 <- Mydata2 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox2 |> 
  survfit(newdata = testdata2) |>
  ggsurvplot(data = testdata2,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))
```

The survival curves depicted in @fig-coxSurv illustrate the survival rates for
leaders of both types. Both the Cox Proportional Hazards (Cox PH) model and the
time-dependent model produce similar plots. Notably, the survival curve for
coup-entry leaders exhibits a significantly lower trajectory compared to that of
autocoup leaders. The steeper drop at the early stage for coup-entry leaders
indicates they are more likely to be ousted shortly after assuming power.
Additionally, the survival curve for coup-entry leaders crosses the median
survival line much earlier than that of autocoup leaders. This disparity suggests
that autocoup leaders tend to remain in power for longer durations than their
coup-entry counterparts.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard ratios and 95% CIs for Leader Ousting
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: !expr format == "html"

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw() 

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |> 
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw() 
```

@fig-coxHR displays the hazard ratios and corresponding 95% confidence intervals
for the variables incorporated in the Cox model. Both the Cox Proportional
Hazards (PH) model and the time-dependent model yield nearly identical plots. The
hazard ratio for coup-entry leaders compared to autocoup leaders is significantly
greater than 1 and statistically significant, indicating that coup-entry leaders
are at a higher risk of being ousted. Most other variables have hazard ratios
close to 1, suggesting that a one-unit increase in these variables does not
significantly affect the risk of being ousted. Although the hazard ratio for GDP
growth trend is considerably less than 1, it is not statistically significant at
the 5% level. However, it is statistically significant at the 10% level,
suggesting that better economic performance may help to consolidate the rule of
the incumbents to some extent.

@fig-coxHR presents the hazard ratios (HRs) and their corresponding 95%
confidence intervals (CIs) for the variables included in the Cox proportional
hazards model (Cox PH) and the time-dependent model. The HR for coup-entry
leaders compared to autocoup leaders is significantly greater than 1, indicating
a substantially higher risk of removal from power for coup-entry leaders. Most
other variables have HRs close to 1, suggesting that a one-unit increase in those
variables has minimal impact on the risk of removal.

The HR for GDP growth trend is noteworthy. While it is considerably less than 1,
it is statistically significant at the 10% level (0.10 \> p-value \> 0.05). This
suggests a possible trend where positive economic performance might reduce the
risk of removal from power for the incumbents, although the evidence is not
conclusive.

#### Assessing the Proportional Hazards Assumption

The validity of the model assumptions significantly influences our confidence in
the results. To assess the proportional hazards assumption of the Cox model, we
use the chi-square test based on Schoenfeld residuals to determine whether the
covariate effects remain constant (proportional) over time. Although the Cox PH
model violates the proportional hazards assumption, our primary analysis relies
on the time-dependent Cox model, which does not show strong evidence of violating
the proportional hazards assumption for any covariate. The global p-value of
0.382 is much greater than the 5% significance level, indicating that the
proportional hazards assumption is reasonably met for the time-dependent Cox
model.

```{r}
#| label: robust_test
#| echo: false
test <- cox.zph(model_cox2)
```

#### Conclusion

This study examined the survival durations of political leaders who come to power
through irregular means, specifically coups and autocoups. We hypothesized that
the mode of accession significantly influences leader tenure. Employing survival
analysis techniques, including the Cox proportional hazards model and a
time-dependent Cox model, we found strong evidence that autocoup leaders
generally enjoy longer tenures than coup-entry leaders.

Our findings revealed a significant difference in average tenure, with
post-autocoup leaders averaging approximately 10.8 years in power compared to 5.4
years for coup-entry leaders. The time-dependent Cox model further indicated that
coup-entry leaders are 2.42 times more likely to be ousted from power at any
given time compared to autocoup leaders, all else being equal.

These results highlight the importance of understanding the phenomenon of
autocoups, where leaders extend their rule by manipulating legal frameworks. Due
to the relative ease and potential benefits of autocoups, this method of power
retention might incentivize more leaders to employ it. Consequently, democratic
backsliding could become more prevalent as autocoups weaken democratic
institutions and constitutional norms, particularly in nascent democracies or
those transitioning from autocracy.

This study contributes to the field of leadership survival by demonstrating that
the mode of accession significantly impacts leader tenure, a factor previously
under-explored in the literature. By utilizing both Cox models, the research
offers robust analytical techniques for studying political leadership survival
and provides strong evidence of divergent tenure lengths between these two types
of irregular-entry leaders.

However, limitations exist. The study relies heavily on the autocoup dataset
collected and coded by the author. The concept and data itself are relatively
novel within academia. Future research should refine and establish wider
recognition for the term "autocoup," leading to more accurate and comprehensive
data collection efforts. Expanding the dataset to include more cases and
integrating it with data on other irregular leadership transitions could yield a
more holistic understanding of political survival in such contexts.

## Conclusion

### Main Findings

This study delves into the dynamics and implications of irregular power
transitions, focusing on coups and autocoups. The findings illuminate the complex
interplay between incumbents and challengers fighting for power.

Firstly, our analysis reveals that the expected success rate of a coup attempt
significantly influences its likelihood. This success rate is heavily influenced
by the balance of power between the incumbent regime and challengers, which is
largely determined by regime type. We find that military regimes, although with
more control over their own military forces, face a higher risk of coups compared
to dominant-party regimes.

Secondly, the study introduces a redefined concept: the autocoup. Defined as an
incumbent leader's refusal to relinquish power as mandated, this research
distinguishes autocoups from broader terms like self-coups. Based on this
definition, we present the first publicly available dataset of autocoup events
from 1945 to 2022, encompassing 110 attempts and 87 successful autocoups. Case
studies and empirical analyses demonstrate the dataset's utility for quantitative
research, providing a robust foundation for further analysis on autocoups.

Thirdly, employing survival analysis techniques, the study finds clear
differences in leader longevity between those who come to power through coups and
those who extend their rule through autocoups. The results indicate that
coup-installed leaders face a significantly higher risk of removal compared to
autocoup leaders who manipulate the system to extend their rule.

### Policy Implications

The findings of this study offer valuable insights for policy-makers concerned
with promoting and protecting global democracy, which has faced increasing
challenges despite a general post-WWII trend towards democratization. Notably,
the "third wave" of democratization [@huntington1991democratization] witnessed a
surge in democratic transitions in the late 20th century. Since the Cold War's
end, democratic nations have outnumbered non-democratic ones (@fig-democracy)
with the gap widening.

```{r}
#| label: fig-democracy
#| fig-cap: Comparison of the number of democratic and non-democratic countries (1945-2020)
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| echo: !expr format == "html"
REIGN |>
  filter(year > 1945 & year < 2021) |> 
  mutate(democracy = case_when(
    gwf_regimetype %in% c("parliamentary", "presidential") ~ "Democracy",
    .default = "Non-democracy")) |> 
  summarise(Count = n(), 
            .by = c(year, democracy)) |> 
  #view()
  ggplot(aes(year, Count, color = democracy)) +
  geom_point() +
  labs(color = 'Regime type',
       x = "Years: 1945 ~ 2020",
       y = "No. of Countries") +
  theme(legend.position = "top") +
  annotate("text",
    label = "Data source: REIGN (Regime data)",
    x = 2010,
    y = 30,
    fontface = "italic",
    size = 3
    )
```

However, a "democratic recession" has emerged in recent years [@diamond2008].
Freedom House reports an 18th consecutive year of global freedom decline in 2023
[@freedomhouse2024freedom]. While few countries have completely regressed to
autocracy, the average global democracy level has fallen back to pre-2000 levels.
Notably, democratic backsliding often occurs within regimes, with democracies
becoming less liberal and autocracies becoming less competitive [@mechkova2017].

This research highlights irregular power transitions as a significant factor in
democratic backsliding within regimes. These transitions, often coups or
autocoups, violate democratic norms and disrupt the path towards stable
democracies. Leaders who gain power through irregular means often resort to
undemocratic tactics to maintain control, creating a vicious cycle of eroding
democratic institutions.

Our findings suggest that the shorter lifespans and potentially severe
consequences associated with coups may deter potential coup leaders. Conversely,
autocoups appear to be a more tempting option for power-hungry leaders due to
their higher success rates, seemingly moderate consequences, and extended leader
tenure after the autocoup. This trend may explain the decline in classic coups
since the 1990s alongside the rise of autocoups [@bermeo2016].

### Limitations and directions for future research

This study offers a novel framework for analysing irregular power transitions,
but some limitations require further exploration:

-   **Data refinement:** Defining and classifying autocoups is a new approach.
    Future research should validate this classification system through additional
    studies and expert evaluations.

-   **Data harmonization:** The current analysis faces challenges due to
    mismatched units (country-year vs. leader) between coup and autocoup
    datasets. Future efforts should explore data harmonization techniques for
    more robust comparisons.

-   **Democratic backsliding:** While this study establishes a connection between
    irregular power transitions and democratic backsliding, further empirical
    evidence is needed to solidify this link.

Several avenues exist for future research:

-   **Terminology and data collection:** Refining the "autocoup" concept and
    achieving wider recognition will facilitate more accurate and comprehensive
    data collection.

-   **Dataset expansion:** Expanding the autocoup dataset with more cases and
    integrating it with data on other irregular leadership transitions can
    provide a more holistic view of political survival after these events.

-   **Power dynamics and long-term impacts:** Utilizing this dataset, future
    studies can delve deeper into power dynamics at play and explore the
    long-term consequences of irregular transitions on political systems,
    particularly regarding democratic backsliding, breakdown, and personalization
    of power.

In conclusion, this study sheds light on the dynamics of irregular power
transitions, specifically focusing on coups and autocoups. By redefining
autocoups, classifying the dataset, analysing determinants, and comparing leader
longevity, we establish a framework for understanding irregular transitions and
leader survival. This work contributes to a deeper understanding of democratic
resilience and political stability. Future research can build upon this
foundation by conducting further empirical analyses based on the novel autocoup
dataset and continuing to refine the framework.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false

cat("\\newpage
## References")
```

[^1]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^2]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^3]: The definitions and concepts of power expansion and power extension can be
    vague. In this study, we define power expansion as an incumbent acquiring
    additional authority from other state apparatuses, whereas power extension
    refers to an incumbent prolonging their tenure beyond the designated term in
    office.

[^4]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^5]: The definitions and concepts of power expansion and power extension can be
    vague. In this study, we define power expansion as an incumbent acquiring
    additional authority from other state apparatuses, whereas power extension
    refers to an incumbent prolonging their tenure beyond the designated term in
    office.

[^6]: Center for Systemic Peace: <https://www.systemicpeace.org>. Accessed on
    `r Sys.Date()`.

[^7]: https://thediplomat.com/2023/08/myanmar-junta-extends-state-of-emergency-for-fourth-time/:
    Myanmar\
    Junta Extends State of Emergency for Fourth Time. Accessed on `r Sys.Date()`.

[^8]: https://www.bbc.co.uk/news/world-asia-china-43361276: China’s Xi allowed to
    remain ‘president for life’ as term limits removed. Accessed on
    `r Sys.Date()`.

[^9]: https://www.ucl.ac.uk/news/2020/jul/analysis-vladimir-putin-secures-constitutional-changes-allowing-him-\
    rule-until-2036: Analysis: Vladimir Putin secures constitutional changes
    allowing him to rule until 2036. Accessed\
    on `r Sys.Date()`.
