---
title: "Irregular Leadership Transitions and Political Survival: An Analysis of Coup-entry leaders and Autocoup leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: survival1.html
categories: [Article, Coups]
bibliography: references.bib

abstract: |
  This study explores and compares the survival time of two categories of irregular-entry political leaders: those who ascend to power through coups and those who overstay their legitimate term limits (autocoups). The central hypothesis posits that the mode of accession to power influences the duration of their tenures. Utilizing survival analysis, the results indicate that leaders who overstay their terms through autocoups generally remain in power longer than those who come to power through coups.

keywords: [Coups, Autocoups, Survival, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(survival)
library(survminer)
library(ggsurvfit)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(stargazer)
library(here)
```

## Introduction

The survival of political leaders has been a well-studied topic in political
science. However, two specific types of leaders - those who come to power through
coups and those who overstay their legitimate terms (autocoup leaders) - have
received comparatively less attention. Examining the tenures of these irregularly
ascended leaders holds particular significance for two main reasons.

Firstly, leaders who ascend through irregular means constitute the majority of
irregular exits from power. According to @goemans2009, between 1945 and 2015, out
of 1,472 leaders who assumed office through regular channels, approximately 213
exited irregularly (14.5%). Conversely, out of 308 leaders who assumed office
through irregular means, roughly 158 (51.3%) experienced irregular exits.

Secondly, among irregularly ascended leaders, the majority gained power through
launching coups or overstaying in office. As per @goemans2009, out of 374 leaders
who exited irregularly, 246 were ousted through coups, constituting 65.8% of
these cases. Additionally, between 1945 and 2020, there were 106 attempts to
overstay in power, of which 86 were successful [@zhu2024]. Leaders who extend
their legitimate term limits essentially stage coups against legitimate
successors, thereby qualifying as autocoup leaders. Hence, it is both relevant
and insightful to delve into and compare the tenures of coup-entry leaders and
autocoup leaders.

While it is expected that autocoup leaders survive longer than coup-entry leaders
when considering their total tenures, a more insightful comparison arises when
examining the tenures following autocoups in contrast to those of coup-entry
leaders. Conducting a log-rank test in survival analysis on the leaders dataset
[@goemans2009] and the author's autocoup dataset @zhu2024\] reveals a distinct
contrast between autocoup leaders and coup-entry leaders.

```{r}
#| label: data
#| echo: !expr format == "html"
#| include: !expr format == "html"

Mydata1 <- read.csv(here("data/mydata1.csv")) |> 
  mutate(survival_year = survival_time / 365)

avg <- Mydata1 |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)

avg
```

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders
#| echo: !expr format == "html"

# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = Mydata1)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = Mydata1, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("entry by autocoup", "entry by coup"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 14)
```

Preliminary findings from survival analysis (see @fig-logrank) indicate a
significant difference. Leaders who consolidate power through autocoups appear to
have a longer average tenure after the autocoup itself (approximately
`r avg$avg[1]`) compared to coup leaders (approximately `r avg$avg[2]`). This
suggests a potential shortfall of 5 years in the average tenure of coup leaders.

Thus, we hypothesize that the method of accession significantly influences
leadership longevity. Coup leaders likely face greater challenges to their
legitimacy, leading to a more precarious initial period in power. They may
encounter higher levels of uncertainty, instability, and pressure to share power,
all of which could contribute to a shorter average tenure compared to autocoup
leaders.

By employing the Cox proportional hazards model and the extended Cox model, the
results suggest that leaders who exceed their term limits generally experience
longer tenures compared to those who come to power through coups.

This study offers two key contributions. First, it highlights a critical factor
often overlooked in leadership survival studies: the impact of accession method.
Leaders' survival is not solely determined by their actions after taking power,
but also by the way they acquired power in the first place. This research
underscores the significant difference in tenure duration between autocoup and
coup leaders. Second, it provides empirical measurements to compare the tenure
duration of these two irregularly ascended leaders, offering insights into their
distinct impacts on leadership longevity.

The remainder of the paper delves deeper into this topic. Chapter 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Chapter 3 explores the factors influencing the survival of
coup and autocoup leaders. Chapter 4 outlines the methodology and data used,
including the application of survival models to analyse the determinants of
leadership longevity. Chapter 5 presents the findings of the analysis and a
detailed discussion of the results. Finally, Chapter 6 concludes by synthesizing
the key takeaways and exploring their broader implications.

## Literature review

The topic of political survival has been a cornerstone of political science
research for decades. This enduring interest stems from the vast variations
observed across regimes, countries, and historical periods. Before delving deeper
into the subject, it is crucial to distinguish between two distinct yet
interconnected concepts: regime survival and individual leader survival.

Regime survival focuses on the longevity of political systems, such as
monarchies, political parties, or specific ideological structures. Leader
survival, on the other hand, is concerned with the duration of individual
leaders' time in office. These two concepts often exhibit contrasting patterns.
For example, in parliamentary democracies like Japan or the UK, specific
political parties may hold power for extended periods while individual leaders
(Prime Ministers) change frequently. Similarly, communist regimes typically see
long-lasting parties in power, with leadership transitions occurring at a slower
pace. In contrast, presidential systems like the United States or some military
regimes experience more frequent changes in both the ruling party and the head of
government. This study specifically investigates the dynamics of individual
leader survival, focusing on the factors influencing how long leaders remain in
power.

The existing literature on leader survival is vast and multifaceted, offering a
wealth of insightful perspectives. Some studies delve into the specific
mechanisms that influence leadership longevity within particular regimes, such as
democracies or autocracies. Others seek to develop more generalizable theoretical
frameworks that can explain leader survival across diverse political systems.
While the development of a universal theory remains an alluring goal, it is
important to acknowledge the inherent challenges in creating a single model that
encompasses the complexities of leadership survival across all regime types.

**Transparent and inclusive vs. Opaque and exclusive**

**First of all**, the mechanisms governing power transitions differ dramatically
in different countries, particularly between democracies and autocracies. The key
distinctions in the leadership selection process can be characterized as follows:
transparent vs. opaque and inclusive vs. exclusive.

In many autocratic systems, leadership selection is a closed affair. Potential
candidates are often restricted to a narrow pool, such as royal families,
military elites, or members of the ruling party. Selection processes are shrouded
in secrecy, and dissent can be met with harsh consequences. Even elections exist
in some autocracies, like Russia, there may have significant barriers to entry
for legitimate challengers. Potential rivals may face threats like assassination,
imprisonment, or exile.

This lack of transparency makes it difficult to gauge true levels of public
support as in democracies. Calculating selectorates or winning coalitions, as
explored by @buenodemesquita2003, becomes a near-impossible task. Leadership
transitions in autocracies often hinge on internal power struggles rather than
public opinion. A dictator may retain power despite low public support.

In contrast, democracies have relatively open competition for leadership
positions. Challengers can openly campaign and garner support through public
speeches, media campaigns, and grassroots movements. If no single candidate
commands a majority, factions can collaborate, leading to power transitions with
broader public backing. This allows for a more accurate assessment of power
dynamics through public opinion polls and electoral performance.

**Regular and predictable vs. Irregular and unpredictable**

**Secondly,** beyond the differences in openness and inclusivity, leadership
transitions in autocracies and some democracies are far more irregular and
unpredictable compared to stable democracies.

Many autocratic regimes lack institutionalized mechanisms for succession or term
limits. A leader's tenure often hinges on their ability to maintain control and
consolidate power. Certain dictators maintain an indefinite grip on power until
their death, subsequently transferring authority to family members, such as sons
in the cases of Syria and North Korea, or brothers in Cuba. This can also lead to
unexpected and irregular leadership changes. Leaders may be ousted through coups,
rebellions, uprisings, or popular protests. The opacity surrounding power
dynamics can also contribute to instability. Even seemingly strong leaders can be
overthrown by a small, well-positioned faction within the regime. Take, for
instance, Henri Namphy, who assumed the presidency of Haiti following a coup in
June 1988, only to be ousted by another coup a mere three months later in
September of the same year. However, Qaddafi, the dictator of Libya, seized power
in a coup in 1969 and ruled for over 40 years before being killed in 2011 by
NATO-backed rebel forces [@goemans2009].

In contrast, stable democracies experience regular and predictable leadership
transitions. Formal institutions, such as term limits, set clear boundaries for
how long leaders can remain in office. For example, US presidents are limited to
a maximum of two terms, regardless of their popularity. Similarly, in autocratic
Mexico from 1919 to 2000, each president served a fixed six-year term without
facing overthrows or overstays, as noted by [@klesner2019].

The study of political survival is particularly captivating due to the
intricacies and uncertainties associated with irregular leadership transitions.
While regular and anticipated leadership changes are important, they offer less
fertile ground for exploring the dynamics of leader longevity. The vast majority
of leaders who assume power through established channels also exit power through
established mechanisms [@goemans2009]. This research delves into the factors
influencing the survival of leaders who achieve power through irregular means, as
these transitions often hinge on complex political calculations, power struggles,
and unforeseen events.

Due to the above discussions, the diverse mechanisms influencing political
leadership survival have prompted scholars to delve deeper into regime-specific
dynamics. For instance, research by @svolik2014 examines democracies, while
@davenport2021 focuses on autocracies. A particular more focus lies on
understanding unexpected tenures, where leaders either fail to complete their
terms or overstay their mandates.

In analysing dterminants of leader survival, two primary perspectives have been
used to explain leader survival. The first considers contextual factors and
resources available to leaders. These include elements like personal competence
[@yu2016], societal stability [@arriola2009], economic development [@palmer1999;
@williams2011], access to natural resources [@smith2004; @quirozflores2012;
@wright2013], and external support networks[@licht2009; @wright2008; @thyne2017].
The second perspective explores the strategies leaders employ to implement their
agendas. This includes both the leaders' political goals and their responses to
opposition, challenges, or even coups and rebellions [@gandhi2007; @morrison2009;
@escribà-folch2013; @davenport2021].

Unsurprisingly, coups have received significant scholarly attention due to their
role in removing authoritarian leaders [@svolik2009; @frantz2016]. Existing
research explores both strategies for thwarting coups [@powell2017; @sudduth2017;
@debruin2020] and how leaders extend their tenures after surviving coup attempts
[@easton2018]. For example, @sudduth2017a examines the post-coup actions of
dictators, focusing on purge strategies. They argue that coup leaders initially
hold more power than elites within the regime, making them less susceptible to
subsequent coups. This challenges the conventional view of new leaders as
inherently weak [@roessler2011]. Despite this distinction, both Sudduth and
Roessler acknowledge that new leaders often consolidate power by purging rivals
(Sudduth suggests this occurs when there is minimal risk, while Roessler argues
it happens when threats are perceived). While the debate on the initial strength
of coup leaders continues, both perspectives agree that early purges are more
likely.

@sudduth2018 investigates how leaders' entry methods affect their removal in
dictatorships. They argue that irregular entry does not necessarily increase the
likelihood of removal. In fact, some forms of irregular entry may offer
protection. However, their focus was on how leaders lose office, not how long
they stay in power. Additionally, they compared six types of irregular leaders
without specifically examining the distinction between coup-entry and autocoup
leaders.

While scholars have extensively analysed leader survival across various contexts,
including universal frameworks, autocratic regimes, and the aftermath of failed
challenges, a significant gap remains. There is a lack of research specifically
exploring and comparing the survival tenures of leaders who extend their tenures
through autocoups compared to those who come to power through coups d'état. This
study aims to address this gap by investigating and comparing the duration of
leadership survival between these two distinct leader types.

## Survival dynamics of autocoup leaders and coup-entry leaders

### Autocoup leader and coup-entry leader

As highlighted in Chapter 2, investigating leadership survival presents inherent
challenges due to factors such as the opacity and varied mechanisms of power
transitions. However, these challenges underscore the importance of this
research, as it provides insights into understudied dynamics.

While the succession of political leaders exhibits complexity and variation, it
is not entirely devoid of patterns. Leaders of similar types often display
significant comparability. This study leverages this comparative approach to
analyse the ruling tenures of two distinct leader categories: coup-entry leaders
and autocoup leaders.

This study defines an autocoup as a situation where an incumbent leader utilizes
illegitimate or unconstitutional methods to extend their tenure and remain in
power. An autocoup is deemed successful if this power extension lasts for at
least six months. Coups d'état, on the other hand, are defined as illegal and
overt attempts by the military or other powerful groups within the state to
unseat the sitting leader [@powell2011]. Similar to coups, successful autocoups
in this study will be measured by a power usurpation lasting at least six months,
differing from the seven-day duration coded by Powell and Thyne.

It is also important to clarify the distinction between an autocoup leader and a
coup-entry leader, as the topic of this study is the survival of leaders.

-   **Autocoup Leader:** This refers to an incumbent leader who successfully uses
    illegitimate or unconstitutional means to extend their tenure in power. In an
    autocoup, the leader orchestrates the power grab and continues to rule
    afterwards.

-   **Coup-Entry Leader:** This term designates the leader who assumes power
    after a successful coup d'état. Unlike in autocoups, coups often involve
    multiple coup leaders (individuals or groups) who overthrow the incumbent
    leader. The coup-entry leader is the individual who ultimately takes the
    reins of power after the coup succeeds. While coup leaders may play a role in
    selecting the coup-entry leader, this is not always the case. In some
    instances, they may support someone outside the coup plot to become the new
    leader. Regardless of the selection process, the coup-entry leader is the one
    who assumes formal leadership following a successful coup.

As autocoup leaders normally exhibit longer overall tenures compared to
coup-entry leaders, this study concentrates on a more nuanced comparison.
Specifically, we will analyze the **post-autocoup** tenure of autocoup leaders,
contrasting it with the tenure of coup-entry leaders. This study examines the
survival tenures of coup-entry leaders and autocoup leaders for three key
reasons:

-   **Illegitimacy:** Both leader types lack legitimacy, though it manifests
    differently. Coup leaders seize power through force or the threat of force,
    making their illegitimacy explicit. Autocoups often employ seemingly legal
    procedures like court rulings or votes, but these processes are frequently
    manipulated by incumbents leveraging their control [@zhu2024]. This
    perception of illegitimacy can be used to justify the removal of such
    leaders, even if the means of removal are themselves illegitimate.

-   **Uncertainty:** The irregular means by which coup-entry and autocoup leaders
    take or retain power create significant uncertainty regarding their long-term
    rule and eventual exit. The established power transition process is
    undermined, raising questions about whether these leaders will enact
    constitutional succession protocols, and if so, whether those protocols will
    be followed. This ambiguity affects not only elites and citizens but also the
    leaders themselves, who face ongoing uncertainties about when, how, and to
    whom power might be transferred. Historical data on irregular leaders
    supports this point. Leader dataset shows that over half of irregular leaders
    experience irregular exits, and coup-related exits constitute more than
    two-thirds of all irregular exits [@goemans2009]. Furthermore, @frantz2016
    show that coup-related exits account for roughly one-third of all exits in
    autocracies, surpassing any other type, including regular transitions.

-   **Instability:** Coup-entry and autocoup leaders are aware of their own
    legitimacy and uncertainty challenges, which can fuel insecurity and a sense
    of crisis. In attempts to consolidate power, they may resort to reshuffling
    power structures or purging potential challengers. However, these efforts to
    maintain stability and strengthen their position often backfire, triggering
    greater instability.

The interplay of illegitimacy, uncertainty, and instability creates a complex
environment that significantly influences leader tenure.

### Hypothesis

While both coup-entry and autocoup leaders face similar challenges – lack of
legitimacy, irregularity, and fostering uncertainty – the severity of these
challenges often differ. Coup-entry leaders typically encounter these obstacles
to a greater degree. Consequently, they are expected to have shorter tenures in
power compared to autocoup leaders.

#### Legitimacy

Coup-entry leaders face an inherent legitimacy gap due to their sudden, often
violent, and unconstitutional rise to power. This approach is widely viewed as
illegal and sets a dangerous precedent, encouraging others to attempt similar
takeovers. Haiti's history exemplifies this, with Henri Namphy's brief presidency
in 1988 being cut short by another coup just three months later [@goemans2009].
Beyond immediate internal power grabs, coups can embolden external actors to
intervene, potentially leading to uprisings, revolutions, and civil wars
[@dahl2023].

In contrast, autocoup leaders manipulate power transition processes to extend
their rule, often through constitutional reinterpretations, engineered
parliamentary votes, or influenced Supreme Court decisions. These actions lack
genuine legitimacy as they exploit power and use illegal measures like silencing
opposition. Although their legitimacy is also contested, autocoup leaders
maintain a superficial veneer of legality, forcing challengers to operate within
legal frameworks. Replicating autocoup is more challenging than staging a coup
because it requires holding a leadership position first.

Furthermore, overthrowing incumbents who overstay in power through autocoups
presents significant hurdles due to the incumbent's power advantage. This
advantage is evident in cases like China (2018), where the National People's
Congress granted Xi Jinping the potential to rule for life[^1], and Russia
(2020), where constitutional changes allow Putin to potentially remain in power
until 2036[^2]. These actions demonstrate the consolidation of power by
incumbents, even if they do not reflect genuine public support.

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts (1950–2023)
#| echo: !expr format == "html"

coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold"
              ) |> 
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

### Uncertainty

Coup-entry and autocoup leaders face different levels of uncertainty immediately
following their rise to or overstay power.

After a coup, at least three major uncertainties arise regarding leadership and
its tenure. First, it is unclear who will assume leadership. Although coup
leaders often take power, some may return or promise to return power to civilian
leaders. Even among coup leaders, determining who will lead can be problematic,
as coup plotters are sometimes a group without a clear core leader. For instance,
following the 1973 Chilean coup, the initial plan for a rotating presidency among
military leaders was abandoned when General Pinochet consolidated control and
remained in power until 1990 [@svolik2014]. Second, the duration of the coup
leader's rule is uncertain. Leaders like Gamal Abdel Nasser in Egypt (1954 coup),
Muammar Gaddafi in Libya (1969 coup), and Idi Amin in Uganda (1971 coup) aimed to
retain power for life [@geddes2018], but their ability to do so was uncertain.
Others promise to transfer power to civilian authorities, but the timing and
fulfillment of these promises are unclear. For example, Myanmar's military junta
(2021 coup) has repeatedly extended a state of emergency, clinging to power
beyond the promised time-frame[^3]. Conversely, after the 2010 coup in Niger, the
military honoured their promise by restoring civilian rule within the same year
[@ginsburg2019]. Third, the successors of coup leaders are uncertain. Some may
designate successors from their inner circle, including family members, while
others may support general elections, though whether this will be fulfilled as
intended remains uncertain.

In contrast, autocoup leaders present a clearer picture regarding leadership and
tenure. There is no uncertainty about who will rule after an autocoup. In the
medium term, autocoup leaders typically hold office themselves. Many, like Putin
in Russia and Xi Jinping in China, seek to extend their rule indefinitely and are
unlikely to relinquish power voluntarily. Others attempt to extend their terms
incrementally, such as President Menem of Argentina, who overstayed until 1993
but failed in his bid for another term in 1999 [@llanos2019].

### Instability

The stability of a regime, particularly in an autocracy, hinges on maintaining a
balance of power. Coups, however, inevitably disrupt this balance, even when they
are bloodless, necessitating the creation of a new equilibrium. The ousting of
previous rulers requires dismantling the established governing structure and
reshuffling high-ranking officials, actions that inherently generate instability
and create adversaries for the new leadership. This makes restoring order and
establishing a balanced power structure notably challenging. Studies show that
new leaders often purge rival elite groups to consolidate their power at the
outset of their tenure [@sudduth2017; @roessler2011]. Such actions can provoke
backlash even from close allies. For instance, in Uganda, President Obote's
attempt to undermine the army commander-in-chief, Idi Amin, led to Amin gaining
the army's support and ultimately ousting Obote in a 1971 coup. Similarly, in
Pakistan in 1999, shortly after Prime Minister Sharif dismissed powerful army
chief General Pervez Musharraf, Sharif himself was ousted in a coup orchestrated
by Musharraf and his supporters [@sudduth2017].

To consolidate power, coup-entry leaders often have to compromise with internal
or external power challengers. However, these compromises are frequently unstable
and easily broken. The situation becomes even more complex when there is a risk
of civil war. Leaders may attempt to reduce the likelihood of subsequent coups,
potentially increasing the chances of societal rebellions and civil wars
[@roessler2011].

Moreover, instability extends beyond leadership to policies. A new leadership
group often brings new policies, and coups are sometimes triggered by
disagreements over significant policies. Major policy shifts can instigate
dissent or grievances from various ruling factions, communities, regions,
ethnicities, or religions.

In contrast, autocoup leaders encounter far fewer of these issues, as their
regimes experience fewer abrupt changes. They face less pressure to dismantle the
existing ruling paradigm and establish a new order. Even when adjustments are
necessary, they have more time to implement changes gradually.

These factors contribute to a shorter expected lifespan for coup-entry leaders
compared to the relatively longer tenures of autocoup leaders. The average
survival period following an autocoup is approximately five years longer than
that of coup-entry leaders (@fig-logrank). Building on these observations, I
propose the following hypothesis:

> ***H1: Political leaders who successfully prolong their tenure in power are
> more likely to experience extended regime survival compared to coup-entry
> leaders.***

In the subsequent section, I will outline the research methodology used in this
paper. I will introduce several control variables to determine if the hypothesis
remains persistent and robust.

## Research Design

### Survival analysis

```{r}
#| label: end
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: https://www.bbc.co.uk/news/world-asia-china-43361276: China’s Xi allowed to
    remain ‘president for life’ as term limits removed. Accessed on
    `r Sys.Date()`.

[^2]: https://www.ucl.ac.uk/news/2020/jul/analysis-vladimir-putin-secures-constitutional-changes-allowing-him-\
    rule-until-2036: Analysis: Vladimir Putin secures constitutional changes
    allowing him to rule until 2036. Accessed\
    on `r Sys.Date()`.

[^3]: https://thediplomat.com/2023/08/myanmar-junta-extends-state-of-emergency-for-fourth-time/:
    Myanmar\
    Junta Extends State of Emergency for Fourth Time. Accessed on `r Sys.Date()`.
