---
title: "Prospective Impact of Expected Coup Outcomes on Coup Attempts: A Selection Mechanism Analysis"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf:
    pdf-engine: pdflatex
  docx: default
  aps-revealjs:
    output-file: coups1.html
categories: [Article, Coups]
bibliography: references.bib
abstract: |
  A substantial body of research has examined coups, with much of it focusing on the factors that lead to coup attempts. However, consensus remains elusive regarding why coups are more prevalent in certain countries while less so in others. Previous scholarship exploring the determinants of coup attempts has often overlooked the crucial aspect of coup success. Given the severe consequences of a failed coup, coup plotters are unlikely to proceed unless they perceive a high chance of success. Thus, the expected outcome of a coup—whether successful or unsuccessful—is not merely incidental but serves as a pivotal determinant of coup attempts. The decision to stage a coup is a self-selected variable contingent upon the anticipated success rate of coups. This study employs a sample selection model (specifically, a two-stage probit model) to elucidate why coups are more common in some autocratic countries but rare in others. I contend that coup attempts are largely shaped by the likelihood of coup success, which, in turn, hinges on the power dynamics between coup perpetrators and incumbents. These power dynamics are influenced by the regime type and their distinct responses to internal and external shocks.
keywords: [Coup, Autocracy, Regime types, Sample selection]
---

```{r}
#| label: setup
#| message: false
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(democracyData)
```

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(coup$country)
succeeded <- length(coup$country[coup$coup == 2])

coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

## Introduction

Coups occur with varying frequency across different countries, with some experiencing them more frequently than others.
According to the Global Instances of Coups (GIC)[^1] dataset [@powell2011], Latin American countries such as Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups between $1950$ and $1984$, while Argentina experienced `r coups |> filter(country == "Argentina") |> pull(2)` during a similar time frame.
However, Mexico's authoritarian period from $1917$ to $2000$ saw no coups at all.
In Africa, Sudan endured `r coups |> filter(country == "Sudan") |> pull(2)` coups between $1955$ and $2023$, whereas South Africa has not experienced any coup since $1950$.
Similar patterns are observed in the Middle East and South Asia.
The question of why coups occur more frequently in certain regions, countries, and periods, while being less common in others, has captivated scholars for decades.
Consequently, scholars have delved into extensive research on coups.
Despite numerous efforts in past studies to shed light on these disparities, a definitive model or set of determinants for analysing coups remains elusive.
As highlighted by @gassebner2016, although approximately one hundred potential determinants of coups have been suggested, the fundamental question remains unanswered.

[^1]: <https://www.uky.edu/~clthyn2/coup_data/home.htm>, accessed on `r Sys.Date()`

However, despite extensive research by scholars, the impact of anticipated outcomes on coup initiation has not received sufficient attention.
When analysing the determinants of coups, it's crucial not to overlook the most significant characteristic of coups themselves.
As noted by @powell2011, coups are "illegal and overt attempts by the military or other elites within the state apparatus to unseat the sitting executive."[@powell2011, P.252].
Due to their illegality, the consequences of a failed coup could be severe, with perpetrators risking imprisonment, exile, or even death.
In some instances, the repercussions extend to the families of the coup perpetrators.

Despite the significant risks associated with coups, as shown in @tbl-coups since $1950$, there have still been as many as `r attempts` coups worldwide.
Furthermore, more importantly, half of these coups have been successful.
At first glance, coups seem to be a high-success-rate, high-reward political venture and speculation.
However, compared to over $12,000$ country-years since $1950$, the occurrence of `r attempts` coups appears rather rare, accounting for less than $4\%$.

The low occurrence rate and high success rate indicate that the initiation of coups is highly selective.
In other words, the likelihood of a coup occurring depends greatly on its potential success rate.
Coup plotters carefully assess their chances before staging a coup.
If they decide to proceed, it suggests that the conditions are relatively ripe, hence the fifty-fifty success rate.
Plans with immature conditions and low chances of success are automatically filtered out.
Fundamentally, launching a coup is not like participating in a general election, where failure in one election allows for another attempt in the next election.
A coup is a high-stakes gamble where success brings substantial rewards, while failure may result in martyrdom.
Failure, put differently, means permanent and absolute defeat.
Furthermore, the outcomes of coups are typically decided within a matter of days, sometimes even mere hours.
Unlike prolonged conflicts where victories or defeats may shift over time, the success of a coup is often sealed at its inception.
Every detail must be meticulously planned and arranged beforehand.

\clearpage

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    #table.width = pct(90),
    column_labels.font.weight = "bold",
    quarto.use_bootstrap = TRUE
              ) |> 
  cols_width(1 ~ px(120),
             2:4 ~ px(88)
  ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

Hence, the factors influencing the success rates of coups play a significant role in shaping coup attempts.
This study employs a sample selection model to examine the factors affecting the success rates of coups and, consequently, the likelihood of coup attempts.
I posit that the power dynamics among coup perpetrators, incumbents, and other ruling elites are pivotal in determining the success of coups.
These dynamics are largely contingent on regime types, highlighting the pivotal role of regime types in shaping coup attempts.

This study offers two potential contributions to the existing literature: firstly, it highlights the significance of power dynamics within various regime types as critical factors influencing coup attempts; secondly, it demonstrates how sample selection models can enhance our understanding of coup attempts by taking into account the success rates of coups.

The subsequent section of this paper delves into previous research on coups.
Following that, in Part $3$, I present the research framework and propose hypotheses.
Part 4 provides insights into the data and variables utilized in the study.
The testing results are discussed in Part $5$, followed by the conclusion in Part $6$.

## Dynamics of coup attempts and outcomes

When coup plotters weigh the decision to stage a coup, they naturally consider two crucial factors, which are also the primary concerns of coup scholars.
One aspect is the disposition, namely, why they would risk their future to instigate a coup.
The other aspect is the capability or opportunity, namely, whether they can garner enough support to succeed in their endeavour.

### Basic framework

A commonly cited framework for assessing the disposition and opportunity of coup attempts is formalized by @leon2013 to evaluate the anticipated benefits for coup plotters.
This framework can be outlined as follows:

$$
\begin{aligned}
\text{Expected Payoff of coups} = \sigma V_S + (1 - \sigma)  (V_I - C)
\end{aligned}
$$ {#eq-eq1}

Here, $V_S$ represents the return of a successful coup, $V_I$ denotes the normal value paid by the incumbent in the status quo, $C$ signifies the cost of the failed coup, and $\sigma$ represents the probability of coup success.
By rearranging the equation, we can derive:

$$
\begin{aligned}
\sigma(V_S - V_I) > (1-\sigma)C
\end{aligned}
$$ {#eq-eq2}

Drawing from these equations, @aidt2019 suggests that the decision to initiate a coup hinges on whether the expected benefits of a successful coup outweigh the potential costs of failure.
Both the disposition to attempt a coup and the opportunity to stage a successful coup are encompassed within the equation.
The disposition is captured in the gain from a successful coup, denoted as $V_S-V_I$.
Meanwhile, the opportunity is represented by the probability of a successful coup, denoted as $\sigma$.
Although this logical framework seems plausible, it immediately poses practical challenges in quantifying the values of $V_S$ and $C$, which are not easily measurable in real-world terms.
How can one accurately measure the cost of losing freedom, life, or even beloved family members if a coup fails?
Similarly, how can the value of assuming leadership of a country be quantified if the coup succeeds?
These are not commodities that can be exchanged in a market.

While $V_S$ and $C$ may be impossible to measure in precise terms, this does not render the equations useless.
Firstly, the underlying logic of the equations remains robust and provides a valuable framework for understanding the decision-making process of coup plotters.
Secondly, there is no need for precise measurements of $V_S$ and $C$.
It is sufficient to consider them as invaluable, with both potentially considered roughly equal.
Additionally, coup plotters are driven by factors beyond mere economic gains or losses.
As long as the coup succeeds, they win anyway.
This suggests that in @eq-eq2, the values of $V_S-V_I$ and $C$ can be disregarded, and the focus can solely be on the probability of the coup's success, $\sigma$:

$$
\begin{aligned}
\sigma > (1-\sigma)
\end{aligned}
$$ {#eq-eq3}

### Disposition

To simplify understanding, coups can be categorized into three types based on their motivations: those driven by personal ambitions, those purportedly for the benefit of the country, and those undertaken for self-preservation.

First and most, coup plotters are motivated by the allure of unrestrained power, revered status, and access to substantial financial resources.
@wintrobe2019 explores various forms of dictatorships, distinguishing between totalitarian leaders and tinpot leaders.
Totalitarian leaders exert control over every aspect of their citizens' lives, relishing the power they wield.
On the other hand, tinpot leaders are primarily motivated by personal consumption, indulging in luxuries such as living in opulent palaces, driving luxury cars like Mercedes, and adorning themselves with prestigious accessories like Rolex watches.

Secondly, at times, coups are justified by motives such as rescuing a country in crisis, upholding constitutional authority, or facilitating democratic transition.
However, this rationale is often met with scepticism, as the first type of coup plotters often use similar excuses to justify their actions.
Nonetheless, there are indeed cases that validate the purported objectives.
For instance, in Niger in $2010$, a coup ousted President Tandja, who sought to secure a third term in office contrary to constitutional limits.
Tandja's attempt to extend his presidency led to the dissolution of the Constitutional Court when it opposed his proposal, prompting a referendum for a new constitution to enable a third term.
This triggered the coup to protect the constitution [@ginsburg2019].

Thirdly, in certain instances, staging a coup serves as a final resort for self-preservation, aimed at evading political persecution by adversaries.
Coup perpetrators may not harbour further ambitions and may be content with the status quo.
However, their mere existence could be perceived as a threat by the incumbent leader, prompting the need for their removal.
Consequently, the coup is triggered rather than initiated by the coup perpetrators themselves.
An illustrative example is Idi Amin, who, as the army commander-in-chief of Uganda in $1971$, staged a coup against President Obote, who was attempting to undermine Amin [@sudduth2017].

Theoretically, these three types of coup plotters can emerge from any type of regimes.
While they may be more prevalent in autocracies, ambitious individuals can also exist within democracies.
Moreover, unstable societies, economic downturns, and democratic backsliding can occur even in democracies.
Although political persecutions may be rare in stable democracies, they are not uncommon in newly established or poorly governed ones.
However, despite these factors, the occurrence of coups remains relatively low, accounting for only $4\%$ of country-years since $1950$.
This is primarily because the disposition to stage coups is heavily contingent upon the opportunity to do so.

### Opportunity

As previously mentioned, while there may be numerous ambitious political figures aspiring to seize supreme power for various reasons, only a select few have the opportunity to orchestrate a successful coup.
When coup plotters conspire to stage a coup, their foremost consideration is the likelihood of success.
According to @eq-eq3, if the probability of success, $\sigma$, exceeds $0.5$, the equation holds true.
This assertion finds some support in empirical data on coups since $1950$.
As indicated in @tbl-coups, the overall success rate of coups stands at $49.9\%$.
Although this figure falls just below the threshold of $0.5$, it is important to acknowledge that this is an average rate calculated post-coups.
In reality, coup plotters typically require more certainty about success than a mere fifty-fifty chance.
Furthermore, outliers do exist, such as irrational political careerists or coup attempts driven more by self-protection instincts than strategic planning.
In such cases, coup perpetrators may not prioritize or even consider the potential outcomes, as the consequences of being purged are no different from those of a failed coup.

Given the pivotal role of coup success in determining coup attempts, the factors influencing coup success inevitably become the determinants of coup attempts themselves.
While the specific determinants may vary, the fundamental factor lies in the balance of power between the coup perpetrators and the incumbent leaders.
I will delve into the factors that could potentially tip the scales of power among different political factions.

### Regime types and balance of power

When discussing the balance of power, the first aspect that comes to mind is military strength.
It's evident that control over the military gives individuals the upper hand in coup attempts.
This explains why military coups are often at the forefront of political science discussions.
In much of the literature on coups, the terms "coup" and "military coup" are used interchangeably, with little exploration into why this is the case.
Consequently, more attention is directed towards preventing military involvement in politics.
Strategies such as "*Keeping the military happy*" [@aidt2019, P. 15] or "*giving them toys*" [@huntington1991democratization, P. 252] have been proposed to mitigate military intervention.
Many coup-proofing strategies, informed by empirical research, aim to either diminish the military's inclination to stage coups or create obstacles to their success.
For example, studies by @leon2013a suggest that nations with lower military spending as a percentage of GDP are more susceptible to coups.
Similarly, @powell2018 argues that increased military expenditures may reduce the likelihood of coups among military factions.

However, previous studies have often oversimplified the intricate balance of power within military forces.
They have frequently assumed, without substantial evidence, that the military operates as a monolithic entity, uniformly and decisively.
Such a notion fails to capture the complexities of real-world dynamics.

It is crucial to acknowledge that the military is not always a unified entity.
Regardless of its size, any military force is composed of various groups or factions, each with its own chain of command.
Within these factions, mutual suspicion, competition, and vigilance are common, while moments of unity are rare.
Due to the clandestine nature of coups and the need for secrecy, coup attempts are often orchestrated within small, tight-knit groups.
Coup plotters are uncertain about the stances and intentions of other factions beforehand, and they are particularly worried that once the coup is initiated, other factions may not only refuse to support it but also actively oppose, intervene in, or even suppress it.
Therefore, the success of a coup heavily depends on the reactions of other military factions [@geddes1999].

Furthermore, the structure of military forces, especially their relationships with other state departments, particularly civilian officials, varies across different regime types.
In democratic countries, the military is nationalized and bound by constitutional authority rather than individual leaders.
For instance, the United States Armed Forces are under the command of the President but remain loyal to the Constitution, refraining from intervening in politics.
In military regimes, power is often centralized in the hands of top generals or a junta, while in personal dictatorships, the leader retains ultimate authority over the military.
In dominant-party regimes, the military serves the interests of the ruling party.
Thus, the balance of power within military forces is deeply influenced by the nature of the regime in question.

To gain a thorough understanding of military power dynamics across diverse political landscapes, it is imperative to explore their intricate nuances.
This section adopts the foundational concepts articulated by @geddes2014 (GWF), which categorize regimes based on the origins of their leadership and the factors influencing their policy decisions.
These regimes are broadly classified into four categories: military regimes, personalist regimes (including monarchies), dominant-party regimes, and democratic regimes.

**Military regimes**, as defined by GWF, are characterized by the dominance of a junta composed of military officers who exercise control over the regime's power structure, including leadership selection and policy formulation.
Examples of such regimes include the Brazilian regime (1964–1985), the Argentine regime (1976–1983), and the Salvadoran regime (1948–1984) [@geddes1999].
It's noteworthy that within military regimes, political parties may persist, but they typically serve as conduits for the military junta or align with the directives of military officers to avoid being categorized as dominant-party regimes.
Additionally, while a junta may appoint a senior officer as the executive or enlist civilian administrators and technocrats as intermediaries, their authority over other officers remains restricted, distinguishing military regimes from personalist ones.

In a typical **personalist regime**, supreme power rests with the dictator, encompassing policy making, control over the military, and the authority to nominate top officials and successors.
Personalist rules often emerge under charismatic leaders, particularly founding fathers of newly independent nations.
Regimes such as Rafael Trujillo’s in the Dominican Republic (1930–1961), Idi Amin’s in Uganda (1971–1979), and Jean-Bédel Bokassa’s in the Central African Republic (1966–1979) serve as illustrative examples of personalist rule [@geddes1999].
In such regimes, either political parties are absent or they are established by or subservient to the dictator.
Importantly, the dictator may or may not be a military officer themselves, but the military is invariably under the dictator's control.

In a typical **dominant-party regime**, supreme power lies within the party, with the incumbent serving as its representative and subject to collective party leadership.
Examples of dominant-party regimes include the Partido Revolucionario Institucional (PRI) in Mexico, the Revolutionary Party of Tanzania (CCM), and Leninist parties in various East European countries [@geddes1999].
The party functions as a well-organized and highly disciplined entity, with its own ideology and political agenda to unite and mobilize its members and supporters.
While powerful leaders may emerge within such regimes, such as Stalin in the Soviet Union (1924-1953 and Mao Zedong in China (1949-1976), they do not wield enough power to supplant or replace the party with a new one.

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and their distribution since 1950 
#| echo: !expr format == "html"

REIGN |> 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant-party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) |> 
  filter(year > 1949) |>  
  summarise(counts = n(),
            .by = regime_type) |>
  arrange(desc(counts)) |>
  mutate(regime_type = str_to_title(regime_type),
         share = counts / sum(counts),
         share = percent(share)) |>
  adorn_totals(where = "row", fill = "100%") |>
  clean_names(case = "title") |> 
  gt() |>
  tab_options(
    column_labels.font.weight = "bold",
    table.width = pct(90)
              ) |>
  # cols_width(1 ~ px(200),
  #            2:3 ~ px(100)
  # ) |>
  tab_source_note(source_note = md("*Source: REIGN dataset*"))
```

## Research Framework and Hypotheses

## Data and Variables

## Results and Discussion

## Conclusion

```{r}
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
