---
title: "Prospective Impact of Expected Coup Outcomes on Coup Attempts: A Selection Mechanism Analysis"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: coups1.html
categories: [Article, Coups]
bibliography: references.bib
abstract: |
  A substantial body of research has examined coups, with much of it focusing on the factors that lead to coup attempts. However, consensus remains elusive regarding why coups are more prevalent in certain countries while less so in others. Previous scholarship exploring the determinants of coup attempts has often overlooked the crucial aspect of coup success. Given the severe consequences of a failed coup, coup plotters are unlikely to proceed unless they perceive a high chance of success. Thus, the expected outcome of a coup—whether successful or unsuccessful—is not merely incidental but serves as a pivotal determinant of coup attempts. The decision to stage a coup is a self-selected variable contingent upon the anticipated success rate of coups. This study employs a sample selection model (specifically, a two-stage probit model) to elucidate why coups are more common in some autocratic countries but rare in others. I contend that coup attempts are largely shaped by the likelihood of coup success, which, in turn, hinges on the power dynamics between coup perpetrators and incumbents. These power dynamics are influenced by the regime type and their distinct responses to internal and external shocks.
keywords: [Coup, Autocracy, Regime types, Sample selection]
---

```{r}
#| label: setup
#| message: false
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(democracyData)
```

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(coup$country)
succeeded <- length(coup$country[coup$coup == 2])

coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

## Introduction

Coups occur with varying frequency across different countries, with some
experiencing them more frequently than others. According to the Global Instances
of Coups (GIC)[^1] dataset [@powell2011], Latin American countries such as
Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups
between $1950$ and $1984$, while Argentina experienced
`r coups |> filter(country == "Argentina") |> pull(2)` during a similar time
frame. However, Mexico's authoritarian period from $1917$ to $2000$ saw no coups
at all. In Africa, Sudan endured
`r coups |> filter(country == "Sudan") |> pull(2)` coups between $1955$ and
$2023$, whereas South Africa has not experienced any coup since $1950$. Similar
patterns are observed in the Middle East and South Asia. The question of why
coups occur more frequently in certain regions, countries, and periods, while
being less common in others, has captivated scholars for decades. Consequently,
scholars have delved into extensive research on coups. Despite numerous efforts
in past studies to shed light on these disparities, a definitive model or set of
determinants for analysing coups remains elusive. As highlighted by
@gassebner2016, although approximately one hundred potential determinants of
coups have been suggested, the fundamental question remains unanswered.

However, despite extensive research by scholars, the impact of anticipated
outcomes on coup initiation has not received sufficient attention. When analysing
the determinants of coups, it's crucial not to overlook the most significant
characteristic of coups themselves. As noted by @powell2011, coups are "illegal
and overt attempts by the military or other elites within the state apparatus to
unseat the sitting executive."[@powell2011, P.252]. Due to their illegality, the
consequences of a failed coup could be severe, with perpetrators risking
imprisonment, exile, or even death. In some instances, the repercussions extend
to the families of the coup perpetrators.

Despite the significant risks associated with coups, as shown in @tbl-coups since
$1950$, there have still been as many as `r attempts` coups worldwide.
Furthermore, more importantly, half of these coups have been successful. At first
glance, coups seem to be a high-success-rate, high-reward political venture and
speculation. However, compared to over $12,000$ country-years since $1950$, the
occurrence of `r attempts` coups appears rather rare, accounting for less than
$4\%$ [@powell2011].

The low occurrence rate and high success rate indicate that the initiation of
coups is highly selective. In other words, the likelihood of a coup occurring
depends greatly on its potential success rate. Coup plotters carefully assess
their chances before staging a coup. If they decide to proceed, it suggests that
the conditions are relatively ripe, hence the fifty-fifty success rate. Plans
with immature conditions and low chances of success are automatically filtered
out. Fundamentally, launching a coup is not like participating in a general
election, where failure in one election allows for another attempt in the next
election. A coup is a high-stakes gamble where success brings substantial
rewards, while failure may result in martyrdom. Failure, put differently, means
permanent and absolute defeat. Furthermore, the outcomes of coups are typically
decided within a matter of days, sometimes even mere hours. Unlike prolonged
conflicts where victories or defeats may shift over time, the success of a coup
is often sealed at its inception. Every detail must be meticulously planned and
arranged beforehand.

Hence, the factors influencing the success rates of coups play a significant role
in shaping coup attempts. This study employs a sample selection model to examine
the factors affecting the success rates of coups and, consequently, the
likelihood of coup attempts. I posit that the power dynamics among coup
perpetrators, incumbents, and other ruling elites are pivotal in determining the
success of coups. These dynamics are largely contingent on regime types,
highlighting the pivotal role of regime types in shaping coup attempts.

This study offers two potential contributions to the existing literature:
firstly, it highlights the significance of power dynamics within various regime
types as critical factors influencing coup attempts; secondly, it demonstrates
how sample selection models can enhance our understanding of coup attempts by
taking into account the success rates of coups.

\clearpage

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold"
              ) |> 
  # cols_width(1 ~ px(120),
  #            2:4 ~ px(88)
  # ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

The subsequent section of this paper delves into previous research on coups.
Following that, in Part $3$, I present the research framework and propose
hypotheses. Part 4 provides insights into the data and variables utilized in the
study. The testing results are discussed in Part $5$, followed by the conclusion
in Part $6$.

## Dynamics of coup attempts and outcomes

When contemplating a coup, plotters grapple with two crucial factors that also
occupy coup scholars: ***disposition*** (why they take the risk) and
***capability*** (whether they can succeed).

### Disposition: Motivations for Coups

Coup motivations can be categorized into three main types:

**Personal Ambition:** The allure of absolute power, prestige, and wealth for
themselves, family, and close associates is a significant motivator.
@wintrobe2019 distinguishes between totalitarian and tinpot dictators based on
their use of power. Totalitarian leaders control every aspect of life, relishing
their authority. Tinpot leaders, however, prioritize personal enrichment,
indulging in extravagant lifestyles.

**Purported National Interest:** Sometimes, coups are justified as rescuing a
nation in crisis, upholding the constitution, or facilitating a democratic
transition. However, scepticism is warranted, as self-serving plotters often use
such rhetoric. Nevertheless, legitimate cases exist. In 2010, a coup in Niger
ousted President Tandja, who attempted an unconstitutional third term. His
actions, including dissolving the opposing Constitutional Court and calling a
self-serving referendum, triggered a coup to protect the constitution
[@ginsburg2019].

**Self-Preservation:** In rare instances, coups serve as a last resort against
imminent political persecution. Coup leaders might not seek power but fear
elimination by the incumbent leader. Consequently, the coup becomes a pre-emptive
strike rather than an ambitious power grab. An example is Idi Amin, who, in 1971,
staged a coup against Ugandan President Obote, who was attempting to remove Amin
from his position as army commander-in-chief [@sudduth2017].

While these motivations can arise in any regime, autocracies are more
susceptible, particularly for the latter two types. Stable democracies rarely
face such constitutional crises or political persecutions that necessitate coups
to protect them. However, new democracies can be vulnerable to instability,
economic downturns, and democratic backsliding which coup plotters can then use
as a pretext to justify their actions.

Despite these potential motivations, coups are relatively uncommon, occurring in
only $4\%$ of country-years since 1950. This is because even the most determined
coup plotters require the capability to succeed, which we will explore in the
next section. No rational actor attempts a guaranteed failure.

### Capability

While there may be numerous ambitious political figures vying to seize supreme
power for various reasons, only a select few have the opportunity to orchestrate
a coup. What truly determines their ability to launch a coup, however, is not
merely their willingness, but their capability. Yet, compared to the incumbent,
coup plotters naturally face disadvantages.

Firstly, as previously mentioned, coups constitute illegal actions against
incumbent leaders. To avoid the severe consequences of exposure or failure, coup
plots must remain clandestine within a tightly knit core group. Coup plotters
cannot openly recruit supporters to bolster their strength, while incumbents can
openly implement coup-proofing strategies to thwart coup attempts.

Secondly, coup plotters are uncertain about how other ruling factions might react
to their actions, particularly those capable of altering the balance of power.
Conversely, incumbent leaders possess a keen understanding of power dynamics and
take proactive measures to tilt the balance in their favour. Though unaware of
specific coup instigators, incumbent leaders are keenly aware of potential
threats and adapt accordingly to those who possess the capability to challenge
them.

Thirdly, coup plotters grapple with a loyalty dilemma and unreliable commitment
among participants. The risks involved are considerable, and anticipated benefits
are highly uncertain. Even in the event of a successful coup, there is no
guarantee that promises made by coup leaders will be honoured. Additionally,
post-coup purges often target capable assistants to pre-empt future challenges.
Consequently, defecting to support the incumbent leader seems less risky and
offers a safer bet with predictable rewards.

Given these challenges, rational coup plotters are unlikely to risk their lives
for a coup with low odds of success. Instead, they may opt to abandon their plans
or bide their time until success becomes more feasible. Therefore, when coup
perpetrators decide to take action, they must have meticulously assessed their
chances of success and deemed the risk worthwhile.

But what constitutes a 'good enough' chance of success for coup plotters to
proceed? Before delving into a theoretical framework to analyze this question,
let's first examine historical data to gain a rough idea. Surprisingly, previous
coups since 1950 suggest a rather satisfactory success rate, nearly $50\%$, as
depicted in @tbl-coups.

### Framework of coup success

An oft-cited framework for assessing the disposition and capability of coup
attempts is formalized by @leon2013, aimed at evaluating the anticipated benefits
for coup plotters. This framework can be represented as follows:

$$
\begin{aligned}
\text{Expected Payoff of coups} = \sigma V_S + (1 - \sigma)  (V_I - C)
\end{aligned}
$$ {#eq-eq1}

Here, $V_S$ represents the return of a successful coup, $V_I$ denotes the normal
value paid by the incumbent in the status quo, $C$ signifies the cost of the
failed coup, and $\sigma$ represents the probability of coup success. Rearranging
the equation yields:

$$
\begin{aligned}
\sigma(V_S - V_I) > (1-\sigma)C
\end{aligned}
$$ {#eq-eq2}

Drawing from these equations, @aidt2019 this equation captures the core
decision-making process: will the potential rewards outweigh the risks? Both the
motivation (disposition) and chance of success (capability) are embedded within
the equation. Motivation is reflected by the difference in benefits $V_S - V_I$,
while opportunity is represented by the probability of success, $\sigma$.

While seemingly clear, the equation faces practical challenges. Quantifying $V_S$
(value of a successful coup) and $C$ (cost of failure) is difficult. How do we
measure the loss of life, freedom, or loved ones in a failed coup? Similarly, how
do we quantify the value of leadership in a successful coup? These are intangible
concepts. However, these challenges don't render the framework useless. First,
the core logic remains valuable. It helps us understand how coup plotters might
think. Second, there is no need to concern the precise values for $V_S$ and $C$.
It's enough to consider them significant, possibly roughly equal. Coup plotters
often have goals beyond mere economic gain. Third, we can potentially shift focus
from $V_S$ and $C$ to the probability of success, $\sigma$:

$$
\begin{aligned}
\sigma > (1-\sigma)
\end{aligned}
$$ {#eq-eq3}

@eq-eq3 suggests that, to hold the equation true, a success probability greater
than $50\%$ is necessary. Interestingly, empirical data on coups since 1950
somewhat supports this notion. As shown in @tbl-coups, the overall success rate
is $49.9\%$. While this falls short of the $50\%$ threshold, it's important to
consider two factors. Firstly, this is an average rate, not necessarily
reflective of the probabilities assessed by coup plotters beforehand. Secondly,
outliers such as irrational actors and coups driven by self-preservation may not
prioritize success probabilities. Taking these points into account, we can
propose our first hypothesis:

> ***H1: The primary determinant of a coup attempt is the perceived chance of
> success. Coup plotters likely require a success threshold of at least***
> $50\%$***.***

This leads us to the next crucial question: what factors determine a coup's
success, influencing the very decision to attempt one? While specifics may vary,
the core element hinges on the power dynamic between coup plotters and the
incumbent regime. Logically, the more powerful entity holds a greater advantage
in this high-stakes struggle for control.

### Regime types and power dynamics

When discussing the balance of power, the first aspect that comes to mind is
military strength. It's evident that control over the military gives individuals
the upper hand in coup attempts. This explains why military coups are often at
the forefront of political science discussions. In much of the literature on
coups, the terms "coup" and "military coup" are used interchangeably, with little
exploration into why this is the case. Consequently, more attention is directed
towards preventing military involvement in politics. Strategies such as "*Keeping
the military happy*" [@aidt2019, P. 15] or "*giving them toys*"
[@huntington1991democratization, P. 252] have been proposed to mitigate military
intervention. Many coup-proofing strategies, informed by empirical research, aim
to either diminish the military's inclination to stage coups or create obstacles
to their success. For example, studies by @leon2013a suggest that nations with
lower military spending as a percentage of GDP are more susceptible to coups.
Similarly, @powell2018 argues that increased military expenditures may reduce the
likelihood of coups among military factions.

However, previous studies have often oversimplified the intricate balance of
power within military forces. They have frequently assumed, without substantial
evidence, that the military operates as a monolithic entity, uniformly and
decisively. Such a notion fails to capture the complexities of real-world
dynamics.

It is crucial to acknowledge that the military is not always a unified entity.
Regardless of its size, any military force is composed of various groups or
factions, each with its own chain of command. Within these factions, mutual
suspicion, competition, and vigilance are common, while moments of unity are
rare. Due to the clandestine nature of coups and the need for secrecy, coup
attempts are often orchestrated within small, tight-knit groups. Coup plotters
are uncertain about the stances and intentions of other factions beforehand, and
they are particularly worried that once the coup is initiated, other factions may
not only refuse to support it but also actively oppose, intervene in, or even
suppress it. Therefore, the success of a coup heavily depends on the reactions of
other military factions [@geddes1999].

Furthermore, the structure of military forces, especially their relationships
with other state departments, particularly civilian officials, varies across
different regime types. In democratic countries, the military is nationalized and
bound by constitutional authority rather than individual leaders. For instance,
the United States Armed Forces are under the command of the President but remain
loyal to the Constitution, refraining from intervening in politics. In military
regimes, power is often centralized in the hands of top generals or a junta,
while in personal dictatorships, the leader retains ultimate authority over the
military. In dominant-party regimes, the military serves the interests of the
ruling party. Thus, the balance of power within military forces is deeply
influenced by the nature of the regime in question.

To gain a thorough understanding of military power dynamics across diverse
political landscapes, it is imperative to explore their intricate nuances. This
section adopts the foundational concepts articulated by @geddes2014 (GWF), which
categorize regimes based on the origins of their leadership and the factors
influencing their policy decisions. These regimes are broadly classified into
four categories: military regimes, personalist regimes (including monarchies),
dominant-party regimes, and democratic regimes.

**Military regimes**, as defined by GWF, are characterized by the dominance of a
junta composed of military officers who exercise control over the regime's power
structure, including leadership selection and policy formulation. Examples of
such regimes include the Brazilian regime (1964–1985), the Argentine regime
(1976–1983), and the Salvadoran regime (1948–1984) [@geddes1999]. It's noteworthy
that within military regimes, political parties may persist, but they typically
serve as conduits for the military junta or align with the directives of military
officers to avoid being categorized as dominant-party regimes. Additionally,
while a junta may appoint a senior officer as the executive or enlist civilian
administrators and technocrats as intermediaries, their authority over other
officers remains restricted, distinguishing military regimes from personalist
ones.

In a typical **personalist regime**, supreme power rests with the dictator,
encompassing policy making, control over the military, and the authority to
nominate top officials and successors. Personalist rules often emerge under
charismatic leaders, particularly founding fathers of newly independent nations.
Regimes such as Rafael Trujillo’s in the Dominican Republic (1930–1961), Idi
Amin’s in Uganda (1971–1979), and Jean-Bédel Bokassa’s in the Central African
Republic (1966–1979) serve as illustrative examples of personalist rule
[@geddes1999]. In such regimes, either political parties are absent or they are
established by or subservient to the dictator. Importantly, the dictator may or
may not be a military officer themselves, but the military is invariably under
the dictator's control.

In a typical **dominant-party regime**, supreme power lies within the party, with
the incumbent serving as its representative and subject to collective party
leadership. Examples of dominant-party regimes include the Partido Revolucionario
Institucional (PRI) in Mexico, the Revolutionary Party of Tanzania (CCM), and
Leninist parties in various East European countries [@geddes1999]. The party
functions as a well-organized and highly disciplined entity, with its own
ideology and political agenda to unite and mobilize its members and supporters.
While powerful leaders may emerge within such regimes, such as Stalin in the
Soviet Union (1924-1953 and Mao Zedong in China (1949-1976), they do not wield
enough power to supplant or replace the party with a new one.

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and their distribution since 1950 
#| echo: !expr format == "html"

REIGN |> 
  mutate(regime_type = case_when(
    gwf_regimetype %in% c("party-based", "party-personal", "party-military", "party-personal-military", "oligarchy") ~ "dominant-party",
    gwf_regimetype == "personal" ~ "personal",
    gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
    gwf_regimetype == "monarchy" ~ "monarchy",
    gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
    TRUE ~ "other"
  )) |> 
  filter(year > 1949) |>  
  summarise(counts = n(),
            .by = regime_type) |>
  arrange(desc(counts)) |>
  mutate(regime_type = str_to_title(regime_type),
         share = counts / sum(counts),
         share = percent(share)) |>
  adorn_totals(where = "row", fill = "100%") |>
  clean_names(case = "title") |> 
  gt() |>
  tab_options(
    column_labels.font.weight = "bold",
    table.width = pct(90)
              ) |>
  # cols_width(1 ~ px(200),
  #            2:3 ~ px(100)
  # ) |>
  tab_source_note(source_note = md("*Source: REIGN dataset*"))
```

## Research Framework and Hypotheses

## Data and Variables

## Results and Discussion

## Conclusion

```{r}
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: <https://www.uky.edu/~clthyn2/coup_data/home.htm>, accessed on
    `r Sys.Date()`
