---
title: "Sample Selection Effects and the Likelihood of Coups"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: coups1.html
categories: [Article, Coups]
bibliography: references.bib
abstract: |
  A substantial body of research has examined coups, with much of it focusing on the factors that lead to coup attempts. However, consensus remains elusive regarding why coups are more prevalent in certain countries while less so in others. Previous scholarship exploring the determinants of coup attempts has often overlooked the crucial aspect of coup success. Given the severe consequences of a failed coup, coup plotters are unlikely to proceed unless they perceive a high chance of success. Thus, the expected outcome of a coup—whether successful or unsuccessful—is not merely incidental but serves as a pivotal determinant of coup attempts. The decision to stage a coup is a self-selected variable contingent upon the anticipated success rate of coups. This study employs a sample selection model (specifically, a two-stage probit model) to elucidate why coups are more common in some autocratic countries but rare in others. I contend that coup attempts are largely shaped by the likelihood of coup success, which, in turn, hinges on the power dynamics between coup perpetrators and incumbents. These power dynamics are influenced by the regime type and their distinct responses to internal and external shocks.
keywords: [Coup, Autocracy, Regime types, Sample selection]
---

```{r}
#| label: setup
#| message: false
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gtsummary)
library(gt)
library(janitor)
library(scales)
```

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| tbl-colwidths: auto
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(coup$country)
succeeded <- length(coup$country[coup$coup == 2])

coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(table.width = pct(90),
              column_labels.font.size = 16,
              column_labels.font.weight = "bold") |>
  opt_row_striping() |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(table.width = pct(90),
              column_labels.font.size = 16,
              column_labels.font.weight = "bold") |>
  opt_row_striping() |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

## Introduction

Coups occur with varying frequency across different countries, with some experiencing them more frequently than others.
According to the Global Instances of Coups (GIC)[^1] dataset [@powell2011], Latin American countries such as Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups between 1950 and 1984, while Argentina experienced `r coups |> filter(country == "Argentina") |> pull(2)` during a similar timeframe.
However, Mexico's authoritarian period from 1917 to 2000 saw no coups at all.
In Africa, Sudan endured `r coups |> filter(country == "Sudan") |> pull(2)` coups between 1955 and 2023, whereas South Africa has not experienced any coup since 1950.
Similar patterns are observed in the Middle East and South Asia.
The question of why coups occur more frequently in certain regions, countries, and periods, while being less common in others, has captivated scholars for decades.
Consequently, scholars have delved into extensive research on coups.
Despite numerous efforts in past studies to shed light on these disparities, a definitive model or set of determinants for analyzing coups remains elusive.
As highlighted by @gassebner2016, although approximately one hundred potential determinants of coups have been suggested, the fundamental question remains unanswered.

[^1]: <https://www.uky.edu/~clthyn2/coup_data/home.htm>, accessed on `r Sys.Date()`

However, despite extensive research by scholars, the impact of anticipated outcomes on coup initiation has not received sufficient attention.
When analyzing the determinants of coups, it's crucial not to overlook the most significant characteristic of coups themselves.
As noted by @powell2011, coups are "illegal and overt attempts by the military or other elites within the state apparatus to unseat the sitting executive."(P.252) Due to their illegality, the consequences of a failed coup could be severe, with perpetrators risking imprisonment, exile, or even death.
In some instances, the repercussions extend to the families of the coup perpetrators.

Despite the significant risks associated with coups, as shown in @tbl-coups since 1950, there have still been as many as `r attempts` coups worldwide.
Furthermore, more importantly, half of these coups have been successful.
At first glance, coups seem to be a high-success-rate, high-reward political venture and speculation.
However, compared to over $12,000$ country-years since 1950, the occurrence of `r attempts` coups appears rather rare, accounting for less than $4\%$.

The low occurrence rate and high success rate indicate that the initiation of coups is highly selective.
In other words, the likelihood of a coup occurring depends greatly on its potential success rate.
Coup plotters carefully assess their chances before staging a coup.
If they decide to proceed, it suggests that the conditions are relatively ripe, hence the fifty-fifty success rate.
Plans with immature conditions and low chances of success are automatically filtered out.
Fundamentally, launching a coup is not like participating in a general election, where failure in one election allows for another attempt in the next election.
A coup is a high-stakes gamble where success brings substantial rewards, while failure may result in martyrdom.
Failure, put differently, means permanent and absolute defeat.
Furthermore, the outcomes of coups are typically decided within a matter of days, sometimes even mere hours.
Unlike prolonged conflicts where victories or defeats may shift over time, the success of a coup is often sealed at its inception.
Every detail must be meticulously planned and arranged beforehand.

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |>
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, accuracy = 0.1)) |>
  clean_names(case = "title") |> 
  gt() |>
  tab_options(table.width = pct(90),
              column_labels.font.weight = "bold") |> 
  tab_source_note(source_note = md("***Source: GIC dataset***")) 
```

Hence, the factors influencing the success rates of coups play a significant role in shaping coup attempts.
This study employs a sample selection model to examine the factors affecting the success rates of coups and, consequently, the likelihood of coup attempts.
I posit that the power dynamics among coup perpetrators, incumbents, and other ruling elites are pivotal in determining the success of coups.
These dynamics are largely contingent on regime types, highlighting the pivotal role of regime types in shaping coup attempts.

This study offers two potential contributions to the existing literature: firstly, it highlights the significance of power dynamics within various regime types as critical factors influencing coup attempts; secondly, it demonstrates how sample selection models can enhance our understanding of coup attempts by taking into account the success rates of coups.

The subsequent section of this paper delves into previous research on coups.
Following that, in Part 3, I present the research framework and propose hypotheses.
Part 4 provides insights into the data and variables utilized in the study.
The testing results are discussed in Part 5, followed by the conclusion in Part 6.

## Literature Review

The majority of studies on coup risks primarily focus on identifying the causes and determinants of coup attempts.
As mentioned earlier, despite the multitude of potential determinants proposed—numbering around one hundred [@gassebner2016]—only a few studies have discussed both coup attempts and coup success.
Moreover, even fewer of these studies have specifically analysed the possible deterrent effect of coup outcomes on coup attempts.
Additionally, the dynamics of power balance among various types of regimes are often overlooked in previous studies.

The oft-cited framework regarding the anticipated gains of coup plotters is formalized by @leon2013 as follows:

$$
\begin{aligned}
\text{Expected Payoff} = \sigma V_S + (1 - \sigma)  (V_I - C)
\end{aligned}
$$ {#eq-eq1}

Here, $V_S$ represents the return of a successful coup, $V_I$ denotes the normal value paid by the incumbent in the status quo, $C$ signifies the cost of the failed coup, and $\sigma$ represents the probability of coup success.
By rearranging the equation, we can derive:

$$
\begin{aligned}
\sigma(V_S - V_I) > (1-\sigma)C
\end{aligned}
$$ {#eq-eq2}

@aidt2019 suggests that a coup is likely to be initiated when the anticipated benefits of a successful coup outweigh the potential costs of failure.
This implies that the decision to orchestrate a coup is heavily influenced by the perceived probability of success.
While the discussion of @eq-eq1 in @leon2013 may appear technical and theoretical, it raises practical challenges in quantifying the values of $V_S$ and $C$, which are not easily measurable in real-world terms and cannot be exchanged in a public market.
Coup plotters are motivated by more than just superficial gains or losses, as I will delve into further in the subsequent chapter.
Nonetheless, the underlying logic of the equation remains robust and offers a valuable framework for comprehending the decision-making dynamics of coup plotters.

When analysing coup outcomes, it is essential to consider three key dimensions.
Firstly, there is the impact coups can have on regime types and socio-economic development.
Many studies in this realm focus on the potential shifts towards democracy following coups.
This debate is exemplified by the arguments between Michael Miller [@miller2016] and Derpanopoulos, Frantz & Gedde [@derpanopoulos2016; @derpanopoulos2017].
The second dimension involves understanding the varying outcomes of coups: why some succeed while others fail [@quinlivan1999; @kebschull1994].
However, the third dimension, which explores how the expected success of coups influences coup attempts, remains underexplored.

Among the limited discussions on the influence of coup success on coup attempts, @powell2012 utilizes a selection model with global data spanning 1961 to 2000.
Powell's findings suggest that coup-proofing strategies, such as balancing and providing resources to the military, reduce the likelihood of coups but increase their chances of success.

However, few of the aforementioned studies have explored the interdependence between coup attempts and coup success.
In other words, as demonstrated in @eq-eq1 and @eq-eq2, they do not consider the likelihood of coup success as a primary determinant of coup attempts.

If the likelihood of coup success is predetermined, what factors influence or potentially determine the success rates of coups?
Much attention is understandably given to military strength, as it is believed that the control of military forces often determines the success of coups.
Consequently, coup-proofing strategies are designed to either decrease the inclination of military forces to stage coups or to erect barriers that impede their success.
For instance, studies by @leon2013a suggest that coups are more probable in nations where military spending as a percentage of GDP is low.
@powell2018 also argue that increased military expenditures may decrease the propensity for coups among military factions.
However, as @powell2012 posits, while measures aimed at pacifying the military may diminish the likelihood of a coup being attempted, greater military resources may simultaneously enhance the probability of a coup's success.

In addition to military factors, research delves into other determinants as well.
For instance, @bell2016 contends that democracies are less susceptible to coups, yet coup attempts against democracies are more likely to succeed.

## Research Framework and Hypotheses

## Data and Variables

## Results and Discussion

## Conclusion

```{r}
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
