---
title: "Leadership Transitions and Survival: Coups, Autocoups, and Power Dynamics"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-07-07
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: coup_autocoup.html

categories: [Article, Political Science]
bibliography: references.bib

abstract: |
  This dissertation examines the dynamics of irregular power transitions, particularly coups and autocoups, and their influence on leader survival. It highlights the critical role of power dynamics, shaped by regime types, in determining coup success rates and attempt frequency. Utilizing Heckman's two-stage selection model, the study reveals that expected coup success significantly influences attempts, with military regimes facing a heightened vulnerability due to their power structure.

  While often understudied, autocoups are shown to have a substantial impact on democratic trends. This research introduces a refined definition of autocoups alongside a novel dataset encompassing events from 1945 to 2022, enabling a more robust quantitative analysis.
  
  Employing survival analysis, the study compares the longevity of leaders who rise to power through coups versus autocoups. The findings demonstrate that coup-installed leaders face a significantly shorter tenure and higher risk of removal. This contrasts with autocoup leaders who manipulate the system to extend their rule, suggesting the potential for autocoups to incentivize power grabs and contribute to democratic backsliding.

  This work contributes significantly to the political science literature by:
   
   - Defining key concepts: It establishes a clear definition of autocoups, a previously understudied phenomenon.
   
   - Introducing a novel dataset: This dataset enables researchers to conduct more comprehensive quantitative analyses of autocoups.
   
   - Establishing a general framework: The framework provides a comparative approach to studying the dynamics of irregular power transitions and their impact on democratic stability.
  
keywords: [Coups, Autocoups, Power transitions, Leadership Survival]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(readxl)
library(here)
library(stargazer)
```

```{r}
#| include: false
Plad <- read_csv("https://raw.githubusercontent.com/reddylee/Datasets/main/PLAD.csv")
Archigos <- read_csv("https://raw.githubusercontent.com/reddylee/Datasets/main/archigos.csv")

Plad |>
  filter(startdate > 1945) |> 
  tabyl(exit)
Archigos |>
  filter(startdate > 1945,
         !exitcode  %in% c("Regular", "Still in Office")) |> 
  tabyl(exitcode) |> 
  adorn_totals()


```

## Introduction

### Research question

Irregular power transitions, marked by a disregard for constitutional procedures,
are a critical area of study in political science. They not only disrupt
established rules but often require unconstitutional tactics to secure power.
Furthermore, these transitions can inspire copycat behaviour among other
ambitious leaders.

Despite their central role in political science and the extensive research
conducted on irregular power transitions, a long-standing question continues to
intrigue political scientists: ***Why are some leaders ousted before their terms
expire, while others complete their full terms or even overstay beyond their
originally mandated limits?*** In other words, why do some leaders survive for
decades while others last for only years, months, or even days? This dissertation
focuses on this question and seeks to provide a comprehensive analysis, dedicated
to understanding how leaders come to power through unconstitutional means and
what factors determine the duration of a leader's rule following an irregular
ascent.

### Analyses on coups and autocoups in a general framework

When discussing irregular power transitions, the concepts that often come to mind
are irregular entries or exits, such as coups, assassinations, rebellions,
protests, and foreign interventions. Among these methods, coups hold a prominent
position due to their frequent occurrence. According to the Archigos dataset
[@goemans2009], from 1945 to 2015, there were approximately 145 instances of
irregular leader exits, with coups accounting for more than half (79 leaders).
The often-cited coup dataset [@powell2011] records even more leaders (245 cases)
removed by coups from 1950 to 2023.

Given their prevalence and substantial influence on political systems, coups have
been extensively studied, particularly since 2000 [@thyne2019]. Consequently, the
concept of a coup is comparatively clear and widely accepted in academic circles.
Many scholars, including this study, follow the definition by @powell2011, which
describes coups as "illegal and overt attempts by the military or other elites
within the state apparatus to unseat the sitting executive... \[a coup is
successful\] if the perpetrators seize and hold power for at least seven days"
(p. 252). Although debates persist, two elements are clear: first, the
perpetrators are elites within the ruling group, and the victims of coups are
incumbent executive leaders. Second, the strategy or aim of a coup involves
completely removing the incumbents, not merely seizing part of their power or
forcing them to concede on specific policies. Beyond defining coups, several
datasets have been developed for quantitative analyses, such as the Global
Instances of Coups [@powell2011], the Cline Centre Coup d’État Project Dataset
[@peyton2024], and the Colpus Dataset [@chin2021]. These datasets are
well-developed and frequently used in political science research.

However, irregular power transitions are not limited to irregular entries and
exits but should also include irregular "overstays." Using illegal means to
overthrow an incumbent leader before their term expires is undoubtedly an
irregular power transition. Similarly, an incumbent using illegitimate means to
extend their term beyond term limits is also an irregular power transition.

Although academic attention to irregular retention of power has increased since
the 1990s, especially after Peru’s President Alberto Fujimori's self-coup in
1992, it remains comparatively understudied and has several shortcomings. First,
there is no universally accepted terminology for this "overstaying in power" type
of irregular power transition, unlike the clear term "coup." Consequently,
various terms such as self-coup, autogolpe, and executive coup are used by
different scholars. This dissertation will use 'autocoup' to refer to this type
of irregular power transition, which will be thoroughly discussed in Chapter 3.
Second, there is no consensus on the definition of an autocoup. Existing
definitions remain vague, often conflating power expansions and power extensions.
For example, @cameron1998a defines an autogolpe as a temporary suspension of the
constitution and dissolution of Congress by the executive, who then rules by
decree. This definition focuses on power expansion instead of power extension,
leading to conceptual confusion and misalignment with the definition of a classic
coup. Third, a consensus autocoup dataset is lacking. While several related
datasets exist, as discussed by @baturo2022 in coding their Incumbent Takeover
dataset, the terminologies, definitions, and coverage years vary, lacking wide
acknowledgment and extensive academic exploration. In summary, autocoup has not
been analysed in a comparative manner connected with coups.

Analysing coups and autocoups separately is less problematic. However, from a
comprehensive framework perspective on irregular power transitions and leader
survival, coups and autocoups should be, and can be, analysed within the same
framework. Both coup and autocoup significantly influence democratic backsliding
and are the most frequent means of irregular power transition. Furthermore, as
both are called "coups," classic coups and autocoups are very similar since a
coup is launched to replace the current leader, while an autocoup is staged to
replace the future leader.

### Academic Contributions

This study addresses a critical gap in the literature by offering a comprehensive
framework for analysing both coups and autocoups, which are the most common forms
of irregular power transitions. While existing research often examines these
topics separately with varying terminologies, definitions, methods, and datasets,
this dissertation integrates these elements to provide a unified perspective on
irregular power transitions and leader survival.

Our contributions are threefold:

-   **Emphasis on power dynamics and regime types**: We highlight the significant
    role of power dynamics, particularly the influence of regime types, in
    determining the success and frequency of coup attempts. Our analysis
    underscores how the expected chances of coup success motivate such attempts,
    with military regimes being notably susceptible.

-   **Refined definition and novel dataset for autocoups:** We introduce a
    refined definition of autocoups and develop a novel dataset covering events
    from 1945 to 2022. This enables a comparative analysis with classic coups,
    providing clearer insights into the nature and impact of autocoups on
    political systems.

-   **Survival analysis of leaders from different entry modes:** By applying
    survival analysis to existing coup data and our new autocoup dataset, we
    demonstrate how different modes of entry into power significantly affect
    leader survival. Our findings reveal that leaders who come to power through
    coups typically have shorter tenures and face higher removal risks compared
    to those who extend their rule through autocoups.

Our analysis of irregular power transitions is particularly relevant to
understanding democratic backsliding. These transitions violate democratic norms
and disrupt the path towards stable democracy. Leaders who gain power through
irregular means often employ undemocratic tactics, such as suppressing
opposition, to consolidate their illegitimate hold on power. This creates a
vicious cycle where the erosion of democratic institutions is both a cause and
consequence of efforts to maintain power.

### Overview of the thesis

This study is structured into three main chapters beyond the introduction, each
addressing key aspects of irregular power transitions and their implications for
political stability and democratic processes.

**Chapter 2** examines the determinants of classic coup attempts. While extensive
research exists on coups, most studies focus on observable factors such as
economic performance, political stability, previous coups, and coup-proofing
strategies. This chapter, however, emphasizes the less observable but crucial
factor of expected coup success chances, which has been often overlooked.
Utilizing Heckman's two-staged sample selection model, the analysis reveals that
expected success rates significantly influence coup attempts. These success rates
are primarily shaped by the balance of power between incumbents and challengers,
which is largely determined by regime types. The findings indicate that military
regimes face a much higher risk of coups compared to dominant-party regimes.

**Chapter 3** focuses on the concept of autocoups, specifically on power
extensions by incumbent leaders. It distinguishes autocoups from broader concepts
like self-coups or executive coups by redefining them as instances where
incumbent leaders refuse to transition power as mandated, thereby overstaying in
office. Based on this refined definition, a novel dataset of autocoup events from
1945 to 2022 is introduced, encompassing 110 attempts and 87 successes. The
chapter includes case studies and empirical analyses that demonstrate the utility
of this dataset for quantitative research, providing a basis for empirical
analysis on autocoups.

**Chapter 4** investigates how the method of power acquisition impacts the
longevity of leaders who come to power through coups versus those who extend
their rule through autocoups. The hypothesis is that the method of accession
significantly affects leader tenure. Using the Cox proportional hazards model and
a time-dependent Cox model, the chapter provides evidence of differing survival
times between these two types of leaders. The results indicate that leaders who
come to power through coups face a significantly higher risk of removal compared
to those who extend their rule through autocoups. This finding highlights the
implications for political stability and democratic processes, suggesting that
the relatively low cost and high returns of autocoups could incentivize
incumbents to seize power in this manner, potentially leading to democratic
backsliding and the personalization of power.

In **Chapter 5**, the study concludes by summarizing the main findings,
discussing policy implications, and acknowledging the limitations of the
research. It also outlines directions for future research, emphasizing the need
for further exploration of irregular power transitions, particularly coups and
autocoups.

## Conclusion

### Main Findings

This study delves into the dynamics and implications of irregular power
transitions, focusing on coups and autocoups. The findings illuminate the complex
interplay between incumbents and challengers fighting for power.

Firstly, our analysis reveals that the expected success rate of a coup attempt
significantly influences its likelihood. This success rate is heavily influenced
by the balance of power between the incumbent regime and challengers, which is
largely determined by regime type. We find that military regimes, although with
more control over their own military forces, face a higher risk of coups compared
to dominant-party regimes.

Secondly, the study introduces a redefined concept: the autocoup. Defined as an
incumbent leader's refusal to relinquish power as mandated, this research
distinguishes autocoups from broader terms like self-coups. Based on this
definition, we present the first publicly available dataset of autocoup events
from 1945 to 2022, encompassing 110 attempts and 87 successful autocoups. Case
studies and empirical analyses demonstrate the dataset's utility for quantitative
research, providing a robust foundation for further analysis on autocoups.

Thirdly, employing survival analysis techniques, the study finds clear
differences in leader longevity between those who come to power through coups and
those who extend their rule through autocoups. The results indicate that
coup-installed leaders face a significantly higher risk of removal compared to
autocoup leaders who manipulate the system to extend their rule.

### Policy Implications

The findings of this study offer valuable insights for policy-makers concerned
with promoting and protecting global democracy, which has faced increasing
challenges despite a general post-WWII trend towards democratization. Notably,
the "third wave" of democratization [@huntington1991democratization] witnessed a
surge in democratic transitions in the late 20th century. Since the Cold War's
end, democratic nations have outnumbered non-democratic ones (@fig-democracy)
with the gap widening.

```{r}
#| label: fig-democracy
#| fig-cap: Comparison of the number of democratic and non-democratic countries (1945-2020)
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| echo: !expr format == "html"
REIGN |>
  filter(year > 1945 & year < 2021) |> 
  mutate(democracy = case_when(
    gwf_regimetype %in% c("parliamentary", "presidential") ~ "Democracy",
    .default = "Non-democracy")) |> 
  summarise(Count = n(), 
            .by = c(year, democracy)) |> 
  #view()
  ggplot(aes(year, Count, color = democracy)) +
  geom_point() +
  labs(color = 'Regime type',
       x = "Years: 1945 ~ 2020",
       y = "No. of Countries") +
  theme(legend.position = "top") +
  annotate("text",
    label = "Data source: REIGN (Regime data)",
    x = 2010,
    y = 30,
    fontface = "italic",
    size = 3
    )
```

However, a "democratic recession" has emerged in recent years [@diamond2008].
Freedom House reports an 18th consecutive year of global freedom decline in 2023
[@freedomhouse2024freedom]. While few countries have completely regressed to
autocracy, the average global democracy level has fallen back to pre-2000 levels.
Notably, democratic backsliding often occurs within regimes, with democracies
becoming less liberal and autocracies becoming less competitive [@mechkova2017].

This research highlights irregular power transitions as a significant factor in
democratic backsliding within regimes. These transitions, often coups or
autocoups, violate democratic norms and disrupt the path towards stable
democracies. Leaders who gain power through irregular means often resort to
undemocratic tactics to maintain control, creating a vicious cycle of eroding
democratic institutions.

Our findings suggest that the shorter lifespans and potentially severe
consequences associated with coups may deter potential coup leaders. Conversely,
autocoups appear to be a more tempting option for power-hungry leaders due to
their higher success rates, seemingly moderate consequences, and extended leader
tenure after the autocoup. This trend may explain the decline in classic coups
since the 1990s alongside the rise of autocoups [@bermeo2016].

### Limitations and directions for future research

This study offers a novel framework for analysing irregular power transitions,
but some limitations require further exploration:

-   **Data refinement:** Defining and classifying autocoups is a new approach.
    Future research should validate this classification system through additional
    studies and expert evaluations.

-   **Data harmonization:** The current analysis faces challenges due to
    mismatched units (country-year vs. leader) between coup and autocoup
    datasets. Future efforts should explore data harmonization techniques for
    more robust comparisons.

-   **Democratic backsliding:** While this study establishes a connection between
    irregular power transitions and democratic backsliding, further empirical
    evidence is needed to solidify this link.

Several avenues exist for future research:

-   **Terminology and data collection:** Refining the "autocoup" concept and
    achieving wider recognition will facilitate more accurate and comprehensive
    data collection.

-   **Dataset expansion:** Expanding the autocoup dataset with more cases and
    integrating it with data on other irregular leadership transitions can
    provide a more holistic view of political survival after these events.

-   **Power dynamics and long-term impacts:** Utilizing this dataset, future
    studies can delve deeper into power dynamics at play and explore the
    long-term consequences of irregular transitions on political systems,
    particularly regarding democratic backsliding, breakdown, and personalization
    of power.

In conclusion, this study sheds light on the dynamics of irregular power
transitions, specifically focusing on coups and autocoups. By redefining
autocoups, classifying the dataset, analysing determinants, and comparing leader
longevity, we establish a framework for understanding irregular transitions and
leader survival. This work contributes to a deeper understanding of democratic
resilience and political stability. Future research can build upon this
foundation by conducting further empirical analyses based on the novel autocoup
dataset and continuing to refine the framework.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false

cat("\\newpage
## References")
```
