---
title: "Leadership Transitions and Survival: Coups, Autocoups, and Power Dynamics"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-07-07
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: coup_autocoup.html

categories: [Article, Political Science]
bibliography: references.bib

abstract: |
  This thesis aims to provide a comprehensive framework for understanding the most common irregular power transitions—coups and autocoups—and the survival of political leadership in their aftermath. 
  Coups have been a central topic in political science literature, but the significant attention of this study is given to how power dynamics and the expected chances of success determine coup attempts. Employing Heck’s two-stage sample selection model, we find that the expected chances of coup success play a crucial role in motivating coup attempts. These chances are significantly influenced by the power dynamics between the incumbent and potential coup plotters, which are, in turn, affected by regime types.
  In contrast, autocoups have received less scholarly attention. This neglect stems partly from the underestimation of autocoups' impact on democratization and democratic resilience, and partly from the lack of a widely accepted definition and dataset. This study addresses these gaps by providing a refined definition of autocoups, reviewing existing terms and definitions such as self-coup, autogolpe, and executive takeover. Based on this new definition, we introduce a novel dataset of autocoups that can facilitate more quantitative studies. Additionally, this study includes several case studies and an empirical analysis to demonstrate the utility of the new dataset in future research.
  
  Beyond power transitions, political leadership survival is another critical topic in political science. This study offers a comparative analysis of leadership survival following coups and autocoups. Using the Cox proportional hazards model and a time-dependent Cox model, we present compelling evidence of differing survival times between leaders who enter through coups and those who extend power via autocoups. The findings indicate that coup-entry leaders have a much shorter average tenure and a higher probability of being ousted compared to autocoup leaders. These results suggest that the relatively low cost and high returns of autocoups could incentivize more incumbents to seize power in this manner, potentially leading to democratic backsliding.
  
  This research makes several contributions to the academic literature. First, it provides a general and comparative framework for the most common irregular power transitions: coups and autocoups. Second, it introduces a refined definition and a dataset of autocoups, enabling comparative studies on the dynamics of irregular power transitions alongside coups. Finally, it highlights the significant influence of power transitions and leadership survival on democratic backsliding, democratic breakdown, and personalization of power.
  
keywords: [Coups, Autocoups, Power transitions, Leadership Survival]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(readxl)
library(here)
library(stargazer)
```

```{r}
#| include: false
Plad <- read_csv("https://raw.githubusercontent.com/reddylee/Datasets/main/PLAD.csv")
Archigos <- read_csv("https://raw.githubusercontent.com/reddylee/Datasets/main/archigos.csv")

Plad |>
  filter(startdate > 1945) |> 
  tabyl(exit)
Archigos |>
  filter(startdate > 1945,
         !exitcode  %in% c("Regular", "Still in Office")) |> 
  tabyl(exitcode) |> 
  adorn_totals()

REIGN |>
  filter(year > 1945 & year <2021) |> 
  mutate(democracy = case_when(
    gwf_regimetype %in% c("parliamentary", "presidential") ~ "Democracy",
    .default = "Non-democracy")) |> 
  summarise(Count = n(), 
            .by = c(year, democracy)) |> 
  #view()
  ggplot(aes(year, Count, color = democracy)) +
  geom_point()
```

## Introduction

### Research question

Democratic process in the world has experienced different types of challenges
since World War II. As the beginning of the “third wave” of democratization
[@huntington1991democratization], we witnessed the grate transitions to democracy
in the last 20 years of 20th century. As the twentieth century ends, elected
officials govern more countries than at any previous time in human history
[@geddes1999]. However, the global democracy is under threat in post-cold war
era. @diamond2008 warned of a democratic recession at the end of the first decade
of 21st century. According to @freedomhouse2024freedom, global freedom declined
for the 18th consecutive year
in 2023. Although the data show relatively few countries backsliding from
democracy all the way to full-blown autocracy, the average level of democracy in
the world has slipped back to where it was before the year 2000, and most
democratic recessions have occurred within regime categories–with democracies
becoming less liberal and autocracies less competitive and more repressive
[@mechkova2017].

The factors cause backsliding vary and debating on which is the 'prime culprit'
is in the processing. This dissertation is not aiming to draw a conclusion on
this question. Instead, we aim to analyse one important factor which greatly
affects democratic backsliding–irregular power transition. Irregular power
transition not only itself break the constitutional rule, it often breaks the
positive process direct to democracy. Even worse, the irregular entry leaders, in
order to strengthen their illegitimate role in power, make more efforts to
repress oppositions or potential power challengers.

Although irregular power transition is so important and has always been a central
topic in political science, a long asked question still intrigued political
scientists: Why some leaders were ousted before their term expires, some can
accomplish their term, while some can even overstay in office beyond their
original mandated term limits. Or put it another way, why do some leaders rule
for decades while others stay in power for only years, months, or even days?

### Coups and autocoups

However, when referring to irregular power transition, first idea bump into our
mind is irregular entry or irregular exit, such as coups, assignations,
rebellions, protests, foreign interventions. Among these irregular means, coups
hold a outstanding position due to their majority occurrence. According to the
Archigos dataset [@goemans2009], from 1945 to 2015, there were approximately 145
instances of irregular leader exits, with coups[^1] accounting for more than half
(79 leaders). The often-cited coup dataset [@powell2011] records even more
leaders (245 cases) removed by coups from 1950 to 2023.

Given their prevalence and substantial influence on political systems, coups have
been extensively studied in previous research, especially since 2000
[@thyne2019]. Consequently, the concept of coup is comparatively clear and has
gained a consensus in academics. Scholars often follow the definition by
@powell2011, which describes coups as "illegal and overt attempts by the military
or other elites within the state apparatus to unseat the sitting executive... \[a
coup is successful\] if the perpetrators seize and hold power for at least seven
days" (p. 252). Although debates persist, two elements are clear: First, the
perpetrators are elites within the ruling group and the victim of coup is
incumbent executive leaders. Second, the strategy, or the aim of coup, involves
completely removing the incumbents, not merely seizing part of their power, or
forcing the incumbent to give in at some special policy or proposal.

Except the definition of coup, coup datasets have also been highly explored for
quantitative analyses, such as the Global Instances of Coups [@powell2011], the
Cline Centre Coup d’État Project Dataset [@peyton2024], and the Colpus Dataset
[@chin2021]. These datasets are well-developed and frequently used in political
science research.

However, when analysing in a comprehensive view on irregular power transition,
irregular entry and exit are not the whole story. It is just the half side of a
ball. Using illegal means to overthrow an incumbent leader before their term
expires is undoubtedly an irregular power transition. Similarly, an incumbent
using illegal or illegitimate means to forcibly remain in power after their term
has expired is also an irregular power transition. The former is irregular
because it happens unexpectedly, while the latter is irregular because it does
not happen as we expected.

Although academic attentions on irregular remaining in office are increasing
since 1990s, especially after the self-coup of Peru’s President Alberto Fujimori
in 1992, they are still comparatively less studied compared to irregular entry or
exit. Firstly, there is not even a terminology to name this "overstaying in
power" type of irregular power transition, which as clear as classic coup.
Consequently, various terms, such as self-coup, autogolpe, and executive
takeover, are used by different scholars. This dissertation will use 'autocoup'
to refer to this kind of irregular power transition, which will be discussed
throughly in Chapter 3. Secondly, no consensus has been made on the definition of
autocoup. Existent definitions remain vague between power expansions and power
extensions[^2]. For example, @cameron1998a defines an autogolpe as a temporary
suspension of the constitution and dissolution of Congress by the executive, who
then rules by decree. This definition focuses on the power expansion instead of
power extension, leading to conceptual confusion and the mis-alignment with the
definition of classic coup. Thirdly, a consensus autocoup dataset is lacking as
well and needs to identify and collect. Although several related datasets exits,
as discussed by @baturo2022 in coding their Incumbent Takeover dataset, the
terminologies, definitions and coverage years vary from one dataset to another,
not to mention the wide acknowledgement from scholars and extensive explore in
academics.

This issue is not as problematic when analysing coups and autocoups separately.
However, from a general framework point of view on irregular power transitions
and leader survival, coups and autocoups should be, and could be analysed in the
same framework. Both coup and autocoup are significant for their influence in
democratic backsliding. Both are the most frequent irregular power transition
means. Furthermore, to some extent, coup and autocoup are very similar. Coup is
to replace the current leader, while autocoup is to replace the future leader.

Extant literature on irregular power transition and leadership survival is mostly
discussed separately, by different scholars, using different concepts and
definitions, employing different methods and datasets. A general and comparative
analysis on coups and autocoups, and coup-entry leaders autocoup leaders, both in
theory and empirical analysis framework, therefore, are both important and
imperative.

This study addresses this gap by creating a comprehensive framework for analysing
coups and autocoups, particularly their determinants and the survival of leaders
who emerge from these events. We contribute to the political science literature
in three main aspects. First, we emphasize the role of power dynamics, primarily
determined by regime types, in determining coup success and, consequently, coup
attempts. Second, we introduce a refined definition of autocoup and a novel
dataset, allowing for comparative analysis with classic coups. Third, using the
existing coup dataset and the new autocoup dataset, we employ survival analysis
to demonstrate how different entry modes significantly affect leadership
survival, providing a starting point for analysing democratic resilience.

By doing so, we gain clearer insights into the factors that determine irregular
power transition attempts and their success or failure, and why leaders entering
power through different methods may have varying survival rates.

### Overview of the thesis

This study includes three main chapters beyond the introduction.

**Chapter 2 d**iscusses the determinants of classic coup attempts. Although coup
research is extensive, most studies focus on the influence of observable factors
like economic performance, political stability, previous coups, or coup-proofing
strategies. This study, however, focuses on the more essential but less
observable factor-expected coup success rates, which is often overlooked by
previous studies. Using Heckman's two-staged sample selection model, we find
evidence that success rates significantly influence coup attempts. The success
rates are primarily influenced by the balance of power between incumbents and
challengers, which is greately determined by regime types. The results indicate
that military regimes face a much higher risk of coups compared to dominant-party
regimes.

**Chapter 3** clarifies the concept of autocoups, focusing on power extensions by
incumbent leaders. By distinguishing autocoups from broader concepts like
self-coups or executive takeovers, we redefine autocoups as instances where
incumbent leaders refuse to transition power as mandated, thereby overstaying in
office. Based on this refined definition, we introduce a novel dataset of
autocoup events from 1945 to 2022, including 110 attempts and 87 successes. Using
this dataset, we present case studies and empirical analysis to demonstrate its
utility for quantitative research.

**Chapter 4** examines how the method of power acquisition impacts the longevity
of coup-entry and autocoup leaders. We hypothesize that the method of accession
significantly affects leader tenure. Using the Cox proportional hazards model and
a time-dependent Cox model, we provide evidence of differing survival times
between these two leader types. The findings indicate that coup-entry leaders
face a significantly higher risk of removal compared to autocoup leaders. This
underscores the implications for political stability and democratic processes,
suggesting that the relatively low cost and high returns of autocoups could
trigger incumbents to seize power in this manner, potentially leading to
democratic backsliding and personalization of power.

## Conclusion

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false

cat("\\newpage
## References")
```

[^1]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable `exitcode` are classified as coups.

[^2]: Power expansion and power extension are somewhat ambiguous terms. In this
    study, we define power expansion as acquiring additional powers beyond those
    originally associated with the leader's role or position. Conversely, we
    define power extension as lengthening a leader’s tenure beyond the originally
    mandated term.
