---
title: "Irregular Leadership Transitions and Political Survival: An Analysis of Coup-entry leaders and Autocoup leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: survival1.html
categories: [Article, Coups]
bibliography: references.bib

abstract: |
  This study explores and compares the survival time of two categories of irregular-entry political leaders: those who ascend to power through coups and those who overstay their legitimate term limits (autocoups). The central hypothesis posits that the mode of accession to power influences the duration of their tenures. Utilizing survival analysis, the results indicate that leaders who overstay their terms through autocoups generally remain in power longer than those who come to power through coups.

keywords: [Coups, Autocoups, Survival, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(survival)
library(survminer)
library(ggsurvfit)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(stargazer)
library(here)
```

## Introduction

The survival of political leaders has been a well-studied topic in political
science. While the general framework has garnered significant attention, specific
types of leaders, particularly coup-entry leaders and autocoup leaders (discussed
in more detail in Chapter 3), have received comparatively less scrutiny. However,
examining the tenures of these leader types is particularly significant.

Leaders who ascend through irregular means, such as coups or autocoups, attract
more interest from both academics and journalists. The processes of entry,
tenure, and exit are more predictable for leaders who come to power through
regular channels, making their survival easier to anticipate and thus less
compelling to study. In contrast, the uncertainty surrounding the tenures of
irregular leaders draws more attention. According to @goemans2009, leaders with
irregular entries or exits are quite common and therefore worth analysing.
Between 1945 and 2015, approximately 14.5% (213 out of 1,472) of leaders who
assumed office through regular channels exited irregularly. The proportion is
even higher among leaders with irregular entries. During the same period,
approximately 51.3% (158 out of 308) of those who assumed office through
irregular means experienced irregular exits.

Coup-entry and autocoup leaders constitute the majority of those who ascend to or
exit from power irregularly. Most leaders who gained power through irregular
means did so by launching coups or overstaying in office. According to
@goemans2009, out of 374 leaders who exited irregularly, 246 were ousted through
coups, accounting for 65.8% of these cases. @frantz2016 show that coup-related
exits account for roughly one-third of all exits in autocracies, surpassing any
other type, including regular transitions. Additionally, between 1945 and 2020,
there were 106 autocoup attempts, with 86 being successful [@zhu2024].

Due to their irregularity and inherent uncertainty, it is challenging to
precisely measure the tenure of coup-entry and autocoup leaders. Nonetheless,
comparing the tenures of these two types of leaders is both possible and
insightful. Conducting a log-rank test in survival analysis on the leaders
dataset [@goemans2009] and the autocoup dataset [@zhu2024] reveals a distinct
contrast in the tenures of autocoup leaders and coup-entry leaders.

```{r}
#| label: data
#| echo: !expr format == "html"
#| include: !expr format == "html"

Mydata1 <- read.csv(here("data/mydata1.csv")) |> 
  mutate(survival_year = survival_time / 365)

avg <- Mydata1 |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)

avg
```

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders
#| echo: !expr format == "html"

# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = Mydata1)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = Mydata1, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("entry by autocoup", "entry by coup"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 14)
```

Preliminary findings from survival analysis (see @fig-logrank) indicate a
significant difference in leadership tenures. Leaders who consolidate power
through autocoups have a longer average tenure post-autocoup (approximately
`r avg$avg[1]` years) compared to coup-entry leaders (approximately
`r avg$avg[2]` years). This suggests a potential shortfall of about 5 years in
the average tenure of coup-entry leaders.

We argue that the method of accession significantly influences leadership
longevity. Coup leaders likely face greater challenges to their legitimacy,
resulting in a more precarious initial period in power. They encounter higher
levels of uncertainty, instability, and pressure to share power, all of which
contribute to a shorter average tenure compared to autocoup leaders.

Employing the Cox proportional hazards model and the extended Cox model, our
results indicate that autocoup leaders generally experience longer post-autocoup
tenures compared to coup-entry leaders.

This study makes two significant contributions. Firstly, it sheds light on a
critical factor often overlooked in leadership survival studies: the impact of
the method of accession to power. Leaders' survival is not solely determined by
their actions after taking power but also by how they acquired power in the first
place. Secondly, utilizing survival models, this research provides empirical
evidence of the significant difference in tenure duration between autocoup and
coup-entry leaders. This finding may help explain the increasing prevalence of
overstaying in power through autocoups since 2000, as more incumbents may be
encouraged by precedents, and ruling elites tend to follow and support such
actions, benefiting from a relatively longer tenure.

The remainder of the paper delves deeper into this topic. Chapter 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Chapter 3 explores the factors influencing the survival of
coup and autocoup leaders. Chapter 4 outlines the methodology and data used,
including the application of survival models to analyze the determinants of
leadership longevity. Chapter 5 presents the findings of the analysis and a
detailed discussion of the results. Finally, Chapter 6 concludes by synthesizing
the key takeaways and exploring their broader implications.

## Literature review

The topic of political survival has been a cornerstone of political science
research for decades. This enduring interest arises from the wide-ranging
variations observed across regimes, countries, and historical periods. Previous
studies have identified two crucial yet interconnected aspects of political
survival: regime survival and individual leader survival.

Regime survival focuses on the longevity of political systems, such as
monarchies, political parties, or specific ideological structures. Leader
survival, on the other hand, is concerned with the duration of individual
leaders' time in office. These two concepts often exhibit contrasting patterns.
For example, in parliamentary democracies like Japan or the UK, specific
political parties may hold power for extended periods while individual leaders
(Prime Ministers) change frequently. Similarly, communist regimes typically see
long-lasting parties in power, with leadership transitions occurring at a slower
pace. In contrast, presidential systems like the United States or some military
regimes experience more frequent changes in both the ruling party or junta and
the leader of the country. This study specifically investigates the dynamics of
individual leader survival, focusing on the factors influencing how long leaders
remain in power.

The existing literature on leader survival is vast and multifaceted, offering a
wealth of insightful perspectives. Some studies delve into the specific
mechanisms that influence leadership longevity within particular regimes, such as
democracies or autocracies [@svolik2014; @davenport2021]. Others seek to develop
more generalizable theoretical frameworks that can explain leader survival across
diverse political systems [@buenodemesquita2003]. While the development of a
universal theory remains an alluring goal, it is important to acknowledge the
inherent challenges in creating a single model that encompasses the complexities
of leadership survival across all regime types.

The mechanisms governing power transitions vary significantly across different
regimes, particularly between democracies and autocracies. In many autocratic
systems, leadership selection is a closed affair. Potential candidates are often
restricted to a narrow pool, such as royal families, military elites, or members
of the ruling party. While political competition and elections may exist in some
autocracies, there are often significant barriers to entry for legitimate
challengers. Potential rivals may face threats such as assassination,
imprisonment, or exile. Moreover, selection processes are shrouded in secrecy,
with outsiders unable to participate or obtain information on the selections, let
alone express dissent or complain about the results. This lack of inclusivity and
transparency makes it challenging to gauge true levels of public support compared
to democracies. Consequently, calculating selectorates or winning coalitions, as
explored by @buenodemesquita2003, becomes a near-impossible task in autocracies.

Therefore, focusing research on more specific regimes or types of leaders may be
more appropriate. While regular and anticipated leadership changes are important
as well, they offer less fertile ground for exploring the dynamics of leader
longevity, as the vast majority of leaders who assume power through established
channels also exit power through established mechanisms [@goemans2009]. In
contrast, the study of political survival among irregular leaders is particularly
captivating due to the intricacies and uncertainties associated with irregular
leadership transitions.

The diverse mechanisms influencing political leadership survival, as discussed
above, have prompted scholars to delve deeper into more specific determinants.
Two primary perspectives have emerged to explain the dynamics of leader survival.

The first perspective considers objective factors and resources available to
leaders. These include elements such as personal competence [@yu2016], societal
stability [@arriola2009], economic development [@palmer1999; @williams2011],
access to natural resources [@smith2004; @quirozflores2012; @wright2013], or
external support networks [@licht2009; @wright2008; @thyne2017].

The second perspective explores subjective factors, focusing on the strategies
leaders employ to consolidate their power. This encompasses both the formulation
and implementation of political policies and the leaders' responses to
opposition, challenges, or even coups and rebellions [@gandhi2007; @morrison2009;
@escribà-folch2013; @davenport2021].

Unsurprisingly, coups have garnered significant scholarly attention due to their
pivotal role in removing leaders [@svolik2009; @frantz2016]. Existing research
delves into strategies for thwarting coups [@powell2017; @sudduth2017;
@debruin2020], as well as how leaders extend their tenures after surviving coup
attempts [@easton2018]. For instance, @sudduth2017 examines the post-coup actions
of dictators, focusing on purge strategies. They argue that coup leaders
initially wield more power than elites within the regime, rendering them less
susceptible to subsequent coups, challenging the conventional view of new leaders
as inherently weak [@roessler2011]. Meanwhile, @sudduth2018 investigates how
leaders' entry methods affect their removal in dictatorships, positing that
irregular entry does not necessarily increase the likelihood of removal; in fact,
some forms of irregular entry may offer protection.

While scholars have extensively analysed leader survival across various contexts,
including universal frameworks, autocratic regimes, and the aftermath of failed
challenges, a significant gap persists. There is a lack of research specifically
exploring and comparing the survival tenures of leaders who extend their reigns
through autocoups compared to coup-entry leaders. This study aims to address this
gap by investigating and comparing the duration of leadership survival between
these two distinct leader types.

## Survival dynamics of autocoup leaders and coup-entry leaders

### Autocoup leader and coup-entry leader

As emphasized in Chapter 2, delving into leadership survival poses inherent
challenges, given factors such as the opacity and diverse mechanisms of power
transitions. However, these challenges underscore the significance of this
research, as it sheds light on understudied dynamics.

While the survival of political leaders manifests complexity and variation, it is
not devoid of patterns entirely. Leaders of similar types often exhibit
significant comparability. Before delving into the comparison, it is necessary to
clarify several relative terminologies.

We define an autocoup as a situation where an incumbent leader utilizes
illegitimate or unconstitutional methods to extend their tenure and remain in
power. An autocoup is deemed successful if this power extension lasts for at
least six months. Coups, on the other hand, are defined as illegal and overt
attempts by the military or other powerful groups within the state to unseat the
sitting leader [@powell2011]. Similar to coups, successful autocoups in this
study will be measured by a power usurpation lasting at least six months,
differing from the seven-day duration coded by Powell and Thyne.

It is also crucial to clarify the distinction between an autocoup leader and a
coup-entry leader, as the survival of leadership is the main concern of this
study.

-   **Autocoup Leader:** This refers to an incumbent leader who successfully uses
    illegitimate or unconstitutional means to extend their tenure in power. In an
    autocoup, the leader orchestrates the power grab and continues to rule
    afterwards.

-   **Coup-Entry Leader:** This term designates the leader who assumes power
    after a successful coup d'état. Unlike in autocoups, coups often involve
    multiple coup leaders (individuals or groups) who overthrow the incumbent
    leader. The coup-entry leader is the individual who ultimately takes the
    reins of power after the coup succeeds. While coup leaders may play a role in
    selecting the coup-entry leader, this is not always the case. In some
    instances, they may support someone outside the coup plot to become the new
    leader. Regardless of the selection process, the coup-entry leader is the one
    who assumes formal leadership following a successful coup.

Given that autocoup leaders typically exhibit longer overall tenures compared to
coup-entry leaders, this study focuses on a more nuanced comparison.
Specifically, we will analyse the **post-autocoup** tenure of autocoup leaders
and contrast it with the **post-coup** tenure of coup-entry leaders. The
examination of the survival tenures of coup-entry leaders and autocoup leaders is
motivated by their relevance and similarity in terms of illegitimacy,
uncertainty, and instability.

-   **Illegitimacy:** Both leader types lack legitimacy, though it manifests
    differently. Coup leaders seize power through force or the threat of force,
    making their illegitimacy explicit. Autocoups often employ seemingly legal
    procedures like court rulings or votes, but these processes are frequently
    manipulated by incumbents leveraging their control [@zhu2024]. This
    perception of illegitimacy can be used to justify the removal of such
    leaders, even if the means of removal are themselves illegitimate.

-   **Uncertainty:** The irregular means by which coup-entry and autocoup leaders
    take or retain power create significant uncertainty regarding their long-term
    rule and eventual exit. The established power transition process is
    undermined, raising questions about whether these leaders will enact
    constitutional succession protocols, and if so, whether those protocols will
    be followed. This ambiguity affects not only elites and citizens but also the
    leaders themselves, who face ongoing uncertainties about when, how, and to
    whom power might be transferred. Historical data on irregular leaders
    supports this point. Leader dataset shows that over half of irregular leaders
    experience irregular exits, and coup-related exits constitute more than
    two-thirds of all irregular exits [@goemans2009]. Furthermore, @frantz2016
    show that coup-related exits account for roughly one-third of all exits in
    autocracies, surpassing any other type, including regular transitions.

-   **Instability:** Coup-entry and autocoup leaders are aware of their own
    legitimacy and uncertainty challenges, which can fuel insecurity and a sense
    of crisis. In attempts to consolidate power, they may resort to reshuffling
    power structures or purging potential challengers. However, these efforts to
    maintain stability and strengthen their position often backfire, triggering
    greater instability.

The interplay of illegitimacy, uncertainty, and instability creates a complex
environment that significantly influences leader tenure.

### Hypothesis

As previous studies have shown, while many factors can affect the survival of
leaders, the fundamental determinant of tenure longevity is the leader's ability
to maintain power. As long as they can hold onto power or balance it in their
favour, they can remain in office. However, as discussed earlier, although both
coup-entry and autocoup leaders face similar challenges related to illegitimacy,
uncertainty, and instability, the severity of these challenges differs. This
difference places the two types of leaders at varying levels of advantage or
disadvantage in terms of power balance. Coup-entry leaders typically encounter
these obstacles to a greater degree, leading to an expectation of shorter tenures
in power compared to autocoup leaders.

#### Illegitimacy

While both coup-entry and autocoup leaders lack genuine legitimacy, the nature of
this illegitimacy differs significantly. Coups represent a more blatant disregard
for legal processes, often involving force or the threat of force to seize power.
In contrast, autocoups employ a façade of legality through tactics such as
manipulating constitutional interpretations, engineering parliamentary votes,
influencing court decisions, and even holding referendums. Despite this veneer,
these maneuvers do not erase the underlying lack of genuine legitimacy, hence
this study categorizes them as autocoups. However, there is considerable debate
among scholars, with some arguing for less severe terms like "incumbent overstay"
or "executive takeover" due to the perceived legality.

This perceived legitimacy can provide a temporary advantage for autocoup leaders.
Challengers are often constrained to operate within legal frameworks, making it
difficult to directly confront the incumbent.

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts (1950–2023)
#| echo: !expr format == "html"

coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold"
              ) |> 
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

#### Uncertainty

Coup-entry and autocoup leaders face different levels of uncertainty immediately
following their rise to or overstay power.

After a coup, at least three major uncertainties arise regarding leadership and
its tenure. First, it is unclear who will assume leadership. Although coup
leaders often take power, some may return or promise to return power to civilian
leaders. Even among coup leaders, determining who will lead can be problematic,
as coup plotters are sometimes a group without a clear core leader. For instance,
following the 1973 Chilean coup, the initial plan for a rotating presidency among
military leaders was abandoned when General Pinochet consolidated control and
remained in power until 1990 [@svolik2014]. Second, the duration of the coup
leader's rule is uncertain. Leaders like Gamal Abdel Nasser in Egypt (1954 coup),
Muammar Gaddafi in Libya (1969 coup), and Idi Amin in Uganda (1971 coup) aimed to
retain power for life [@geddes2018], but their ability to do so was uncertain.
Others promise to transfer power to civilian authorities, but the timing and
fulfillment of these promises are unclear. For example, Myanmar's military junta
(2021 coup) has repeatedly extended a state of emergency, clinging to power
beyond the promised time-frame[^1]. Conversely, after the 2010 coup in Niger, the
military honoured their promise by restoring civilian rule within the same year
[@ginsburg2019]. Third, the successors of coup leaders are uncertain. Some may
designate successors from their inner circle, including family members, while
others may support general elections, though whether this will be fulfilled as
intended remains uncertain.

In contrast, autocoup leaders present a clearer picture regarding leadership and
tenure. There is no ambiguity about who will rule after an autocoup. In the
medium term, autocoup leaders typically hold office themselves. Many, like Putin
in Russia and Xi Jinping in China, seek to extend their rule indefinitely and are
unlikely to relinquish power voluntarily. Others attempt to extend their terms
incrementally, such as President Menem of Argentina, who overstayed until 1993
but failed in his bid for another term in 1999 [@llanos2019].

#### Instability

The stability of a regime, particularly in an autocracy, hinges on maintaining a
balance of power. Coups, however, inevitably disrupt this balance, even when they
are bloodless, necessitating the creation of a new equilibrium. The ousting of
previous rulers requires dismantling the established governing structure and
reshuffling high-ranking officials, actions that inherently generate instability
and create adversaries for the new leadership. This makes restoring order and
establishing a balanced power structure notably challenging. Studies show that
new leaders often purge rival elite groups to consolidate their power at the
outset of their tenure [@sudduth2017; @roessler2011]. Such actions can provoke
backlash even from close allies. For instance, in Uganda, President Obote's
attempt to undermine the army commander-in-chief, Idi Amin, led to Amin gaining
the army's support and ultimately ousting Obote in a 1971 coup. Similarly, in
Pakistan in 1999, shortly after Prime Minister Sharif dismissed powerful army
chief General Pervez Musharraf, Sharif himself was ousted in a coup orchestrated
by Musharraf and his supporters [@sudduth2017].

To consolidate power, coup-entry leaders often have to compromise with internal
or external power challengers. However, these compromises are frequently unstable
and easily broken. The situation becomes even more complex when there is a risk
of civil war. Leaders may attempt to reduce the likelihood of subsequent coups,
potentially increasing the chances of societal rebellions and civil wars
[@roessler2011].

Moreover, instability extends beyond leadership to policies. A new leadership
group often brings new policies, and coups are sometimes triggered by
disagreements over significant policies. Major policy shifts can instigate
dissent or grievances from various ruling factions, communities, regions,
ethnicities, or religions.

In contrast, autocoup leaders encounter far fewer of these issues, as their
regimes experience fewer abrupt changes. They face less pressure to dismantle the
existing ruling paradigm and establish a new order. Even when adjustments are
necessary, they have more time to implement changes gradually.

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-entry leaders 
#| echo: !expr format == "html"
#| results: asis
# library(kableExtra)
read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(100),
    2:3 ~ px(250)
  ) |>
   data_color(rows = c(2, 4), colors = "#edede9") 
```

In conclusion,

Furthermore, overthrowing incumbents who overstay in power through autocoups
presents significant hurdles due to the incumbent's power advantage. This
advantage is evident in cases like China (2018), where the National People's
Congress granted Xi Jinping the potential to rule for life[^2], and Russia
(2020), where constitutional changes allow Putin to potentially remain in power
until 2036[^3]. These actions demonstrate the consolidation of power by
incumbents, even if they do not reflect genuine public support.

These factors contribute to a shorter expected lifespan for coup-entry leaders
compared to the relatively longer tenures of autocoup leaders. The average
survival period following an autocoup is approximately five years longer than
that of coup-entry leaders (@fig-logrank). Building on these observations, I
propose the following hypothesis:

> ***H1: Political leaders who successfully prolong their tenure in power are
> more likely to experience extended regime survival compared to coup-entry
> leaders.***

In the subsequent section, I will outline the research methodology used in this
paper. I will introduce several control variables to determine if the hypothesis
remains persistent and robust.

## Research Design

### Survival analysis

```{r}
#| label: end
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: https://thediplomat.com/2023/08/myanmar-junta-extends-state-of-emergency-for-fourth-time/:
    Myanmar\
    Junta Extends State of Emergency for Fourth Time. Accessed on `r Sys.Date()`.

[^2]: https://www.bbc.co.uk/news/world-asia-china-43361276: China’s Xi allowed to
    remain ‘president for life’ as term limits removed. Accessed on
    `r Sys.Date()`.

[^3]: https://www.ucl.ac.uk/news/2020/jul/analysis-vladimir-putin-secures-constitutional-changes-allowing-him-\
    rule-until-2036: Analysis: Vladimir Putin secures constitutional changes
    allowing him to rule until 2036. Accessed\
    on `r Sys.Date()`.
