---
title: "Leadership Transitions and Survival:\n Coups, Autocoups, and Power Dynamics"

author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk
  
format:
  aps-html: default
  aps-pdf: 
    documentclass: report
    keep-tex: true
    colorlinks: true
    urlcolor: blue
    link-citations: true
    linkcolor: blue
    fig-asp: 0.618
    number-sections: true
    shift-heading-level-by: -1
    toc: true
    toc-title: Contents
    lof: true
    lot: true
    crossref: 
      lof-title: Figures
      lot-title: Tables
    linestretch: 1.618
    indent: true
    fontsize: 12pt
    mainfont: "Times New Roman"
    sansfont: "Arial"
    monofont: "Courier New"
    #latex-engine: pdflatex
    pdf-engine: xelatex
    geometry:
      # - showframe
      - top = 3cm
      - bottom = 3cm
      - left = 3cm
      - right = 2.7cm
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
      
  docx: default
    
categories: [Article, Coups, Autocoups, Leadership Survival]
bibliography: references.bib
#include-before-body: title.tex
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

pacman::p_load(tidyverse, data.table, gt, gtsummary, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects)
```

## Acknowledgements {.unnumbered}

The completion of this thesis marks the end of a significant journey, one filled
with hard work, continuous learning, and moments of profound joy. This
dissertation would not have been possible without the support and encouragement
of numerous individuals to whom I am deeply indebted.

First and foremost, I extend my heartfelt gratitude to my esteemed supervisor,
Professor Kristian Skrede Gleditsch. His invaluable guidance, unwavering support,
and insightful feedback have been instrumental in shaping this research. His
expertise and encouragement have truly been the cornerstone of this academic
endeavor.

I am equally grateful to Professor Han Dorussen, the chair of my board panel, for
his continuous support and constructive criticism. His thoughtful input has
significantly enhanced the quality and depth of my research.

I would like to acknowledge the vital contributions of my initial co-supervisors,
Dr. Saurabh Pant and Professor David Siroky. Their instruction and guidance
during the early stages of my research, prior to their departure from the
University of Essex, laid a strong foundation for this work.

This dissertation has benefited immensely from the expertise of several esteemed
scholars in the field. I am particularly indebted to Dr. Brian J Phillips, Dr.
Prabin Khadka, and Dr. Winnie Xia for their insightful comments, advice, and
suggestions. Their contributions have undoubtedly enriched this research.

On a personal note, I owe a debt of gratitude to my family for their unwavering
support and love. To my beloved wife, Ji Zhi, who has been my steadfast pillar of
strength throughout this journey - your patience and encouragement have been
immeasurable. To my dear children, Siyan and Sisheng, your joy and curiosity have
been a constant source of motivation and inspiration.

I am profoundly thankful to my father for his enduring support and belief in my
abilities. To the cherished memory of my late mother - your love, guidance, and
the values you instilled in me continue to shape my path and inspire my
endeavors. And to my three brothers, whose support enabled me to pursue my PhD
without worries - your contribution to this achievement cannot be overstated.

While many have contributed to the success of this work, any errors or
shortcomings remain entirely my own.

## Abstract {.unnumbered}

This dissertation examines the dynamics of irregular power transitions,
specifically coups and autocoups, and their influence on leader survival. Through
a rigorous analysis of historical data, case studies, and quantitative modeling,
this dissertation contributes to the broader understanding of political power
dynamics and the intricate factors that shape leadership survival in the wake of
irregular transitions.

The study first highlights the critical role of power dynamics, shaped by regime
type, in determining coup success rates and attempt frequency. Utilizing a double
probit model with sample selection, the research reveals that the expected
chances of coup success significantly influence coup attempts, with military
regimes facing heightened vulnerability due to their power structure.

While often understudied, autocoups are shown to have a substantial impact on
democratic backsliding. This research introduces a refined definition of
autocoups alongside a novel dataset encompassing events from 1945 to 2023,
enabling more robust quantitative analysis.

Employing survival analysis, the study compares the longevity of leaders who rise
to power through coups versus autocoups. The findings demonstrate that
coup-installed leaders face significantly shorter tenures and higher risks of
removal. This contrasts with autocoup leaders who manipulate the system to extend
their rule, suggesting the potential for autocoups to motivate power grabs and
contribute to democratic backsliding.

This work contributes significantly to the political science literature by:

-   Defining key concepts: Establishing a clear definition of autocoups, a
    previously understudied phenomenon.

-   Introducing a novel dataset: Enabling researchers to conduct more
    comprehensive quantitative analyses of autocoups.

-   Establishing a general framework: Providing a comparative approach to
    studying both coups and autocoups, enabling analysis in a unified framework
    for these two highly relevant political events and their potential effects on
    democratic backsliding.

***Keywords:*** Coups, Autocoups, Power transitions, Leadership survival,
Democratic backsliding

## Introduction

### Research Question and Context

Irregular power transitions, characterised by a disregard for constitutional
procedures, represent a critical area of study in political science. These
transitions not only disrupt established rules but also often employ
unconstitutional tactics to consolidate power post-transition. Moreover, such
events can inspire emulation among other ambitious leaders, potentially
triggering a cascade of similar actions across different political contexts.

Despite extensive research on irregular power transitions, a fundamental question
continues to challenge political scientists:

**Why do some leaders face premature ousting while others extend their tenure
beyond constitutionally mandated limits?**

This inquiry extends to understanding the stark disparities in leadership
longevity, where some rulers maintain power for decades while others' tenures are
measured in years, months, or even days.

This dissertation focuses on this central question, aiming to provide a
comprehensive analysis dedicated to understanding:

-   How leaders ascend to or overstay in power through unconstitutional means.

-   What factors determine the duration of a leader's rule following an irregular
    ascent to power.

### Coups and Autocoups: A Comparative Unified Framework for Irregular Power Transitions

In the study of irregular power transitions, coups have traditionally dominated
academic discourse due to their frequency and visibility. According to the
Archigos dataset [@goemans2009], coups[^1] accounted for more than half of the
approximately 145 irregular leader exits from 1945 to 2015. The Global Instances
of Coups (GIC) dataset [@powell2011] records an even higher number, with 245
coup-related removals from 1950 to 2024.

The prevalence of coups has led to a well-established definition and extensive
research, particularly since the turn of the millennium [@thyne2019]. This
scholarly focus has resulted in a general consensus on the definition of coups,
as well as the development of comprehensive datasets for quantitative analyses.
While debates on the precise definition persist, most scholars, including this
study, adhere to the definition proposed by @powell2011. This widely accepted
definition encompasses two fundamental elements:

-   **Perpetrators and Victims**: Perpetrators of coups are elites within the
    existing power structure and the victims are incumbent executive leaders.

-   **Strategy and Aim**: The primary objective is the complete removal of
    incumbents from power.

It is important to note that this definition excludes actions that only partially
challenge the incumbent's authority or seek policy changes without aiming for a
complete transfer of power.

The scholarly consensus on coup definition has facilitated the development of
several comprehensive datasets, enabling rigorous quantitative analyses in
political science research. Notable among these are: GIC Dataset, Cline Centre
Coup d'État Project Dataset [@peyton2024], and Colpus Dataset [@chin2021]. These
datasets have become invaluable resources in political science, enabling
researchers to conduct large-scale, comparative studies on the causes, dynamics,
and consequences of coups across different political contexts and historical
periods.

However, another form of irregular power transition has been largely overlooked
in power transition studies: leaders who refuse to relinquish power and extend
their mandated terms. At least three challenges still exist in the analyses of
this type of irregular power transition:

-   **Terminology**: Unlike the more widely accepted term "coup," various terms
    (e.g., self-coup, autogolpe, executive coup, incumbent takeover) are used to
    describe this phenomenon.

-   **Definition**: Existing definitions often conflate power expansion and power
    extension[^2]. For instance, @cameron1998a defines an autogolpe primarily in
    terms of power expansion, which does not align with the definition of coup.

-   **Data**: A consensus dataset for autocoups is lacking, with existing
    datasets varying in terminology, definitions, and coverage years @baturo2022.

To address this gap in the literature and provide a framework for analysis, we
propose the term "autocoup" to describe this type of transition. This concept
will be extensively discussed and defined in Chapter 3, providing a basis for
integrating these events into the broader study of irregular power transitions
and their impact on political systems and leader longevity.

Analysing coups and autocoups within a unified framework is crucial for a
comprehensive understanding of irregular power transitions and leader survival
for three reasons:

-   Both phenomena significantly influence democratic backsliding and represent
    the most frequent means of irregular power transition.
-   Their similar nomenclature reflects a fundamental similarity: while a coup
    aims to replace the current leader, an autocoup seeks to prevent the
    succession of a future leader.
-   This comparative approach will enable a more nuanced analysis of irregular
    power transitions, contributing to our understanding of political stability,
    democratic erosion, and leadership longevity in various regime types.

### Academic Contributions

This study addresses a critical gap in the literature by offering a unified
framework for analysing both coups and autocoups. My contributions are threefold:

-   **Emphasis on Power Dynamics and Regime Types**: This study highlights the
    significant role of power dynamics, particularly the influence of regime
    types, in determining the success and frequency of coup attempts, underscores
    how the expected chances of coup success motivate such attempts, with
    military regimes being notably susceptible.
-   **Refined Definition and Novel Dataset for Autocoups**: I introduce a refined
    definition of autocoups, develop a novel dataset covering events from 1945 to
    2024, and fill a significant gap in the existing literature and enable a
    comparative analysis with classic coups.
-   **Survival Analysis of Leaders from Different Entry Modes**: This research
    applys survival analysis to existing coup data and the new autocoup dataset,
    demonstrates how different modes of entry into power significantly affect
    leader survival, and reveals that leaders who come to power through coups
    typically have shorter tenures and face higher removal risks compared to
    those who extend their rule through autocoups.

These contributions collectively advance the field by providing a more holistic
understanding of irregular power transitions, offering new tools and data for
quantitative analysis of autocoups, and demonstrating the interconnectedness of
power acquisition methods and leadership survival.

This work lays the foundation for future research on the dynamics of political
power, regime stability, and democratic backsliding, offering both theoretical
insights and practical implications for policy-makers and scholars alike.

### Implications

The examination of irregular power transitions provides a crucial perspective on
the interrelated phenomena of democratic backsliding, breakdown, and autocratic
intensification. The findings of this study provide logical explanations for
several political phenomena:

-   **Regression of Global Democracy Levels**: This study can explain why global
    democracy levels have regressed to pre-2000 levels. For example, Freedom
    House reports an 18th consecutive year of global freedom decline in 2023
    [@freedomhouse2024freedom]. Irregular power transitions inevitably violate
    democratic or constitutional norms and disrupt the trajectory towards stable
    democracies or democratization from autocracies. Leaders who ascend through
    irregular means often undermine constitutional norms to seize or overstay in
    power, creating a vicious cycle of eroding democratic institutions.

-   **Within-Regime Democratic Erosion**: This study explains why democratic
    backsliding often occurs within regimes, with democracies becoming less
    liberal and autocracies becoming less competitive [@mechkova2017]. This is
    particularly because autocoups have become more prevalent than traditional
    coups since 2000 [@bermeo2016]. Importantly, autocoups typically do not
    result in immediate regime change, contributing to the observed pattern of
    within-regime democratic erosion.

-   **Prevalence of Autocoups Since 2000**: This research explains why autocoups
    are more prevalent since 2000. Firstly, autocoups have a significantly higher
    probability of success compared to traditional coups. Additionally, even when
    unsuccessful, the repercussions of autocoups are generally less severe than
    those of failed traditional coups. Moreover, successful autocoup leaders tend
    to maintain power for substantially longer periods than leaders who enter
    through traditional coups.

### Overview of the Thesis

This study investigates irregular power transitions and their implications for
leadership survival and democratic processes. I examine three key aspects:
classic coup attempts, autocoups, and how the method of power acquisition impacts
leader longevity.

#### Chapter 2: Determinants of Classic Coup Attempts

This chapter delves into the factors influencing classic coup attempts, with a
novel focus on the less observable but crucial factor of expected coup success
rates. Key features include the utilization of a double probit model with sample
selection and the analysis of how expected success rates significantly influence
coup attempts. It also examines how regime types shape the balance of power
between incumbents and challengers, determining the chances of success of coups.
Findings indicate a substantially higher coup risk in military regimes compared
to dominant-party regimes.

#### Chapter 3: Conceptualising and Analysing Autocoups

This chapter refines and analyses the concept of autocoups, with a specific focus
on power extensions by incumbent leaders. Notable elements include the
redefinition of autocoups as instances where incumbent leaders refuse mandated
power transitions. Introduction of a novel dataset covering autocoup events from
1945 to 2024, encompassing 110 attempts and 87 successes. Presentation of case
studies and empirical analyses demonstrating the utility of this dataset for more
quantitative research in political science.

#### Chapter 4: Impact of Power Acquisition Methods on Leadership Longevity

This chapter investigates how the method of power acquisition affects the tenure
of leaders who ascend through coups versus those who extend their tenure via
autocoups. Key aspects include hypothesis testing on the significant impact of
the accession method on leader tenure. Evidence of differing survival times
between coup-entry and autocoup leaders, employing the Cox proportional hazards
model and time-dependent Cox model. Analysis of how the risk-reward profile of
autocoups might motivate democratic backsliding and power personalization.

#### Chapter 5: Conclusion and Future Directions

The final chapter synthesizes the summary of key findings from each substantive
chapter. Acknowledges study limitations and their potential impact on findings.
Outlines future research directions, emphasizing the need for further exploration
of irregular power transitions, particularly coups and autocoups.

## Power Dynamics and Coup Attempts: A Selection Mechanism Analysis {#sec-chapter2}

### Abstract {.unnumbered}

Despite extensive research identifying around one hundred potential determinants
of coup attempts, no consensus has been reached. This study introduces a novel
approach that prioritizes determinants based on their impact on coup success. By
analysing coup success rates, the study hypothesizes that the expected outcomes
of coups are critical determinants of their occurrence. Utilizing a double probit
model with sample selection, the research investigates the relationship between
regime types and coup attempts. The findings confirm that regime type, by shaping
internal power dynamics, is a crucial determinant of coup likelihood. Military
and personalist regimes, characterized by weaker institutional frameworks and
higher vulnerability, demonstrate significantly higher susceptibility to coups
compared to dominant-party regimes.

\newpage

### Introduction

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data

Coup <- fread("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt")

attempts <- length(Coup$country)
succeeded <- length(Coup$country[Coup$coup == 2])

coups <- Coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )



Coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

Coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

Coups d'état, defined by @powell2011 as "illegal and overt attempts by the
military or other elites within the state apparatus to unseat the sitting
executive" (p. 252), represent a critical challenge to political stability and
democratic governance worldwide. This chapter examines the complex dynamics of
coup attempts, their success rates, and the factors that influence both their
occurrence and outcomes.

Coups occur with varying frequency across countries and regions. For instance: In
Latin America, Bolivia experienced
`r coups |> filter(country == "Bolivia") |> pull(2)` coups between 1950 and 1984,
while Argentina saw `r coups |> filter(country == "Argentina") |> pull(2)` during
a similar period. In Africa, Sudan endured
`r coups |> filter(country == "Sudan") |> pull(2)` coups between 1955 and 2023.
Contrastingly, countries like Mexico (1917-2000) and South Africa (since 1950)
have not experienced any coups.

This variability raises a fundamental question: **Why are coups more frequent in
some countries than others?**

Despite decades of research, political scientists have yet to reach a consensus
on the key determinants of coups. @gassebner2016 highlight this challenge, noting
that approximately 100 potential factors have been proposed, with their study
testing 66 of these across three million model permutations. This proliferation
of variables presents a critical issue: How can we establish a framework that
allows scholars to focus on the most relevant factors, rather than navigating an
ever-expanding list of potential determinants?

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold",
    table.layout = "fixed"
              ) |> 
  # cols_width(1 ~ px(120),
  #            2:4 ~ px(88)
  # ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

My analysis reveals a significant oversight in previous research: the exclusive
focus on pre-coup conditions without adequate consideration of post-coup factors.
This approach has neglected a critical element in coup dynamics – the expected
probability of success. Key observations supporting this perspective, as shown in
@tbl-coups, include:

-   **High-stakes nature of coups**: Failed coups often result in severe
    consequences for perpetrators, including imprisonment, exile, or death.

-   **Selectivity in coup attempts**: Despite 491 coup attempts since 1950, they
    represent only about 4% of over 12,000 country-years in the same period (GIC
    dataset).

-   **Satisfactory success rate**: Approximately half of all coup attempts
    succeed, suggesting careful selection of opportunities by plotters.

The low occurrence rate and high success rate clearly indicate that the
initiation of coups is highly selective. In other words, the likelihood of a coup
occurring depends greatly on its probability of success. However, this
probability is not directly observable to outsiders, including researchers, prior
to a coup attempt.

Given the limitations in directly observing coup success probabilities, I propose
focusing on regime type as a crucial proxy for coup outcomes. This approach is
based on the following reasoning: Coup outcomes are ultimately determined by the
balance of power within a regime and regime types are classified based on control
over the military, policy-making authority, and appointment of officials. By
analysing power dynamics across different regime types, we can gain valuable
insights into the structural factors shaping coup attempts and their outcomes.

To address the selection bias inherent in studying coup attempts, I employ a
double probit model with sample selection. This approach allows for simultaneous
analysis of factors influencing both coup initiation and success.

Through this comprehensive analysis, I aim to provide a more nuanced
understanding of the factors driving coup attempts and shaping their outcomes,
contributing to both scholarly discourse and practical efforts in two key ways:

-   **Emphasis on expected chances of success**: By focusing on the probability
    of success as a driver of coup attempts, I offer a more targeted approach to
    understanding coup dynamics.

-   **Highlighting the significance of regime type**: I demonstrate how regime
    type influences coup likelihood, even when researchers lack perfect knowledge
    of internal balance of power.

The remainder of this chapter is organized as follows: Section 2 explores the
dynamics of coup attempts and their outcomes. Section 3 outlines the research
design, methodology, and variables. Section 4 presents and discusses the
empirical findings. Section 5 concludes with key insights and their implications
for understanding and potentially mitigating coup risks.

### Dynamics of coup attempts and outcomes

Coup attempts are driven by a complex interplay of factors, encompassing both the
motivations of potential challengers (**disposition**) and the resources and
opportunities available to them (**capability**). This section delves into the
dynamics of coup attempts, exploring the motivations behind them, the factors
influencing their success, and the role of regime types in shaping coup
susceptibility.

#### Motivations for coups

This section focuses on the motivations that compel actors to undertake coups,
categorizing them into three main types:

-   **Personal Ambition:** The allure of absolute power, prestige, and wealth
    serves as a significant motivator for many coup plotters. Seizing control
    promises the ability to shape national policies, control resources, and make
    impactful decisions without constraints. The pursuit of prestige,
    recognition, potential economic gain, and the desire to leave a lasting
    legacy can further incentivize individuals to undertake this risky endeavour.

-   **Purported National Interest:** Coups are sometimes justified as necessary
    interventions to address national crises, uphold the constitution, or
    facilitate a transition to democracy. While such claims require scrutiny,
    genuine examples exist. For instance, the 2010 coup in Niger ousted President
    Tandja, who attempted an unconstitutional third term by dissolving the
    opposing court and calling a self-serving referendum [@ginsburg2019].

-   **Self-Preservation:** In certain cases, coups act as pre-emptive strikes
    against perceived threats. Coup leaders might not necessarily seek power for
    themselves but rather fear elimination or political persecution by the
    incumbent leaders. An example is Idi Amin's 1971 coup against Ugandan
    President Obote, who was attempting to remove Amin from his military command
    position [@sudduth2017].

These motivations are often most prevalent in autocratic regimes, where
justifications under the guise of national interest or self-preservation can mask
personal agendas. Stable democracies rarely face the same level of constitutional
crises or political persecution that might necessitate a coup. However, newly
established democracies can be vulnerable to instability, economic downturns, and
democratic backsliding, creating opportunities for coup plotters to exploit these
weaknesses and justify their actions.

Despite these potential motivations, coups remain relatively uncommon, occurring
in only about 4% of country-years since 1950. This rarity highlights the
importance of capability – even the most motivated actors need the resources and
opportunities to succeed.

#### Capability for coups

The decision to attempt a coup hinges not only on motivation but also on a
calculated assessment of the chances of success. Several factors influence this
assessment:

-   **Military Strength**: A clear advantage in military capabilities compared to
    the incumbent regime significantly increases the odds of a successful coup.

-   **Internal Divisions within the Regime**: Existing fractures within the
    government's power structure can be exploited by coup plotters to gain
    support from disgruntled factions.

-   **Public Support**: Widespread discontent with the incumbent regime,
    especially within the military or key sectors of society, can create a ripe
    environment for a successful coup.

-   **Foreign Backing**: External support from powerful nations can provide
    resources, legitimacy, and even direct military intervention to tip the
    scales in favor of the coup plotters.

While historical data might suggest a high success rate for coups, it's crucial
to consider selection bias. We only observe attempted coups, not the numerous
plots that never materialize. Analysing launched coup data alone can be
misleading. To understand coup attempts and their likelihood comprehensively, we
need a theoretical framework that accounts for this selection bias.

#### Framework of coup success

A frequently cited framework [@gassebner2016; @aidt2019] offers a structured
approach to assess the disposition and capability of coup attempts by evaluating
the anticipated benefits for coup plotters. The expected payoff of a coup can be
represented by the equation:

$$
\begin{aligned}
E(U) = p \times B + (1 - p) \times (-C)
\end{aligned}
$$ {#eq-eq1}

Where:

-   $E(U)$**:** Expected utility or pay-off of the coup attempt

<!-- -->

-   $B$ represents the return of a successful coup

-   $C$ signifies the cost of a failed coup

-   $p$ represents the probability of coup success

The condition for staging a coup is when the expected benefit is positive
($E(U) > 0$). Rearranging the equation, we get:

$$
\begin{aligned}
p \times B > (1 - p) \times C
\end{aligned}
$$ {#eq-eq2}

This implies that for a coup to be attempted, the expected benefits of success
must outweigh the expected costs of failure.

Quantifying $B$ and $C$ is inherently difficult. The loss of life, freedom, or
loved ones after a failed coup, as well as the value of assuming leadership after
a successful coup, are challenging to measure precisely.

However, the framework's core logic remains valuable. Given the difficulty in
precisely quantifying $B$ and $C$, we can treat them as roughly equal. This
allows us to shift our focus to the probability of success ($p$). The simplified
@eq-eq2 becomes:

$$
\begin{aligned}
p > (1-p)
\end{aligned}
$$ {#eq-eq3}

This suggests that a success probability greater than 50% is necessary for a coup
to be attempted. While empirical data shows a slightly lower overall success rate
for coups since 1950 (49.9%, as shown in @tbl-coups), it's crucial to remember
that this is an average and does not reflect the specific probabilities assessed
by coup plotters beforehand.

Therefore, I can propose the first hypothesis:

> ***H1: The fundamental determinant of a coup attempt is the perceived chance of
> success. Coup plotters likely require a success threshold of at least 50%.***

This leads to the next question: What factors determine coup success and
influence the decision to attempt one? The answer lies in understanding regime
types and their inherent power dynamics.

#### Regime types and power dynamics

Historical coup success rates do not dictate individual coup attempts. Coup
plotters assess their chances based on their unique context. While military
strength is undeniably crucial, often leading to an oversimplification of coups
as solely military events, it is vital to recognize the complex internal dynamics
within the military itself [@singh2016].

The clandestine nature of coups necessitates small, secretive groups, making it
difficult to gauge the stance of other factions within the military. The success
of a coup often hinges on the reactions of these other factions [@geddes1999].

Furthermore, factors beyond military force, such as internal divisions within the
ruling elites, public support, and foreign backing, significantly shape the
balance of power.

A useful framework for understanding coup susceptibility is to analyse regime
types, as their classification is based on power structures [@geddes2014]. We can
categorize autocracies into three main types:

-   **Military Regimes:** Characterized by a junta – a group of military officers
    controlling leadership selection and policy formulation. Examples include
    regimes in Brazil (1964-1985), Argentina (1976-1983), and El Salvador
    (1948-1984) [@geddes1999].

-   **Personalist Regimes:** Power is concentrated in a single, charismatic
    leader who controls the military, policy, and succession. Examples include
    Rafael Trujillo's regime in the Dominican Republic (1930-1961), Idi Amin's
    regime in Uganda (1971-1979), and Jean-Bédel Bokassa's regime in the Central
    African Republic (1966-1979) [@geddes1999].

-   **Dominant-Party Regimes:** Power resides within a well-organized ruling
    party, with leaders acting as its representatives. The party structure and
    ideology foster internal cohesion and a long-term vision. Examples include
    the Partido Revolucionario Institucional (PRI) in Mexico, the Revolutionary
    Party of Tanzania (CCM), and Leninist parties in various Eastern European
    countries [@geddes1999].

These regime types exhibit distinct power dynamics that influence their
susceptibility to coups (@tbl-regimes1):

-   **Military Regimes:** Despite concentrated military control, they are
    surprisingly unstable due to internal power struggles within the junta. The
    lack of a clear final authority and the presence of multiple military
    factions increase the likelihood of resorting to force to resolve disputes,
    making these regimes the most vulnerable to coups.

-   **Personalist Regimes:** Relatively stable during the leader's tenure, but
    face a higher risk of coups due to unclear succession plans and
    vulnerabilities associated with the leader's personal weaknesses, health, and
    mortality.

-   **Dominant-Party Regimes:** Exhibit the greatest resilience against coups due
    to their institutionalized structures, unified leadership, clear ideology,
    and internal discipline.

Empirical data supports this framework. While military regimes represent only
5.6% of country-years since 1950, they experience a disproportionate share of
coups (over 22%). Personalist regimes, constituting 13% of country-years, account
for 23% of coups. Conversely, dominant-party regimes, representing 22.6% of
country-years, account for only 16.7% of coups (@tbl-regimes).

This leads to our second hypothesis:

> ***H2: Due to their balance of power dynamics, military regimes are the most
> prone to coups, followed by personalist regimes, while dominant-party regimes
> are the least likely to experience coups among the three.***

\newpage

\blandscape

```{r}
#| label: tbl-regimes1
#| tbl-cap: Main features of different types of regimes 
#| echo: !expr format == "html"

read.csv(here("data/regimes_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center",
              page.orientation = "landscape"
              ) |>
  cols_width(
    1:6 ~ px(100)
  ) |>
  tab_source_note(source_note = md("*Source: GWF & Author*"))
```

\newpage

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and coups since 1950 
#| echo: !expr format == "html"

regimes <- REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949)

coup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/coup_model.csv")

regime_total <- coup_model |>
  select(regime_type = regime) |> 
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(country_year =n,
         share = percent(percent),
         .keep = "unused") 
  

coup1 <- Coup |>
  select(country, ccode, year, coup)

coup_by_regime <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup)) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  mutate(num_of_coups = n,
         percent_of_coups = percent(percent,accuracy = 0.1),
         .keep = "unused") 

coup_by_succeed <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup),
         coup == 2) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  select(regime, succeeded_coup = n) 

coup_regimes <-  coup_by_regime |> 
  left_join(coup_by_succeed) |> 
  mutate(success_rate = percent(succeeded_coup / num_of_coups, accuracy = 0.1))

regime_total |>
  left_join(coup_regimes, join_by(regime_type == regime)) |>
  mutate(regime_type = str_to_title(regime_type),
         coup_likelihood = percent(num_of_coups / country_year, accuracy = 0.1)) |>
  select(-succeeded_coup) |> 
  clean_names(case = "title") |>
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN and GIC Datasets*"))
```

\elandscape

### Research Design

#### Double probit with sample selection model {#sec-coup231}

This study employs a sophisticated statistical approach to account for the
selective nature of coup attempts. While coup attempt rates vary across regimes,
success rates tend to be surprisingly consistent, hovering around 50% (as shown
in @tbl-regimes). This suggests that coup attempts are not random acts, but
rather strategically planned and undertaken only when the odds of success appear
favorable. A standard statistical model would not account for this selectivity,
potentially leading to biased results.

To address this issue, we utilize a double probit with sample selection model,
similar to the approach used by @powell2012. This model, known as a Heckman
probit model or bivariate probit model with sample selection, consists of two
parts:

-   **Selection Equation (Stage 1)**: This stage analyses the factors influencing
    whether a coup attempt occurs in a particular country-year.

-   **Outcome Equation (Stage 2)**: This stage focuses on the probability of
    success for those coup attempts that actually take place.

The selection equation (first stage) models the probability that a coup attempt
occurs:

$$
\begin{aligned}
y_{1i}^*&=\alpha_0 + \alpha_1 Regime_i + \mathbf{X}_i \boldsymbol{A} + \mu_{1i}
\\
\\
y_{1i} &= 
\begin{cases} 
1 &\text{if $y_{1i}^*>0$ (coup attempt occurs)} \\
\\
0 &\text{if $y_{1i}^*\le0$ (no coup attempt)}
\end{cases}
\end{aligned}
$$

The outcome equation (second stage) models the probability that a coup attempt is
successful, given that it occurs:

$$
\begin{aligned}
y_{2i}^*&=\beta_0 + \beta_1 Regime_i + \mathbf{Z}_i \boldsymbol{B} + \mu_{2i}
\\
\\
y_{2i} &= 
\begin{cases} 
1 &\text{if $y_{2i}^*>0$ (coup succeeds)} \\
\\
0 &\text{if $y_{2i}^*\le0$ (coup failes)}
\end{cases}
\end{aligned}
$$

Where:

-   $y_{1i}^*$ *and* $y_{2i}^*$ are latent variables

-   $\text{Regime}_i$ is a categorical variable (military, personalist, or
    dominant-party)

-   $\mathbf{X}_i$ and $\mathbf{Z}_i$ are vectors of control variables

-   $\mu_{1i}$ and $\mu_{2i}$ are error terms, assumed to follow a bivariate
    normal distribution with correlation $\rho$

The model assumes:

$$
\binom{\mu_{1 i}}{\mu_{2 i}} \sim N\left(\binom{0}{0},\left(\begin{array}{ll}
1 & \rho \\
\rho & 1
\end{array}\right)\right)
$$

The probability equations are:

$$
\begin{aligned}
P\left(y_{1 i}=1\right) & =\Phi\left(\alpha_0+\alpha_1 \operatorname{Regime}_i+\mathbf{X}_i \boldsymbol{A}\right) \\
\\
P\left(y_{2 i}=1 \mid y_{1 i}=1\right) & =\Phi\left(\beta_0+\beta_1 \operatorname{Regime}_i+\mathbf{Z}_i \boldsymbol{B}\right)
\end{aligned}
$$

Where $\Phi(\cdot)$ is the cumulative distribution function of the standard
normal distribution.

#### Variables

##### Dependent variable

Our analysis utilizes data on coup attempts and outcomes from @powell2011. A
successful coup is defined as one where the incumbent leader is removed from
power for more than seven days. The dataset covers the period from 1950 to 2023
and includes information on 491 coup attempts, with roughly half (245) being
successful. Descriptive statistics for these coup attempts and regime types can
be found in @tbl-coups and @tbl-regimes.

-   **Coup Attempt**: Binary variable indicating whether a coup attempt
    occurred (1) or not (0) in a given country-year.

-   **Coup Success**: Binary variable indicating whether a coup attempt was
    successful (1) or failed (0), conditional on a coup attempt occurring.

##### Key Independent Variable: Regime Type {#sec-chapter2322}

I categorize regime types following @geddes2014 (GWF), focusing on military,
personalist, and dominant-party regimes, with democracies and monarchies included
for comparison. Descriptive statistics for regime types are presented in
@tbl-regimes.

##### Control variables

Our control variables are chosen based on the research of @gassebner2016. They
analyzed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

-   **Economic Level:** Represented by GDP per capita. This measure provides an
    indication of the overall economic health and standard of living in a
    country. We use GDP per capita data (in constant 2017 international 1000
    dollars, PPP) from the V-Dem dataset by @fariss2022.

-   **Economic Performance:** Measured using the current-trend ($CT$) ratio
    developed by @krishnarajan2019. This ratio compares a country's current GDP
    per capita to the average GDP per capita over the previous five years. A
    higher $CT$ ratio indicates stronger economic performance. For a country $i$
    at year $t$, the $CT$ ratio is calculated as follows: $$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
    $$ {#eq-eq6}

-   **Political Stability**: This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    variable "actotal" in the Major Episodes of Political Violence dataset
    [@marshall2005current], with 0 representing the most stable conditions (no
    violence at all) and 18 representing the most unstable.

-   **Previous coups:** Included in the selection equation as either: a) The
    number of previous coups in a country (Model 1), or b) The time since the
    last coup attempt (Model 2 for robustness check).

### Results and Discussion

```{r}
#| label: tbl-coupmodel
#| tbl-cap: "Sample Selection Model of Regime Type and Coup Success, 1950-2019"
#| results: asis
#| echo: !expr format == "html"

model_1 <- coup_model|>
  filter(regime != "other") |>
  select(coup, regime, GDP_trend, GDP_pc, violence, coup_dummy = pre_coups) |> 
  mutate(regime = fct_relevel(regime, "dominant-party")) |> 
  selection(coup > 0 ~ regime + GDP_trend + GDP_pc + violence + coup_dummy,
            coup > 1 ~ regime + GDP_trend + GDP_pc + violence,
            #method = "2step",
            data = _) 

model_1_selection <- model_1
model_1_selection$param$index$betaO <- model_1$param$index$betaS
model_1_selection$param$index$betaS <- model_1$param$index$betaO

model_2 <- coup_model |>
  filter(regime != "other") |>
  select(Coup = coup, regime, GDP_trend, GDP_pc, violence, coup_dummy, time_since_last_coup) |> 
  mutate(regime = fct_relevel(regime, "dominant-party")) |> 
  selection(Coup > 0 ~ regime + GDP_trend + GDP_pc + violence + coup_dummy*time_since_last_coup,
            Coup > 1 ~ regime + GDP_trend + GDP_pc + violence,
            data = _) 

model_2_selection <- model_2
model_2_selection$param$index$betaO <- model_2$param$index$betaS
model_2_selection$param$index$betaS <- model_2$param$index$betaO

stargazer(model_1_selection, model_1, model_2_selection, model_2,
          #type = "text",
          float = F,
          header = F,
          dep.var.caption = "",
          dep.var.labels = c("Model 1", "Model 2"),
          multicolumn = T,
          dep.var.labels.include = T, 
          column.labels = c("Coup Attempts", "Coup Outcome","Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant", "Regime: Democracy", "\\hspace{1.6cm}Military", "\\hspace{1.6cm}Monarchy", "\\hspace{1.6cm}Personalist", "Economic trend", "GDP per capita", "Political violence", "Previous coups (P)", "Yrs since coup (Y)", "Interaction term: P * Y"),
          summary.logical = F,
          column.sep.width = "7pt",
          font.size = "small"
          )
```

The double probit model with sample selection, estimated using the
***sampleSelection*** package [@sampleSelection-2] in R, provides valuable
insights into the factors influencing coup attempts and their outcomes across
different regime types from 1950 to 2019 (@tbl-coupmodel). I present two models
that differ slightly in their treatment of previous coups: Model 1 incorporates
the number of previous coups, while Model 2 utilizes the time elapsed since the
last coup.

#### Selection Model: Coup Attempts

In the selection model (Column 1), military and personalist regimes exhibit
significant positive coefficients at the 1% level, indicating a higher likelihood
of experiencing coup attempts compared to dominant-party regimes. This aligns
with our theoretical expectations regarding internal power struggles within
military juntas and succession vulnerabilities in personalist regimes.

```{r}
#| label: tbl-mfx1
#| tbl-cap: Average marginal effects of coup attempts (Selection of Model 1)
#| echo: false

avg_comparisons(model_1) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_1, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups",
                          term == "GDP_trend" ~ "Economic trend",
                          term == "GDP_pc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Regime: Democracy",
                          str_detect(contrast, "military") ~ "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Regime: Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_style(
    style = cell_text(indent = px(63)),
    locations = cells_body(
      columns = Term,
      rows = 2:4
    )
  ) |> 
  tab_options(
    table.width = pct(95),
    # table.font.size = px(14),
    column_labels.font.weight = "bold"
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  )
```

@tbl-mfx1 clarifies the regime effects using Average Marginal Effects (AME) and
ratio percentages. The military regime's marginal effect of 0.07 indicates that
the probability of coup attempts in military regimes is 7 percentage points (pp)
higher than in dominant-party regimes, ceteris paribus. This translates to
military regimes being about 277.7% more likely to encounter coups than
dominant-party regimes. Similarly, personalist regimes show a 2.4 pp higher
probability, about 94% more likely compared to dominant-party regimes.

Control variables show effects in expected directions but with weaker magnitudes.
Stronger economic performance, indicated by higher economic growth trends and GDP
per capita levels, correlates with a lower risk of coup attempts. Political
violence shows a positive effect, indicating that higher levels of instability
increase the likelihood of coups. The positive coefficient for the number of
previous coups suggests a "copycat" effect from earlier incidents.

#### Outcome Model: Coup Success

The outcome model (Columns 2 and 4 in @tbl-coupmodel) reveals determinants of
coup success. Military regimes demonstrate a higher probability of coup success
compared to dominant-party regimes, aligning with expectations that military
regimes face higher coup risks due to their increased chances of success.
Personalist and monarchical regimes show slight positive effects on coup success,
but these effects are not statistically significant.

Control variables exhibit different patterns in the outcome model compared to the
selection model. Both GDP per capita and political violence maintain a weak
influence, similar to their effects in the selection model. However, the economic
trend shows a less significant negative effect on coup success.

#### Model Comparison (Model 1 vs Model 2)

Model 2 employs years since the last coup instead of the number of previous
coups, with an interaction term between previous coups (as a binary variable) and
years since the last coup. Generally, Model 2 shows results in the same direction
as Model 1, albeit with relatively lower coefficients (@tbl-mfx2).

```{r}
#| label: tbl-mfx2
#| tbl-cap: Average marginal effects of coup attempts (Selection of Model 2)
#| echo: false


avg_comparisons(model_2) |> 
  select(1:3) |> 
  select(1:2, ame = estimate) |>
  cbind(avg_comparisons(model_2, comparison = "ratio") |> select(3)) |> 
  mutate('ratio(%)' = round((estimate - 1) * 100, 2),
         ame = round(ame, 3)) |> 
  select(-estimate) |>
  mutate(term = case_when(term == "coup_dummy" ~ "Previous coups (P)",
                          term == "time_since_last_coup" ~ "Yrs since coup (P)" ,
                          term == "coup_dummy*time_since_last_coup" ~ "Interaction term: P * Y",
                          term == "GDP_trend" ~ "Economic trend",
                          term == "GDP_pc" ~ "GDP per capita",
                          term == "violence" ~ "Political violence",
                          str_detect(contrast, "democracy") ~ "Regime: Democracy",
                          str_detect(contrast, "military") ~ "Military",
                          str_detect(contrast, "monarchy") ~ "Monarchy",
                          str_detect(contrast, "personal") ~ "Personal",
                          TRUE ~ term
                          )) |> 
  mutate(term = factor(term, levels = c(
  "Regime: Democracy",
  "Military",
  "Monarchy",
  "Personal",
  "Economic trend",
  "GDP per capita",
  "Political violence",
  "Previous coups (P)",
  "Yrs since coup (Y)",
   "Interaction term: P * Y"
)),
contrast = str_replace(
    contrast, 
    "mean\\((.+?)\\) - mean\\((.+?)\\)", 
    "mean(\\1 - \\2)"
  )) |> 
  arrange(term) |> 
  clean_names(case = "title") |>
  rename(AME = Ame) |> 
  gt() |> 
  cols_align(align = "left", columns = Term) |> 
  tab_style(
    style = cell_text(indent = px(63)),
    locations = cells_body(
      columns = Term,
      rows = 2:4
    )
  ) |> 
  tab_options(
    table.width = pct(95),
    # table.font.size = px(14),
    column_labels.font.weight = "bold"
    ) |> 
  fmt_number(
    decimals = 3          # specify the number of decimal places
  ) |> 
  tab_footnote(
    footnote = "AME: Average Marginal Effect",
    locations = cells_column_labels(columns = AME)
  )
```

The differences between Model 1 and Model 2 suggest that while the recency of
coups matters, the overall history of coups in a country may have a stronger
influence on future coup attempts.

#### Discussion of key findings

The $\rho$ values of 0.898 in Model 1 and 0.386 in Model 2, significant at 1% and
10% levels respectively, indicate strong correlation between unobserved factors
influencing coup attempts and coup success. This supports the appropriateness of
the sample selection model and underscores the importance of considering both
stages in the analysis.

The significant coefficients with theoretically consistent directions suggest the
model effectively captures key aspects of coup dynamics. The observed disparity
between coup attempt rates and success rates across regimes points towards
selection bias, further validating the use of the sample selection model.

#### Theoretical Implications

These results strongly support my theoretical framework, highlighting the crucial
role of regime structure in determining coup vulnerability. The findings
underscore that coups are strategic actions undertaken when odds appear
favourable, rather than random events.

### Conclusion

This study addresses the lack of consensus in empirical research on coup
predictors by introducing a novel approach that prioritizes determinants based on
their impact on coup success. By analysing coup success rates, I posit that the
expected outcomes of coups are critical determinants of their occurrence.
Employing a double probit model with sample selection, the research investigates
and confirms a strong and robust relationship between regime types and coup
attempts.

The main findings reveal that regime type plays a pivotal role in the likelihood
of coup attempts. Military and personalist regimes, characterized by weaker
institutional frameworks and higher vulnerability, demonstrate significantly
higher susceptibility to coups compared to dominant-party regimes. This
underscores the importance of supporting initiatives that establish regimes with
constitutional institutions rather than those dependent on military power or
personal authority, as the latter prove more volatile and coup-prone.

The research also indicates that stronger economic performance, while not as
influential as regime type, is associated with a lower risk of coups. This
suggests that policies promoting economic development can be effective in
reducing coup risk. Given that regime type is often determined during regime
formation and is difficult to alter, promoting economic growth might be the most
viable coup-proofing strategy available to incumbents.

However, these findings present a paradox for autocratic leaders, particularly
dictators and military juntas. While institutionalization of regimes could
enhance stability and reduce coup risk, few such leaders are willing to implement
these reforms. This reluctance stems from the potential constraints on their
power or shortened terms that such changes might entail. Thus, while institutions
may benefit the regime's longevity, they do not necessarily serve the immediate
interests of individual leaders.

This study opens avenues for further research, including:

-   Investigating the long-term effects of regime types on political stability,
    including democratization or personalization.

-   Examining how international factors might influence regime choices and coup
    dynamics when internal factors are too weak to prevent coups.

In conclusion, this chapter provides robust evidence for the critical role of
regime type in determining coup risk, while also acknowledging the complex
realities of political power. By illuminating the strategic calculus behind coup
attempts, it offers valuable insights for scholars seeking to understand
political instability and policymakers working to promote democratic transitions
and political stability in diverse global contexts.

## Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders {#sec-chapter3}

### Abstract {.unnumbered}

This study aims to clarify the concept of autocoups, specifically focusing on
power extensions by incumbent leaders. By distinguishing autocoups from the
broader and more ambiguous concepts of self-coups or executive takeovers, which
encompass both executive power aggrandizement and power extension, this research
redefines the concept of autocoups. Based on this refined definition, I introduce
a novel dataset of autocoup events from 1945 to 2024. Using this newly compiled
dataset, the research includes three types of case studies that provide
qualitative insights into the dynamics of autocoups. Additionally, an empirical
analysis of the determinants of autocoup attempts and success is offered to
demonstrate how the autocoup dataset can be employed for more quantitative
research. This study contributes to the existing literature by providing a
clearer conceptual framework and a novel dataset of autocoups. It enhances our
understanding of the mechanisms and motivations behind power extensions by
incumbent leaders and examines the implications for democratic backsliding,
democratic breakdown, and autocratic deterioration. The insights gained from this
study could draw more attention to the effects of autocoups on power transitions,
political stability, and democratic resilience.

keywords: *Autocoups, Coups, Power transitions*

\newpage

### Introduction

The study of irregular power transitions has predominantly focused on coups due
to their frequency and significant impact. This emphasis has led to substantial
scholarly attention and the creation of comprehensive datasets. However, as noted
by @gassebner2016, the multitude of potential factors proposed to explain coups
has often resulted in increased confusion rather than a clearer understanding of
coup dynamics.

In contrast, another form of irregular power transition—the incumbent leader's
refusal to relinquish power—has received relatively less attention despite its
significance. Recent decades have witnessed a decline in traditional coups and a
rise in this "incumbent retention or overstay" type of irregular power
transition, particularly since the end of the Cold War (@ginsburg2010evasion;
@baturo2014; @versteeg2020law).

This chapter will redefine and clarify this type of irregular power transition,
where leaders overstay their mandated term limits, as "autocoup." Although
analyses related to autocoups are not rare, the existing literature exhibits
several notable shortcomings:

-   **Terminological Ambiguity:** Terms such as self-coups, autocoups,
    autogolpes, incumbent takeovers, executive aggrandizement, overstay, and
    continuismo are used without a clear, universally accepted definition. This
    lack of standardization leads to confusion and inconsistent application
    [@marsteintredet2019; @baturo2022].

-   **Limited Data:** Due to the conceptual ambiguity surrounding autocoups, data
    collection remains in its early stages compared to the rich datasets
    available for traditional coups.

-   **Methodological Gaps:** Research on autocoups has largely relied on case
    studies [@cameron1998; @antonio2021; @pion-berlin2022], with few studies
    employing quantitative analysis.

Moreover, analyses of autocoups are often not integrated with those of
traditional coups. As a distinct category of coup, autocoups lack a corresponding
definition in relation to classic coups. Consequently, traditional coups and
autocoups are frequently analysed separately, despite their related nature, and
there is a paucity of comparative analyses.

Studying autocoups is important for several reasons. Firstly, autocoups can
undermine the rule of law, weaken institutions, and contribute to democratic
backsliding or authoritarian personalization. Secondly, like traditional coups,
successful autocoups increase the likelihood of future irregular power
transitions. For example, since 1945, approximately 62% of leaders who extended
their terms through autocoups in non-democratic countries were either ousted or
assassinated while in office [@baturo2019]. Thirdly, failed autocoups often lead
to instability, inciting protests, violence, and even civil wars.

This chapter aims to address these gaps by focusing on autocoups, aiming to
clarify terminology, refine concepts and definitions, enhance data collection,
and explore determinants through empirical analysis, contributing in three key
areas:

-   **Conceptual Clarification:** The term "autocoup" will be redefined and
    clarified, with a focus on power extension.

-   **Data Collection:** A new dataset of autocoups since 1945 will be introduced
    based on this refined definition.

-   **Empirical Analysis:** Utilizing this dataset, a quantitative analysis of
    the factors influencing leaders' decisions to attempt autocoups will be
    conducted.

The structure of this chapter is as follows: Section 2 will review definitions
related to power expansions and extensions, leading to a precise definition of
autocoups. Section 3 will present the new autocoup dataset. Sections 4 and 5 will
explore the determinants of autocoup attempts through case studies and
demonstrate the application of the dataset in empirical analysis. The conclusion
will summarize key findings and suggest directions for future research.

"Definitions of power expansion and power extension in political science can
often be ambiguous or overlapping. To ensure clarity in this study, we propose
distinct definitions for these concepts:

-   Power Expansion: This refers to the process by which an incumbent leader
    acquires additional authority or control over state apparatuses beyond their
    original mandate. This may involve centralizing power, reducing checks and
    balances, or assuming roles typically held by other branches of government.

-   Power Extension: This concept specifically relates to the temporal aspect of
    an incumbent's rule. It describes situations where a leader prolongs their
    tenure beyond the originally mandated term in office, often through
    constitutional amendments, cancellation of elections, or other means of
    circumventing term limits.

By distinguishing between these two concepts, we aim to provide a more nuanced
analysis of how leaders consolidate and maintain their positions. While power
expansion and extension can often occur simultaneously, differentiating between
them allows for a more precise examination of the strategies employed by leaders
to entrench their rule.

This distinction is crucial for our study as it enables us to: 1. Analyze the
different mechanisms leaders use to solidify their power 2. Examine the potential
consequences of each type of power consolidation 3. Explore the relationship
between power expansion and extension in various political contexts

Understanding these nuances is essential for comprehending the complex dynamics
of political power and leadership longevity in both democratic and authoritarian
systems."

### Autocoups: A literature review and clarification of definitions

The concept of autocoups, or the illegitimate seizure of power by incumbent
leaders, remains a complex and contested area of study. Unlike coups, which are
generally understood as illegal attempts by elites to overthrow the government,
autocoups lack a clear and consistent definition. This ambiguity hinders our
ability to accurately identify, analyze, and compare instances of this
phenomenon.

#### Terminology

The most common term in autocoup literature is "self-coup," or "autogolpe" in
Spanish [@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017;
@marsteintredet2019]. This term gained academic prominence after Peruvian
President Alberto Fujimori dissolved Congress, temporarily suspended the
constitution, and ruled by decree in 1992 [@mauceri1995; @cameron1998]. However,
as @marsteintredet2019 point out, the term "self-coup" can be misleading, as it
implies a coup against oneself, which is inaccurate since it targets other state
institutions or apparatus.

Another approach to describe coups staged by incumbents is to use terms with
adjectives or modifiers, such as "presidential coup," "executive coup,"
"constitutional coup," "electoral coup," "judicial coup," "slow-motion coup,"
"soft coup," and "parliamentary coup" [@marsteintredet2019]. While these terms
can be helpful in specific contexts, their proliferation often adds to the
overall confusion rather than providing clarification. Most of these terms focus
on the specific methods used by coup perpetrators but fail to clearly identify
the perpetrator, necessitating further explanation. In fact, many of these
methods could be employed either by or against executive leaders.

A third alternative involves terms like "incumbent takeover," "executive
takeover," or "overstay." Incumbent takeover refers to "an event perpetuated by a
ruling executive that significantly reduces the formal and/or informal
constraints on his/her power" [@baturo2022, p.374], based on earlier research
[@svolik2014]. Meanwhile, overstay is defined as "staying longer than the maximum
term as it stood when the candidate originally came into office"
[@ginsburg2011evasion, p.1844]. These terms identify the perpetrator (the
incumbent) and/or the nature of the event (overstaying/extending power). However,
they do not highlight the illegality or illegitimacy of these actions. Therefore,
they cannot serve as a direct counterpart to "coup," which clearly denotes the
illegality of leadership ousters, while "takeover" or "overstay" diminish the
severity.

As these terms often lack precision, focusing on specific methods rather than the
core act of power usurpation, this study proposes "autocoup" as the most suitable
term for this phenomenon. Unlike other terms, 'autocoup' clearly identifies the
perpetrator and the illegitimate nature of the power grab, distinguishing it from
traditional coups.

#### Definition

While terminology is important, another issue arises with the previous definition
of autocoups: What is the emphasis of an autocoup—power expansion, power
extension, or both?

Definitions of power expansion and power extension in political science can often
be ambiguous or overlapping. To ensure clarity in the study of autocoups, I
propose distinct definitions for these concepts:

-   **Power Expansion:** This refers to the process by which an incumbent leader
    acquires additional authority or control over state apparatuses beyond their
    original mandate. This may involve centralizing power, reducing checks and
    balances, or encroaching on the authority of other branches like the
    legislature or judiciary.

-   **Power Extension:** This describes situations where a leader prolongs their
    tenure beyond the originally mandated term in office, often through
    constitutional amendments, cancellation of elections, or other means of
    circumventing term limits.

Existing definitions of autocoups or related concepts often either are ambiguous
between power expansion and extension, or focus more on power expansion, which
has several drawbacks.

Firstly, defining autocoups primarily in terms of power expansion does not align
well with the definition of a coup. The focus of a classical coup is clearly on
the ouster of the current leader, not merely a limitation or restriction on their
power. Using the same logic, a more appropriate definition of an autocoup should
prioritize the tenure extension of executive leadership. Power restriction on
incumbents would not be coded as a coup as long as they remain in office.
Similarly, an executive leader acquiring more power from other branches could be
coded as power aggrandizement, but not an autocoup, as long as they step down
when their term expires.

Secondly, emphasizing power expansion in autocoups often neglects the ultimate
purpose of incumbents. It is irrational for an incumbent to expand executive
power and then pass the powerful role to future leaders. Although the term
"self-coup" gained prominence from the 1992 Fujimori case in Peru, which
initially involved seizing power from other institutions, it is important to note
that Fujimori ultimately extended his term limits through constitutional
amendments. The 1993 Constitution allowed Fujimori to run for a second term,
which he won with popularity in April 1995. Shortly after Fujimori began his
second term, his supporters in Congress passed a law of "authentic
interpretation" which effectively allowed him to run for another term in 2000,
which he won amid suspicions and rumors. However, he did not survive the third
term. In 2000, facing charges of corruption and human rights abuses, Fujimori
fled Peru and took refuge in Japan [@ezrow2019].

Thirdly, measuring the extent of power expansion to qualify as an autocoup can be
challenging. As @cameron1998a defined, a self-coup is "a temporary suspension of
the constitution and dissolution of congress by the executive, who rules by
decree until new legislative elections and a referendum can be held to ratify a
political system with broader executive power" (p. 220). "Broader executive
power" is difficult to define, and it would be problematic and disputable no
matter how it is defined.

Therefore, this study argues that a more accurate definition of autocoups should
prioritize power extension as the core characteristic. It is easy to identify the
event and the outcome in the first place. In most cases, autocoups in terms of
power extension involve power expansions as their prerequisite and foreshadowing.

Based on these criteria, I define **an autocoup as the illegitimate extension of
an incumbent leader's term in office beyond the originally mandated limits
through unconstitutional means**. This definition emphasizes the core
characteristic of power extension while acknowledging the potential for power
expansion as a related phenomenon. Three key points need to be highlighted for
this definition.

Firstly, this definition refers to the actual leaders of the country, regardless
of their official titles. Typically, this would be the president; however, in
some cases, such as in Germany, the primary leader is the premier, as the
president is a nominal head of state.

Secondly, while the primary characteristic of an autocoup is extending the term
in office, this definition does not exclude instances of power expansion. Both
aspects can coexist, but the extension of the term is the central element.

Thirdly, autocoups, by their nature, subvert legal norms and established power
transfer mechanisms. No matter how legitimate they claim to be, their
illegitimacy is not beyond a reasonable doubt as long as the incumbents are the
direct beneficiaries. This critical aspect will be explored further in Section 3.

### Introduction to the Autocoup Dataset

#### Defining the scope

Categorizing political events as autocoups inevitably involves challenging
borderline cases. To maintain consistency and avoid ambiguity, this study adopts
a broad coding approach: All instances of incumbents extending their original
mandated term in office are coded as autocoups, regardless of the apparent
legality of the extension.

This approach is justified because truly legitimate amendments to power
transition institutions should apply only to subsequent leaders, not the
incumbent. Even when extension procedures appear legal, the legitimacy is
questionable when the incumbent is the direct beneficiary.

#### Classifying autocoups {#sec-classify}

Autocoups manifest in various forms. I categorize them based on several key
factors:

-   **Methods Employed**: Specific strategies used by incumbents (e.g.,
    constitutional amendments, election cancellation).

-   **Degree of Legality**: Extent of deviation from established legal norms.

-   **Duration of Extension**: Length of time the incumbent remains in office
    beyond designated term limits.

-   **Outcomes**: Whether the autocoup attempt succeeds or fails.

This study primarily focuses on the methods employed, while coding for other
aspects when information is available.

##### Evasion of term limits {.unnumbered}

Evasion of term limits is a common tactic employed in autocoups. Incumbents often
resort to seemingly legal manoeuvres to extend their hold on power. These
manoeuvres primarily involve manipulating constitutional provisions through
various means. The incumbents may pressure legislative bodies (congress) or
judicial institutions (Supreme Court) to reinterpret existing term limits, amend
the constitution to extend terms, or even replace the constitution altogether.
This might also involve popular vote through referendums, or a combination of
these approaches. The extension can range from a single term to indefinite rule.

These manoeuvres primarily involve manipulating constitutional provisions through
various means.

-   **Changing Term Length:** Incumbents might lengthen the official term
    duration (e.g., from 4 to 6 years) to stay in office longer, even if the
    number of allowed terms remains unchanged. Examples, in the dataset, include
    Presidents Dacko (CAR, 1962), Kayibanda (Rwanda, 1973), and Pinochet (Chile,
    1988).

-   **Enabling re-election:** This approach involves incumbents modifying legal
    or constitutional frameworks to permit themselves to run for leadership
    again, despite initial restrictions. These restrictions might include
    prohibitions on re-election, bans on immediate re-election, or term limits
    that the incumbents have already reached. An illustrative example is
    President Menem of Argentina in 1993, who leveraged this tactic to extend his
    tenure.

-   **Removing Term Limits Altogether:** This approach, as seen with President Xi
    Jinping of China in 2018, technically allows the leader to rule for life,
    although they may still need to participate in elections (a formality in such
    cases).

-   **Declaring** **Leader for Life:** This differs from removing term limits as
    the leader still faces elections (although potentially rigged or
    uncontested). An example is Indonesia's President Sukarno, who attempted to
    declare himself president for life in 1963 (ultimately unsuccessful).

These methods are often used in combination. Initially, the duration of a term is
extended, followed by amendments to allow re-election, then the removal of term
limits, and finally, the declaration of the leader for life. For example, Haitian
President François Duvalier amended the constitution in 1961 to permit immediate
re-election and then declared himself president for life in 1964.

##### Election Manipulation or Rigging {.unnumbered}

Election manipulation or rigging is the second most commonly used tactic to
extend an incumbent's tenure.

-   **Delaying or Removing Elections:** Delaying or removing scheduled elections
    without legitimate justification is a frequent method used by incumbents to
    maintain power. For instance, Chadian President François Tombalbaye delayed
    general elections until 1969 after assuming power in 1960. Similarly, Angolan
    President José Eduardo dos Santos suspended elections throughout his rule
    from 1979 to 2017.

-   **Refusing Unfavourable Election Results:** Incumbents may refuse to accept
    unfavourable election results and attempt to overturn them through
    illegitimate means. For example, President Donald Trump of the United States
    refused to accept the results of the 2020 election and tried to overturn
    them.

-   **Rigging Elections:** Winning elections with an extraordinarily high
    percentage of votes is highly questionable. This study will code elections
    where the incumbent wins more than 90% of the vote as autocoups. For
    instance, President Teodoro Obiang of Equatorial Guinea has consistently won
    elections with over 95% of the vote in multi-party elections since 1996,
    indicating election rigging.

-   **Excluding Opposition in Elections:** Manipulating the electoral process by
    excluding opposition parties or candidates from participation, effectively
    creating a one-candidate race, clearly signifies an autocoup.

##### **Use of Figurehead** {.unnumbered}

To circumvent term limits, some incumbents might choose a close associate to act
as a figurehead, taking the office publicly while the incumbent retains real
power behind the scenes. This can be achieved through seemingly subordinate
positions.

One example is Russia in 2008. Facing term limits, President Putin selected
Dmitry Medvedev to run for president. After the election, Medvedev appointed
Putin as Prime Minister. However, most analysts believe Putin wielded the true
power throughout this period.

##### Reassigning supreme authority to a new role {.unnumbered}

This tactic involves an incumbent leader manipulating the constitution or legal
framework to create a new position of power, or elevate an existing one, before
stepping down from their current role. They then strategically take on this new
position, effectively retaining significant control despite appearing to
relinquish power. For example, in 2017, Recep Tayyip Erdoğan, the Prime Minister
of Turkey, spearheaded a constitutional referendum that transitioned the country
from a parliamentary system to a presidential one. This new system concentrated
significant executive power in the presidency. Following the referendum's
approval, Erdoğan successfully ran for the newly established presidency,
effectively retaining control under a different title.

##### One-Time Arrangement for Current Leaders {.unnumbered}

This strategy involves special arrangements that extend the term or tenure of
current leaders without altering the underlying institutions. For example,
Lebanon extended President Émile Lahoud's term by three years in 2004 through a
one-time arrangement.

#### Data Coding

```{r}
#| label: tbl-source
#| tbl-cap: Main Data Sources for Coding the Autocoup Dataset
#| echo: false
#| results: asis

read.csv(here("data/autocoup_source.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    2 ~ px(200)
  )
```

The autocoup dataset is built upon existing studies and datasets, ensuring a
comprehensive and reliable foundation. Table 1 outlines the main sources used for
coding the autocoup dataset.

The Archigos dataset [@goemans2009] and the Political Leaders' Affiliation
Database (PLAD) [@bomprezzi2024wedded] provide comprehensive data on all leaders
from 1875 to 2023, although our coding only includes autocoups since 1945. These
datasets are invaluable for identifying actual rulers, distinguishing them from
nominal heads of state.

The Incumbent Takeover dataset [@baturo2022] integrates data from 11 related
datasets, offering a broad spectrum of cases where leaders significantly reduced
constraints on their power. This dataset includes both power expansions and
extensions, necessitating cross-referencing with Archigos to verify
qualifications for autocoups.

```{r}
#| label: data-autocoups
#| echo: false

Autocoup <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_clean.csv")
# nrow(Autocoup)
# nrow(Autocoup |> filter(!is.na(source)))
```

In total, `r nrow(Autocoup)` observations were coded, with
`r nrow(Autocoup |> filter(!is.na(source)))` overlapping with the candidate data
from Incumbent Takeover. The remaining
`r nrow(Autocoup) - nrow(Autocoup |> filter(!is.na(source)))` events were newly
coded by the author through verification with other sources such as Archigos,
PLAD and news reports.

The main deviation from the Incumbent Takeover dataset arises from excluding
power expansions that do not involve attempts to extend tenure.

The dataset includes the following main variables:

-   **Country Identification:** Country code (*ccode*) and country name
    (*country*) from Correlates of War project [@stinnett2002].

-   **Leader Information:** Name of the de facto leader (*leader_name*,coded
    following Archigos and PLAD datasets).

-   **Timeline Variables:** Date the leader assumed power (*entry_date*), date
    the leader left office(*exit_date*), date of the significant event marking
    the autocoup (*autocoup_date*), and Start date of the leader's additional
    term acquired through the autocoup (*extending_date*).

-   **Power Transition Methods:** Categorical variable for how the leader entered
    power (*entry_method*), categorical variable for how the leader exited power
    (*exit_method*), dummy variable indicating regular (1) or irregular (0) entry
    (*entry_regular*), and dummy variable indicating regular (1) or irregular (0)
    exit (*exit_regular*).

-   **Autocoup Details:** Key variable capturing methods used to extend power
    (*autocoup_method*) and outcome of the autocoup attempt (*autocoup_outcome*,
    "fail and lose power", "fail but complete original tenure", or "successful").
    For successful coups, the additional term length can be calculated from the
    difference between *exit_date* and *extending_date*.

-   **Data Source:** Identifies the dataset source used for coding (*source*).

-   **Additional Notes:** Provides context for exceptional cases (*notes*).

The dataset encompasses a total of 14 variables along with the `notes` field.

There are a few coding challenges and decisions worth mention.

-   Determining Autocoup Date: For cases where extensions happen incrementally,
    the autocoup_date reflects a significant event marking the extension, such as
    a legislative vote or successful referendum.

-   Multiple Autocoup Attempts: In cases where a leader undertook multiple
    autocoup attempts, details are recorded in the notes field.

-   Distinguishing Autocoups from Power Expansions: Care was taken to
    differentiate between cases of power expansion and actual attempts to extend
    tenure, which required cross-referencing multiple sources.

-   Verifying Outcomes: Determining the success or failure of an autocoup attempt
    often required in-depth research, especially for less documented cases.

#### Data descriptions

```{r}
#| label: data_descriptions
#| echo: !expr format == "html"

country <- Autocoup |> 
  distinct(ccode) |> 
  summarise(count = n())

autocoup_desc <- Autocoup |> 
  summarise(autocoup_attempted = n(),
            autocoup_succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method) |> 
  arrange(-autocoup_attempted)
```

The primary coding has identified `r nrow(Autocoup)` autocoup cases from 1945 to
2024, involving `r country$count` countries. This comprehensive dataset provides
a rich source of information for analysing trends and patterns in autocoup
attempts across different political contexts.

@tbl-autocoup_method presents a breakdown of the autocoup methods employed by
leaders:

The most common autocoup method is "enabling re-election", accounting for
`r autocoup_desc$autocoup_attempted[1]` events. This is followed by "removing
term limits altogether" (`r autocoup_desc$autocoup_attempted[2]` cases), and then
"delaying elections" and "declaring the leader for life" (each with
`r autocoup_desc$autocoup_attempted[3]` cases).

The overall success rate of autocoups is 79%, which is significantly higher than
the approximately 50% success rate of classical coups. This high success rate can
be attributed to several factors:

-   Incumbent Advantage: Leaders already in power have access to resources and
    institutional mechanisms that can be leveraged to their advantage.

-   Gradual Implementation: Unlike sudden coups, autocoups can be implemented
    gradually, allowing leaders to build support and legitimacy over time.

-   Legal Facade: Many autocoup methods operate within a veneer of legality,
    making them harder to oppose openly.

-   Control of State Apparatus: Incumbents often have significant control over
    state institutions, which can be used to facilitate their autocoup attempts.

However, success rates vary significantly across different methods.

-   **100% Success Rate**: Removing term limits, delaying elections, declaring
    the leader for life, and cancelling elections all have perfect success rates.
    This suggests that once these processes are set in motion, they are difficult
    to reverse.

-   **Lower Success Rates:** Refusing to accept election results has the lowest
    success rate, with only 1 out of 4 attempts succeeding. Although the sample
    size is limited (only 4 cases in total), this trend might suggest several
    factors at play. These include greater democratic resilience in systems where
    general elections are regularly held, heightened international scrutiny and
    pressure in response to blatant manipulation of election results, and
    stronger domestic opposition to such overt power grabs.

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup methods and success rates (1945-2021)
#| echo: !expr format == "html"

autocoup_desc |> 
  adorn_totals() |> 
  mutate(Success_rate = percent(autocoup_succeeded/autocoup_attempted)) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(180),
    2:3 ~ px(140),
    4 ~ px(100)
  )

#autocoup_desc$autocoup_attempted[1]
```

### Determinants of Autocoup Attempts: Case Studies

#### High Frequency and Success Rate of Autocoups in Post-Communist Countries

Analysis of our dataset reveals a notably high frequency and success rate of
autocoups in post-communist countries. These nations, formerly communist regimes
prior to the collapse of the Soviet Union, have largely evolved into 'hybrid
regimes' [@nurumov2019], with only a few retaining their communist status. The
data documents 12 cases of autocoups aimed at prolonging incumbency in these
countries, with only two attempts failing. Examination of these cases highlights
several distinctive characteristics:

-   **Inherited Authoritarian Systems**: Despite most of these 12 countries
    transitioning from communist to non-communist governments (with the exception
    of China), they retained many authoritarian systems from their communist
    past.

-   **Continuity of Former Elites**: The transitions did not result in the
    removal or overthrow of previous ruling groups. Instead, former communist
    elites often maintained their positions of power.

-   **Subverted Democratic Processes**: While general elections and term limits
    were introduced in most of these countries, the legacy of former communist
    regimes frequently led to the circumvention of term limits and manipulation
    of elections [@nurumov2019].

##### Case 1: Lifelong Ruler–Alexander Lukashenko in Belarus

Alexander Lukashenko, a former member of the Supreme Soviet of the Byelorussian
Soviet Socialist Republic, became the head of the interim anti-corruption
committee of the Supreme Council of Belarus following the dissolution of the
Soviet Union. Elected as Belarus's first president in 1994, he has maintained
this position ever since. Initially, the 1994 constitution limited presidents to
two successive terms. However, Lukashenko removed this restriction in 2004.
International monitors have not regarded Belarusian elections as free and fair
since his initial victory. Despite significant protests, Lukashenko has
consistently claimed to win with a high vote share, often exceeding 80% in each
election. This pattern is evident across all five Central Asian countries of the
former Soviet Union, where post-dissolution leaders were typically high officials
or heads of the former Soviet republics who continued their leadership in the
presidency.

##### Case 2: Transferring Power to a Handpicked Successor–Nursultan Nazarbayev in Kazakhstan

Nursultan Nazarbayev served as the first president of Kazakhstan from 1991 until
2019. Prior to the dissolution of the Soviet Union, he held de facto leadership
as the First Secretary of the Communist Party of Kazakhstan. Following
independence, he was elected as the first president and retained office until
2019 through various means, including resetting term limits due to the
implementation of new constitutions. Notably, Nazarbayev did not officially
eliminate term limits but instead created an exemption for the "First President"
[@nurumov2019]. Unlike Lukashenko, who remains the incumbent of Belarus,
Nazarbayev transferred the presidency to a designated successor, Kassym-Jomart
Tokayev, in 2019. However, he retained significant influence as the Chairman of
the Security Council of Kazakhstan until 2022.

```{r}
#| label: latin
#| include: false

library(countrycode)
latin <- codelist |> 
  filter(region == "Latin America & Caribbean") |> 
  select(region, cown) |>
  drop_na() |> 
  pull(cown)

Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = country) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = autocoup_method) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% latin) |>
  summarise(count = n(),
            .by = c(autocoup_method,autocoup_outcome)) |>
  adorn_totals()
```

#### Autocoups for immediate re-election: Cases of Latin America

Latin America has a long-standing tradition of maintaining term limit
conventions. Simón Bolívar, the founding father of Bolivia, was initially a
strong advocate for term limits, stating in 1819, "Nothing is as dangerous as
allowing the same citizen to remain in power for a long time... That's the origin
of usurpation and tyranny" [@ginsburg2019, p.38]. Although Bolívar eventually
modified his stance, arguing in his 1826 Constitution Assembly speech that "a
president for life with the right to choose the successor is the most sublime
inspiration for the republican order," term limits became a convention in Latin
America. Approximately 81% of Latin American constitutions between independence
and 1985 imposed some form of term limits on the presidency
[@marsteintredet2019a].

An analysis of cases in Latin American countries reveals two notable patterns.

##### Often Successful at Breaking Non-re-election or Non-immediate Re-election Restrictions

Unlike other presidential systems where two terms are more common,
non-re-election or non-immediate re-election used to be prevalent in Latin
America. According to @marsteintredet2019a, non-consecutive re-election was
mandated in about 64.9% of all constitutions between independence and 1985, while
5.9% banned re-election entirely.

However, adherence to these conventions has varied across the region. Since
Mexico introduced non-re-election institutions in 1911 at the start of the
Mexican Revolution, they have remained inviolate [@klesner2019]. Similarly,
Panama and Uruguay have never altered their re-election rules, and Costa Rica has
only experienced a brief period (1897-1913) permitting immediate presidential
re-election since prohibiting it in 1859 [@marsteintredet2019a]. In many other
countries, however, constitutions have been frequently amended or violated.

The pursuit of re-election or consecutive re-election, therefore, has been a
significant trigger for autocoups aimed at power extension in this region. Our
research documents 32 autocoup cases, with over 50% (17 cases) attempting to
enable re-election or immediate re-election, and about 59% (10 cases out of 17)
being successful.

Unlike those who attempt to overstay in office indefinitely, many Latin American
leaders exit after their second term expires. Examples include President Fernando
Henrique Cardoso of Brazil (1995-2003), President Danilo Medina of the Dominican
Republic (2012-2020), and President Juan Orlando Hernández of Honduras
(2014-2022) [@ginsburg2019; @marsteintredet2019a; @landau2019; @baturo2019;
@neto2019].

##### Failing to Further Extend Tenure

This trend does not imply that none of these leaders attempted further
extensions, but rather that most accepted their unsuccessful outcomes without
abusing their power to manipulate the process. While autocoups aimed at securing
one additional term are often successful, attempts to overstay beyond this are
frequently unsuccessful.

Two contrasting cases illustrate the varied outcomes of term limit challenges:

-   **Unsuccessful Extension – Carlos Menem (Argentina)**: President Menem
    successfully extended his tenure by one term through a 1994 constitutional
    amendment allowing one executive re-election. He was subsequently re-elected
    in 1995. However, his attempt to reset his term count, arguing that his first
    term (1988-1995) should not count as it was under previous constitutions, was
    unanimously rejected by the Supreme Court in March 1999 [@llanos2019]. A
    similar scenario unfolded with President Álvaro Uribe of Colombia (2002-2010)
    [@baturo2019].

-   **Successful Extension – Daniel Ortega (Nicaragua)**: In contrast, Daniel
    Ortega, the incumbent president of Nicaragua, successfully extended his
    presidency. In 2009, the Supreme Court of Justice of Nicaragua permitted his
    re-election in 2011. Subsequently, in 2014, the National Assembly of
    Nicaragua approved constitutional amendments abolishing presidential term
    limits, allowing Ortega to run for an unlimited number of five-year terms. As
    a result, he has held the presidency since 2007 [@close2019].

#### As common as classical coups: Cases of African countries

```{r}
#| label: africa
#| include: false

africa <- codelist |> 
  filter(continent == "Africa") |> 
  select(continent, cown) |>
  drop_na() |> 
  pull(cown)

Autocoup |>
  filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = country) |>
  adorn_totals()
Autocoup |>
  filter(ccode %in% africa) |>
  distinct(country)
Autocoup |>
  filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = autocoup_method) |>
  adorn_totals()
Autocoup |>
  #filter(ccode %in% africa) |>
  summarise(count = n(),
            .by = autocoup_outcome) |>
  adorn_totals()
```

Classical coups have been prevalent in Africa, accounting for approximately 45%
of all global coups (219 out of 491 cases) since 1950, involving 45 out of 54
African countries (GIC dataset). While autocoups are less frequent compared to
traditional coups, they maintain a significant presence in Africa. Among 110
documented autocoup cases globally, 46% (51 cases) occurred in Africa, involving
36 countries. Notably, the success rate of autocoups in Africa is over 84% (43
out of 51 attempts), which surpasses both the success rate of classical coups in
the region (roughly 50%) and the global average success rate of autocoups (79%).

Identifying a clear pattern of autocoups in Africa is challenging, mirroring the
complexity observed with classical coups. Various factors have been proposed to
explain this phenomenon:

-   **Natural Resources**: Countries rich in natural resources, particularly oil
    or diamonds, may see leaders more likely to attempt and succeed in extending
    their terms [@posner; @cheeseman2015; @cheeseman2019a].

-   **Quality of Democracy**: The quality of democracy is a critical factor
    influencing respect for term limits [@reyntjens2016].

-   **International Influence**: International aid or donor influence can play a
    significant role in discouraging attempts at power extension [@brown2001;
    @tangri2010].

-   **Organized Opposition and Party Unity**: The extent of organized opposition
    and the president's ability to enforce unity within the ruling party are
    crucial factors [@cheeseman2019].

Utilizing the Africa Executive Term Limits (AETL) dataset, @cassani2020
highlights human rights abuses and the desire for impunity as main drivers for
incumbents to cling to power. The more authoritarian a leader, the more likely
they are to attempt to break term limits and overstay in office. A leader's
ability to secure the loyalty of the armed forces through public investment
increases the chances of success in overstaying.

Despite both coups and autocoups being prevalent, there has been a noticeable
shift since the end of the Cold War in 1991: Traditional coups have decreased in
frequency while autocoups have become more prevalent.

This trend can be partially attributed to the introduction of multi-party
elections in Africa in the 1990s, which also brought in term limits for
executives [@cassani2020; @cheeseman2019]. Before 1991, personal or military rule
was more common, and term limits were less frequent. Post-1991, with more term
limits introduced, challenges to these limits have increased. However, it is
crucial to note that this increase in challenges does not necessarily imply that
violations are more common than adherence to term limits, because total power
transitions have increased compared to the past.

### Empirical Analysis: An Example of Utilizing the Autocoup Dataset

The availability of the autocoup dataset has made it feasible to conduct
quantitative analyses that extend beyond traditional case studies. This section
provides a straightforward example of how to utilize this dataset effectively. To
analyze the determinants of autocoup attempts, I employ a probit regression
model. This approach differs from the double probit model with sample selection
used in Chapter 2 for coup attempts and success analyses. Instead, I use two
separate probit models. Due to the high probability of success in autocoups, they
do not exhibit the typical sample selection characteristics that necessitated the
use of a sample selection model in our earlier analysis of traditional coups.

#### Dependent Variables

-   **Autocoup Attempt**: Binary variable indicating whether an autocoup attempt
    occurred (1) or not (0) during the tenure of an incumbent leader.

-   **Autocoup Success**: Binary variable indicating whether an autocoup attempt
    was successful (1) or failed (0), conditional on an autocoup attempt
    occurring.

#### Independent Variables

The selection of independent variables are based on the analysis of traditional
coups, plus the population size and the leader's age.

-   **Regime types**: Following @geddes2014 (GWF), focusing on military,
    personalist, and dominant-party regimes, with democracies included for
    comparison.

-   **Economic Level**: Represented by GDP per capita. This measure provides an
    indication of the overall economic health and standard of living in a
    country. We use GDP per capita data (in constant 2017 international 1000
    dollars, PPP) from the V-Dem dataset by @fariss2022.

-   **Economic Performance**: Measured using the current-trend ($CT$) ratio
    developed by @krishnarajan2019, which is consistent with @eq-eq6 in
    Chapter 2.

-   **Political Stability**: This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    variable "actotal" in the Major Episodes of Political Violence dataset
    [@marshall2005current], with 0 representing the most stable conditions (no
    violence at all) and 18 representing the most unstable.

-   **Population Size:** To account for its potential impact on leaders' tenures,
    we consider the log of the population size. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on power transitions. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's Age:** The age of the leader is included as an additional variable
    in the analysis, offering insights into potential correlations with
    leadership strength. Older leaders may have different experiences, networks,
    and health considerations that could influence their ability to maintain
    power. This data is sourced from Archigos and PLAD datasets.

Unlike the analysis of coup determinants, which could theoretically occur in any
given year, I assume that an autocoup happens only once during an incumbent
leader's tenure, as a successful autocoup negates the need for another attempt.
However, this assumption does not always reflect reality, as leaders might
attempt further extensions or try again after a failed attempt. For simplicity, I
overlook these possibilities in our analysis.

Therefore, in our probit model, the unit of analysis for autocoups is the entire
tenure of a leader, rather than a country-year. I establish a base year for the
variables: for leaders who staged an autocoup, we use the year of their first
attempt as the base year; for leaders who did not attempt to overstay, I use the
middle year of their tenure as the base year.

#### Results and discussions

```{r}
#| label: tbl-autocoupmodel
#| tbl-cap: "Determinants of autocoup attempts and success (1945-2018)"
#| results: asis
#| echo: false

autocoup_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/autocoup_model.csv")


model_attempt <- autocoup_model |> 
  filter(!regime %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential")) |>
  glm(autocoup_attempt ~ regime + GDP_pc + GDP_trend + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

model_success <- autocoup_model |> 
  filter(!regime  %in% c("monarchy", "other")) |> 
  mutate(regime = fct_relevel(regime, "presidential"),
         autocoup_outcome = factor(autocoup_outcome)) |>
  filter(autocoup_attempt == 1) |> 
  glm(autocoup_outcome ~ regime + GDP_pc + GDP_trend + violence + age + pop_log, 
      family = binomial(link = "probit"),
      data = _)

stargazer(model_attempt, model_success, 
          header = F,
          float = F,
          column.labels = c("Autocoup Attempts", "Autocoup Outcome"),
          intercept.bottom = F,
          covariate.labels = c("Constant",
                               "Regime: Dominant-party",
                               "\\hspace{1.6cm}Military",
                               "\\hspace{1.6cm}Personalist",
                               "GDP per capita",
                               "Economic trend",
                               "Political stability",
                               "Age",
                               "Population(log)"),
          dep.var.caption = "",
          dep.var.labels.include = F,
          summary.logical = T,
          column.sep.width = "50pt"
          )
#view(autocoup_model)
```

@tbl-autocoupmodel summarizes the findings from the probit regression models
based on our analysis of the determinants of autocoup attempts and their success.

Model 1, which examines autocoup attempts, reveals only one significant predictor
besides the constant term. Among the regime types, personalist regimes
significantly increase the likelihood of autocoup attempts, all else being equal.
This suggests that leaders in personalist regimes are more prone to attempt to
extend their power through autocoups compared to leaders in democratic regimes
(reference regime). Leaders in dominant-party and military regimes, however, show
no significant difference in the likelihood of attempting an autocoup compared to
democratic leaders.

The model for autocoup success (Model 2) shows similar dynamics. Personalist
regimes again have a strong positive and significant effect on the success of
autocoups compared to democratic leaders. Dominant-party regimes also show a
positive and marginally significant effect. However, a detailed examination
reveals that about half of the successful autocoups in dominant-party regimes (9
out of 20) exhibit a personalist style, such as "party-personal-military"
regimes.

This outcome is logical since personalist leaders are typically much more
powerful than other types of leaders, making them more inclined and capable of
overstaying in power.

Other factors play an insignificant role in determining the attempts and outcomes
of autocoups. This aligns with our conclusions on the determinants of classic
coups. Both coups and autocoups are significantly affected by power dynamics. As
power transitions involve the struggle between seizing and maintaining power, the
balance of power status quo inevitably matters in both coups and autocoups. This
also explains the high success rate of autocoups. Compared to power challengers,
incumbents are in an obviously advantageous position. Incumbent leaders can use
state power to their benefit, which is difficult to counteract. Even the abuse of
power is often unchecked under a powerful leader's rule.

The empirical analysis of autocoups yields significant implications for
real-world politics. In particular, the high overall success rate of autocoups
highlights the vulnerability of democratic institutions to gradual erosion by
incumbent leaders. The threshold for ousting or impeaching an incumbent leader
through constitutional means is exceptionally high, with success often requiring
more than a simple majority and substantial support across various sectors.
Resorting to illegal means, such as a coup, presents even greater challenges due
to high costs, severe consequences, and a low likelihood of success.

Conversely, political dynamics, whether in democracies or autocracies, tend to
favour incumbents even when they act unconstitutionally. Incumbents can leverage
state resources to achieve their political ambitions, benefiting from a high
probability of success and minimal consequences in case of failure. This
asymmetry in power and risk creates a concerning scenario: for incumbents who do
not respect constitutional institutions, the opportunity to launch an autocoup
appears sufficiently low-risk to warrant an attempt.

### Conclusion

This chapter presents a comprehensive analysis of autocoups, focusing
specifically on political events where incumbent leaders illegitimately extend
their tenure in power. By refining the concept and distinguishing it from broader
definitions such as 'self-coups', 'autogolpes', and 'executive takeovers', I
introduce a novel dataset of autocoups spanning from 1945 to 2024. Through this
more precise definition and dataset, this research expands the analysis of
irregular power dynamics from irregular transition to irregular retention,
offering a broader and more nuanced review of the phenomenon.

The findings reveal that personalist regimes are significantly more likely to
experience autocoup attempts and succeed in these attempts compared to
democracies. Dominant-party systems, often exhibiting personalist
characteristics, also show an association with successful autocoups. While regime
type significantly influences autocoups, other factors appear less impactful,
mirroring classic coups where the balance of power is a more essential
determinant. The high success rate of autocoups can be attributed to the inherent
advantages incumbents possess, such as control over or abuse of state power and
the difficulty of removing or impeaching them through legal or illegal means.

However, several limitations warrant consideration for future research. Firstly,
the definition of an autocoup requires further commentary and discussion to gain
wider acceptance in the academic community. Despite efforts to maintain
objectivity, some coding decisions may involve subjective judgements,
particularly in borderline cases. Secondly, due to the nature of autocoups, which
are less frequent than classic coups (491 coups versus 110 autocoups during the
same period), the quantitative analysis cannot be conducted as a country-year
variable as in coup studies. This raises the issue of choosing an appropriate
base year. For instance, when analysing how GDP level or growth rate, political
stability, or the age of the leader affects autocoup attempts, we must decide
which year's value should be used. In this study, I chose the middle year of a
leader's tenure or the year they staged the autocoup, but determining the most
appropriate year requires further discussion and potentially sensitivity
analyses.

Despite these limitations, this research significantly enhances our understanding
of the mechanisms and motivations behind autocoups, contributing to the
literature on political stability and democratic resilience. The findings
highlight the vulnerability of political systems, particularly democracies, to
erosion from within by incumbent leaders.

Future studies could build on this work by employing the dataset to explore more
nuanced power dynamics or examine the long-term impacts of these events on
political systems. Particularly fruitful areas for investigation include the
relationship between autocoups and democratic backsliding, democratic breakdown,
and the personalization of power. Additionally, comparative analyses between
autocoups and traditional coups could yield insights into the evolving nature of
power consolidation strategies in different political contexts.

In conclusion, this study not only provides a valuable resource for future
research but also contributes to our understanding of the complex interplay
between leadership, institutional structures, and political stability. As
autocratic tendencies continue to challenge democratic norms globally, the
insights gained from this analysis of autocoups become increasingly relevant for
both scholars and policy-makers concerned with preserving and strengthening
democratic institutions.

## Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-Entry and Autocoup Leaders

### Abstract {.unnumbered}

This chapter examines the relationship between methods of power acquisition and
the longevity of leaders who assume power through irregular means, specifically
coup-entry and autocoup leaders. The central hypothesis posits that the mode of
accession significantly influences leader tenure. Employing Cox proportional
hazards and time-dependent Cox models, this study provides robust evidence of
divergent survival times between these two leader types. Findings reveal that
coup-entry leaders face a substantially higher risk of removal from office
compared to their autocoup counterparts. These results have significant
implications for political stability and democratic processes, suggesting that
the perceived low costs and high rewards associated with autocoups may
incentivize incumbents to extend their tenure through this mechanism, potentially
contributing to democratic backsliding. This research makes a notable
contribution to the academic literature by leveraging a newly developed dataset
on autocoups. By doing so, it offers valuable insights into the dynamics of
irregular leadership transitions and enhances our understanding of the complex
interplay between power acquisition methods and political longevity.

**keywords**: *Coups, Autocoups, Leadership Survival, Cox Model*

\newpage

### Introduction

The longevity of political leaders has long captivated scholars and observers.
Why do some rulers endure for decades while others are ousted within months or
even days? This question has spurred extensive research in political science,
primarily focusing on the survival of political leaders. However, within this
broader framework, a specific subset of leaders—those who ascend to power through
coups or extend their tenure through autocoups—has received comparatively less
attention. Examining the tenures of these leaders is crucial, as it illuminates
the dynamics of irregular leadership transitions and their implications for
political stability and democratic processes.

Leaders who attain power through conventional means often follow predictable
patterns of ascension, tenure, and departure. In contrast, those who rise through
irregular channels, such as coups or autocoups, present more complex and
intriguing cases for study. Archigos dataset underscores the prevalence of such
irregular transitions: between 1945 and 2015, over half of leaders who assumed
power irregularly (158 out of 308) also exited irregularly. This rate
significantly surpasses that of leaders who accessed office through regular
channels, of whom only 14.5% (213 out of 1,472) experienced irregular departures.

Coup-entry and autocoup leaders constitute a substantial portion of these
irregular cases. Archigos dataset notes that of 374 leaders who exited
irregularly, 246 (65.8%) were ousted through coups. @frantz2016 further
demonstrate that coup-related exits account for approximately one-third of all
exits in autocracies, surpassing any other transition type. Additionally, the
autocoup dataset, introduced in [Chapter 3](#sec-chapter3), documents 110
autocoup attempts between 1945 and 2024, of which 87 were successful.

```{r}
#| label: data
#| include: false
survival_cox_ph_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_ph_model.csv") |> 
  mutate(survival_year = survival_days / 365)

survival_cox_td_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_td_model.csv")

avg <- survival_cox_ph_model |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)
```

Measuring the survival tenure of coup-entry and autocoup leaders presents
challenges due to the inherent irregularity and uncertainty of their positions.
Nevertheless, a comparative analysis reveals that leaders who extend power
through autocoups tend to have longer average post-autocoup tenures
(approximately `r avg$avg[1]` years) compared to coup-entry leaders
(approximately `r avg$avg[2]` years), suggesting a potential tenure gap of over
five years.

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders
#| echo: !expr format == "html"

# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = survival_cox_ph_model)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = survival_cox_ph_model, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("Autocoup leaders", "Coup-entry leaders"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 12)
```

A preliminary log-rank test in survival analysis, as illustrated in @fig-logrank,
demonstrates a statistically significant difference between the tenures of
autocoup and coup-entry leaders. The survival curve for autocoup leaders
consistently exceeds that of coup-entry leaders, indicating longer survival times
and a reduced risk of ouster for autocoup leaders.

This study posits that the method of accession significantly influences
leadership longevity. Coup-entry leaders likely confront greater challenges to
their rule, resulting in shorter average tenures compared to autocoup leaders.
The analysis, employing Cox proportional hazards and time-dependent Cox models,
supports this hypothesis, demonstrating that autocoup leaders generally
experience longer tenures than coup-entry leaders.

This research offers two primary contributions to the field. First, it highlights
an understudied factor in leadership survival analysis: the impact of the method
of accession to power. My findings suggest that leader survival is influenced not
only by ruling strategies but also by the initial method of acquiring power.
Second, by employing survival models, this study provides empirical evidence of
the significant difference in tenure duration between autocoup and coup-entry
leaders. This insight may explain the increasing prevalence of tenure extensions
through autocoups since 2000, as more incumbents observe and potentially emulate
successful precedents.

The remainder of this chapter is structured as follows: Section 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Section 3 explores the factors influencing the survival of
coup and autocoup leaders. Section 4 outlines the methodology and data used,
including the application of survival models to analyse the determinants of
leadership longevity. Section 5 presents the analysis findings and a detailed
discussion of the results. Finally, Section 6 concludes by synthesizing key
takeaways and exploring their broader implications for political stability and
democratic processes.

### Literature review

The longevity of political leaders has been a central focus in political science
research for decades, driven by the wide-ranging variations observed across
regimes, countries, and historical periods. This field encompasses two
interconnected aspects: regime survival and individual leader survival. Regime
survival focuses on the endurance of political systems, such as monarchies,
political parties, or specific ideological structures, while leader survival
concerns the duration of individual leaders' time in office.

These concepts often exhibit contrasting patterns. For instance, parliamentary
democracies like Japan or the United Kingdom may experience prolonged periods of
party dominance coupled with frequent leadership changes, whereas communist
regimes typically demonstrate enduring party rule with more frequent leadership
transitions. Presidential systems, such as the United States or many military
regimes, tend to exhibit more frequent changes in both ruling party or junta and
leader. This study specifically investigates the dynamics of individual leader
survival, focusing on factors influencing the duration of leaders' time in
office.

The existing literature on leader survival is extensive and multifaceted. Some
studies explore specific mechanisms influencing leadership longevity within
particular regimes, such as democracies [@svolik2014] or autocracies
[@davenport2021], while others aim to develop more generalizable theoretical
frameworks explaining leader survival across different political systems
[@buenodemesquita2003]. Although a universal theory remains an aspirational goal,
the complexities of leadership survival across diverse regime types present
significant challenges.

Power transition mechanisms vary substantially across different regimes,
particularly between democracies and autocracies. Autocratic systems often
feature closed leadership selection processes, restricted to a narrow pool of
individuals such as royal families, military elites, or ruling party members.
While some autocracies may hold elections, significant barriers to entry for
legitimate challengers typically persist. Potential rivals may face threats like
assassination, imprisonment, or exile. Moreover, the opacity of selection
processes makes it difficult to assess genuine levels of public support compared
to democracies. Consequently, conceptualizing selectorates or winning coalitions,
as proposed by @buenodemesquita2003, becomes problematic in many autocratic
contexts.

Given these complexities, focusing research on specific regimes or leader types
may be more fruitful. While regular leadership changes offer valuable insights,
they provide limited opportunities to explore the dynamics of leader survival. In
contrast, the study of irregular leaders, such as those who ascend to power
through coups or overstay in office through autocoups, offers a more compelling
avenue for research due to the inherent complexities and uncertainties
surrounding their leadership trajectories.

Two primary perspectives have emerged to explain the dynamics of leader survival.
The first emphasizes objective factors and resources, such as personal competence
[@yu2016], societal stability [@arriola2009], economic development [@palmer1999;
@williams2011], natural resource endowments [@smith2004; @quirozflores2012;
@wright2013], and external support [@licht2009; @wright2008; @thyne2017]. The
second focuses on subjective factors and strategies, including political
policies, responses to opposition, and tactics for consolidating power
[@gandhi2007; @morrison2009; @escribà-folch2013; @davenport2021].

Coups have received considerable scholarly attention, with research examining
coup prevention strategies, the impact of coups on leadership, and the subsequent
actions of coup leaders. Existing research delves into strategies for thwarting
coups [@powell2017; @sudduth2017; @debruin2020] and how leaders extend their
tenures after surviving coup attempts [@easton2018]. @sudduth2017 examines
post-coup actions of dictators, focusing on purge strategies, while @sudduth2018
investigates how leaders' entry methods affect their removal in dictatorships.

However, a significant gap exists in the literature regarding the comparison of
leadership survival between coup-entry and autocoup leaders. This study aims to
address this gap by investigating and comparing the duration of leadership
survival for these two leader types, contributing to a more nuanced understanding
of political survival in irregular leadership transitions.

### Survival dynamics of autocoup and coup-entry leaders

#### Autocoup leaders versus coup-entry leaders

As highlighted in the previous section, investigating leadership survival
presents inherent challenges due to factors such as the opacity and diverse
mechanisms of power transitions. However, these challenges underscore the
significance of this research, as it illuminates understudied dynamics in
political leadership.

While the survival of political leaders exhibits complexity and variation, it is
not entirely devoid of patterns. Leaders of similar types often display
significant comparability. Before delving into this comparison, it is essential
to clarify several key terminologies.

Firstly, the concepts of coup and autocoup adhere to the definitions established
in 1 and @sec-chapter3 respectively. However, a significant disparity exists in
the tenure lengths of coup-entry leaders and autocoup leaders. Many coup-entry
leaders rule for only a few months or even days, while autocoup leaders typically
enjoy longer tenures after their power grab. To ensure meaningful analysis, this
chpater focuses on leaders with more substantial periods in power. As defined in
@sec-chapter3, an autocoup is considered successful if the power extension lasts
for at least six months. For consistency and to allow for more robust comparison,
this chapter applies the same six-month threshold to both autocoup and coup-entry
leaders. This approach differs from the seven-day duration used by @powell2011
for traditional coups, allowing us to examine more established leadership periods
and their associated dynamics.

Secondly, it is crucial to distinguish between an autocoup leader and a
coup-entry leader, as leadership survival is the primary focus of this study:

-   **Autocoup leader:** This refers to an incumbent leader who successfully uses
    illegitimate or unconstitutional means to extend their tenure in power. In an
    autocoup, the leader orchestrates the power grab and continues to rule
    afterwards.

-   **Coup-entry leader:** This term designates the individual who assumes power
    after a successful coup. The coup leader and the coup-entry leader may or may
    not be the same person. Unlike autocoups, coups often involve multiple
    leaders (individuals or groups) who overthrow the incumbent leader, but
    typically only one of them assumes supreme power. In some instances, coup
    leaders may support someone outside the coup plot to become the new leader,
    such as military officers returning power to civilians or supporting a new
    general election. Regardless of the specific scenario, a coup-entry leader in
    this study refers to the individual who assumes formal leadership following a
    successful coup.

Given that autocoup leaders typically exhibit longer overall tenures compared to
coup-entry leaders, this study focuses on a more nuanced comparison.
Specifically, we will analyse the **post-autocoup** tenure of autocoup leaders
and contrast it with the **post-coup** tenure of coup-entry leaders. This
examination is motivated by the relevance and similarity of these leader types in
terms of illegitimacy, uncertainty, and instability.

#### Different challenges and tactics to consolidate power

Previous research has established that the ability to skillfully retain power is
the primary determinant of leader longevity. Leaders who can maintain control or
manipulate the balance of power tend to have longer tenures. However, coup-entry
and autocoup leaders face distinct challenges in consolidating their power,
primarily due to differences in the intensity of issues related to illegitimacy,
uncertainty, and instability. This disparity creates an uneven playing field in
terms of power dynamics, with coup-entry leaders at a significant disadvantage.
The following analysis examines these challenges and their impact on leader
tenure.

##### Illegitimacy: Both coup-entry and autocoup leaders suffer from a legitimacy deficit, but the nature of this deficit differs

While both coup-entry and autocoup leaders suffer from a legitimacy deficit, the
nature of this deficit differs. Coup leaders openly seize power, making their
illegitimacy explicit. Conversely, autocoup leaders employ a deceptive strategy,
manipulating legal processes to create a facade of democratic legitimacy. Despite
this veneer, their actions fundamentally undermine democratic principles.
Scholars have suggested alternative terms like "incumbent overstay" or "executive
takeover" to reflect the less overt nature of power grabs in these cases. This
perceived legitimacy can provide autocoup leaders with a temporary advantage as
opponents are often restricted to legal challenges.

##### **Uncertainty:** The irregular paths to power taken by both types of leaders create uncertainty regarding their reigns and eventual departures

The irregular paths to power taken by coup-initiators and autocoup leaders create
uncertainty regarding their reigns and eventual departures. Their ascension
through irregular means undermines established power transition norms, leaving
doubts about their commitment to constitutional succession protocols. This
uncertainty not only unsettles elites and citizens but also plagues the leaders
themselves, who grapple with the ambiguity surrounding the transfer of power –
when, how, and to whom. Historical analyses underscore this predicament, with
data revealing that more than two-thirds of irregular exits from leadership stem
from coup-related upheavals [@goemans2009].

Coup-entry and autocoup leaders face different levels of uncertainty following
their rise to power. After a coup, three major uncertainties arise:

-   **Leadership Assumption**: It is unclear who will assume leadership. Although
    coup leaders often take power, some may return or promise to return power to
    civilian leaders. Even among coup leaders, determining who will lead can be
    problematic, as coup plotters are sometimes a group without a clear core
    leader. For instance, following the 1973 Chilean coup, the initial plan for a
    rotating presidency among military leaders was abandoned when General
    Pinochet consolidated control and remained in power until 1990 [@svolik2014].

-   **Duration of Rule**: The duration of the coup leader's rule is uncertain.
    Leaders like Gamal Abdel Nasser in Egypt (1954 coup), Muammar Gaddafi in
    Libya (1969 coup), and Idi Amin in Uganda (1971 coup) aimed to retain power
    for life [@geddes2018], but their ability to do so was uncertain. Others
    promise to transfer power to civilian authorities, but the timing and
    fulfilment of these promises are unclear. For example, Myanmar's military
    junta (2021 coup) has repeatedly extended a state of emergency, clinging to
    power beyond the promised time-frame[^3]. Conversely, after the 2010 coup in
    Niger, the military honoured their promise by restoring civilian rule within
    the same year [@ginsburg2019].

-   **Succession**: The successors of coup leaders are uncertain. Some may
    designate successors from their inner circle, including family members, while
    others may support general elections, though whether this will be fulfilled
    as intended remains uncertain.

In contrast, autocoup leaders present a clearer picture regarding leadership and
tenure. There is no ambiguity about who will rule after an autocoup. In the
medium term, autocoup leaders typically hold office themselves. Many, like Putin
in Russia and Xi Jinping in China, seek to extend their rule indefinitely and are
unlikely to relinquish power voluntarily. Others attempt to extend their terms
incrementally, such as President Menem of Argentina, who overstayed successfully
in 1993 but failed in his bid for another term in 1999 [@llanos2019].

##### **Instability:** The awareness of shaky legitimacy and persistent uncertainty breeds insecurity and a sense of crisis

The awareness of their shaky legitimacy and the persistent uncertainty breeds
insecurity and a perpetual sense of crisis among coup-entry and autocoup leaders.
In a bid to solidify their grip on power, they often resort to reshaping power
dynamics or purging potential adversaries. Paradoxically, these attempts to
bolster stability frequently backfire, unleashing greater turmoil and
instability.

The stability of a regime, particularly in an autocracy, hinges on maintaining a
balance of power. Coups, however, inevitably disrupt this balance, even when they
are bloodless, necessitating the creation of a new equilibrium. The ousting of
previous rulers requires dismantling the established governing structure and
reshuffling high-ranking officials, actions that inherently generate instability
and create adversaries for the new leadership. This makes restoring order and
establishing a balanced power structure notably challenging. Studies show that
new leaders often purge rival elite groups to consolidate their power at the
outset of their tenure [@sudduth2017; @roessler2011].

Such actions can provoke backlash even from close allies. For instance, in
Uganda, President Obote's attempt to undermine the army commander-in-chief, Idi
Amin, led to Amin gaining the army's support and ultimately ousting Obote in a
1971 coup. Similarly, in Pakistan in 1999, shortly after Prime Minister Sharif
dismissed powerful army chief General Pervez Musharraf, Sharif himself was ousted
in a coup orchestrated by Musharraf and his supporters [@sudduth2017].

To consolidate power, coup-entry leaders often have to compromise with internal
or external power challengers. However, these compromises are frequently unstable
and easily broken. The situation becomes even more complex when there is a risk
of civil war. Leaders may attempt to reduce the likelihood of subsequent coups,
potentially increasing the chances of societal rebellions and civil wars
[@roessler2011].

Moreover, instability extends beyond leadership to policies. A new leadership
group often brings new policies, and coups are sometimes triggered by
disagreements over significant policies. Major policy shifts can instigate
dissent or grievances from various ruling factions, communities, regions,
ethnicities, or religions. In contrast, autocoup leaders encounter fewer of these
issues, as their regimes experience fewer abrupt changes. They face less pressure
to dismantle the existing ruling paradigm and establish a new order. Even when
adjustments are necessary, they have more time to implement changes gradually.

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-entry leaders 
#| echo: !expr format == "html"
#| results: asis
# library(kableExtra)
read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(100),
    2:3 ~ px(250)
  ) |>
   data_color(rows = c(2, 4), colors = "#edede9") 
```

Coup-entry leaders face significantly greater challenges in consolidating power
compared to autocoup leaders. This disadvantage creates a self-perpetuating
cycle. Weaker leaders struggle to attract and retain strong support, making them
more vulnerable to internal and external challenges. The perception of risk
discourages potential allies, further eroding their power base.

Empirical evidence supports this dynamic. Data reveals a correlation between the
frequency of coup attempts in a country and the likelihood of future coups. For
example, @tbl-coups shows over a third of coups occurring in the top ten
countries with the most attempts since 1950. This suggests that the more coups
occur in a country, the more likely additional coups are to happen in the future.

Conversely, autocoup leaders, often benefiting from a veneer of legitimacy and a
stronger initial position, are better able to consolidate power and attract
supporters. This advantage can be self-reinforcing, as a strong power base
discourages challenges and fosters loyalty. This dynamic is evident in cases like
China (2018), where the National People's Congress granted Xi Jinping the
potential to rule for life[^4], and Russia (2020), where constitutional changes
allow Putin to potentially remain in power until 2036[^5].

These factors contribute to a shorter expected tenure for coup-entry leaders
compared to the relatively longer tenures of autocoup leaders. The average
survival period following an autocoup is approximately five years longer than
that of coup-entry leaders (@fig-logrank). Based on these observations, I propose
the following hypothesis:

> ***H1: Political leaders who successfully extend their tenure through autocoups
> are more likely to survive longer compared to coup-entry leaders.***

### Research Design

#### Methodology: Survival analysis

To test the hypothesis, I will employ two Cox models to analyze the survival
tenures of coup-entry leaders and autocoup leaders. Unlike the Kaplan-Meier
model, the Cox model allows for the estimation of the impacts of multiple
factors. Although it does not directly estimate the duration of tenure in office,
it evaluates the hazard rate associated with being ousted from power.
Essentially, this represents different facets of the same phenomenon: as a
leader's cumulative hazard of being ousted increases, their probability of
survival in office decreases.

The first model will utilize the Cox proportional hazards model (Cox PH model),
using only the variables present at the entry year, without considering changes
in these variables over the leaders' survival times.

However, apart from the primary variable of interest in this research—the leader
type—control variables such as economic performance, Polity5 scores, and
political stability do change over time. Therefore, the second model will account
for these variations by using the time-dependent Cox model.

#### Dependent variables

-   **Survival Time:** The duration of a leader's tenure, measured in days. For
    coup-entry leaders, the survival time begins on the day they assume power
    through a coup. For autocoup leaders, the survival time starts on the
    expiration date of their original legitimate term. For example, Xi Jinping
    assumed power in 2013 and removed term limits in 2018. His original
    legitimate tenure was set to end in 2023, so his survival time begins in
    2023, marking the start of his post-autocoup tenure. The survival time
    concludes on the day the leader exits office, applicable to both coup-entry
    and autocoup leaders.

-   **End point status:** This variable indicates the manner in which the
    leader's tenure concluded, categorized as follows:

    -   **0 = Censored:** This status is assigned to leaders who leave office
        through regular means other than being ousted. This includes leaders
        transferring power to their designated successors, leaving office as
        their terms expire, losing in general elections, voluntarily leaving
        office due to health issues, or dying of natural causes.

    -   **1 = Ousted:** This status is assigned to leaders who are forced to
        leave office. This includes leaders resigning under pressure, being
        ousted by coups or other forces, or being assassinated.

#### Key Independent variable: Leader type

This variable categorizes leaders into two distinct groups:

-   **Group A = Autocoup Leader**: Leaders who extend their tenure through
    autocoups.
-   **Group B = Coup-Entry Leader**: Leaders who assume power through coups.

This variable is the primary independent variable of interest, serving as the
basis for comparing the survival time between these two types of leaders.

The data for both dependent and independent variables are sourced from the
autocoup dataset introduced in this study, Archigos, and PLAD.

#### Control variables

-   **Economic Performance:** This variable is measured using two indicators:
    economic level and economic growth trend.

    -   **Economic Level:** Represented by GDP per capita. This measure provides
        an indication of the overall economic health and standard of living in a
        country.

    -   **Economic Growth Trend:** Assessed using the current-trend ratio,
        developed by @krishnarajan2019, which is consistent with @eq-eq6 in
        chapters 2 and 3.

The GDP per capita data, expressed in constant 2017 international dollars (PPP)
and measured in units of \$10,000, is sourced from the V-Dem dataset by
@fariss2022. To account for the economic impact of the previous year, this data
is lagged by one year.

-   **Political Stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    Major Episodes of Political Violence dataset by @marshall2005current. This
    index provides a comprehensive measure of the level of violence and conflict
    within a country, which can significantly impact leadership survival.

-   **Degree of Democracy:** The level of democracy is gauged using Polity 5
    scores at the entry year for each respective country. These scores range from
    -10 (fully autocratic) to +10 (fully democratic), capturing the extent of
    democratic versus autocratic governance. This dataset is sourced from the
    Center for Systemic Peace (CSP) and provides an essential measure of
    political regime type, which can influence the stability and survival of
    leaders in power.

-   **Population Size:** To account for its potential impact on leaders' tenures,
    the log of the population size is considered. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on leadership survival. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's Age:** The age of the leader at the entry year is included as an
    additional variable in the analysis, offering insights into potential
    correlations with leadership survival. Older leaders may have different
    experiences, networks, and health considerations that could influence their
    ability to maintain power. This data is sourced from the leaders dataset by
    Archigos and PLAD.

### Results and discussion

#### Model results

Using the `survival` package in R [@survival], I present the regression results
for both the Cox Proportional Hazards (Cox PH) model and the time-dependent Cox
model in @tbl-cox.

```{r}
#| label: model
#| echo: !expr format == "html"

Mydata1 <- survival_cox_ph_model |> 
   mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")))

model_cox1 <- Mydata1 |> 
  coxph(
    Surv(survival_days, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age,
    data = _
  ) 

tbl1 <- model_cox1|> 
   tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T,
                         hide_se = F, 
                         hide_ci = T,
                         thresholds = c(0.01, 0.05, 0.1),
                         ) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 

Mydata2 <- survival_cox_td_model |> 
  mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")),
         survt = (tstop - tstart) * 365)

model_cox2 <- Mydata2 |> 
  coxph(
    Surv(T1, T2, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age + cluster(id),
    data = _
  ) 
tbl2 <- model_cox2|>
  tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T,
                         hide_se = F, 
                         hide_ci = T,
                         thresholds = c(0.01, 0.05, 0.1),
                         ) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox models for survival time of different types of leaders
#| echo: !expr format == "html"

tbl_merge(list(tbl1, tbl2),
          tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**")) |> 
  as_gt()
```

Both the Cox PH model and the time-dependent Cox model analyses revealed a
statistically significant association between leadership type and the hazard of
removal from power. Since time-dependent Cox model use the control variables
which change over time, I interpret the main findings based on time-dependent
model.

Coup-entry leaders were found to have a hazard ratio of 2.23 in the
time-dependent model compared to autocoup leaders (reference group), assuming all
other variables in the model are held constant. This suggests that coup-entry
leaders face a significantly greater risk of removal from power compared to
autocoup leaders. At any given time during their tenure, coup-entry leaders are
2.23 times more likely to be ousted from power compared to autocoup leaders, all
else being equal in the model.

The control variables perform differently in the two models. Economic level (GDP
per capita) exhibits statistically significant effects in both models. In the
time-dependent model, the hazard ratio of 0.95 indicates that for each unit
increase in GDP per capita (measured in units of \$10,000), the hazard (or risk)
of being ousted at any given time is reduced by 5%, assuming all other variables
in the model are held constant.

GDP growth trend demonstrates a more substantial effect in reducing the risk of
coups. Specifically, a 1 percentage point higher economic growth trend is
associated with an 80% reduction in the risk of being ousted, although this
effect is only statistically significant at the 10% level. This suggests a
possible trend where positive economic performance might mitigate the risk of
removal from power, but the evidence is not robust enough to confirm this
conclusively.

Political stability, as measured by the violence index, shows that a 1-point
increase in the index correlates with an 11% higher risk of being ousted.
However, this effect is also only statistically significant at the 10% level,
indicating a weaker but potentially important relationship between increased
violence and the risk of removal from office.

#### Discussion

```{r}
#| label: fig-coxSurv
#| fig-cap: Survival curves for Cox Model
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

testdata1 <- Mydata1 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox1 |> 
  survfit(newdata = testdata1) |>
  ggsurvplot(data = testdata1,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))

testdata2 <- Mydata2 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox2 |> 
  survfit(newdata = testdata2) |>
  ggsurvplot(data = testdata2,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))
```

The survival curves depicted in @fig-coxSurv illustrate the survival rates for
leaders of both types. Both the Cox PH model and the time-dependent Cox model
produce similar plots. Notably, the survival curve for coup-entry leaders
exhibits a significantly lower trajectory compared to that of autocoup leaders.
The steeper drop at the early stage for coup-entry leaders indicates they are
more likely to be ousted shortly after assuming power. Additionally, the survival
curve for coup-entry leaders crosses the median survival line much earlier (about
3,000 days) than that of autocoup leaders (about 8,500 days). This disparity
suggests that autocoup leaders tend to remain in power for longer durations than
their coup-entry counterparts.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard ratios and 95% CIs for Leader Ousting
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw() 

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |> 
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw() 
```

@fig-coxHR displays the hazard ratios and corresponding 95% confidence intervals
for the variables incorporated in the Cox model. Both the Cox Proportional
Hazards (PH) model and the time-dependent model produce similar plots,
reinforcing the robustness of the findings. Key points to note include:

-   The closer the hazard ratio (represented by the dots) is to 1, the less
    impact the variable has on the risk of being ousted. A hazard ratio of 1
    indicates no effect.

-   The whiskers extending from the dots represent the 95% confidence intervals.
    If these whiskers cross the vertical blue line at 1, it indicates that the
    variable is not statistically significant at the 5% level.

-   The hazard ratio for coup-entry leaders is significantly greater than 1 and
    statistically significant at the 5% level. This indicates that coup-entry
    leaders face a substantially higher risk of being ousted compared to autocoup
    leaders.

-   Most other variables have hazard ratios close to 1, suggesting that a
    one-unit increase in these variables does not significantly affect the risk
    of being ousted.

-   Although the hazard ratio for GDP growth trend is considerably less than 1 in
    the time-dependent model, indicating a potential protective effect, it is not
    statistically significant at the 5% level. However, it is statistically
    significant at the 10% level, suggesting that better economic performance may
    help to consolidate the rule of the incumbents to some extent, albeit the
    evidence is not as strong.

#### Assessing the Proportional Hazards Assumption

Assessing the proportional hazards assumption is crucial for the validity of the
Cox model results. To evaluate this, we used the chi-square test based on
Schoenfeld residuals to determine whether the covariate effects remain constant
(proportional) over time. Although the Cox PH model violates the proportional
hazards assumption, our primary analysis relies on the time-dependent Cox model,
which does not show strong evidence of violating the proportional hazards
assumption for any covariate. The global p-value of 0.416 is much greater than
the 5% significance level, indicating that the proportional hazards assumption is
reasonably met for the time-dependent Cox model.

```{r}
#| label: robust_test
#| include: false
test <- cox.zph(model_cox2)
```

### Conclusion

This chapter examined the survival durations of political leaders who come to
power through irregular means, specifically coups and autocoups. I hypothesized
that the mode of accession significantly influences leader tenure. Employing
survival analysis techniques, including the Cox proportional hazards model and a
time-dependent Cox model, I found strong evidence that autocoup leaders generally
enjoy longer tenures than coup-entry leaders.

The findings revealed a significant difference in average tenure, with
post-autocoup leaders averaging approximately 11 years in power compared to 5.6
years for coup-entry leaders. The time-dependent Cox model further indicated that
coup-entry leaders are 2.23imes more likely to be ousted from power at any given
time compared to autocoup leaders, all else being equal.

These results highlight the importance of understanding the phenomenon of
autocoups, where leaders extend their rule by manipulating legal frameworks. Due
to the relative ease and potential benefits of autocoups, this method of power
retention might incentivize more leaders to employ it. Consequently, democratic
backsliding could become more prevalent as autocoups weaken democratic
institutions and constitutional norms, particularly in nascent democracies or
those transitioning from autocracy.

This study contributes to the field of leadership survival by demonstrating that
the mode of accession significantly impacts leader tenure, a factor previously
under-explored in the literature. By utilizing both Cox models, the research
offers robust analytical techniques for studying political leadership survival
and provides strong evidence of divergent tenure lengths between these two types
of irregular-entry leaders.

However, limitations exist. The study relies heavily on the autocoup dataset
collected and coded by the author. The concept and data itself are relatively
novel within academia. Future research should refine and establish wider
recognition for the term "autocoup," leading to more accurate and comprehensive
data collection efforts. Expanding the dataset to include more cases and
integrating it with data on other irregular leadership transitions could yield a
more holistic understanding of political survival in such contexts.

Overall, this chapter underscores the need for more nuanced approaches to
studying political tenure and the mechanisms of irregular power retention,
contributing valuable insights into the dynamics of political stability and the
risks associated with different forms of non-democratic leadership succession.

## Conclusion

### Main Findings

This study delves into the dynamics and implications of irregular power
transitions, focusing on coups and autocoups. The findings illuminate the complex
interplay between incumbents and challengers fighting for power.

Firstly, our analysis reveals that the expected success rate of a coup attempt
significantly influences its likelihood. This success rate is heavily influenced
by the balance of power between the incumbent regime and challengers, which is
largely determined by regime type. We find that military regimes, although with
more control over their own military forces, face a higher risk of coups compared
to dominant-party regimes.

Secondly, the study introduces a redefined concept: the autocoup. Defined as an
incumbent leader's refusal to relinquish power as mandated, this research
distinguishes autocoups from broader terms like self-coups. Based on this
definition, we present the first publicly available dataset of autocoup events
from 1945 to 2022, encompassing 110 attempts and 87 successful autocoups. Case
studies and empirical analyses demonstrate the dataset's utility for quantitative
research, providing a robust foundation for further analysis on autocoups.

Thirdly, employing survival analysis techniques, the study finds clear
differences in leader longevity between those who come to power through coups and
those who extend their rule through autocoups. The results indicate that
coup-installed leaders face a significantly higher risk of removal compared to
autocoup leaders who manipulate the system to extend their rule.

### Limitations and directions for future research

This study offers a novel framework for analysing irregular power transitions,
but some limitations require further exploration:

-   **Data refinement:** Defining and classifying autocoups is a new approach.
    Future research should validate this classification system through additional
    studies and expert evaluations.

-   **Data harmonization:** The current analysis faces challenges due to
    mismatched units (country-year vs. leader) between coup and autocoup
    datasets. Future efforts should explore data harmonization techniques for
    more robust comparisons.

-   **Democratic backsliding:** While this study establishes a connection between
    irregular power transitions and democratic backsliding, further empirical
    evidence is needed to solidify this link.

Several avenues exist for future research:

-   **Terminology and data collection:** Refining the "autocoup" concept and
    achieving wider recognition will facilitate more accurate and comprehensive
    data collection.

-   **Dataset expansion:** Expanding the autocoup dataset with more cases and
    integrating it with data on other irregular leadership transitions can
    provide a more holistic view of political survival after these events.

-   **Power dynamics and long-term impacts:** Utilizing this dataset, future
    studies can delve deeper into power dynamics at play and explore the
    long-term consequences of irregular transitions on political systems,
    particularly regarding democratic backsliding, breakdown, and personalization
    of power.

In conclusion, this study sheds light on the dynamics of irregular power
transitions, specifically focusing on coups and autocoups. By redefining
autocoups, classifying the dataset, analysing determinants, and comparing leader
longevity, we establish a framework for understanding irregular transitions and
leader survival. This work contributes to a deeper understanding of democratic
resilience and political stability. Future research can build upon this
foundation by conducting further empirical analyses based on the novel autocoup
dataset and continuing to refine the framework.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false

cat("\\newpage
## References")
```

[^1]: According to the Archigos dataset, "Removed by Military, without Foreign
    Support" and "Removed by Other Government Actors, without Foreign Support" in
    the variable exitcode are classified as coups.

[^2]: The definitions and concepts of power expansion and power extension can
    often be ambiguous. In this study, we define power expansion as an incumbent
    acquiring additional authority from other branches or apparatuses of the
    state. Conversely, power extension refers to an incumbent prolonging their
    tenure beyond the originally mandated term in office.

[^3]: https://thediplomat.com/2023/08/myanmar-junta-extends-state-of-emergency-for-fourth-time/:
    Myanmar\
    Junta Extends State of Emergency for Fourth Time. Accessed on `r Sys.Date()`.

[^4]: https://www.bbc.co.uk/news/world-asia-china-43361276: China’s Xi allowed to
    remain ‘president for life’ as\
    term limits removed. Accessed on `r Sys.Date()`.

[^5]: https://www.ucl.ac.uk/news/2020/jul/analysis-vladimir-putin-secures-constitutional-changes-allowing-him-rule-until-2036:
    Analysis: Vladimir Putin secures constitutional changes allowing him to rule
    until 2036. Accessed on `r Sys.Date()`.
