---
title: "Power Dynamics and Coup Attempts: A Selection Mechanism Analysis"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: "`r Sys.Date()`"
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: coups1.html
categories: [Article, Coups]
bibliography: references.bib

abstract: |
  Despite extensive research identifying around one hundred potential determinants of coup attempts, no consensus has been reached. This study introduces a novel approach that prioritizes determinants based on their impact on coup success. By analyzing coup success rates, the study hypothesizes that the expected outcomes of coups are critical determinants of their occurrence. Utilizing a double probit model with sample selection, the research investigates the relationship between regime types and coup attempts. The findings suggest that regime type, by shaping internal power dynamics, is a crucial determinant of coup likelihood.

keywords: [Coup attempts, Coup outcomes, Regime types, Sample selection]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
```

```{r}
#| label: tbl-coup_year_counts
#| tbl-cap: "Coups by year and by country"
#| tbl-subcap:
#|   - "by year frequency"
#|   - "by country frequency"
#| layout-ncol: 2
#| echo: !expr format == "html"
#| include: !expr format == "html"
#load coup data
coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

attempts <- length(coup$country)
succeeded <- length(coup$country[coup$coup == 2])

coup |> 
  summarise(counts = n(),
            .by = year) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))

coup |> 
  summarise(counts = n(),
            .by = country) |> 
  arrange(desc(counts)) |> 
  head(15) |>
  clean_names(case = "title") |> 
  gt() |> 
  cols_align(
    align = "right",
    columns = 2
  ) |>
  tab_options(
    table.width = pct(95),
    column_labels.font.weight = "bold") |>
  tab_source_note(source_note = md("***Source: GIC dataset***"))
```

## Introduction

Coups occur with varying frequency across different countries, with some
experiencing them more frequently than others. According to the Global Instances
of Coups (GIC)[^1] dataset [@powell2011], Latin American countries such as
Bolivia witnessed `r coups |> filter(country == "Bolivia") |> pull(2)` coups
between $1950$ and $1984$, while Argentina experienced
`r coups |> filter(country == "Argentina") |> pull(2)` during a similar time
frame. However, Mexico's authoritarian period from $1917$ to $2000$ saw no coups
at all. In Africa, Sudan endured
`r coups |> filter(country == "Sudan") |> pull(2)` coups between $1955$ and
$2023$, whereas South Africa has not experienced any coup since $1950$. Similar
patterns are observed in the Middle East and South Asia.

The varying frequency of coup attempts has captivated political scientists for
decades, leading to extensive research on the subject. As highlighted by
@gassebner2016, despite approximately one hundred potential determinants of coups
being suggested, no consensus has been reached. In an effort to address this
issue, @gassebner2016 tested 66 factors proposed in previous literature using
three million model permutations in an extreme bounds analysis.

Examining previous research, which has tested around 100 variables as potential
determinants of coups, raises an important question beyond simply understanding
why coups are more frequent in some countries than others. The critical question
is: Can we establish a method to help scholars focus on the most relevant factors
of coups, rather than sifting through over 100 variables without reaching a
consensus?

Reviewing previously proposed variables of coups, it is evident that all focus on
pre-coup conditions, with no consideration given to post-coup factors. However,
coups are high-stakes gambles with an all-or-nothing nature. As defined by
@powell2011, coups are "illegal and overt attempts by the military or other
elites within the state apparatus to unseat the sitting executive" [@powell2011,
P.252]. Due to their illegality, the consequences of a failed coup can be severe,
with perpetrators risking imprisonment, exile, or even death. In some instances,
repercussions extend to the families of the coup perpetrators. Therefore, no coup
plotters would stage a coup without some assurance of success.

Historical coup attempts and their success rates provide valuable insights.
Despite the significant risks associated with coups since 1950, as shown in
@tbl-coups, there have been `r attempts` coups worldwide. Importantly, about half
of these coups have been successful. At first glance, coups appear to be a
high-success-rate political venture. However, compared to over 12,000
country-years since 1950, the occurrence of `r attempts` coups is relatively
rare, accounting for about 4% [@powell2011].

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts
#| echo: !expr format == "html"

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |> 
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`, 
                                accuracy = 0.1
                                )) |>   
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold"
              ) |> 
  # cols_width(1 ~ px(120),
  #            2:4 ~ px(88)
  # ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |> 
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

The low occurrence rate and high success rate clearly indicate that the
initiation of coups is highly selective. In other words, the likelihood of a coup
occurring depends greatly on its potential success rate. Since coup plotters
meticulously assess potential outcomes, we should also analyze what factors most
affect these outcomes when discussing the key determinants of coups. This
approach allows us to focus on the most relevant factors and disregard those less
related.

When considering the factors that most affect the outcomes of coups, the current
literature predominantly identifies military power as the decisive factor in the
success of coups. This necessitates an analysis of power dynamics within regimes,
as military power is ultimately shaped by power dynamics.

Because coup attempts are self-selective rather than random, this study employs a
double `probit` model with sample selection to examine factors influencing coup
success rates and, consequently, the likelihood of coup attempts. I posit that
regime type, by shaping internal power dynamics among coup plotters, incumbents,
and other ruling elites, is a crucial determinant of coup likelihood.

This study makes two key contributions to the existing literature. First, it
underscores the importance of regime type as a crucial determinant of coup
attempts. Previous studies often treat regime type as a control variable,
overlooking that variations in many other variables are fundamentally rooted in
different regime types. More importantly, this study establishes a systematic
approach for identifying the most relevant factors, thereby avoiding sifting
through over 100 variables.

The subsequent sections of this paper explore the dynamics of coup attempts and
their outcomes. In Part 3, I detail the research design, outlining the
methodology and variables used in the analysis. Part 4 presents and discusses the
empirical findings. Finally, Part 5 concludes the study, summarizing the key
insights and their implications.

## Dynamics of coup attempts and outcomes

Coup attempts are driven by a complex interplay of factors, with two key elements
attracting significant scholarly attention: **disposition** (the motivations
behind the attempt) and **capability** (the resources and opportunities to
succeed).

### Motivations for coups

This section focuses on disposition, exploring the primary motivations that
compel individuals to undertake the significant risks associated with a coup. We
can categorize coup motivations into three main types:

**Personal Ambition:** The allure of absolute power, prestige, and wealth is a
significant motivator for some coup plotters. For example, @wintrobe2019
distinguishes between totalitarian and tinpot dictators based on their use of
power. While both prioritize personal gain, totalitarian leaders seek complete
control over every aspect of society, whereas tinpot leaders focus on enriching
themselves through extravagant lifestyles.

**Purported National Interest:** Coups are sometimes justified as necessary
interventions to save a nation from crisis, uphold the constitution, or
facilitate a transition to democracy. While scepticism is warranted due to the
potential for self-serving justifications, legitimate cases do exist. For
instance, the 2010 coup in Niger ousted President Tandja, who attempted an
unconstitutional third term by dissolving the opposing court and calling a
self-serving referendum [@ginsburg2019].

**Self-Preservation:** In some instances, coups are pre-emptive strikes against
imminent political persecution or repression. Coup leaders might not be motivated
by a desire for power, but rather a fear of elimination by the incumbent regime.
A notable example is Idi Amin's 1971 coup against Ugandan President Obote, who
was attempting to remove Amin from his military command position [@sudduth2017].

These motivations can arise in any regime, but autocracies are particularly
susceptible, especially for coups framed under the guise of national interest or
self-preservation. Stable democracies, on the other hand, rarely face the same
level of constitutional crises or political persecution that might necessitate a
coup. However, new democracies can be vulnerable to instability, economic
downturns, and democratic backsliding, creating opportunities for coup plotters
to exploit these weaknesses and justify their actions.

Despite the potential motivations outlined above, coups remain relatively
uncommon events, occurring in only about 4% of country-years since 1950. This low
frequency highlights the importance of the second key element – capability. Even
the most motivated plotters require the resources and opportunities to succeed.
No rational actor attempts a guaranteed failure; the next section will explore
the concept of capability in greater detail.

### Capability for coups

While many ambitious individuals may covet supreme power, only a select few
possess the capability to orchestrate a successful coup. This capability hinges
not just on their desire, but on overcoming inherent disadvantages compared to
the incumbent leaders.

Firstly, coups are inherently clandestine operations due to their illegality.
Plotters require a tight-knit group to minimize leaks and maximize the element of
surprise. This secrecy restricts their ability to openly recruit supporters, a
privilege enjoyed by incumbents who can implement "coup-proofing" measures.

Secondly, coup plotters face uncertainty about the reactions of other powerful
factions within the regime, those who could tip the scales of power. Incumbents,
however, have a deeper understanding of these dynamics and proactively work to
solidify their own position. While they may not know who exactly might attempt a
coup, they are attuned to potential threats and adapt their strategies
accordingly.

Thirdly, coup plotters face a significant challenge in securing unwavering
loyalty from potential co-conspirators. The risks associated with a coup are
substantial, with uncertain rewards even in the event of success. Promises made
by coup leaders might not be kept, and post-coup purges are a common tactic to
eliminate future coup threats. Defecting to the incumbent leader can often be a
safer option, offering predictable rewards and less risk.

Given these inherent obstacles, rational coup plotters are unlikely to gamble on
a low-probability attempt. They may choose to abandon their plans altogether or
bide their time for a more opportune timing. Therefore, when coup plotters do
take action, it is because they have meticulously assessed their chances of
success and believe the risks are outweighed by the potential gains.

But what is the threshold for a "good enough" chance of success? Before diving
into a theoretical framework, let's examine historical data to gain some
perspective. Surprisingly, coups since 1950 boast a rather high success rate,
with nearly half ending in victory (as shown in @tbl-coups).

### Framework of coup success

An oft-cited framework [@gassebner2016; @aidt2019] provides a structured approach
to assess the disposition and capability of coup attempts by evaluating the
anticipated benefits for coup plotters. The expected payoff of coups can be
represented by the equation:

$$
\begin{aligned}
E(U) = p \times B + (1 - p) \times (-C)
\end{aligned}
$$ {#eq-eq1}

Here, $\mathbf B$ represents the return of a successful coup, $\mathbf C$
signifies the cost of a failed coup, and $p$ represents the probability of coup
success. The condition for staging a coup is when the expected benefit is
positive, meaning that the expected pay-off is greater than 0. Rearranging the
equation, we get:

$$
\begin{aligned}
p \times B > (1 - p) \times C
\end{aligned}
$$ {#eq-eq2}

@eq-eq2 implies that for @eq-eq1 to hold, the expected benefits earned from
successful coups must outweigh the expected cost of failed coups.

While seemingly clear, the equation faces practical challenges. Quantifying
$\mathbf B$ (the value of a successful coup) and $\mathbf C$ (the cost of
failure) is difficult. The loss of life, freedom, or loved ones after a failed
coup, as well as the value of assuming leadership after a successful coup, are
intangible concepts that defy precise measurement. As evidenced by the 1979 coup
in Ghana[^2], the fate of the coup leader(s) hangs in the balance; they are high
likely to be killed if the coup fails, or to execute others if the coup succeeds.

However, these challenges do not render the framework useless. Firstly, its core
logic remains valuable, offering insights into how coup plotters might assess the
return and cost of their actions. Secondly, given the significant and elusive
nature of precise values for $\mathbf B$ and $\mathbf C$, they can be treated as
roughly equal. Consequently, there is no need to fret over how to measure and
compare these values precisely. Instead, we can shift our focus from $\mathbf B$
and $\mathbf C$, to the probability of success ($p$), simplifying @eq-eq2 to:

$$
\begin{aligned}
p > (1-p)
\end{aligned}
$$ {#eq-eq3}

@eq-eq3 suggests that, to hold @eq-eq2 true, a success probability greater than
$50\%$ is necessary. Interestingly, empirical data on coups since 1950 somewhat
supports this notion. As shown in @tbl-coups, the overall success rate is
$49.9\%$. While this falls short of the $50\%$ threshold, it's important to
consider two factors. Firstly, this is an average rate, not necessarily
reflective of the probabilities assessed by coup plotters beforehand. Secondly,
outliers such as irrational actors and coups driven by self-preservation may not
prioritize success probabilities. Taking these points into account, we can
propose our first hypothesis:

> ***H1: The fundamental determinant of a coup attempt is the perceived chance of
> success. Coup plotters likely require a success threshold of at least 50%.***

This leads us to the next crucial question: what factors determine a coup's
success, influencing the very decision to attempt one? While specifics may vary,
the core element hinges on the power dynamic between coup plotters and the
incumbent leaders. Logically, the more powerful entity holds a greater advantage
in this high-stakes struggle for control.

### Regime types and power dynamics

When discussing the balance of power, the first aspect that springs to mind is
military prowess. It is evident that control over the military grants individuals
the upper hand in coup attempts. This elucidates why military coups often take
center stage in coup discussions. In much of the literature on coups, the terms
"coup" and "military coup" are frequently used interchangeably, as many scholars
argue that coups are orchestrated by military personnel through the use or threat
of armed forces. @powell2011 summarise 14 studies on coups, with half of them
attributing coup attempts solely to 'the armed forces'. Consequently, most
attention, both from academic researchers and political leaders, is focused on
the balance of power between civilian and military authorities, or among military
factions. Strategies such as "Keeping the military content" [@aidt2019, P. 15] or
"providing them with resources" [@huntington1991democratization, P. 252] have
been proposed to mitigate military intervention. Many coup-proofing strategies,
informed by empirical research, aim to either decrease the military's inclination
to stage coups or erect barriers to their success. For instance, studies by
@leon2013a suggest that nations with lower military spending as a percentage of
GDP are more vulnerable to coups. Similarly, @powell2018 argue that increased
military expenditures may diminish the likelihood of coups among military
factions.

However, previous studies have often oversimplified the intricate balance of
power within military forces. They have frequently assumed that the military
operates as a monolithic entity, uniformly and decisively. Such a notion fails to
capture the complexities of real-world dynamics.

It is crucial to acknowledge that the military is not always a unified entity.
Regardless of its size, any military force is composed of various groups or
factions, each with its own chain of command. Within these factions, mutual
suspicion, competition, and vigilance are common, while moments of unity are
rare. Due to the clandestine nature of coups and the need for secrecy, coup
attempts are often orchestrated within small, tight-knit groups. Coup plotters
are uncertain about the stances and intentions of other factions beforehand, and
they are particularly worried that once the coup is initiated, other factions may
not only refuse to support it but also actively oppose, intervene in, or even
suppress it. As in the attempted coup in Niger in 2021, which occurred just two
days before the new presidential inauguration, the military unit that staged the
coup was swiftly thwarted by Niger's security forces, resulting in the failure of
the coup within an hour[^3]. Therefore, the success of a coup heavily depends on
the reactions of other military factions [@geddes1999].

More importantly, the relationship between a government and its military varies
significantly across different regime types. In democracies, civilian authority
reigns supreme. The military is considered a national institution, bound by the
constitution rather than individual leaders, whether military or civilian. For
instance, the United States Armed Forces serve under the President's command but
ultimately answer to the Constitution, ensuring they remain politically neutral.

In contrast, non-democracies present a more complex picture. The lines of
authority are often either poorly defined in written documents or blatantly
disregarded by those in actual power. Identifying the true leader of the military
hinges on the specific regime type. To explore this further, we will utilize the
foundational concepts outlined by by @geddes2014 (GWF), which categorise
autocratic regimes based on leadership origins and decision-making factors. These
regimes fall into three broad categories: military regimes, personalist regimes,
and dominant-party regimes. Democracies and monarchies will also be included in
our analysis for comparative purposes.

**Military Regimes**

As defined by GWF, military regimes are characterized by the dominance of a junta
– a group of military officers who control the regime's power structure,
including leadership selection and policy formulation. Examples include the
Brazilian regime (1964-1985), the Argentine regime (1976-1983), and the
Salvadoran regime (1948-1984) [@geddes1999]. Notably, political parties might
still exist within military regimes, but they typically act as mouthpieces for
the junta or align themselves with the military's directives to avoid being
categorized as dominant-party regimes. Additionally, while a junta may appoint a
senior officer as the head of state or utilize civilian administrators as
intermediaries, the authority of these intermediaries over other officers remains
limited, differentiating military regimes from personalist ones.

**Personalist Regimes**

In a typical personalist regime, absolute power rests with a single dictator.
This encompasses policy making, control over the military, and the authority to
appoint top officials and successors. Personalist regimes often emerge under
charismatic leaders, particularly founding figures of newly independent nations.
Regimes like Rafael Trujillo's in the Dominican Republic (1930-1961), Idi Amin's
in Uganda (1971-1979), and Jean-Bédel Bokassa's in the Central African Republic
(1966-1979) exemplify personalist rule [@geddes1999]. In such regimes, political
parties are either absent or completely subservient to the dictator. Importantly,
the dictator may or may not have a military background, but the military itself
falls under their control.

**Dominant-Party Regimes**

In a dominant-party regime, supreme power resides within the ruling party, with
the leader acting as its representative and subject to its collective leadership.
Examples include the Partido Revolucionario Institucional (PRI) in Mexico, the
Revolutionary Party of Tanzania (CCM), and Leninist parties in various Eastern
European countries [@geddes1999]. The dominant party is a well-organized and
disciplined entity with its own ideology and political agenda used to unite and
mobilize its members and supporters. While powerful leaders may emerge within
such regimes, like Stalin in the Soviet Union (1924-1953) and Mao Zedong in China
(1949-1976), they lack the absolute power to dismantle or replace the party
altogether.

According to @geddes1999, different factions within a regime – military officers,
party cadres in dominant-party systems, and members of cliques in personalist
regimes – have distinct interests. For instance, professional soldiers prioritize
military survival and efficacy, party cadres seek to hold office, and members of
cliques depend on supporting the incumbent leader for their own survival and
relationships. While their strategies may differ, all ruling groups ultimately
prioritize self-preservation and advancement.

The key distinction between military, dominant-party, and personalist regimes
lies in the unique balance of power established by each. This power dynamic
emerges during the power seizure process. The most competent group, be it a
military junta, a political party, or a strongman, typically rises to power due
to the demanding nature of seizing control. @sudduth2017 and @roessler2011
analyse the purging of co-conspirators after power seizures, arguing that these
purges are more likely in the early stages of new regimes.

Following internal purges and external challenges, a new power dynamic emerges
among the factions within a regime, typically solidifying into one of three main
types: military regimes, personalist regimes, or dominant-party regimes.

Dominant-party regimes are the most stable due to their institutionalized
structure. A dominant party is a highly organized group with shared political
beliefs, goals, and ideologies. This shared ideology fosters internal cohesion
and a long-term vision, resulting in a robust structure with no single individual
wielding absolute power. The military force aligns with the party itself rather
than individual leaders, contributing significantly to greater regime stability
compared to personalist or military regimes. Dominant-party regimes also manage
leadership succession more effectively, as they typically regulate the process
and enforce term limits. As demonstrated by @frantz2017, 97% of country-years in
dominant-party regimes have formalized institutional succession rules, compared
to 77% in personalist regimes and a mere 59% in military regimes.

Personalist regimes also exhibit a degree of stability, as dictators usually
emerge from intense internal and external competition, with the most competent,
tough, and powerful individuals ultimately prevailing. Potential challengers have
often been purged, resulting in a relatively stable status quo within the
dictator's close circle. However, personalist regimes are vulnerable due to the
sudden death of incumbents. Rarely is there a universally accepted successor
during the chaos of power transitions, as incumbent dictators often purge
potential rivals. This uncertainty can trigger coups. Thus, while personalist
regimes maintain relative stability as long as succession is not an immediate
issue.

In contrast, military regimes are typically the least stable. These regimes often
rule through a junta, where power is shared equally among members. Mistrust and
suspicion are common, leading to frequent conflicts over benefits and policies.
Without a single authority figure, resolving these conflicts is challenging, as
no member is willing to concede, and no senior authority can enforce resolutions.
For example, General Augusto Pinochet was chosen to lead the junta in Chile after
the 1973 coup because his colleagues saw him as a safe choice due to his
professionalism, respect for rules, and uncharismatic demeanor
[@arriagadaherrera1988]. Although Pinochet later surprised them, this example
illustrates that juntas are generally unstable leadership groups.

```{r}
#| label: tbl-regimes1
#| tbl-cap: Main features of different types of regimes 
#| echo: !expr format == "html"

read.csv("regimes_comparison.csv") |> 
  mutate(institutionalized_leadership_succession = percent(institutionalized_leadership_succession)) |> 
  clean_names(case = "title") |> 
  gt() |> 
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    4 ~ px(180),
    3 ~ px(80),
    1:2 ~ px(30)
  ) |> 
  tab_source_note(source_note = md("*Source: GWF & Author*"))
```

These differing power dynamics significantly impact the likelihood of coup
attempts. Military regimes, in particular, have an inherent predisposition
towards coups, as junta members typically command their own military factions,
eliminating the need to organize an army for a coup. According to @tbl-regimes,
military regimes face the highest risk, representing only 5.6% of country-years
yet accounting for over 22% of coups. Personalist regimes follow, with 23% of
coups despite comprising only 13% of country-years. Dominant-party regimes, with
their stronger institutions and unified leadership, have the lowest incidence of
coups, constituting 22.6% of country-years but responsible for only 16.7% of
coups.

\cleardoublepage

```{r}
#| label: tbl-regimes
#| tbl-cap: Regime types and coups since 1950 
#| echo: !expr format == "html"

regimes <- REIGN |>
  distinct(reign_cowcode, year, .keep_all = T) |>
  mutate(
    regime_type = case_when(
      gwf_regimetype %in% c(
        "party-based",
        "party-personal",
        "party-military",
        "party-personal-military",
        "oligarchy"
      ) ~ "dominant-party",
      gwf_regimetype == "personal" ~ "personal",
      gwf_regimetype %in% c("military", "military personal", "indirect military") ~ "military",
      gwf_regimetype == "monarchy" ~ "monarchy",
      gwf_regimetype %in% c("presidential", "parliamentary") ~ "democracy",
      TRUE ~ "other"
    )
  ) |>
  filter(year > 1949)

regime_total <- regimes |>
  tabyl(regime_type) |> 
  arrange(-n) |> 
  adorn_totals("row") |> 
  mutate(country_year =n,
         share = percent(percent),
         .keep = "unused") 
  

coup1 <- coup |>
  select(country, ccode, year, coup)

coup_by_regime <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup)) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  mutate(num_of_coups = n,
         percent_of_coups = percent(percent,accuracy = 0.1),
         .keep = "unused") 

coup_by_succeed <- regimes |>
  select(ccode = reign_cowcode, extended_country_name, year, regime_type) |>
  full_join(coup1, by = join_by(ccode, year)) |>
  mutate(regime = na.locf(regime_type, .by = ccode)) |>
  filter(!is.na(coup),
         coup == 2) |>
  tabyl(regime) |> 
  adorn_totals("row") |> 
  select(regime, succeeded_coup = n) 

coup_regimes <-  coup_by_regime |> 
  left_join(coup_by_succeed) |> 
  mutate(success_rate = percent(succeeded_coup / num_of_coups, accuracy = 0.1))

regime_total |>
  left_join(coup_regimes, join_by(regime_type == regime)) |>
  mutate(regime_type = str_to_title(regime_type)) |>
  select(-succeeded_coup) |> 
  clean_names(case = "title") |>
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  tab_source_note(source_note = md("*Source: REIGN and GIC Datasets*"))
```

> ***H2: Due to their balance of power dynamics, military regimes are more prone
> to coups, followed by personalist regimes, while dominant-party regimes are the
> least likely to experience coups among the three.***

## Research Design

### Double probit with sample selection model

To account for the determinants of coup attempts, this study utilizes a sample
selection model [@heckman1979]. While coup attempt rates vary across regimes,
success rates tend to converge around 50% (@tbl-regimes). This convergence cannot
be attributed to chance and suggests that coup attempts are strategically planned
and self-selected, targeting situations with a perceived high chance of success.
Consequently, a standard regression model would yield biased results. Therefore,
we use a double `probit` model with sample selection, as employed by @powell2012,
for empirical analysis. The first stage (selection equation) examines the
probability that a regime experiences a coup attempt, while the second stage
(outcome equation) evaluates the probability of the success of those coups.

The primary explanatory variables are regime types, as previously discussed.
Control variables are included in $\mathbf{XB}$. The selection equation (first
stage) models the probability that a coup attempt occurs and can be expressed as
follows:

$$
\begin{aligned}
y_1^*=\alpha_0 + \alpha_1 Regime_i + \mathbf {XA} + \mu_{1i}
\end{aligned}
$$ {#eq-eq4}

Here, ${y_1}^*$ is an unobserved variable, which may be known to coup plotters.
$Regime_i$ is a categorical variable (*military*, *personalist*, or
*dominant-party*). ***XB*** captures other control variables, such as the
economic crisis index, previous coups, military expenditure, etc.

The observed binary outcome $\mathbf{y_1}$ is:

$$
\begin{aligned}
y_1 = 
\begin{cases} 
1 &\text{if $y_1^*>0$ (coup attempt occurs)} \\
0 &\text{if $y_1^*\le0$ (no coup attempt)}
\end{cases}
\end{aligned}
$$\
In the first stage, if $y_1^*\le0$, no coup attempt occurs in a given
country-year, indicating that the unobserved variable does not reach the
threshold. If $y_1^*>0$, at least one coup attempt is made in a country-year,
indicating that the unobserved variable surpasses the threshold. The probability
is expressed as:

$$
\begin{aligned}
Prob(y_1 =1)&=Prob(y_{1}^*>0) \\
&=\Phi(\alpha_0 + \alpha_1 Regime_i + \mathbf{XA})
\end{aligned}
$$ {#eq-eq4a}

Similarly, the outcome equation (second stage) models the probability that a coup
attempt is successful, given that it occurs:

$$
\begin{aligned}
y_2^*=\beta_0 + \beta_1 Regime_i + \mathbf {XB} + \mu_{2i}
\end{aligned}
$$ {#eq-eq5}

The observed outcome $y_2$ is:

$$
\begin{aligned}
y_2 = 
\begin{cases} 
1 &\text{if $y_2^*>0$ (coup succeeds)} \\
0 &\text{if $y_2^*\le0$ (coup fails)}
\end{cases}
\end{aligned}
$$

The probability equations is:

$$
\begin{aligned}
Prob(y_2=1|y_1 =1)=\Phi(\beta_0 + \beta_1 Regime_i + \mathbf{XB})
\end{aligned}
$$ {#eq-eq5a}

### Variables

#### Dependent variable

To test our hypotheses, we use coup attempts and outcomes data from @powell2011
as the dependent variable. Successful coups are those in which the incumbent is
removed from office for more than seven days. The dataset spans from 1950 to
2023, accounting for 491 coup attempts, of which 245 were successful. Descriptive
statistics are presented in @tbl-coups and @tbl-regimes.

#### Independent variables

The primary independent variable in our analysis is the regime type. We follow
the categories defined by @geddes2014 (GWF). Our main focus is on military,
personalist, and dominant-party regimes, with democracy and monarchy included as
reference categories. Descriptive statistics are presented in @tbl-regimes.

#### Control variables

In selecting control variables, we draw on the research by @gassebner2016, who
test the robustness of 66 factors proposed in the empirical literature using a
monthly sample of 164 countries from 1952 to 2011. They find that slow economic
growth rates, previous coup experiences, and other forms of political violence
are particularly conducive to inciting coups. Therefore, our main control
variables are economic performance, political violence, and the number of
previous coups.

**Economic performance:** We use the current-trend ($CT$) ratio developed by
@krishnarajan2019 to measure economic performance. This metric calculates the
ratio between the current level of gross domestic product per capita (GDP/cap)
and the average GDP/cap level of the previous five years. For a country $i$ at
year $t$, the $CT$ ratio is calculated as follows: $$
\begin{aligned}
CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
\end{aligned}
$$

Since a given year's situation is primarily determined by the previous year, the
variable is lagged by one year. The GDP per capita (in constant 2017
international dollars, PPP) is sourced from the V-Dem dataset provided by
@fariss2022.

**Violence:** The violence index captures all types of violence and wars
(internal or interstate), indicating the stability of the regime. This data is
sourced from the Major Episodes of Political Violence [@marshall2005current].

**Previous coups:** This variable captures the number of previous coups in each
country. In the double probit with sample selection model, the second-stage model
cannot be identical to the first-stage model. Therefore, the first-stage model
includes the number of previous coups, while the second-stage does not. Previous
coups may have a greater impact on coup attempts, while once a coup is launched,
the number of previous coups might not significantly explain the outcome.

## Results and Discussion

### Interpretation and Discussion

```{r}
#| results: asis
#| echo: false

library(readxl)
political_violence <- read_excel("MEPVv2018.xls") |> 
  dplyr::select(ccode, year, violence = actotal)

gdp <- vdem |> 
  dplyr::select(ccode = COWcode, year, gdppc = e_gdppc) |> 
  filter(year > 1944) |> 
  mutate(ct = gdppc / lag(rollapply(gdppc, width = 5, FUN = mean,
                               align = 'right', fill = NA)),
         .by = c(ccode)) 

coup_all <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_ccode_year.txt", sep = "\t") |> 
  mutate(coup = case_when(
    coup1 == 0 ~ 0,
    coup1 == 2 | coup2 ==2 | coup3 == 2 | coup4 == 2 ~ 2,
    TRUE ~ 1
  )) |>
  mutate(pre_coups = cumsum(lag(coup)  %in% c(1,2))) |> 
  dplyr::select(ccode, year, coup, pre_coups)

reg_data <- regimes |> 
  dplyr::select(country = extended_country_name,
         ccode = reign_cowcode, year, regime_type) |>
  left_join(coup_all, by = join_by(ccode, year)) |> 
  left_join(political_violence, by = join_by(ccode, year)) |>
  left_join(gdp, by = join_by(ccode, year)) |> 
  filter(!is.na(gdppc)) |> 
  mutate(violence = na.locf(violence, .by = ccode)) 

model_2step <- reg_data |>
  filter(regime_type != "other") |>
  mutate(regime = relevel(factor(regime_type), "dominant-party")) |>
  selection(coup > 0 ~ regime + ct + gdppc + violence + pre_coups,
            coup > 1 ~ regime + ct + gdppc + violence,
            data = _) 

model_selection <- model_2step
model_selection$param$index$betaO <- model_2step$param$index$betaS
model_selection$param$index$betaS <- model_2step$param$index$betaO

stargazer(model_selection, model_2step,
          header = F,
          column.labels = c("Coup Attempts", "Coup Outcome"),
          intercept.bottom = F,
          title = "Sample Selection Model of Regime Types and Coups, 1950-2019",
          label = "results",
          covariate.labels = c("Constant", "Regime: Democracy", "\\hspace{1.6cm}Military", "\\hspace{1.6cm}Monarchy", "\\hspace{1.6cm}Personalist", "Economic trend", " GDP per capita", "Political violence",  "Previous coups"),
          dep.var.caption = "",
          dep.var.labels.include = F,
          summary.logical = T,
          column.sep.width = "50pt"
          )
```

Using the R package `sampleSelection` [@sampleSelection], the findings from the
double probit model, as presented in Table \ref{results}, provide valuable
insights into the dynamics of coup attempts and their likelihood of success
across different regime types:

As hypothesized, both the coefficients for military and personalist regimes are
positive and significant at the 1% level, indicating a higher likelihood of
experiencing coup attempts compared to dominant-party regimes, holding other
factors constant. This higher propensity for coup attempts in military and
personalist regimes aligns with our hypothesis that these regimes face distinct
internal power dynamics, increasing the likelihood of coup attempts.

All control variables are statistically significant at the 1% level. As expected,
better economic performance and higher levels of GDP per capita help reduce the
risk of coups. Surprisingly, more previous coups appear to reduce the risk of
coups, but this finding is not substantially significant. This indicates that
even after accounting for various relevant factors, regime type remains a
critical determinant of coup likelihood. This underscores the robustness of our
findings and the importance of regime type in understanding coup dynamics.

In the outcome equation, most variables are not significant at the 5% level. This
supports our hypothesis that coup attempts are highly selective at the attempt
stage, thereby diminishing the influence of these variables on the coup outcome.
Higher GDP per capita is negatively associated with the likelihood of coup
success at the 5% significance level. This suggests that better economic
conditions may bolster support for incumbent leaders, thereby reducing the
chances of a successful coup against them.

The correlation coefficient ($\rho$) between the error terms of the selection and
outcome equations is 0.7086 with a p-value of 0.111. Although positive, this
correlation is not statistically significant, indicating that there may not be a
strong correlation between the errors in the selection and outcome equations.
This suggests that the selection bias might not be as severe as initially
hypothesized.

Overall, the regression results indicate that the sample selection model is a
well-specified choice for this research. The significant coefficients and
theoretically consistent directions suggest that the model effectively captures
key aspects of coup dynamics. Specifically:

-   The direction and significance of the coefficients align with theoretical
    expectations. Regimes with less institutional stability, such as military and
    personalist regimes, are more likely to experience coup attempts.

-   Better economic conditions not only reduce the likelihood of coup attempts
    but also decrease the likelihood of coup success.

-   Coup attempts are not random events; they occur under specific conditions.
    Thus, treating the selection process (whether a coup attempt occurs)
    separately from the outcome (whether the coup is successful) effectively
    captures this non-randomness.

-   The double probit with samle selection model is designed to correct for
    selection bias. The similar success rates of coups across different regimes,
    despite varying attempt rates, indicate potential selection bias. This
    supports the need for a sample selection model, as coup plotters are likely
    to attempt coups only when they perceive a high likelihood of success.

In conclusion, the use of a sample selection model is appropriate for this
research, providing robust insights into the factors influencing coup attempts
and their outcomes across different regime types. The results highlight the
significant role of regime type in coup dynamics and the selective nature of coup
attempts, thereby validating the theoretical framework and empirical strategy
employed in this study.

## Conclusion

The findings of this study underscore the critical role of regime type in
determining the likelihood of coup attempts. Through the application of a double
probit with sample selection model, this research highlights that coups are
significantly more probable in military and personalist regimes compared to
dominant-party regimes. The results suggest that the power dynamics inherent in
different regime types influence the propensity for coups, with less
institutional stability and more fragmented power structures increasing the risk.

Moreover, the study reveals that better economic conditions, while reducing the
overall likelihood of coup attempts, paradoxically enhance the chances of coup
success. This implies that economic resources can provide the necessary support
for successful coups, even as they deter initial attempts.

The significant coefficients and the alignment of results with theoretical
expectations validate the use of the sample selection model in this context. By
accounting for the selection bias in coup attempts, the model offers robust
insights into the factors influencing both the occurrence and success of coups.
These findings contribute to a deeper understanding of coup dynamics, emphasizing
the importance of regime type and economic conditions in shaping political
stability.

In conclusion, this study provides a nuanced perspective on coup attempts and
their outcomes, reinforcing the need to consider regime-specific characteristics
and economic contexts in analyses of political instability. Future research could
further explore the interplay between these factors and examine additional
variables that may influence coup dynamics.

```{r}
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: <https://www.uky.edu/~clthyn2/coup_data/home.htm>, accessed on
    `r Sys.Date()`

[^2]: In the case of the Ghanaian coup, flight lieutenant Jerry John Rawlings
    narrowly avoided execution after his initial failure, being freed by mutinous
    soldiers. Three weeks later, following Rawlings' successful overthrow of the
    government, the deposed leader, General Fred Akuffo, was executed along with
    many other senior members of his government.

[^3]: Niger: Attack on presidential palace an 'attempted coup'. Source: Al
    Jazeera. Retrieved from
    <https://www.aljazeera.com/news/2021/3/31/heavy-gunfire-heard-near-nigers-presidency.>
    Accessed on `r Sys.Date()`.
