---
title: "Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-Entry and Autocoup Leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-10
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: survival1.html
categories: [Article, Coups]
bibliography: references.bib

abstract: |
  This study examines how the method of power acquisition impacts the longevity of two main types of irregular-entry political leaders: coup-entry leaders and autocoup leaders. We hypothesize that the method of accession significantly affects leader tenure. Utilizing the Cox proportional hazards model and a time-dependent Cox model, we provide compelling evidence of differing survival times between these two leader types. The findings indicate that coup-entry leaders face a significantly higher risk of removal compared to autocoup leaders. Specifically, coup-entry leaders are found to have a much shorter average tenure and a higher probability of being ousted. This study underscores the implications of these findings for political stability and democratic processes, suggesting that the relatively low cost and high returns of autocoups could incentivize more incumbents to seize power in this manner, potentially leading to democratic backsliding. Furthermore, this research contributes to the academic literature by effectively utilizing a newly developed dataset on autocoups, offering valuable insights into the dynamics of irregular leadership transitions.
  
keywords: [Coups, Autocoups, Survival, Political Leadership, Cox Model]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(survival)
library(survminer)
library(ggsurvfit)
library(gt)
library(gtsummary)
library(broom)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(stargazer)
library(here)
```

## Introduction

Why do some leaders rule for decades while others stay in power for only years,
months, or even days? This question has garnered substantial attention, making
the survival of political leaders a well-explored topic in political science.
Although the general framework of political survival has been extensively studied
and numerous theories have been proposed, specific types of leaders—particularly
coup-entry leaders and autocoup leaders (defined in more detail in Chapter
3)—have received comparatively less scrutiny. Nonetheless, examining the tenures
of these specific leader types is particularly significant, as it provides
insights into the dynamics of irregular leadership transitions and their impact
on political stability and democratic processes.

The processes of entry, tenure, and exit are more predictable for leaders who
come to power through regular channels, making their survival easier to
anticipate and thus less compelling to study. In contrast, leaders who ascend
through irregular means, such as coups or autocoups, attract more interest from
both academics and journalists due to the unpredictable and often tumultuous
nature of their tenures. According to @goemans2009, leaders with irregular
entries or exits are quite common. Between 1945 and 2015, more than half of the
leaders who entered power irregularly (158 out of 308) eventually exited
irregularly as well. This rate is significantly higher than that of leaders who
assumed office through regular channels, of whom only 14.5% (213 out of 1,472)
experienced irregular exits. While this percentage is still notable, it is
considerably lower than that for irregular-entry leaders.

Among leaders with irregular entry or exit, coup-entry and autocoup leaders
constitute the majority. According to @goemans2009, out of 374 leaders who exited
irregularly, 246 were ousted through coups, accounting for 65.8% of these cases.
@frantz2016 show that coup-related exits account for roughly one-third of all
exits in autocracies, surpassing any other type, including regular transitions.
Additionally, between 1945 and 2020, there were 106 autocoup attempts, with 86
being successful [@zhu2024].

```{r}
#| label: data
#| include: false
Mydata1 <- read.csv(here("data/mydata1.csv")) |> 
  mutate(survival_year = survival_time / 365)

avg <- Mydata1 |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)
```

Due to their irregularity and inherent uncertainty, precisely measuring the
survival tenure of coup-entry and autocoup leaders presents a significant
challenge. Nonetheless, a comparative analysis of the tenures of these two types
of leaders is both feasible and valuable. Leaders who consolidate power through
autocoups have a longer average tenure post-autocoup (approximately
`r avg$avg[1]` years) compared to coup-entry leaders (approximately
`r avg$avg[2]` years). This suggests a potential shortfall of about 5 years in
the average tenure of coup-entry leaders.

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of overstaying and coup-entry leaders
#| echo: !expr format == "html"

# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = Mydata1)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = Mydata1, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("Autocoup leaders", "Coup-entry leaders"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-entry leaders",
           font.title = 12)
```

A preliminary analysis using a log-rank test in survival analysis, as shown in
@fig-logrank, reveals a distinct contrast in the tenures of autocoup leaders
versus coup-entry leaders. The survival curve for autocoup leaders is
consistently higher than that for coup-entry leaders, and the difference is
statistically significant. This indicates that autocoup leaders generally have a
longer survival tenure and a lower risk of being ousted compared to coup-entry
leaders.

We argue that the method of accession significantly influences leadership
longevity. Coup-entry leaders are likely to face greater challenges to their
rule, resulting in a shorter average tenure compared to autocoup leaders.

Using the Cox proportional hazards model and the time-dependent Cox model, our
results indicate that autocoup leaders generally experience longer tenures
compared to coup-entry leaders.

This study makes two potential contributions. Firstly, it sheds light on an
understudied factor in leadership survival analysis: the impact of the method of
accession to power. Leaders' survival is influenced not only by their ruling
strategies after taking power but also by how they acquired power initially.
Secondly, by utilizing survival models, this research provides empirical evidence
of the significant difference in tenure duration between autocoup and coup-entry
leaders. This finding may help explain the increasing prevalence of overstaying
in power through autocoups since 2000. As more incumbents observe these
precedents, they may be encouraged to adopt similar tactics, with ruling elites
tending to follow and support such actions to benefit from a relatively longer
tenure.

The remainder of the paper delves deeper into this topic. Chapter 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Chapter 3 explores the factors influencing the survival of
coup and autocoup leaders. Chapter 4 outlines the methodology and data used,
including the application of survival models to analyze the determinants of
leadership longevity. Chapter 5 presents the findings of the analysis and a
detailed discussion of the results. Finally, Chapter 6 concludes by synthesizing
the key takeaways and exploring their broader implications.

## Literature review

The topic of political survival has been a cornerstone of political science
research for decades. This enduring interest arises from the wide-ranging
variations observed across regimes, countries, and historical periods. Previous
studies have identified two crucial yet interconnected aspects of political
survival: regime survival and individual leader survival.

Regime survival focuses on the longevity of political systems, such as
monarchies, political parties, or specific ideological structures. Leader
survival, on the other hand, is concerned with the duration of individual
leaders' time in office. These two concepts often exhibit contrasting patterns.
For example, in parliamentary democracies like Japan or the UK, specific
political parties may hold power for extended periods while individual leaders
(Prime Ministers) change frequently. Similarly, communist regimes typically see
long-lasting parties in power, with leadership transitions occurring at a more
frequent pace. In contrast, presidential systems like the United States or some
military regimes experience more frequent changes in both the ruling party or
junta and the leader of the country. This study specifically investigates the
dynamics of individual leader survival, focusing on the factors influencing how
long leaders remain in power.

The existing literature on leader survival is vast and multifaceted, offering a
wealth of insightful perspectives. Some studies delve into the specific
mechanisms that influence leadership longevity within particular regimes, such as
democracies [@svolik2014] or autocracies [@davenport2021]. Others seek to develop
more generalizable theoretical frameworks that can explain leader survival across
diverse political systems [@buenodemesquita2003]. While the development of a
universal theory remains an alluring goal, it is important to acknowledge the
inherent challenges in creating a single model that encompasses the complexities
of leadership survival across all regime types.

The mechanisms governing power transitions vary significantly across different
regimes, particularly between democracies and autocracies. In many autocratic
systems, leadership selection is a closed affair. Potential candidates are often
restricted to a narrow pool, such as royal families, military elites, or members
of the ruling party. While political competition and elections may exist in some
autocracies, there are often significant barriers to entry for legitimate
challengers. Potential rivals may face threats such as assassination,
imprisonment, or exile. Moreover, selection processes are shrouded in secrecy,
with outsiders unable to participate or obtain information on the selections, let
alone express dissent or complain about the results. This lack of inclusivity and
transparency makes it challenging to gauge true levels of public support compared
to democracies. Consequently, calculating selectorates or winning coalitions, as
explored by @buenodemesquita2003, becomes a near-impossible task in autocracies.

Therefore, focusing research on more specific regimes or types of leaders may be
more appropriate. While regular and anticipated leadership changes are important
as well, they offer less fertile ground for exploring the dynamics of leader
longevity, as the vast majority of leaders who assume power through established
channels also exit power through established mechanisms [@goemans2009]. In
contrast, the study of political survival among irregular leaders is particularly
captivating due to the intricacies and uncertainties associated with irregular
leadership transitions.

The diverse mechanisms influencing political leadership survival, as discussed
above, have prompted scholars to delve deeper into more specific determinants.
Two primary perspectives have emerged to explain the dynamics of leader survival.

The first perspective considers objective factors and resources available to
leaders. These include elements such as personal competence [@yu2016], societal
stability [@arriola2009], economic development [@palmer1999; @williams2011],
access to natural resources [@smith2004; @quirozflores2012; @wright2013], or
external support networks [@licht2009; @wright2008; @thyne2017].

The second perspective explores subjective factors, focusing on the strategies
leaders employ to consolidate their power. This encompasses both the formulation
and implementation of political policies and the leaders' responses to
opposition, challenges, or even coups and rebellions [@gandhi2007; @morrison2009;
@escribà-folch2013; @davenport2021].

Unsurprisingly, coups have garnered significant scholarly attention due to their
pivotal role in removing leaders [@svolik2009; @frantz2016]. Existing research
delves into strategies for thwarting coups [@powell2017; @sudduth2017;
@debruin2020], as well as how leaders extend their tenures after surviving coup
attempts [@easton2018]. For instance, @sudduth2017 examines the post-coup actions
of dictators, focusing on purge strategies. They argue that coup leaders
initially wield more power than elites within the regime, rendering them less
susceptible to subsequent coups, challenging the conventional view of new leaders
as inherently weak [@roessler2011]. Meanwhile, @sudduth2018 investigates how
leaders' entry methods affect their removal in dictatorships, positing that
irregular entry does not necessarily increase the likelihood of removal; in fact,
some forms of irregular entry may offer protection.

While scholars have extensively analysed leader survival across various contexts,
including universal frameworks, autocratic regimes, and the aftermath of failed
challenges, a significant gap persists. There is a lack of research specifically
exploring and comparing the survival tenures of leaders who extend their reigns
through autocoups compared to coup-entry leaders. This study aims to address this
gap by investigating and comparing the duration of leadership survival between
these two distinct leader types.

## Survival dynamics of autocoup and coup-entry leaders

### Autocoup leaders versus coup-entry leaders

As emphasized in Chapter 2, delving into leadership survival poses inherent
challenges, given factors such as the opacity and diverse mechanisms of power
transitions. However, these challenges underscore the significance of this
research, as it sheds light on understudied dynamics.

While the survival of political leaders manifests complexity and variation, it is
not devoid of patterns entirely. Leaders of similar types often exhibit
significant comparability. Before delving into the comparison, it is necessary to
clarify several relative terminologies.

Firstly, we define an autocoup as a situation where an incumbent leader utilizes
illegitimate or unconstitutional methods to extend their tenure and remain in
power. An autocoup is deemed successful if this power extension lasts for at
least six months. Coups, on the other hand, are defined as illegal and overt
attempts by the military or other powerful groups within the state to unseat the
sitting leader [@powell2011]. Similar to coups, successful autocoups in this
study will be measured by a power usurpation lasting at least six months,
differing from the seven-day duration coded by Powell and Thyne.

Secondly, it is also crucial to clarify the distinction between an autocoup
leader and a coup-entry leader, as the survival of leadership is the main concern
of this study.

-   **Autocoup leader:** This refers to an incumbent leader who successfully uses
    illegitimate or unconstitutional means to extend their tenure in power. In an
    autocoup, the leader orchestrates the power grab and continues to rule
    afterwards.

-   **Coup-entry leader:** This term designates the individual who assumes power
    after a successful coup. The coup leader and the coup-entry leader may or may
    not be the same person. Unlike in autocoups, coups often involve multiple
    leaders (individuals or groups) who overthrow the incumbent leader, but
    typically only one of them assumes supreme power. In some instances, coup
    leaders may support someone outside the coup plot to become the new leader.
    For example, military officers might return power to civilians after a coup
    or support a new general election. Regardless of the specific scenario, a
    coup-entry leader in this study refers to the individual who assumes formal
    leadership following a successful coup.

Given that autocoup leaders typically exhibit longer overall tenures compared to
coup-entry leaders, this study focuses on a more nuanced comparison.
Specifically, we will analyse the **post-autocoup** tenure of autocoup leaders
and contrast it with the **post-coup** tenure of coup-entry leaders. The
examination of the survival tenures of coup-entry leaders and autocoup leaders is
motivated by their relevance and similarity in terms of illegitimacy,
uncertainty, and instability.

### Hypothesis

Previous research emphasizes that skilful power retention is the ultimate
determinant of leader longevity. Leaders who can maintain control or manipulate
the balance of power tend to stay in office longer. However, as discussed
earlier, although both coup-entry and autocoup leaders face similar challenges,
the intensity of their challenges related to illegitimacy, uncertainty, and
instability differs. These discrepancies create an uneven playing field in terms
of power dynamics, with coup-entry leaders at a significant disadvantage. This
disparity in power equilibrium shapes a nuanced landscape that profoundly impacts
the duration of leader tenures.

#### Illegitimacy

Both leader types lack legitimacy, though it manifests differently. Coup leaders
seize power through force or the threat of force, making their illegitimacy
explicit. Autocoups often employ seemingly legal procedures, but these processes
are frequently manipulated by incumbents leveraging their control [@zhu2024].
This perception of illegitimacy can be used to justify the removal of autocoup
and coup-entry leaders, even if the means of removal are themselves illegitimate.

While both coup-entry and autocoup leaders lack genuine legitimacy, the nature of
this illegitimacy differs significantly. Coups represent a more blatant disregard
for legal processes, often involving force or the threat of force to seize power.
In contrast, autocoups employ a façade of legality through tactics such as
manipulating constitutional interpretations, engineering parliamentary votes,
influencing court decisions, and even holding referendums. Despite this veneer,
these manoeuvres do not erase the underlying lack of genuine legitimacy, hence
this study categorizes them as autocoups. However, there is considerable debate
among scholars, with some arguing for less severe terms like "incumbent overstay"
or "executive takeover" due to the perceived legality.

This perceived legitimacy can provide a temporary advantage for autocoup leaders.
Challengers are often constrained to operate within legal frameworks, making it
difficult to directly confront the incumbent.

#### Uncertainty

The tumultuous paths to power undertaken by coup-initiators and autocratic
leaders cast a shadow of uncertainty over their reigns and eventual departure.
Their ascension through irregular means undermines established power transition
norms, leaving doubts lingering over their commitment to constitutional
succession protocols. This uncertainty not only unsettles elites and citizens but
also plagues the leaders themselves, who grapple with the perpetual ambiguity
surrounding the transfer of power – when, how, and to whom. Historical analyses
underscore this predicament, with data revealing that more than two-thirds of
irregular exits from leadership stem from coup-related upheavals [@goemans2009].

Coup-entry and autocoup leaders face different levels of uncertainty immediately
following their rise to or overstay power.

After a coup, at least three major uncertainties arise regarding leadership and
its tenure. First, it is unclear who will assume leadership. Although coup
leaders often take power, some may return or promise to return power to civilian
leaders. Even among coup leaders, determining who will lead can be problematic,
as coup plotters are sometimes a group without a clear core leader. For instance,
following the 1973 Chilean coup, the initial plan for a rotating presidency among
military leaders was abandoned when General Pinochet consolidated control and
remained in power until 1990 [@svolik2014]. Second, the duration of the coup
leader's rule is uncertain. Leaders like Gamal Abdel Nasser in Egypt (1954 coup),
Muammar Gaddafi in Libya (1969 coup), and Idi Amin in Uganda (1971 coup) aimed to
retain power for life [@geddes2018], but their ability to do so was uncertain.
Others promise to transfer power to civilian authorities, but the timing and
fulfillment of these promises are unclear. For example, Myanmar's military junta
(2021 coup) has repeatedly extended a state of emergency, clinging to power
beyond the promised time-frame[^1]. Conversely, after the 2010 coup in Niger, the
military honoured their promise by restoring civilian rule within the same year
[@ginsburg2019]. Third, the successors of coup leaders are uncertain. Some may
designate successors from their inner circle, including family members, while
others may support general elections, though whether this will be fulfilled as
intended remains uncertain.

In contrast, autocoup leaders present a clearer picture regarding leadership and
tenure. There is no ambiguity about who will rule after an autocoup. In the
medium term, autocoup leaders typically hold office themselves. Many, like Putin
in Russia and Xi Jinping in China, seek to extend their rule indefinitely and are
unlikely to relinquish power voluntarily. Others attempt to extend their terms
incrementally, such as President Menem of Argentina, who overstayed until 1993
but failed in his bid for another term in 1999 [@llanos2019].

#### Instability

Moreover, the awareness of their shaky legitimacy and the persistent uncertainty
breeds insecurity and a perpetual sense of crisis among coup-entry leaders and
autocoup leaders. In a bid to solidify their grip on power, they often resort to
reshaping power dynamics or purging potential adversaries. Paradoxically, these
attempts to bolster stability and fortify their rule frequently boomerang,
unleashing even greater turmoil and instability.

The stability of a regime, particularly in an autocracy, hinges on maintaining a
balance of power. Coups, however, inevitably disrupt this balance, even when they
are bloodless, necessitating the creation of a new equilibrium. The ousting of
previous rulers requires dismantling the established governing structure and
reshuffling high-ranking officials, actions that inherently generate instability
and create adversaries for the new leadership. This makes restoring order and
establishing a balanced power structure notably challenging. Studies show that
new leaders often purge rival elite groups to consolidate their power at the
outset of their tenure [@sudduth2017; @roessler2011]. Such actions can provoke
backlash even from close allies. For instance, in Uganda, President Obote's
attempt to undermine the army commander-in-chief, Idi Amin, led to Amin gaining
the army's support and ultimately ousting Obote in a 1971 coup. Similarly, in
Pakistan in 1999, shortly after Prime Minister Sharif dismissed powerful army
chief General Pervez Musharraf, Sharif himself was ousted in a coup orchestrated
by Musharraf and his supporters [@sudduth2017].

To consolidate power, coup-entry leaders often have to compromise with internal
or external power challengers. However, these compromises are frequently unstable
and easily broken. The situation becomes even more complex when there is a risk
of civil war. Leaders may attempt to reduce the likelihood of subsequent coups,
potentially increasing the chances of societal rebellions and civil wars
[@roessler2011].

Moreover, instability extends beyond leadership to policies. A new leadership
group often brings new policies, and coups are sometimes triggered by
disagreements over significant policies. Major policy shifts can instigate
dissent or grievances from various ruling factions, communities, regions,
ethnicities, or religions.

In contrast, autocoup leaders encounter far fewer of these issues, as their
regimes experience fewer abrupt changes. They face less pressure to dismantle the
existing ruling paradigm and establish a new order. Even when adjustments are
necessary, they have more time to implement changes gradually.

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-entry leaders 
#| echo: !expr format == "html"
#| results: asis
# library(kableExtra)
read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(100),
    2:3 ~ px(250)
  ) |>
   data_color(rows = c(2, 4), colors = "#edede9") 
```

In conclusion, coup-entry leaders face a significantly greater degree of the
challenges discussed earlier, placing them at a substantial disadvantage in the
power dynamic compared to autocoup leaders. This disadvantage creates a
self-perpetuating cycle. Weaker leaders struggle to attract and retain strong
support, making them more vulnerable to internal and external challenges. The
perception of risk discourages potential allies, further eroding their power
base.

Empirical evidence bolsters this dynamic. Data reveals a correlation between the
frequency of coup attempts in a country and the likelihood of future coups (e.g.,
@tbl-coups shows over a third of coups occurring in the top ten countries with
the most attempts since 1950). This suggests that the more coups occur in a
country, the more likely additional coups are to happen in the future.

```{r}
#| label: tbl-coups
#| tbl-cap: Top 10 countries with the most coup attempts (1950–2023)
#| echo: !expr format == "html"

coup <- read.csv("https://www.uky.edu/~clthyn2/coup_data/powell_thyne_coups_final.txt",
                 sep = "\t")

coups <- coup |>
  summarise(
    `Coup attempted` = n(),
    `Coup succeeded` = sum(coup == 2),
    .by = country
  )

coups |>
  # add one row which adds up the total
  arrange(desc(`Coup attempted`)) |>
  head(11) |>
  add_row(
    country = "Total",
    `Coup attempted` = sum(coups[, 2]),
    `Coup succeeded` = sum(coups[, 3])
  ) |>
  mutate(Success_rate = percent(`Coup succeeded` / `Coup attempted`,
                                accuracy = 0.1
                                )) |>
  clean_names(case = "title") |>
  gt() |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold"
              ) |>
  cols_align(
    align = "center",
    columns = 2:3
  ) |>
  tab_source_note(source_note = md("*Source: GIC dataset*"))
```

Conversely, autocoup leaders, often benefiting from a veneer of legitimacy and a
stronger initial position, are better able to consolidate power and attract
supporters. This advantage can be self-reinforcing, as a strong power base
discourages challenges and fosters loyalty. This dynamic is evident in cases like
China (2018), where the National People's Congress granted Xi Jinping the
potential to rule for life[^2], and Russia (2020), where constitutional changes
allow Putin to potentially remain in power until 2036[^3].

These features and the balance of power contribute to a shorter expected tenure
for coup-entry leaders compared to the relatively longer tenures of autocoup
leaders. The average survival period following an autocoup is approximately five
years longer than that of coup-entry leaders (@fig-logrank). Based on these
observations and discussions, I propose the following hypothesis:

> ***H1: Political leaders who successfully extend their tenure through autocoups
> are more likely to experience longer leadership survival compared to coup-entry
> leaders.***

In the subsequent section, I will outline the research methodology used in this
paper. I will introduce several control variables to determine if the hypothesis
remains persistent and robust.

## Research Design

### Methodology: Survival analysis

To test the hypothesis, I will employ two Cox models to analyse the survival
tenures of coup-entry leaders and autocoup leaders. Unlike the Kaplan-Meier
model, the Cox model enables the estimation of the impacts of multiple factors.
Although it does not directly estimate the duration of tenure in office, it
evaluates the hazard rate associated with being ousted from power. Essentially,
this represents different facets of the same phenomenon: as a leader's cumulative
hazard of being ousted increases, their probability of survival in office
decreases.

The first model will utilize the Cox proportional hazards model (referred to as
the Cox PH model), where I will only use the variables present at the entry year,
without considering the changes in these variables over the leaders' survival
times.

However, apart from the primary variable of interest in this research—the leader
type—control variables such as economic performance, Polity5 scores, and
political stability do change over time. Therefore, the second model will account
for these variations by using the time-dependent Cox model.

### Data

The main variables are as follows.

#### Dependent variables

-   **Survival Time:** The duration of a leader's tenure, measured in days. For
    coup-entry leaders, the survival time begins on the day they assume power
    through a coup. For autocoup leaders, the survival time starts on the
    expiration date of their original legitimate term. For example, Xi Jinping
    assumed power in 2013 and removed term limits in 2018. His original
    legitimate tenure was set to end in 2023, so his survival time begins in
    2023, not 2013 or 2018, marking the start of his post-autocoup tenure. The
    survival time concludes on the day the leader exits office, applicable to
    both coup-entry and autocoup leaders.

-   **End point status:** This variable indicates the manner in which the
    leader's tenure concluded, categorized as follows:

    -   **0 = Censored:** This status is assigned to leaders who leave office
        through means other than being ousted. This includes leaders who appoint
        their successors, opt for democratic transitions, or leave office due to
        health issues or natural death.

    -   **1 = Ousted:** This status is assigned to leaders who are forced out of
        office. This includes cases where leaders "voluntarily" resign under
        pressure, reflecting a de facto ousting by force.

#### Key Independent variable: Leader type

This variable categorizes leaders into two distinct groups:

-   Group A = Autocoup Leader: Leaders who extend their tenure through autocoups.
-   Group B = Coup-Entry Leader: Leaders who assume power through coups.This
    variable is the primary independent variable of interest, serving as the
    basis for comparing the survival time between these two types of leaders.

The data for both dependent and independent variables are sourced from @zhu2024
and @goemans2009.

#### Control variables

-   **Economic Performance:** This variable is measured using two indicators:
    economic level and economic growth rate.

    -   **Economic Level:** Represented by GDP per capita. This measure provides
        an indication of the overall economic health and standard of living in a
        country.

    -   **Economic Growth Trend:** Assessed using the current-trend (CT) ratio,
        developed by @krishnarajan2019. The CT ratio compares a country's current
        GDP per capita to the average GDP per capita over the previous five
        years. A higher CT ratio signifies stronger economic performance.

$$
\begin{aligned}
CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
\end{aligned}
$$

The GDP per capita data, expressed in constant 2017 international dollars (PPP)
and measured in units of \$10,000, is sourced from the V-Dem dataset by
@fariss2022. To account for the economic impact of the previous year, this data
is lagged by one year.

-   **Political Stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    Major Episodes of Political Violence dataset by Marshall. This index provides
    a comprehensive measure of the level of violence and conflict within a
    country, which can significantly impact leadership survival.
    [@marshall2005current].

-   **Degree of Democracy:** The level of democracy is gauged using Polity 5
    scores at the entry year for each respective country. These scores range from
    -10 (fully autocratic) to +10 (fully democratic), capturing the extent of
    democratic versus autocratic governance. This dataset is sourced from the
    Center for Systemic Peace (CSP)[^4] and provides an essential measure of
    political regime type, which can influence the stability and survival of
    leaders in power.

-   **Population Size:** To account for its potential impact on leaders' tenures,
    the log of the population size is considered. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on leadership survival. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's Age:** The age of the leader at the entry year is included as an
    additional variable in the analysis, offering insights into potential
    correlations with leadership survival. Older leaders may have different
    experiences, networks, and health considerations that could influence their
    ability to maintain power. This data is sourced from the leaders dataset by
    [@goemans2009].

Based on the methods and variables, we have the model as:

$$
\begin{aligned}
Surv(survival\_time, status) &= Group + \\
&GDP\_growth\_trend + GDP\_per\_capita + \\
&pop\_log + polity5 + Stability + age
\end{aligned}
$$

## Results and discussion

### Model results

Using the @survival package in R, we present the regression results for both the
Cox PH model and the Time-dependent Cox model in Table @tbl-cox.

\clearpage

```{r}
#| label: model
#| echo: !expr format == "html"

Mydata1 <- read.csv(here("data/mydata1.csv")) |> 
   mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")))

model_cox1 <- Mydata1 |> 
  coxph(
    Surv(survival_time, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + polstab + age,
    data = _
  ) 

tbl1 <- model_cox1|> 
   tbl_regression(exponentiate = T,
                 conf.int = NULL,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   polstab ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T, hide_se = F, hide_ci = T) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 

Mydata2 <- read.csv(here("data/mydata2.csv")) |> 
  mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-entry leaders")),
         survt = (tstop - tstart) * 365) 
model_cox2 <- Mydata2 |>
  coxph(
    Surv(T1, T2, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + polstab + age + cluster(id),
    data = _
  ) 
tbl2 <- model_cox2|>
  tbl_regression(exponentiate = T,
                 conf.int = NULL,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   polstab ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T, hide_se = F, hide_ci = T) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 
  
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox models for survival time of different types of leaders
#| echo: !expr format == "html"

tbl_merge(list(tbl1, tbl2),
          tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**")) |> 
  as_gt()
```

Generally speaking, both the Cox PH model and time-dependent Cox model analyses
revealed a statistically significant association between leadership type and the
hazard of removal from power. Coup-entry leaders were found to have a hazard
ratio of 2.42 (in time-dependent Cox model) compared to autocoup leaders
(reference group), assuming all other variables in the model are held constant.
The analysis suggests that coup-entry leaders face a significantly greater risk
of removal from power compared to autocoup leaders. At any given time during
their tenure, coup-entry leaders are 2.42 times more likely to be ousted from
power compared to autocoup leaders, all else being equal in the model.

All other control variables in Cox PH model are not statistically significant. In
time-dependent Cox model, however, GDP per capita and the log of population are
both statistically significant at 5% level.

-   **GDP per capita:** The hazard ratio of 0.94 indicates that for each unit
    increase in GDP per capita (measured in constant 2017 international dollars,
    PPP, in units of \$10,000), the hazard (or risk) of being ousted at any given
    time is reduced by 6%, assuming all other variables in the model are held
    constant. Similarly, it indicates that for each unit increase in the log of
    population, the hazard (or risk) of being oustedat any given time is reduced
    by 14%, assuming all other variables in the model are held constant.

-   **Population:** The hazard ratio of 0.86 indicates that for each unit
    increase in the logarithm of the population (one percent increase of
    population), the hazard (or risk) of being ousted at any given time decreases
    by 14%, assuming all other variables in the model are held constant.

### Discussion

```{r}
#| label: fig-coxSurv
#| fig-cap: Survival curves for Cox Model
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: !expr format == "html"

testdata1 <- Mydata1 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    polstab = mean(polstab),
    age = mean(age)
  )

model_cox1 |> 
  survfit(newdata = testdata1) |>
  ggsurvplot(data = testdata1,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))

testdata2 <- Mydata2 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    polstab = mean(polstab),
    age = mean(age)
  )

model_cox2 |> 
  survfit(newdata = testdata2) |>
  ggsurvplot(data = testdata2,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-entry leaders"))
```

The survival curves depicted in @fig-coxSurv illustrate the survival rates for
leaders of both types. Both the Cox Proportional Hazards (Cox PH) model and the
time-dependent model produce similar plots. Notably, the survival curve for
coup-entry leaders exhibits a significantly lower trajectory compared to that of
autocoup leaders. The steeper drop at the early stage for coup-entry leaders
indicates they are more likely to be ousted shortly after assuming power.
Additionally, the survival curve for coup-entry leaders crosses the median
survival line much earlier than that of autocoup leaders. This disparity suggests
that autocoup leaders tend to remain in power for longer durations than their
coup-entry counterparts.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard ratios and 95% CIs for Leader Ousting
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: !expr format == "html"

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw() 

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |> 
  mutate(term = c("Coup-entry leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw() 
```

@fig-coxHR displays the hazard ratios and corresponding 95% confidence intervals
for the variables incorporated in the Cox model. Both the Cox Proportional
Hazards (PH) model and the time-dependent model yield nearly identical plots. The
hazard ratio for coup-entry leaders compared to autocoup leaders is significantly
greater than 1 and statistically significant, indicating that coup-entry leaders
are at a higher risk of being ousted. Most other variables have hazard ratios
close to 1, suggesting that a one-unit increase in these variables does not
significantly affect the risk of being ousted. Although the hazard ratio for GDP
growth trend is considerably less than 1, it is not statistically significant at
the 5% level. However, it is statistically significant at the 10% level,
suggesting that better economic performance may help to consolidate the rule of
the incumbents to some extent.

@fig-coxHR presents the hazard ratios (HRs) and their corresponding 95%
confidence intervals (CIs) for the variables included in the Cox proportional
hazards model (Cox PH) and the time-dependent model. The HR for coup-entry
leaders compared to autocoup leaders is significantly greater than 1, indicating
a substantially higher risk of removal from power for coup-entry leaders. Most
other variables have HRs close to 1, suggesting that a one-unit increase in those
variables has minimal impact on the risk of removal.

The HR for GDP growth trend is noteworthy. While it is considerably less than 1,
it is statistically significant at the 10% level (0.10 \> p-value \> 0.05). This
suggests a possible trend where positive economic performance might reduce the
risk of removal from power for the incumbents, although the evidence is not
conclusive.

### Assessing the Proportional Hazards Assumption

The validity of the model assumptions significantly influences our confidence in
the results. To assess the proportional hazards assumption of the Cox model, we
use the chi-square test based on Schoenfeld residuals to determine whether the
covariate effects remain constant (proportional) over time. Although the Cox PH
model violates the proportional hazards assumption, our primary analysis relies
on the time-dependent Cox model, which does not show strong evidence of violating
the proportional hazards assumption for any covariate. The global p-value of
0.382 is much greater than the 5% significance level, indicating that the
proportional hazards assumption is reasonably met for the time-dependent Cox
model.

```{r}
#| echo: false
test <- cox.zph(model_cox2)
```

## Conclusion

This study examined the survival durations of political leaders who come to power
through irregular means, specifically coups and autocoups. We hypothesized that
the mode of accession significantly influences leader tenure. Employing survival
analysis techniques, including the Cox proportional hazards model and a
time-dependent Cox model, we found strong evidence that autocoup leaders
generally enjoy longer tenures than coup-entry leaders.

Our findings revealed a significant difference in average tenure, with
post-autocoup leaders averaging approximately 10.8 years in power compared to 5.4
years for coup-entry leaders. The time-dependent Cox model further indicated that
coup-entry leaders are 2.42 times more likely to be ousted from power at any
given time compared to autocoup leaders, all else being equal.

These results highlight the importance of understanding the phenomenon of
autocoups, where leaders extend their rule by manipulating legal frameworks. Due
to the relative ease and potential benefits of autocoups, this method of power
retention might incentivize more leaders to employ it. Consequently, democratic
backsliding could become more prevalent as autocoups weaken democratic
institutions and constitutional norms, particularly in nascent democracies or
those transitioning from autocracy.

This study contributes to the field of leadership survival by demonstrating that
the mode of accession significantly impacts leader tenure, a factor previously
under-explored in the literature. By utilizing both Cox models, the research
offers robust analytical techniques for studying political leadership survival
and provides strong evidence of divergent tenure lengths between these two types
of irregular-entry leaders.

However, limitations exist. The study relies heavily on the autocoup dataset
collected and coded by the author. The concept and data itself are relatively
novel within academia. Future research should refine and establish wider
recognition for the term "autocoup," leading to more accurate and comprehensive
data collection efforts. Expanding the dataset to include more cases and
integrating it with data on other irregular leadership transitions could yield a
more holistic understanding of political survival in such contexts.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: https://thediplomat.com/2023/08/myanmar-junta-extends-state-of-emergency-for-fourth-time/:
    Myanmar\
    Junta Extends State of Emergency for Fourth Time. Accessed on `r Sys.Date()`.

[^2]: https://www.bbc.co.uk/news/world-asia-china-43361276: China’s Xi allowed to
    remain ‘president for life’ as term limits removed. Accessed on
    `r Sys.Date()`.

[^3]: https://www.ucl.ac.uk/news/2020/jul/analysis-vladimir-putin-secures-constitutional-changes-allowing-him-\
    rule-until-2036: Analysis: Vladimir Putin secures constitutional changes
    allowing him to rule until 2036. Accessed\
    on `r Sys.Date()`.

[^4]: Center for Systemic Peace: <https://www.systemicpeace.org>. Accessed on
    `r Sys.Date()`.
