---
title: "Power Acquisition and Leadership Survival: A Comparative Analysis of Coup-installed and Autocoup Leaders"


date: last-modified
format:
  aps-pdf: 
    keep-tex: true
    header-includes:
      - \usepackage{sectsty}
      - \chapterfont{\centering}
      - \usepackage{lscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}

bibliography: references.bib

abstract: |
  This study examines the relationship between the methods of power acquisition and the tenure of leaders who ascend to power through unconventional means, with a particular focus on coup-installed and autocoup leaders. The central hypothesis posits that the mode of accession has a profound impact on leadership longevity. Utilizing Cox proportional hazards and time-dependent Cox models, this study provides robust empirical evidence of disparate survival times between these two leader types. The findings reveal that, on average, coup-installed leaders are 2.23 times more likely to be ousted from power than autocoup leaders, all else being equal. These results have far-reaching implications for political stability and democratic processes, suggesting that the perceived low costs and high rewards associated with autocoups may incentivize incumbents to prolong their tenure through this means, potentially contributing to democratic erosion. This research makes a notable contribution to the academic literature by offering nuanced insights into the dynamics of irregular leadership transitions and enhances our understanding of the complex interplay between power acquisition methods and leadership longevity.
  
keywords: [Coups, Autocoups, Leadership Survival, Cox Model]

execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
pacman::p_load(tidyverse, data.table, gt, gtsummary, gtExtras, janitor, scales, zoo, democracyData, vdemdata, readxl, here, stargazer, survival, survminer, ggsurvfit, sampleSelection, marginaleffects, glue)
```

## Introduction

The enduring fascination with the longevity of political leaders has sparked
extensive research in political science, with scholars seeking to understand why
some leaders maintain power for decades while others are ousted in a matter of
months or even days. However, a specific subset of leaders—those who ascend to
power through coups or extend their tenure through autocoups—has received
relatively limited attention. Examining the tenures of these leaders is crucial,
as it sheds light on the dynamics of irregular leadership transitions and their
implications for political stability and democratic processes.

In contrast to leaders who attain power through conventional means, those who
rise through irregular channels, such as coups or autocoups, present more complex
and intriguing cases for study. The Archigos dataset highlights the prevalence of
irregular power transitions. Between 1945 and 2015, over half of leaders who
assumed power irregularly also exited irregularly, a rate significantly higher
than that of leaders who accessed office through regular channels.

Coup-installed and autocoup leaders constitute a substantial portion of these
irregular cases. The Archigos dataset notes that of 374 leaders who exited
irregularly, 246 (65.8%) were ousted through coups. Furthermore, research by
@frantz2016 demonstrates that coup-related exits account for approximately
one-third of all exits in autocracies, surpassing any other transition type.
Additionally, the autocoup dataset, introduced by @zhu2024, documents 110
autocoup attempts between 1945 and 2023, of which 87 were successful.

```{r}
#| label: data
#| include: false
survival_cox_ph_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_ph_model.csv") |> 
  mutate(survival_year = survival_days / 365)

survival_cox_td_model <- fread("https://raw.githubusercontent.com/reddylee/Datasets/main/survival_cox_td_model.csv")

avg <- survival_cox_ph_model |>
  summarise(avg=round(mean(survival_year),1),
            .by = group)
```

Measuring the tenure of coup-installed and autocoup leaders poses challenges due
to the inherent irregularity and uncertainty of their positions. Nevertheless, a
comparative analysis reveals that leaders who extend power through autocoups tend
to have longer average post-autocoup tenures (approximately `r avg$avg[1]` years)
compared to coup-installed leaders (approximately `r avg$avg[2]` years),
suggesting a potential tenure gap of over five years.

```{r}
#| label: fig-logrank
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
#| fig-cap: Survival curves of autocoup and coup-installed leaders


# Fit survival curves
fit <- survfit(Surv(survival_year, status) ~ group, data = survival_cox_ph_model)
# Create log-rank plot (Kaplan-Meier plot)
ggsurvplot(fit, 
           data = survival_cox_ph_model, 
           risk.table = TRUE, 
           risk.table.height = 0.3, 
           pval = TRUE,
           surv.median.line = "hv",
           break.x.by = 5, 
           legend.title = "Leader Type",
           legend.labs = c("Autocoup leaders", "coup-installed leaders"),
           xlab = "Survival Years",
           #title = "Survival Curves of overstay and coup-installed leaders",
           font.title = 12)
```

A preliminary log-rank test in survival analysis, as illustrated in @fig-logrank,
demonstrates a statistically significant difference between the tenures of
autocoup and coup-installed leaders. The survival curve for autocoup leaders
consistently exceeds that of coup-installed leaders, indicating longer survival
times and a reduced risk of ouster for autocoup leaders.

This study posits that the method of accession significantly influences
leadership longevity. Coup-installed leaders likely confront greater challenges
to their rule, resulting in shorter average tenures compared to autocoup leaders.
The analysis, employing Cox proportional hazards and time-dependent Cox models,
supports this hypothesis, demonstrating that autocoup leaders generally
experience longer tenures than coup-installed leaders.

This research offers two primary contributions to the field. First, it highlights
an understudied factor in leadership survival analysis: the impact of the method
of accession to power. The findings suggest that leader survival is influenced
not only by ruling strategies but also by the initial method of acquiring power.
Second, by employing survival models, this study provides empirical evidence of
the significant difference in tenure duration between autocoup and coup-installed
leaders. This insight may explain the increasing prevalence of tenure extensions
through autocoups since 2000, as more incumbents observe and potentially emulate
successful precedents.

The remainder of this article is structured as follows: Section 2 provides a
comprehensive literature review on political survival, establishing the context
for this research. Section 3 explores the factors influencing the survival of
coup and autocoup leaders. Section 4 outlines the methodology and data used,
including the application of survival models to analyze the determinants of
leadership longevity. Section 5 presents the analysis findings and a detailed
discussion of the results. Finally, Section 6 concludes by synthesizing key
takeaways and exploring their broader implications for political stability and
democratic processes.

## Literature review

The longevity of political leaders, which varies widely across different regimes,
countries, and historical periods, has been a longstanding focus of research in
political science. This field can be divided into two interconnected areas:
regime survival and individual leader survival, which are distinct but related
concepts. Regime survival focuses on the endurance of political systems, such as
monarchies, political parties, or specific ideological structures, while leader
survival concerns the duration of individual leaders' time in office.

Political survival patterns vary widely across different systems. Parliamentary
democracies (e.g., Japan, United Kingdom) often experience prolonged periods of
party dominance coupled with frequent leadership changes. Similarly, communist
regimes (e.g., China) typically demonstrate enduring party rule with more
frequent leadership transitions. Presidential systems (e.g., United States) and
many military regimes tend to exhibit more frequent changes in both ruling party
or junta and leader.

The existing literature on leader survival is extensive and multifaceted. Some
studies explore specific mechanisms influencing leadership longevity within
particular regimes, such as democracies [@svolik2014] or autocracies
[@davenport2021]. Others aim to develop more generalizable theoretical frameworks
explaining leader survival across different political systems
[@buenodemesquita2003]. While a universal theory remains an aspirational goal,
the complexities of leadership survival across diverse regime types present
significant challenges.

Power transition mechanisms vary substantially across different types of regimes,
particularly between democracies and autocracies. Autocratic systems often
feature closed leadership selection processes, restricted to a narrow pool of
individuals. While some autocracies may hold elections, significant barriers to
entry for legitimate challengers typically persist. The opacity of selection
processes in autocracies makes it difficult to assess genuine levels of public
support compared to democracies. Conceptualizing selectorates or winning
coalitions, as proposed by @buenodemesquita2003, becomes problematic in many
autocratic contexts.

Given these complexities, focusing research on specific regimes or leader types
may be more fruitful. The study of irregular leaders, such as those who ascend to
power through coups or extend their tenures through autocoups, offers a
compelling avenue for research due to the inherent complexities and uncertainties
surrounding their leadership trajectories.

Two primary perspectives have emerged to explain the dynamics of leader survival.
The first emphasizes objective factors and resources, such as personal competence
[@yu2016], societal stability [@arriola2009], economic development [@palmer1999;
@williams2011], natural resource endowments [@smith2004; @quirozflores2012;
@wright2013], and external support [@licht2009; @wright2008; @thyne2017]. The
second focuses on subjective factors and strategies, including political
policies, responses to opposition, and tactics for consolidating power
[@gandhi2007; @morrison2009; @escribà-folch2013; @davenport2021].

Coups, a significant aspect of irregular leadership transitions, have received
considerable scholarly attention. Research has examined coup prevention
strategies [@powell2017; @sudduth2017; @debruin2020]. Studies have explored the
impact of coups on leadership and the subsequent actions of coup leaders
[@sudduth2017, @sudduth2018, @easton2018].

However, a significant gap remains in the literature regarding the comparison of
leadership survival between coup-installed and autocoup leaders. This study aims
to address this gap by investigating and comparing the duration of leadership
survival for these two leader types.

By focusing on the comparison between coup-installed and autocoup leaders, this
study seeks to contribute to a more nuanced understanding of political survival
in irregular leadership transitions. This approach may offer valuable insights
into the complex dynamics of leader longevity across different political
contexts.

### Survival dynamics of autocoup and coup-installed leaders

Studying leadership survival in political systems is challenging due to the
opacity and diverse mechanisms of power transitions. These challenges, however,
underscore the significance of this research, as it illuminates understudied
dynamics in political leadership. Although the survival of political leaders is
complex and varied, some patterns do emerge. Leaders of similar types often
exhibit similar characteristics, allowing for meaningful analysis.

### Key definitions and scope

Before delving into the comparison, it is essential to clarify several key
terminologies:

-   **Coup and autocoup**: Coup is defined as "illegal and overt attempts by the
    military or other elites within the state apparatus to unseat the sitting
    executive" [@powell2011], and autocoup is defined as "the illegitimate
    extension of an incumbent leader's term in office beyond the originally
    mandated limits" [@zhu2024]. These definitions provide a clear distinction
    between the two types of irregular leadership transitions.

-   **Tenure length threshold**: To ensure meaningful analysis, this study
    focuses on leaders with substantial periods in power, applying a six-month
    threshold to both autocoup and coup-installed leaders. This criterion filters
    out ephemeral leadership episodes, allowing for a more robust examination of
    survival dynamics.

-   **Autocoup leader**: An incumbent leader who successfully employs
    illegitimate or unconstitutional means to extend their tenure in power. This
    definition encompasses various methods of power consolidation that circumvent
    established democratic processes or constitutional limits.

-   **Coup-installed leader**: The individual who assumes power after a
    successful coup, regardless of their role in the coup itself. This broad
    definition allows for the inclusion of both coup instigators and those
    selected to lead post-coup, providing a comprehensive view of leadership
    dynamics following forceful regime change.

This study focuses on comparing the post-autocoup tenure of autocoup leaders with
the post-coup tenure of coup-installed leaders. This comparative approach is
motivated by the relevance and similarity of these leader types in terms of
illegitimacy, uncertainty, and instability. By examining these parallel yet
distinct paths to power, we can gain insights into the factors that influence
leadership longevity in irregular leadership transitions.

### Challenges in power consolidation

Both autocoup and coup-installed leaders confront distinct challenges in
consolidating their power, primarily stemming from the varying intensity of
issues related to illegitimacy, uncertainty, and instability. This disparity
creates an uneven playing field in terms of power dynamics, placing
coup-installed leaders at a significant disadvantage. @tbl-leaders provides a
comparative overview of the main features of autocoup and coup-installed leaders,
highlighting these key differences.

\blandscape

```{r}
#| label: tbl-leaders
#| tbl-cap: Main features of autocoup and coup-installed leaders 
#| echo: false
#| results: asis

read.csv(here("data/leader_comparison.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(150),
    2:3 ~ px(300)
  )
```

\elandscape

#### Illegitimacy

While both types of leaders suffer from a legitimacy deficit, the nature and
perception of this deficit differ significantly:

-   **Coup-installed leaders**: Their illegitimacy is blatant and unambiguous,
    stemming from the overt and often violent seizure of power. This overt act
    undermines pre-existing norms and institutions, generating immediate domestic
    and international condemnation.

-   **Autocoup leaders**: In contrast, autocoup leaders employ a more subtle and
    deceptive strategy, manipulating legal processes and institutions to create a
    façade of democratic legitimacy. This veneer of legality, while often thin,
    can provide a degree of cover and buy time for consolidating power.

#### Uncertainty

The irregular paths to power for both types of leaders create uncertainty about
the longevity of their rule and the mechanisms of their eventual departure.
However, the levels and sources of this uncertainty differ significantly.

-   **Coup-installed leaders:** These leaders face a trifecta of uncertainties.
    First, the immediate aftermath of a coup often involves a struggle for power
    within the junta or ruling coalition, creating ambiguity about who will
    ultimately consolidate control. Second, the tenure of coup-installed leaders
    is inherently precarious, subject to internal rivalries, popular uprisings,
    or counter-coups. Third, the lack of established succession mechanisms
    further amplifies uncertainty, making it difficult to predict the transfer of
    power and potentially triggering future instability.

-   **Autocoup leaders:** While not immune to uncertainty, autocoup leaders
    generally present a clearer picture. The question of who will rule
    post-autocoup is largely settled, as the incumbent retains power.
    Furthermore, many autocoup leaders openly aspire to extend their rule
    indefinitely or incrementally, attempting to establish a sense of permanence.
    This perceived stability, whether real or manufactured, can contribute to a
    more predictable political environment, at least in the short term.

#### Instability

The awareness of shaky legitimacy and persistent uncertainty inevitably breeds
insecurity and a sense of crisis, forcing both autocoup and coup-installed
leaders to prioritize stabilization measures. However, the nature and intensity
of these challenges differ:

-   **Coup-installed leaders:** These leaders face the daunting task of rapidly
    reshaping power dynamics, often resorting to purges and crackdowns to
    eliminate potential adversaries and consolidate control. This process of
    dismantling existing structures and building new ones generates significant
    instability, potentially alienating former allies and triggering resistance
    from various segments of society. The need to appease powerful actors both
    domestically and internationally further limits their options, forcing them
    into compromises that can undermine their authority and long-term stability.

-   **Autocoup leaders:** In contrast, autocoup leaders often benefit from a
    degree of continuity in regime personnel and institutions. This relative
    stability allows them to implement changes gradually, minimizing disruptions
    and mitigating potential backlash. While they may still face opposition, they
    are less likely to confront immediate and existential threats to their rule,
    providing them with more time and leverage to consolidate power.

By understanding these contrasting challenges, we can better appreciate the
relative advantages and disadvantages faced by autocoup and coup-installed
leaders. This comparative perspective provides a nuanced framework for analyzing
the strategies these leaders employ to consolidate power and navigate the
perilous terrain of irregular leadership transitions.

### Empirical evidence and hypothesis

Empirical evidence substantiates the disadvantage faced by coup-installed
leaders, revealing a complex interplay between historical precedent, power
consolidation challenges, and leadership longevity. This section presents key
data points and introduces the central hypothesis guiding this study.

Data analysis shows a significant correlation between the frequency of coup
attempts in a country and the likelihood of future coups. Notably, over a third
of coups have occurred in the top ten countries with the most attempts since 1950
[@powell2011]. This pattern suggests a self-reinforcing cycle of political
instability, where each successful coup increases the probability of subsequent
attempts, creating an environment of persistent uncertainty for coup-installed
leaders.

The disparity in leadership longevity between autocoup and coup-installed leaders
is starkly illustrated by survival data. As depicted in @fig-logrank, the average
survival period following an autocoup is approximately five years longer than
that of coup-installed leaders. This substantial difference in tenure length
underscores the divergent challenges faced by these two types of leaders in
maintaining their grip on power.

The distinct challenges faced by autocoup leaders and coup-installed leaders in
consolidating power create a self-perpetuating cycle that significantly
influences their tenure length:

-   Coup-installed leaders: Face greater legitimacy challenges and internal
    instability; Struggle to attract and retain strong support; More vulnerable
    to internal and external challenges; Shorter average tenures reinforce
    perception of instability.

-   Autocoup leaders: Often benefit from a veneer of legitimacy and a stronger
    initial position; Better able to consolidate power and attract supporters;
    Face less immediate threat of overthrow; Longer average tenures contribute to
    perception of stability.

This cycle suggests that the initial method of power acquisition or extension has
far-reaching consequences for a leader's ability to maintain their position over
time.

Based on these observations and the theoretical framework outlined earlier, I
propose the following hypothesis:

***H1: Political leaders who successfully extend their tenure through autocoups
are more likely to survive longer extended tenure compared to coup-installed
leaders.***

This hypothesis encapsulates the expected outcome of the divergent challenges and
advantages faced by autocoup and coup-installed leaders. By testing this
hypothesis, I aim to quantify the impact of the method of power acquisition or
extension on leadership longevity, contributing to a more nuanced understanding
of political survival in contexts of irregular transitions.

## Research design

This section employs survival analysis to test the hypothesis that autocoup
leaders have longer survival times in office compared to coup-installed leaders.
This study uses Cox models to analyze the survival tenures of autocoup and
coup-installed leaders, controlling for various factors that may affect their
time in office.

### Methodology: Survival analysis

Two Cox models will be employed to analyze the survival tenures of coup-installed
and autocoup leaders:

-   **Cox proportional hazards (PH) model**: This model uses only the variables
    present at the entry year, without considering changes over time.

-   **Time-dependent Cox model**: This model accounts for variations in
    time-dependent control variables such as economic performance and political
    stability.

The Cox model is preferred over the Kaplan-Meier model because it enables the
estimation of the impact of multiple factors. Although it does not directly
estimate the duration of tenure, it assesses the hazard rate associated with
being ousted from power. This approach captures different facets of the same
phenomenon: as a leader's cumulative hazard of being ousted increases, their
probability of survival in office decreases.

### Dependent variables[^1]

The dependent variables include survival time and end point status:

-   **Survival time:** Survival time refers to the duration of a leader's tenure,
    measured in days. For coup-installed leaders, the survival time begins on the
    day they assume power through a coup. For autocoup leaders, the survival time
    starts on the expiration date of their original legitimate term. For example,
    Russia's president Vladimir Putin assumed power in 2000 and, after serving
    two terms, stepped down in 2008. However, he remained in a powerful position
    as the prime minister and hand-picked Dmitry Medvedev to succeed him as
    president, while continuing to control the power behind the scenes. In this
    case, Putin's survival time begins in 2008, marking the start of his
    post-autocoup tenure. The survival time concludes on the day the leader
    finally exits office, applicable to both coup-installed and autocoup leaders.

-   **End point status:** This variable indicates the manner in which the
    leader's tenure concluded, categorized as follows:

    **0 = Censored:** This status is assigned to leaders who leave office through
    regular means other than being ousted. This includes leaders transferring
    power to their designated successors, leaving office as their terms expire,
    losing in general elections, voluntarily leaving office due to health issues,
    or dying of natural causes.

    **1 = Ousted:** This status is assigned to leaders who are forced to leave
    office. This includes leaders resigning under pressure, being ousted by coups
    or other forces, or being assassinated.

### Key independent variable: leader type

The key independent variable is the leader type, which categorizes leaders into
two distinct groups:

-   **Group A = Autocoup leader**: Leaders who extend their tenure through
    autocoups.
-   **Group B = Coup-installed leader**: Leaders who assume power through coups.

This variable is the primary independent variable of interest, serving as the
basis for comparing the survival time between these two types of leaders.

The data for both dependent and independent variables are sourced from the
Archigos dataset [@goemans2009] and PLAD dataset [@bomprezzi2024wedded]. The
autocoup dataset is introduced by @zhu2024.

### Control variables

The control variables are chosen based on the research of @gassebner2016. They
analysed 66 factors potentially influencing coups and found that slow economic
growth, prior coup attempts, and other forms of political violence are
particularly significant factors. Therefore, we include economic performance,
political violence, and the number of previous coups as our main control
variables.

```{r}
#| label: variables
#| results: asis
gdppc <- "-   **Economic Level:** Represented by GDP per capita. This measure provides an
    indication of the overall economic health and standard of living in a
    country. We use GDP per capita data (in constant 2017 international 1000
    dollars, PPP) from the V-Dem dataset by @fariss2022."

ct <- "-   **Economic Performance:** Measured using the current-trend ($CT$) ratio
    developed by @krishnarajan2019. This ratio compares a country's current GDP
    per capita to the average GDP per capita over the previous five years. A
    higher $CT$ ratio indicates stronger economic performance. For a country $i$
    at year $t$, the $CT$ ratio is calculated as follows:"

glue(gdppc)
glue(ct)
```

$$
    \begin{aligned}
    CT_{i,t} = {GDP/cap_{i,t} \over {1 \over 5} {\sum_{k=1}^5GDP/cap_{i,t-k}}}
    \end{aligned}
$$ {#eq-eq6}

-   **Political stability:** This variable captures overall regime stability by
    including a violence index that encompasses all types of internal and
    interstate wars and violence. The data for this index is sourced from the
    variable “actotal” in the Major Episodes of Political Violence dataset
    [@marshall2005current], with 0 representing the most stable conditions (no
    violence at all) and 18 representing the most unstable.

-   **Previous coups:** Included in the selection equation as either: a) The
    number of previous coups in a country (Model 1), or b) The time since the
    last coup attempt (Model 2 for robustness check).

-   **Population size:** To account for its potential impact on leaders' tenures,
    we consider the log of the population size. This transformation helps in
    managing the wide range of population sizes across different countries. The
    data is sourced from the V-Dem dataset and is evaluated to understand its
    influence on power transitions. Larger populations may present more
    governance challenges and potential sources of opposition, thereby affecting
    the stability and longevity of a leader's tenure.

-   **Leader's age:** The age of the leader is included as an additional variable
    in the analysis, offering insights into potential correlations with
    leadership strength. Older leaders may have different experiences, networks,
    and health considerations that could influence their ability to maintain
    power. This data is sourced from Archigos and PLAD datasets.

Unlike the analysis of classic coup determinants, which could theoretically occur
in any given year, I assume that an autocoup happens only once during an
incumbent leader's tenure, as a successful autocoup negates the need for another
attempt. However, this assumption does not always reflect reality, as leaders
might attempt further extensions or try again after a failed attempt. For
simplicity, I overlook these possibilities in our analysis.

Therefore, in our probit model, the unit of analysis for autocoups is the entire
tenure of a leader, rather than a country-year. I establish a base year for the
variables: for leaders who staged an autocoup, we use the year of their first
attempt as the base year; for leaders who did not attempt to overstay, I use the
middle year of their tenure as the base year.

## Results and discussion

### Model results

Using the ***surviavl*** package in R [@survival], I present the regression
results for both the Cox Proportional Hazards model (Cox PH) and the
time-dependent Cox model in @tbl-cox.

```{r}
#| label: model

Mydata1 <- survival_cox_ph_model |> 
   mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-installed leaders")))

model_cox1 <- Mydata1 |> 
  coxph(
    Surv(survival_days, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age,
    data = _
  ) 

tbl1 <- model_cox1|> 
   tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T,
                         hide_se = F, 
                         hide_ci = T,
                         thresholds = c(0.01, 0.05, 0.1),
                         ) |> 
  bold_p(t = 0.05) |> 
  bold_labels() 

Mydata2 <- survival_cox_td_model |> 
  mutate(group = factor(group, 
                        labels = c("Autocoup leaders", "Coup-installed leaders")),
         survt = (tstop - tstart) * 365)

model_cox2 <- Mydata2 |> 
  coxph(
    Surv(T1, T2, status) ~ 
      group + GDP_trend + GDP_pc + pop_log + 
      polity5 + violence + age + cluster(id),
    data = _
  ) 
tbl2 <- model_cox2|>
  tbl_regression(exponentiate = T,
                 conf.int = F,
                 add_estimate_to_reference_rows = T,
                 label = list(group ~ "Leader Type",
                   GDP_trend ~ "GDP Growth Trend",
                   GDP_pc ~ "GDP per capita",
                   pop_log ~ "Population: log",
                   polity5 ~ "Polity 5",
                   violence ~ "Political stability",
                   age ~ "Age")) |> 
  add_n(location = "level") |> 
  add_nevent(location = "level") |> 
  add_significance_stars(hide_p = T,
                         hide_se = F, 
                         hide_ci = T,
                         thresholds = c(0.01, 0.05, 0.1),
                         ) |> 
  bold_p(t = 0.05) |> 
  bold_labels()
```

```{r}
#| label: tbl-cox
#| tbl-cap: Cox models for survival time of different types of leaders


tbl_merge(list(tbl1, tbl2),
          tab_spanner = c("**Cox PH Model**", "**Time-dependent Cox Model**"))
```

Both models showed a statistically significant relationship between leadership
type and the hazard of removal from power. Since time-dependent Cox model use the
control variables which change over time, I interpret the main findings based on
time-dependent model.

Coup-installed leaders were found to have a hazard ratio of 2.23 in the
time-dependent model compared to autocoup leaders (reference group), assuming all
other variables in the model are held constant. This suggests that coup-installed
leaders face a significantly greater risk of removal from power compared to
autocoup leaders. At any given time during their tenure, coup-installed leaders
are 2.23 times more likely to be ousted from power compared to autocoup leaders,
all else being equal in the model.

The control variables perform differently in the two models. Economic level (GDP
per capita) exhibits statistically significant effects in both models. In the
time-dependent model, the hazard ratio of 0.95 indicates that for each unit
increase in GDP per capita (measured in units of \$10,000), the hazard (or risk)
of being ousted at any given time is reduced by 5%, assuming all other variables
in the model are held constant.

GDP growth trend demonstrates a more substantial effect in reducing the risk of
coups. Specifically, a 1 percentage point higher economic growth trend is
associated with an 80% reduction in the risk of being ousted, although this
effect is only statistically significant at the 10% level. This suggests a
possible trend where positive economic performance might mitigate the risk of
removal from power, but the evidence is not robust enough to confirm this
conclusively.

Political stability, as measured by the violence index, shows that a 1-point
increase in the index correlates with an 11% higher risk of being ousted.
However, this effect is also only statistically significant at the 10% level,
indicating a weaker but potentially important relationship between increased
violence and the risk of removal from office.

### Discussion

```{r}
#| label: fig-coxSurv
#| fig-cap: Survival curves for Cox Model
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

testdata1 <- Mydata1 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox1 |> 
  survfit(newdata = testdata1) |>
  ggsurvplot(data = testdata1,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-installed leaders"))

testdata2 <- Mydata2 |> 
  summarise(
    .by = group,
    GDP_trend = mean(GDP_trend),
    GDP_pc = mean(GDP_pc),
    pop_log = mean(pop_log),
    polity5 = mean(polity5),
    violence = mean(violence),
    age = mean(age)
  )

model_cox2 |> 
  survfit(newdata = testdata2) |>
  ggsurvplot(data = testdata2,
             xlab = "Days",
             surv.median.line = "hv",
             legend.title = "Leader Type",
             legend.labs = c("Autocoup leaders", "Coup-installed leaders"))
```

The survival curves in @fig-coxSurv displays the survival rates for both types of
leaders, highlighting the differences in their survival curves. Both the Cox PH
model and the time-dependent Cox model yield similar results. Notably, the
survival curve for coup-installed leaders has a significantly lower trajectory
than that of autocoup leaders. The steeper drop at the early stage for
coup-installed leaders indicates they are more likely to be ousted shortly after
assuming power. Additionally, the survival curve for coup-installed leaders
crosses the median survival line much earlier (about 3,000 days) than that of
autocoup leaders (about 8,500 days). This disparity suggests that autocoup
leaders tend to remain in power for longer durations than their coup-installed
counterparts.

```{r}
#| label: fig-coxHR
#| fig-cap: Hazard ratios and 95% CIs for Leader Ousting
#| fig-subcap: 
#|  - Cox PH Model
#|  - Time-dependent Cox Model
#| layout-ncol: 2
#| echo: false

coxplottidy <- tidy(model_cox1, exponentiate = T, conf.int = T) |>
  mutate(term = c("Coup-installed leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = NULL, title = "") +
  theme_bw() 

coxplottidy <- tidy(model_cox2, exponentiate = T, conf.int = T) |> 
  mutate(term = c("Coup-installed leaders",
                      "GDP Growth Trend",
                      "GDP per capita",
                      "Population: log",
                      "Polity 5",
                      "Political stability",
                      "Age"))

coxplottidy |>
  ggplot(aes(
    y = term,
    x = estimate,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_pointrange() +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "skyblue") +
  labs(x = "Hazard ratios", y = "", title = "") +
  theme_bw() 
```

@fig-coxHR displays the hazard ratios and corresponding 95% confidence intervals
for the variables incorporated in the Cox model. Both the Cox Proportional
Hazards (PH) model and the time-dependent model produce similar plots,
reinforcing the robustness of the findings. Key points to note include:

-   The closer the hazard ratio (represented by the dots) is to 1, the less
    impact the variable has on the risk of being ousted. A hazard ratio of 1
    indicates no effect.

-   The whiskers extending from the dots represent the 95% confidence intervals.
    If these whiskers cross the vertical blue line at 1, it indicates that the
    variable is not statistically significant at the 5% level.

-   The hazard ratio for coup-installed leaders is significantly greater than 1
    and statistically significant at the 5% level. This indicates that
    coup-installed leaders face a substantially higher risk of being ousted
    compared to autocoup leaders.

-   Most other variables have hazard ratios close to 1, suggesting that a
    one-unit increase in these variables does not significantly affect the risk
    of being ousted.

-   Although the hazard ratio for GDP growth trend is considerably less than 1 in
    the time-dependent model, indicating a potential protective effect, it is not
    statistically significant at the 5% level. However, it is statistically
    significant at the 10% level, suggesting that better economic performance may
    help to consolidate the rule of the incumbents to some extent, albeit the
    evidence is not as strong.

### Assessing the proportional hazards assumption

Assessing the proportional hazards assumption is crucial for the validity of the
Cox model results. To evaluate this, we used the chi-square test based on
Schoenfeld residuals to determine whether the covariate effects remain constant
(proportional) over time. Although the Cox PH model violates the proportional
hazards assumption, our primary analysis relies on the time-dependent Cox model,
which does not show strong evidence of violating the proportional hazards
assumption for any covariate. The global p-value of 0.416 is much greater than
the 5% significance level, indicating that the proportional hazards assumption is
reasonably met for the time-dependent Cox model.

```{r}
#| label: robust_test
#| include: false
test <- cox.zph(model_cox2)
```

## Conclusion

This study explores the survival durations of political leaders who come to power
through unconventional means, specifically focusing on coups and autocoups. Based
on the hypothesis that the mode of accession affects leader tenure, I use
survival analysis techniques, including the Cox proportional hazards model and a
time-dependent Cox model, to investigate this phenomenon. The findings suggest
that autocoup leaders tend to have longer tenures than coup-installed leaders.

Empirical analysis reveals a significant disparity in tenure length: leaders who
assume power via autocoup remain in office for an average of 11 years, compared
to just 5.6 years for those installed by coups. Moreover, the time-dependent Cox
model indicates that coup-installed leaders are 2.23 times more likely to be
ousted from power at any given time compared to their autocoup counterparts, all
other factors being equal. These findings underscore the importance of
understanding the autocoup as a mechanism through which leaders extend their rule
by manipulating legal frameworks and weakening institutional constraints.

The implications of these findings are profound. The relative ease and potential
rewards of an autocoup could incentivize more leaders to resort to this method of
power retention, particularly in fragile democracies or transitioning regimes.
Consequently, democratic backsliding may become more prevalent, as autocoups
erode democratic institutions and undermine constitutional norms.

This study contributes significantly to the literature on political leadership
survival by showing that the mode of accession has a substantial impact on leader
tenure, an aspect that has received limited attention in previous research.
Methodologically, this work advances the field by applying robust survival
analysis techniques, including both Cox models, to provide a nuanced
understanding of the dynamics that influence leadership stability.

However, the study is not without limitations. The analysis relies on an autocoup
dataset that was collected and coded by the author, a relatively novel concept in
the academic sphere. As the understanding and recognition of "autocoup" as a term
continue to evolve, future research should refine and expand the dataset.
Incorporating additional cases and cross-referencing with other forms of
irregular leadership transitions would contribute to a more comprehensive view of
political survival under such conditions.

In conclusion, this study highlights the need for more refined approaches to
studying political tenure and irregular power retention. By offering valuable
insights into the dynamics of political stability and the risks associated with
non-democratic leadership transitions, this research emphasizes the importance of
continued investigation into the complex relationships between power, legitimacy,
and survival in political leadership.

```{r}
#| label: references
#| results: asis
cat("\\newpage
## References")
```

[^1]: The dataset used in this article is publicly available and can be accessed
    at: <https://github.com/reddylee/Datasets>.
