---
title: "Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-14
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: autocoup.html

categories: [Article, Coups, Autocoups]
bibliography: references.bib

abstract: |
  This study aims to clarify the concept of overstay coups, focusing on power extensions by incumbent leaders. By distinguishing overstay coups from the broader, more ambiguous concepts of self-coups or autocoups, which encompass both executive power aggrandizement and power extension, this research introduces a refined understanding of these specific political events.

  The research involves coding overstay coup events and compiling a comprehensive dataset covering the period from 1945 to the present. Using this dataset, two detailed case studies provide qualitative insights into the dynamics and outcomes of overstay coups. Additionally, a regression analysis identifies the determinants of overstay coup attempts, shedding light on the factors that influence incumbents' decisions to extend their power illegitimately.

  The findings contribute to the existing literature by providing a clearer conceptual framework and empirical evidence on overstay coups. This research enhances our understanding of the mechanisms and motivations behind power extensions by incumbent leaders, and it examines the implications for democratic resilience, democratic backsliding, democratic breakdown, personalization of power, and autocratic deterioration. The insights gained from this study inform policy and theoretical discussions on political stability and the challenges facing contemporary democracies.
  
keywords: [Coups, Autocoups, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

## Introduction

Understanding how political leaders come to power and leave office is crucial in
political science. Traditionally, research has focused on two main exit routes:
leaders who are forced out before their term ends (e.g., coups) or those who
complete their terms and leave voluntarily. However, a growing trend of leaders
attempting to extend their term limits through illegitimate means has been
understudied. This research examines these power extensions by political leaders,
or autocoup as defined in this study.

It is not surprising that coups, as the primary form of irregular exits, have
garnered significant academic attention in previous studies. According to
@goemans2009, out of 374 leaders who exited irregularly, 246 were ousted through
coups, accounting for 65.8% of these cases. Frantz and Stein
([2016](https://reddylee.quarto.pub/ps/autocoups.html#ref-frantz2016)) shows that
coup-related exits account for roughly one-third of all exits in autocracies,
surpassing any other type, including regular transitions.

In recent decades, however, once the most common means of leadership change,
traditional coups have become less frequent [@bermeo2016; @thyne2019]. In
contrast, autocoups, which are perpetrated by incumbent leaders instead of
against the incumbents, become more prevalence [@ginsburg2010evasion;
@baturo2014; @versteeg2020law]. While traditional coups are well-documented
[@thyne2019] due to its significant impact on politics, the equally effects of
autocoups, if not more important, have received much less research attention
compared to coups.

Firstly, autocoups are typically accompanied by a disregard for the rule of law,
weakened institutions, and potential democratic backsliding or authoritarian
personalization. Secondly, like traditional coups, successful autocoups increase
the risk of irregular power transitions in the future. Research indicates that
since 1945, approximately 62 percent of leaders who successfully overstayed their
term limits through autocoups in non-democratic countries were either ousted or
assassinated while in office [@baturo2019]. Thirdly, failed overstay coups often
trigger instability that fuels protests, violence, and even civil wars.

Despite its significance in politics, autocoup is understudied compared to coups.
Although several relevant concepts or terminologies such as self-coups,
autocoups, autogolpes, incumbent takeovers, executive aggrandizement, overstay,
and continuismo have been proposed and discussed [@marsteintredet2019;
@baturo2022], there is no clear and widely accepted term or terms to describe
this type of coups. Sometimes, several different terms are used to describe the
same, or at least similar, situations or events, while sometimes the same term is
used to describe two different situations without clear distinction. For
instance, self-coups, autocoups, and autogolpes are normally used to refer to
"when a president closes the courts and the legislature, suspends the
constitution, and rules by decree until a referendum and new legislative
elections are held to approve broader executive powers" [@cameron1998], but power
extensions are also included in their discussion although they should not be by
their definitions. Moreover, due to the widely accepted concepts, the collection
of autocoup datasets are in a very premitive stage compared to the rich datasets
of coups, hence the limited existing studies on autocoups have primarily relied
on case studies [@cameron1998; @antonio2021; @pion-berlin2022], with few focusing
on quantitative analyses.

In order to fully understand irregular political leadership transition and
survival, it is crucial to redefine and clarify the term of autocoup in the first
place. Based on the clear definition of the concept and term, we can continue to
introduce the autocoup dataset which can be used for further empirical analysis.

This study aims to address the knowledge gap surrounding autocoups and contribute
in three areas. First, it clarifies the terminology by defining autocoups
dependent mainly on power extension. Second, it introduces a new dataset of
autocoups since 1945 based on this refined definition. Finally, the research
utilizes this dataset for a quantitative analysis of the factors that influence
leaders' decisions to attempt autocoups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of
autocoups. This will be followed by an introduction to the new autocoup dataset
in Part III. In Parts IV and V, I will explain the determinants of autocoup
attempts through two case studies and demonstrate how the novel autocoup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude in Part VI, summarizing the key insights gained from the study and
suggesting avenues for future research.

Understanding how political leaders come to power and leave office is crucial in
political science. Traditionally, research has focused on two main exit routes:
leaders who are forced out before their term ends (e.g., coups) or those who
complete their terms and leave voluntarily. However, an increasing trend of
leaders extending their term limits through illegitimate means has been
understudied. This research examines these power extensions by political leaders,
specifically focusing on the concept of autocoups as defined in this study.

It is not surprising that coups, as the primary form of irregular exits, have
garnered significant academic attention. According to @goemans2009, out of 374
leaders who exited irregularly, 246 were ousted through coups, accounting for
65.8% of these cases. @frantz2016 show that coup-related exits account for
roughly one-third of all exits in autocracies, surpassing any other type,
including regular transitions.

In recent decades, however, traditional coups have become less frequent
[@bermeo2016; @thyne2019]. In contrast, autocoups—perpetrated by incumbent
leaders to extend their own rule—are becoming more prevalent
[@ginsburg2010evasion; @baturo2014; @versteeg2020law]. While traditional coups
are well-documented due to their significant political impact [@thyne2019], the
equally important effects of autocoups have received much less research
attention.

Firstly, autocoups are typically accompanied by a disregard for the rule of law,
weakened institutions, and potential democratic backsliding or authoritarian
personalization. Secondly, like traditional coups, successful autocoups increase
the risk of future irregular power transitions. Research indicates that since
1945, approximately 62 percent of leaders who overstayed their term limits
through autocoups in non-democratic countries were either ousted or assassinated
while in office [@baturo2019]. Thirdly, failed autocoups often trigger
instability that fuels protests, violence, and even civil wars.

Despite their significance, autocoups are understudied compared to traditional
coups. Although several relevant concepts and terminologies such as self-coups,
autocoups, autogolpes, incumbent takeovers, executive aggrandizement, overstay,
and continuismo have been proposed and discussed [@marsteintredet2019;
@baturo2022], there is no clear and widely accepted term to describe these
events. Different terms are often used interchangeably or inconsistently, leading
to confusion. For instance, self-coups, autocoups, and autogolpes are normally
used to refer to "when a president closes the courts and the legislature,
suspends the constitution, and rules by decree until a referendum and new
legislative elections are held to approve broader executive powers"
[@cameron1998], but power extensions are also included in their discussion,
despite not fitting the definitions. Moreover, due to the lack of clear concepts,
the collection of autocoup datasets is still in its early stages compared to the
rich datasets of traditional coups. Hence, the limited existing studies on
autocoups have primarily relied on case studies [@cameron1998; @antonio2021;
@pion-berlin2022], with few focusing on quantitative analyses.

To fully understand irregular political leadership transitions and survival, we
need to redefine and clarify the term "autocoup" first. Based on a clear
definition, we can introduce an autocoup dataset that can be used for further
empirical analysis. This study aims to address these knowledge gap surrounding
autocoups and contribute in three areas. First, it clarifies the terminology by
defining autocoups based mainly on power extension. Second, it introduces a new
dataset of autocoups since 1945 based on this refined definition. Finally, the
research utilizes this dataset for a quantitative analysis of the factors that
influence leaders' decisions to attempt autocoups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of
autocoups. This will be followed by an introduction to the new autocoup dataset
in Part III. In Parts IV and V, I will explain the determinants of autocoup
attempts through two case studies and demonstrate how the novel autocoup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude in Part VI, summarizing the key insights gained from the study and
suggesting avenues for future research.

## Overstay coup: Relevant terms and a new definition

### Relevant terms on power expansions and power extensions

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
