---
title: "Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-14
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: autocoup.html

categories: [Article, Coups, Autocoups]
bibliography: references.bib

abstract: |
  This study aims to clarify the concept of overstay coups, focusing on power extensions by incumbent leaders. By distinguishing overstay coups from the broader, more ambiguous concepts of self-coups or autocoups, which encompass both executive power aggrandizement and power extension, this research introduces a refined understanding of these specific political events.

  The research involves coding overstay coup events and compiling a comprehensive dataset covering the period from 1945 to the present. Using this dataset, two detailed case studies provide qualitative insights into the dynamics and outcomes of overstay coups. Additionally, a regression analysis identifies the determinants of overstay coup attempts, shedding light on the factors that influence incumbents' decisions to extend their power illegitimately.

  The findings contribute to the existing literature by providing a clearer conceptual framework and empirical evidence on overstay coups. This research enhances our understanding of the mechanisms and motivations behind power extensions by incumbent leaders, and it examines the implications for democratic resilience, democratic backsliding, democratic breakdown, personalization of power, and autocratic deterioration. The insights gained from this study inform policy and theoretical discussions on political stability and the challenges facing contemporary democracies.
  
keywords: [Coups, Autocoups, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

## Introduction

Understanding how political leaders come to power and leave office is crucial in
political science. Traditionally, research has focused on two main exit routes:
leaders who are forced out before their term ends (e.g., coups) or those who
complete their terms and leave voluntarily. However, an increasing trend of
leaders extending their term limits through illegitimate means has been
understudied. This research examines these power extensions by political leaders,
specifically focusing on the concept of autocoups as defined in this study.

It is not surprising that coups, as the primary form of irregular exits, have
garnered significant academic attention. According to @goemans2009, out of 374
leaders who exited irregularly, 246 were ousted through coups, accounting for
65.8% of these cases. @frantz2016 show that coup-related exits account for
roughly one-third of all exits in autocracies, surpassing any other type,
including regular transitions.

In recent decades, however, traditional coups have become less frequent
[@bermeo2016; @thyne2019]. In contrast, autocoups—perpetrated by incumbent
leaders to extend their own rule—are becoming more prevalent
[@ginsburg2010evasion; @baturo2014; @versteeg2020law]. While traditional coups
are well-documented due to their significant political impact [@thyne2019], the
equally important effects of autocoups have received much less research
attention.

Firstly, autocoups are typically accompanied by a disregard for the rule of law,
weakened institutions, and potential democratic backsliding or authoritarian
personalization. Secondly, like traditional coups, successful autocoups increase
the risk of future irregular power transitions. Research indicates that since
1945, approximately 62 percent of leaders who overstayed their term limits
through autocoups in non-democratic countries were either ousted or assassinated
while in office [@baturo2019]. Thirdly, failed autocoups often trigger
instability that fuels protests, violence, and even civil wars.

Despite their significance, autocoups are understudied compared to traditional
coups. Although several relevant concepts and terminologies such as self-coups,
autocoups, autogolpes, incumbent takeovers, executive aggrandizement, overstay,
and continuismo have been proposed and discussed [@marsteintredet2019;
@baturo2022], there is no clear and widely accepted term to describe these
events. Different terms are often used interchangeably or inconsistently, leading
to confusion. Moreover, due to the lack of clear concepts, the collection of
autocoup datasets is still in its early stages compared to the rich datasets of
traditional coups. Hence, the limited existing studies on autocoups have
primarily relied on case studies [@cameron1998; @antonio2021; @pion-berlin2022],
with few focusing on quantitative analyses.

To fully understand irregular political leadership transitions and survival, we
need to redefine and clarify the term "autocoup" first. Based on a clear
definition, we can introduce an autocoup dataset that can be used for further
empirical analysis. This study aims to address these knowledge gap surrounding
autocoups and contribute in three areas. First, it clarifies the terminology by
defining autocoups based mainly on power extension. Second, it introduces a new
dataset of autocoups since 1945 based on this refined definition. Finally, the
research utilizes this dataset for a quantitative analysis of the factors that
influence leaders' decisions to attempt autocoups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of
autocoups. This will be followed by an introduction to the new autocoup dataset
in Part III. In Parts IV and V, I will explain the determinants of autocoup
attempts through two case studies and demonstrate how the novel autocoup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude in Part VI, summarizing the key insights gained from the study and
suggesting avenues for future research.

## Autocoup: A review of relevant terms and a new definition

### Relevant terms on power expansions and power extensions

Compared to coups d'état, which are clearly defined as illegal attempts by elites
within the ruling group to overthrow the leadership [@powell2011], the concept of
autocoups (also known as autogolpes and self-coups) is mired in inconsistency,
ambiguity, and confusion.

The Spanish term "autogolpe," "self-coup" in English, gained prominence after
Peruvian President Alberto Fujimori dissolved Congress, temporarily suspended the
constitution, and ruled by decree in 1992 [@mauceri1995; @cameron1998]. It is
commonly defined as "a temporary suspension of the constitution and dissolution
of congress by the executive, who rules by decree until new legislative elections
and a referendum can be held to ratify a political system with broader executive
power" [@cameron1998a, p.220]. This definition, or variations of it, is widely
used in the literature on coups and autocoups [@przeworski2000; @cameron1998a;
@bermeo2016; @helmke2017; @marsteintredet2019]. Other terms, such as
"presidential coup" [@roberts1995; @marshall2019coup], "executive coup"
[@bermeo2016], and "constitutional coup" [@helmke2017], have also been used.

However, as @marsteintredet2019 point out, the term "self-coup" derived from
"autogolpe" can be misleading. It implies self-inflicted harm, which is
inaccurate since the leader orchestrating the autocoup benefits from it.
Autocoups are not self-directed; they are launched by the current leader against
other state institutions. To address this confusion, this study advocates for the
use of "autocoup" to denote coups staged by incumbents.

Another issue with the term "self-coup" is its focus on executive aggrandizement
or power expansion rather than power extension. It primarily describes situations
where executive leaders seize more power from other branches of government rather
than extending their leadership tenure. While power extensions might be included
to some extent in past literature, they are not the central concern.

@baturo2022 proposed "incumbent takeover" as an alternative term, referring to
"an event perpetuated by a ruling executive that significantly reduces the formal
and/or informal constraints on his/her power," drawing upon earlier research
[@svolik2014]. This term has merit as it encompasses both power expansions and
power extensions. However, "incumbent/executive takeover" lacks clarity regarding
the illegal nature of these events. Just as "coup" clearly denotes the illegal
ousting of leadership, a term like "leadership replacement" would not convey the
same gravity. Similarly, the legality of "incumbent/executive takeover" remains
ambiguous. While most takeovers employ a façade of legality, a clear counterpart
to "coup" is valuable. This is another reason why "autocoup" is preferred to
describe situations where incumbents extend their tenure through illegitimate
means.

Another approach involves using "coup" with adjectives or modifiers
[@marsteintredet2019]. Examples include presidential coup, executive coup,
constitutional coup, democratic coup, promissory coups [@bermeo2016], electoral
coup, judicial coup, slow-motion coup, soft coup, and parliamentary coup
[@marsteintredet2019]. While these terms can help analyse specific political
events, they often contribute to overall confusion. Returning to the core
definition of a classic coup, which focuses on the illegal replacement of
incumbent leaders, offers greater clarity.

While coups aim to replace incumbents, autocoups aim to prevent legal successors
from assuming power. Coups are overt illegal actions, whereas autocoups might
have a veneer of legality achieved through the manipulation of state institutions
and officials.

In light of this discussion, we formally define an autocoup as the practice of
current political leaders extending their time in office beyond their originally
mandated term limits, often achieved through seemingly legal means like
constitutional amendments or referendums, but ultimately illegitimate in essence.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
