---
title: "Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-14
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: autocoup.html

categories: [Article, Coups, Autocoups]
bibliography: references.bib

abstract: |
  This study aims to clarify the concept of overstay coups, focusing on power extensions by incumbent leaders. By distinguishing overstay coups from the broader, more ambiguous concepts of self-coups or autocoups, which encompass both executive power aggrandizement and power extension, this research introduces a refined understanding of these specific political events.

  The research involves coding overstay coup events and compiling a comprehensive dataset covering the period from 1945 to the present. Using this dataset, two detailed case studies provide qualitative insights into the dynamics and outcomes of overstay coups. Additionally, a regression analysis identifies the determinants of overstay coup attempts, shedding light on the factors that influence incumbents' decisions to extend their power illegitimately.

  The findings contribute to the existing literature by providing a clearer conceptual framework and empirical evidence on overstay coups. This research enhances our understanding of the mechanisms and motivations behind power extensions by incumbent leaders, and it examines the implications for democratic resilience, democratic backsliding, democratic breakdown, personalization of power, and autocratic deterioration. The insights gained from this study inform policy and theoretical discussions on political stability and the challenges facing contemporary democracies.
  
keywords: [Coups, Autocoups, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

## Introduction

The study of irregular power transitions, encompassing both irregular entries and
exits from office, has long been a central topic in political science. However,
one form of irregular power transition - the incumbent leader's refusal to
relinquish power - remains relatively understudied. This scenario, where leaders
overstay their mandated term limits for one or more terms, or even indefinitely,
lacks a universally accepted term in the existing literature. To address this
gap, this study adopts the term "autocoup" to denote this phenomenon.

Coups, being the primary form of irregular power transitions, have received
significant scholarly attention. Research by @goemans2009 indicate that coups
account for a substantial portion of such transitions, with roughly two-thirds
(65.8%) of 374 irregular exits occurring through coups. Additionally, @frantz2016
demonstrate that coups are the most common form of exit in autocracies,
surpassing regular transitions by a third.

However, while coups were once frequent, recent decades have witnessed a decline
in their occurrence [@bermeo2016; @thyne2019]. Conversely, autocoups appear to be
on the rise, particularly since the Cold War's end [@ginsburg2010evasion;
@baturo2014; @versteeg2020law]. Despite the significant political impact and
rising prevalence of autocoups, their study lags behind that of classical coups,
which are extensively documented due to their historical frequency [@thyne2019].

This research aims to bridge this gap by focusing on the understudied phenomenon
of autocoups as they are at least equally important and warrant comprehensive
analysis. Firstly, autocoups are typically accompanied by a disregard for the
rule of law, weakened institutions, and potential democratic backsliding or
authoritarian personalization. Secondly, like traditional coups, successful
autocoups increase the risk of future irregular power transitions. Research
indicates that since 1945, approximately 62 percent of leaders who overstayed
their term limits through autocoups in non-democratic countries were either
ousted or assassinated while in office [@baturo2019]. Thirdly, failed autocoups
often trigger instability that fuels protests, violence, and even civil wars.

Despite their significance, autocoups are understudied compared to traditional
coups. Although several relevant concepts and terminologies such as self-coups,
autocoups, autogolpes, incumbent takeovers, executive aggrandizement, overstay,
and continuismo have been proposed and discussed [@marsteintredet2019;
@baturo2022], there is no clear and widely accepted term to describe these
events. Different terms are often used interchangeably or inconsistently, leading
to confusion. Moreover, due to the lack of clear concepts, the collection of
autocoup datasets is still in its early stages compared to the rich datasets of
traditional coups. Hence, the limited existing studies on autocoups have
primarily relied on case studies [@cameron1998; @antonio2021; @pion-berlin2022],
with few focusing on quantitative analyses.

To fully understand irregular political leadership transitions and survival, we
need to redefine and clarify the term "autocoup" first. Based on a clear
definition, we can introduce an autocoup dataset that can be used for further
empirical analysis. This study aims to address these knowledge gap surrounding
autocoups and contribute in three areas. First, it clarifies the terminology by
defining autocoups based mainly on power extension. Second, it introduces a new
dataset of autocoups since 1945 based on this refined definition. Finally, the
research utilizes this dataset for a quantitative analysis of the factors that
influence leaders' decisions to attempt autocoups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of
autocoups. This will be followed by an introduction to the new autocoup dataset
in Part III. In Parts IV and V, I will explain the determinants of autocoup
attempts through two case studies and demonstrate how the novel autocoup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude in Part VI, summarizing the key insights gained from the study and
suggesting avenues for future research.

## Autocoups: A literature review and clarification of definitions

Compared to coups, which are clearly defined and widely accepted as illegal
attempts by elites within the ruling group to overthrow the leadership
[@powell2011], the concept of autocoups (also known as self-coups or autogolpes)
suffers from a lack of consistency and clarity. This ambiguity hinders our
understanding and study of a critical phenomenon in irregular power transitions.
To address this gap, we need to tackle two key issues: terminology and
definition.

### Terminology

The most common term in autocoup literature is self-coup, or autogolpe in Spanish
[@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017; @marsteintredet2019].
This term gained academic prominence after Peruvian President Alberto Fujimori
dissolved Congress, temporarily suspended the constitution, and ruled by decree
in 1992 [@mauceri1995; @cameron1998]. However, as @marsteintredet2019 point out,
the term "self-coup" can be misleading, as it implies a coup against oneself,
which is inaccurate since it is not self-directed but targets other state
institutions or apparatus.

Another approach to describe coups staged by incumbents is to use terms with
adjectives or modifiers, such as presidential coup, executive coup,
constitutional coup, electoral coup, judicial coup, slow-motion coup, soft coup,
and parliamentary coup [@marsteintredet2019]. While these terms can be helpful in
specific contexts, their proliferation often adds to the overall confusion rather
than providing clarification. Most of these terms focus on the specific methods
used by coup perpetrators but fail to clearly identify the perpetrator,
necessitating further explanation. In fact, many of these methods could be
employed either by or against executive leaders.

A third alternative involves terms like "incumbent takeover" or "overstay".
Incumbent takeover referrs to "an event perpetuated by a ruling executive that
significantly reduces the formal and/or informal constraints on his/her power"
[@baturo2022, p.374], based on earlier research [@svolik2014]. While overstay is
defined as "staying longer than the maximum term as it stood when the candidate
originally came in office” [@ginsburg2011evasion, p.1844]. These terms identify
the perpetrator (the incumbent) and/or the nature of the event
(overstaying/extending power). However, they do not highlight the illegality or
illegitimacy of these actions. Therefore, they cannot serve as a direct
counterpart to "coup," which clearly denotes the illegality of leadership
ousters, while "takeover" or "overstay" diminish the severity.

Therefore, this study strongly advocates for "autocoup" as the most accurate and
appropriate term. It avoids the pitfalls of other terms by clearly conveying the
essence of the phenomenon as a coup-like power grab and effectively
distinguishing autocoups from classical coups by highlighting the perpetrator.

### Definition

While terminology is important, another issue arises with the previous definition
of autocoups: should we emphasize power expansion or power extension? Power
Expansion refers to situations where an executive branch expands its power beyond
its original boundaries, often by encroaching on the authority of other branches
like the legislature or judiciary. Power Extension, on the other hand, refers to
situations where a leader extends their tenure in office beyond the originally
mandated term limits. Existing definitions of self-coup or autocoup sometimes
touch upon power extensions, but they primarily focus on power expansions, which
has several drawbacks.

Firstly, defining autocoups primarily in terms of power expansion does not align
well with the definition of a coup. When we define a classical coup, the focus is
clearly on the ouster of the current leader, not merely a limitation or
restriction on their power. Using the same logic, a more appropriate definition
of an autocoup should prioritize the extension of executive leadership. We would
not classify an event as a coup if a vice president seizes some power from the
president, as long as the president remains in office. Similarly, an executive
leader acquiring more power from other branches should not automatically qualify
as an autocoup if their term limits remain unchanged.

Secondly, emphasizing power expansion raises issues regarding the purpose of
launching an autocoup. As @cameron1998a defined, a self-coup is "a temporary
suspension of the constitution and dissolution of congress by the executive, who
rules by decree until new legislative elections and a referendum can be held to
ratify a political system with broader executive power" [@cameron1998a, p.220].
However, without extending their time in office, the leader faces significant
risks once their term expires, as they could face legal or political
repercussions for the autocoup itself, leaving behind an overly powerful
executive branch for the successor. Although the term "self-coup" gained
prominence from the 1992 Fujimori case in Peru, which initially involved seizing
power from other institutions, it is important to note that Fujimori ultimately
extended his term limits through constitutional amendments. The 1993 Constitution
allowed Fujimori to run for a second term, which he won with popularity in April
1995. Shortly after Fujimori began his second term, his supporters in Congress
passed a law of "authentic interpretation" which effectively allowed him to run
for another term in 2000, which he won amid suspicions and rumors. However, he
did not survive the third term. In 2000, facing charges of corruption and human
rights abuses, Fujimori fled Peru and took refuge in Japan [@ezrow2019].

Thirdly, measuring the extent of power expansion to qualify as an autocoup can be
challenging. Power expansion might occur as a sudden, significant event, as seen
with Fujimori in 1992, or it could be gradual and incremental, as in the case of
Putin since assuming power in 1999 [@chaisty2019]. In the latter case, it is
unclear whether each step should be considered an autocoup or if the entire
process together should be labelled as such. If the latter, determining the point
at which it becomes an autocoup is again problematic.

Therefore, this study argues that a more accurate definition of autocoups should
prioritize power extension as the core characteristic.

In light of this discussion, we formally define an autocoup as *the practice of
current political leaders extending their time in office beyond their originally
mandated term limits*. Three key points need to be highlighted for this
definition. Firstly, this definition refers to the actual leaders of the country,
regardless of their official titles. Typically, this would be the president;
however, in some cases, such as in Germany, the primary leader is the premier, as
the president is a nominal head of state. Secondly, while the primary
characteristic of an autocoup is extending the term in office, this definition
does not exclude instances of power expansion. Both aspects can coexist, but the
extension of the term is the central element. Thirdly, autocoups, by their
nature, subvert legal norms and established power transfer mechanisms. While they
may employ seemingly legal tactics, their essence is illegitimacy. This critical
aspect will be explored further in Chapter III.

## Autocoup dataset

### Defining the scope

Defining any concept or term inevitably involves borderline cases that are
difficult to categorize. In the context of autocoups, it is particularly
challenging to determine whether a specific instance of overstaying should be
coded as an autocoup. To avoid ambiguity and dilemmas, we will code all instances
of incumbents extending their original mandated term in office as autocoups,
regardless of whether the extension is patently illegal or superficially legal.

This approach is taken because a truly legitimate amendment to power transition
institutions should only apply to subsequent leaders, not the incumbent. If the
incumbent changes the rules to their own benefit, the legitimacy of such changes
is questionable, even if the amendment procedures appear legal. We have valid
reasons to question their motivations since they are the beneficiaries of these
reforms.

Even so, some cases remain unclear. For example, consider a two-term limit
presidency where the leader wins a second term. Allegations of election rigging
might surface, raising suspicions about the fairness of the election results. If
such suspicions can be verified with evidence or judicial verdict, the case would
undoubtedly be coded as an autocoup. However, proving election rigging in
countries with powerful executives can be difficult. In such instances, we will
not classify it as an autocoup unless a formal judicial verdict confirms the
rigging.

Even in contexts where leaders lack clear term limits or the constitution allows
indefinite terms, identifying autocoups remains possible. We will discuss and
explain this in the next section.

### **Classifying autocoups** {#sec-classify}

Autocoups can manifest in various forms. To understand this phenomenon better, we
can categorize them based on several key factors:

-   **Methods Employed:** This refers to the specific strategies incumbents use
    to extend their hold on power. Examples include constitution amending or
    reinterpreting, election cancelling, delaying or rigging.

-   **Degree of Legality:** This dimension captures the extent to which an
    autocoup deviates from established legal norms. It can range from minor
    rule-bending to outright constitutional violations.

-   **Duration of Extension:** This category focuses on the length of time the
    incumbent leader remains in office beyond their designated term limits. It
    could be a single additional term, multiple terms, or even a lifetime hold on
    power.

-   **Outcomes:** This factor examines whether the autocoup attempt is successful
    or ultimately fails.

This study will primarily focus on the **methods** employed by incumbents to
stage autocoups. However, we will also code for other relevant aspects like the
degree of legality, duration of extension, and outcomes whenever information is
available. This multifaceted approach allows for a more nuanced understanding of
the variations within autocoups.

#### Evasion of term limits

Evasion of term limits is a common tactic employed in autocoups. Incumbents often
resort to seemingly legal maneuvers to extend their hold on power. These
maneuvers primarily involve manipulating constitutional provisions through
various means. The incumbents may pressure legislative bodies (congress) or
judicial institutions (Supreme Court) to reinterpret existing term limits, amend
the constitution to extend terms, or even replace the constitution altogether.
This might also involve popular vote through referendums, or a combination of
these approaches. The extension can range from a single term to indefinite rule.

These manoeuvres primarily involve manipulating constitutional provisions through
various means.

-   **Changing Term Length:** Incumbents might lengthen the official term
    duration (e.g., from 4 to 6 years) to stay in office longer, even if the
    number of allowed terms remains unchanged. Examples, in the dataset, include
    Presidents Dacko (CAR, 1962), Kayibanda (Rwanda, 1973), and Pinochet (Chile,
    1988).

-   **Enabling re-election:** This approach involves incumbents modifying legal
    or constitutional frameworks to permit themselves to run for leadership
    again, despite initial restrictions. These restrictions might include
    prohibitions on re-election, bans on immediate re-election, or term limits
    that the incumbents have already reached. An illustrative example is
    President Menem of Argentina in 1993, who leveraged this tactic to extend his
    tenure.

-   **Removing Term Limits Altogether:** This approach, as seen with President Xi
    Jinping of China in 2018, technically allows the leader to rule for life,
    although they may still need to participate in elections (a formality in such
    cases).

-   **Leader for Life:** This differs from removing term limits as the leader
    still faces elections (although potentially rigged or uncontested). An
    example is Indonesia's President Sukarno, who attempted to declare himself
    president for life in 1963 (ultimately unsuccessful).

These methods are often used in combination. Initially, the duration of a term is
extended, followed by amendments to allow re-election, then the removal of term
limits, and finally, the declaration of the leader for life. For example, Haitian
President François Duvalier amended the constitution in 1961 to permit immediate
re-election and then declared himself president for life in 1964.

#### Election Manipulation or Rigging

Election manipulation or rigging is the second most commonly used tactic to
extend an incumbent's tenure.

-   **Delaying or Removing Elections:** Delaying or removing scheduled elections
    without legitimate justification is a frequent method used by incumbents to
    maintain power. For instance, Chadian President François Tombalbaye delayed
    general elections until 1969 after assuming power in 1960. Similarly, Angolan
    President José Eduardo dos Santos suspended elections throughout his rule
    from 1979 to 2017.

-   **Refusing Unfavourable Election Results:** Incumbents may refuse to accept
    unfavourable election results and attempt to overturn them through
    illegitimate means. For example, President Donald Trump of the United States
    refused to accept the results of the 2020 election and tried to overturn
    them.

-   **Rigging Elections:** Winning elections with an extraordinarily high
    percentage of votes is highly questionable. This study will code elections
    where the incumbent wins more than 90% of the vote as autocoups. For
    instance, President Teodoro Obiang of Equatorial Guinea has consistently won
    elections with over 95% of the vote in multi-party elections since 1996,
    indicating election rigging.

-   **Excluding Opposition in Elections:** Manipulating the electoral process by
    excluding opposition parties or candidates from participation, effectively
    creating a one-candidate race, clearly signifies an autocoup.

#### **Figurehead**

To circumvent term limits, some incumbents might choose a close associate to act
as a figurehead, taking the office publicly while the incumbent retains real
power behind the scenes. This can be achieved through seemingly subordinate
positions.

One example is Russia in 2008. Facing term limits, President Putin selected
Dmitry Medvedev to run for president. After the election, Medvedev appointed
Putin as Prime Minister. However, most analysts believe Putin wielded the true
power throughout this period.

#### **Constitutional power transferring**

This tactic involves an incumbent leader manipulating the constitution or legal
framework to create a new position of power, or elevate an existing one, before
stepping down from their current role. They then strategically take on this new
position, effectively retaining significant control despite appearing to
relinquish power. For example, in 2017, Recep Tayyip Erdoğan, the Prime Minister
of Turkey, spearheaded a constitutional referendum that transitioned the country
from a parliamentary system to a presidential one. This new system concentrated
significant executive power in the presidency. Following the referendum's
approval, Erdoğan successfully ran for the newly established presidency,
effectively retaining control under a different title.

#### One-Time Arrangement for Current Leaders

This strategy involves special arrangements that extend the term or tenure of
current leaders without altering the underlying institutions. For example,
Lebanon extended President Émile Lahoud's term by three years in 2004 through a
one-time arrangement.

### Data Coding

#### Data sources

```{r}
#| label: tbl-source
#| tbl-cap: Sources of coding autocoup dataset 
#| echo: !expr format == "html"
#| results: asis
# library(kableExtra)

read.csv(here("data/autocoup_source.csv")) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1:4 ~ px(80),
    5 ~ px(200)
  ) |>
   data_color(rows = c(1, 3), colors = "#edede9") 
```

The autocoup dataset, like most dataset coding procedures, is based on existing
studies and datasets. @tbl-source outlines the main sources for coding the
autocoup dataset. The Archigos dataset [@goemans2009] and the REIGN dataset
[@bell2016reign] provide comprehensive data on all leaders from 1875 to 2021,
although our coding only includes autocoups since 1945. These datasets help
identify the actual rulers of countries, saving time in distinguishing real
leaders from nominal heads of state.

The Incumbent Takeover dataset [@baturo2022], which integrates data from 11
related datasets, offers a broad spectrum of cases where leaders significantly
reduce the constraints on their power, encompassing both power expansions and
extensions. Since some executive takeovers do not qualify as autocoups due to the
lack of term extensions, I cross-referenced the Archigos dataset, which includes
detailed trajectories of leaders' entries and exits from power, to verify the
qualifications for autocoups.

As the Archigos dataset concludes at the end of 2015, I used trusted news reports
to validate data from the Incumbent Takeover dataset. In total, I coded 113
observations, with 75 overlapping with the candidate data from Incumbent
Takeover. The remaining 38 events were newly coded by the author through
verification with other sources such as Archigos, REIGN, and news reports.The
main deviation from the Incumbent Takeover dataset arises from excluding power
expansions that do not involve attempts to extend tenure.

#### Definitions of main variables

This section describes the variables included in the autocoup dataset.

-   **Country Identification:**

    -   `ccode` and `country`: These variables come from the Correlates of War
        project [@stinnett2002] and identify the countries included in the
        dataset, which are widely used in political science datasets, ensuring
        consistency and compatibility across different studies and datasets.

-   **Leader Information:**

    -   `leader_name`: This variable follows the Archigos dataset's coding and
        records the name of the de facto leader who wielded power in the country.

-   **Timeline Variables:**

    -   `entry_date`: Captures the date the leader assumed power.

    -   `exit_date`: Records the date the leader left office.

    -   `autocoup_date`: Indicates the date the autocoup is considered to have
        occurred. Since extensions often happen incrementally, this date reflects
        a significant event marking the extension, such as a legislative vote or
        successful referendum.

    -   `extending_date`: Represents the start date of the leader's additional
        term acquired through the autocoup.

    The `entry_date` and `exit_date` come from the Archigos and REIGN datasets,
    while the other two dates are coded by the dataset creator.

-   **Power Transition Methods:**

    -   `entry_method` and `exit_method`: These categorical variables record how
        the leader entered and exited power (e.g., election, coup d'état, death).

    -   `entry_regular` and `exit_regular`: These dummy variables indicate
        whether the entry and exit methods were regular (e.g., election) or
        irregular (e.g., coup).

-   **Autocoup Details:**

    -   `autocoup_method`: This key variable captures the various methods leaders
        use to extend their stay in power (see @sec-classify for details).

    -   `autocoup_outcome`: This variable indicates the outcome of the autocoup
        attempt: "fail and lose power", "fail but complete original tenure", or
        "successful". For successful coups, the additional term length can be
        calculated from the difference between `exit_date` and `extending_date`.

-   **Data Source:**

    -   `source`: This variable identifies the dataset source used for coding,
        primarily differentiating among "Incumbent Takeovers" or other sources.

-   **Additional Notes:**

    -   `notes`: This variable provides context for exceptional cases. For
        instance, if a leader undertook multiple autocoup attempts, details are
        recorded here.

The dataset encompasses a total of 14 variables along with the `notes` field.

### Data descriptions

```{r}
#| label: data_descriptions
#| echo: false
autocoup <- read_excel(here("data/autocoup.xlsx")) |> 
  mutate(exit_date = ifelse(exit_date != "Incumbent", exit_date, NA)) |> 
  mutate(exit_date = case_when(is.na(exit_date) ~ Sys.Date(), 
                               .default = convert_to_date(exit_date)),
         extending_date = make_date(year = extending_date, month(entry_date), day(entry_date)))
country <- autocoup |> 
  distinct(ccode) |> 
  summarise(count = n())
```

```{r}
#| label: tbl-autocoup_method
#| tbl-cap: Autocoup methods and success rates (1945-2021)
#| eval: !expr format == "latex"
#| echo: false

autocoup |> 
  summarise(autocoup_attempted = n(),
            autocoup_succeeded = sum(autocoup_outcome == "Yes"),
    .by = autocoup_method) |> 
  arrange(-autocoup_attempted) |> 
  adorn_totals() |> 
  mutate(Success_rate = percent (autocoup_succeeded/autocoup_attempted)) |> 
  clean_names(case = "title") |> 
  gt() |>
  tab_options(column_labels.font.weight = "bold",
              table.width = pct(100),
              table.align = "center") |>
  cols_width(
    1 ~ px(180),
    2:3 ~ px(140),
    4 ~ px(100)
  )
```

According to our primary coding, we have identified 113 autocoup cases from 1945
to 2021, involving `r country$count` countries. As shown in @tbl-autocoup_method,
the most common autocoup method is enabling re-election, accounting for 49
events. This is followed by removing term limits altogether, with 14 cases, and
then delaying elections and declaring the leader for life, each with 9 cases.

Examining the success rates of autocoups, the total success rate is 78%, which is
significantly higher than the roughly 50% success rate of classical coups. This
suggests that incumbents are in an advantageous position to expand or extend
their powers as executive leaders. They can do so openly and gradually, whereas
coup plotters must operate in secrecy and face numerous challenges such as
promissory issues and the risk of betrayal or exposure.

However, the success rates vary significantly across different methods. Removing
term limits, delaying elections, declaring the leader for life, and canceling
elections are all 100% successful. In contrast, there are only 4 cases of
refusing election results, with just one succeeding. Although this sample is
small, it suggests that in regimes where general elections are held and
incumbents lose, the system is relatively more democratic. In such cases,
incumbents must run for office without the ability to rig elections, making it
less likely for them to overturn election results.

## Determinants of Autocoup Attempts: Case Studies

### High attempts and high success rate of autocoups in post-communist countries

From the dataset, we observe that in post-communist countries, both the frequency
and success rate of autocoups are notably high. Post-communist countries refer to
those that were communist at the end of the Cold War, regardless of whether they
have since transitioned to other types of regimes. In these countries, there are
12 documented cases[^1] of autocoups aimed at prolonging incumbency, with only 2
of these attempts failing. Look into the cases in post-communist countries, we
can see several characteristics in these countries. First of all, although most
of these countries transferred from communist regimes to non-communist countries
(except China), they inherit the authoritarian system of communist. Secondly, the
transitions did not result in the removal or overthrow of the previous ruling
groups and the former communist elites completely remain in power. Thirdly,
although most of the countries introduced general elections and term limits, due
to the inertia of former communist regimes, the term limits are often broken and
the elections are rigged.

For example, Lukashenko was the member Supreme Soviet of the Byelorussian Soviet
Socialist Republic before the dissolution of the Soviet Union. After the
dissolution, he assumed the position of head of the interim anti-corruption
committee of the Supreme Council of Belarus. He was elected as the first
president of Belarus in 1994, then holding the office ever since. The same
pattern happens to all the five Central Asian countries of the former Soviet
Union. In the 1994 constitutions, the presidency has a maxium of two successive
terms, but this was removed by Lukashenko in 2004. Further more, International
monitors have not regarded Belarusian elections as free and fair, except for his
initial win. Although a lot of protests against Lukashenko, he claimed to win
with a pretty high votes, almost more than 80% in each election. The
post-dissolution leaders of these countries are all high officials or heads of
the former Soviet republics and continued their leadership in presidency.

## Determinants of Autocoup Attempts: Case Studies

### High Frequency and Success Rate of Autocoups in Post-Communist Countries

From the dataset, we observe that in post-communist countries, both the frequency
and success rate of autocoups are notably high. Post-communist countries refer to
those that were communist at the end of the Cold War, regardless of whether they
have since transitioned to other types of regimes. In these countries, there are
12 documented cases[^2] of autocoups aimed at prolonging incumbency, with only 2
of these attempts failing.

Examining the cases in post-communist countries, several characteristics stand
out:

-   **Inherited Authoritarian Systems**: Although most of these countries
    transitioned from communist regimes to non-communist governments (with the
    exception of China), they inherited the authoritarian systems of their
    communist past.
-   **Continuity of Former Elites**: The transitions did not result in the
    removal or overthrow of the previous ruling groups. Instead, the former
    communist elites remained in power.
-   **Subverted Democratic Processes**: Despite the introduction of general
    elections and term limits in most of these countries, the legacy of the
    former communist regimes often led to term limits being ignored and elections
    being rigged.

For example, Alexander Lukashenko was a member of the Supreme Soviet of the
Byelorussian Soviet Socialist Republic before the dissolution of the Soviet
Union. After the dissolution, he became head of the interim anti-corruption
committee of the Supreme Council of Belarus. Elected as the first president of
Belarus in 1994, he has held the office ever since. Initially, the 1994
constitutions set a maximum of two successive presidential terms, but Lukashenko
removed this limit in 2004. Furthermore, international monitors have not regarded
Belarusian elections as free and fair, except for his initial win. Despite
significant protests against him, Lukashenko claimed to win with a high vote
share, often exceeding 80% in each election. This pattern is evident in all five
Central Asian countries of the former Soviet Union. Similarly, the
post-dissolution leaders of these countries were high officials or heads of the
former Soviet republics who continued their leadership in the presidency.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```

[^1]: The 12 cases involve 11 countries, with Azerbaijan experiencing two
    autocoup attempts.

[^2]: The 12 cases involve 11 countries, with Azerbaijan experiencing two
    autocoup attempts.
