---
title: "Overstay Coups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-14
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: autocoup.html

categories: [Article, Coups, Autocoups]
bibliography: references.bib

abstract: |
  This study aims to clarify the concept of overstay coups, focusing on power extensions by incumbent leaders. By distinguishing overstay coups from the broader, more ambiguous concepts of self-coups or autocoups, which encompass both executive power aggrandizement and power extension, this research introduces a refined understanding of these specific political events.

  The research involves coding overstay coup events and compiling a comprehensive dataset covering the period from 1945 to the present. Using this dataset, two detailed case studies provide qualitative insights into the dynamics and outcomes of overstay coups. Additionally, a regression analysis identifies the determinants of overstay coup attempts, shedding light on the factors that influence incumbents' decisions to extend their power illegitimately.

  The findings contribute to the existing literature by providing a clearer conceptual framework and empirical evidence on overstay coups. This research enhances our understanding of the mechanisms and motivations behind power extensions by incumbent leaders, and it examines the implications for democratic resilience, democratic backsliding, democratic breakdown, personalization of power, and autocratic deterioration. The insights gained from this study inform policy and theoretical discussions on political stability and the challenges facing contemporary democracies.
  
keywords: [Coups, Autocoups, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

## Introduction

Understanding how political leaders come to power and leave office is crucial in
political science. Traditionally, research has focused on two main exit routes:
leaders who are forced out before their term ends (e.g., coups) or those who
complete their terms and leave voluntarily. However, a growing trend of leaders
attempting to extend their term limits through illegitimate means has been
understudied. This research examines these irregular stays in power, specifically
focusing on the concept of overstay coups, as defined in this study.

Coups, once a common means of leadership change, have become less frequent in
recent decades [@bermeo2016; @thyne2019]. In contrast, irregular stays in
power—often referred to by various terms like self-coups, autocoups/autogolpes,
incumbent takeovers, or continuismo—are on the rise [@ginsburg2010evasion;
@baturo2014; @versteeg2020law]. Leaders like Vladimir Putin and Xi Jinping
exemplify this trend, manipulating legal frameworks to extend their rule. This
phenomenon is not limited to non-democratic regimes; even well-established
democracies have witnessed similar attempts [@antonio2021; @pion-berlin2022].

Overstay coups pose a significant threat to democratic stability and the trend of
democratization in autocratic countries. While the negative impacts of coups are
well-documented [@thyne2019], the detrimental effects of overstay coups are no
less severe. Firstly, overstay coups are typically accompanied by a disregard for
the rule of law, weakened institutions, and potential democratic backsliding or
authoritarian personalization. Secondly, like traditional coups, successful
overstay coups increase the risk of irregular power transitions in the future.
Research indicates that since 1945, approximately 62 percent of leaders who
overstayed in non-democratic countries were either ousted or assassinated while
in office [@baturo2019]. Thirdly, failed overstay coups often trigger instability
that fuels protests, violence, and even civil wars.

Despite their prevalence and negative consequences, overstay coups have received
significantly less research attention compared to coups. Furthermore, there are
two issues in previous studies that need to be addressed or clarified. Firstly,
as irregular events or actions perpetrated by incumbent leaders become more
common post-Cold War, several relevant concepts or terminologies such as
self-coups, autocoups, autogolpes, incumbent takeovers, executive aggrandizement,
overstay, and continuismo have become rich and popular topics of discussion
[@marsteintredet2019; @baturo2022]. However, there is no clear and widely
accepted term or terms to describe these political situations or events.
Additionally, these terms are used by different scholars in a puzzling context.
Sometimes, several different terms are used to describe the same, or at least
similar, situations or events. For example, coups, autocoups, autogolpes, and
executive aggrandizement are used to describe power expansions by executive
leaders in different literatures, while overstay and continuismo are used to
describe power extensions by incumbent leaders. Sometimes the same term is used
to describe two different situations without clear distinction. For instance,
self-coups, autocoups, and autogolpes are normally used to refer to "when a
president closes the courts and the legislature, suspends the constitution, and
rules by decree until a referendum and new legislative elections are held to
approve broader executive powers" [@cameron1998], but power extensions are also
included in their discussion although they should not be by their definitions.
Secondly, due to the lack of datasets compared to the rich datasets of coups, the
limited existing studies on irregular overstays have primarily relied on case
studies [@cameron1998; @antonio2021; @pion-berlin2022], with few focusing on
quantitative analyses.

In order to fully understand irregular political leadership transition and
survival, it is crucial to clarify these terms in the first place, distinguish
power expansion and power extension, and then introduce a relevant dataset for
further empirical analysis.

This study aims to address the knowledge gap surrounding overstay coups and
contribute in three areas. First, it clarifies the terminology by defining
overstay coups as a specific type of irregular stay characterized by power
extension. Second, it introduces a new dataset of overstay coups since 1945 based
on this refined definition. Finally, the research utilizes this dataset for a
quantitative analysis of the factors that influence leaders' decisions to attempt
overstay coups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of overstay
coups. This will be followed by an introduction to the dataset of overstay coups.
In Parts IV and V, I will explain the determinants of overstay coup attempts
through two case studies and demonstrate how the novel overstay coup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude in Part VI, summarizing the key insights gained from the study and
suggesting avenues for future research.

## Overstay coup: Relevant terms and a new definition

### Relevant terms on power expansions and power extensions

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
