---
title: "Autocoups: Conceptual Clarification and Analysis of Power Extensions by Incumbent Leaders"
author:
  - name: Zhu Qi
    affiliations: University of Essex
    email: qz21485@essex.ac.uk

date: 2024-06-14
date-modified: last-modified
format:
  aps-html: default
  aps-pdf: default
  docx: default
  aps-revealjs:
    output-file: autocoup.html

categories: [Article, Coups, Autocoups]
bibliography: references.bib

abstract: |
  This study aims to clarify the concept of overstay coups, focusing on power extensions by incumbent leaders. By distinguishing overstay coups from the broader, more ambiguous concepts of self-coups or autocoups, which encompass both executive power aggrandizement and power extension, this research introduces a refined understanding of these specific political events.

  The research involves coding overstay coup events and compiling a comprehensive dataset covering the period from 1945 to the present. Using this dataset, two detailed case studies provide qualitative insights into the dynamics and outcomes of overstay coups. Additionally, a regression analysis identifies the determinants of overstay coup attempts, shedding light on the factors that influence incumbents' decisions to extend their power illegitimately.

  The findings contribute to the existing literature by providing a clearer conceptual framework and empirical evidence on overstay coups. This research enhances our understanding of the mechanisms and motivations behind power extensions by incumbent leaders, and it examines the implications for democratic resilience, democratic backsliding, democratic breakdown, personalization of power, and autocratic deterioration. The insights gained from this study inform policy and theoretical discussions on political stability and the challenges facing contemporary democracies.
  
keywords: [Coups, Autocoups, Political Leadership]

execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(zoo)
library(democracyData)
library(vdemdata)
library(sampleSelection)
library(stargazer)
library(readxl)
library(here)
```

## Introduction

The study of irregular power transitions, encompassing both irregular entries and
exits from office, has long been a central topic in political science. However,
one form of irregular power transition - the incumbent leader's refusal to
relinquish power - remains relatively understudied. This scenario, where leaders
overstay their mandated term limits for one or more terms, or even indefinitely,
lacks a universally accepted term in the existing literature. To address this
gap, this study adopts the term "autocoup" to denote this phenomenon.

Coups, being the primary form of irregular power transitions, have received
significant scholarly attention. Research by @goemans2009 indicate that coups
account for a substantial portion of such transitions, with roughly two-thirds
(65.8%) of 374 irregular exits occurring through coups. Additionally, @frantz2016
demonstrate that coups are the most common form of exit in autocracies,
surpassing regular transitions by a third.

However, while coups were once frequent, recent decades have witnessed a decline
in their occurrence [@bermeo2016; @thyne2019]. Conversely, autocoups appear to be
on the rise, particularly since the Cold War's end [@ginsburg2010evasion;
@baturo2014; @versteeg2020law]. Despite the significant political impact and
rising prevalence of autocoups, their study lags behind that of classical coups,
which are extensively documented due to their historical frequency [@thyne2019].

This research aims to bridge this gap by focusing on the understudied phenomenon
of autocoups as they are at least equally important and warrant comprehensive
analysis. Firstly, autocoups are typically accompanied by a disregard for the
rule of law, weakened institutions, and potential democratic backsliding or
authoritarian personalization. Secondly, like traditional coups, successful
autocoups increase the risk of future irregular power transitions. Research
indicates that since 1945, approximately 62 percent of leaders who overstayed
their term limits through autocoups in non-democratic countries were either
ousted or assassinated while in office [@baturo2019]. Thirdly, failed autocoups
often trigger instability that fuels protests, violence, and even civil wars.

Despite their significance, autocoups are understudied compared to traditional
coups. Although several relevant concepts and terminologies such as self-coups,
autocoups, autogolpes, incumbent takeovers, executive aggrandizement, overstay,
and continuismo have been proposed and discussed [@marsteintredet2019;
@baturo2022], there is no clear and widely accepted term to describe these
events. Different terms are often used interchangeably or inconsistently, leading
to confusion. Moreover, due to the lack of clear concepts, the collection of
autocoup datasets is still in its early stages compared to the rich datasets of
traditional coups. Hence, the limited existing studies on autocoups have
primarily relied on case studies [@cameron1998; @antonio2021; @pion-berlin2022],
with few focusing on quantitative analyses.

To fully understand irregular political leadership transitions and survival, we
need to redefine and clarify the term "autocoup" first. Based on a clear
definition, we can introduce an autocoup dataset that can be used for further
empirical analysis. This study aims to address these knowledge gap surrounding
autocoups and contribute in three areas. First, it clarifies the terminology by
defining autocoups based mainly on power extension. Second, it introduces a new
dataset of autocoups since 1945 based on this refined definition. Finally, the
research utilizes this dataset for a quantitative analysis of the factors that
influence leaders' decisions to attempt autocoups.

The subsequent section, based on a comprehensive review of the definitions of
power expansions and power extensions, proposes a precise definition of
autocoups. This will be followed by an introduction to the new autocoup dataset
in Part III. In Parts IV and V, I will explain the determinants of autocoup
attempts through two case studies and demonstrate how the novel autocoup dataset
collected in this study can be used in empirical analysis. Finally, I will
conclude in Part VI, summarizing the key insights gained from the study and
suggesting avenues for future research.

## Autocoups: A literature review and clarification of definitions

Compared to coups, which are clearly defined and widely accepted as illegal
attempts by elites within the ruling group to overthrow the leadership
[@powell2011], the concept of autocoups (also known as autogolpes and self-coups)
suffers from a lack of consistency and clarity. This ambiguity hinders our
understanding and studying of a critical phenomenon in irregular power
transitions, hence necessitates addressing two key issues: terminology and core
characteristics.

### Choosing the Right Term

The most common terms in autocoup literature are *self-coup* or *autogolpe* in
Spanish [@przeworski2000; @cameron1998a; @bermeo2016; @helmke2017;
@marsteintredet2019]. The usage gained academic prominence after Peruvian
President Alberto Fujimori dissolved Congress, temporarily suspended the
constitution, and ruled by decree in 1992 [@mauceri1995; @cameron1998]. It is
commonly defined as "a temporary suspension of the constitution and dissolution
of congress by the executive, who rules by decree until new legislative elections
and a referendum can be held to ratify a political system with broader executive
power" [@cameron1998a, p.220]. However, as @marsteintredet2019 point out, the
term "self-coup" can be misleading, as it implies a coup against oneself, which
is inaccurate since it is not self-directed but to other state institutions or
apparatus.

Another approach to describe coups staged by incumbents is to use coups with
adjectives or modifiers, such as presidential coup, executive coup,
constitutional coup, electoral coup, judicial coup, slow-motion coup, soft coup,
and parliamentary coup [@marsteintredet2019]. While these terms can be helpful in
specific power expansions or power grabs perpetrated by incumbents, the
proliferation of such terms can add to the overall confusion rather than
providing clarification. Essentially, most of these coups with adjective terms
primarily focus on the specific methods used by incumbents to expand their power
or lessen restrictions, often failing to clearly identify the incumbent leader as
the perpetrator. Even in specific contexts, using a term like "constitutional
autocoup" instead of "constitutional coup" would be more informative due to its
explicit identification of the incumbent as the actor.

The third alternative involves terms like like "incumbent takeover" or "executive
takeover," referring to "an event perpetuated by a ruling executive that
significantly reduces the formal and/or informal constraints on his/her power"
[@baturo2022, p.374] based on earlier research [@svolik2014]. This term
identifies the perpetrator (incumbent) and the nature of the event (power grab).
However, "incumbent takeover" lacks clarity regarding the illegal or illegitimate
nature of these actions. While "coup" clearly denotes the illegality of
leadership ousters, replacing it with "takeover" diminishes the severity.
Therefore, "incumbent takeover" cannot be a direct counterpart to "coup" due to
its ambiguity concerning illegality.

Based on this analysis, "autocoup" emerges as the most accurate and appropriate
term to describe situations where incumbents orchestrate actions to extend and/or
expand their power through illegitimate means.

### Key characteristics of autocoups

Beyond terminology, a clear definition of an autocoup is crucial. It is important
to distinguish autocoups from other power grabs.

Another issue with the previous definition of *self-coup* is its emphasis on
executive aggrandizement or power expansion rather than power extension. It
primarily describes situations where executive leaders seize more power from
other branches of government. While power extensions might be included in some
analyses as well, they are not taken as the determinant feature of self-coups.
When we refer to a classical coup, we clearly mean the ouster of the sitting
executive leader instead of the power shrinking from the executive leader, or the
ouster of any other leader. Due to the same logic, it is more appropriate to
define autocoup based first on the executive leadership instead of other state
apparatus, second on the holding or relinquishing the leadership instead of power
expansions of the executive role.

the power extension instead of power expansion. For instance, we would not code a
situation, say a vice-president, a primer, or a chair of the congress grabbing
more power from the president, a coup, as long as the president is still in the
office. Then, we should not code the power grabs by the president from a
vice-president, a primer, or a chair of the congress as a autocoup.

I argue that power extension should be the main feature of autocoups.

Returning to the core definition of a classic coup, which focuses on the illegal
replacement of incumbent leaders, offers greater clarity.

While coups aim to replace incumbents, autocoups aim to prevent legal successors
from assuming power. Coups are overt illegal actions, whereas autocoups might
have a veneer of legality achieved through the manipulation of state institutions
and officials.

In light of this discussion, we formally define an autocoup as the practice of
current political leaders extending their time in office beyond their originally
mandated term limits, often achieved through seemingly legal means like
constitutional amendments or referendums, but ultimately illegitimate in essence.

-   **Focus on Power Extension:** While autocoups might involve power expansion
    (taking more power from other branches), their primary characteristic is
    extending the incumbent's term in office.

-   **Illegality or Illegitimacy:** Autocoups often employ seemingly legal
    mechanisms like referendums or constitutional amendments, but their essence
    is illegitimate.

This review highlights the need for a clear understanding of autocoups. By
advocating for the term "autocoup" and establishing its core characteristics, we
can improve research and analysis of this critical phenomenon in irregular power
transitions.

This definition, or slight variations of it, Some relevant terms with similar
definition, such as "presidential coup" [@roberts1995; @marshall2019coup],
"executive coup" [@bermeo2016], and "constitutional coup" [@helmke2017], have
also been used.

```{r}
#| label: references
#| results: asis
#| eval: !expr format == "latex"
#| echo: false
cat("\\newpage
## References")
```
